### 2021-06-04 doc/manuscript/ESSDmanuscript.Rmd
start setting up article
authors, reference library, front matter...

---
title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'ESSD')`"
author:
  - given_name: Madison 
    surname: Williams
    affiliation: "1"
  - given_name: Valentine
    surname: Herrmann
    affiliation: "1"
  - given_name: Rebecca
    surname: Banbury Morgan
    affiliation: "2" 
  - given_name: Ben 
    surname: Bond-Lamberty (confirm)
    affiliation: "3" 
  - given_name: Susan 
    surname: Cook-Patton (confirm)
    affiliation: "4" 
  - given_name: Sandro 
    surname: Federici (confirm)
    affiliation: "5" 
  - given_name: Helene
    surname: Muller-Landau
    affiliation: "6" 
  - given_name: Camille
    surname: Piponiot
    affiliation: "7" 
  - given_name: Teagan
    surname: Rogers
    affiliation: "1"
  - given_name: Valentyna
    surname: Slivinska (confirm)
    affiliation: "5"
  - given_name: Kristina J.
    surname: Anderson-Teixeira
    affiliation: "1,6 *"
    email: teixeirak@si.edu
    corresponding: true

#affiliation numbering will need to be fixed

affiliation:
  - code: 1
    address: Center for Conservatiton Ecology, Smithsonian Conservation Biology Institute, Front Royal, VA, United States
  - code: 2
    address: 
  - code: 3
    address: 
  - code: 4
    address: 
  - code: 5
    address: 
  - code: 6
    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
  - code: 7
    address: CIRAD, Montpellier, France

abstract: |
  The abstract goes here.
  It can also be on _multiple lines_.
bibliography: references.bib
running:
  title: IPCC forest C accounting with ForC
  author: Williams et al.
# This section is mandatory even if you declare that no competing interests are present.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
copyrightstatement: |
  The author's copyright for this publication is transferred to institution/company. 
### The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
### It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
availability:
  #code: |
  #  use this to add a statement when having only software code available
  #data: |
  #  use this to add a statement when having only data sets available
  codedata: |
    use this to add a statement when having data sets and software code available
  sample: |
    use this section when having geoscientific samples available
videosupplement: |
  use this section when having video supplements available
authorcontribution: |
  Daniel wrote the package. Josiah thought about poterry. Markus filled in for a second author.
disclaimer: |
  We like Copernicus.
acknowledgements: |
  Thanks to the rticles contributors!
appendix: |
  \section{Figures and tables in appendices}
  Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
  \subsection{Option 1}
  If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
  They will be correctly named automatically.
  \subsection{Option 2}
  If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.

  To rename them correctly to A1, A2, etc., please add the following commands in front of them:
  `\appendixfigures` needs to be added in front of appendix figures
  `\appendixtables` needs to be added in front of appendix tables

  Please add `\clearpage` between each table and/or figure. Further guidelines on figures and tables can be found below.
output:
  rticles::copernicus_article: default
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
---

\introduction[Introduction]

Introduction text goes here.
You can change the name of the section if neccessary using `\introduction[modified heading]`.

The following settings can or must be configured in the header of this file and are bespoke for Copernicus manuscripts:

- The `journal` you are submitting to using the official abbreviation. You can use the function `rticles::copernicus_journal_abbreviations(name = '...')` to search the existing journals.

- Specific sections of the manuscript:

  - `running` with `title` and `author`

  - `competinginterests`

  - `copyrightstatement` (optional)

  - `availability` (strongly recommended if any used), one of `code`, `data`, or `codedata`

  - `authorcontribution`

  - `disclaimer`

  - `acknowledgements`

See the defaults and examples from the skeleton and the official Copernicus documentation for details.

**Important**: Always double-check with the official manuscript preparation guidelines at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html), especially the sections "Technical instructions for LaTeX" and "Manuscript composition".
Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.

# Content section one

## Subsection Heading Here

Subsection text here.

### Subsubsection Heading Here

Subsubsection text here.

# Content section with citations

See the [R Markdown docs for bibliographies and citations](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html).

Copernicus supports biblatex and a sample bibliography is in file `sample.bib`.
Example citation [@anderson-teixeira_forc_2018]

# Content section with R code chunks

```{r, echo = FALSE}
sum <- 1 + 41
```

You should always use `echo = FALSE` on R Markdown code blocks as they add formatting and styling not desired by Copernicus.
The hidden workflow results in `r sum`.

You can add verbatim code snippets without extra styles by using ` ``` ` without additional instructions.

```
sum <- 1 + 41
```

# Content section with list

If you want to insert a list, you must

- leave

- empty lines

- between each list item

because the `\tightlist` format used by R Markdown is not supported in the Copernicus template. Example:

```
- leave

- empty lines

- between each list item
```

# Examples from the official template

## FIGURES

When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage between bibliography and first table and/or figure as well as between each table and/or figure.

### ONE-COLUMN FIGURES

Include a 12cm width figure of Nikolaus Copernicus from [Wikipedia](https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg) with caption using R Markdown.

```{r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
knitr::include_graphics("Nikolaus_Kopernikus.jpg")
```

### TWO-COLUMN FIGURES

You can also include a larger figure.

```{r, out.width = "12cm", echo = FALSE, fig.cap = "two column figure"}
knitr::include_graphics("Nikolaus_Kopernikus.jpg")
```

## TABLES

You can ad \LaTeX table in an R Markdown document to meet the template requirements.

### ONE-COLUMN TABLE

\begin{table}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table Footnotes}
\end{table}

### TWO-COLUMN TABLE

\begin{table*}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table footnotes}
\end{table*}

## MATHEMATICAL EXPRESSIONS

All papers typeset by Copernicus Publications follow the math typesetting regulations given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry, 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).

Physical quantities/variables are typeset in italic font (t for time, T for Temperature)

Indices which are not defined are typeset in italic font (x, y, z, a, b, c)

Items/objects which are defined are typeset in roman font (Car A, Car B)

Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)

Abbreviations from 2 letters are typeset in roman font (RH, LAI)

Vectors are identified in bold italic font using \vec{x}

Matrices are identified in bold roman font

Multiplication signs are typeset using the LaTeX commands `\times` (for vector products, grids, and exponential notations) or `\cdot`

The character * should not be applied as mutliplication sign

## EQUATIONS

### Single-row equation

Unnumbered equations (i.e. using `$$` and getting inline preview in RStudio) are not supported by Copernicus.

\begin{equation}
1 \times 1 \cdot 1 = 42
\end{equation}

\begin{equation}
A = \pi r^2
\end{equation}

\begin{equation}
x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
\end{equation}

### Multiline equation

\begin{align}
& 3 + 5 = 8\\
& 3 + 5 = 8\\
& 3 + 5 = 8
\end{align}

## MATRICES

$$
\begin{matrix}
x & y & z\\
x & y & z\\
x & y & z\\
\end{matrix}
$$

## ALGORITHM

If you want to use algorithms, you can either enable the required packages in the header (the default, see `algorithms: true`), or make sure yourself that the \LaTeX packages `algorithms` and `algorithmicx` are installed so that `algorithm.sty` respectively `algorithmic.sty` can be loaded by the Copernicus template.
Copernicus staff will remove all undesirable packages from your LaTeX source code, so please stick to using the header option, which only adds the two acceptable packages.

\begin{algorithm}
\caption{Algorithm Caption}
\label{a1}
\begin{algorithmic}
\STATE $i\gets 10$
\IF {$i\geq 5$} 
        \STATE $i\gets i-1$
\ELSE
        \IF {$i\leq 3$}
                \STATE $i\gets i+2$
        \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

## CHEMICAL FORMULAS AND REACTIONS

For formulas embedded in the text, please use `\chem{}`, e.g. \chem{A \rightarrow B}.

The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.

- `\rightarrow` should be used for normal (one-way) chemical reactions

- `\rightleftharpoons` should be used for equilibria

- `\leftrightarrow` should be used for resonance structures

\begin{reaction}
A \rightarrow B \\
\end{reaction}
\begin{reaction}
Coper \rightleftharpoons nicus \\
\end{reaction}
\begin{reaction}
Publi \leftrightarrow cations
\end{reaction}

## PHYSICAL UNITS

Please use `\unit{}` (allows to save the math/`$` environment) and apply the exponential notation, for example \( 3.14\,\unit{km\,h^{-1}} \) (using LaTeX mode: `\( 3.14\,\unit{...} \)`) or \unit{0.872\,m\,s^{-1}} (using only `\unit{0.872\,m\,s^{-1}}`).

\conclusions[Conclusions]

The conclusion goes here.
You can modify the section name with  `\conclusions[modified heading if necessary]`.

### 2021-06-04 doc/manuscript/ESSDmanuscript.tex
start setting up article
authors, reference library, front matter...

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% copernicus_rticles_template (flag for rticles template detection - do not remove!)

%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.

%% 2-column papers and discussion papers
\documentclass[, manuscript]{copernicus}

%% Journal abbreviations (please use the same for discussion papers and final revised papers)

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% ASTRA Proceedings (ap)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Bone and Joint Infection (jbji)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Polarforschung - Journal of the German Society for Polar Research (polf)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

% Pandoc citation processing

% The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
%
\usepackage{algorithmic}
\usepackage{algorithm}

\begin{document}

\title{Informing IPCC accounting of forest carbon using the global
forest carbon database (ForC v4.0)}

\Author[1]{Madison}{Williams}
\Author[1]{Valentine}{Herrmann}
\Author[2]{Rebecca}{Banbury Morgan}
\Author[3]{Ben}{Bond-Lamberty (confirm)}
\Author[4]{Susan}{Cook-Patton (confirm)}
\Author[5]{Sandro}{Federici (confirm)}
\Author[6]{Helene}{Muller-Landau}
\Author[7]{Camille}{Piponiot}
\Author[1]{Teagan}{Rogers}
\Author[5]{Valentyna}{Slivinska (confirm)}
\Author[1,6 *]{Kristina J.}{Anderson-Teixeira}

\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation
Biology Institute, Front Royal, VA, United States}
\affil[2]{}
\affil[3]{}
\affil[4]{}
\affil[5]{}
\affil[6]{Forest Global Earth Observatory, Smithsonian Tropical Research
Institute, Panama, Republic of Panama}
\affil[7]{CIRAD, Montpellier, France}

\runningtitle{IPCC forest C accounting with ForC}

\runningauthor{Williams et al.}

\correspondence{Kristina J.\ Anderson-Teixeira\ (teixeirak@si.edu)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
The abstract goes here. It can also be on \emph{multiple lines}.
\end{abstract}

\copyrightstatement{The author's copyright for this publication is
transferred to institution/company.}

\introduction[Introduction]

Introduction text goes here. You can change the name of the section if
neccessary using
\texttt{\textbackslash{}introduction{[}modified\ heading{]}}.

The following settings can or must be configured in the header of this
file and are bespoke for Copernicus manuscripts:

\begin{itemize}
\item
  The \texttt{journal} you are submitting to using the official
  abbreviation. You can use the function
  \texttt{rticles::copernicus\_journal\_abbreviations(name\ =\ \textquotesingle{}...\textquotesingle{})}
  to search the existing journals.
\item
  Specific sections of the manuscript:

  \begin{itemize}
  \item
    \texttt{running} with \texttt{title} and \texttt{author}
  \item
    \texttt{competinginterests}
  \item
    \texttt{copyrightstatement} (optional)
  \item
    \texttt{availability} (strongly recommended if any used), one of
    \texttt{code}, \texttt{data}, or \texttt{codedata}
  \item
    \texttt{authorcontribution}
  \item
    \texttt{disclaimer}
  \item
    \texttt{acknowledgements}
  \end{itemize}
\end{itemize}

See the defaults and examples from the skeleton and the official
Copernicus documentation for details.

\textbf{Important}: Always double-check with the official manuscript
preparation guidelines at
\url{https://publications.copernicus.org/for_authors/manuscript_preparation.html},
especially the sections ``Technical instructions for LaTeX'' and
``Manuscript composition''. Please contact Daniel Nüst,
\texttt{daniel.nuest@uni-muenster.de}, with any problems.

\section{Content section one}

\subsection{Subsection Heading Here}

Subsection text here.

\subsubsection{Subsubsection Heading Here}

Subsubsection text here.

\section{Content section with citations}

See the
\href{http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html}{R
Markdown docs for bibliographies and citations}.

Copernicus supports biblatex and a sample bibliography is in file
\texttt{sample.bib}. Example citation
\citep{anderson-teixeira_forc_2018}

\section{Content section with R code chunks}

You should always use \texttt{echo\ =\ FALSE} on R Markdown code blocks
as they add formatting and styling not desired by Copernicus. The hidden
workflow results in 42.

You can add verbatim code snippets without extra styles by using
\texttt{\textasciigrave{}\textasciigrave{}\textasciigrave{}} without
additional instructions.

\begin{verbatim}
sum <- 1 + 41
\end{verbatim}

\section{Content section with list}

If you want to insert a list, you must

\begin{itemize}
\item
  leave
\item
  empty lines
\item
  between each list item
\end{itemize}

because the \texttt{\textbackslash{}tightlist} format used by R Markdown
is not supported in the Copernicus template. Example:

\begin{verbatim}
- leave

- empty lines

- between each list item
\end{verbatim}

\section{Examples from the official template}

\subsection{FIGURES}

When figures and tables are placed at the end of the MS (article in
one-column style), please add \clearpage between bibliography and first
table and/or figure as well as between each table and/or figure.

\subsubsection{ONE-COLUMN FIGURES}

Include a 12cm width figure of Nikolaus Copernicus from
\href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia}
with caption using R Markdown.

\begin{figure}
\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:unnamed-chunk-2}
\end{figure}

\subsubsection{TWO-COLUMN FIGURES}

You can also include a larger figure.

\begin{figure}
\includegraphics[width=12cm]{Nikolaus_Kopernikus} \caption{two column figure}\label{fig:unnamed-chunk-3}
\end{figure}

\subsection{TABLES}

You can ad \LaTeX table in an R Markdown document to meet the template
requirements.

\subsubsection{ONE-COLUMN TABLE}

\begin{table}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table Footnotes}
\end{table}

\subsubsection{TWO-COLUMN TABLE}

\begin{table*}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table footnotes}
\end{table*}

\subsection{MATHEMATICAL EXPRESSIONS}

All papers typeset by Copernicus Publications follow the math
typesetting regulations given by the IUPAC Green Book (IUPAC:
Quantities, Units and Symbols in Physical Chemistry, 2nd Edn., Blackwell
Science, available at:
http://old.iupac.org/publications/books/gbook/green\_book\_2ed.pdf,
1993).

Physical quantities/variables are typeset in italic font (t for time, T
for Temperature)

Indices which are not defined are typeset in italic font (x, y, z, a, b,
c)

Items/objects which are defined are typeset in roman font (Car A, Car B)

Descriptions/specifications which are defined by itself are typeset in
roman font (abs, rel, ref, tot, net, ice)

Abbreviations from 2 letters are typeset in roman font (RH, LAI)

Vectors are identified in bold italic font using \vec{x}

Matrices are identified in bold roman font

Multiplication signs are typeset using the LaTeX commands
\texttt{\textbackslash{}times} (for vector products, grids, and
exponential notations) or \texttt{\textbackslash{}cdot}

The character * should not be applied as mutliplication sign

\subsection{EQUATIONS}

\subsubsection{Single-row equation}

Unnumbered equations (i.e.~using \texttt{\$\$} and getting inline
preview in RStudio) are not supported by Copernicus.

\begin{equation}
1 \times 1 \cdot 1 = 42
\end{equation}

\begin{equation}
A = \pi r^2
\end{equation}

\begin{equation}
x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
\end{equation}

\subsubsection{Multiline equation}

\begin{align}
& 3 + 5 = 8\\
& 3 + 5 = 8\\
& 3 + 5 = 8
\end{align}

\subsection{MATRICES}

\[
\begin{matrix}
x & y & z\\
x & y & z\\
x & y & z\\
\end{matrix}
\]

\subsection{ALGORITHM}

If you want to use algorithms, you can either enable the required
packages in the header (the default, see \texttt{algorithms:\ true}), or
make sure yourself that the \LaTeX packages \texttt{algorithms} and
\texttt{algorithmicx} are installed so that \texttt{algorithm.sty}
respectively \texttt{algorithmic.sty} can be loaded by the Copernicus
template. Copernicus staff will remove all undesirable packages from
your LaTeX source code, so please stick to using the header option,
which only adds the two acceptable packages.

\begin{algorithm}
\caption{Algorithm Caption}
\label{a1}
\begin{algorithmic}
\STATE $i\gets 10$
\IF {$i\geq 5$} 
        \STATE $i\gets i-1$
\ELSE
        \IF {$i\leq 3$}
                \STATE $i\gets i+2$
        \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

\subsection{CHEMICAL FORMULAS AND REACTIONS}

For formulas embedded in the text, please use
\texttt{\textbackslash{}chem\{\}}, e.g.~\chem{A \rightarrow B}.

The reaction environment creates labels including the letter R,
i.e.~(R1), (R2), etc.

\begin{itemize}
\item
  \texttt{\textbackslash{}rightarrow} should be used for normal
  (one-way) chemical reactions
\item
  \texttt{\textbackslash{}rightleftharpoons} should be used for
  equilibria
\item
  \texttt{\textbackslash{}leftrightarrow} should be used for resonance
  structures
\end{itemize}

\begin{reaction}
A \rightarrow B \\
\end{reaction}
\begin{reaction}
Coper \rightleftharpoons nicus \\
\end{reaction}
\begin{reaction}
Publi \leftrightarrow cations
\end{reaction}

\subsection{PHYSICAL UNITS}

Please use \texttt{\textbackslash{}unit\{\}} (allows to save the
math/\texttt{\$} environment) and apply the exponential notation, for
example \(3.14\,\unit{km\,h^{-1}}\) (using LaTeX mode:
\texttt{\textbackslash{}(\ 3.14\textbackslash{},\textbackslash{}unit\{...\}\ \textbackslash{})})
or \unit{0.872\,m\,s^{-1}} (using only
\texttt{\textbackslash{}unit\{0.872\textbackslash{},m\textbackslash{},s\^{}\{-1\}\}}).

\conclusions[Conclusions]

The conclusion goes here. You can modify the section name with
\texttt{\textbackslash{}conclusions{[}modified\ heading\ if\ necessary{]}}.

\codedataavailability{use this to add a statement when having data sets
and software code
available} %% use this section when having data sets and software code available

\sampleavailability{use this section when having geoscientific samples
available} %% use this section when having geoscientific samples available

\videosupplement{use this section when having video supplements
available} %% use this section when having geoscientific samples available

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Figures and tables in appendices}

Regarding figures and tables in appendices, the following two options
are possible depending on your general handling of figures and tables in
the manuscript environment:

\subsection{Option 1}

If you sorted all figures and tables into the sections of the text,
please also sort the appendix figures and appendix tables into the
respective appendix sections. They will be correctly named
automatically.

\subsection{Option 2}

If you put all figures after the reference list, please insert appendix
tables and figures after the normal tables and figures.

To rename them correctly to A1, A2, etc., please add the following
commands in front of them: \texttt{\textbackslash{}appendixfigures}
needs to be added in front of appendix figures
\texttt{\textbackslash{}appendixtables} needs to be added in front of
appendix tables

Please add \texttt{\textbackslash{}clearpage} between each table and/or
figure. Further guidelines on figures and tables can be found below.
\noappendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontribution{Daniel wrote the package. Josiah thought about
poterry. Markus filled in for a second author.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\competinginterests{The authors declare no competing
interests.} %% this section is mandatory even if you declare that no competing interests are present

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\disclaimer{We like Copernicus.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
Thanks to the rticles contributors!
\end{acknowledgements}

%% REFERENCES
%% DN: pre-configured to BibTeX for rticles

%% The reference list is compiled as follows:
%%
%% \begin{thebibliography}{}
%%
%% \bibitem[AUTHOR(YEAR)]{LABEL1}
%% REFERENCE 1
%%
%% \bibitem[AUTHOR(YEAR)]{LABEL2}
%% REFERENCE 2
%%
%% \end{thebibliography}

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{references.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

%% LITERATURE CITATIONS
%%
%% command                        & example result
%% \citet{jones90}|               & Jones et al. (1990)
%% \citep{jones90}|               & (Jones et al., 1990)
%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
%% \citeauthor{jones90}|          & Jones et al.
%% \citeyear{jones90}|            & 1990

\end{document}

### 2021-06-04 doc/manuscript/ESSDmanuscript.Rmd
add some content
(transferring over from notes)

@@
 
-\introduction[Introduction]
-
-Introduction text goes here.
-You can change the name of the section if neccessary using `\introduction[modified heading]`.
-
 The following settings can or must be configured in the header of this file and are bespoke for Copernicus manuscripts:
@@
 
-# Content section one
+\introduction[Introduction]
 
-## Subsection Heading Here
+*(Importance of forests for climate change mitigation)*
 
-Subsection text here.
+*(Need for good data in international carbon accounting)*
 
-### Subsubsection Heading Here
+*(Introduce EFDb & forc)*
 
-Subsubsection text here.
+Here, we:
+(1) clarify C cycle terminology
+(2) describe mapping of ForC to IPCC's EFDB,
+(3) describe updates to ForC  (ForC v4.0)
+(4) summarize the data in ForC that's relevant to EFDB, identifying gaps
+(5) provide recommendations for improving data collection, analysis, database, and accounting
+
+# Defining carbon stocks and incremenets
+
+For quantifying forest role in global C cycle, we ultimately care about:
+(1) C stocks --stores of C that would be released to the atmosphere upon and use change
+(2) C increments -- changes in those C stocks.
+
+## Carbon stocks
+
+Forest ecosystem C stocks may be parsed into pools in various ways. 
+IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table 1).
+Quantifying these requires a one-time measurement.
+
+\newpage 
+**Table 1: variables with definitions and measurement methods.** Definitions from IPCC Table 1.1.
+(See Table 1.1 in IPCC guidance).
+(*Currently adding this as a figure (generated from original draft) because kableExtra doesn't seem to work in this template.
+Table that we want here is "figures_tables/C_pools.csv"*)
+
+```{r, out.width = "12cm", echo = FALSE, fig.cap = "Table 1"}
+knitr::include_graphics("figures_tables/C_pools_fig.png")
+```
+
+## Carbon increments
+
+C increments are defined as the change over time, in annual increments, in each C pool.
+These may be estimated as the difference between C stocks at two time points, or as the difference between inputs and outputs to the pool (i.e., fluxes).
+Quantifying these requires at least two measurements.
+
+Fluxes are the inputs and outputs to each pool.
+
+**Figure: schematic illustrating fluxes in and out of each pool**
 

### 2021-06-04 doc/manuscript/ESSDmanuscript.tex
add some content
(transferring over from notes)

@@
 
-\introduction[Introduction]
-
-Introduction text goes here. You can change the name of the section if
-neccessary using
-\texttt{\textbackslash{}introduction{[}modified\ heading{]}}.
-
 The following settings can or must be configured in the header of this
@@
 
-\section{Content section one}
+\introduction[Introduction]
 
-\subsection{Subsection Heading Here}
+\emph{(Importance of forests for climate change mitigation)}
 
-Subsection text here.
+\emph{(Need for good data in international carbon accounting)}
 
-\subsubsection{Subsubsection Heading Here}
+\emph{(Introduce EFDb \& forc)}
 
-Subsubsection text here.
+Here, we: (1) clarify C cycle terminology (2) describe mapping of ForC
+to IPCC's EFDB, (3) describe updates to ForC (ForC v4.0) (4) summarize
+the data in ForC that's relevant to EFDB, identifying gaps (5) provide
+recommendations for improving data collection, analysis, database, and
+accounting
+
+\section{Defining carbon stocks and incremenets}
+
+For quantifying forest role in global C cycle, we ultimately care about:
+(1) C stocks --stores of C that would be released to the atmosphere upon
+and use change (2) C increments -- changes in those C stocks.
+
+\subsection{Carbon stocks}
+
+Forest ecosystem C stocks may be parsed into pools in various ways. IPCC
+parses into biomass (aboveground and belowground), dead organic matter
+(dead wood and litter), and soil organic matter (Table 1). Quantifying
+these requires a one-time measurement.
+
+\newpage
+
+\textbf{Table 1: variables with definitions and measurement methods.}
+Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).
+(\emph{Currently adding this as a figure (generated from original draft)
+because kableExtra doesn't seem to work in this template. Table that we
+want here is ``figures\_tables/C\_pools.csv''})
+
+\begin{figure}
+\includegraphics[width=12cm]{figures_tables/C_pools_fig} \caption{Table 1}\label{fig:unnamed-chunk-1}
+\end{figure}
+
+\subsection{Carbon increments}
+
+C increments are defined as the change over time, in annual increments,
+in each C pool. These may be estimated as the difference between C
+stocks at two time points, or as the difference between inputs and
+outputs to the pool (i.e., fluxes). Quantifying these requires at least
+two measurements.
+
+Fluxes are the inputs and outputs to each pool.
+
+\textbf{Figure: schematic illustrating fluxes in and out of each pool}
 
@@
 \begin{figure}
-\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:unnamed-chunk-2}
+\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:unnamed-chunk-3}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=12cm]{Nikolaus_Kopernikus} \caption{two column figure}\label{fig:unnamed-chunk-3}
+\includegraphics[width=12cm]{Nikolaus_Kopernikus} \caption{two column figure}\label{fig:unnamed-chunk-4}
 \end{figure}

### 2021-06-04 doc/manuscript/ESSDmanuscript.Rmd
trying (failing) to make table formatting work

@@
 \newpage
+
 **Table 1: variables with definitions and measurement methods.** Definitions from IPCC Table 1.1.
 (See Table 1.1 in IPCC guidance).
-(*Currently adding this as a figure (generated from original draft) because kableExtra doesn't seem to work in this template.
+(*Currently adding this as a figure (generated from original draft) because kableExtra doesn't seem to work in this template, and I can't quickly get the template format to work.
 Table that we want here is "figures_tables/C_pools.csv"*)
@@
 ```{r, out.width = "12cm", echo = FALSE, fig.cap = "Table 1"}
+#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
 knitr::include_graphics("figures_tables/C_pools_fig.png")
 ```
+
+\begin{table*}[t]
+\caption{This is a start at table 1 using the template format. }
+\begin{tabular}{l c c c r}
+\tophline
+
+pool & subpool & definition & major sources of estimate variation & IPCC guidance \\
+\middlehline
+biomass & aboveground & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory  \\
+
+\bottomhline
+\end{tabular}
+\belowtable{I don't know how to adjust so that it doesn't run off the page.}
+\end{table*}
 

### 2021-06-04 doc/manuscript/ESSDmanuscript.tex
trying (failing) to make table formatting work

@@
 (\emph{Currently adding this as a figure (generated from original draft)
-because kableExtra doesn't seem to work in this template. Table that we
-want here is ``figures\_tables/C\_pools.csv''})
+because kableExtra doesn't seem to work in this template, and I can't
+quickly get the template format to work. Table that we want here is
+``figures\_tables/C\_pools.csv''})
 
@@
 \end{figure}
+
+\begin{table*}[t]
+\caption{This is a start at table 1 using the template format. }
+\begin{tabular}{l c c c r}
+\tophline
+
+pool & subpool & definition & major sources of estimate variation & IPCC guidance \\
+\middlehline
+biomass & aboveground & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory  \\
+
+\bottomhline
+\end{tabular}
+\belowtable{I don't know how to adjust so that it doesn't run off the page.}
+\end{table*}
 

### 2021-06-04 doc/manuscript/ESSDmanuscript.Rmd
delete extra content

@@
 
+Example citation [@anderson-teixeira_forc_2018]
+
 Here, we:
@@
 
-# Content section with citations
-
-See the [R Markdown docs for bibliographies and citations](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html).
-
-Copernicus supports biblatex and a sample bibliography is in file `sample.bib`.
-Example citation [@anderson-teixeira_forc_2018]
-
-# Content section with R code chunks
-
-```{r, echo = FALSE}
-sum <- 1 + 41
-```
-
-You should always use `echo = FALSE` on R Markdown code blocks as they add formatting and styling not desired by Copernicus.
-The hidden workflow results in `r sum`.
-
-You can add verbatim code snippets without extra styles by using ` ``` ` without additional instructions.
-
-```
-sum <- 1 + 41
-```
-
-# Content section with list
-
-If you want to insert a list, you must
-
-- leave
-
-- empty lines
-
-- between each list item
-
-because the `\tightlist` format used by R Markdown is not supported in the Copernicus template. Example:
-
-```
-- leave
-
-- empty lines
-
-- between each list item
-```
-
-# Examples from the official template
-
-## FIGURES
-
-When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage between bibliography and first table and/or figure as well as between each table and/or figure.
-
-### ONE-COLUMN FIGURES
-
-Include a 12cm width figure of Nikolaus Copernicus from [Wikipedia](https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg) with caption using R Markdown.
-
-```{r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
-knitr::include_graphics("Nikolaus_Kopernikus.jpg")
-```
-
-### TWO-COLUMN FIGURES
-
-You can also include a larger figure.
-
-```{r, out.width = "12cm", echo = FALSE, fig.cap = "two column figure"}
-knitr::include_graphics("Nikolaus_Kopernikus.jpg")
-```
-
-## TABLES
-
-You can ad \LaTeX table in an R Markdown document to meet the template requirements.
-
-### ONE-COLUMN TABLE
-
-\begin{table}[t]
-\caption{TEXT}
-\begin{tabular}{l c r}
-\tophline
-
-a & b & c \\
-\middlehline
-1 & 2 & 3 \\
-
-\bottomhline
-\end{tabular}
-\belowtable{Table Footnotes}
-\end{table}
-
-### TWO-COLUMN TABLE
-
-\begin{table*}[t]
-\caption{TEXT}
-\begin{tabular}{l c r}
-\tophline
-
-a & b & c \\
-\middlehline
-1 & 2 & 3 \\
-
-\bottomhline
-\end{tabular}
-\belowtable{Table footnotes}
-\end{table*}
-
-## MATHEMATICAL EXPRESSIONS
-
-All papers typeset by Copernicus Publications follow the math typesetting regulations given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry, 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
-
-Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
-
-Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
-
-Items/objects which are defined are typeset in roman font (Car A, Car B)
-
-Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
-
-Abbreviations from 2 letters are typeset in roman font (RH, LAI)
-
-Vectors are identified in bold italic font using \vec{x}
-
-Matrices are identified in bold roman font
-
-Multiplication signs are typeset using the LaTeX commands `\times` (for vector products, grids, and exponential notations) or `\cdot`
-
-The character * should not be applied as mutliplication sign
-
-## EQUATIONS
-
-### Single-row equation
-
-Unnumbered equations (i.e. using `$$` and getting inline preview in RStudio) are not supported by Copernicus.
-
-\begin{equation}
-1 \times 1 \cdot 1 = 42
-\end{equation}
-
-\begin{equation}
-A = \pi r^2
-\end{equation}
-
-\begin{equation}
-x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
-\end{equation}
-
-### Multiline equation
-
-\begin{align}
-& 3 + 5 = 8\\
-& 3 + 5 = 8\\
-& 3 + 5 = 8
-\end{align}
-
-## MATRICES
-
-$$
-\begin{matrix}
-x & y & z\\
-x & y & z\\
-x & y & z\\
-\end{matrix}
-$$
-
-## ALGORITHM
-
-If you want to use algorithms, you can either enable the required packages in the header (the default, see `algorithms: true`), or make sure yourself that the \LaTeX packages `algorithms` and `algorithmicx` are installed so that `algorithm.sty` respectively `algorithmic.sty` can be loaded by the Copernicus template.
-Copernicus staff will remove all undesirable packages from your LaTeX source code, so please stick to using the header option, which only adds the two acceptable packages.
-
-\begin{algorithm}
-\caption{Algorithm Caption}
-\label{a1}
-\begin{algorithmic}
-\STATE $i\gets 10$
-\IF {$i\geq 5$} 
-        \STATE $i\gets i-1$
-\ELSE
-        \IF {$i\leq 3$}
-                \STATE $i\gets i+2$
-        \ENDIF
-\ENDIF
-\end{algorithmic}
-\end{algorithm}
-
-## CHEMICAL FORMULAS AND REACTIONS
-
-For formulas embedded in the text, please use `\chem{}`, e.g. \chem{A \rightarrow B}.
-
-The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.
-
-- `\rightarrow` should be used for normal (one-way) chemical reactions
-
-- `\rightleftharpoons` should be used for equilibria
-
-- `\leftrightarrow` should be used for resonance structures
-
-\begin{reaction}
-A \rightarrow B \\
-\end{reaction}
-\begin{reaction}
-Coper \rightleftharpoons nicus \\
-\end{reaction}
-\begin{reaction}
-Publi \leftrightarrow cations
-\end{reaction}
-
-## PHYSICAL UNITS
-
-Please use `\unit{}` (allows to save the math/`$` environment) and apply the exponential notation, for example \( 3.14\,\unit{km\,h^{-1}} \) (using LaTeX mode: `\( 3.14\,\unit{...} \)`) or \unit{0.872\,m\,s^{-1}} (using only `\unit{0.872\,m\,s^{-1}}`).
-
 \conclusions[Conclusions]

### 2021-06-04 doc/manuscript/ESSDmanuscript.tex
delete extra content

@@
 
+Example citation \citep{anderson-teixeira_forc_2018}
+
 Here, we: (1) clarify C cycle terminology (2) describe mapping of ForC
@@
 \textbf{Figure: schematic illustrating fluxes in and out of each pool}
-
-\section{Content section with citations}
-
-See the
-\href{http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html}{R
-Markdown docs for bibliographies and citations}.
-
-Copernicus supports biblatex and a sample bibliography is in file
-\texttt{sample.bib}. Example citation
-\citep{anderson-teixeira_forc_2018}
-
-\section{Content section with R code chunks}
-
-You should always use \texttt{echo\ =\ FALSE} on R Markdown code blocks
-as they add formatting and styling not desired by Copernicus. The hidden
-workflow results in 42.
-
-You can add verbatim code snippets without extra styles by using
-\texttt{\textasciigrave{}\textasciigrave{}\textasciigrave{}} without
-additional instructions.
-
-\begin{verbatim}
-sum <- 1 + 41
-\end{verbatim}
-
-\section{Content section with list}
-
-If you want to insert a list, you must
-
-\begin{itemize}
-\item
-  leave
-\item
-  empty lines
-\item
-  between each list item
-\end{itemize}
-
-because the \texttt{\textbackslash{}tightlist} format used by R Markdown
-is not supported in the Copernicus template. Example:
-
-\begin{verbatim}
-- leave
-
-- empty lines
-
-- between each list item
-\end{verbatim}
-
-\section{Examples from the official template}
-
-\subsection{FIGURES}
-
-When figures and tables are placed at the end of the MS (article in
-one-column style), please add \clearpage between bibliography and first
-table and/or figure as well as between each table and/or figure.
-
-\subsubsection{ONE-COLUMN FIGURES}
-
-Include a 12cm width figure of Nikolaus Copernicus from
-\href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia}
-with caption using R Markdown.
-
-\begin{figure}
-\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:unnamed-chunk-3}
-\end{figure}
-
-\subsubsection{TWO-COLUMN FIGURES}
-
-You can also include a larger figure.
-
-\begin{figure}
-\includegraphics[width=12cm]{Nikolaus_Kopernikus} \caption{two column figure}\label{fig:unnamed-chunk-4}
-\end{figure}
-
-\subsection{TABLES}
-
-You can ad \LaTeX table in an R Markdown document to meet the template
-requirements.
-
-\subsubsection{ONE-COLUMN TABLE}
-
-\begin{table}[t]
-\caption{TEXT}
-\begin{tabular}{l c r}
-\tophline
-
-a & b & c \\
-\middlehline
-1 & 2 & 3 \\
-
-\bottomhline
-\end{tabular}
-\belowtable{Table Footnotes}
-\end{table}
-
-\subsubsection{TWO-COLUMN TABLE}
-
-\begin{table*}[t]
-\caption{TEXT}
-\begin{tabular}{l c r}
-\tophline
-
-a & b & c \\
-\middlehline
-1 & 2 & 3 \\
-
-\bottomhline
-\end{tabular}
-\belowtable{Table footnotes}
-\end{table*}
-
-\subsection{MATHEMATICAL EXPRESSIONS}
-
-All papers typeset by Copernicus Publications follow the math
-typesetting regulations given by the IUPAC Green Book (IUPAC:
-Quantities, Units and Symbols in Physical Chemistry, 2nd Edn., Blackwell
-Science, available at:
-http://old.iupac.org/publications/books/gbook/green\_book\_2ed.pdf,
-1993).
-
-Physical quantities/variables are typeset in italic font (t for time, T
-for Temperature)
-
-Indices which are not defined are typeset in italic font (x, y, z, a, b,
-c)
-
-Items/objects which are defined are typeset in roman font (Car A, Car B)
-
-Descriptions/specifications which are defined by itself are typeset in
-roman font (abs, rel, ref, tot, net, ice)
-
-Abbreviations from 2 letters are typeset in roman font (RH, LAI)
-
-Vectors are identified in bold italic font using \vec{x}
-
-Matrices are identified in bold roman font
-
-Multiplication signs are typeset using the LaTeX commands
-\texttt{\textbackslash{}times} (for vector products, grids, and
-exponential notations) or \texttt{\textbackslash{}cdot}
-
-The character * should not be applied as mutliplication sign
-
-\subsection{EQUATIONS}
-
-\subsubsection{Single-row equation}
-
-Unnumbered equations (i.e.~using \texttt{\$\$} and getting inline
-preview in RStudio) are not supported by Copernicus.
-
-\begin{equation}
-1 \times 1 \cdot 1 = 42
-\end{equation}
-
-\begin{equation}
-A = \pi r^2
-\end{equation}
-
-\begin{equation}
-x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
-\end{equation}
-
-\subsubsection{Multiline equation}
-
-\begin{align}
-& 3 + 5 = 8\\
-& 3 + 5 = 8\\
-& 3 + 5 = 8
-\end{align}
-
-\subsection{MATRICES}
-
-\[
-\begin{matrix}
-x & y & z\\
-x & y & z\\
-x & y & z\\
-\end{matrix}
-\]
-
-\subsection{ALGORITHM}
-
-If you want to use algorithms, you can either enable the required
-packages in the header (the default, see \texttt{algorithms:\ true}), or
-make sure yourself that the \LaTeX packages \texttt{algorithms} and
-\texttt{algorithmicx} are installed so that \texttt{algorithm.sty}
-respectively \texttt{algorithmic.sty} can be loaded by the Copernicus
-template. Copernicus staff will remove all undesirable packages from
-your LaTeX source code, so please stick to using the header option,
-which only adds the two acceptable packages.
-
-\begin{algorithm}
-\caption{Algorithm Caption}
-\label{a1}
-\begin{algorithmic}
-\STATE $i\gets 10$
-\IF {$i\geq 5$} 
-        \STATE $i\gets i-1$
-\ELSE
-        \IF {$i\leq 3$}
-                \STATE $i\gets i+2$
-        \ENDIF
-\ENDIF
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{CHEMICAL FORMULAS AND REACTIONS}
-
-For formulas embedded in the text, please use
-\texttt{\textbackslash{}chem\{\}}, e.g.~\chem{A \rightarrow B}.
-
-The reaction environment creates labels including the letter R,
-i.e.~(R1), (R2), etc.
-
-\begin{itemize}
-\item
-  \texttt{\textbackslash{}rightarrow} should be used for normal
-  (one-way) chemical reactions
-\item
-  \texttt{\textbackslash{}rightleftharpoons} should be used for
-  equilibria
-\item
-  \texttt{\textbackslash{}leftrightarrow} should be used for resonance
-  structures
-\end{itemize}
-
-\begin{reaction}
-A \rightarrow B \\
-\end{reaction}
-\begin{reaction}
-Coper \rightleftharpoons nicus \\
-\end{reaction}
-\begin{reaction}
-Publi \leftrightarrow cations
-\end{reaction}
-
-\subsection{PHYSICAL UNITS}
-
-Please use \texttt{\textbackslash{}unit\{\}} (allows to save the
-math/\texttt{\$} environment) and apply the exponential notation, for
-example \(3.14\,\unit{km\,h^{-1}}\) (using LaTeX mode:
-\texttt{\textbackslash{}(\ 3.14\textbackslash{},\textbackslash{}unit\{...\}\ \textbackslash{})})
-or \unit{0.872\,m\,s^{-1}} (using only
-\texttt{\textbackslash{}unit\{0.872\textbackslash{},m\textbackslash{},s\^{}\{-1\}\}}).
 

### 2021-06-04 doc/manuscript/ESSDmanuscript.Rmd
finish filling in content from planning notes

@@
   - given_name: Susan 
-    surname: Cook-Patton (confirm)
+    surname: Cook-Patton
     affiliation: "4" 
   - given_name: Sandro 
-    surname: Federici (confirm)
+    surname: Federici 
     affiliation: "5" 
@@
   - given_name: Valentyna
-    surname: Slivinska (confirm)
+    surname: Slivinska 
     affiliation: "5"
@@
   - code: 5
-    address: 
+    address:  IPCC Task Force on National Greenhouse Gas Inventories Technical Support Unit, Institute for Global Environmental Strategies, Hayama, Japan
   - code: 6
@@
     use this to add a statement when having data sets and software code available
-  sample: |
-    use this section when having geoscientific samples available
-videosupplement: |
-  use this section when having video supplements available
 authorcontribution: |
-  Daniel wrote the package. Josiah thought about poterry. Markus filled in for a second author.
+  (fill this in)
 disclaimer: |
-  We like Copernicus.
 acknowledgements: |
-  Thanks to the rticles contributors!
+  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have helped to build the database.
+  Funding for this study was provided by The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
 appendix: |
-  \section{Figures and tables in appendices}
-  Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
-  \subsection{Option 1}
-  If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
-  They will be correctly named automatically.
-  \subsection{Option 2}
-  If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
 
-  To rename them correctly to A1, A2, etc., please add the following commands in front of them:
-  `\appendixfigures` needs to be added in front of appendix figures
-  `\appendixtables` needs to be added in front of appendix tables
-
-  Please add `\clearpage` between each table and/or figure. Further guidelines on figures and tables can be found below.
 output:
@@
 
-The following settings can or must be configured in the header of this file and are bespoke for Copernicus manuscripts:
-
-- The `journal` you are submitting to using the official abbreviation. You can use the function `rticles::copernicus_journal_abbreviations(name = '...')` to search the existing journals.
-
-- Specific sections of the manuscript:
-
-  - `running` with `title` and `author`
-
-  - `competinginterests`
-
-  - `copyrightstatement` (optional)
-
-  - `availability` (strongly recommended if any used), one of `code`, `data`, or `codedata`
-
-  - `authorcontribution`
-
-  - `disclaimer`
-
-  - `acknowledgements`
-
-See the defaults and examples from the skeleton and the official Copernicus documentation for details.
-
 **Important**: Always double-check with the official manuscript preparation guidelines at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html), especially the sections "Technical instructions for LaTeX" and "Manuscript composition".
@@
 
-*(Importance of forests for climate change mitigation)*
+**(Forest are critical for climate change mitigation)**
 
-*(Need for good data in international carbon accounting)*
+**(Need for good data in international carbon accounting)**
 
-*(Introduce EFDb & forc)*
-
-Example citation [@anderson-teixeira_forc_2018]
+**(Introduce EFDB & ForC)**
 
 Here, we:
-(1) clarify C cycle terminology
+(1) clarify definitions of relevant carbon stocks and increments
 (2) describe mapping of ForC to IPCC's EFDB,
-(3) describe updates to ForC  (ForC v4.0)
-(4) summarize the data in ForC that's relevant to EFDB, identifying gaps
-(5) provide recommendations for improving data collection, analysis, database, and accounting
+(3) describe updates to ForC  (ForC v4.0),
+(4) summarize the data in ForC that's relevant to EFDB, identifying gaps, and
+(5) provide recommendations for improving data collection, analysis, database, and accounting.
+We note that
 
@@
 
+# Mapping ForC to EFDB
+
+## Carbon cycle variables
+
+**Table: variable  mapping and equations-- give equations to calculate IPCC variables from C cycle variables**
+
+Define relationship among NEE, NEP, and delta.C., especially noting role of harvest.
+
+## Land use categories
+
+# Updates to ForC  (ForC v4.0)
+
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we implemented some modest restructuring, resolved duplicate records, and conducted quality control.
+
+This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].
+
+## ForC restructuring
+
+**Table: changes to ForC fields**.  
+Use  this table : https://github.com/forc-db/ForC/blob/master/database_management_records/record_of_changes.csv.
+
+## Resolving duplicates
+
+## Quality control
+
+# Results
+
+**figure: map of relevant ForC data with underlying FAO ecozones**
+
+summarize the data in ForC that's relevant to EFDB, identifying gaps
+
+dead wood and litter comparisons will be particularly interesting, as IPCC values are based on just a handful of references for each climate zone (table 2.2 in 2019 guidelines)
+
+# Recommendations
+
+## Data collection and analysis needs
+
+**(Paragraph highlighting important gaps in variables / regions)**
+
+Several variables of value to IPCC, including standing dead wood, woody mortality, delta.agb, are not calculated and presented as frequently as are AGB and ANPP_woody, even though they can readily be derived from the same census data.
+We recommend that researchers calculate and report these, as specified below.
+Furthermore, there is an opportunity to fill data gaps by calculating these from existing census data.
+For example, the core census protocol of the Forest Global Earth Observatory [ForestGEO; REFS] collects the data required to calculate standing dead wood, woody mortality, and delta.agb, but these have not been calculated and reported for all sites for which the appropriate number of censuses are available (n=1 for standing dead wood, n=2 for woody mortality and delta.agb) [but see REFS].
+
+**A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometries to convert DBH measurements to biomass.**
+*(Camille/Helene can write this paragraph easily.)*
+
+## Data reporting needs
+
+**We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards:**
+
+  - adopt common standards for variables like min diameter of deadwood, select soil sampling increments to include a cutoff at 30.
+
+  - report 95% CIs, SE, or STD and n
+
+  - report C variables in article text, table, or SI table. EFDB cannot accept data digitized from figures
+
+## Database needs
+
+There are plenty of relevant, published data that are not included in ForC. 
+Systematic review of the literature could vastly improve data coverage.
+*(There are some efforts underway, including a few that Susan can specify.)*
+
+## IPCC
+
+An important challenge is that forests are changing rapidly, and data collected a decaade ago may no longer be relevant, particularly in the cases of C increments and fluxes.
+
 \conclusions[Conclusions]

### 2021-06-04 doc/manuscript/ESSDmanuscript.tex
finish filling in content from planning notes

@@
 \Author[3]{Ben}{Bond-Lamberty (confirm)}
-\Author[4]{Susan}{Cook-Patton (confirm)}
-\Author[5]{Sandro}{Federici (confirm)}
+\Author[4]{Susan}{Cook-Patton}
+\Author[5]{Sandro}{Federici}
 \Author[6]{Helene}{Muller-Landau}
@@
 \Author[1]{Teagan}{Rogers}
-\Author[5]{Valentyna}{Slivinska (confirm)}
+\Author[5]{Valentyna}{Slivinska}
 \Author[1,6 *]{Kristina J.}{Anderson-Teixeira}
@@
 \affil[4]{}
-\affil[5]{}
+\affil[5]{IPCC Task Force on National Greenhouse Gas Inventories
+Technical Support Unit, Institute for Global Environmental Strategies,
+Hayama, Japan}
 \affil[6]{Forest Global Earth Observatory, Smithsonian Tropical Research
@@
 
-The following settings can or must be configured in the header of this
-file and are bespoke for Copernicus manuscripts:
-
-\begin{itemize}
-\item
-  The \texttt{journal} you are submitting to using the official
-  abbreviation. You can use the function
-  \texttt{rticles::copernicus\_journal\_abbreviations(name\ =\ \textquotesingle{}...\textquotesingle{})}
-  to search the existing journals.
-\item
-  Specific sections of the manuscript:
-
-  \begin{itemize}
-  \item
-    \texttt{running} with \texttt{title} and \texttt{author}
-  \item
-    \texttt{competinginterests}
-  \item
-    \texttt{copyrightstatement} (optional)
-  \item
-    \texttt{availability} (strongly recommended if any used), one of
-    \texttt{code}, \texttt{data}, or \texttt{codedata}
-  \item
-    \texttt{authorcontribution}
-  \item
-    \texttt{disclaimer}
-  \item
-    \texttt{acknowledgements}
-  \end{itemize}
-\end{itemize}
-
-See the defaults and examples from the skeleton and the official
-Copernicus documentation for details.
-
 \textbf{Important}: Always double-check with the official manuscript
@@
 
-\emph{(Importance of forests for climate change mitigation)}
+\textbf{(Forest are critical for climate change mitigation)}
 
-\emph{(Need for good data in international carbon accounting)}
+\textbf{(Need for good data in international carbon accounting)}
 
-\emph{(Introduce EFDb \& forc)}
+\textbf{(Introduce EFDB \& ForC)}
 
-Example citation \citep{anderson-teixeira_forc_2018}
-
-Here, we: (1) clarify C cycle terminology (2) describe mapping of ForC
-to IPCC's EFDB, (3) describe updates to ForC (ForC v4.0) (4) summarize
-the data in ForC that's relevant to EFDB, identifying gaps (5) provide
-recommendations for improving data collection, analysis, database, and
-accounting
+Here, we: (1) clarify definitions of relevant carbon stocks and
+increments (2) describe mapping of ForC to IPCC's EFDB, (3) describe
+updates to ForC (ForC v4.0), (4) summarize the data in ForC that's
+relevant to EFDB, identifying gaps, and (5) provide recommendations for
+improving data collection, analysis, database, and accounting. We note
+that
 
@@
 
+\section{Mapping ForC to EFDB}
+
+\subsection{Carbon cycle variables}
+
+\textbf{Table: variable mapping and equations-- give equations to
+calculate IPCC variables from C cycle variables}
+
+Define relationship among NEE, NEP, and delta.C., especially noting role
+of harvest.
+
+\subsection{Land use categories}
+
+\section{Updates to ForC (ForC v4.0)}
+
+To support export of data to EFDB, and to improve the overall quality of
+the ForC database, we implemented some modest restructuring, resolved
+duplicate records, and conducted quality control.
+
+This section describes changes relative to ForC v2.0
+\citep{anderson-teixeira_forc_2018}.
+
+\subsection{ForC restructuring}
+
+\textbf{Table: changes to ForC fields}.\\
+Use this table :
+https://github.com/forc-db/ForC/blob/master/database\_management\_records/record\_of\_changes.csv.
+
+\subsection{Resolving duplicates}
+
+\subsection{Quality control}
+
+\section{Results}
+
+\textbf{figure: map of relevant ForC data with underlying FAO ecozones}
+
+summarize the data in ForC that's relevant to EFDB, identifying gaps
+
+dead wood and litter comparisons will be particularly interesting, as
+IPCC values are based on just a handful of references for each climate
+zone (table 2.2 in 2019 guidelines)
+
+\section{Recommendations}
+
+\subsection{Data collection and analysis needs}
+
+\textbf{(Paragraph highlighting important gaps in variables / regions)}
+
+Several variables of value to IPCC, including standing dead wood, woody
+mortality, delta.agb, are not calculated and presented as frequently as
+are AGB and ANPP\_woody, even though they can readily be derived from
+the same census data. We recommend that researchers calculate and report
+these, as specified below. Furthermore, there is an opportunity to fill
+data gaps by calculating these from existing census data. For example,
+the core census protocol of the Forest Global Earth Observatory
+{[}ForestGEO; REFS{]} collects the data required to calculate standing
+dead wood, woody mortality, and delta.agb, but these have not been
+calculated and reported for all sites for which the appropriate number
+of censuses are available (n=1 for standing dead wood, n=2 for woody
+mortality and delta.agb) {[}but see REFS{]}.
+
+\textbf{A universal challenge in estimating biomass (living or dead)
+from forest census data is applying appropriate allometries to convert
+DBH measurements to biomass.} \emph{(Camille/Helene can write this
+paragraph easily.)}
+
+\subsection{Data reporting needs}
+
+\textbf{We recommend that, unless they have some specific reason to do
+otherwise, researchers calculate and report the values according to IPCC
+standards:}
+
+\begin{itemize}
+\item
+  adopt common standards for variables like min diameter of deadwood,
+  select soil sampling increments to include a cutoff at 30.
+\item
+  report 95\% CIs, SE, or STD and n
+\item
+  report C variables in article text, table, or SI table. EFDB cannot
+  accept data digitized from figures
+\end{itemize}
+
+\subsection{Database needs}
+
+There are plenty of relevant, published data that are not included in
+ForC. Systematic review of the literature could vastly improve data
+coverage. \emph{(There are some efforts underway, including a few that
+Susan can specify.)}
+
+\subsection{IPCC}
+
+An important challenge is that forests are changing rapidly, and data
+collected a decaade ago may no longer be relevant, particularly in the
+cases of C increments and fluxes.
+
 \conclusions[Conclusions]
@@
 
-\sampleavailability{use this section when having geoscientific samples
-available} %% use this section when having geoscientific samples available
-
-\videosupplement{use this section when having video supplements
-available} %% use this section when having geoscientific samples available
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-\section{Figures and tables in appendices}
-
-Regarding figures and tables in appendices, the following two options
-are possible depending on your general handling of figures and tables in
-the manuscript environment:
-
-\subsection{Option 1}
-
-If you sorted all figures and tables into the sections of the text,
-please also sort the appendix figures and appendix tables into the
-respective appendix sections. They will be correctly named
-automatically.
-
-\subsection{Option 2}
-
-If you put all figures after the reference list, please insert appendix
-tables and figures after the normal tables and figures.
-
-To rename them correctly to A1, A2, etc., please add the following
-commands in front of them: \texttt{\textbackslash{}appendixfigures}
-needs to be added in front of appendix figures
-\texttt{\textbackslash{}appendixtables} needs to be added in front of
-appendix tables
-
-Please add \texttt{\textbackslash{}clearpage} between each table and/or
-figure. Further guidelines on figures and tables can be found below.
-\noappendix
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontribution{Daniel wrote the package. Josiah thought about
-poterry. Markus filled in for a second author.} %% optional section
+\authorcontribution{(fill this in)} %% optional section
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\disclaimer{We like Copernicus.} %% optional section
 
@@
 \begin{acknowledgements}
-Thanks to the rticles contributors!
+Thank you to all researchers who collected and published the data
+contained in ForC, and to all research assistants and collaborators who
+have helped to build the database. Funding for this study was provided
+by The Nature Conservancy, the Institute for Global Environmental
+Strategies, WLS(?)
 \end{acknowledgements}

### 2021-06-04 doc/manuscript/ESSDmanuscript.Rmd
misc updates

@@
     surname: Banbury Morgan
-    affiliation: "2" 
+    affiliation: "1,2" 
   - given_name: Ben 
@@
 (5) provide recommendations for improving data collection, analysis, database, and accounting.
-We note that
 
@@
 To support export of data to EFDB, and to improve the overall quality of the ForC database, we implemented some modest restructuring, resolved duplicate records, and conducted quality control.
-
 This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].
@@
 
-**Table: changes to ForC fields**.  
-Use  this table : https://github.com/forc-db/ForC/blob/master/database_management_records/record_of_changes.csv.
+```{r, out.width = "12cm", echo = FALSE,fig.pos="H", fig.cap = "Table of changes to ForC fields (placeholder) "}
+#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
+knitr::include_graphics("figures_tables/ForC_updates_placeholder_fig.png")
+```
+*(The above is a placeholder for the table located at  https://github.com/forc-db/ForC/blob/master/database_management_records/record_of_changes.csv, which we'll need to format.)*
+
+## Quality control measures
+Prior to releasing ForC v4.0, we executed several quality control measures.
+First, to improve information on geographic coordinates, we flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
+Second, to identify erroneous climate data... *(Issue #212)[https://github.com/forc-db/ForC/issues/212]*.
 
 ## Resolving duplicates
-
-## Quality control
 
@@
 
-summarize the data in ForC that's relevant to EFDB, identifying gaps
+**(summarize the data in ForC that's relevant to EFDB, identifying gaps)**
 
@@
 
-**A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometries to convert DBH measurements to biomass.**
+A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometries to convert DBH measurements to biomass.
 *(Camille/Helene can write this paragraph easily.)*
@@
 
-**We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards:**
+We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards:
 
@@
   - report C variables in article text, table, or SI table. EFDB cannot accept data digitized from figures
+
+**Contributing data to ForC and/or EFDB directly will ensure its broader impact.**
+The latter is more efficient for getting data to EFDB, but does not get the data into ForC, where it can be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].
 
@@
 
+Remote sensing biomass estimates include standing dead wood [@duncanson_aboveground_2021].
+
 \conclusions[Conclusions]

### 2021-06-04 doc/manuscript/ESSDmanuscript.tex
misc updates

@@
 \Author[1]{Valentine}{Herrmann}
-\Author[2]{Rebecca}{Banbury Morgan}
+\Author[1,2]{Rebecca}{Banbury Morgan}
 \Author[3]{Ben}{Bond-Lamberty (confirm)}
@@
 relevant to EFDB, identifying gaps, and (5) provide recommendations for
-improving data collection, analysis, database, and accounting. We note
-that
+improving data collection, analysis, database, and accounting.
 
@@
 the ForC database, we implemented some modest restructuring, resolved
-duplicate records, and conducted quality control.
-
-This section describes changes relative to ForC v2.0
-\citep{anderson-teixeira_forc_2018}.
+duplicate records, and conducted quality control. This section describes
+changes relative to ForC v2.0 \citep{anderson-teixeira_forc_2018}.
 
@@
 
-\textbf{Table: changes to ForC fields}.\\
-Use this table :
-https://github.com/forc-db/ForC/blob/master/database\_management\_records/record\_of\_changes.csv.
+\begin{figure}[H]
+\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-2}
+\end{figure}
+
+\emph{(The above is a placeholder for the table located at
+https://github.com/forc-db/ForC/blob/master/database\_management\_records/record\_of\_changes.csv,
+which we'll need to format.)}
+
+\subsection{Quality control measures}
+
+Prior to releasing ForC v4.0, we executed several quality control
+measures. First, to improve information on geographic coordinates, we
+flagged and reviewed records with suspected low precision \emph{(Issue
+\#29){[}https://github.com/forc-db/ForC/issues/229{]}}. Second, to
+identify erroneous climate data\ldots{} \emph{(Issue
+\#212){[}https://github.com/forc-db/ForC/issues/212{]}}.
 
 \subsection{Resolving duplicates}
-
-\subsection{Quality control}
 
@@
 
-summarize the data in ForC that's relevant to EFDB, identifying gaps
+\textbf{(summarize the data in ForC that's relevant to EFDB, identifying
+gaps)}
 
@@
 
-\textbf{A universal challenge in estimating biomass (living or dead)
-from forest census data is applying appropriate allometries to convert
-DBH measurements to biomass.} \emph{(Camille/Helene can write this
-paragraph easily.)}
+A universal challenge in estimating biomass (living or dead) from forest
+census data is applying appropriate allometries to convert DBH
+measurements to biomass. \emph{(Camille/Helene can write this paragraph
+easily.)}
 
@@
 
-\textbf{We recommend that, unless they have some specific reason to do
+We recommend that, unless they have some specific reason to do
 otherwise, researchers calculate and report the values according to IPCC
-standards:}
+standards:
 
@@
 \end{itemize}
+
+\textbf{Contributing data to ForC and/or EFDB directly will ensure its
+broader impact.} The latter is more efficient for getting data to EFDB,
+but does not get the data into ForC, where it can be more broadly
+useful--for example, being used for basic science
+\citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021}
+or model benchmarking \citep{fer_ecosystem_2021}.
 
@@
 cases of C increments and fluxes.
+
+Remote sensing biomass estimates include standing dead wood
+\citep{duncanson_aboveground_2021}.
 

### 2021-06-06 doc/manuscript/ESSDmanuscript.Rmd

@@
   - given_name: Ben 
-    surname: Bond-Lamberty (confirm)
+    surname: Bond-Lamberty
     affiliation: "3" 
@@
   - code: 3
-    address: 
+    address: Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
   - code: 4

### 2021-06-07 doc/manuscript/ESSDmanuscript.Rmd
rough start at intro, fill in author affiliations

@@
   - code: 2
-    address: 
+    address: School of Geography, University of Leeds, Leeds, UK
   - code: 3
@@
   - code: 4
-    address: 
+    address: The Nature Conservancy; Arlington VA 22203, USA
   - code: 5
@@
 
-**(Forest are critical for climate change mitigation)**
+**As we seek to mitigate climate change [@unfccc_adoption_2015], forests are critical.**
+*(See intros to other papers for relevant refs)*
+Under Paris Agreement, **##**% of Nationally Determined Contributions relate to forests [@grassi_key_2017].
 
-**(Need for good data in international carbon accounting)**
+**The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories [(REFS for older guidelines), @ipcc_2019_2019].**
+These include guidance for accounting for forest land [@ipcc_agriculture_2006; @ipcc_2019_2019a].
+This guidance has improved as more of the relevant underlying data has become available [@requenasuarez_estimating_2019], but there remains room for continuous improvement as the science advances.
+For example, @cook-patton_mapping_2020 *(summarize IPCC comparison results)*. 
+*Moreover, it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.*
+To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
 
-**(Introduce EFDB & ForC)**
+**The Global Forest Carbon Database, ForC, is the largest collection of published estimates of forest carbon stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].**
+*(add stats/ details)*
+As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
+The purpose of this publication is to document that process and provide recommendations for future improvements.
 
-Here, we:
+**Here, we**
 (1) clarify definitions of relevant carbon stocks and increments

### 2021-06-07 doc/manuscript/ESSDmanuscript.tex
rough start at intro, fill in author affiliations

@@
 \Author[1,2]{Rebecca}{Banbury Morgan}
-\Author[3]{Ben}{Bond-Lamberty (confirm)}
+\Author[3]{Ben}{Bond-Lamberty}
 \Author[4]{Susan}{Cook-Patton}
@@
 Biology Institute, Front Royal, VA, United States}
-\affil[2]{}
-\affil[3]{}
-\affil[4]{}
+\affil[2]{School of Geography, University of Leeds, Leeds, UK}
+\affil[3]{Joint Global Change Research Institute, Pacific Northwest
+National Laboratory, College Park, MD, United States}
+\affil[4]{The Nature Conservancy; Arlington VA 22203, USA}
 \affil[5]{IPCC Task Force on National Greenhouse Gas Inventories
@@
 
-\textbf{(Forest are critical for climate change mitigation)}
+\textbf{As we seek to mitigate climate change
+\citep{unfccc_adoption_2015}, forests are critical.} \emph{(See intros
+to other papers for relevant refs)} Under Paris Agreement,
+\textbf{\#\#}\% of Nationally Determined Contributions relate to forests
+\citep{grassi_key_2017}.
 
-\textbf{(Need for good data in international carbon accounting)}
+\textbf{The International Panel on Climate Change (IPCC) provides
+guidance for national greenhouse gas inventories \citep[(REFS for older
+guidelines),][]{ipcc_2019_2019}.} These include guidance for accounting
+for forest land \citep{ipcc_agriculture_2006, ipcc_2019_2019a}. This
+guidance has improved as more of the relevant underlying data has become
+available \citep{requenasuarez_estimating_2019}, but there remains room
+for continuous improvement as the science advances. For example,
+\citet{cook-patton_mapping_2020} \emph{(summarize IPCC comparison
+results)}. \emph{Moreover, it is useful for those compiling national
+greenhouse gas inventories to have access to locally-specific
+information, when available.} To improve the data accessible for C
+accounting, the IPCC created the Emission Factor Database (EFDB;
+https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a
+recognized library of emission factors and other parameters that can be
+used for estimating greenhouse gas emissions and removals.
 
-\textbf{(Introduce EFDB \& ForC)}
+\textbf{The Global Forest Carbon Database, ForC, is the largest
+collection of published estimates of forest carbon stocks, increments,
+and annual fluxes
+\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}.}
+\emph{(add stats/ details)} As such, ForC is positioned to improve
+forest C accounting through the transfer of data to EFDB. The purpose of
+this publication is to document that process and provide recommendations
+for future improvements.
 
-Here, we: (1) clarify definitions of relevant carbon stocks and
+\textbf{Here, we} (1) clarify definitions of relevant carbon stocks and
 increments (2) describe mapping of ForC to IPCC's EFDB, (3) describe

### 2021-06-14 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd
#15

@@
 
+For data synthesis projects, compilation can only be useful to the EFDB if they include all the required, along with transparent description on the methodology applied to derive emission factors (or have a brief description and a reference to the original source) and the original emission factor values are present (not modified/rounded).
+
 **Contributing data to ForC and/or EFDB directly will ensure its broader impact.**

### 2021-06-15 doc/manuscript/ESSDmanuscript.Rmd
add some info

@@
 
+Documented at https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/defining_land_subcategory.md, https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/IPCC_LandUse_mapping.csv, and in [issue #8](https://github.com/forc-db/IPCC_database_integration/issues/8).
+
+Land classification under UNFCCC (e.g., https://cbmjournal.biomedcentral.com/articles/10.1186/s13021-018-0095-3) is (i) defined differently across countries and (ii) probably not defined the same way as it might be defined in publications included in ForC.
+Therefore, we do not transfer any classification of land as managed/unmanaged, but do provide any auxiliary info that may be useful in making this determination.
+
 # Updates to ForC  (ForC v4.0)

### 2021-06-15 doc/manuscript/ESSDmanuscript.tex
add some info

@@
 
+Documented at
+https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB\_mapping/defining\_land\_subcategory.md,
+https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB\_mapping/IPCC\_LandUse\_mapping.csv,
+and in
+\href{https://github.com/forc-db/IPCC_database_integration/issues/8}{issue
+\#8}.
+
+Land classification under UNFCCC (e.g.,
+https://cbmjournal.biomedcentral.com/articles/10.1186/s13021-018-0095-3)
+is (i) defined differently across countries and (ii) probably not
+defined the same way as it might be defined in publications included in
+ForC. Therefore, we do not transfer any classification of land as
+managed/unmanaged, but do provide any auxiliary info that may be useful
+in making this determination.
+
 \section{Updates to ForC (ForC v4.0)}
@@
 \end{itemize}
+
+For data synthesis projects, compilation can only be useful to the EFDB
+if they include all the required, along with transparent description on
+the methodology applied to derive emission factors (or have a brief
+description and a reference to the original source) and the original
+emission factor values are present (not modified/rounded).
 

### 2021-06-16 doc/manuscript/ESSDmanuscript.tex
trying to make table work

@@
 % The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
+%%\usepackage{multirow}
+\usepackage{multicol}
+\usepackage{colortbl}
+\usepackage{hhline}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{hyperref}
 %
@@
 
+\providecommand{\docline}[3]{\noalign{\global\setlength{\arrayrulewidth}{#1}}\arrayrulecolor[HTML]{#2}\cline{#3}}
+
+\setlength{\tabcolsep}{2pt}
+
+\renewcommand*{\arraystretch}{1.5}
+
+\begin{longtable}[c]{|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}}
+
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}-}
+
+\multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{pool}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{subpool}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{definition}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{major sources of estimate variation}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}!{\color[HTML]{000000}\vrule width 0pt}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{IPCC guidance}}} \\
+
+\noalign{\global\setlength{\arrayrulewidth}{2pt}}\arrayrulecolor[HTML]{666666}\cline{1-5}
+
+\endfirsthead
+
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=2pt}-}
+
+\multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{pool}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{subpool}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{definition}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{major sources of estimate variation}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}!{\color[HTML]{000000}\vrule width 0pt}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{IPCC guidance}}} \\
+
+\noalign{\global\setlength{\arrayrulewidth}{2pt}}\arrayrulecolor[HTML]{666666}\cline{1-5}\endhead
+
+\multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{biomass }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{aboveground }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{all biomass of living vegetation, both woody and herbaceous, above the soil }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ allometry, min dbh }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}!{\color[HTML]{000000}\vrule width 0pt}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ acceptable to exclude understory}}} \\
+
+\multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{  }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{belowground }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{all biomass of live roots }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ allometry, min dbh, assumed ratio of belowground to aboveground biomass (IPCC table 4.4) }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}!{\color[HTML]{000000}\vrule width 0pt}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter}}} \\
+
+\multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{dead organic matter }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{dead wood }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ min dbh,  ...
+}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}!{\color[HTML]{000000}\vrule width 0pt}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ default min dbh = 10cm, but may be chosen by country}}} \\
+
+\multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{litter }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ min dbh for dead wood , ..}}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}!{\color[HTML]{000000}\vrule width 0pt}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{}}} \\
+
+\multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{soils }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{soil organic matter }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{organic carbon in mineral soils to a specified depth }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ sampling depth }}} & \multicolumn{1}{!{\color[HTML]{000000}\vrule width 0pt}>{\raggedright}p{\dimexpr 0.75in+0\tabcolsep+0\arrayrulewidth}!{\color[HTML]{000000}\vrule width 0pt}}{\fontsize{11}{11}\selectfont{\textcolor[HTML]{000000}{ default sampling depth = 30cm, but may be chosen by country}}} \\
+
+\noalign{\global\setlength{\arrayrulewidth}{2pt}}\arrayrulecolor[HTML]{666666}\cline{1-5}
+
+\end{longtable}
+
 \begin{table*}[t]
@@
 \begin{figure}[H]
-\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-2}
+\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-3}
 \end{figure}

### 2021-06-16 doc/manuscript/ESSDmanuscript.Rmd
trying to make table work
updates related to land management categories

--- 2021-06-14 doc/manuscript/ESSDmanuscript.Rmd
@@
 
-**The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories [(REFS for older guidelines), @ipcc_2019_2019].**
-These include guidance for accounting for forest land [@ipcc_agriculture_2006; @ipcc_2019_2019a].
+**The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas (GHG) inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, (REFS for older guidelines), @ipcc_2019_2019].**
+Under this guidance, GHG inventories are to include all managed land, including most of the world's forest land [@ogle_delineating_2018].
+The IPCC inventory guidelines include specific instructions for accounting for forest land ... [@ipcc_agriculture_2006; @ipcc_2019_2019a].
 This guidance has improved as more of the relevant underlying data has become available [@requenasuarez_estimating_2019], but there remains room for continuous improvement as the science advances.
@@
 For quantifying forest role in global C cycle, we ultimately care about:
-(1) C stocks --stores of C that would be released to the atmosphere upon and use change
+(1) C stocks -- stores of C that would be vulnerable to release to the atmosphere upon land use change
 (2) C increments -- changes in those C stocks.
@@
 knitr::include_graphics("figures_tables/C_pools_fig.png")
+```
+
+```{r eval = FALSE, echo=FALSE, warning=FALSE}
+#another failed attempt at the table
+library(flextable)
+C_pools_table <- read.csv("figures_tables/C_pools.csv")
+ft <- flextable(C_pools_table)
+ft
+
 ```
@@
 ## Land use categories
+
+Documented at https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/defining_land_subcategory.md, https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/IPCC_LandUse_mapping.csv, and in [issue #8](https://github.com/forc-db/IPCC_database_integration/issues/8).
+
+The UNFCCC requires GHG reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*].
+This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 
+Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).
 

### 2021-06-16 doc/manuscript/ESSDmanuscript.tex
updates related to land management categories

--- 2021-06-15 doc/manuscript/ESSDmanuscript.tex
@@
 \textbf{The International Panel on Climate Change (IPCC) provides
-guidance for national greenhouse gas inventories \citep[(REFS for older
-guidelines),][]{ipcc_2019_2019}.} These include guidance for accounting
-for forest land \citep{ipcc_agriculture_2006, ipcc_2019_2019a}. This
-guidance has improved as more of the relevant underlying data has become
+guidance for national greenhouse gas (GHG) inventories for reporting to
+the United Nations Framework Convention on Climate Change \citep[UNFCCC,
+(REFS for older guidelines),][]{ipcc_2019_2019}.} Under this guidance,
+GHG inventories are to include all managed land, including most of the
+world's forest land \citep{ogle_delineating_2018}. The IPCC inventory
+guidelines include specific instructions for accounting for forest land
+\ldots{} \citep{ipcc_agriculture_2006, ipcc_2019_2019a}. This guidance
+has improved as more of the relevant underlying data has become
 available \citep{requenasuarez_estimating_2019}, but there remains room
@@
 For quantifying forest role in global C cycle, we ultimately care about:
-(1) C stocks --stores of C that would be released to the atmosphere upon
-and use change (2) C increments -- changes in those C stocks.
+(1) C stocks -- stores of C that would be vulnerable to release to the
+atmosphere upon land use change (2) C increments -- changes in those C
+stocks.
 
@@
 
-Land classification under UNFCCC (e.g.,
-https://cbmjournal.biomedcentral.com/articles/10.1186/s13021-018-0095-3)
-is (i) defined differently across countries and (ii) probably not
-defined the same way as it might be defined in publications included in
-ForC. Therefore, we do not transfer any classification of land as
-managed/unmanaged, but do provide any auxiliary info that may be useful
-in making this determination.
+The UNFCCC requires GHG reporting for all managed lands in a country,
+where management is defined as ``human interventions and practices have
+been applied to perform production, ecological or social functions''
+{[}\emph{IPCC 2006 full report REF}{]}. This definition is applied
+differently across countries, and is not clearly defined by the majority
+of governments \citep{ogle_delineating_2018}. Given this, and because
+the IPCC definition of management does not necessarily match that which
+would be reported in scientific publications and hence in ForC, we do
+not transfer any classification of land management status from ForC to
+the EFDB, but do provide auxiliary info that may be useful in making
+this determination (e.g., geographical location).
 

### 2021-06-16 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd

@@
 **The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas (GHG) inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, (REFS for older guidelines), @ipcc_2019_2019].**
-Under this guidance, GHG inventories are to include all managed land, including most of the world's forest land [@ogle_delineating_2018].
+Under this guidance, GHG inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
 The IPCC inventory guidelines include specific instructions for accounting for forest land ... [@ipcc_agriculture_2006; @ipcc_2019_2019a].

### 2021-06-16 doc/manuscript/ESSDmanuscript.Rmd
add increment variables
#6

@@
 **Here, we**
-(1) clarify definitions of relevant carbon stocks and increments
+(1) review definitions of relevant carbon stocks and increments
 (2) describe mapping of ForC to IPCC's EFDB,
@@
 
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we implemented some modest restructuring, resolved duplicate records, and conducted quality control.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added ## increment variables, implemented some modest restructuring, resolved duplicate records, and conducted quality control.
 This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].
+
+## Increment variables
+We added ## increment variables to the set of named and defined variables. 
+*(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)*
+These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
+
+Although these variables currently lack records, the structure exists such that records can be populated over time.
 

### 2021-06-16 doc/manuscript/ESSDmanuscript.tex
add increment variables
#6

@@
 (REFS for older guidelines),][]{ipcc_2019_2019}.} Under this guidance,
-GHG inventories are to include all managed land, including most of the
-world's forest land \citep{ogle_delineating_2018}. The IPCC inventory
-guidelines include specific instructions for accounting for forest land
-\ldots{} \citep{ipcc_agriculture_2006, ipcc_2019_2019a}. This guidance
-has improved as more of the relevant underlying data has become
-available \citep{requenasuarez_estimating_2019}, but there remains room
-for continuous improvement as the science advances. For example,
+GHG inventories include all managed land, including most of the world's
+forest land \citep{ogle_delineating_2018}. The IPCC inventory guidelines
+include specific instructions for accounting for forest land \ldots{}
+\citep{ipcc_agriculture_2006, ipcc_2019_2019a}. This guidance has
+improved as more of the relevant underlying data has become available
+\citep{requenasuarez_estimating_2019}, but there remains room for
+continuous improvement as the science advances. For example,
 \citet{cook-patton_mapping_2020} \emph{(summarize IPCC comparison
@@
 
-\textbf{Here, we} (1) clarify definitions of relevant carbon stocks and
+\textbf{Here, we} (1) review definitions of relevant carbon stocks and
 increments (2) describe mapping of ForC to IPCC's EFDB, (3) describe
@@
 To support export of data to EFDB, and to improve the overall quality of
-the ForC database, we implemented some modest restructuring, resolved
-duplicate records, and conducted quality control. This section describes
-changes relative to ForC v2.0 \citep{anderson-teixeira_forc_2018}.
+the ForC database, we added \#\# increment variables, implemented some
+modest restructuring, resolved duplicate records, and conducted quality
+control. This section describes changes relative to ForC v2.0
+\citep{anderson-teixeira_forc_2018}.
+
+\subsection{Increment variables}
+
+We added \#\# increment variables to the set of named and defined
+variables.
+\emph{(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)} These
+are directly related to C stocks as previously defined in ForC, with
+``\emph{delta.}'' added in front of the variable name.
+
+Although these variables currently lack records, the structure exists
+such that records can be populated over time.
 

### 2021-06-16 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd
#6

@@
 
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added ## increment variables, implemented some modest restructuring, resolved duplicate records, and conducted quality control.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added ten increment variables, implemented some modest restructuring, resolved duplicate records, and conducted quality control.
 This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].
@@
 ## Increment variables
-We added ## increment variables to the set of named and defined variables. 
+We added ten increment variables to the set of named and defined variables (or 20, counting _OM and _C versions), which previously included only one (aboveground biomass increment, *delta.agb*). 
 *(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)*

### 2021-06-17 doc/manuscript/ESSDmanuscript.Rmd

@@
 Forest ecosystem C stocks may be parsed into pools in various ways. 
-IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table 1).
+IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table&nbsp;\@ref(table_variables)).
 Quantifying these requires a one-time measurement.
 
-\newpage
+```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
+C_pools_table <- read.csv("figures_tables/C_pools.csv")
+names(C_pools_table) <- stringr::str_replace_all(names(C_pools_table), "\\.", " ")
 
-**Table 1: variables with definitions and measurement methods.** Definitions from IPCC Table 1.1.
-(See Table 1.1 in IPCC guidance).
-(*Currently adding this as a figure (generated from original draft) because kableExtra doesn't seem to work in this template, and I can't quickly get the template format to work.
-Table that we want here is "figures_tables/C_pools.csv"*)
-
-```{r, out.width = "12cm", echo = FALSE, fig.cap = "Table 1"}
-#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
-knitr::include_graphics("figures_tables/C_pools_fig.png")
+library(knitr)
+library(kableExtra)
+knitr::kable(C_pools_table, format = "latex",
+             longtable = FALSE,
+             caption = "\\textbf{Variables with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
+  kableExtra::kable_styling(full_width = TRUE) %>%
+  row_spec(0, bold = TRUE)
 ```
-
-```{r eval = FALSE, echo=FALSE, warning=FALSE}
-#another failed attempt at the table
-library(flextable)
-C_pools_table <- read.csv("figures_tables/C_pools.csv")
-ft <- flextable(C_pools_table)
-ft
-
-```
-
-\begin{table*}[t]
-\caption{This is a start at table 1 using the template format. }
-\begin{tabular}{l c c c r}
-\tophline
-
-pool & subpool & definition & major sources of estimate variation & IPCC guidance \\
-\middlehline
-biomass & aboveground & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory  \\
-
-\bottomhline
-\end{tabular}
-\belowtable{I don't know how to adjust so that it doesn't run off the page.}
-\end{table*}
 

### 2021-06-17 doc/manuscript/ESSDmanuscript.Rmd
update+knit

@@
 title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
-journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'ESSD')`"
+journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
 author:
@@
 
+### Biomass
+Biomass includes living vegetation, above- and below-ground.
+
+The IPCC defines aboveground biomass as "all biomass of living vegetation, both woody and herbaceous, above the soil including stems, stumps, branches, bark, seeds, and foliage" [].
+
+Belowground biomass is defined as "all biomass of live roots" [].
+
+### Dead Organic Matter
+### Soil Organic Matter
+
 ```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}

### 2021-06-17 doc/manuscript/ESSDmanuscript.tex
update+knit

@@
 % The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
+%%\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 %
@@
 parses into biomass (aboveground and belowground), dead organic matter
-(dead wood and litter), and soil organic matter (Table 1). Quantifying
-these requires a one-time measurement.
+(dead wood and litter), and soil organic matter
+(Table~@ref(table\_variables)). Quantifying these requires a one-time
+measurement.
 
-\newpage
+\subsubsection{Biomass}
 
-\textbf{Table 1: variables with definitions and measurement methods.}
-Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).
-(\emph{Currently adding this as a figure (generated from original draft)
-because kableExtra doesn't seem to work in this template, and I can't
-quickly get the template format to work. Table that we want here is
-``figures\_tables/C\_pools.csv''})
+Biomass includes living vegetation, above- and below-ground.
 
-\begin{figure}
-\includegraphics[width=12cm]{figures_tables/C_pools_fig} \caption{Table 1}\label{fig:unnamed-chunk-1}
-\end{figure}
+The IPCC defines aboveground biomass as ``all biomass of living
+vegetation, both woody and herbaceous, above the soil including stems,
+stumps, branches, bark, seeds, and foliage'' {[}{]}.
 
-\begin{table*}[t]
-\caption{This is a start at table 1 using the template format. }
-\begin{tabular}{l c c c r}
-\tophline
+Belowground biomass is defined as ``all biomass of live roots'' {[}{]}.
 
-pool & subpool & definition & major sources of estimate variation & IPCC guidance \\
-\middlehline
-biomass & aboveground & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory  \\
+\subsubsection{Dead Organic Matter}
 
-\bottomhline
-\end{tabular}
-\belowtable{I don't know how to adjust so that it doesn't run off the page.}
-\end{table*}
+\subsubsection{Soil Organic Matter}
+
+\begin{table}
+
+\caption{\label{tab:table_variables}\textbf{Variables with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
+\centering
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\hline
+\textbf{pool} & \textbf{subpool} & \textbf{definition} & \textbf{major sources of estimate variation} & \textbf{IPCC guidance}\\
+\hline
+biomass & aboveground & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory\\
+\hline
+ & belowground & all biomass of live roots & allometry, min dbh, assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
+\hline
+dead organic matter & dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min dbh,  ...
+& default min dbh = 10cm, but may be chosen by country\\
+\hline
+ & litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & min dbh for dead wood , ..
+& \\
+\hline
+soils & soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & default sampling depth = 30cm, but may be chosen by country\\
+\hline
+\end{tabu}
+\end{table}
 
@@
 To support export of data to EFDB, and to improve the overall quality of
-the ForC database, we added \#\# increment variables, implemented some
+the ForC database, we added ten increment variables, implemented some
 modest restructuring, resolved duplicate records, and conducted quality
@@
 
-We added \#\# increment variables to the set of named and defined
-variables.
+We added ten increment variables to the set of named and defined
+variables (or 20, counting \_OM and \_C versions), which previously
+included only one (aboveground biomass increment, \emph{delta.agb}).
 \emph{(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)} These
@@
 \begin{figure}[H]
-\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-3}
+\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-1}
 \end{figure}

### 2021-06-17 doc/manuscript/ESSDmanuscript.Rmd
add C variable mapping figure

@@
 
-**Figure: schematic illustrating fluxes in and out of each pool**
+```{r, out.width = "12cm", echo = FALSE, fig.cap = "Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.
+In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
+Additional caveats are as follows: 1- assumes minimal non-respiratory CO2 fluxes,...
+(Chapin et al.
+2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7-foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year)."}
+#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
+knitr::include_graphics("figures_tables/C_variable_mapping.png")
+```
 

### 2021-06-17 doc/manuscript/ESSDmanuscript.tex
add C variable mapping figure

@@
 
-\textbf{Figure: schematic illustrating fluxes in and out of each pool}
+\begin{figure}
+\includegraphics[width=12cm]{figures_tables/C_variable_mapping} \caption{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.
+In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
+Additional caveats are as follows: 1- assumes minimal non-respiratory CO2 fluxes,...
+(Chapin et al.
+2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7-foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year).}\label{fig:unnamed-chunk-1}
+\end{figure}
 
@@
 \begin{figure}[H]
-\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-1}
+\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-2}
 \end{figure}

### 2021-06-17 doc/manuscript/ESSDmanuscript.Rmd
add table of ForC changes

@@
 In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
-Additional caveats are as follows: 1- assumes minimal non-respiratory CO2 fluxes,...
+Additional caveats are as follows: 1- assumes minimal non-respiratory C fluxes,...
 (Chapin et al.
@@
 
-Define relationship among NEE, NEP, and delta.C., especially noting role of harvest.
+Define relationship among NEE, NEP, and delta.C., especially noting role of harvest [@chapin_reconciling_2006].
 
@@
 
-```{r, out.width = "12cm", echo = FALSE,fig.pos="H", fig.cap = "Table of changes to ForC fields (placeholder) "}
-#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
-knitr::include_graphics("figures_tables/ForC_updates_placeholder_fig.png")
+```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE}
+#"figures_tables/record_of_changes.csv" is copied over from the table located at  https://github.com/forc-db/ForC/blob/master/database_management_records/record_of_changes.csv, and I deleted the date field.
+
+record_of_changes <- read.csv("figures_tables/record_of_changes.csv")
+names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
+
+library(knitr)
+library(kableExtra)
+knitr::kable(record_of_changes, format = "latex",
+             longtable = FALSE,
+             caption = "\\textbf{Table of changes to ForC fields.}") %>%
+  kableExtra::kable_styling(full_width = TRUE) %>%
+  row_spec(0, bold = TRUE)
 ```
-*(The above is a placeholder for the table located at  https://github.com/forc-db/ForC/blob/master/database_management_records/record_of_changes.csv, which we'll need to format.)*
 

### 2021-06-17 doc/manuscript/ESSDmanuscript.tex
add table of ForC changes

@@
 Define relationship among NEE, NEP, and delta.C., especially noting role
-of harvest.
+of harvest \citep{chapin_reconciling_2006}.
 
@@
 
-\begin{figure}[H]
-\includegraphics[width=12cm]{figures_tables/ForC_updates_placeholder_fig} \caption{Table of changes to ForC fields (placeholder) }\label{fig:unnamed-chunk-2}
-\end{figure}
+\begin{table}
 
-\emph{(The above is a placeholder for the table located at
-https://github.com/forc-db/ForC/blob/master/database\_management\_records/record\_of\_changes.csv,
-which we'll need to format.)}
+\caption{\label{tab:table_ForCchanges}\textbf{Table of changes to ForC fields.}}
+\centering
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+Sites & coordinates.precision & Precision of geographic coordinates, as reported by source or estimated from maps. & field added & allow identification of records with poor coordinate precision\\
+\hline
+Measuremennts & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
+\hline
+ & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
+& replaces `stat` and `stat.name` & cleaner format; ability to handle assymetrical 95 percent confidence intervals\\
+\hline
+ & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
+\hline
+ & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
+& field added & track units conversion, allow back-calculation of OM if conversion factor deemed inappropriate\\
+\hline
+PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
+\hline
+ & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
+\hline
+Citations & (several fields) &  &  & \\
+\hline
+\end{tabu}
+\end{table}
 

### 2021-06-17 doc/manuscript/ESSDmanuscript.Rmd
improving figures & tables

@@
 
-```{r, out.width = "12cm", echo = FALSE, fig.cap = "Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.
-In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
+```{r, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
 Additional caveats are as follows: 1- assumes minimal non-respiratory C fluxes,...
@@
 
-**Table: variable  mapping and equations-- give equations to calculate IPCC variables from C cycle variables**
+Mapping is shown in Fig. 1
 

### 2021-06-17 doc/manuscript/ESSDmanuscript.tex
improving figures & tables

@@
 \centering
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
 \hline
-\textbf{pool} & \textbf{subpool} & \textbf{definition} & \textbf{major sources of estimate variation} & \textbf{IPCC guidance}\\
+\textbf{pool} & \textbf{definition} & \textbf{major sources of estimate variation} & \textbf{IPCC guidance}\\
 \hline
-biomass & aboveground & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory\\
+aboveground biomass & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory\\
 \hline
- & belowground & all biomass of live roots & allometry, min dbh, assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
+belowground biomass & all biomass of live roots & allometry, min dbh, assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
 \hline
-dead organic matter & dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min dbh,  ...
-& default min dbh = 10cm, but may be chosen by country\\
+dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min dbh,  ... & default min dbh = 10cm, but may be chosen by country\\
 \hline
- & litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & min dbh for dead wood , ..
+litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & min dbh for dead wood , ..
 & \\
 \hline
-soils & soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & default sampling depth = 30cm, but may be chosen by country\\
+soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & default sampling depth = 30cm, but may be chosen by country\\
 \hline
@@
 \begin{figure}
-\includegraphics[width=12cm]{figures_tables/C_variable_mapping} \caption{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.
-In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
+\includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
 Additional caveats are as follows: 1- assumes minimal non-respiratory C fluxes,...
@@
 
-\textbf{Table: variable mapping and equations-- give equations to
-calculate IPCC variables from C cycle variables}
+Mapping is shown in Fig. 1
 

### 2021-06-17 doc/manuscript/ESSDmanuscript.Rmd
modest structural updates

@@
 **Here, we**
-(1) review definitions of relevant carbon stocks and increments
+(1) review IPCC definitions of relevant carbon stocks and increments
 (2) describe mapping of ForC to IPCC's EFDB,
@@
 
-# Defining carbon stocks and incremenets
+# IPCC definitions of carbon stocks and incremenets
 
@@
 Quantifying these requires a one-time measurement.
-
-### Biomass
-Biomass includes living vegetation, above- and below-ground.
-
-The IPCC defines aboveground biomass as "all biomass of living vegetation, both woody and herbaceous, above the soil including stems, stumps, branches, bark, seeds, and foliage" [].
-
-Belowground biomass is defined as "all biomass of live roots" [].
-
-### Dead Organic Matter
-### Soil Organic Matter
 
@@
 
+### Biomass
+Biomass includes living vegetation, above- and below-ground.
+
+The IPCC defines aboveground biomass as "all biomass of living vegetation, both woody and herbaceous, above the soil including stems, stumps, branches, bark, seeds, and foliage" [].
+
+Belowground biomass is defined as "all biomass of live roots" [].
+
+### Dead Organic Matter
+### Soil Organic Matter
+
 ## Carbon increments
@@
 
+\clearpage
 ```{r, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).

### 2021-06-17 doc/manuscript/ESSDmanuscript.tex
modest structural updates

@@
 
-\textbf{Here, we} (1) review definitions of relevant carbon stocks and
-increments (2) describe mapping of ForC to IPCC's EFDB, (3) describe
+\textbf{Here, we} (1) review IPCC definitions of relevant carbon stocks
+and increments (2) describe mapping of ForC to IPCC's EFDB, (3) describe
 updates to ForC (ForC v4.0), (4) summarize the data in ForC that's
@@
 
-\section{Defining carbon stocks and incremenets}
+\section{IPCC definitions of carbon stocks and incremenets}
 
@@
 measurement.
-
-\subsubsection{Biomass}
-
-Biomass includes living vegetation, above- and below-ground.
-
-The IPCC defines aboveground biomass as ``all biomass of living
-vegetation, both woody and herbaceous, above the soil including stems,
-stumps, branches, bark, seeds, and foliage'' {[}{]}.
-
-Belowground biomass is defined as ``all biomass of live roots'' {[}{]}.
-
-\subsubsection{Dead Organic Matter}
-
-\subsubsection{Soil Organic Matter}
 
@@
 
+\subsubsection{Biomass}
+
+Biomass includes living vegetation, above- and below-ground.
+
+The IPCC defines aboveground biomass as ``all biomass of living
+vegetation, both woody and herbaceous, above the soil including stems,
+stumps, branches, bark, seeds, and foliage'' {[}{]}.
+
+Belowground biomass is defined as ``all biomass of live roots'' {[}{]}.
+
+\subsubsection{Dead Organic Matter}
+
+\subsubsection{Soil Organic Matter}
+
 \subsection{Carbon increments}
@@
 
+\clearpage
 \begin{figure}

### 2021-06-21 doc/manuscript/ESSDmanuscript.Rmd
clarify definition of "litter"

@@
 ### Dead Organic Matter
+
+Litter is defined as including " all  non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil.
+This includes litter,  fumic,  and  humic  layers.
+Live  fine  roots  (of  less  than  the  suggested  diameter  limit for  belowground  biomass) are included in litter where they cannot be distinguished from it empirically." (2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
+
 ### Soil Organic Matter

### 2021-06-21 doc/manuscript/ESSDmanuscript.tex
clarify definition of "litter"

@@
 litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & min dbh for dead wood , ..
-& \\
+& includes entire O horizon\\
 \hline
@@
 \subsubsection{Dead Organic Matter}
+
+Litter is defined as including " all non-living biomass with a diameter
+less than a minimum diameter chosen by the country (for example 10 cm),
+lying dead, in various states of decomposition above the mineral or
+organic soil. This includes litter, fumic, and humic layers. Live fine
+roots (of less than the suggested diameter limit for belowground
+biomass) are included in litter where they cannot be distinguished from
+it empirically." (2003 IPCc GPG for LULUCF
+(https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
 

### 2021-06-21 doc/manuscript/ESSDmanuscript.Rmd
add a bit of text
Update ESSDmanuscript.Rmd

@@
 
+Dead organic matter includes all non-living biomass that is not within the mineral soil layer and smaller than the litter size threshold.
+
+Dead wood is defined as...
+
 Litter is defined as including " all  non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil.
@@
 ### Soil Organic Matter
+
+Soil organic matter is defined as...
 
@@
 
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added ten increment variables, implemented some modest restructuring, resolved duplicate records, and conducted quality control.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we defined ## new variables, implemented some modest restructuring, resolved duplicate records, and conducted quality control.
 This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].
 
-## Increment variables
-We added ten increment variables to the set of named and defined variables (or 20, counting _OM and _C versions), which previously included only one (aboveground biomass increment, *delta.agb*). 
+## Defining new variables
+
+We added eleven increment variables to the set of named and defined variables (or 22, counting _OM and _C versions), which previously included only one (aboveground biomass increment, *delta.agb*). 
 *(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)*
@@
 Although these variables currently lack records, the structure exists such that records can be populated over time.
+
+We clarified the definition of *organic.layer* (O-horizon, including *litter*) and added the variable *litter* (relatively undecomposed plant material/ OL horizon).
 

### 2021-06-21 doc/manuscript/ESSDmanuscript.Rmd
litter/ O.horizon/ organic.layer distinction
https://github.com/forc-db/ForC/issues/240

#19

@@
 Litter is defined as including " all  non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil.
-This includes litter,  fumic,  and  humic  layers.
+This includes litter (OL),  fumic (OF),  and  humic (OH) layers.
 Live  fine  roots  (of  less  than  the  suggested  diameter  limit for  belowground  biomass) are included in litter where they cannot be distinguished from it empirically." (2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
@@
 
-We clarified the definition of *organic.layer* (O-horizon, including *litter*) and added the variable *litter* (relatively undecomposed plant material/ OL horizon).
+To provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including *litter* (OL)).
 

### 2021-06-21 doc/manuscript/ESSDmanuscript.tex
add a bit of text
litter/ O.horizon/ organic.layer distinction
https://github.com/forc-db/ForC/issues/240

#19

@@
 
+Dead organic matter includes all non-living biomass that is not within
+the mineral soil layer and smaller than the litter size threshold.
+
+Dead wood is defined as\ldots{}
+
 Litter is defined as including " all non-living biomass with a diameter
@@
 lying dead, in various states of decomposition above the mineral or
-organic soil. This includes litter, fumic, and humic layers. Live fine
-roots (of less than the suggested diameter limit for belowground
-biomass) are included in litter where they cannot be distinguished from
-it empirically." (2003 IPCc GPG for LULUCF
+organic soil. This includes litter (OL), fumic (OF), and humic (OH)
+layers. Live fine roots (of less than the suggested diameter limit for
+belowground biomass) are included in litter where they cannot be
+distinguished from it empirically." (2003 IPCc GPG for LULUCF
 (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
@@
 \subsubsection{Soil Organic Matter}
+
+Soil organic matter is defined as\ldots{}
 
@@
 To support export of data to EFDB, and to improve the overall quality of
-the ForC database, we added ten increment variables, implemented some
+the ForC database, we defined \#\# new variables, implemented some
 modest restructuring, resolved duplicate records, and conducted quality
@@
 
-\subsection{Increment variables}
+\subsection{Defining new variables}
 
-We added ten increment variables to the set of named and defined
-variables (or 20, counting \_OM and \_C versions), which previously
+We added eleven increment variables to the set of named and defined
+variables (or 22, counting \_OM and \_C versions), which previously
 included only one (aboveground biomass increment, \emph{delta.agb}).
@@
 such that records can be populated over time.
+
+To provide better definition of the previously existing variable
+\emph{organic.layer}, which has a nebulous definition that reflects the
+varied definitions adopted by original studies, we added two clearly
+defined variables: \emph{litter} (relatively undecomposed plant
+material/ OL horizon), and \emph{O.horizon} (entire O-horizon, including
+\emph{litter} (OL)).
 

### 2021-06-22 scripts/1_ForC_to_IPCC_formatting.R
drop Technologies/Practices mapping #8

@@ -538,7 +538,7 @@ EFDB <- data.frame("EF ID" = "",
                    "Fuel 2006 (ID)" = "",
                    "C pool" = V_mapping$IPCC.C_pool[m_vmap],
                    "Description" =  V_mapping$description[m_vmap],
-                   "Technologies/Practices" = generate_subfields("Technologies/Practices"),
+                   "Technologies/Practices" = "", #generate_subfields("Technologies/Practices"),
                    "Abatement/Control technologies" =  generate_subfields( "Abatement/Control technologies"),
                    "Parameters/Conditions" = generate_subfields("Parameters/Conditions"),
                    "Region/Regional conditions" = generate_subfields("Region/Regional conditions"),

### 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
start variable mapping table

@@
 
+```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
+
+ForC_variables_included <- read.csv("figures_tables/ForC_variables_included.csv")
+names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
+
+library(knitr)
+library(kableExtra)
+knitr::kable(ForC_variables_included, format = "latex",
+             longtable = FALSE,
+             caption = "\\textbf{Mapping of ForC fields to EFDB.}") %>%
+  kableExtra::kable_styling(full_width = TRUE) %>%
+  row_spec(0, bold = TRUE)
+```
+
 ## Carbon cycle variables
 
-Mapping is shown in Fig. 1
+Mapping of variables is shown in Fig. 1
 

### 2021-06-23 doc/manuscript/ESSDmanuscript.tex
start variable mapping table

@@
 
+\begin{table}
+
+\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.}}
+\centering
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\hline
+\textbf{Table} & \textbf{Variable} & \textbf{EFDB field} & \textbf{Use} & \textbf{Required}\\
+\hline
+Sites &  & NA &  & \\
+\hline
+Measuremennts &  & NA &  & \\
+\hline
+ & sd, se, lower95\%CI, upper 95\%CI & NA &  & y\\
+\hline
+ & mean.in.original.units, original.units & NA &  & y\\
+\hline
+ & data.location.within.source & NA & confirm that data weren't digitized, facilitate finding data in original publication & y\\
+\hline
+PFT & description & NA &  & \\
+\hline
+ &  & NA &  & \\
+\hline
+Citations & (several fields) & NA &  & \\
+\hline
+\end{tabu}
+\end{table}
+
 \subsection{Carbon cycle variables}
 
-Mapping is shown in Fig. 1
+Mapping of variables is shown in Fig. 1
 

### 2021-06-23 doc/manuscript/ESSDmanuscript.tex
updating table

--- 2021-06-21 doc/manuscript/ESSDmanuscript.tex
@@
 
+\begin{table}
+
+\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.}}
+\centering
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\hline
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Use} & \textbf{Required}\\
+\hline
+Sites &  &  &  & \\
+\hline
+Measuremennts & variable.name & - & link to variable info in ForC\_variables table & y\\
+\hline
+ & mean & Value & (direct) & y\\
+\hline
+ & mean.in.original.units & Value in Common Units & (direct) & y\\
+\hline
+ & original.units & Common Unit & (direct) & y\\
+\hline
+ & sd, se, lower95\%CI, upper 95\%CI &  &  & y\\
+\hline
+ &  &  &  & \\
+\hline
+ & data.location.within.source &  & confirm that data weren't digitized, facilitate finding data in original publication & y\\
+\hline
+PFT & description &  &  & \\
+\hline
+variables & units &  &  & y\\
+\hline
+Citations & (several fields) &  &  & \\
+\hline
+\end{tabu}
+\end{table}
+
 \subsection{Carbon cycle variables}
 
-Mapping is shown in Fig. 1
+Mapping of variables is shown in Fig. 1
 

### 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
add comments from Susan

@@
   Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have helped to build the database.
-  Funding for this study was provided by The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
+  Funding for this study was provided by Bezos Earth Fund to The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
 appendix: |
@@
 **The Global Forest Carbon Database, ForC, is the largest collection of published estimates of forest carbon stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].**
-*(add stats/ details)*
+*(add stats/ details, maybe record of how ForC has grown over time)*
 As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
@@
 Quantifying these requires at least two measurements.
-
+(*But, Can carbon increments be inferred from a single measure and a known age (i.e., the approach we used in GROA)?*)
 Fluxes are the inputs and outputs to each pool.
@@
 
+(*strongly flag both useful variables that the EFDB does not track and useful variables that papers fail to include that EFDB needs*)
+
 ## Data collection and analysis needs

### 2021-06-23 doc/manuscript/ESSDmanuscript.tex
add comments from Susan

@@
 \citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}.}
-\emph{(add stats/ details)} As such, ForC is positioned to improve
-forest C accounting through the transfer of data to EFDB. The purpose of
-this publication is to document that process and provide recommendations
-for future improvements.
+\emph{(add stats/ details, maybe record of how ForC has grown over
+time)} As such, ForC is positioned to improve forest C accounting
+through the transfer of data to EFDB. The purpose of this publication is
+to document that process and provide recommendations for future
+improvements.
 
@@
 outputs to the pool (i.e., fluxes). Quantifying these requires at least
-two measurements.
-
+two measurements. (\emph{But, Can carbon increments be inferred from a
+single measure and a known age (i.e., the approach we used in GROA)?})
 Fluxes are the inputs and outputs to each pool.
@@
 
+(\emph{strongly flag both useful variables that the EFDB does not track
+and useful variables that papers fail to include that EFDB needs})
+
 \subsection{Data collection and analysis needs}
@@
 have helped to build the database. Funding for this study was provided
-by The Nature Conservancy, the Institute for Global Environmental
-Strategies, WLS(?)
+by Bezos Earth Fund to The Nature Conservancy, the Institute for Global
+Environmental Strategies, WLS(?)
 \end{acknowledgements}

### 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
switch to reading record_of_changes direct from ForC repo
previously, was reading a copy manually added to this repo. This change simplifies/ ensures we have the latest version.

@@
 ```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE}
-#"figures_tables/record_of_changes.csv" is copied over from the table located at  https://github.com/forc-db/ForC/blob/master/database_management_records/record_of_changes.csv, and I deleted the date field.
-
-record_of_changes <- read.csv("figures_tables/record_of_changes.csv")
+library (readr)
+urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
+record_of_changes<-read_csv(url(urlfile))
 names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
+record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
+record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
 

### 2021-06-23 doc/manuscript/ESSDmanuscript.tex
switch to reading record_of_changes direct from ForC repo
previously, was reading a copy manually added to this repo. This change simplifies/ ensures we have the latest version.

@@
 \hline
-Measuremennts & variable.name & - & link to variable info in ForC\_variables table & y\\
+Measurements & variable.name & - & link to variable info in ForC\_variables table & y\\
 \hline
@@
 
+\begin{verbatim}
+## 
+## -- Column specification --------------------------------------------------------
+## cols(
+##   Table = col_character(),
+##   Column = col_character(),
+##   Description = col_character(),
+##   Changes = col_character(),
+##   Motivation = col_character(),
+##   `Date changed` = col_character()
+## )
+\end{verbatim}
+
 \begin{table}
@@
 \hline
-Measuremennts & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
+Measurements & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
 \hline
@@
 \hline
-Citations & (several fields) &  &  & \\
+Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC\\
+\hline
+ & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
+\hline
+ & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
+\hline
+ & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
+\hline
+ & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
 \hline

### 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
split record_of_changes table across pages
previously ran off the page.

This change moved caption to center.

@@
 knitr::kable(record_of_changes, format = "latex",
-             longtable = FALSE,
+             longtable = TRUE,
              caption = "\\textbf{Table of changes to ForC fields.}") %>%
-  kableExtra::kable_styling(full_width = TRUE) %>%
+  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
   row_spec(0, bold = TRUE)

### 2021-06-23 doc/manuscript/ESSDmanuscript.tex
split record_of_changes table across pages
previously ran off the page.

This change moved caption to center.

@@
 
-\begin{table}
-
-\caption{\label{tab:table_ForCchanges}\textbf{Table of changes to ForC fields.}}
-\centering
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\caption{\label{tab:table_ForCchanges}\textbf{Table of changes to ForC fields.}}\\
 \hline
@@
 \hline
+\endfirsthead
+\caption[]{\textbf{Table of changes to ForC fields.} \textit{(continued)}}\\
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+\endhead
 Sites & coordinates.precision & Precision of geographic coordinates, as reported by source or estimated from maps. & field added & allow identification of records with poor coordinate precision\\
@@
 \hline
-\end{tabu}
-\end{table}
+\end{longtabu}
 

### 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
suppress column specification message

@@
 
-```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
 library (readr)
 urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
-record_of_changes<-read_csv(url(urlfile))
+record_of_changes <-read_csv(url(urlfile))
 names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
 record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
-record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
+#record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
 

### 2021-06-23 doc/manuscript/ESSDmanuscript.tex
suppress column specification message

@@
 
-\begin{verbatim}
-## 
-## -- Column specification --------------------------------------------------------
-## cols(
-##   Table = col_character(),
-##   Column = col_character(),
-##   Description = col_character(),
-##   Changes = col_character(),
-##   Motivation = col_character(),
-##   `Date changed` = col_character()
-## )
-\end{verbatim}
-
 \begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
@@
 \hline
- & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
+NA & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
 & replaces `stat` and `stat.name` & cleaner format; ability to handle assymetrical 95 percent confidence intervals\\
 \hline
- & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
+NA & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
 \hline
- & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
+NA & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
 & field added & track units conversion, allow back-calculation of OM if conversion factor deemed inappropriate\\
 \hline
-PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
+PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & NA\\
 \hline
- & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
+NA & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & NA\\
 \hline
@@
 \hline
- & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
+NA & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
 \hline
- & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
+NA & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
 \hline
- & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
+NA & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
 \hline
- & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
+NA & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
 \hline

### 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
reinstate removal of NAs from table

--- 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
@@
 
-```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
 library (readr)
 urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
-record_of_changes<-read_csv(url(urlfile))
+record_of_changes <-read_csv(url(urlfile))
 names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")

### 2021-06-24 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd
@mawilliams99 ,  in line 251 I'm trying (with my inadequate R skills!) to exclude rows for which Motivation = "housekeeping" from the table, but this is not working.  If this is fast for you, it would be great if you could help.

--- 2021-06-23 doc/manuscript/ESSDmanuscript.Rmd
@@
 
-```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
 library (readr)
 urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
-record_of_changes<-read_csv(url(urlfile))
+record_of_changes <-read_csv(url(urlfile))
 names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
 record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
+#record_of_changes<- record_of_changes[which(!grep("housekeeping",record_of_changes$Motivation)) , c(1:5)]  #trying to exclude rows for which Motivation = "housekeeping", but this is not working
 record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks

### 2021-06-24 doc/manuscript/ESSDmanuscript.Rmd
updating table 2

@@
 knitr::kable(ForC_variables_included, format = "latex",
-             longtable = FALSE,
+             longtable = TRUE,
              caption = "\\textbf{Mapping of ForC fields to EFDB.}") %>%
-  kableExtra::kable_styling(full_width = TRUE) %>%
+  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
   row_spec(0, bold = TRUE)

### 2021-06-24 doc/manuscript/ESSDmanuscript.tex
updating table 2

@@
 
-\begin{table}
-
-\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.}}
-\centering
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.}}\\
 \hline
@@
 \hline
-Sites &  &  &  & \\
+\endfirsthead
+\caption[]{\textbf{Mapping of ForC fields to EFDB.} \textit{(continued)}}\\
 \hline
-Measurements & variable.name & - & link to variable info in ForC\_variables table & y\\
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Use} & \textbf{Required}\\
 \hline
- & mean & Value & (direct) & y\\
+\endhead
+Measurements & dominant.life.form & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & yes\\
 \hline
- & mean.in.original.units & Value in Common Units & (direct) & y\\
+ & stand.age & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Parameters/ Conditions & used to determine land subcategories (see defining\_land\_subcategory.md), directly listed in Parameters/ Conditions & (yes)\\
 \hline
- & original.units & Common Unit & (direct) & y\\
+ & dominant.veg, veg.notes, min.dbh & Parameters/ Conditions &  & no\\
 \hline
- & sd, se, lower95\%CI, upper 95\%CI &  &  & y\\
+ & variable.name & - & link to variable info in ForC\_variables table & yes\\
+\hline
+ & mean & Value & direct mapping & yes\\
+\hline
+ & mean.in.original.units & Value in Common Units & direct mapping & yes\\
+\hline
+ & original.units & Common Unit & direct mapping & yes\\
+\hline
+ & lower95\%CI, upper 95\%CI, se, sd and n & Lower Confidence Limit, Upper Confidence Limit & direct or calculated & (yes)\\
 \hline
@@
 \hline
- & data.location.within.source &  & confirm that data weren't digitized, facilitate finding data in original publication & y\\
+ & data.location.within.source &  & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
 \hline
-PFT & description &  &  & \\
+Sites & soil.texture, sand, silt, clay, soil.classification, FAO.ecozone & Parameters/ Conditions & direct mapping & no\\
 \hline
-variables & units &  &  & y\\
+ & country, state, city, lat, lon, masl, continent, Koeppen, mat, map & Region/Regional conditions & direct mapping, and lat-lon used to extract continent, Koeppen, and FAO.ecozone & no\\
+\hline
+History & (need to detail this) & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine distmrs.type & (yes)\\
+\hline
+ & (need to detail this) & Abatement/Control technologies & generate list management events & no\\
+\hline
+Plots & distmrs.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & auto\\
+\hline
+PFT & description &  &  & auto\\
+\hline
+variables & variable.type & Gases & For stocks in unit of organic matter, gases include CO2, CO, CH4, NO, NO2, N2O. For increments, fluxes, and stocks in units of C, gases includes only CO2. & auto\\
+\hline
+ & variable.name & C pool &  & auto\\
+\hline
+ &  & Description &  & auto\\
+\hline
+ & units &  &  & auto\\
 \hline
@@
  & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
+\hline
+ & EFDB.ready & Indicates whether data have been checked for export to EFDB. & field added & housekeeping\\
 \hline

### 2021-06-24 doc/manuscript/ESSDmanuscript.Rmd
finish filling table 2

@@
 names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
+ForC_variables_included <- ForC_variables_included[c(1:5)] #include all fields but TO_DO
 

### 2021-06-24 doc/manuscript/ESSDmanuscript.tex
finish filling table 2

--- 2021-06-23 doc/manuscript/ESSDmanuscript.tex
@@
 \hline
-Sites &  &  &  & \\
+\endfirsthead
+\caption[]{\textbf{Mapping of ForC fields to EFDB.} \textit{(continued)}}\\
 \hline
-Measurements & variable.name & - & link to variable info in ForC\_variables table & y\\
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Use} & \textbf{Required}\\
 \hline
- & mean & Value & (direct) & y\\
+\endhead
+Measurements & measurement.ID & Other Properties & direct mapping & (no)\\
 \hline
- & mean.in.original.units & Value in Common Units & (direct) & y\\
+ & dominant.life.form & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & yes\\
 \hline
- & original.units & Common Unit & (direct) & y\\
+ & stand.age & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Parameters/ Conditions & used to determine land subcategories (see defining\_land\_subcategory.md), directly listed in Parameters/ Conditions & (yes)\\
 \hline
- & sd, se, lower95\%CI, upper 95\%CI &  &  & y\\
+ & dominant.veg, veg.notes, min.dbh & Parameters/ Conditions & direct mapping/ linking to dominant.veg description & no\\
 \hline
- &  &  &  & \\
+ & variable.name & - & link to variable info in ForC\_variables table & yes\\
 \hline
- & data.location.within.source &  & confirm that data weren't digitized, facilitate finding data in original publication & y\\
+ & date / start.date, end.date & Other Properties & direct mapping & no\\
 \hline
-PFT & description &  &  & \\
+ & mean & Value & direct mapping & yes\\
 \hline
-variables & units &  &  & y\\
+ & mean.in.original.units & Value in Common Units & direct mapping & yes\\
 \hline
-Citations & (several fields) &  &  & \\
+ & original.units & Common Unit & direct mapping & yes\\
 \hline
-\end{tabu}
-\end{table}
+ & lower95\%CI, upper 95\%CI, se, sd and n & Lower Confidence Limit, Upper Confidence Limit & direct or calculated & (yes)\\
+\hline
+ & (covariates) & Other Properties & direct mapping & no\\
+\hline
+ & allometry & Comments from Data Provider & link to biomass allometry source & \\
+\hline
+ & data.location.within.source &  & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
+\hline
+ & ForC.investigator & Data Provider, Data Provider Contact & link to Data Provider, Data Provider Contact info & yes\\
+\hline
+Sites & site.ID, sites.sitename & Other Properties & direct mapping & (no)\\
+\hline
+ & soil.texture, sand, silt, clay, soil.classification, FAO.ecozone & Parameters/ Conditions & direct mapping & no\\
+\hline
+ & country, state, city, lat, lon, masl, continent, Koeppen, mat, map & Region/Regional conditions & direct mapping, and lat-lon used to extract continent, Koeppen, and FAO.ecozone & no\\
+\hline
+History & (need to detail this) & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine distmrs.type & (yes)\\
+\hline
+ & (need to detail this) & Abatement/Control technologies & generate list management events & no\\
+\hline
+ & plot.area & Other Properties & direct mapping & no\\
+\hline
+Plots & plot.ID, plot.name & Other Properties & direct mapping & (no)\\
+\hline
+ & distmrs.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & auto\\
+\hline
+ & distmrs.type, distmrs.year, regrowth.type, regrowth.year & Other Properties & direct mapping & auto\\
+\hline
+PFT & description & Parameters/ Conditions & direct mapping & auto\\
+\hline
+variables & variable.type & Gases & For stocks in unit of organic matter, gases include CO2, CO, CH4, NO, NO2, N2O. For increments, fluxes, and stocks in units of C, gases includes only CO2. & auto\\
+\hline
+ & variable.name & C pool, Equation & link to C pool, Equation & auto\\
+\hline
+ & description & Description & direct mapping & auto\\
+\hline
+ & extended.description & Other Properties & direct mapping & auto\\
+\hline
+ & units & Unit (ID) & link to IPCC units & auto\\
+\hline
+Citations & citation.citation & Full Technical Reference & direct mapping & yes/auto\\
+\hline
+ & citation.language & Reference Language & direct mapping & yes/auto\\
+\hline
+ & citation.url & URL & direct mapping & no/auto\\
+\hline
+ & citation.abstract & Abstract in English & direct mapping & no/auto\\
+\hline
+ & source.type & Source of Data & direct mapping & yes\\
+\hline
+\end{longtabu}
 
@@
 \hline
+ & source.type & citation source type & field added & field required by IPCC\\
+\hline
  & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
+\hline
+ & EFDB.ready & Indicates whether data have been checked for export to EFDB. & field added & housekeeping\\
 \hline

### 2021-06-24 doc/manuscript/ESSDmanuscript.tex
worked on Table 2

@@
 \hline
- & (covariates) & Other Properties & direct mapping & no\\
+ & depth, covariate\_1, cov\_1.value, covariate\_2, cov\_2.value & Other Properties & direct mapping & no\\
 \hline
- & allometry & Comments from Data Provider & link to biomass allometry source & \\
+ & allometry\_1, allometry\_2 & Comments from Data Provider & link to biomass allometry source, when provided & no\\
 \hline
- & data.location.within.source &  & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
+ & data.location.within.source & - & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
 \hline
@@
 \hline
-History & (need to detail this) & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine distmrs.type & (yes)\\
-\hline
- & (need to detail this) & Abatement/Control technologies & generate list management events & no\\
+History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)\\
 \hline

### 2021-06-25 doc/manuscript/ESSDmanuscript.Rmd
update table 2
I believe all the info is now complete/ correct, but some formatting is required

@@
 
+\clearpage
 ```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
 
@@
 
-library(knitr)
-library(kableExtra)
 knitr::kable(ForC_variables_included, format = "latex",
              longtable = TRUE,
-             caption = "\\textbf{Mapping of ForC fields to EFDB.}") %>%
-  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
-  row_spec(0, bold = TRUE)
+             caption = "\\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). ") %>%
+  row_spec(0, bold = TRUE) %>%
+kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
+
 ```
+'Required' field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
+
+** '(yes)' for most recent severe disturbance; 'no' for other history events
 

### 2021-06-25 doc/manuscript/ESSDmanuscript.tex
update table 2
I believe all the info is now complete/ correct, but some formatting is required

@@
 
+\clearpage
+
 \begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.}}\\
+\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). }\\
 \hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Use} & \textbf{Required}\\
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
 \hline
 \endfirsthead
-\caption[]{\textbf{Mapping of ForC fields to EFDB.} \textit{(continued)}}\\
+\caption[]{\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted).  \textit{(continued)}}\\
 \hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Use} & \textbf{Required}\\
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
 \hline
@@
 \hline
- & soil.texture, sand, silt, clay, soil.classification, FAO.ecozone & Parameters/ Conditions & direct mapping & no\\
+ & lat, lon & Region/Regional conditions & direct mapping; used to extract continent, Koeppen, and FAO.ecozone & (no)\\
 \hline
- & country, state, city, lat, lon, masl, continent, Koeppen, mat, map & Region/Regional conditions & direct mapping, and lat-lon used to extract continent, Koeppen, and FAO.ecozone & no\\
+ & country, state, city, masl,  mat, map & Region/Regional conditions & direct mapping & no\\
 \hline
-History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)\\
+ & continent, Koeppen & Region/Regional conditions & direct mapping & auto\\
+\hline
+ & soil.texture, sand, silt, clay, soil.classification & Parameters/ Conditions & direct mapping & no\\
+\hline
+ & FAO.ecozone & Parameters/ Conditions & direct mapping & auto\\
+\hline
+History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)/no**\\
 \hline
@@
 \end{longtabu}
+
+`Required' field indicates whether the field is required by EFDB: yes =
+value required; (yes) = input required, missing value acceptable if not
+reported; auto = present within ForC infrasructure, and therefore will
+always be exported to EFDB ; (no) = not required for EFDB, but required
+for ForC and therefore will always be exported to EFDB; no = not
+required, but exported to EFDB when a value is present.
+
+** `(yes)' for most recent severe disturbance; `no' for other history
+events
 

### 2021-06-25 scripts/1_ForC_to_IPCC_formatting.R
use source.type and EFDC.ready fields

@@ -549,7 +549,7 @@ EFDB <- data.frame("EF ID" = "",
                    "Common Unit" = ForC_simplified$original.units,
                    "Equation" =  ifelse(my_is.na(V_mapping$Equation[m_vmap]), "", V_mapping$Equation[m_vmap]),
                    "IPCC Worksheet Number" = ifelse(my_is.na(V_mapping$IPCC.Worksheet.Number[m_vmap]), "", my_is.na(V_mapping$IPCC.Worksheet.Number[m_vmap])),
-                   "Source of Data" = "Peer-reviewed journal", #  Peer-reviewed journal (usually), maybe not if no ref? to check
+                   "Source of Data" = CITATIONS$source.type[m_citations], # "Peer-reviewed journal", #  Peer-reviewed journal (usually), maybe not if no ref? to check
                    "Full Technical Reference" = CITATIONS$citation.citation[m_citations],
                    "URL" = CITATIONS$citation.url[m_citations],
                    "Reference Language" = CITATIONS$citation.language[m_citations],
@@ -585,23 +585,28 @@ EFDB <- data.frame("EF ID" = "",
 ## first delete any file so that if slightly different number of records, still overwrites.
 file.remove(list.files("data/1-to-review/", pattern = ".csv", full.names = T))
 
-for(c_id in   c("Archibald_2009_doiv",
-                "Chave_2008_aepa",
-                "Goulden_1996_eocd",
-                "Keith_2009_mmce",
-                "Keller_2001_beit",
-                "Lin_2012_tvia",
-                "Meakem_2018_rots",
-                "Ngo_2013_csip",
-                "Orihuela-Belmonte_2013_csaa",
-                "Rice_2004_cbav",
-                "Saleska_2003_ciaf",
-                "Toky_1983_ssfs",
-                "Uhl_1984_sand",
-                "Lutz_2018_giol",
-                "Johnson_2018_csss",
-                "Gonzalez-Akre_2016_potm")
-   ) { # unique(ForC_simplified$citation.ID )
+
+
+for(c_id in CITATIONS$citation.ID[CITATIONS$EFDB.ready %in% 1]) { 
+  # unique(ForC_simplified$citation.ID )
+  
+  # c("Archibald_2009_doiv",
[…]

### 2021-08-10 doc/manuscript/ESSDmanuscript.Rmd
remove Valentyna & Sandro from author list

@@
     affiliation: "4" 
-  - given_name: Sandro 
-    surname: Federici 
-    affiliation: "5" 
   - given_name: Helene
     surname: Muller-Landau
-    affiliation: "6" 
+    affiliation: "5" 
   - given_name: Camille
     surname: Piponiot
-    affiliation: "7" 
+    affiliation: "6" 
   - given_name: Teagan
@@
     affiliation: "1"
-  - given_name: Valentyna
-    surname: Slivinska 
-    affiliation: "5"
   - given_name: Kristina J.
     surname: Anderson-Teixeira
-    affiliation: "1,6 *"
+    affiliation: "1,5 *"
     email: teixeirak@si.edu
@@
   - code: 5
-    address:  IPCC Task Force on National Greenhouse Gas Inventories Technical Support Unit, Institute for Global Environmental Strategies, Hayama, Japan
+    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
   - code: 6
-    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
-  - code: 7
     address: CIRAD, Montpellier, France

### 2021-08-10 doc/manuscript/ESSDmanuscript.tex
remove Valentyna & Sandro from author list

@@
 
-%% \usepackage commands included in the copernicus.cls:
+%% Please DO NOT add additional packages or LaTeX commands to the template. They
+%% are not supported by Coperncius. LaTeX packages already
+%% included in the copernicus.cls are:
 %\usepackage[german, english]{babel}
@@
 %
-\usepackage{algorithmic}
-\usepackage{algorithm}
 
@@
 \Author[4]{Susan}{Cook-Patton}
-\Author[5]{Sandro}{Federici}
-\Author[6]{Helene}{Muller-Landau}
-\Author[7]{Camille}{Piponiot}
+\Author[5]{Helene}{Muller-Landau}
+\Author[6]{Camille}{Piponiot}
 \Author[1]{Teagan}{Rogers}
-\Author[5]{Valentyna}{Slivinska}
-\Author[1,6 *]{Kristina J.}{Anderson-Teixeira}
+\Author[1,5 *]{Kristina J.}{Anderson-Teixeira}
 
@@
 \affil[4]{The Nature Conservancy; Arlington VA 22203, USA}
-\affil[5]{IPCC Task Force on National Greenhouse Gas Inventories
-Technical Support Unit, Institute for Global Environmental Strategies,
-Hayama, Japan}
-\affil[6]{Forest Global Earth Observatory, Smithsonian Tropical Research
+\affil[5]{Forest Global Earth Observatory, Smithsonian Tropical Research
 Institute, Panama, Republic of Panama}
-\affil[7]{CIRAD, Montpellier, France}
+\affil[6]{CIRAD, Montpellier, France}
 

### 2021-09-08 doc/manuscript/ESSDmanuscript.Rmd
mention new CI system

@@
 Prior to releasing ForC v4.0, we executed several quality control measures.
-First, to improve information on geographic coordinates, we flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
-Second, to identify erroneous climate data... *(Issue #212)[https://github.com/forc-db/ForC/issues/212]*.
+First, we implemented a system of continuous integration using GitHub Actions (*sensu* Kim et al. in prep) to run some automatic checks any time the master data files are updated.
+Second, to improve information on geographic coordinates, we flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
+Third, to identify erroneous climate data... *(Issue #212)[https://github.com/forc-db/ForC/issues/212]*.
 

### 2021-09-08 doc/manuscript/ESSDmanuscript.tex
mention new CI system

@@
 Prior to releasing ForC v4.0, we executed several quality control
-measures. First, to improve information on geographic coordinates, we
-flagged and reviewed records with suspected low precision \emph{(Issue
-\#29){[}https://github.com/forc-db/ForC/issues/229{]}}. Second, to
+measures. First, we implemented a system of continuous integration using
+GitHub Actions (\emph{sensu} Kim et al.~in prep) to run some automatic
+checks any time the master data files are updated. Second, to improve
+information on geographic coordinates, we flagged and reviewed records
+with suspected low precision \emph{(Issue
+\#29){[}https://github.com/forc-db/ForC/issues/229{]}}. Third, to
 identify erroneous climate data\ldots{} \emph{(Issue

### 2021-09-13 doc/manuscript/ESSDmanuscript.Rmd
update variable mapping fig

@@
 (Chapin et al.
-2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7-foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year)."}
+2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 8 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year)."}
 #this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}

### 2021-09-13 doc/manuscript/ESSDmanuscript.tex
update variable mapping fig

@@
 (Chapin et al.
-2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7-foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year).}\label{fig:unnamed-chunk-1}
+2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 8 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year).}\label{fig:unnamed-chunk-1}
 \end{figure}

### 2021-09-13 doc/manuscript/ESSDmanuscript.Rmd
update figure & footnotes

--- 2021-09-08 doc/manuscript/ESSDmanuscript.Rmd
@@
 (Chapin et al.
-2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7-foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year)."}
+2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 7- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 8 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 9 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
 #this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}

### 2021-09-13 doc/manuscript/ESSDmanuscript.tex
update figure & footnotes

--- 2021-09-08 doc/manuscript/ESSDmanuscript.tex
@@
 (Chapin et al.
-2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7-foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year).}\label{fig:unnamed-chunk-1}
+2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 7- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 8 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 9 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:unnamed-chunk-1}
 \end{figure}

### 2021-09-13 scripts/X_generate_list_of_studies_for_review.R
bring back ready_to_rerun_and_send and notes

@@ -47,7 +47,21 @@ citation_list <- MEASUREMENTS %>%
             ready_to_rerun_and_send = "") %>% arrange(desc(review_priority_score))
 
 
+# merge notes from previous file if it exists ####
+filename = "data/citations_ordered_by_priority_score.csv"
+
+if(file.exists(filename)) {
+  old_citation_list <- read.csv(filename)
+  
+  m <- match(citation_list$citation.ID, old_citation_list$citation.ID)
+  
+  citation_list$ready_to_rerun_and_send <- old_citation_list$ready_to_rerun_and_send[m]
+  citation_list$notes <- old_citation_list$notes[m]
+  
+  
+}
+
 # save ####
-write.csv(citation_list, file = "data/citations_ordered_by_priority_score.csv", row.names = F)
+write.csv(citation_list, file = filename, row.names = F)
 
 

### 2021-09-15 scripts/1_ForC_to_IPCC_formatting.R
ignore case for "Table S"

@@ -384,7 +384,7 @@ any(is.na(m_meas)) # should be FALSE
 
 ForC_simplified$data.location.within.source <- MEASUREMENTS$data.location.within.source[m_meas]
 
-ForC_simplified$send_SI <- ifelse(grepl("Table S", ForC_simplified$data.location.within.source), 1, 0)
+ForC_simplified$send_SI <- ifelse(grepl("Table S", ForC_simplified$data.location.within.source, ignore.case = T), 1, 0)
 
 ## import and/or modify COVARIATES ####
 m_meas <- match(ForC_simplified$measurement.ID, MEASUREMENTS$measurement.ID)

### 2021-09-15 scripts/1_ForC_to_IPCC_formatting.R
fix order of start and end date #27

@@ -513,7 +513,7 @@ idx_start_end_date_NA <- my_is.na(ForC_simplified$start.date) | my_is.na(ForC_si
 
 
 ForC_simplified$mean_date <- round((as.numeric(ForC_simplified$start.date) +  as.numeric(ForC_simplified$end.date))/2, 1)
-ForC_simplified$start_end_date <- paste(ForC_simplified$end.date,   ForC_simplified$start.date, sep = "-")
+ForC_simplified$start_end_date <- paste(ForC_simplified$start.date,   ForC_simplified$end.date, sep = "-")
 
 # for stocks `date`, or average of start.date and end.date
   

### 2021-09-16 scripts/2_From_approved_csv_to_EFDB_form.R
add code to write list of files to send in the batch

@@ -54,4 +54,15 @@ for(file_to_export in approved_files) {
   
 }
 
+
+# add a text file with list of files sent in the batch ####
+batch_list_files <- list.files("data/4-transferred-to-EFDB/", pattern = ".txt")
+last_batch <- max(as.numeric(regmatches(batch_list_files, regexpr("\\d{1}", batch_list_files))))
+
+new_batch <- last_batch + 1
+write.table(gsub(".csv", "", approved_files), row.names = F, quote = F, col.names = F,
+            file = paste0("data/3-EFDB-forms-ready/list_files_batch_", new_batch, "_", gsub("-", "", Sys.Date()), ".txt"))
+
+
+# restartR ####
 .rs.restartR()

### 2021-09-21 doc/manuscript/ESSDmanuscript.Rmd
note that calculated variables are not acceptable

@@
 
+  - present calculations of all variables that would be useful to IPCC.
+EFDB requires that data in the database be presented in the original article, and as such cannot accept subsequent calculations.
+For example, if aboveground biomass and total biomass are presented, but root biomass is not presented, root biomass cannot be subsequently calculated and sent to EFDB.
+
 For data synthesis projects, compilation can only be useful to the EFDB if they include all the required, along with transparent description on the methodology applied to derive emission factors (or have a brief description and a reference to the original source) and the original emission factor values are present (not modified/rounded).

### 2021-09-21 doc/manuscript/ESSDmanuscript.Rmd
remove NEE, total.ecosystem from variable mapping diagram & manuscript

@@
 ```{r, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
-Additional caveats are as follows: 1- assumes minimal non-respiratory C fluxes,...
-(Chapin et al.
-2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 7- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 8 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 9 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
+Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
 #this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
@@
 Mapping of variables is shown in Fig. 1
-
-Define relationship among NEE, NEP, and delta.C., especially noting role of harvest [@chapin_reconciling_2006].
 

### 2021-09-21 doc/manuscript/ESSDmanuscript.tex
remove NEE, total.ecosystem from variable mapping diagram & manuscript

--- 2021-09-08 doc/manuscript/ESSDmanuscript.tex
@@
 \includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
-Additional caveats are as follows: 1- assumes minimal non-respiratory C fluxes,...
-(Chapin et al.
-2006); 2- assumes that change in foliage biomass is negligible (see note 7); 3- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 4- incomplete: excludes belowground components;  5-incomplete: excludes breakage into pieces less than dead wood threshold size; 6-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter; 7-foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year).}\label{fig:unnamed-chunk-1}
+Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:unnamed-chunk-1}
 \end{figure}
@@
   accept data digitized from figures
+\item
+  present calculations of all variables that would be useful to IPCC.
+  EFDB requires that data in the database be presented in the original
+  article, and as such cannot accept subsequent calculations. For
+  example, if aboveground biomass and total biomass are presented, but
+  root biomass is not presented, root biomass cannot be subsequently
+  calculated and sent to EFDB.
 \end{itemize}

### 2021-10-05 scripts/1_ForC_to_IPCC_formatting.R
add "sum" "difference", "+", and "-" as terms to avoid
@teixeirak, FYI

@@ -79,7 +79,7 @@ ForC_simplified$source.notes <-  MEASUREMENTS$source.notes[m_meas]
 ForC_simplified$data.location.within.source <- MEASUREMENTS$data.location.within.source[m_meas]
 
 
-ForC_simplified <- ForC_simplified[!grepl("Fig|Figure|digitized|not found|calculated", paste(ForC_simplified$source.notes, ForC_simplified$data.location.within.source), ignore.case = T), ]
+ForC_simplified <- ForC_simplified[!grepl("Fig|Figure|digitized|not found|calculated|sum|difference|\\+|\\-", paste(ForC_simplified$source.notes, ForC_simplified$data.location.within.source), ignore.case = T), ]
 
 
 # Generate/modify fields we need ####

### 2021-10-06 scripts/1_ForC_to_IPCC_formatting.R
add an error if current_LU did not map #29

@@ -100,6 +100,7 @@ ForC_simplified$current_LU [ForC_simplified$dominant.life.form %in% "crop"  & !m
 
 unique(ForC_simplified$dominant.life.form)
 unique(ForC_simplified$current_LU) # should be no ""
+if(any(ForC_simplified$current_LU %in% "")) stop("some current_LU did not map!")
 unique(ForC_simplified[ForC_simplified$current_LU %in% "", c("dominant.life.form", "stand.age", "current_LU")]) # should be empty
 
 

### 2021-10-07 scripts/1_ForC_to_IPCC_formatting.R
update to deal with past_LU better
change stand.age as numeric, deal with 999 differently
+ give error if some past LU other than Burned or StandClearing or empty did not map.

@@ -70,6 +70,10 @@ ForC_simplified$variable.name <-  MEASUREMENTS$variable.name[m_meas]
 
 ForC_simplified <- ForC_simplified[ForC_simplified$variable.name %in% V_mapping$variable.name[V_mapping$provide.to.IPCC %in% 1], ]
 
+# consider stand.age as numeric ####
+ForC_simplified$stand.age <- as.numeric(ForC_simplified$stand.age)
+
+
 ## exclude any records for which data.location.within.source includes "Figure" or "Fig", or source.notes includes "digitized".
 m_meas <- match(ForC_simplified$measurement.ID, MEASUREMENTS$measurement.ID)
 any(is.na(m_meas)) # should be FALSE
@@ -121,11 +125,16 @@ ForC_simplified$past_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$
 ForC_simplified$past_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age <20) & ForC_simplified$distmrs.type %in% c("No disturbance", "No severe disturbance", "Flood", "Forest dieback", "Landslide","Major Storm")] <- ForC_simplified$current_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age <20) & ForC_simplified$distmrs.type %in% c("No disturbance", "No severe disturbance", "Flood", "Forest dieback", "Landslide","Major Storm")]
 ForC_simplified$past_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age <20) & ForC_simplified$distmrs.type %in% c("Cut", "Harvest")] <- ForC_simplified$current_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age <20) & ForC_simplified$distmrs.type %in% c("Cut", "Harvest")]
 
-ForC_simplified$past_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age <20) & ForC_simplified$distmrs.type %in% c("Burned", "StandClearing") | my_is.na( ForC_simplified$distmrs.type)] <- ""
+ForC_simplified$past_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age <20) & ForC_simplified$distmrs.type %in% c("Burned", "StandClearing", "") | my_is.na( ForC_simplified$distmrs.type)] <- ""
 
[…]

### 2021-10-08 scripts/1_ForC_to_IPCC_formatting.R
change "stop" to "restart session" to make sure it stops when error

@@ -102,9 +102,10 @@ ForC_simplified$current_LU [ForC_simplified$dominant.life.form %in% "grass" & !m
 ForC_simplified$current_LU [ForC_simplified$dominant.life.form %in% "grass" & !my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age == 0] <- "Grassland"
 ForC_simplified$current_LU [ForC_simplified$dominant.life.form %in% "crop"  & !my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age == 0] <- "Cropland"
 
+
 unique(ForC_simplified$dominant.life.form)
 unique(ForC_simplified$current_LU) # should be no ""
-if(any(ForC_simplified$current_LU %in% "")) stop("some current_LU did not map!")
+if(any(ForC_simplified$current_LU %in% "")) rstudioapi::restartSession(command = "some current_LU did not map!")
 unique(ForC_simplified[ForC_simplified$current_LU %in% "", c("dominant.life.form", "stand.age", "current_LU")]) # should be empty
 
 
@@ -131,7 +132,7 @@ unique(ForC_simplified$past_LU)
 sum(ForC_simplified$past_LU[(!my_is.na(ForC_simplified$stand.age) & ForC_simplified$stand.age <20)]%in%"")
 unique(ForC_simplified[ForC_simplified$past_LU %in% "" & !( ForC_simplified$distmrs.type %in% c("Burned", "StandClearing", "") | my_is.na( ForC_simplified$distmrs.type)), c("distmrs.type", "stand.age", "past_LU")]) # should be empty
 
-if(sum(ForC_simplified$past_LU %in% "" & !( ForC_simplified$distmrs.type %in% c("Burned", "StandClearing", "") | my_is.na( ForC_simplified$distmrs.type))) > 0 ) stop("some Past LU did not map")
+if(sum(ForC_simplified$past_LU %in% "" & !( ForC_simplified$distmrs.type %in% c("Burned", "StandClearing", "") | my_is.na( ForC_simplified$distmrs.type))) > 0 ) rstudioapi::restartSession(command = "some current_LU did not map!")
 
 # temp <- unique(ForC_simplified[ForC_simplified$past_LU %in% "", c("distmrs.type", "stand.age", "past_LU")])
 # aggregate(stand.age~distmrs.type+past_LU, data = temp, FUN = range)

### 2021-12-08 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd

@@
 \introduction[Introduction]
+
+We are currently in a climate crisis and as we seek to find answers climate data is increasingly important.**something about climate change and freenhouse gasses** The influence the forests have on asmosphereic carbon has made forest data increasingly valuable.
 
@@
 For example, if aboveground biomass and total biomass are presented, but root biomass is not presented, root biomass cannot be subsequently calculated and sent to EFDB.
+Similarly, fine and coarse root biomass can't be summed; soil carbon can't be summed across depth increments, etc.
 

### 2022-01-28 doc/manuscript/ESSDmanuscript.Rmd
add a reference
paper edits

--- 2021-09-21 doc/manuscript/ESSDmanuscript.Rmd
@@
 
-**As we seek to mitigate climate change [@unfccc_adoption_2015], forests are critical.**
-*(See intros to other papers for relevant refs)*
-Under Paris Agreement, **##**% of Nationally Determined Contributions relate to forests [@grassi_key_2017].
+Forests are critical to our management of greenhouse gases. The influence the forests have on asmosphereic carbon has made forest data increasingly valuable.
 
@@
 For example, @cook-patton_mapping_2020 *(summarize IPCC comparison results)*. 
+Another example is @cuni-sanchez_high_2021.
 *Moreover, it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.*

### 2022-01-28 doc/manuscript/ESSDmanuscript.tex
paper edits

@@
 
-%% Journal abbreviations (please use the same for discussion papers and final revised papers)
+%% Journal abbreviations (please use the same for preprints and final revised papers)
 
@@
 % Archives Animal Breeding (aab)
-% ASTRA Proceedings (ap)
 % Atmospheric Chemistry and Physics (acp)
@@
 % Proceedings of the International Association of Hydrological Sciences (piahs)
+% Safety of Nuclear Waste Disposal (sand)
 % Scientific Drilling (sd)
@@
 % The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
+% 
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 %%\usepackage{booktabs}
@@
 %
-
 \begin{document}
 
-\title{Informing IPCC accounting of forest carbon using the global
-forest carbon database (ForC v4.0)}
+\title{Informing IPCC accounting of forest carbon using the global forest
+carbon database (ForC v4.0)}
 
@@
 
-\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation
-Biology Institute, Front Royal, VA, United States}
+\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation Biology
+Institute, Front Royal, VA, United States}
 \affil[2]{School of Geography, University of Leeds, Leeds, UK}
-\affil[3]{Joint Global Change Research Institute, Pacific Northwest
-National Laboratory, College Park, MD, United States}
+\affil[3]{Joint Global Change Research Institute, Pacific Northwest National
+Laboratory, College Park, MD, United States}
 \affil[4]{The Nature Conservancy; Arlington VA 22203, USA}
@@
 
-\copyrightstatement{The author's copyright for this publication is
-transferred to institution/company.}
+\copyrightstatement{The author's copyright for this publication is transferred to
+institution/company.}
 
@@
 
-\textbf{As we seek to mitigate climate change
-\citep{unfccc_adoption_2015}, forests are critical.} \emph{(See intros
-to other papers for relevant refs)} Under Paris Agreement,
-\textbf{\#\#}\% of Nationally Determined Contributions relate to forests
-\citep{grassi_key_2017}.
+Currently the world is in a climate crisis and as we seek to find
+answers climate data is increasingly important.Forests are critical to
+our management of greenhouse gases. The influence the forests have on
+asmosphereic carbon has made forest data increasingly valuable.
 
@@
 \citet{cook-patton_mapping_2020} \emph{(summarize IPCC comparison
-results)}. \emph{Moreover, it is useful for those compiling national
-greenhouse gas inventories to have access to locally-specific
-information, when available.} To improve the data accessible for C
-accounting, the IPCC created the Emission Factor Database (EFDB;
+results)}. Another example is \citet{cuni-sanchez_high_2021}.
+\emph{Moreover, it is useful for those compiling national greenhouse gas
+inventories to have access to locally-specific information, when
+available.} To improve the data accessible for C accounting, the IPCC
+created the Emission Factor Database (EFDB;
 https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a
@@
 \hline
-\textbf{pool} & \textbf{definition} & \textbf{major sources of estimate variation} & \textbf{IPCC guidance}\\
+\textbf{ï  pool} & \textbf{definition} & \textbf{major sources of estimate variation} & \textbf{IPCC guidance}\\
 \hline
@@
 \hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
+\textbf{ï  ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
 \hline
@@
 Mapping of variables is shown in Fig. 1
-
-Define relationship among NEE, NEP, and delta.C., especially noting role
-of harvest \citep{chapin_reconciling_2006}.
 
@@
   root biomass is not presented, root biomass cannot be subsequently
-  calculated and sent to EFDB.
+  calculated and sent to EFDB. Similarly, fine and coarse root biomass
+  can't be summed; soil carbon can't be summed across depth increments,
+  etc.
 \end{itemize}
@@
 
-\codedataavailability{use this to add a statement when having data sets
-and software code
+\codedataavailability{use this to add a statement when having data sets and software code
 available} %% use this section when having data sets and software code available
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\competinginterests{The authors declare no competing
-interests.} %% this section is mandatory even if you declare that no competing interests are present
+\competinginterests{The authors declare no competing interests.} %% this section is mandatory even if you declare that no competing interests are present
 

### 2022-02-17 doc/manuscript/ESSDmanuscript.Rmd
paper

@@
     corresponding: true
-
-#affiliation numbering will need to be fixed
-
 affiliation:
@@
     address: CIRAD, Montpellier, France
-
 abstract: |
@@
   author: Williams et al.
-# This section is mandatory even if you declare that no competing interests are present.
 competinginterests: |
@@
   The author's copyright for this publication is transferred to institution/company. 
-### The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
-### It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
 ### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
 availability:
+  #  use this to add a statement when having only software code available
+  #  use this to add a statement when having only data sets available
   #code: |
-  #  use this to add a statement when having only software code available
   #data: |
-  #  use this to add a statement when having only data sets available
   codedata: |
@@
 appendix: |
-
 output:

### 2022-02-17 doc/manuscript/ESSDmanuscript.tex
paper

@@
 
-\title{Informing IPCC accounting of forest carbon using the global forest
-carbon database (ForC v4.0)}
+\title{Informing IPCC accounting of forest carbon using the global
+forest carbon database (ForC v4.0)}
 
@@
 
-\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation Biology
-Institute, Front Royal, VA, United States}
+\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation
+Biology Institute, Front Royal, VA, United States}
 \affil[2]{School of Geography, University of Leeds, Leeds, UK}
-\affil[3]{Joint Global Change Research Institute, Pacific Northwest National
-Laboratory, College Park, MD, United States}
+\affil[3]{Joint Global Change Research Institute, Pacific Northwest
+National Laboratory, College Park, MD, United States}
 \affil[4]{The Nature Conservancy; Arlington VA 22203, USA}
@@
 
-\copyrightstatement{The author's copyright for this publication is transferred to
-institution/company.}
+\copyrightstatement{The author's copyright for this publication is
+transferred to institution/company.}
 
@@
 
-Currently the world is in a climate crisis and as we seek to find
-answers climate data is increasingly important.Forests are critical to
-our management of greenhouse gases. The influence the forests have on
-asmosphereic carbon has made forest data increasingly valuable.
+Forests are critical to our management of greenhouse gases. The
+influence the forests have on asmosphereic carbon has made forest data
+increasingly valuable.
 
@@
 
-Litter is defined as including " all non-living biomass with a diameter
+Litter is defined as including '' all non-living biomass with a diameter
 less than a minimum diameter chosen by the country (for example 10 cm),
@@
 belowground biomass) are included in litter where they cannot be
-distinguished from it empirically." (2003 IPCc GPG for LULUCF
+distinguished from it empirically.'' (2003 IPCc GPG for LULUCF
 (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
@@
 
-\codedataavailability{use this to add a statement when having data sets and software code
+\codedataavailability{use this to add a statement when having data sets
+and software code
 available} %% use this section when having data sets and software code available
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\competinginterests{The authors declare no competing interests.} %% this section is mandatory even if you declare that no competing interests are present
+\competinginterests{The authors declare no competing
+interests.} %% this section is mandatory even if you declare that no competing interests are present
 

### 2022-02-23 doc/manuscript/ESSDmanuscript.Rmd
worked on abstract & intro
@mawilliams99 , FYI:

- first pass at abstract
- added a bit of content to intro

@@
 abstract: |
-  The abstract goes here.
-  It can also be on _multiple lines_.
+  Forests are critical for climate change mitigation and consitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
+  Yet, the efficacy of greenhouse gas (GHG) mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
+  Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
+  Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
+  Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector GHG emissions and removals.
+  We begin by reconciling terminology and mapping ForC fields into EFDB.
+  This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
+  At the time of writing, ForC contained ## values that would qualify for inclusion in the EFDB, ## of which have been transferred to date.
+  (Some analysis of representation/ gaps)
+  <!-- -->
+  In the future, forest C estimates in EFDB can be improved through targetted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
+
 bibliography: references.bib
@@
 
-**Important**: Always double-check with the official manuscript preparation guidelines at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html), especially the sections "Technical instructions for LaTeX" and "Manuscript composition".
-Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.
+<!--**Important**: Always double-check with the official manuscript preparation guidelines at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html), especially the sections "Technical instructions for LaTeX" and "Manuscript composition".
+Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.-->
 
@@
 
-Forests are critical to our management of greenhouse gases. The influence the forests have on asmosphereic carbon has made forest data increasingly valuable.
+Forests are critical to our management of greenhouse gases. 
+The influence the forests have on atmospheric carbon has made forest data increasingly valuable.
+
+**Forests play a central role in international plans for climate change mitigation [@unfccc_adoption_2015].**
+As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
+Effective forest-based climate change mitigation requires solid scientific data [@anderson-teixeira_effective_2022].
 
@@
 To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
+
+**To ensure that planned emissions reductions are realistic, high-quality estimates of forest C stocks and fluxes must be publicly accessible.**
+The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished GHG reductions, or as a basis of comparison for external parties to evaluate these inventories.
+<!-- check Washington Post article on Malaysia NDC, @deng_comparing_2021 -->
 

### 2022-02-23 doc/manuscript/ESSDmanuscript.tex
worked on abstract & intro
@mawilliams99 , FYI:

- first pass at abstract
- added a bit of content to intro

--- 2021-09-21 doc/manuscript/ESSDmanuscript.tex
@@
 \begin{abstract}
-The abstract goes here. It can also be on \emph{multiple lines}.
+Forests are critical for climate change mitigation and consitute a
+substantial portion of planned emissions reductions under the 2015 Paris
+Agreement. Yet, the efficacy of greenhouse gas (GHG) mitigation planning
+and reporting is dependent upon the quality of available emission
+factors data, including forest carbon (C) stocks and changes therein.
+Tens of thousands of relevant forest C estimates have been published,
+yet are not readily accesible to the practitioners compiling national
+greenhouse gas inventories. Many of these data have, however, been
+compiled in the Global Forest C database (ForC;
+https://forc-db.github.io/) and stand to be of value to greenhouse gas
+accounting if made available through the Emission Factor Database (EFDB)
+of the International Panel on Climate Change (IPCC). Here, we develop
+and document a process for semi-automated transfer of data from ForC
+into the EFDB, assess the data available and transferred to date, and
+provide recommendations for improving forest data collection, analysis,
+and reporting to improve accounting of forest-sector GHG emissions and
+removals. We begin by reconciling terminology and mapping ForC fields
+into EFDB. This process required some updates to the ForC database
+structure, leading to the release of a new version of ForC (v4.0;
+described here). At the time of writing, ForC contained \#\# values that
+would qualify for inclusion in the EFDB, \#\# of which have been
+transferred to date. (Some analysis of representation/ gaps) In the
+future, forest C estimates in EFDB can be improved through targetted
+research to fill critical gaps, reporting of information required by
+IPCC, and continued submission of data from scientific publications to
+the EFDB.
 \end{abstract}
@@
 
-\textbf{Important}: Always double-check with the official manuscript
-preparation guidelines at
-\url{https://publications.copernicus.org/for_authors/manuscript_preparation.html},
-especially the sections ``Technical instructions for LaTeX'' and
-``Manuscript composition''. Please contact Daniel Nüst,
-\texttt{daniel.nuest@uni-muenster.de}, with any problems.
-
 \introduction[Introduction]
 
-\textbf{As we seek to mitigate climate change
-\citep{unfccc_adoption_2015}, forests are critical.} \emph{(See intros
-to other papers for relevant refs)} Under Paris Agreement,
-\textbf{\#\#}\% of Nationally Determined Contributions relate to forests
-\citep{grassi_key_2017}.
+Forests are critical to our management of greenhouse gases. The
+influence the forests have on atmospheric carbon has made forest data
+increasingly valuable.
+
+\textbf{Forests play a central role in international plans for climate
+change mitigation \citep{unfccc_adoption_2015}.} As of 2016,
+forest-based mitigation accounted for 26\% of total planned greenhouse
+gas mitigation within Nationally Determined Contributions under the
+Paris Agreement \citep{grassi_key_2017}. Effective forest-based climate
+change mitigation requires solid scientific data
+\citep{anderson-teixeira_effective_2022}.
 
@@
 used for estimating greenhouse gas emissions and removals.
+
+\textbf{To ensure that planned emissions reductions are realistic,
+high-quality estimates of forest C stocks and fluxes must be publicly
+accessible.} The EFDB is intended as a recognized library, and can be
+used both for efforts to tally a nation's intended or accomplished GHG
+reductions, or as a basis of comparison for external parties to evaluate
+these inventories.
 
@@
 
-Define relationship among NEE, NEP, and delta.C., especially noting role
-of harvest \citep{chapin_reconciling_2006}.
-
 \subsection{Land use categories}

### 2022-02-23 doc/manuscript/ESSDmanuscript.Rmd
minor formatting

@@
   At the time of writing, ForC contained ## values that would qualify for inclusion in the EFDB, ## of which have been transferred to date.
-  (Some analysis of representation/ gaps)
+  (Some analysis of representation/ gaps.)
   <!-- -->
@@
 
-The conclusion goes here.
-You can modify the section name with  `\conclusions[modified heading if necessary]`.
+<!--The conclusion goes here.
+You can modify the section name with  `\conclusions[modified heading if necessary]`.-->

### 2022-02-23 doc/manuscript/ESSDmanuscript.tex
minor formatting

@@
 would qualify for inclusion in the EFDB, \#\# of which have been
-transferred to date. (Some analysis of representation/ gaps) In the
+transferred to date. (Some analysis of representation/ gaps.) In the
 future, forest C estimates in EFDB can be improved through targetted
@@
 
-The conclusion goes here. You can modify the section name with
-\texttt{\textbackslash{}conclusions{[}modified\ heading\ if\ necessary{]}}.
+\newpage
 

### 2022-02-24 doc/manuscript/ESSDmanuscript.Rmd
added appendix content
... but so far I don't know how to put in the tables

@@
 appendix: |
+ \clearpage 
+ \section{Instructions (TEMPORARY)}
+  Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
+  \subsection{Option 1}
+
+  If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
+  They will be correctly named automatically.
+  \subsection{Option 2}
+  If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
+
+  To rename them correctly to A1, A2, etc., please add the following commands in front of them:
+  `\appendixfigures` needs to be added in front of appendix figures
+  `\appendixtables` needs to be added in front of appendix tables
+
+  Please add `\clearpage` between each table and/or figure. Further guidelines on figures and tables can be found below.
+
+   \clearpage
+   \section{Mapping ForC to EFDB}
+
+    CURRENT TABLE 2 GOES HERE
+
+   \clearpage
+   \section{Updates to ForC}
+
+   CURRENT TABLE 2 GOES HERE
+
 output:

### 2022-02-24 doc/manuscript/ESSDmanuscript.tex
added appendix content
... but so far I don't know how to put in the tables

@@
 
-\newpage
-
 \codedataavailability{use this to add a statement when having data sets
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+\clearpage 
+\section{Instructions (TEMPORARY)}
+
+Regarding figures and tables in appendices, the following two options
+are possible depending on your general handling of figures and tables in
+the manuscript environment:
+
+\subsection{Option 1}
+
+If you sorted all figures and tables into the sections of the text,
+please also sort the appendix figures and appendix tables into the
+respective appendix sections. They will be correctly named
+automatically.
+
+\subsection{Option 2}
+
+If you put all figures after the reference list, please insert appendix
+tables and figures after the normal tables and figures.
+
+To rename them correctly to A1, A2, etc., please add the following
+commands in front of them: \texttt{\textbackslash{}appendixfigures}
+needs to be added in front of appendix figures
+\texttt{\textbackslash{}appendixtables} needs to be added in front of
+appendix tables
+
+Please add \texttt{\textbackslash{}clearpage} between each table and/or
+figure. Further guidelines on figures and tables can be found below.
+
+\clearpage
+
+\section{Mapping ForC to EFDB}
+
+CURRENT TABLE 2 GOES HERE
+
+\clearpage
+
+\section{Updates to ForC}
+
+CURRENT TABLE 2 GOES HERE
+\noappendix
 

### 2022-02-24 doc/manuscript/ESSDmanuscript.Rmd
temporary solution for Appendix tables
moved big tables to the end so that they're out of the way, but not yet following the template/ does not get the numbering right

--- 2022-02-23 doc/manuscript/ESSDmanuscript.Rmd
@@
 appendix: |
+   \section{Mapping ForC to EFDB}
+
+   CURRENT TABLE 2 GOES HERE
+
+   \section{Updates to ForC}
+
+   CURRENT TABLE 3 GOES HERE
+
 output:
@@
 
-\clearpage
-```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-
-ForC_variables_included <- read.csv("figures_tables/ForC_variables_included.csv")
-names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
-ForC_variables_included <- ForC_variables_included[c(1:5)] #include all fields but TO_DO
-
-knitr::kable(ForC_variables_included, format = "latex",
-             longtable = TRUE,
-             caption = "\\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). ") %>%
-  row_spec(0, bold = TRUE) %>%
-kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
-
-```
-'Required' field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
-
-** '(yes)' for most recent severe disturbance; 'no' for other history events
-
 ## Carbon cycle variables
@@
 ## ForC restructuring
-
-```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
-library (readr)
-urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
-record_of_changes <-read_csv(url(urlfile))
-names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
-record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
-#record_of_changes<- record_of_changes[which(!grep("housekeeping",record_of_changes$Motivation)) , c(1:5)]  #trying to exclude rows for which Motivation = "housekeeping", but this is not working
-record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
-
-library(knitr)
-library(kableExtra)
-knitr::kable(record_of_changes, format = "latex",
-             longtable = TRUE,
-             caption = "\\textbf{Table of changes to ForC fields.}") %>%
-  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
-  row_spec(0, bold = TRUE)
-```
 
@@
 
-\conclusions[Conclusions]
+# Conclusions
 
@@
 You can modify the section name with  `\conclusions[modified heading if necessary]`.-->
+
+\clearpage
+# Appendix A 
+<!--remove number-->
+
+```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+
+ForC_variables_included <- read.csv("figures_tables/ForC_variables_included.csv")
+names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
+ForC_variables_included <- ForC_variables_included[c(1:5)] #include all fields but TO_DO
+
+knitr::kable(ForC_variables_included, format = "latex",
+             longtable = TRUE,
+             caption = "\\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). ") %>%
+  row_spec(0, bold = TRUE) %>%
+kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
+
+```
+'Required' field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
+
+** '(yes)' for most recent severe disturbance; 'no' for other history events
+
+\clearpage
+# Appendix B
+<!--remove number-->
+
+```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
+library (readr)
+urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
+record_of_changes <-read_csv(url(urlfile))
+names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
+record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
+#record_of_changes<- record_of_changes[which(!grep("housekeeping",record_of_changes$Motivation)) , c(1:5)]  #trying to exclude rows for which Motivation = "housekeeping", but this is not working
+record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
+
+library(knitr)
+library(kableExtra)
+knitr::kable(record_of_changes, format = "latex",
+             longtable = TRUE,
+             caption = "\\textbf{Table of changes to ForC fields.}") %>%
+  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
+  row_spec(0, bold = TRUE)
+```

### 2022-03-08 doc/manuscript/ESSDmanuscript.tex
temporary solution for Appendix tables
moved big tables to the end so that they're out of the way, but not yet following the template/ does not get the numbering right
manuscript

--- 2022-02-23 doc/manuscript/ESSDmanuscript.tex
@@
 % E&G Quaternary Science Journal (egqsj)
+% EGUsphere (egusphere) | This is only for EGUsphere preprints submitted without relation to an EGU journal.
 % European Journal of Mineralogy (ejm)
@@
 
-%% Please DO NOT add additional packages or LaTeX commands to the template. They
-%% are not supported by Coperncius. LaTeX packages already
-%% included in the copernicus.cls are:
-%\usepackage[german, english]{babel}
-%\usepackage{tabularx}
-%\usepackage{cancel}
-%\usepackage{multirow}
-%\usepackage{supertabular}
-%\usepackage{algorithmic}
-%\usepackage{algorithm}
-%\usepackage{amsthm}
-%\usepackage{float}
-%\usepackage{subfig}
-%\usepackage{rotating}
-
 % Pandoc citation processing
@@
 %
-
 \begin{document}
@@
 \section{Mapping ForC to EFDB}
-
-\clearpage
-
-\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). }\\
-\hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
-\hline
-\endfirsthead
-\caption[]{\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted).  \textit{(continued)}}\\
-\hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
-\hline
-\endhead
-Measurements & measurement.ID & Other Properties & direct mapping & (no)\\
-\hline
- & dominant.life.form & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & yes\\
-\hline
- & stand.age & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Parameters/ Conditions & used to determine land subcategories (see defining\_land\_subcategory.md), directly listed in Parameters/ Conditions & (yes)\\
-\hline
- & dominant.veg, veg.notes, min.dbh & Parameters/ Conditions & direct mapping/ linking to dominant.veg description & no\\
-\hline
- & variable.name & - & link to variable info in ForC\_variables table & yes\\
-\hline
- & date / start.date, end.date & Other Properties & direct mapping & no\\
-\hline
- & mean & Value & direct mapping & yes\\
-\hline
- & mean.in.original.units & Value in Common Units & direct mapping & yes\\
-\hline
- & original.units & Common Unit & direct mapping & yes\\
-\hline
- & lower95\%CI, upper 95\%CI, se, sd and n & Lower Confidence Limit, Upper Confidence Limit & direct or calculated & (yes)\\
-\hline
- & depth, covariate\_1, cov\_1.value, covariate\_2, cov\_2.value & Other Properties & direct mapping & no\\
-\hline
- & allometry\_1, allometry\_2 & Comments from Data Provider & link to biomass allometry source, when provided & no\\
-\hline
- & data.location.within.source & - & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
-\hline
- & ForC.investigator & Data Provider, Data Provider Contact & link to Data Provider, Data Provider Contact info & yes\\
-\hline
-Sites & site.ID, sites.sitename & Other Properties & direct mapping & (no)\\
-\hline
- & lat, lon & Region/Regional conditions & direct mapping; used to extract continent, Koeppen, and FAO.ecozone & (no)\\
-\hline
- & country, state, city, masl,  mat, map & Region/Regional conditions & direct mapping & no\\
-\hline
- & continent, Koeppen & Region/Regional conditions & direct mapping & auto\\
-\hline
- & soil.texture, sand, silt, clay, soil.classification & Parameters/ Conditions & direct mapping & no\\
-\hline
- & FAO.ecozone & Parameters/ Conditions & direct mapping & auto\\
-\hline
-History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)/no**\\
-\hline
- & plot.area & Other Properties & direct mapping & no\\
-\hline
-Plots & plot.ID, plot.name & Other Properties & direct mapping & (no)\\
-\hline
- & distmrs.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & auto\\
-\hline
- & distmrs.type, distmrs.year, regrowth.type, regrowth.year & Other Properties & direct mapping & auto\\
-\hline
-PFT & description & Parameters/ Conditions & direct mapping & auto\\
-\hline
-variables & variable.type & Gases & For stocks in unit of organic matter, gases include CO2, CO, CH4, NO, NO2, N2O. For increments, fluxes, and stocks in units of C, gases includes only CO2. & auto\\
-\hline
- & variable.name & C pool, Equation & link to C pool, Equation & auto\\
-\hline
- & description & Description & direct mapping & auto\\
-\hline
- & extended.description & Other Properties & direct mapping & auto\\
-\hline
- & units & Unit (ID) & link to IPCC units & auto\\
-\hline
-Citations & citation.citation & Full Technical Reference & direct mapping & yes/auto\\
-\hline
- & citation.language & Reference Language & direct mapping & yes/auto\\
-\hline
- & citation.url & URL & direct mapping & no/auto\\
-\hline
- & citation.abstract & Abstract in English & direct mapping & no/auto\\
-\hline
- & source.type & Source of Data & direct mapping & yes\\
-\hline
-\end{longtabu}
-
-`Required' field indicates whether the field is required by EFDB: yes =
-value required; (yes) = input required, missing value acceptable if not
-reported; auto = present within ForC infrasructure, and therefore will
-always be exported to EFDB ; (no) = not required for EFDB, but required
-for ForC and therefore will always be exported to EFDB; no = not
-required, but exported to EFDB when a value is present.
-
-** `(yes)' for most recent severe disturbance; `no' for other history
-events
 
@@
 \subsection{ForC restructuring}
-
-\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\caption{\label{tab:table_ForCchanges}\textbf{Table of changes to ForC fields.}}\\
-\hline
-\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
-\hline
-\endfirsthead
-\caption[]{\textbf{Table of changes to ForC fields.} \textit{(continued)}}\\
-\hline
-\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
-\hline
-\endhead
-Sites & coordinates.precision & Precision of geographic coordinates, as reported by source or estimated from maps. & field added & allow identification of records with poor coordinate precision\\
-\hline
-Measurements & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
-\hline
- & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
-& replaces `stat` and `stat.name` & cleaner format; ability to handle assymetrical 95 percent confidence intervals\\
-\hline
- & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
-\hline
- & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
-& field added & track units conversion, allow back-calculation of OM if conversion factor deemed inappropriate\\
-\hline
-PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
-\hline
- & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
-\hline
-Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC\\
-\hline
- & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
-\hline
- & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
-\hline
- & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
-\hline
- & source.type & citation source type & field added & field required by IPCC\\
-\hline
- & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
-\hline
- & EFDB.ready & Indicates whether data have been checked for export to EFDB. & field added & housekeeping\\
-\hline
-\end{longtabu}
 
@@
 
-\conclusions[Conclusions]
+\section{Conclusions}
 
-\newpage
+\clearpage
+
+\section{Appendix A}
+
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). }\\
+\hline
+\textbf{ï  ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
+\hline
+\endfirsthead
+\caption[]{\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted).  \textit{(continued)}}\\
+\hline
+\textbf{ï  ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
+\hline
+\endhead
+Measurements & measurement.ID & Other Properties & direct mapping & (no)\\
+\hline
+ & dominant.life.form & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & yes\\
+\hline
+ & stand.age & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Parameters/ Conditions & used to determine land subcategories (see defining\_land\_subcategory.md), directly listed in Parameters/ Conditions & (yes)\\
+\hline
+ & dominant.veg, veg.notes, min.dbh & Parameters/ Conditions & direct mapping/ linking to dominant.veg description & no\\
+\hline
+ & variable.name & - & link to variable info in ForC\_variables table & yes\\
+\hline
+ & date / start.date, end.date & Other Properties & direct mapping & no\\
+\hline
+ & mean & Value & direct mapping & yes\\
+\hline
+ & mean.in.original.units & Value in Common Units & direct mapping & yes\\
+\hline
+ & original.units & Common Unit & direct mapping & yes\\
+\hline
+ & lower95\%CI, upper 95\%CI, se, sd and n & Lower Confidence Limit, Upper Confidence Limit & direct or calculated & (yes)\\
+\hline
+ & depth, covariate\_1, cov\_1.value, covariate\_2, cov\_2.value & Other Properties & direct mapping & no\\
+\hline
+ & allometry\_1, allometry\_2 & Comments from Data Provider & link to biomass allometry source, when provided & no\\
+\hline
+ & data.location.within.source & - & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
+\hline
+ & ForC.investigator & Data Provider, Data Provider Contact & link to Data Provider, Data Provider Contact info & yes\\
+\hline
+Sites & site.ID, sites.sitename & Other Properties & direct mapping & (no)\\
+\hline
+ & lat, lon & Region/Regional conditions & direct mapping; used to extract continent, Koeppen, and FAO.ecozone & (no)\\
+\hline
+ & country, state, city, masl,  mat, map & Region/Regional conditions & direct mapping & no\\
+\hline
+ & continent, Koeppen & Region/Regional conditions & direct mapping & auto\\
+\hline
+ & soil.texture, sand, silt, clay, soil.classification & Parameters/ Conditions & direct mapping & no\\
+\hline
+ & FAO.ecozone & Parameters/ Conditions & direct mapping & auto\\
+\hline
+History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)/no**\\
+\hline
+ & plot.area & Other Properties & direct mapping & no\\
+\hline
+Plots & plot.ID, plot.name & Other Properties & direct mapping & (no)\\
+\hline
+ & distmrs.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & auto\\
+\hline
+ & distmrs.type, distmrs.year, regrowth.type, regrowth.year & Other Properties & direct mapping & auto\\
+\hline
+PFT & description & Parameters/ Conditions & direct mapping & auto\\
+\hline
+variables & variable.type & Gases & For stocks in unit of organic matter, gases include CO2, CO, CH4, NO, NO2, N2O. For increments, fluxes, and stocks in units of C, gases includes only CO2. & auto\\
+\hline
+ & variable.name & C pool, Equation & link to C pool, Equation & auto\\
+\hline
+ & description & Description & direct mapping & auto\\
+\hline
+ & extended.description & Other Properties & direct mapping & auto\\
+\hline
+ & units & Unit (ID) & link to IPCC units & auto\\
+\hline
+Citations & citation.citation & Full Technical Reference & direct mapping & yes/auto\\
+\hline
+ & citation.language & Reference Language & direct mapping & yes/auto\\
+\hline
+ & citation.url & URL & direct mapping & no/auto\\
+\hline
+ & citation.abstract & Abstract in English & direct mapping & no/auto\\
+\hline
+ & source.type & Source of Data & direct mapping & yes\\
+\hline
+\end{longtabu}
+
+`Required' field indicates whether the field is required by EFDB: yes =
+value required; (yes) = input required, missing value acceptable if not
+reported; auto = present within ForC infrasructure, and therefore will
+always be exported to EFDB ; (no) = not required for EFDB, but required
+for ForC and therefore will always be exported to EFDB; no = not
+required, but exported to EFDB when a value is present.
+
+** `(yes)' for most recent severe disturbance; `no' for other history
+events
+
+\clearpage
+
+\section{Appendix B}
+
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\caption{\label{tab:table_ForCchanges}\textbf{Table of changes to ForC fields.}}\\
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+\endfirsthead
+\caption[]{\textbf{Table of changes to ForC fields.} \textit{(continued)}}\\
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+\endhead
+Sites & coordinates.precision & Precision of geographic coordinates, as reported by source or estimated from maps. & field added & allow identification of records with poor coordinate precision\\
+\hline
+Measurements & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
+\hline
+ & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
+& replaces `stat` and `stat.name` & cleaner format; ability to handle assymetrical 95 percent confidence intervals\\
+\hline
+ & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
+\hline
+ & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
+& field added & track units conversion, allow back-calculation of OM if conversion factor deemed inappropriate\\
+\hline
+PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
+\hline
+ & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
+\hline
+Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC\\
+\hline
+ & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
+\hline
+ & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
+\hline
+ & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
+\hline
+ & source.type & citation source type & field added & field required by IPCC\\
+\hline
+ & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
+\hline
+ & EFDB.ready & Indicates whether data have been checked for export to EFDB. & field added & housekeeping\\
+\hline
+\end{longtabu}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+\section{Mapping ForC to EFDB}
+
+CURRENT TABLE 2 GOES HERE
+
+\section{Updates to ForC}
+
+CURRENT TABLE 3 GOES HERE
+\noappendix
 

### 2022-03-14 doc/manuscript/ESSDmanuscript.Rmd
work
revert commit
manuscript
Update ESSDmanuscript.Rmd

@@
 ## Carbon stocks
-
 Forest ecosystem C stocks may be parsed into pools in various ways. 
@@
 
-Soil organic matter is defined as...
+Soil organic matter is defined as "Includes organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and
+applied consistently through the time series.
+Live fine roots (of less than the suggested diameter limit for belowground biomass) are included with soil organic matter where they cannot be distinguished from it empirically."(2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
 

### 2022-03-14 doc/manuscript/ESSDmanuscript.Rmd
manuscript
finally got it to knit!

@@
 # Mapping ForC to EFDB
+ForC data is incredibly valuable to EFDB and there is data which is included in the ForC database that does not meet EFDB standards.
+There were two main EFDB guidelines which limits the amount of data we could transfer.
+EFDB will not accept data which has been digitized(from graph) and ForC does.
 

### 2022-03-14 doc/manuscript/ESSDmanuscript.tex
trying 2 knit
manuscript
finally got it to knit!

@@
 
-Soil organic matter is defined as\ldots{}
+Soil organic matter is defined as ``Includes organic carbon in mineral
+and organic soils (including peat) to a specified depth chosen by the
+country and applied consistently through the time series. Live fine
+roots (of less than the suggested diameter limit for belowground
+biomass) are included with soil organic matter where they cannot be
+distinguished from it empirically.''(2003 IPCc GPG for LULUCF
+(https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
 
@@
 \section{Mapping ForC to EFDB}
+
+ForC data is incredibly valuable to EFDB and there is data which is
+included in the ForC database that does not meet EFDB standards. There
+were two main EFDB guidelines which limits the amount of data we could
+transfer. EFDB will not accept data which has been digitized(from graph)
+and ForC does.
 

### 2022-03-18 doc/manuscript/ESSDmanuscript.tex
knit on my laptop
@mawilliams99 , this is great that you got it to knit! I noticed that the tables at the end didn't knit properly, though. I have no idea what's going on, but pushing my version for you to see the difference

--- 2022-02-24 doc/manuscript/ESSDmanuscript.tex
@@
 
-\clearpage
-
-\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). }\\
-\hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
-\hline
-\endfirsthead
-\caption[]{\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted).  \textit{(continued)}}\\
-\hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
-\hline
-\endhead
-Measurements & measurement.ID & Other Properties & direct mapping & (no)\\
-\hline
- & dominant.life.form & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & yes\\
-\hline
- & stand.age & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Parameters/ Conditions & used to determine land subcategories (see defining\_land\_subcategory.md), directly listed in Parameters/ Conditions & (yes)\\
-\hline
- & dominant.veg, veg.notes, min.dbh & Parameters/ Conditions & direct mapping/ linking to dominant.veg description & no\\
-\hline
- & variable.name & - & link to variable info in ForC\_variables table & yes\\
-\hline
- & date / start.date, end.date & Other Properties & direct mapping & no\\
-\hline
- & mean & Value & direct mapping & yes\\
-\hline
- & mean.in.original.units & Value in Common Units & direct mapping & yes\\
-\hline
- & original.units & Common Unit & direct mapping & yes\\
-\hline
- & lower95\%CI, upper 95\%CI, se, sd and n & Lower Confidence Limit, Upper Confidence Limit & direct or calculated & (yes)\\
-\hline
- & depth, covariate\_1, cov\_1.value, covariate\_2, cov\_2.value & Other Properties & direct mapping & no\\
-\hline
- & allometry\_1, allometry\_2 & Comments from Data Provider & link to biomass allometry source, when provided & no\\
-\hline
- & data.location.within.source & - & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
-\hline
- & ForC.investigator & Data Provider, Data Provider Contact & link to Data Provider, Data Provider Contact info & yes\\
-\hline
-Sites & site.ID, sites.sitename & Other Properties & direct mapping & (no)\\
-\hline
- & lat, lon & Region/Regional conditions & direct mapping; used to extract continent, Koeppen, and FAO.ecozone & (no)\\
-\hline
- & country, state, city, masl,  mat, map & Region/Regional conditions & direct mapping & no\\
-\hline
- & continent, Koeppen & Region/Regional conditions & direct mapping & auto\\
-\hline
- & soil.texture, sand, silt, clay, soil.classification & Parameters/ Conditions & direct mapping & no\\
-\hline
- & FAO.ecozone & Parameters/ Conditions & direct mapping & auto\\
-\hline
-History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)/no**\\
-\hline
- & plot.area & Other Properties & direct mapping & no\\
-\hline
-Plots & plot.ID, plot.name & Other Properties & direct mapping & (no)\\
-\hline
- & distmrs.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & auto\\
-\hline
- & distmrs.type, distmrs.year, regrowth.type, regrowth.year & Other Properties & direct mapping & auto\\
-\hline
-PFT & description & Parameters/ Conditions & direct mapping & auto\\
-\hline
-variables & variable.type & Gases & For stocks in unit of organic matter, gases include CO2, CO, CH4, NO, NO2, N2O. For increments, fluxes, and stocks in units of C, gases includes only CO2. & auto\\
-\hline
- & variable.name & C pool, Equation & link to C pool, Equation & auto\\
-\hline
- & description & Description & direct mapping & auto\\
-\hline
- & extended.description & Other Properties & direct mapping & auto\\
-\hline
- & units & Unit (ID) & link to IPCC units & auto\\
-\hline
-Citations & citation.citation & Full Technical Reference & direct mapping & yes/auto\\
-\hline
- & citation.language & Reference Language & direct mapping & yes/auto\\
-\hline
- & citation.url & URL & direct mapping & no/auto\\
-\hline
- & citation.abstract & Abstract in English & direct mapping & no/auto\\
-\hline
- & source.type & Source of Data & direct mapping & yes\\
-\hline
-\end{longtabu}
-
-`Required' field indicates whether the field is required by EFDB: yes =
-value required; (yes) = input required, missing value acceptable if not
-reported; auto = present within ForC infrasructure, and therefore will
-always be exported to EFDB ; (no) = not required for EFDB, but required
-for ForC and therefore will always be exported to EFDB; no = not
-required, but exported to EFDB when a value is present.
-
-** `(yes)' for most recent severe disturbance; `no' for other history
-events
+ForC data is incredibly valuable to EFDB and there is data which is
+included in the ForC database that does not meet EFDB standards. There
+were two main EFDB guidelines which limits the amount of data we could
+transfer. EFDB will not accept data which has been digitized(from graph)
+and ForC does.
 
@@
 
-\conclusions[Conclusions]
+\section{Conclusions}
+
+\clearpage
+
+\section{Appendix A}
+
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). }\\
+\hline
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
+\hline
+\endfirsthead
+\caption[]{\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted).  \textit{(continued)}}\\
+\hline
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
+\hline
+\endhead
+Measurements & measurement.ID & Other Properties & direct mapping & (no)\\
+\hline
+ & dominant.life.form & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & yes\\
+\hline
+ & stand.age & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Parameters/ Conditions & used to determine land subcategories (see defining\_land\_subcategory.md), directly listed in Parameters/ Conditions & (yes)\\
+\hline
+ & dominant.veg, veg.notes, min.dbh & Parameters/ Conditions & direct mapping/ linking to dominant.veg description & no\\
+\hline
+ & variable.name & - & link to variable info in ForC\_variables table & yes\\
+\hline
+ & date / start.date, end.date & Other Properties & direct mapping & no\\
+\hline
+ & mean & Value & direct mapping & yes\\
+\hline
+ & mean.in.original.units & Value in Common Units & direct mapping & yes\\
+\hline
+ & original.units & Common Unit & direct mapping & yes\\
+\hline
+ & lower95\%CI, upper 95\%CI, se, sd and n & Lower Confidence Limit, Upper Confidence Limit & direct or calculated & (yes)\\
+\hline
+ & depth, covariate\_1, cov\_1.value, covariate\_2, cov\_2.value & Other Properties & direct mapping & no\\
+\hline
+ & allometry\_1, allometry\_2 & Comments from Data Provider & link to biomass allometry source, when provided & no\\
+\hline
+ & data.location.within.source & - & confirm that data weren't digitized, facilitate finding data in original publication & yes\\
+\hline
+ & ForC.investigator & Data Provider, Data Provider Contact & link to Data Provider, Data Provider Contact info & yes\\
+\hline
+Sites & site.ID, sites.sitename & Other Properties & direct mapping & (no)\\
+\hline
+ & lat, lon & Region/Regional conditions & direct mapping; used to extract continent, Koeppen, and FAO.ecozone & (no)\\
+\hline
+ & country, state, city, masl,  mat, map & Region/Regional conditions & direct mapping & no\\
+\hline
+ & continent, Koeppen & Region/Regional conditions & direct mapping & auto\\
+\hline
+ & soil.texture, sand, silt, clay, soil.classification & Parameters/ Conditions & direct mapping & no\\
+\hline
+ & FAO.ecozone & Parameters/ Conditions & direct mapping & auto\\
+\hline
+History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)/no**\\
+\hline
+ & plot.area & Other Properties & direct mapping & no\\
+\hline
+Plots & plot.ID, plot.name & Other Properties & direct mapping & (no)\\
+\hline
+ & distmrs.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories & used to determine land subcategories (see defining\_land\_subcategory.md) & auto\\
+\hline
+ & distmrs.type, distmrs.year, regrowth.type, regrowth.year & Other Properties & direct mapping & auto\\
+\hline
+PFT & description & Parameters/ Conditions & direct mapping & auto\\
+\hline
+variables & variable.type & Gases & For stocks in unit of organic matter, gases include CO2, CO, CH4, NO, NO2, N2O. For increments, fluxes, and stocks in units of C, gases includes only CO2. & auto\\
+\hline
+ & variable.name & C pool, Equation & link to C pool, Equation & auto\\
+\hline
+ & description & Description & direct mapping & auto\\
+\hline
+ & extended.description & Other Properties & direct mapping & auto\\
+\hline
+ & units & Unit (ID) & link to IPCC units & auto\\
+\hline
+Citations & citation.citation & Full Technical Reference & direct mapping & yes/auto\\
+\hline
+ & citation.language & Reference Language & direct mapping & yes/auto\\
+\hline
+ & citation.url & URL & direct mapping & no/auto\\
+\hline
+ & citation.abstract & Abstract in English & direct mapping & no/auto\\
+\hline
+ & source.type & Source of Data & direct mapping & yes\\
+\hline
+\end{longtabu}
+
+`Required' field indicates whether the field is required by EFDB: yes =
+value required; (yes) = input required, missing value acceptable if not
+reported; auto = present within ForC infrasructure, and therefore will
+always be exported to EFDB ; (no) = not required for EFDB, but required
+for ForC and therefore will always be exported to EFDB; no = not
+required, but exported to EFDB when a value is present.
+
+** `(yes)' for most recent severe disturbance; `no' for other history
+events
+
+\clearpage
+
+\section{Appendix B}
+
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\caption{\label{tab:table_ForCchanges}\textbf{Table of changes to ForC fields.}}\\
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+\endfirsthead
+\caption[]{\textbf{Table of changes to ForC fields.} \textit{(continued)}}\\
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+\endhead
+Sites & coordinates.precision & Precision of geographic coordinates, as reported by source or estimated from maps. & field added & allow identification of records with poor coordinate precision\\
+\hline
+Measurements & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
+\hline
+ & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
+& replaces `stat` and `stat.name` & cleaner format; ability to handle assymetrical 95 percent confidence intervals\\
+\hline
+ & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
+\hline
+ & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
+& field added & track units conversion, allow back-calculation of OM if conversion factor deemed inappropriate\\
+\hline
+PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
+\hline
+ & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
+\hline
+Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC\\
+\hline
+ & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
+\hline
+ & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
+\hline
+ & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
+\hline
+ & source.type & citation source type & field added & field required by IPCC\\
+\hline
+ & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
+\hline
+ & EFDB.ready & Indicates whether data have been checked for export to EFDB. & field added & housekeeping\\
+\hline
+\end{longtabu}
 
@@
 \appendix
-\clearpage 
-\section{Instructions (TEMPORARY)}
-
-Regarding figures and tables in appendices, the following two options
-are possible depending on your general handling of figures and tables in
-the manuscript environment:
-
-\subsection{Option 1}
-
-If you sorted all figures and tables into the sections of the text,
-please also sort the appendix figures and appendix tables into the
-respective appendix sections. They will be correctly named
-automatically.
-
-\subsection{Option 2}
-
-If you put all figures after the reference list, please insert appendix
-tables and figures after the normal tables and figures.
-
-To rename them correctly to A1, A2, etc., please add the following
-commands in front of them: \texttt{\textbackslash{}appendixfigures}
-needs to be added in front of appendix figures
-\texttt{\textbackslash{}appendixtables} needs to be added in front of
-appendix tables
-
-Please add \texttt{\textbackslash{}clearpage} between each table and/or
-figure. Further guidelines on figures and tables can be found below.
-
-\clearpage
-
 \section{Mapping ForC to EFDB}
@@
 
-\clearpage
-
 \section{Updates to ForC}
 
-CURRENT TABLE 2 GOES HERE
+CURRENT TABLE 3 GOES HERE
 \noappendix

### 2022-04-11 doc/manuscript/ESSDmanuscript.Rmd
work on intro
@mawilliams99 , FYI, I'm working on this today.

@@
   Forests are critical for climate change mitigation and consitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
-  Yet, the efficacy of greenhouse gas (GHG) mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
+  Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
   Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
   Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-  Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector GHG emissions and removals.
+  Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
   We begin by reconciling terminology and mapping ForC fields into EFDB.
@@
 
-Forests are critical to our management of greenhouse gases. 
-The influence the forests have on atmospheric carbon has made forest data increasingly valuable.
+Forests are critical to management of atmospheric concentrations of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
+In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
+Table 2--> [forests only, @harris_global_2021].
+This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
+Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
+Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
+Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@refs], and on human conservation, restoration, and management of forests [@refs].
 
-**Forests play a central role in international plans for climate change mitigation [@unfccc_adoption_2015].**
+Reflecting their strong influence on Earth's climate, forests play a central role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
+Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options, with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
 As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
-Effective forest-based climate change mitigation requires solid scientific data [@anderson-teixeira_effective_2022].
+Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [@refs].
+One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning, underlain by solid scientific data [@anderson-teixeira_effective_2022].
 
-**The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas (GHG) inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, (REFS for older guidelines), @ipcc_2019_2019].**
-Under this guidance, GHG inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
-The IPCC inventory guidelines include specific instructions for accounting for forest land ... [@ipcc_agriculture_2006; @ipcc_2019_2019a].
-This guidance has improved as more of the relevant underlying data has become available [@requenasuarez_estimating_2019], but there remains room for continuous improvement as the science advances.
-For example, @cook-patton_mapping_2020 *(summarize IPCC comparison results)*. 
-Another example is @cuni-sanchez_high_2021.
-*Moreover, it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.*
+The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2019_2019].
+Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
+The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
+This guidance has improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
+For example, the year following the release of the latest IPCC guidelines, @cook-patton_mapping_2020 found that the latest default rates may underestimate rates of C accumulation in regrowth forests by 32% on average and fail to capture eight-fold variation within ecozones.
+In addition, @cuni-sanchez_high_2021 found that aboveground C stocks in mature African tropical montane forests were two-thirds higher than the IPCC default values for these forests.
+This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
+Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @refs] implies that it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.
 To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
@@
 **To ensure that planned emissions reductions are realistic, high-quality estimates of forest C stocks and fluxes must be publicly accessible.**
-The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished GHG reductions, or as a basis of comparison for external parties to evaluate these inventories.
+The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories.
 <!-- check Washington Post article on Malaysia NDC, @deng_comparing_2021 -->
@@
 
-The UNFCCC requires GHG reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*].
+The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*].
 This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 

### 2022-04-11 doc/manuscript/ESSDmanuscript.tex
work on intro
@mawilliams99 , FYI, I'm working on this today.

@@
 substantial portion of planned emissions reductions under the 2015 Paris
-Agreement. Yet, the efficacy of greenhouse gas (GHG) mitigation planning
-and reporting is dependent upon the quality of available emission
-factors data, including forest carbon (C) stocks and changes therein.
-Tens of thousands of relevant forest C estimates have been published,
-yet are not readily accesible to the practitioners compiling national
-greenhouse gas inventories. Many of these data have, however, been
-compiled in the Global Forest C database (ForC;
-https://forc-db.github.io/) and stand to be of value to greenhouse gas
-accounting if made available through the Emission Factor Database (EFDB)
-of the International Panel on Climate Change (IPCC). Here, we develop
-and document a process for semi-automated transfer of data from ForC
-into the EFDB, assess the data available and transferred to date, and
-provide recommendations for improving forest data collection, analysis,
-and reporting to improve accounting of forest-sector GHG emissions and
-removals. We begin by reconciling terminology and mapping ForC fields
-into EFDB. This process required some updates to the ForC database
-structure, leading to the release of a new version of ForC (v4.0;
-described here). At the time of writing, ForC contained \#\# values that
-would qualify for inclusion in the EFDB, \#\# of which have been
-transferred to date. (Some analysis of representation/ gaps.) In the
-future, forest C estimates in EFDB can be improved through targetted
-research to fill critical gaps, reporting of information required by
-IPCC, and continued submission of data from scientific publications to
-the EFDB.
+Agreement. Yet, the efficacy of greenhouse gas mitigation planning and
+reporting is dependent upon the quality of available emission factors
+data, including forest carbon (C) stocks and changes therein. Tens of
+thousands of relevant forest C estimates have been published, yet are
+not readily accesible to the practitioners compiling national greenhouse
+gas inventories. Many of these data have, however, been compiled in the
+Global Forest C database (ForC; https://forc-db.github.io/) and stand to
+be of value to greenhouse gas accounting if made available through the
+Emission Factor Database (EFDB) of the International Panel on Climate
+Change (IPCC). Here, we develop and document a process for
+semi-automated transfer of data from ForC into the EFDB, assess the data
+available and transferred to date, and provide recommendations for
+improving forest data collection, analysis, and reporting to improve
+accounting of forest-sector greenhoouse gas emissions and removals. We
+begin by reconciling terminology and mapping ForC fields into EFDB. This
+process required some updates to the ForC database structure, leading to
+the release of a new version of ForC (v4.0; described here). At the time
+of writing, ForC contained \#\# values that would qualify for inclusion
+in the EFDB, \#\# of which have been transferred to date. (Some analysis
+of representation/ gaps.) In the future, forest C estimates in EFDB can
+be improved through targetted research to fill critical gaps, reporting
+of information required by IPCC, and continued submission of data from
+scientific publications to the EFDB.
 \end{abstract}
@@
 
-Forests are critical to our management of greenhouse gases. The
-influence the forests have on atmospheric carbon has made forest data
-increasingly valuable.
+Forests are critical to management of atmospheric concentrations of the
+greenhouse gas carbon dioxide (CO\textsubscript{2}), and thereby climate
+change. In recent decades, CO\textsubscript{2} uptake by forests,
+woodlands, and savannas has exceeded releases from deforestation and
+other severe disturbances, resulting in a net carbon CO\textsubscript{2}
+sink of \textasciitilde0.88 Gt C yr\textsuperscript{-1} \citep[all
+biomes with trees,][]{xu_changes_2021} to \textasciitilde1.6 Gt C
+yr\textsuperscript{-1} \citep[forests only,][]{harris_global_2021}. This
+has offset an estimated 10\% to 18\% of anthropogenic
+CO\textsubscript{2} emissions from fossil fuels and cement
+\citep{xu_changes_2021, harris_global_2021}, dramatically slowing the
+pace of atmospheric CO\textsubscript{2} accumulation and climate change.
+Going into the future, the fate of this important CO\textsubscript{2}
+sink is highly uncertain, depending both upon forest responses to
+climate change, which are likely to reduce the sink strength
+\citep{refs}, and on human conservation, restoration, and management of
+forests \citep{refs}.
 
-\textbf{Forests play a central role in international plans for climate
-change mitigation \citep{unfccc_adoption_2015}.} As of 2016,
-forest-based mitigation accounted for 26\% of total planned greenhouse
-gas mitigation within Nationally Determined Contributions under the
-Paris Agreement \citep{grassi_key_2017}. Effective forest-based climate
-change mitigation requires solid scientific data
-\citep{anderson-teixeira_effective_2022}.
+Reflecting their strong influence on Earth's climate, forests play a
+central role in international plans for climate change mitigation under
+the Paris Agreement \citep{unfccc_adoption_2015}. Forest conservation,
+reforestation, and improved sustainable management all have significant
+-- and relatively cost-effective -- potential as climate change
+mitigation options, with conservation and reforestation having the
+fourth and fifth largest net emission reduction potentials or all
+mitigation options \citep{ipcc_summary_2022}. As of 2016, forest-based
+mitigation accounted for 26\% of total planned greenhouse gas mitigation
+within Nationally Determined Contributions under the Paris Agreement
+\citep{grassi_key_2017}. Yet, envisioned forest-based climate change
+mitigation initiatives do not always correspond to actual emission
+reductions through on-the-ground implementation \citep{refs}. One
+critical need for ensuring that forest-based climate change mitigation
+initiatives are effective is realistic planning, underlain by solid
+scientific data \citep{anderson-teixeira_effective_2022}.
 
-\textbf{The International Panel on Climate Change (IPCC) provides
-guidance for national greenhouse gas (GHG) inventories for reporting to
-the United Nations Framework Convention on Climate Change \citep[UNFCCC,
-(REFS for older guidelines),][]{ipcc_2019_2019}.} Under this guidance,
-GHG inventories include all managed land, including most of the world's
+The International Panel on Climate Change (IPCC) provides guidance for
+national greenhouse gas inventories for reporting to the United Nations
+Framework Convention on Climate Change
+\citep[UNFCCC,][]{ipcc_2019_2019}. Under this guidance, greenhouse gas
+inventories include all managed land, including most of the world's
 forest land \citep{ogle_delineating_2018}. The IPCC inventory guidelines
-include specific instructions for accounting for forest land \ldots{}
-\citep{ipcc_agriculture_2006, ipcc_2019_2019a}. This guidance has
-improved as more of the relevant underlying data has become available
-\citep{requenasuarez_estimating_2019}, but there remains room for
-continuous improvement as the science advances. For example,
-\citet{cook-patton_mapping_2020} \emph{(summarize IPCC comparison
-results)}. Another example is \citet{cuni-sanchez_high_2021}.
-\emph{Moreover, it is useful for those compiling national greenhouse gas
-inventories to have access to locally-specific information, when
-available.} To improve the data accessible for C accounting, the IPCC
-created the Emission Factor Database (EFDB;
-https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a
-recognized library of emission factors and other parameters that can be
-used for estimating greenhouse gas emissions and removals.
+include specific instructions for accounting for greenhouse gas (mainly
+CO\textsubscript{2}) exchanges between forest land and the atmosphere
+\citep{ipcc_agriculture_2006, ipcc_2019_2019}. This guidance has
+improved over the years as more of the relevant underlying data has
+become available
+\citep{requenasuarez_estimating_2019, rozendaal_aboveground_2022}, but
+there remains room for continuous improvement as the science advances.
+For example, the year following the release of the latest IPCC
+guidelines, \citet{cook-patton_mapping_2020} found that the latest
+default rates may underestimate rates of C accumulation in regrowth
+forests by 32\% on average and fail to capture eight-fold variation
+within ecozones. In addition, \citet{cuni-sanchez_high_2021} found that
+aboveground C stocks in mature African tropical montane forests were
+two-thirds higher than the IPCC default values for these forests. This
+rapid evolution of scientific information on the climate mitigation
+potential of forests is beneficial to climate mitigation efforts, but
+requires improved mechanisms for communicating the latest information
+from scientific researchers to the practitioners who need reliable
+estimates for greenhouse gas mitigation planning. Moreover, high
+variability of forest C cycling within ecozones
+\citep[e.g.,][]{cook-patton_mapping_2020, refs} implies that it is
+useful for those compiling national greenhouse gas inventories to have
+access to locally-specific information, when available. To improve the
+data accessible for C accounting, the IPCC created the Emission Factor
+Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which
+is intended as a recognized library of emission factors and other
+parameters that can be used for estimating greenhouse gas emissions and
+removals.
 
@@
 accessible.} The EFDB is intended as a recognized library, and can be
-used both for efforts to tally a nation's intended or accomplished GHG
-reductions, or as a basis of comparison for external parties to evaluate
-these inventories.
+used both for efforts to tally a nation's intended or accomplished
+greenhouse gas reductions, or as a basis of comparison for external
+parties to evaluate these inventories.
 
@@
 
-The UNFCCC requires GHG reporting for all managed lands in a country,
-where management is defined as ``human interventions and practices have
-been applied to perform production, ecological or social functions''
-{[}\emph{IPCC 2006 full report REF}{]}. This definition is applied
-differently across countries, and is not clearly defined by the majority
-of governments \citep{ogle_delineating_2018}. Given this, and because
-the IPCC definition of management does not necessarily match that which
-would be reported in scientific publications and hence in ForC, we do
-not transfer any classification of land management status from ForC to
-the EFDB, but do provide auxiliary info that may be useful in making
-this determination (e.g., geographical location).
+The UNFCCC requires greenhouse gas reporting for all managed lands in a
+country, where management is defined as ``human interventions and
+practices have been applied to perform production, ecological or social
+functions'' {[}\emph{IPCC 2006 full report REF}{]}. This definition is
+applied differently across countries, and is not clearly defined by the
+majority of governments \citep{ogle_delineating_2018}. Given this, and
+because the IPCC definition of management does not necessarily match
+that which would be reported in scientific publications and hence in
+ForC, we do not transfer any classification of land management status
+from ForC to the EFDB, but do provide auxiliary info that may be useful
+in making this determination (e.g., geographical location).
 

### 2022-04-11 doc/manuscript/ESSDmanuscript.Rmd
work on intro

@@
 copyrightstatement: |
-  The author's copyright for this publication is transferred to institution/company. 
+### The author's copyright for this publication is transferred to institution/company. 
 ### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
@@
 As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
-Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [@refs].
-One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning, underlain by solid scientific data [@anderson-teixeira_effective_2022].
+Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022]<!--could be good to add some more references here-->.
+One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
 
@@
 This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
-Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @refs] implies that it is useful for those compiling national greenhouse gas inventories to have access to locally-specific information, when available.
+Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @refs] implies that it is useful for practitioners to have access to locally-specific information, when available.
+
 To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
-
-**To ensure that planned emissions reductions are realistic, high-quality estimates of forest C stocks and fluxes must be publicly accessible.**
 The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories.
-<!-- check Washington Post article on Malaysia NDC, @deng_comparing_2021 -->
+The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
+In the case of forests, emission factors refer to ... <!--C stocks and increments, and check other GHGs-->
 

### 2022-04-11 doc/manuscript/ESSDmanuscript.tex
work on intro

@@
 
-\copyrightstatement{The author's copyright for this publication is
-transferred to institution/company.}
-
 \introduction[Introduction]
@@
 mitigation initiatives do not always correspond to actual emission
-reductions through on-the-ground implementation \citep{refs}. One
-critical need for ensuring that forest-based climate change mitigation
-initiatives are effective is realistic planning, underlain by solid
-scientific data \citep{anderson-teixeira_effective_2022}.
+reductions through on-the-ground implementation
+\citep[e.g.,][]{badgley_systematic_2022}. One critical need for ensuring
+that forest-based climate change mitigation initiatives are effective is
+realistic planning and reporting, underlain by solid scientific data
+\citep{anderson-teixeira_effective_2022, deng_comparing_2021}.
 
@@
 \citep[e.g.,][]{cook-patton_mapping_2020, refs} implies that it is
-useful for those compiling national greenhouse gas inventories to have
-access to locally-specific information, when available. To improve the
-data accessible for C accounting, the IPCC created the Emission Factor
-Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which
-is intended as a recognized library of emission factors and other
-parameters that can be used for estimating greenhouse gas emissions and
-removals.
+useful for practitioners to have access to locally-specific information,
+when available.
 
-\textbf{To ensure that planned emissions reductions are realistic,
-high-quality estimates of forest C stocks and fluxes must be publicly
-accessible.} The EFDB is intended as a recognized library, and can be
-used both for efforts to tally a nation's intended or accomplished
-greenhouse gas reductions, or as a basis of comparison for external
-parties to evaluate these inventories.
+To improve the data accessible for C accounting, the IPCC created the
+Emission Factor Database (EFDB;
+https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a
+recognized library of emission factors and other parameters that can be
+used for estimating greenhouse gas emissions and removals. The EFDB is
+intended as a recognized library, and can be used both for efforts to
+tally a nation's intended or accomplished greenhouse gas reductions, or
+as a basis of comparison for external parties to evaluate these
+inventories. The EFDB encourages researchers to submit estimates of
+emission factors or other related parameters from peer-reviewed journal
+papers or other accepted sources for inclusion in the database. In the
+case of forests, emission factors refer to \ldots{}
 

### 2022-04-11 doc/manuscript/ESSDmanuscript.Rmd
working on intro

@@
 
-To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
-The EFDB is intended as a recognized library, and can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories.
+To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
+The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
 The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors refer to ... <!--C stocks and increments, and check other GHGs-->
+In the case of forests, emission factors include carbon stocks and increments within various pools (Table&nbsp;\@ref(table_variables)).
 
-**The Global Forest Carbon Database, ForC, is the largest collection of published estimates of forest carbon stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].**
-*(add stats/ details, maybe record of how ForC has grown over time)*
+The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest carbon stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
+For C currently contains **##** records from **##** plots in **##** distinct geographical areas, along with records of stand age and disturbance history.
 As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
@@
 **Here, we**
-(1) review IPCC definitions of relevant carbon stocks and increments
-(2) describe mapping of ForC to IPCC's EFDB,
-(3) describe updates to ForC  (ForC v4.0),
-(4) summarize the data in ForC that's relevant to EFDB, identifying gaps, and
+(1) review IPCC definitions of relevant carbon stocks and increments;
+(2) describe mapping of ForC to IPCC's EFDB;
+(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data transfer to EFDB;
+(4) summarize the data in ForC that's relevant to EFDB and records that have been transferred to date; and
 (5) provide recommendations for improving data collection, analysis, database, and accounting.

### 2022-04-11 doc/manuscript/ESSDmanuscript.tex
working on intro

@@
 Emission Factor Database (EFDB;
-https://www.ipcc-nggip.iges.or.jp/EFDB/main.php), which is intended as a
-recognized library of emission factors and other parameters that can be
-used for estimating greenhouse gas emissions and removals. The EFDB is
-intended as a recognized library, and can be used both for efforts to
-tally a nation's intended or accomplished greenhouse gas reductions, or
-as a basis of comparison for external parties to evaluate these
-inventories. The EFDB encourages researchers to submit estimates of
-emission factors or other related parameters from peer-reviewed journal
-papers or other accepted sources for inclusion in the database. In the
-case of forests, emission factors refer to \ldots{}
+\url{https://www.ipcc-nggip.iges.or.jp/EFDB/main.php}), which is
+intended as a recognized library of emission factors and other
+parameters that can be used for estimating greenhouse gas emissions and
+removals. The EFDB can be used both for efforts to tally a nation's
+intended or accomplished greenhouse gas reductions, or as a basis of
+comparison for external parties to evaluate these inventories. The EFDB
+encourages researchers to submit estimates of emission factors or other
+related parameters from peer-reviewed journal papers or other accepted
+sources for inclusion in the database. In the case of forests, emission
+factors include carbon stocks and increments within various pools
+(Table~@ref(table\_variables)).
 
-\textbf{The Global Forest Carbon Database, ForC, is the largest
-collection of published estimates of forest carbon stocks, increments,
-and annual fluxes
-\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}.}
-\emph{(add stats/ details, maybe record of how ForC has grown over
-time)} As such, ForC is positioned to improve forest C accounting
-through the transfer of data to EFDB. The purpose of this publication is
-to document that process and provide recommendations for future
-improvements.
+The Global Forest Carbon Database, ForC
+(\url{https://forc-db.github.io/}), is the largest collection of
+published estimates of forest carbon stocks, increments, and annual
+fluxes
+\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. For
+C currently contains \textbf{\#\#} records from \textbf{\#\#} plots in
+\textbf{\#\#} distinct geographical areas, along with records of stand
+age and disturbance history. As such, ForC is positioned to improve
+forest C accounting through the transfer of data to EFDB. The purpose of
+this publication is to document that process and provide recommendations
+for future improvements.
 
 \textbf{Here, we} (1) review IPCC definitions of relevant carbon stocks
-and increments (2) describe mapping of ForC to IPCC's EFDB, (3) describe
-updates to ForC (ForC v4.0), (4) summarize the data in ForC that's
-relevant to EFDB, identifying gaps, and (5) provide recommendations for
-improving data collection, analysis, database, and accounting.
+and increments; (2) describe mapping of ForC to IPCC's EFDB; (3)
+describe updates to ForC (ForC v4.0), most of which were implemented to
+facilitate data transfer to EFDB; (4) summarize the data in ForC that's
+relevant to EFDB and records that have been transferred to date; and (5)
+provide recommendations for improving data collection, analysis,
+database, and accounting.
 

### 2022-04-11 doc/manuscript/ESSDmanuscript.Rmd
planning/organizing (+ misc changes)
@mawilliams99 , FYI. Changes include the following:
- inserted upfront r chunk to define table & figure numbers, insert relevant statistics (to be automated)
-inserted placeholder figures and tables
- added/ edited some content

@@
 
-   CURRENT TABLE 2 GOES HERE
+   CURRENT TABLE 3 GOES HERE
 
@@
 
-   CURRENT TABLE 3 GOES HERE
+   CURRENT TABLE 4 GOES HERE
 
@@
 Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.-->
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+#Tables
+table_pools =1
+table_variables =2
+
+#Figures
+fig_variable_mapping=1
+fig_map=2
+fig_dominant_veg=3
+
+#Statistics
+n_records_ForC= NA # total n records in ForC
+n_independent_records_ForC= NA #suspected duplicates removed
+n_records_EFDB_variables= NA #n independent records relevant to EFDB
+n_records_sent_to_EFDB = NA # n records sent to EFDB
+n_records_posted_EFDB = NA # n records accepted and posted in EFDB
+```
 
@@
 The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include carbon stocks and increments within various pools (Table&nbsp;\@ref(table_variables)).
+In the case of forests, emission factors include carbon stocks and increments within various pools [@ipcc_2019_2019].
 
@@
 
-**Here, we**
+Here, we
 (1) review IPCC definitions of relevant carbon stocks and increments;
@@
 Forest ecosystem C stocks may be parsed into pools in various ways. 
-IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table&nbsp;\@ref(table_variables)).
+IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
 Quantifying these requires a one-time measurement.
 
-```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table_pools, eval = TRUE, echo=FALSE, warning=FALSE}
 C_pools_table <- read.csv("figures_tables/C_pools.csv")
@@
              longtable = FALSE,
-             caption = "\\textbf{Variables with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
+             caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
   kableExtra::kable_styling(full_width = TRUE) %>%
@@
 \clearpage
-```{r, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
+```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
 Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
@@
 
-Mapping of variables is shown in Fig. 1
+Mapping of variables is shown in Fig. `r `fig_variable_mapping`.
 
@@
 
+Because ForC v4.0 contained known duplicate records, we used R scripts to remove likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
+Henceforth, we refer to the records with duplicates removed as "independent records".
+
 # Results
 
-**figure: map of relevant ForC data with underlying FAO ecozones**
+## ForC v4.0 contents
 
-**(summarize the data in ForC that's relevant to EFDB, identifying gaps)**
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_records_EFDB_variables` of which were for variables relevant to EFDB (Fig.
+`r fig_variable_mapping`).
+These records were distributed across all forested continents and ecozones, with particularly high concentrations in *[ecozone(s)]* and low concentrations in *[ecozone(s)]* (Fig. `r fig_map`).
+The most widely represented forest type was *[type]*, followed by *[type]* and *[type]* (Fig. `r fig_dominant_veg`).
 
-dead wood and litter comparisons will be particularly interesting, as IPCC values are based on just a handful of references for each climate zone (table 2.2 in 2019 guidelines)
+```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} PLACEHOLDER.
+We need to add records transferred to EFDB.
+We also need to limit map to variables relavent to EFDB.
+It would also be good to add an underlying map of FAP ecozones"}
+#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
+knitr::include_graphics("figures_tables/PLACEHOLDER_World_Map_of_sites.png")
+```
+
+```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Historgram of number of records in ForC relavent to (grey) and transferred to (black) EFDB.} PLACEHOLDER.
+We need to add axis labels and the two classes of records.
+"}
+#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
+knitr::include_graphics("figures_tables/PLACEHOLDER_Williams_Histogram_of_dominant_vegetation.png")
+```
+
+ForC contained data for **##** of the **##** variables relevant to EFDB (Table `r table_variables`, Fig. `r fig_variable_mapping`).
+
+```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
+C_variables_table <- read.csv("figures_tables/C_variables.csv")
+names(C_variables_table) <- stringr::str_replace_all(names(C_variables_table), "\\.", " ")
+
+library(knitr)
+library(kableExtra)
+knitr::kable(C_variables_table, format = "latex",
+             longtable = FALSE,
+             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
+") %>%
+  kableExtra::kable_styling(full_width = TRUE) %>%
+  row_spec(0, bold = TRUE)
+```
+
+## Data transfers to EFDB
+
+As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB`, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Fig.
+`r fig_map`-`r fig_dominant_veg`, `Table r table_variables`).
+*[DETAILS]*
 
@@
 
-**(Paragraph highlighting important gaps in variables / regions)**
+*(Paragraph highlighting important gaps in variables / regions)*
+
+<!--dead wood and litter comparisons will be particularly interesting, as IPCC values are based on just a handful of references for each climate zone (table 2.2 in 2019 guidelines)-->
 
 Several variables of value to IPCC, including standing dead wood, woody mortality, delta.agb, are not calculated and presented as frequently as are AGB and ANPP_woody, even though they can readily be derived from the same census data.
-We recommend that researchers calculate and report these, as specified below.
+We recommend that researchers calculate and report these, following the reporting guidelines specified below.
 Furthermore, there is an opportunity to fill data gaps by calculating these from existing census data.
-For example, the core census protocol of the Forest Global Earth Observatory [ForestGEO; REFS] collects the data required to calculate standing dead wood, woody mortality, and delta.agb, but these have not been calculated and reported for all sites for which the appropriate number of censuses are available (n=1 for standing dead wood, n=2 for woody mortality and delta.agb) [but see REFS].
+For example, the core census protocol of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfsforestgeo_2015,@davies_forestgeo_2021] collects the data required to calculate standing dead wood, woody mortality, and delta.agb, but these have not been calculated and reported for all sites for which the appropriate number of censuses are available <!--"(n=1 for standing dead wood, n=2 for woody mortality and delta.agb)" -- previous text commented out because I think it's out of date-->[but see @piponiot_distribution_2022, @refs<!--insert other refs including these variables-->].
 
@@
 
-## IPCC
+## IPCC protocol considerations
 
-An important challenge is that forests are changing rapidly, and data collected a decaade ago may no longer be relevant, particularly in the cases of C increments and fluxes.
+An important challenge is that forests are changing rapidly, and data collected a decade ago may no longer be relevant, particularly in the cases of C increments and fluxes.
 

### 2022-04-11 doc/manuscript/ESSDmanuscript.tex
planning/organizing (+ misc changes)
@mawilliams99 , FYI. Changes include the following:
- inserted upfront r chunk to define table & figure numbers, insert relevant statistics (to be automated)
-inserted placeholder figures and tables
- added/ edited some content

@@
 factors include carbon stocks and increments within various pools
-(Table~@ref(table\_variables)).
+\citep{ipcc_2019_2019}.
 
@@
 
-\textbf{Here, we} (1) review IPCC definitions of relevant carbon stocks
-and increments; (2) describe mapping of ForC to IPCC's EFDB; (3)
-describe updates to ForC (ForC v4.0), most of which were implemented to
+Here, we (1) review IPCC definitions of relevant carbon stocks and
+increments; (2) describe mapping of ForC to IPCC's EFDB; (3) describe
+updates to ForC (ForC v4.0), most of which were implemented to
 facilitate data transfer to EFDB; (4) summarize the data in ForC that's
@@
 parses into biomass (aboveground and belowground), dead organic matter
-(dead wood and litter), and soil organic matter
-(Table~@ref(table\_variables)). Quantifying these requires a one-time
-measurement.
+(dead wood and litter), and soil organic matter (Table 1). Quantifying
+these requires a one-time measurement.
 
@@
 
-\caption{\label{tab:table_variables}\textbf{Variables with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
+\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
 \centering
@@
 \includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
-Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:unnamed-chunk-1}
+Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:fig_variable_mapping}
 \end{figure}
@@
 
-Mapping of variables is shown in Fig. 1
+Mapping of variables is shown in Fig. \texttt{r}fig\_variable\_mapping`.
 
@@
 
+Because ForC v4.0 contained known duplicate records, we used R scripts
+to remove likely duplicates, as detailed in
+\citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the
+records with duplicates removed as ``independent records''.
+
 \section{Results}
 
-\textbf{figure: map of relevant ForC data with underlying FAO ecozones}
+\subsection{ForC v4.0 contents}
 
-\textbf{(summarize the data in ForC that's relevant to EFDB, identifying
-gaps)}
+As of April 11, 2022, ForC (v4.0) contained NA independent records (NA
+total), NA of which were for variables relevant to EFDB (Fig. 1). These
+records were distributed across all forested continents and ecozones,
+with particularly high concentrations in \emph{{[}ecozone(s){]}} and low
+concentrations in \emph{{[}ecozone(s){]}} (Fig. 2). The most widely
+represented forest type was \emph{{[}type{]}}, followed by
+\emph{{[}type{]}} and \emph{{[}type{]}} (Fig. 3).
 
-dead wood and litter comparisons will be particularly interesting, as
-IPCC values are based on just a handful of references for each climate
-zone (table 2.2 in 2019 guidelines)
+\begin{figure}
+\includegraphics[width=15cm]{figures_tables/PLACEHOLDER_World_Map_of_sites} \caption{\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} PLACEHOLDER.
+We need to add records transferred to EFDB.
+We also need to limit map to variables relavent to EFDB.
+It would also be good to add an underlying map of FAP ecozones}\label{fig:fig_map}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=15cm]{figures_tables/PLACEHOLDER_Williams_Histogram_of_dominant_vegetation} \caption{\textbf{Historgram of number of records in ForC relavent to (grey) and transferred to (black) EFDB.} PLACEHOLDER.
+We need to add axis labels and the two classes of records.
+}\label{fig:fig_dominant_veg}
+\end{figure}
+
+ForC contained data for \textbf{\#\#} of the \textbf{\#\#} variables
+relevant to EFDB (Table 2, Fig. 1).
+
+\begin{table}
+
+\caption{\label{tab:table_variables}\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
+}
+\centering
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\hline
+\textbf{variable} & \textbf{n in ForC} & \textbf{n independent records in ForC} & \textbf{n reviewed} & \textbf{n sent to EFDB} & \textbf{n posted to EFDB}\\
+\hline
+STOCKS & NA & NA & NA & NA & NA\\
+\hline
+biomass & NA & NA & NA & NA & NA\\
+\hline
+biomass\_ag & NA & NA & NA & NA & NA\\
+\hline
+biomass\_ag\_foliage & NA & NA & NA & NA & \vphantom{1} NA\\
+\hline
+biomass\_ag\_woody & NA & NA & NA & NA & NA\\
+\hline
+biomass\_root & NA & NA & NA & NA & NA\\
+\hline
+biomass\_root\_fine & NA & NA & NA & NA & NA\\
+\hline
+biomass\_root\_coarse & NA & NA & NA & NA & NA\\
+\hline
+deadwood & NA & NA & NA & NA & NA\\
+\hline
+deadwood\_standing & NA & NA & NA & NA & NA\\
+\hline
+deadwood\_down & NA & NA & NA & NA & NA\\
+\hline
+O.horizon & NA & NA & NA & NA & NA\\
+\hline
+SOM / SOC & NA & NA & NA & NA & NA\\
+\hline
+INCREMENTS & NA & NA & NA & NA & NA\\
+\hline
+delta.biomass & NA & NA & NA & NA & NA\\
+\hline
+delta.agb & NA & NA & NA & NA & NA\\
+\hline
+biomass\_ag\_foliage & NA & NA & NA & NA & NA\\
+\hline
+delta.biomass\_root & NA & NA & NA & NA & NA\\
+\hline
+delta.deadwood & NA & NA & NA & NA & NA\\
+\hline
+delta.deadwood\_standing & NA & NA & NA & NA & NA\\
+\hline
+delta.deadwood\_down & NA & NA & NA & NA & NA\\
+\hline
+delta.O.horizon & NA & NA & NA & NA & NA\\
+\hline
+delta.SOM / delta.SOC & NA & NA & NA & NA & NA\\
+\hline
+FLUXES & NA & NA & NA & NA & NA\\
+\hline
+NPP\_woody & NA & NA & NA & NA & NA\\
+\hline
+ANPP\_woody\_stem & NA & NA & NA & NA & NA\\
+\hline
+ANPP\_foliage & NA & NA & NA & NA & NA\\
+\hline
+BNPP\_root & NA & NA & NA & NA & NA\\
+\hline
+BNPP\_root\_fine & NA & NA & NA & NA & NA\\
+\hline
+BNPP\_root\_coarse & NA & NA & NA & NA & NA\\
+\hline
+woody\_mortality & NA & NA & NA & NA & NA\\
+\hline
+NPP\_litterfall & NA & NA & NA & NA & NA\\
+\hline
+R\_het\_litter & NA & NA & NA & NA & NA\\
+\hline
+R\_het\_soil & NA & NA & NA & NA & NA\\
+\hline
+TOTAL & NA & NA & NA & NA & NA\\
+\hline
+\end{tabu}
+\end{table}
+
+\subsection{Data transfers to EFDB}
+
+As of April 11, 2022, we had reviewed and sent NA, NA of which have been
+reviewed, accepted, and posted (Fig. 2-3,
+\texttt{Table\ r\ table\_variables}). \emph{{[}DETAILS{]}}
 
@@
 
-\textbf{(Paragraph highlighting important gaps in variables / regions)}
+\emph{(Paragraph highlighting important gaps in variables / regions)}
 
@@
 the same census data. We recommend that researchers calculate and report
-these, as specified below. Furthermore, there is an opportunity to fill
-data gaps by calculating these from existing census data. For example,
-the core census protocol of the Forest Global Earth Observatory
-{[}ForestGEO; REFS{]} collects the data required to calculate standing
-dead wood, woody mortality, and delta.agb, but these have not been
-calculated and reported for all sites for which the appropriate number
-of censuses are available (n=1 for standing dead wood, n=2 for woody
-mortality and delta.agb) {[}but see REFS{]}.
+these, following the reporting guidelines specified below. Furthermore,
+there is an opportunity to fill data gaps by calculating these from
+existing census data. For example, the core census protocol of the
+Forest Global Earth Observatory {[}ForestGEO;
+\citet{anderson-teixeira_ctfsforestgeo_2015},\citet{davies_forestgeo_2021}{]}
+collects the data required to calculate standing dead wood, woody
+mortality, and delta.agb, but these have not been calculated and
+reported for all sites for which the appropriate number of censuses are
+available \citep[but see][\citet{refs}]{piponiot_distribution_2022}.
 
@@
 
-\subsection{IPCC}
+\subsection{IPCC protocol considerations}
 
 An important challenge is that forests are changing rapidly, and data
-collected a decaade ago may no longer be relevant, particularly in the
+collected a decade ago may no longer be relevant, particularly in the
 cases of C increments and fluxes.
@@
 
-CURRENT TABLE 2 GOES HERE
+CURRENT TABLE 3 GOES HERE
 
@@
 
-CURRENT TABLE 3 GOES HERE
+CURRENT TABLE 4 GOES HERE
 \noappendix

### 2022-04-11 doc/manuscript/ESSDmanuscript.Rmd
misc changes

@@
 n_records_ForC= NA # total n records in ForC
+n_plots_ForC = NA # total n plots in ForC
+n_areas_ForC = NA # total n geographic areas in ForC
 n_independent_records_ForC= NA #suspected duplicates removed
@@
 
-The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2019_2019].
+The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
 Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
@@
 The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest carbon stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
-For C currently contains **##** records from **##** plots in **##** distinct geographical areas, along with records of stand age and disturbance history.
+For C currently contains `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
 As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
@@
 
-For quantifying forest role in global C cycle, we ultimately care about:
-(1) C stocks -- stores of C that would be vulnerable to release to the atmosphere upon land use change
-(2) C increments -- changes in those C stocks.
+The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly for the purpose of national inventories [@ipcc_2006_2006;@ipcc_2019_2019].
+More proximately, IPCC guidelines focus on:
+(1) C stocks -- i.e., stores of C that would be vulnerable to release to the atmosphere upon land use change
+and (2) C increments -- i.e., changes in those C stocks over one or more years.
+*[detail how these are used]*
 
 ## Carbon stocks
-Forest ecosystem C stocks may be parsed into pools in various ways. 
+Forest ecosystem C stocks may be parsed into pools in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies.
 IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).

### 2022-04-11 doc/manuscript/ESSDmanuscript.tex
misc changes

@@
 Framework Convention on Climate Change
-\citep[UNFCCC,][]{ipcc_2019_2019}. Under this guidance, greenhouse gas
-inventories include all managed land, including most of the world's
-forest land \citep{ogle_delineating_2018}. The IPCC inventory guidelines
-include specific instructions for accounting for greenhouse gas (mainly
-CO\textsubscript{2}) exchanges between forest land and the atmosphere
-\citep{ipcc_agriculture_2006, ipcc_2019_2019}. This guidance has
-improved over the years as more of the relevant underlying data has
-become available
+\citep[UNFCCC,][]{ipcc_2006_2006, ipcc_2019_2019}. Under this guidance,
+greenhouse gas inventories include all managed land, including most of
+the world's forest land \citep{ogle_delineating_2018}. The IPCC
+inventory guidelines include specific instructions for accounting for
+greenhouse gas (mainly CO\textsubscript{2}) exchanges between forest
+land and the atmosphere \citep{ipcc_agriculture_2006, ipcc_2019_2019}.
+This guidance has improved over the years as more of the relevant
+underlying data has become available
 \citep{requenasuarez_estimating_2019, rozendaal_aboveground_2022}, but
@@
 \citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. For
-C currently contains \textbf{\#\#} records from \textbf{\#\#} plots in
-\textbf{\#\#} distinct geographical areas, along with records of stand
-age and disturbance history. As such, ForC is positioned to improve
-forest C accounting through the transfer of data to EFDB. The purpose of
-this publication is to document that process and provide recommendations
-for future improvements.
+C currently contains NA records from NA plots in NA distinct
+geographical areas, along with records of stand age and disturbance
+history. As such, ForC is positioned to improve forest C accounting
+through the transfer of data to EFDB. The purpose of this publication is
+to document that process and provide recommendations for future
+improvements.
 
@@
 
-For quantifying forest role in global C cycle, we ultimately care about:
-(1) C stocks -- stores of C that would be vulnerable to release to the
-atmosphere upon land use change (2) C increments -- changes in those C
-stocks.
+The end goal of IPCC greenhouse gas inventories is to quantify
+greenhouse gas emissions to, or withdrawals from, the atmosphere on an
+annual basis, most commonly for the purpose of national inventories
+\citep{ipcc_2006_2006, ipcc_2019_2019}. More proximately, IPCC
+guidelines focus on: (1) C stocks -- i.e., stores of C that would be
+vulnerable to release to the atmosphere upon land use change and (2) C
+increments -- i.e., changes in those C stocks over one or more years.
+\emph{{[}detail how these are used{]}}
 
@@
 
-Forest ecosystem C stocks may be parsed into pools in various ways. IPCC
-parses into biomass (aboveground and belowground), dead organic matter
-(dead wood and litter), and soil organic matter (Table 1). Quantifying
-these requires a one-time measurement.
+Forest ecosystem C stocks may be parsed into pools in various ways, and
+while certain definitions and thresholds are more common than others,
+there is no single standard for measuring or reporting that is adhered
+to by all -- or even most -- studies. IPCC parses into biomass
+(aboveground and belowground), dead organic matter (dead wood and
+litter), and soil organic matter (Table 1). Quantifying these requires a
+one-time measurement.
 

### 2022-04-12 doc/manuscript/ESSDmanuscript.Rmd
work on IPCC methods & definitions section

@@
 The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include carbon stocks and increments within various pools [@ipcc_2019_2019].
+In the case of forests, emission factors include carbon stocks, increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
 
@@
 Here, we
-(1) review IPCC definitions of relevant carbon stocks and increments;
+(1) review IPCC methods and definitions for tallying forest C;
 (2) describe mapping of ForC to IPCC's EFDB;
@@
 
-# IPCC definitions of carbon stocks and incremenets
+# IPCC methods and definitions
 
-The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly for the purpose of national inventories [@ipcc_2006_2006;@ipcc_2019_2019].
-More proximately, IPCC guidelines focus on:
-(1) C stocks -- i.e., stores of C that would be vulnerable to release to the atmosphere upon land use change
-and (2) C increments -- i.e., changes in those C stocks over one or more years.
-*[detail how these are used]*
-
-## Carbon stocks
-Forest ecosystem C stocks may be parsed into pools in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies.
-IPCC parses into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
-Quantifying these requires a one-time measurement.
+The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
+For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; Mg C yr^-1^) are calculated as the sum of changes in various pools, including any harvested wood products.
+IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
+For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig.
+`r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5-->[@ipcc_2006_2006].
+Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools (Fig. `r fig_variable_mapping`).
 
@@
 ```
+
+## Carbon pools
 
@@
 
+Not required for forest land remaining forest land under Tier 1 methodology [@ipcc_2006_2006].
+
 ### Soil Organic Matter
@@
 
-## Carbon increments
-
-C increments are defined as the change over time, in annual increments, in each C pool.
-These may be estimated as the difference between C stocks at two time points, or as the difference between inputs and outputs to the pool (i.e., fluxes).
-Quantifying these requires at least two measurements.
-(*But, Can carbon increments be inferred from a single measure and a known age (i.e., the approach we used in GROA)?*)
-Fluxes are the inputs and outputs to each pool.
-
-\clearpage
 ```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
@@
 
-Mapping of variables is shown in Fig. `r `fig_variable_mapping`.
+Mapping of variables is shown in Fig. `r fig_variable_mapping`.
 
@@
 
+Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies.
+
 ## Data reporting needs

### 2022-04-12 doc/manuscript/ESSDmanuscript.tex
work on IPCC methods & definitions section

@@
 sources for inclusion in the database. In the case of forests, emission
-factors include carbon stocks and increments within various pools
-\citep{ipcc_2019_2019}.
+factors include carbon stocks, increments (``stock changes''), and
+fluxes (``gains'' and ``losses'') for various pools
+\citep{ipcc_2006_2006, ipcc_2019_2019}.
 
@@
 
-Here, we (1) review IPCC definitions of relevant carbon stocks and
-increments; (2) describe mapping of ForC to IPCC's EFDB; (3) describe
-updates to ForC (ForC v4.0), most of which were implemented to
-facilitate data transfer to EFDB; (4) summarize the data in ForC that's
-relevant to EFDB and records that have been transferred to date; and (5)
-provide recommendations for improving data collection, analysis,
-database, and accounting.
+Here, we (1) review IPCC methods and definitions for tallying forest C;
+(2) describe mapping of ForC to IPCC's EFDB; (3) describe updates to
+ForC (ForC v4.0), most of which were implemented to facilitate data
+transfer to EFDB; (4) summarize the data in ForC that's relevant to EFDB
+and records that have been transferred to date; and (5) provide
+recommendations for improving data collection, analysis, database, and
+accounting.
 
-\section{IPCC definitions of carbon stocks and incremenets}
+\section{IPCC methods and definitions}
 
@@
 greenhouse gas emissions to, or withdrawals from, the atmosphere on an
-annual basis, most commonly for the purpose of national inventories
-\citep{ipcc_2006_2006, ipcc_2019_2019}. More proximately, IPCC
-guidelines focus on: (1) C stocks -- i.e., stores of C that would be
-vulnerable to release to the atmosphere upon land use change and (2) C
-increments -- i.e., changes in those C stocks over one or more years.
-\emph{{[}detail how these are used{]}}
-
-\subsection{Carbon stocks}
-
-Forest ecosystem C stocks may be parsed into pools in various ways, and
-while certain definitions and thresholds are more common than others,
-there is no single standard for measuring or reporting that is adhered
-to by all -- or even most -- studies. IPCC parses into biomass
-(aboveground and belowground), dead organic matter (dead wood and
-litter), and soil organic matter (Table 1). Quantifying these requires a
-one-time measurement.
+annual basis, most commonly on a national level
+\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
+within a land-use category, annual stock changes (\(\Delta C\); Mg C
+yr\textsuperscript{-1}) are calculated as the sum of changes in various
+pools, including any harvested wood products. IPCC parses forest C pools
+into biomass (aboveground and belowground), dead organic matter (dead
+wood and litter), and soil organic matter (Table 1). For each pool,
+\(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
+takes the difference between gains and losses (influx and outflux
+variables in Fig. 1), or using the ``Stock-Difference Method'', which
+computes \(\Delta C\) based on C stocks at two points in
+time\citep{ipcc_2006_2006}. Thus, C cycle variables relevant to the IPCC
+methodology and to EFDB include C stocks, increments, and fluxes in the
+IPCC-defined pools (Fig. 1).
 
@@
 \hline
-\textbf{pool} & \textbf{definition} & \textbf{major sources of estimate variation} & \textbf{IPCC guidance}\\
+\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
 \hline
-aboveground biomass & all biomass of living vegetation, both woody and herbaceous, above the soil & allometry, min dbh & acceptable to exclude understory\\
+aboveground biomass & all biomass of living vegetation, both woody and herbaceous & min dbh & acceptable to exclude understory if minor component\\
 \hline
-belowground biomass & all biomass of live roots & allometry, min dbh, assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
+ &  & biomass allometry & \\
 \hline
-dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min dbh,  ... & default min dbh = 10cm, but may be chosen by country\\
+ &  & include dead standing? & no\\
 \hline
-litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & min dbh for dead wood , ..
-& includes entire O horizon\\
+belowground biomass & all biomass of live roots & min dbh (coarse roots) & \\
 \hline
-soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & default sampling depth = 30cm, but may be chosen by country\\
+ &  & allometry or assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & \\
+\hline
+ &  & min root diameter & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
+\hline
+ &  & sampling depth & \\
+\hline
+dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min diameter & 10 cm default, but may be chosen by country\\
+\hline
+ &  & include belowground? & \vphantom{1} yes\\
+\hline
+litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter & 10 cm default, but may be chosen by country\\
+\hline
+ &  & layers included & includes entire O horizon\\
+\hline
+ &  & include belowground? & yes\\
+\hline
+soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & depth = 30 cm default, but may be chosen by country\\
 \hline
@@
 \end{table}
+
+\subsection{Carbon pools}
 
@@
 
+Not required for forest land remaining forest land under Tier 1
+methodology \citep{ipcc_2006_2006}.
+
 \subsubsection{Soil Organic Matter}
@@
 
-\subsection{Carbon increments}
-
-C increments are defined as the change over time, in annual increments,
-in each C pool. These may be estimated as the difference between C
-stocks at two time points, or as the difference between inputs and
-outputs to the pool (i.e., fluxes). Quantifying these requires at least
-two measurements. (\emph{But, Can carbon increments be inferred from a
-single measure and a known age (i.e., the approach we used in GROA)?})
-Fluxes are the inputs and outputs to each pool.
-
-\clearpage
 \begin{figure}
@@
 
-Mapping of variables is shown in Fig. \texttt{r}fig\_variable\_mapping`.
+Mapping of variables is shown in Fig. 1.
 
@@
 
-As of April 11, 2022, ForC (v4.0) contained NA independent records (NA
+As of April 12, 2022, ForC (v4.0) contained NA independent records (NA
 total), NA of which were for variables relevant to EFDB (Fig. 1). These
@@
 
-As of April 11, 2022, we had reviewed and sent NA, NA of which have been
+As of April 12, 2022, we had reviewed and sent NA, NA of which have been
 reviewed, accepted, and posted (Fig. 2-3,
@@
 easily.)}
+
+Forest ecosystem C pools may be parsed in various ways, and while
+certain definitions and thresholds are more common than others, there is
+no single standard for measuring or reporting that is adhered to by all
+-- or even most -- studies.
 

### 2022-04-12 doc/manuscript/ESSDmanuscript.Rmd
work on C pools

@@
 
+#Cited sections (listed here because these can change)
+section_C_pools = 2.1
+
 #Statistics
@@
 The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; Mg C yr^-1^) are calculated as the sum of changes in various pools, including any harvested wood products.
+For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; Mg C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
+For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig.
+`r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006].
+Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools (Fig. `r fig_variable_mapping`).
+
+## Carbon pools
+
+Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies.
 IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
-For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig.
-`r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5-->[@ipcc_2006_2006].
-Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools (Fig. `r fig_variable_mapping`).
+While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
 
@@
 
-## Carbon pools
-
 ### Biomass
@@
 
-The IPCC defines aboveground biomass as "all biomass of living vegetation, both woody and herbaceous, above the soil including stems, stumps, branches, bark, seeds, and foliage" [].
+The IPCC defines aboveground biomass as "all biomass of living vegetation, both woody and herbaceous, above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].
 
-Belowground biomass is defined as "all biomass of live roots" [].
+Belowground biomass is defined as "all biomass of live roots" [@ipcc_2006_2006].
 
@@
 
-Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies.
-
 ## Data reporting needs

### 2022-04-12 doc/manuscript/ESSDmanuscript.tex
work on C pools

@@
 yr\textsuperscript{-1}) are calculated as the sum of changes in various
-pools, including any harvested wood products. IPCC parses forest C pools
-into biomass (aboveground and belowground), dead organic matter (dead
-wood and litter), and soil organic matter (Table 1). For each pool,
+pools (section 2.1), plus any harvested wood products. For each pool,
 \(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
@@
 variables in Fig. 1), or using the ``Stock-Difference Method'', which
-computes \(\Delta C\) based on C stocks at two points in
-time\citep{ipcc_2006_2006}. Thus, C cycle variables relevant to the IPCC
+computes \(\Delta C\) based on C stocks at two points in time
+\citep{ipcc_2006_2006}. Thus, C cycle variables relevant to the IPCC
 methodology and to EFDB include C stocks, increments, and fluxes in the
 IPCC-defined pools (Fig. 1).
+
+\subsection{Carbon pools}
+
+Forest ecosystem C pools may be parsed in various ways, and while
+certain definitions and thresholds are more common than others, there is
+no single standard for measuring or reporting that is adhered to by all
+-- or even most -- studies. IPCC parses forest C pools into biomass
+(aboveground and belowground), dead organic matter (dead wood and
+litter), and soil organic matter (Table 1). While there is some
+flexibility around the components included in each pool, each national
+inventory must apply these in a consistent manner.
 
@@
 \hline
-aboveground biomass & all biomass of living vegetation, both woody and herbaceous & min dbh & acceptable to exclude understory if minor component\\
+aboveground biomass & all biomass of living vegetation & min dbh & acceptable to exclude understory if minor component\\
 \hline
@@
 \hline
-litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter & 10 cm default, but may be chosen by country\\
+litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter (= max diameter for deadwood) & 10 cm default, but may be chosen by country\\
 \hline
@@
 
-\subsection{Carbon pools}
-
 \subsubsection{Biomass}
@@
 vegetation, both woody and herbaceous, above the soil including stems,
-stumps, branches, bark, seeds, and foliage'' {[}{]}.
+stumps, branches, bark, seeds, and foliage'' \citep{ipcc_2006_2006}.
 
-Belowground biomass is defined as ``all biomass of live roots'' {[}{]}.
+Belowground biomass is defined as ``all biomass of live roots''
+\citep{ipcc_2006_2006}.
 
@@
 easily.)}
-
-Forest ecosystem C pools may be parsed in various ways, and while
-certain definitions and thresholds are more common than others, there is
-no single standard for measuring or reporting that is adhered to by all
--- or even most -- studies.
 

### 2022-04-12 doc/manuscript/ESSDmanuscript.Rmd
work on C pools

@@
 The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; Mg C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
+For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
 For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig.
@@
 While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
+In this section, we define and review the IPCC definitions in the context of typical forest C estimation methodologies.
 
@@
 ### Biomass
-Biomass includes living vegetation, above- and below-ground.
+Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].
 
-The IPCC defines aboveground biomass as "all biomass of living vegetation, both woody and herbaceous, above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].
+Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].
+IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" is not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
+A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
+In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs].
+Beyond the minimum DBH sampled, forest censuses and biomass estimates also differ in their inclusion of life forms other than dicot trees -- including lianas, ferns, palms, and bamboo -- which in some places can reach large sizes and/or constitute a large fraction of forest C.
+*[IPCC guidance]*
+Further, it is important to note that this excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
 
-Belowground biomass is defined as "all biomass of live roots" [@ipcc_2006_2006].
+Belowground biomass is defined as "all biomass of live roots" [@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
+The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain.
+Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
 

### 2022-04-12 doc/manuscript/ESSDmanuscript.tex
work on C pools

@@
 \citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
-within a land-use category, annual stock changes (\(\Delta C\); Mg C
+within a land-use category, annual stock changes (\(\Delta C\); t C
 yr\textsuperscript{-1}) are calculated as the sum of changes in various
@@
 flexibility around the components included in each pool, each national
-inventory must apply these in a consistent manner.
+inventory must apply these in a consistent manner. In this section, we
+define and review the IPCC definitions in the context of typical forest
+C estimation methodologies.
 
@@
 \hline
- &  & biomass allometry & \\
+ &  & include non-dicot trees & ?\\
 \hline
@@
 \hline
-belowground biomass & all biomass of live roots & min dbh (coarse roots) & \\
+ &  & biomass allometry & ?\\
+\hline
+belowground biomass & all biomass of live roots & all factors relevant to aboveground biomass & see above\\
 \hline
@@
 
-Biomass includes living vegetation, above- and below-ground.
+Biomass includes living vegetation, above- and below-ground, both woody
+and herbaceous, but with a focus on woody plants and trees given their
+much greater potential to sequester large amounts of C
+\citep{ipcc_2006_2006}.
 
-The IPCC defines aboveground biomass as ``all biomass of living
-vegetation, both woody and herbaceous, above the soil including stems,
-stumps, branches, bark, seeds, and foliage'' \citep{ipcc_2006_2006}.
+Aboveground biomass, which is typically \textless200 t C
+ha\textsuperscript{-1} but can exceed 700 t C ha\textsuperscript{-1}
+\citep{anderson-teixeira_carbon_2021}, is defined by the IPCC as ``all
+biomass of living vegetation above the soil including stems, stumps,
+branches, bark, seeds, and foliage'' \citep{ipcc_2006_2006}. IPCC's
+guidance is that the understory may be excluded the understory if it
+constitutes a ``minor'' component, \emph{where quantitative definitions
+of ``understory'' and ``minor'' is not provided}, but where a commonly
+applied minimum size sampling threshold for mature forests would be 10
+cm stem diameter at breast height (DBH). A recent study characterizing
+the contributions of trees in different DBH classes to ecosystem C
+stocks and fluxes found that trees 1 - 10 cm DBH contributed up to
+\textasciitilde8\% aboveground biomass, \textasciitilde17\% aboveground
+woody net primary productivity (\(ANPP_{woody.stem}\)), and
+\textasciitilde20\% woody mortality (\(M_{woody}\)) of mature
+closed-canopy forests worldwide \citep{piponiot_distribution_2022}. In
+regrowth forests, woodlands, or savannas, small trees and shrubs
+contribute a much larger proportion of C stocks and fluxes
+\citep{piponiot_distribution_2022, refs}, and, correspondingly, biomass
+estimates for these ecosystems tend include smaller size classes
+\citep[e.g.,][]{refs}. Beyond the minimum DBH sampled, forest censuses
+and biomass estimates also differ in their inclusion of life forms other
+than dicot trees -- including lianas, ferns, palms, and bamboo -- which
+in some places can reach large sizes and/or constitute a large fraction
+of forest C. \emph{{[}IPCC guidance{]}} Further, it is important to note
+that this excludes standing dead wood, which is included in remote
+sensing biomass estimates \citep{duncanson_aboveground_2021}.
 
 Belowground biomass is defined as ``all biomass of live roots''
-\citep{ipcc_2006_2006}.
+\citep{ipcc_2006_2006}, a definition including both coarse roots, whose
+biomass is typically estimated based on stem censuses and allometries or
+belowground to aboveground biomass ratios, and fine roots, whose biomass
+is typically estimated via extraction of roots from soil samples. The
+former, which is typically \textless40 t C ha\textsuperscript{-1}
+\citep{anderson-teixeira_carbon_2021}, is methodologically linked to
+aboveground biomass estimates, sharing the same methodological sources
+of variation, but tending to be far more uncertain. Fine root biomass
+generally constitutes a much smaller C pool \citep[typically \textless5
+t C ha\textsuperscript{-1},][]{anderson-teixeira_carbon_2021}, and IPCC
+guidance is that it can be excluded when fine roots cannot be
+distinguished empirically from soil organic matter or litter
+\citep{ipcc_2006_2006}, which can be a painstaking process.
 

### 2022-04-12 doc/manuscript/ESSDmanuscript.Rmd
work on section 2

@@
 `r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006].
-Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools (Fig. `r fig_variable_mapping`).
+Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools.
 
@@
   kableExtra::kable_styling(full_width = TRUE) %>%
+  #column_spec(1, width = ".7cm") %>% #This doesn't seem to work
   row_spec(0, bold = TRUE)
@@
 Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" is not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
+IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
 A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
@@
 Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
+Field methods for estimating root biomass are highly variable.
+IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].
 
@@
 ```
+
+## Land use categories
+IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
+Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
+Definitions of forest are allowed to vary by country, but must be applied consistently.
 

### 2022-04-12 doc/manuscript/ESSDmanuscript.tex
work on section 2

@@
 methodology and to EFDB include C stocks, increments, and fluxes in the
-IPCC-defined pools (Fig. 1).
+IPCC-defined pools.
 
@@
 \hline
- &  & include non-dicot trees & ?\\
+ &  & include non-dicot trees? & ?\\
 \hline
@@
 \hline
- &  & allometry or assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & \\
+ &  & allometry or assumed ratio of below- to above-ground biomass (R) & can estimate based on R\\
 \hline
- &  & min root diameter & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
+ &  & min root diameter & fine roots may be excluded\\
 \hline
- &  & sampling depth & \\
+ &  & sampling depth & ?\\
 \hline
@@
 constitutes a ``minor'' component, \emph{where quantitative definitions
-of ``understory'' and ``minor'' is not provided}, but where a commonly
+of ``understory'' and ``minor'' are not provided}, but where a commonly
 applied minimum size sampling threshold for mature forests would be 10
@@
 distinguished empirically from soil organic matter or litter
-\citep{ipcc_2006_2006}, which can be a painstaking process.
+\citep{ipcc_2006_2006}, which can be a painstaking process. Field
+methods for estimating root biomass are highly variable. IPCC's default
+method for Tier 1 estimates is to apply a ratio of belowground to
+aboveground biomass, with default factors defined based on ecological
+zone, continent, and forest age \citep{ipcc_2006_2006, ipcc_2019_2019}.
 
@@
 \end{figure}
+
+\subsection{Land use categories}
+
+IPCC defines land-use categories to include six categories -- Forest
+Land, Grassland, Wetlands, Cropland, Settlements, and Other Land
+\citep{ipcc_2006_2006}. Sub-divisions include land that has remained in
+a particular category for \textgreater20 years (e.g., Forest Land
+remaining Forest Land) and land that has been converted from one
+category to another in the past 20 years (e.g., Cropland converted to
+Forest Land). Definitions of forest are allowed to vary by country, but
+must be applied consistently.
 

### 2022-04-13 doc/manuscript/ESSDmanuscript.Rmd
complete references in intro

@@
 Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
-Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@refs], and on human conservation, restoration, and management of forests [@refs].
+Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and on human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
 
@@
 As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
-Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022]<!--could be good to add some more references here-->.
+Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
 One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
@@
 This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
-Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @refs] implies that it is useful for practitioners to have access to locally-specific information, when available.
+Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_202] implies that it is useful for practitioners to have access to locally-specific information, when available.
 
@@
 
+Managed land...
+
 # Mapping ForC to EFDB
@@
 As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB`, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Fig.
-`r fig_map`-`r fig_dominant_veg`, `Table r table_variables`).
+`r fig_map`-`r fig_dominant_veg`, Table `r table_variables`).
 *[DETAILS]*

### 2022-04-13 doc/manuscript/ESSDmanuscript.tex
complete references in intro

@@
 climate change, which are likely to reduce the sink strength
-\citep{refs}, and on human conservation, restoration, and management of
-forests \citep{refs}.
+\citep{mcdowell_pervasive_2020, hammond_global_2022}, and on human
+conservation, restoration, and management of forests
+\citep{ipcc_climate_2019, ipcc_climate_2022}.
 
@@
 variability of forest C cycling within ecozones
-\citep[e.g.,][]{cook-patton_mapping_2020, refs} implies that it is
-useful for practitioners to have access to locally-specific information,
-when available.
+\citep[e.g.,][]{cook-patton_mapping_2020, cuni-sanchez_high_202} implies
+that it is useful for practitioners to have access to locally-specific
+information, when available.
 
@@
 
+Managed land\ldots{}
+
 \section{Mapping ForC to EFDB}
@@
 
-As of April 12, 2022, ForC (v4.0) contained NA independent records (NA
+As of April 13, 2022, ForC (v4.0) contained NA independent records (NA
 total), NA of which were for variables relevant to EFDB (Fig. 1). These
@@
 
-As of April 12, 2022, we had reviewed and sent NA, NA of which have been
-reviewed, accepted, and posted (Fig. 2-3,
-\texttt{Table\ r\ table\_variables}). \emph{{[}DETAILS{]}}
+As of April 13, 2022, we had reviewed and sent NA, NA of which have been
+reviewed, accepted, and posted (Fig. 2-3, Table 2). \emph{{[}DETAILS{]}}
 

### 2022-04-13 doc/manuscript/ESSDmanuscript.Rmd
allometry paragraph

@@
 Beyond the minimum DBH sampled, forest censuses and biomass estimates also differ in their inclusion of life forms other than dicot trees -- including lianas, ferns, palms, and bamboo -- which in some places can reach large sizes and/or constitute a large fraction of forest C.
-*[IPCC guidance]*
+*[explain IPCC guidance]*
 Further, it is important to note that this excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
+
+A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometries to convert DBH measurements to biomass.
+Selection of allometries has an enormous influence on estimates of biomass stocks, increments, of fluxes [@clark_landscapescale_2000;@clark_net_2001].
+While trusted and standardized allometric equations are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
+*[explain IPCC guidance]*
 
@@
 
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometries to convert DBH measurements to biomass.
-*(Camille/Helene can write this paragraph easily.)*
-
 ## Data reporting needs

### 2022-04-13 doc/manuscript/ESSDmanuscript.tex
allometry paragraph

@@
 in some places can reach large sizes and/or constitute a large fraction
-of forest C. \emph{{[}IPCC guidance{]}} Further, it is important to note
-that this excludes standing dead wood, which is included in remote
-sensing biomass estimates \citep{duncanson_aboveground_2021}.
+of forest C. \emph{{[}explain IPCC guidance{]}} Further, it is important
+to note that this excludes standing dead wood, which is included in
+remote sensing biomass estimates \citep{duncanson_aboveground_2021}.
+
+A universal challenge in estimating biomass (living or dead) from forest
+census data is applying appropriate allometries to convert DBH
+measurements to biomass. Selection of allometries has an enormous
+influence on estimates of biomass stocks, increments, of fluxes
+\citep{clark_landscapescale_2000, clark_net_2001}. While trusted and
+standardized allometric equations are becoming increasingly available
+\citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022},
+large uncertainties remain. \emph{{[}explain IPCC guidance{]}}
 
@@
 
-A universal challenge in estimating biomass (living or dead) from forest
-census data is applying appropriate allometries to convert DBH
-measurements to biomass. \emph{(Camille/Helene can write this paragraph
-easily.)}
-
 \subsection{Data reporting needs}

### 2022-04-13 doc/manuscript/ESSDmanuscript.Rmd
dead wood

@@
 section_C_pools = 2.1 
+section_land_use_categories = 2.2
 
@@
 Dead organic matter includes all non-living biomass that is not within the mineral soil layer and smaller than the litter size threshold. 
+It's inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
 
-Dead wood is defined as...
+Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_2006_2006].
+This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
+While dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance; @carmona_coarse_2002], aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
+In turn, they are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
 
@@
 Live  fine  roots  (of  less  than  the  suggested  diameter  limit for  belowground  biomass) are included in litter where they cannot be distinguished from it empirically." (2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
-
-Not required for forest land remaining forest land under Tier 1 methodology [@ipcc_2006_2006].
+IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
 
@@
 
-<!--dead wood and litter comparisons will be particularly interesting, as IPCC values are based on just a handful of references for each climate zone (table 2.2 in 2019 guidelines)-->
-
 Several variables of value to IPCC, including standing dead wood, woody mortality, delta.agb, are not calculated and presented as frequently as are AGB and ANPP_woody, even though they can readily be derived from the same census data.
@@
 For example, the core census protocol of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfsforestgeo_2015,@davies_forestgeo_2021] collects the data required to calculate standing dead wood, woody mortality, and delta.agb, but these have not been calculated and reported for all sites for which the appropriate number of censuses are available <!--"(n=1 for standing dead wood, n=2 for woody mortality and delta.agb)" -- previous text commented out because I think it's out of date-->[but see @piponiot_distribution_2022, @refs<!--insert other refs including these variables-->].
+
+Given widespread trends of increasing tree mortality[@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.
 

### 2022-04-13 doc/manuscript/ESSDmanuscript.tex
dead wood

@@
 \hline
-aboveground biomass & all biomass of living vegetation & min dbh & acceptable to exclude understory if minor component\\
+aboveground biomass & all biomass of living vegetation & min dbh & may exclude understory if minor component\\
 \hline
@@
 \hline
- &  & min root diameter & fine roots may be excluded\\
+ &  & min root diameter & may exclude fine roots\\
 \hline
@@
 \hline
-soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & depth = 30 cm default, but may be chosen by country\\
+soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & 30 cm default, but may be chosen by country\\
 \hline
@@
 Dead organic matter includes all non-living biomass that is not within
-the mineral soil layer and smaller than the litter size threshold.
+the mineral soil layer and smaller than the litter size threshold. It's
+inclusion in inventories is not required under Tier 1 methodology for
+Forest Land remaining Forest Land (see section 2.2), but is required for
+land that has transitioned to or from forest within the past 20 years
+\citep{ipcc_2006_2006}.
 
-Dead wood is defined as\ldots{}
+Dead wood, which is typically \textless50 t C ha\textsuperscript{-1} but
+can exceed 150 t C ha\textsuperscript{-1}
+\citep{anderson-teixeira_carbon_2021}, is defined by IPCC as ``all
+non-living woody biomass not contained in the litter, either standing,
+lying on the ground, or in the soil'' \citep{ipcc_2006_2006}. This pool
+includes standing and fallen dead wood, stumps, and dead roots of
+diameter ≥10 cm (or a diameter specified by the country). While dead
+wood stocks and fluxes can be quite variable across forests
+\citep{anderson-teixeira_carbon_2021}, and can at times be the dominant
+pool in a forest ecosystem {[}e.g., following a severe natural
+disturbance; \citet{carmona_coarse_2002}{]}, aboveground dead wood
+remains relatively poorly characterized at a global scale
+\citep{anderson-teixeira_carbon_2021}, and belowground dead wood is
+rarely studied \citep{merganicova_dadwood_2012}. In turn, they are
+poorly characterized in large-scale forest C budgets
+\citep{pan_large_2011, harris_global_2021}, and IPCC's latest Tier 1
+default values are based on just 1-31 references per climate zone
+\citep[Table 2.2 in][]{ipcc_2019_2019}.
 
@@
 (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
-
-Not required for forest land remaining forest land under Tier 1
-methodology \citep{ipcc_2006_2006}.
+IPCC's latest Tier 1 default values are based on just 1-7 references per
+climate zone \citep[Table 2.2 in][]{ipcc_2019_2019}.
 
@@
 available \citep[but see][\citet{refs}]{piponiot_distribution_2022}.
+
+Given widespread trends of increasing tree mortality\citep{refs},
+including through severe natural disturbance \citep{refs}, better
+characterization of dead wood will be critical.
 

### 2022-04-13 doc/manuscript/ESSDmanuscript.tex
knit

@@
 \hline
-dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min diameter & 10 cm default, but may be chosen by country\\
+dead wood & all non-living woody biomass above a specified diameter, aboveground or belowground & min diameter & 10 cm default, but may be chosen by country\\
 \hline

### 2022-04-13 scripts/3_create_list_of_ all_data_sources_sent.R
update list of submitted data

@@ -19,9 +19,10 @@ list_files_sent <- list.files("data/4-transferred-to-EFDB/", pattern = ".txt", f
 # Create the table with all we need ####
 summary_table <- NULL
 
-for(x in list_files_sent) {
-  dates_sent_yyyymmdd = regmatches(x, regexpr("\\d{8}", x))
-  x <- readLines(x)
+for(f in list_files_sent) {
+  dates_sent_yyyymmdd = regmatches(f, regexpr("\\d{8}", f))
+  x <- readLines(f)
+  x <- x[file.exists(paste0("data/2-approved/processed/", x, ".csv"))] # remove the ones that were canceled
   
   # load each data.frame we sent
   data_sent <- lapply(x, function(y) read.csv(paste0("data/2-approved/processed/", y, ".csv")))

### 2022-04-13 scripts/1_ForC_to_IPCC_formatting.R
start histogram for publication #34

@@ -48,6 +48,9 @@ generate_subfields <- function(field, x = ForC_simplified, y = V_mapping_subfiel
 # subset records KEEP RECORDS WE WANT TO SEND OVER ####
 
 ## only keep records we have not already processed ####
+
+ForC_simplified_already_sent <- ForC_simplified[ForC_simplified$measurement.ID %in% trace_of_measurement_IDs$measurement.ID, ] # keep a record of what we sent for the histogram later
+
 ForC_simplified <- ForC_simplified[!ForC_simplified$measurement.ID %in% trace_of_measurement_IDs$measurement.ID, ]
 
 
@@ -69,6 +72,26 @@ ForC_simplified$variable.name <-  MEASUREMENTS$variable.name[m_meas]
 
 
 ForC_simplified <- ForC_simplified[ForC_simplified$variable.name %in% V_mapping$variable.name[V_mapping$provide.to.IPCC %in% 1], ]
+
+# ---- right, here, make the histogram for the paper ---- #
+## see this issue:https://github.com/forc-db/IPCC-EFDB-integration/issues/34
+## grey is number of records in ForC that are relevant to EFDB (n records for the variables we send)
+## black is what is already transferred to EFDB
+## x axis is dominant vegetation type
+png("doc/manuscript/figures_tables/Histogram_n_Relevant_and_Transferred_Records.png", width = 8, height = 4, units = "in", res = 300)
+
+ForBarplot <- data.frame(data_type = rep(c("relevant", "transferred"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg))
+ForBarplot <- table(ForBarplot$data_type, ForBarplot$dominant.veg)
+
+ForBarplot <- ForBarplot[, order(colSums(ForBarplot), decreasing = T)]
+
+b <- barplot(ForBarplot, col = c("grey", "black"), xaxt = "n", legend.text = T, args.legend = list(x = "topright", bty = "n"), ylab = "Number of records",, las = 1)
+text(x= b, y = -500, labels = colnames(ForBarplot), srt = 90, xpd = NA, adj= 1)
+mtext("Dominant vegetation", side = 1, line = 3.5)
+
+dev.off()
+
+# ---- END OF FIGURE FOR PAPER ----#
 
 # consider stand.age as numeric ####
 ForC_simplified$stand.age <- as.numeric(ForC_simplified$stand.age)

### 2022-04-13 scripts/1_ForC_to_IPCC_formatting.R
new version of histogram(s) #34
@mawilliams99 and @teixeirak , FYI.

@@ -78,16 +78,60 @@ ForC_simplified <- ForC_simplified[ForC_simplified$variable.name %in% V_mapping$
 ## grey is number of records in ForC that are relevant to EFDB (n records for the variables we send)
 ## black is what is already transferred to EFDB
 ## x axis is dominant vegetation type
-png("doc/manuscript/figures_tables/Histogram_n_Relevant_and_Transferred_Records.png", width = 8, height = 4, units = "in", res = 300)
+png("doc/manuscript/figures_tables/Histogram_n_Relevant_and_Transferred_Records.png", width = 8, height = 9, units = "in", res = 300)
 
-ForBarplot <- data.frame(data_type = rep(c("relevant", "transferred"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg))
-ForBarplot <- table(ForBarplot$data_type, ForBarplot$dominant.veg)
 
-ForBarplot <- ForBarplot[, order(colSums(ForBarplot), decreasing = T)]
+ForBarplot <- data.frame(data_type = rep(c("relevant", "transferred"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg), FAO.ecozone = c(ForC_simplified$FAO.ecozone, ForC_simplified_already_sent$FAO.ecozone), continent = c(ForC_simplified$continent, ForC_simplified_already_sent$continent), stand.age = c(ForC_simplified$stand.age, ForC_simplified_already_sent$stand.age))
 
-b <- barplot(ForBarplot, col = c("grey", "black"), xaxt = "n", legend.text = T, args.legend = list(x = "topright", bty = "n"), ylab = "Number of records",, las = 1)
-text(x= b, y = -500, labels = colnames(ForBarplot), srt = 90, xpd = NA, adj= 1)
-mtext("Dominant vegetation", side = 1, line = 3.5)
[…]

### 2022-04-14 doc/manuscript/ESSDmanuscript.Rmd
knit in new figures

@@
 section_land_use_categories = 2.2
+
+#READING IN FILES TO GET FIGURES/ TABLES/ STATS 
+path_to_dir <- dirname(dirname(getwd())) # get path to repo
+
+# MOVE FIGURE FILES 
+#traditional comparison
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub//Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/IPCC-EFDB-integration/doc/manuscript/figures_tables", "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE))) # first remove old files
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/ForC/figures/", "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/IPCC-EFDB-integration/doc/manuscript/figures_tables")
 
@@
 #this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
-knitr::include_graphics("figures_tables/PLACEHOLDER_World_Map_of_sites.png")
+knitr::include_graphics("figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png")
 ```
@@
 #this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
-knitr::include_graphics("figures_tables/PLACEHOLDER_Williams_Histogram_of_dominant_vegetation.png")
+knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
 ```

### 2022-04-14 doc/manuscript/ESSDmanuscript.tex
knit in new figures

@@
 
-As of April 13, 2022, ForC (v4.0) contained NA independent records (NA
+As of April 14, 2022, ForC (v4.0) contained NA independent records (NA
 total), NA of which were for variables relevant to EFDB (Fig. 1). These
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/PLACEHOLDER_World_Map_of_sites} \caption{\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} PLACEHOLDER.
+\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} PLACEHOLDER.
 We need to add records transferred to EFDB.
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/PLACEHOLDER_Williams_Histogram_of_dominant_vegetation} \caption{\textbf{Historgram of number of records in ForC relavent to (grey) and transferred to (black) EFDB.} PLACEHOLDER.
+\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Historgram of number of records in ForC relavent to (grey) and transferred to (black) EFDB.} PLACEHOLDER.
 We need to add axis labels and the two classes of records.
@@
 
-As of April 13, 2022, we had reviewed and sent NA, NA of which have been
+As of April 14, 2022, we had reviewed and sent NA, NA of which have been
 reviewed, accepted, and posted (Fig. 2-3, Table 2). \emph{{[}DETAILS{]}}

### 2022-04-14 doc/manuscript/ESSDmanuscript.Rmd
setup to automatically retrieve latest map figure from ForC
@mawilliams99 , this requires a path to the ForC directory. I put in a placeholder (currently line 122) where you can enter that path so that the code will work on your machine.

@@
 path_to_dir <- dirname(dirname(getwd())) # get path to repo 
+path_to_ForC= "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db" #Krista's laptop
+# path_to_ForC= "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db" #Madison, to knit, you'll need to enter the path to ForC-db on your computer, then un-comment this line.
 
 # MOVE FIGURE FILES 
-#traditional comparison
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub//Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/IPCC-EFDB-integration/doc/manuscript/figures_tables", "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE))) # first remove old files
-flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/ForC/figures/", "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/IPCC-EFDB-integration/doc/manuscript/figures_tables")
+#world map from GitHub
+do.call(file.remove, list(list.files(paste0(path_to_dir, "/doc/manuscript/figures_tables"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE))) # first remove old file
+flist <- list.files(paste0(path_to_ForC, "/ForC/figures/"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE)
+file.copy(flist, paste0(path_to_dir, "/doc/manuscript/figures_tables"))
 

### 2022-04-14 doc/manuscript/ESSDmanuscript.Rmd
update figure captions
#33, #34

@@
 
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} PLACEHOLDER.
-We need to add records transferred to EFDB.
-We also need to limit map to variables relavent to EFDB.
-It would also be good to add an underlying map of FAP ecozones"}
+```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+Underlying map shows FAO ecozones, which are coded as follows: ...."}
 #this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
@@
 
-```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Historgram of number of records in ForC relavent to (grey) and transferred to (black) EFDB.} PLACEHOLDER.
-We need to add axis labels and the two classes of records.
+```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} Codes are as follows: ....
 "}

### 2022-04-14 doc/manuscript/ESSDmanuscript.tex
update figure captions
#33, #34

@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} PLACEHOLDER.
-We need to add records transferred to EFDB.
-We also need to limit map to variables relavent to EFDB.
-It would also be good to add an underlying map of FAP ecozones}\label{fig:fig_map}
+\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+Underlying map shows FAO ecozones, which are coded as follows: ....}\label{fig:fig_map}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Historgram of number of records in ForC relavent to (grey) and transferred to (black) EFDB.} PLACEHOLDER.
-We need to add axis labels and the two classes of records.
+\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} Codes are as follows: ....
 }\label{fig:fig_dominant_veg}

### 2022-04-14 doc/manuscript/ESSDmanuscript.Rmd
misc small changes

@@
 Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
-#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
+#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
 knitr::include_graphics("figures_tables/C_variable_mapping.png")
@@
 
-## ForC restructuring
-
 ## Quality control measures
 Prior to releasing ForC v4.0, we executed several quality control measures.
-First, we implemented a system of continuous integration using GitHub Actions (*sensu* Kim et al. in prep) to run some automatic checks any time the master data files are updated.
+First, we implemented a system of continuous integration using GitHub Actions (*sensu* Kim et al. in review) to run some automatic checks any time the master data files are updated.
 Second, to improve information on geographic coordinates, we flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
@@
 Henceforth, we refer to the records with duplicates removed as "independent records".
+
+## Manual review of records to be sent to EFDB
+
+EFDB data submissions require information that was not recorded in previous versions of ForC.
+It was therefore necessary to return to original publications to retrieve relevant information, including ...
 
@@
 
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
 Underlying map shows FAO ecozones, which are coded as follows: ...."}
-#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
+#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
 knitr::include_graphics("figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png")
@@
 "}
-#this is for a two-column figure. for one-column figure, use {r, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
+#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
 knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")

### 2022-04-14 doc/manuscript/ESSDmanuscript.tex
misc small changes

@@
 
-\subsection{ForC restructuring}
-
 \subsection{Quality control measures}
@@
 measures. First, we implemented a system of continuous integration using
-GitHub Actions (\emph{sensu} Kim et al.~in prep) to run some automatic
+GitHub Actions (\emph{sensu} Kim et al.~in review) to run some automatic
 checks any time the master data files are updated. Second, to improve
@@
 
+\subsection{Manual review of records to be sent to EFDB}
+
+EFDB data submissions require information that was not recorded in
+previous versions of ForC. It was therefore necessary to return to
+original publications to retrieve relevant information, including
+\ldots{}
+
 \section{Results}
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relavent to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
 Underlying map shows FAO ecozones, which are coded as follows: ....}\label{fig:fig_map}

### 2022-04-14 doc/manuscript/ESSDmanuscript.Rmd
update variables table template
#35

@@
 names(C_variables_table) <- stringr::str_replace_all(names(C_variables_table), "\\.", " ")
+C_variables_table[is.na(C_variables_table)]<- "" #replace NAs with blanks
 
@@
 knitr::kable(C_variables_table, format = "latex",
-             longtable = FALSE,
+             longtable = TRUE,
              caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
@@
   kableExtra::kable_styling(full_width = TRUE) %>%
-  row_spec(0, bold = TRUE)
+  row_spec(0, bold = TRUE)  %>%
+  row_spec(1, bold = TRUE)  %>% #biomass
+  row_spec(5, bold = TRUE)  %>% #agb
+  row_spec(13, bold = TRUE) %>% #root
+  row_spec(21, bold = TRUE)  %>% #dead wood
+  row_spec(28, bold = TRUE)  %>% #litter
+  row_spec(33, bold = TRUE)  %>% #som
+  row_spec(37, bold = TRUE) #total
 ```

### 2022-04-14 doc/manuscript/ESSDmanuscript.tex
update variables table template
#35

@@
 
-\begin{table}
-
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
 \caption{\label{tab:table_variables}\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
-}
-\centering
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+}\\
 \hline
-\textbf{variable} & \textbf{n in ForC} & \textbf{n independent records in ForC} & \textbf{n reviewed} & \textbf{n sent to EFDB} & \textbf{n posted to EFDB}\\
+\textbf{variable} & \textbf{n in ForC} & \textbf{n independent records in ForC} & \textbf{n reviewed} & \textbf{n sent to EFDB}\\
 \hline
-STOCKS & NA & NA & NA & NA & NA\\
+\textbf{Biomass} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
 \hline
-biomass & NA & NA & NA & NA & NA\\
+biomass &  &  &  & \\
 \hline
-biomass\_ag & NA & NA & NA & NA & NA\\
+delta.biomass &  &  &  & \\
 \hline
-biomass\_ag\_foliage & NA & NA & NA & NA & \vphantom{1} NA\\
+NPP\_woody &  &  &  & \\
 \hline
-biomass\_ag\_woody & NA & NA & NA & NA & NA\\
+\textbf{Aboveground biomass} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
 \hline
-biomass\_root & NA & NA & NA & NA & NA\\
+biomass\_ag &  &  &  & \\
 \hline
-biomass\_root\_fine & NA & NA & NA & NA & NA\\
+biomass\_ag\_woody &  &  &  & \\
 \hline
-biomass\_root\_coarse & NA & NA & NA & NA & NA\\
+biomass\_ag\_foliage &  &  &  & \\
 \hline
-deadwood & NA & NA & NA & NA & NA\\
+delta.agb &  &  &  & \\
 \hline
-deadwood\_standing & NA & NA & NA & NA & NA\\
+ANPP\_woody\_stem &  &  &  & \\
 \hline
-deadwood\_down & NA & NA & NA & NA & NA\\
+ANPP\_foliage &  &  &  & \\
 \hline
-O.horizon & NA & NA & NA & NA & NA\\
+woody\_mortality &  &  &  & \\
 \hline
-SOM / SOC & NA & NA & NA & NA & NA\\
+\textbf{Belowground biomass} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
 \hline
-INCREMENTS & NA & NA & NA & NA & NA\\
+biomass\_root &  &  &  & \\
 \hline
-delta.biomass & NA & NA & NA & NA & NA\\
+biomass\_root\_fine &  &  &  & \\
 \hline
-delta.agb & NA & NA & NA & NA & NA\\
+biomass\_root\_coarse &  &  &  & \\
 \hline
-biomass\_ag\_foliage & NA & NA & NA & NA & NA\\
+delta.biomass\_root &  &  &  & \\
 \hline
-delta.biomass\_root & NA & NA & NA & NA & NA\\
+BNPP\_root &  &  &  & \\
 \hline
-delta.deadwood & NA & NA & NA & NA & NA\\
+BNPP\_root\_fine &  &  &  & \\
 \hline
-delta.deadwood\_standing & NA & NA & NA & NA & NA\\
+BNPP\_root\_coarse &  &  &  & \\
 \hline
-delta.deadwood\_down & NA & NA & NA & NA & NA\\
+\textbf{Dead wood} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
 \hline
-delta.O.horizon & NA & NA & NA & NA & NA\\
+deadwood &  &  &  & \\
 \hline
-delta.SOM / delta.SOC & NA & NA & NA & NA & NA\\
+deadwood\_standing &  &  &  & \\
 \hline
-FLUXES & NA & NA & NA & NA & NA\\
+deadwood\_down &  &  &  & \\
 \hline
-NPP\_woody & NA & NA & NA & NA & NA\\
+delta.deadwood &  &  &  & \\
 \hline
-ANPP\_woody\_stem & NA & NA & NA & NA & NA\\
+delta.deadwood\_standing &  &  &  & \\
 \hline
-ANPP\_foliage & NA & NA & NA & NA & NA\\
+delta.deadwood\_down &  &  &  & \\
 \hline
-BNPP\_root & NA & NA & NA & NA & NA\\
+\textbf{Litter} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
 \hline
-BNPP\_root\_fine & NA & NA & NA & NA & NA\\
+O.horizon &  &  &  & \\
 \hline
-BNPP\_root\_coarse & NA & NA & NA & NA & NA\\
+delta.O.horizon &  &  &  & \\
 \hline
-woody\_mortality & NA & NA & NA & NA & NA\\
+NPP\_litterfall &  &  &  & \\
 \hline
-NPP\_litterfall & NA & NA & NA & NA & NA\\
+R\_het\_litter &  &  &  & \\
 \hline
-R\_het\_litter & NA & NA & NA & NA & NA\\
+\textbf{Soil organic matter} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
 \hline
-R\_het\_soil & NA & NA & NA & NA & NA\\
+SOM / SOC &  &  &  & \\
 \hline
-TOTAL & NA & NA & NA & NA & NA\\
+delta.SOM / delta.SOC &  &  &  & \\
 \hline
-\end{tabu}
-\end{table}
+R\_het\_soil &  &  &  & \\
+\hline
+\textbf{TOTAL} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
+\hline
+\end{longtabu}
 

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
litter section

@@
 This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
-Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_202] implies that it is useful for practitioners to have access to locally-specific information, when available.
+Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.
 
@@
 
-Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].
+Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
 IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
@@
 
-Belowground biomass is defined as "all biomass of live roots" [@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
+Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
 The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain.
@@
 
-Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_2006_2006].
+Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_good_2003;@ipcc_2006_2006].
 This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
@@
 
-Litter is defined as including " all  non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil.
-This includes litter (OL),  fumic (OF),  and  humic (OH) layers.
-Live  fine  roots  (of  less  than  the  suggested  diameter  limit for  belowground  biomass) are included in litter where they cannot be distinguished from it empirically." (2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
-IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
+Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
+As noted above, live fine roots may be included in litter when difficult to separate empirically.
+The definition includes the entire O horizion, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
+This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
+Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
 
@@
 applied consistently through the time series.
-Live fine roots (of less than the suggested diameter limit for belowground biomass) are included with soil organic matter where they cannot be distinguished from it empirically."(2003 IPCc GPG for LULUCF (https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf_files/Glossary_Acronyms_BasicInfo/Glossary.pdf)
+Live fine roots (of less than the suggested diameter limit for belowground biomass) are included with soil organic matter where they cannot be distinguished from it empirically." [@ipcc_good_2003;@ipcc_2006_2006].
 
@@
   row_spec(28, bold = TRUE)  %>% #litter
-  row_spec(33, bold = TRUE)  %>% #som
-  row_spec(37, bold = TRUE) #total
+  row_spec(35, bold = TRUE)  %>% #som
+  row_spec(39, bold = TRUE) #total
 ```

### 2022-04-15 doc/manuscript/ESSDmanuscript.tex
litter section

@@
 variability of forest C cycling within ecozones
-\citep[e.g.,][]{cook-patton_mapping_2020, cuni-sanchez_high_202} implies
-that it is useful for practitioners to have access to locally-specific
-information, when available.
+\citep[e.g.,][]{cook-patton_mapping_2020, cuni-sanchez_high_2021}
+implies that it is useful for practitioners to have access to
+locally-specific information, when available.
 
@@
 \hline
-litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter (= max diameter for deadwood) & 10 cm default, but may be chosen by country\\
+litter & all non-living biomass, within specified size limits, in various states of decomposition above or within the mineral or organic soil & max diameter (= min diameter for deadwood) & 10 cm default, but may be chosen by country\\
 \hline
- &  & layers included & includes entire O horizon\\
+ &  & min size (= size limit for soil organic matter) & ?\\
+\hline
+ &  & layers included & entire O horizon: litter (OL),  fumic (OF),  and  humic (OH) layers\\
 \hline
@@
 biomass of living vegetation above the soil including stems, stumps,
-branches, bark, seeds, and foliage'' \citep{ipcc_2006_2006}. IPCC's
-guidance is that the understory may be excluded the understory if it
-constitutes a ``minor'' component, \emph{where quantitative definitions
-of ``understory'' and ``minor'' are not provided}, but where a commonly
-applied minimum size sampling threshold for mature forests would be 10
-cm stem diameter at breast height (DBH). A recent study characterizing
-the contributions of trees in different DBH classes to ecosystem C
-stocks and fluxes found that trees 1 - 10 cm DBH contributed up to
-\textasciitilde8\% aboveground biomass, \textasciitilde17\% aboveground
-woody net primary productivity (\(ANPP_{woody.stem}\)), and
-\textasciitilde20\% woody mortality (\(M_{woody}\)) of mature
-closed-canopy forests worldwide \citep{piponiot_distribution_2022}. In
-regrowth forests, woodlands, or savannas, small trees and shrubs
-contribute a much larger proportion of C stocks and fluxes
-\citep{piponiot_distribution_2022, refs}, and, correspondingly, biomass
-estimates for these ecosystems tend include smaller size classes
-\citep[e.g.,][]{refs}. Beyond the minimum DBH sampled, forest censuses
-and biomass estimates also differ in their inclusion of life forms other
-than dicot trees -- including lianas, ferns, palms, and bamboo -- which
-in some places can reach large sizes and/or constitute a large fraction
-of forest C. \emph{{[}explain IPCC guidance{]}} Further, it is important
-to note that this excludes standing dead wood, which is included in
-remote sensing biomass estimates \citep{duncanson_aboveground_2021}.
+branches, bark, seeds, and foliage''
+\citep{ipcc_good_2003, ipcc_2006_2006}. IPCC's guidance is that the
+understory may be excluded the understory if it constitutes a ``minor''
+component, \emph{where quantitative definitions of ``understory'' and
+``minor'' are not provided}, but where a commonly applied minimum size
+sampling threshold for mature forests would be 10 cm stem diameter at
+breast height (DBH). A recent study characterizing the contributions of
+trees in different DBH classes to ecosystem C stocks and fluxes found
+that trees 1 - 10 cm DBH contributed up to \textasciitilde8\%
+aboveground biomass, \textasciitilde17\% aboveground woody net primary
+productivity (\(ANPP_{woody.stem}\)), and \textasciitilde20\% woody
+mortality (\(M_{woody}\)) of mature closed-canopy forests worldwide
+\citep{piponiot_distribution_2022}. In regrowth forests, woodlands, or
+savannas, small trees and shrubs contribute a much larger proportion of
+C stocks and fluxes \citep{piponiot_distribution_2022, refs}, and,
+correspondingly, biomass estimates for these ecosystems tend include
+smaller size classes \citep[e.g.,][]{refs}. Beyond the minimum DBH
+sampled, forest censuses and biomass estimates also differ in their
+inclusion of life forms other than dicot trees -- including lianas,
+ferns, palms, and bamboo -- which in some places can reach large sizes
+and/or constitute a large fraction of forest C. \emph{{[}explain IPCC
+guidance{]}} Further, it is important to note that this excludes
+standing dead wood, which is included in remote sensing biomass
+estimates \citep{duncanson_aboveground_2021}.
 
@@
 Belowground biomass is defined as ``all biomass of live roots''
-\citep{ipcc_2006_2006}, a definition including both coarse roots, whose
-biomass is typically estimated based on stem censuses and allometries or
-belowground to aboveground biomass ratios, and fine roots, whose biomass
-is typically estimated via extraction of roots from soil samples. The
-former, which is typically \textless40 t C ha\textsuperscript{-1}
-\citep{anderson-teixeira_carbon_2021}, is methodologically linked to
-aboveground biomass estimates, sharing the same methodological sources
-of variation, but tending to be far more uncertain. Fine root biomass
-generally constitutes a much smaller C pool \citep[typically \textless5
-t C ha\textsuperscript{-1},][]{anderson-teixeira_carbon_2021}, and IPCC
+\citep{ipcc_good_2003, ipcc_2006_2006}, a definition including both
+coarse roots, whose biomass is typically estimated based on stem
+censuses and allometries or belowground to aboveground biomass ratios,
+and fine roots, whose biomass is typically estimated via extraction of
+roots from soil samples. The former, which is typically \textless40 t C
+ha\textsuperscript{-1} \citep{anderson-teixeira_carbon_2021}, is
+methodologically linked to aboveground biomass estimates, sharing the
+same methodological sources of variation, but tending to be far more
+uncertain. Fine root biomass generally constitutes a much smaller C pool
+\citep[typically \textless5 t C
+ha\textsuperscript{-1},][]{anderson-teixeira_carbon_2021}, and IPCC
 guidance is that it can be excluded when fine roots cannot be
@@
 non-living woody biomass not contained in the litter, either standing,
-lying on the ground, or in the soil'' \citep{ipcc_2006_2006}. This pool
-includes standing and fallen dead wood, stumps, and dead roots of
-diameter ≥10 cm (or a diameter specified by the country). While dead
-wood stocks and fluxes can be quite variable across forests
+lying on the ground, or in the soil''
+\citep{ipcc_good_2003, ipcc_2006_2006}. This pool includes standing and
+fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a
+diameter specified by the country). While dead wood stocks and fluxes
+can be quite variable across forests
 \citep{anderson-teixeira_carbon_2021}, and can at times be the dominant
@@
 
-Litter is defined as including " all non-living biomass with a diameter
-less than a minimum diameter chosen by the country (for example 10 cm),
-lying dead, in various states of decomposition above the mineral or
-organic soil. This includes litter (OL), fumic (OF), and humic (OH)
-layers. Live fine roots (of less than the suggested diameter limit for
-belowground biomass) are included in litter where they cannot be
-distinguished from it empirically." (2003 IPCc GPG for LULUCF
-(https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
-IPCC's latest Tier 1 default values are based on just 1-7 references per
-climate zone \citep[Table 2.2 in][]{ipcc_2019_2019}.
+Litter, which is typically \textless40 t C ha\textsuperscript{-1} but
+can exceed 100 t C ha\textsuperscript{-1}
+\citep{anderson-teixeira_carbon_2021}, is defined by IPCCC as including
+``all non-living biomass with a diameter less than a minimum diameter
+chosen by the country (for example 10 cm), lying dead, in various states
+of decomposition above the mineral or organic soil''
+\citep{ipcc_good_2003, ipcc_2006_2006}. As noted above, live fine roots
+may be included in litter when difficult to separate empirically. The
+definition includes the entire O horizion, including litter (OL), fumic
+(OF), and humic (OH) layers, in addition to litter embedded within the
+soil. This definition contrasts with empirical studies that focus on
+aboveground litter, often including only the OL layer in the definition
+of litter, and do not always specify the components included. Similar to
+dead wood, litter is poorly characterized in large-scale forest C
+budgets \citep{pan_large_2011, harris_global_2021}, and IPCC's latest
+Tier 1 default values are based on just 1-7 references per climate zone
+\citep[Table 2.2 in][]{ipcc_2019_2019}.
 
@@
 biomass) are included with soil organic matter where they cannot be
-distinguished from it empirically.''(2003 IPCc GPG for LULUCF
-(https://www.ipcc-nggip.iges.or.jp/public/gpglulucf/gpglulucf\_files/Glossary\_Acronyms\_BasicInfo/Glossary.pdf)
+distinguished from it empirically.''
+\citep{ipcc_good_2003, ipcc_2006_2006}.
 
@@
 
-As of April 14, 2022, ForC (v4.0) contained NA independent records (NA
+As of April 15, 2022, ForC (v4.0) contained NA independent records (NA
 total), NA of which were for variables relevant to EFDB (Fig. 1). These
@@
 \hline
+litter &  &  &  & \\
+\hline
+delta.litter &  &  &  & \\
+\hline
 NPP\_litterfall &  &  &  & \\
@@
 
-As of April 14, 2022, we had reviewed and sent NA, NA of which have been
+As of April 15, 2022, we had reviewed and sent NA, NA of which have been
 reviewed, accepted, and posted (Fig. 2-3, Table 2). \emph{{[}DETAILS{]}}

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
update variable mapping
added woody.mortality and woody.mortality_root. Still need to add these to ForC_variables.

@@
 
-```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
-Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
+```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., min DBH).
+Additional caveats are as follows: 1- often excludes branch fall, assumes that change in foliage biomass is negligible (see note 7); 2- excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- excludes belowground components;  4-excludes breakage into pieces less than dead wood threshold size; 5- may exclude woody mortality of stems <10 cm DBH and branch fall, decomposition of dead wood into sizes classified as litter; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
 #this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
@@
   row_spec(1, bold = TRUE)  %>% #biomass
-  row_spec(5, bold = TRUE)  %>% #agb
-  row_spec(13, bold = TRUE) %>% #root
-  row_spec(21, bold = TRUE)  %>% #dead wood
-  row_spec(28, bold = TRUE)  %>% #litter
-  row_spec(35, bold = TRUE)  %>% #som
-  row_spec(39, bold = TRUE) #total
+  row_spec(6, bold = TRUE)  %>% #agb
+  row_spec(14, bold = TRUE) %>% #root
+  row_spec(25, bold = TRUE)  %>% #dead wood
+  row_spec(33, bold = TRUE)  %>% #litter
+  row_spec(40, bold = TRUE)  %>% #som
+  row_spec(44, bold = TRUE) #total
 ```

### 2022-04-15 doc/manuscript/ESSDmanuscript.tex
update variable mapping
added woody.mortality and woody.mortality_root. Still need to add these to ForC_variables.

@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., minimum DBH).
-Additional caveats are as follows: 1- assumes that change in foliage biomass is negligible (see note 7); 2- incomplete: excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- incomplete: excludes belowground components;  4-incomplete: excludes breakage into pieces less than dead wood threshold size; 5-incomplete: excludes woody mortality of stems <10 cm DBH, decomposition of dead wood (aboveground and coarse roots) into sizes classified as litter, may exclude branch fall; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:fig_variable_mapping}
+\includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., min DBH).
+Additional caveats are as follows: 1- often excludes branch fall, assumes that change in foliage biomass is negligible (see note 7); 2- excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- excludes belowground components;  4-excludes breakage into pieces less than dead wood threshold size; 5- may exclude woody mortality of stems <10 cm DBH and branch fall, decomposition of dead wood into sizes classified as litter; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:fig_variable_mapping}
 \end{figure}
@@
 \hline
+woody.mortality &  &  &  & \\
+\hline
 \textbf{Aboveground biomass} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
@@
 \hline
-ANPP\_woody\_stem &  &  &  & \\
+ANPP\_woody &  &  &  & \\
 \hline
@@
 \hline
-woody\_mortality &  &  &  & \\
+woody.mortality\_ag &  &  &  & \\
 \hline
@@
 \hline
-BNPP\_root &  &  &  & \\
+delta.biomass\_root\_coarse &  &  &  & \\
+\hline
+delta.biomass\_root\_fine &  &  &  & \\
+\hline
+woody.mortality\_root &  &  &  & \\
 \hline
 BNPP\_root\_fine &  &  &  & \\
+\hline
+BNPP\_root.turnover\_fine &  &  &  & \\
 \hline
@@
 delta.deadwood\_down &  &  &  & \\
+\hline
+R\_het\_deadwood & 0 &  &  & \\
 \hline

### 2022-04-15 scripts/1_ForC_to_IPCC_formatting.R
update histogram #34

@@ -78,7 +78,6 @@ ForC_simplified <- ForC_simplified[ForC_simplified$variable.name %in% V_mapping$
 ## grey is number of records in ForC that are relevant to EFDB (n records for the variables we send)
 ## black is what is already transferred to EFDB
 ## x axis is dominant vegetation type
-png("doc/manuscript/figures_tables/Histogram_n_Relevant_and_Transferred_Records.png", width = 8, height = 9, units = "in", res = 300)
 
 
 ForBarplot <- data.frame(data_type = rep(c("relevant", "transferred"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg), FAO.ecozone = c(ForC_simplified$FAO.ecozone, ForC_simplified_already_sent$FAO.ecozone), continent = c(ForC_simplified$continent, ForC_simplified_already_sent$continent), stand.age = c(ForC_simplified$stand.age, ForC_simplified_already_sent$stand.age))
@@ -94,10 +93,11 @@ ForBarplot$FAO.ecozone[is.na(ForBarplot$FAO.ecozone)] <- "Unclassified"
 # ForBarplot$continent <- gsub(" ", "\n", ForBarplot$continent)
 
 labels <- list(A = list("dominant.veg", "Dominant vegetation", F),
-               B = list("FAO.ecozone", "FAO ecozone", F),
-               C = list("continent", "Continent", T),
+               B = list("FAO.ecozone", "FAO ecozone", T),
+               C = list("continent", "Continent", F),
                D = list("stand.age", "Stand age", F))
 
+png("doc/manuscript/figures_tables/Histogram_n_Relevant_and_Transferred_Records.png", width = 8, height = 9, units = "in", res = 300)
 
 par(mfrow = c(2,2), mar = c(6,3,2.5,1), oma = c(1,2,0,0))
 for(i in 1:length(labels)){
@@ -106,32 +106,37 @@ for(i in 1:length(labels)){
   leg = labels[[i]][[3]]
   
   bp <- table(ForBarplot$data_type, ForBarplot[,n])
+  
+  if(n %in% "dominant.veg") {
[…]

### 2022-04-15 scripts/1_ForC_to_IPCC_formatting.R
change Broadleaf to Broadleaf (mixed E/D) in histogram
 (and use PFT description instead of description.individual)
FYI @teixeirak
#34

@@ -108,11 +108,13 @@ for(i in 1:length(labels)){
   bp <- table(ForBarplot$data_type, ForBarplot[,n])
   
   if(n %in% "dominant.veg") {
-    colnames(bp) <- PFT$description.individual[match(colnames(bp), PFT$pftcode)]
+    colnames(bp) <- PFT$description[match(colnames(bp), PFT$pftcode)]
     colnames(bp)[is.na(colnames(bp))] <- "NAC"
-    colnames(bp) <-  gsub("Tree, ", "", colnames(bp))
+    colnames(bp) <-  gsub("Trees, |conifer \\(|\\)|\\(.*\\)|broadleaf and/or conifer, | or.*$", "", colnames(bp))
+    colnames(bp) <-  gsub("conifer", "needleaf", colnames(bp))
     colnames(bp) <-  gsub("\\b(\\pL)", "\\U\\1", colnames(bp), perl = T)
-    colnames(bp) <-  gsub(", ", "\n", colnames(bp))
+    colnames(bp) <-  gsub(", | ", "\n", colnames(bp))
+    colnames(bp)[colnames(bp) %in% "Broadleaf"] <-  "Broadleaf\n(mixed E/D)"
     
   }
   

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
knit in new histogram, update caption
#34

@@
 
-```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} Codes are as follows: ....
-"}
+```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), "Other" includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
 #this is for a two-column figure. for one-column figure, use out.width = "8.3cm"

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
update variable mapping figure

--- 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
@@
 
+```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, and that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon."}
+#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
+knitr::include_graphics("figures_tables/C_variable_mapping.png")
+```
+
 ### Biomass
@@
 
-### Soil Organic Matter
+### Soil Organic Matter/ Carbon
 
-Soil organic matter is defined as "Includes organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and
-applied consistently through the time series.
-Live fine roots (of less than the suggested diameter limit for belowground biomass) are included with soil organic matter where they cannot be distinguished from it empirically." [@ipcc_good_2003;@ipcc_2006_2006].
-
-```{r fig_variable_mapping, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., min DBH).
-Additional caveats are as follows: 1- often excludes branch fall, assumes that change in foliage biomass is negligible (see note 7); 2- excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- excludes belowground components;  4-excludes breakage into pieces less than dead wood threshold size; 5- may exclude woody mortality of stems <10 cm DBH and branch fall, decomposition of dead wood into sizes classified as litter; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/C_variable_mapping.png")
-```
+Soil organic matter/ carbon (SOM/ SOC), which *(statistic on how much C is typial)* [@ref], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
+Live fine roots (suggested diameter cutoff of 2 mm) may be included with soil organic matter when it is not feasible to distinguish them empirically.
+The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.
 
@@
 Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
+Forest Land
 Definitions of forest are allowed to vary by country, but must be applied consistently.
 
-Managed land...
+Inventories are conducted only for managed land, which is defined by IPCC to include "all forests subject to some kind of human interactions (notably commercial management, harvest of industrial round-wood (logs) and fuelwood, production and use of wood commodities, and forest managed for amenity value or environmental protection if specified by the country), with defined geographical boundaries" [@ipcc_good_2003.
 
@@
 
-```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} Codes are as follows: ....
-"}
+```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
 #this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
@@
   row_spec(44, bold = TRUE)  #total
+
 ```

### 2022-04-15 doc/manuscript/ESSDmanuscript.tex
update variable mapping figure

@@
 \hline
-litter & all non-living biomass, within specified size limits, in various states of decomposition above or within the mineral or organic soil & max diameter (= min diameter for deadwood) & 10 cm default, but may be chosen by country\\
+litter & all non-living biomass smaller than dead wood but larger than soil organic matter, in various states of decomposition both above or within the mineral or organic soil & max diameter (= min diameter for deadwood) & 10 cm default, but may be chosen by country\\
 \hline
@@
 \end{table}
+
+\begin{figure}
+\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, and that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.}\label{fig:fig_variable_mapping}
+\end{figure}
 
@@
 
-\subsubsection{Soil Organic Matter}
-
-Soil organic matter is defined as ``Includes organic carbon in mineral
-and organic soils (including peat) to a specified depth chosen by the
-country and applied consistently through the time series. Live fine
-roots (of less than the suggested diameter limit for belowground
-biomass) are included with soil organic matter where they cannot be
-distinguished from it empirically.''
-\citep{ipcc_good_2003, ipcc_2006_2006}.
-
-\begin{figure}
-\includegraphics[width=15cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} In many cases, the match of ForC variables to IPCC criteria depends upon measurement protocols (e.g., min DBH).
-Additional caveats are as follows: 1- often excludes branch fall, assumes that change in foliage biomass is negligible (see note 7); 2- excludes large branch fall; also, under IPCC definitions, outflux from aboveground biomass should include all sizes, influx to deadwood should include only above the minimum diameter chosen for dead wood; 3- excludes belowground components;  4-excludes breakage into pieces less than dead wood threshold size; 5- may exclude woody mortality of stems <10 cm DBH and branch fall, decomposition of dead wood into sizes classified as litter; 6- measurements often limited to decomposition of relatively fine litter and may exclude branches and stems below the dead wood size threshold and/or the more decomposed layers of the O horizon; 7 - foliage production is generally measured by collecting leaf-fall, a method that assumes that the influx = outflux (foliage biomass is roughly constant year-to-year); 8 - excludes branch fall, which is necessary for a full accounting of woody productivity but is typically assumed negligible for calculations of net biomass change.}\label{fig:fig_variable_mapping}
-\end{figure}
+\subsubsection{Soil Organic Matter/ Carbon}
+
+Soil organic matter/ carbon (SOM/ SOC), which \emph{(statistic on how
+much C is typial)} \citep{ref}, is defined by IPCC as ``organic carbon
+in mineral and organic soils (including peat) to a specified depth
+chosen by the country and applied consistently through the time series''
+\citep{ipcc_good_2003, ipcc_2006_2006}. Live fine roots (suggested
+diameter cutoff of 2 mm) may be included with soil organic matter when
+it is not feasible to distinguish them empirically. The greatest source
+of methodological variation in measuring SOM/ SOC is sampling depth,
+which has a suggested default of 30 cm but may vary by country provided
+that consistent criteria are applied.
 
@@
 category to another in the past 20 years (e.g., Cropland converted to
-Forest Land). Definitions of forest are allowed to vary by country, but
-must be applied consistently.
-
-Managed land\ldots{}
+Forest Land). Forest Land Definitions of forest are allowed to vary by
+country, but must be applied consistently.
+
+Inventories are conducted only for managed land, which is defined by
+IPCC to include ``all forests subject to some kind of human interactions
+(notably commercial management, harvest of industrial round-wood (logs)
+and fuelwood, production and use of wood commodities, and forest managed
+for amenity value or environmental protection if specified by the
+country), with defined geographical boundaries''
+{[}\citet{ipcc_good_2003}.
 
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} Codes are as follows: ....
-}\label{fig:fig_dominant_veg}
+\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_dominant_veg}
 \end{figure}

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
finish first draft of section 2

@@
 Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
-Forest Land
+Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ* [@ipcc_good_2003].
 Definitions of forest are allowed to vary by country, but must be applied consistently.
+Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
 
-Inventories are conducted only for managed land, which is defined by IPCC to include "all forests subject to some kind of human interactions (notably commercial management, harvest of industrial round-wood (logs) and fuelwood, production and use of wood commodities, and forest managed for amenity value or environmental protection if specified by the country), with defined geographical boundaries" [@ipcc_good_2003.
+Inventories are conducted only for managed land, which is defined by IPCC to include "all forests subject to some kind of human interactions (notably commercial management, harvest of industrial round-wood (logs) and fuelwood, production and use of wood commodities, and forest managed for amenity value or environmental protection if specified by the country), with defined geographical boundaries" [@ipcc_good_2003].
+As such, managed land constitutes all or the majority of Forest Land in most countries; however, many countries have not yet reported their approach for defining managed land [@ogle_delineating_2018; @deng_comparing_2021].
 

### 2022-04-15 doc/manuscript/ESSDmanuscript.tex
finish first draft of section 2

@@
 category to another in the past 20 years (e.g., Cropland converted to
-Forest Land). Forest Land Definitions of forest are allowed to vary by
-country, but must be applied consistently.
+Forest Land). Forest Land is defined as at least 10-30\% crown cover of
+trees with potential to reach a minimum height of 2-5 m \emph{in situ}
+\citep{ipcc_good_2003}. Definitions of forest are allowed to vary by
+country, but must be applied consistently. Forest Land includes land
+where vegetation temporarily falls below the threshold values for forest
+(e.g., due to disturbance), but is expected to exceed those thresholds
+in the future \citep{ipcc_good_2003}.
 
@@
 for amenity value or environmental protection if specified by the
-country), with defined geographical boundaries''
-{[}\citet{ipcc_good_2003}.
+country), with defined geographical boundaries'' \citep{ipcc_good_2003}.
+As such, managed land constitutes all or the majority of Forest Land in
+most countries; however, many countries have not yet reported their
+approach for defining managed land
+\citep{ogle_delineating_2018, deng_comparing_2021}.
 

### 2022-04-15 doc/manuscript/ESSDmanuscript.Rmd
start on section 3

@@
 # Mapping ForC to EFDB
-ForC data is incredibly valuable to EFDB and there is data which is included in the ForC database that does not meet EFDB standards.
-There were two main EFDB guidelines which limits the amount of data we could transfer.
-EFDB will not accept data which has been digitized(from graph) and ForC does.
+
+The majority of information required by EFDB was present in previous versions of ForC [].
+We made modest structural changes to ForC (Table ##) to allow complete records to be restructured for automated population of EFDB data entry forms using R ().
 
@@
 EFDB data submissions require information that was not recorded in previous versions of ForC.
-It was therefore necessary to return to original publications to retrieve relevant information, including ...
+It was therefore necessary to return to original publications to retrieve relevant information, including the following:
+
+- Data location within source - Because EFDB will not accept data which has been digitized (from graphs) while ForC does, it was necessary to review and note the location of the data within the original source
 

### 2022-04-15 doc/manuscript/ESSDmanuscript.tex
start on section 3

@@
 
-ForC data is incredibly valuable to EFDB and there is data which is
-included in the ForC database that does not meet EFDB standards. There
-were two main EFDB guidelines which limits the amount of data we could
-transfer. EFDB will not accept data which has been digitized(from graph)
-and ForC does.
+The majority of information required by EFDB was present in previous
+versions of ForC {[}{]}. We made modest structural changes to ForC
+(Table \#\#) to allow complete records to be restructured for automated
+population of EFDB data entry forms using R ().
 
@@
 previous versions of ForC. It was therefore necessary to return to
-original publications to retrieve relevant information, including
-\ldots{}
+original publications to retrieve relevant information, including the
+following:
+
+\begin{itemize}
+\tightlist
+\item
+  Data location within source - Because EFDB will not accept data which
+  has been digitized (from graphs) while ForC does, it was necessary to
+  review and note the location of the data within the original source
+\end{itemize}
 

### 2022-04-16 doc/manuscript/ESSDmanuscript.Rmd
rearrangement - switch sections 3 & 4

@@
 
-# Mapping ForC to EFDB
-
-The majority of information required by EFDB was present in previous versions of ForC [].
-We made modest structural changes to ForC (Table ##) to allow complete records to be restructured for automated population of EFDB data entry forms using R ().
-
-## Carbon cycle variables
-
-Mapping of variables is shown in Fig. `r fig_variable_mapping`.
-
-## Land use categories
-
-Documented at https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/defining_land_subcategory.md, https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/IPCC_LandUse_mapping.csv, and in [issue #8](https://github.com/forc-db/IPCC_database_integration/issues/8).
-
-The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*].
-This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 
-Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).
-
 # Updates to ForC  (ForC v4.0)
 
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we defined ## new variables, implemented some modest restructuring, resolved duplicate records, and conducted quality control.
+Previous versions of ForC [@refs] contained most of the information required for transfer to EFDB, and, more broadly, to inform C accounting under IPCC guidelines. 
+However, modest changes to the structure and contents of ForC were needed in order to provide all information fequired by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added ## new fields, defined ## new variables, implemented some modest restructuring (Table ##), resolved duplicate records, and conducted quality control.
 This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].
+
+## Adding new fields
 
@@
 
-- Data location within source - Because EFDB will not accept data which has been digitized (from graphs) while ForC does, it was necessary to review and note the location of the data within the original source
+*Data location within source*-  Because EFDB will not accept data which has been digitized (from graphs) while ForC does, it was necessary to review and note the location of the data within the original source
+
+# ForC to EFDB data transfer
+
+To transfer data from ForC into the EFDB, we populated EFDB's bulk import form ("EFDB bulk import.xlsx"), using R scripts to restructure complete, reviewed records for entry into this form.
+
+## Mapping ForC to EFDB
+### Carbon cycle variables
+
+Mapping of variables is shown in Fig. `r fig_variable_mapping`.
+
+### Land use categories
+
+Documented at https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/defining_land_subcategory.md, https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/IPCC_LandUse_mapping.csv, and in [issue #8](https://github.com/forc-db/IPCC_database_integration/issues/8).
+
+The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*].
+This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 
+Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).
+
+## Review and posting of data
 
@@
   row_spec(44, bold = TRUE)  #total
-
 ```

### 2022-04-16 doc/manuscript/ESSDmanuscript.tex
rearrangement - switch sections 3 & 4

@@
 
-\section{Mapping ForC to EFDB}
-
-The majority of information required by EFDB was present in previous
-versions of ForC {[}{]}. We made modest structural changes to ForC
-(Table \#\#) to allow complete records to be restructured for automated
-population of EFDB data entry forms using R ().
-
-\subsection{Carbon cycle variables}
+\section{Updates to ForC (ForC v4.0)}
+
+Previous versions of ForC \citep{refs} contained most of the information
+required for transfer to EFDB, and, more broadly, to inform C accounting
+under IPCC guidelines. However, modest changes to the structure and
+contents of ForC were needed in order to provide all information
+fequired by EFDB and to improve ForC's capacity to serve as a repository
+of valuable information for forest C accounting under IPCC guidelines.
+To support export of data to EFDB, and to improve the overall quality of
+the ForC database, we added \#\# new fields, defined \#\# new variables,
+implemented some modest restructuring (Table \#\#), resolved duplicate
+records, and conducted quality control. This section describes changes
+relative to ForC v2.0 \citep{anderson-teixeira_forc_2018}.
+
+\subsection{Adding new fields}
+
+\subsection{Defining new variables}
+
+We added eleven increment variables to the set of named and defined
+variables (or 22, counting \_OM and \_C versions), which previously
+included only one (aboveground biomass increment, \emph{delta.agb}).
+\emph{(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)} These
+are directly related to C stocks as previously defined in ForC, with
+``\emph{delta.}'' added in front of the variable name.
+
+Although these variables currently lack records, the structure exists
+such that records can be populated over time.
+
+To provide better definition of the previously existing variable
+\emph{organic.layer}, which has a nebulous definition that reflects the
+varied definitions adopted by original studies, we added two clearly
+defined variables: \emph{litter} (relatively undecomposed plant
+material/ OL horizon), and \emph{O.horizon} (entire O-horizon, including
+\emph{litter} (OL)).
+
+\subsection{Quality control measures}
+
+Prior to releasing ForC v4.0, we executed several quality control
+measures. First, we implemented a system of continuous integration using
+GitHub Actions (\emph{sensu} Kim et al.~in review) to run some automatic
+checks any time the master data files are updated. Second, to improve
+information on geographic coordinates, we flagged and reviewed records
+with suspected low precision \emph{(Issue
+\#29){[}https://github.com/forc-db/ForC/issues/229{]}}. Third, to
+identify erroneous climate data\ldots{} \emph{(Issue
+\#212){[}https://github.com/forc-db/ForC/issues/212{]}}.
+
+\subsection{Resolving duplicates}
+
+Because ForC v4.0 contained known duplicate records, we used R scripts
+to remove likely duplicates, as detailed in
+\citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the
+records with duplicates removed as ``independent records''.
+
+\subsection{Manual review of records to be sent to EFDB}
+
+EFDB data submissions require information that was not recorded in
+previous versions of ForC. It was therefore necessary to return to
+original publications to retrieve relevant information, including the
+following:
+
+\emph{Data location within source}- Because EFDB will not accept data
+which has been digitized (from graphs) while ForC does, it was necessary
+to review and note the location of the data within the original source
+
+\section{ForC to EFDB data transfer}
+
+To transfer data from ForC into the EFDB, we populated EFDB's bulk
+import form (``EFDB bulk import.xlsx''), using R scripts to restructure
+complete, reviewed records for entry into this form.
+
+\subsection{Mapping ForC to EFDB}
+
+\subsubsection{Carbon cycle variables}
 
@@
 
-\subsection{Land use categories}
+\subsubsection{Land use categories}
 
@@
 
-\section{Updates to ForC (ForC v4.0)}
-
-To support export of data to EFDB, and to improve the overall quality of
-the ForC database, we defined \#\# new variables, implemented some
-modest restructuring, resolved duplicate records, and conducted quality
-control. This section describes changes relative to ForC v2.0
-\citep{anderson-teixeira_forc_2018}.
-
-\subsection{Defining new variables}
-
-We added eleven increment variables to the set of named and defined
-variables (or 22, counting \_OM and \_C versions), which previously
-included only one (aboveground biomass increment, \emph{delta.agb}).
-\emph{(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)} These
-are directly related to C stocks as previously defined in ForC, with
-``\emph{delta.}'' added in front of the variable name.
-
-Although these variables currently lack records, the structure exists
-such that records can be populated over time.
-
-To provide better definition of the previously existing variable
-\emph{organic.layer}, which has a nebulous definition that reflects the
-varied definitions adopted by original studies, we added two clearly
-defined variables: \emph{litter} (relatively undecomposed plant
-material/ OL horizon), and \emph{O.horizon} (entire O-horizon, including
-\emph{litter} (OL)).
-
-\subsection{Quality control measures}
-
-Prior to releasing ForC v4.0, we executed several quality control
-measures. First, we implemented a system of continuous integration using
-GitHub Actions (\emph{sensu} Kim et al.~in review) to run some automatic
-checks any time the master data files are updated. Second, to improve
-information on geographic coordinates, we flagged and reviewed records
-with suspected low precision \emph{(Issue
-\#29){[}https://github.com/forc-db/ForC/issues/229{]}}. Third, to
-identify erroneous climate data\ldots{} \emph{(Issue
-\#212){[}https://github.com/forc-db/ForC/issues/212{]}}.
-
-\subsection{Resolving duplicates}
-
-Because ForC v4.0 contained known duplicate records, we used R scripts
-to remove likely duplicates, as detailed in
-\citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the
-records with duplicates removed as ``independent records''.
-
-\subsection{Manual review of records to be sent to EFDB}
-
-EFDB data submissions require information that was not recorded in
-previous versions of ForC. It was therefore necessary to return to
-original publications to retrieve relevant information, including the
-following:
-
-\begin{itemize}
-\tightlist
-\item
-  Data location within source - Because EFDB will not accept data which
-  has been digitized (from graphs) while ForC does, it was necessary to
-  review and note the location of the data within the original source
-\end{itemize}
+\subsection{Review and posting of data}
 
@@
 
-As of April 15, 2022, ForC (v4.0) contained NA independent records (NA
+As of April 16, 2022, ForC (v4.0) contained NA independent records (NA
 total), NA of which were for variables relevant to EFDB (Fig. 1). These
@@
 
-As of April 15, 2022, we had reviewed and sent NA, NA of which have been
+As of April 16, 2022, we had reviewed and sent NA, NA of which have been
 reviewed, accepted, and posted (Fig. 2-3, Table 2). \emph{{[}DETAILS{]}}

### 2022-04-16 doc/manuscript/ESSDmanuscript.Rmd
section 3

@@
 However, modest changes to the structure and contents of ForC were needed in order to provide all information fequired by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added ## new fields, defined ## new variables, implemented some modest restructuring (Table ##), resolved duplicate records, and conducted quality control.
-This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified ## fields (Table ##), defined ## new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
 
-## Adding new fields
+*This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].*
 
-## Defining new variables
+## New or modified fields
+
+## New variables
 
@@
 
-## Resolving duplicates
-
 Because ForC v4.0 contained known duplicate records, we used R scripts to remove likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
-Henceforth, we refer to the records with duplicates removed as "independent records".
+Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
+All records sent to EFCB were ensured to be original through manual review, as detailed below.
 
@@
 *Data location within source*-  Because EFDB will not accept data which has been digitized (from graphs) while ForC does, it was necessary to review and note the location of the data within the original source
+
+## Addition of new records
 

### 2022-04-16 doc/manuscript/ESSDmanuscript.tex
section 3

@@
 To support export of data to EFDB, and to improve the overall quality of
-the ForC database, we added \#\# new fields, defined \#\# new variables,
-implemented some modest restructuring (Table \#\#), resolved duplicate
-records, and conducted quality control. This section describes changes
-relative to ForC v2.0 \citep{anderson-teixeira_forc_2018}.
-
-\subsection{Adding new fields}
-
-\subsection{Defining new variables}
+the ForC database, we added or modified \#\# fields (Table \#\#),
+defined \#\# new variables, implemented enhanced quality control,
+manually reviewed \textgreater\# records to obtain additional required
+information, and added \# new records.
+
+\emph{This section describes changes relative to ForC v2.0
+\citep{anderson-teixeira_forc_2018}.}
+
+\subsection{New or modified fields}
+
+\subsection{New variables}
 
@@
 
-\subsection{Resolving duplicates}
-
 Because ForC v4.0 contained known duplicate records, we used R scripts
 to remove likely duplicates, as detailed in
-\citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the
-records with duplicates removed as ``independent records''.
+\citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the set
+of records with likely duplicates removed as ``independent records''.
+All records sent to EFCB were ensured to be original through manual
+review, as detailed below.
 
@@
 to review and note the location of the data within the original source
+
+\subsection{Addition of new records}
 

### 2022-04-16 doc/manuscript/ESSDmanuscript.Rmd
describe new variables

@@
 n_records_posted_EFDB = NA # n records accepted and posted in EFDB
+
+n_new_variables = 11 + 2 + 2 #delta. + woody.mortality + O horizon
 ```
@@
 Previous versions of ForC [@refs] contained most of the information required for transfer to EFDB, and, more broadly, to inform C accounting under IPCC guidelines. 
-However, modest changes to the structure and contents of ForC were needed in order to provide all information fequired by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified ## fields (Table ##), defined ## new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
+However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified ## fields (Table ##), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
 
@@
 
-We added eleven increment variables to the set of named and defined variables (or 22, counting _OM and _C versions), which previously included only one (aboveground biomass increment, *delta.agb*). 
-*(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)*
+We added a total of `r n_new_variables` new variables to the set of named and defined variables, counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
+All of these variables are relevant to C accounting under IPCC, and hence to EFDB (Fig. `r fig_variable_mapping`)
+This included eleven increment variables, adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
 These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
+Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
+Although most of these variables currently lack records in ForC, the structure exists such that records can be populated over time.
 
-Although these variables currently lack records, the structure exists such that records can be populated over time.
-
-To provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including *litter* (OL)).
+Further, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including *litter* (OL)).
 

### 2022-04-16 doc/manuscript/ESSDmanuscript.tex
describe new variables

@@
 contents of ForC were needed in order to provide all information
-fequired by EFDB and to improve ForC's capacity to serve as a repository
+required by EFDB and to improve ForC's capacity to serve as a repository
 of valuable information for forest C accounting under IPCC guidelines.
@@
 the ForC database, we added or modified \#\# fields (Table \#\#),
-defined \#\# new variables, implemented enhanced quality control,
-manually reviewed \textgreater\# records to obtain additional required
+defined 15 new variables, implemented enhanced quality control, manually
+reviewed \textgreater\# records to obtain additional required
 information, and added \# new records.
@@
 
-We added eleven increment variables to the set of named and defined
-variables (or 22, counting \_OM and \_C versions), which previously
-included only one (aboveground biomass increment, \emph{delta.agb}).
-\emph{(https://github.com/forc-db/IPCC-EFDB-integration/issues/6)} These
-are directly related to C stocks as previously defined in ForC, with
-``\emph{delta.}'' added in front of the variable name.
-
-Although these variables currently lack records, the structure exists
-such that records can be populated over time.
-
-To provide better definition of the previously existing variable
-\emph{organic.layer}, which has a nebulous definition that reflects the
-varied definitions adopted by original studies, we added two clearly
-defined variables: \emph{litter} (relatively undecomposed plant
-material/ OL horizon), and \emph{O.horizon} (entire O-horizon, including
-\emph{litter} (OL)).
+We added a total of 15 new variables to the set of named and defined
+variables, counting each pair of variables with units in C (ending in
+\texttt{\_C}) or organic matter (ending in \texttt{\_OM}) as one. All of
+these variables are relevant to C accounting under IPCC, and hence to
+EFDB (Fig. 1) This included eleven increment variables, adding to only
+one previously defined increment variable (aboveground biomass
+increment, \emph{delta.agb}). These are directly related to C stocks as
+previously defined in ForC, with ``\emph{delta.}'' added in front of the
+variable name. Further, we added variables capturing the belowground
+component of woody mortality (\emph{woody.mortality\_root}) and the
+combined aboveground and belowground components of woody mortality
+(\emph{woody.mortality}). Although most of these variables currently
+lack records in ForC, the structure exists such that records can be
+populated over time.
+
+Further, to provide better definition of the previously existing
+variable \emph{organic.layer}, which has a nebulous definition that
+reflects the varied definitions adopted by original studies, we added
+two clearly defined variables: \emph{litter} (relatively undecomposed
+plant material/ OL horizon), and \emph{O.horizon} (entire O-horizon,
+including \emph{litter} (OL)).
 

### 2022-04-16 doc/manuscript/ESSDmanuscript.Rmd
appendix arrangement

@@
   Funding for this study was provided by Bezos Earth Fund to The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
-appendix: |
-   \section{Mapping ForC to EFDB}
-
-   CURRENT TABLE 3 GOES HERE
-
-   \section{Updates to ForC}
-
-   CURRENT TABLE 4 GOES HERE
-
+appendix: 
 output:
@@
     base_format: rticles::copernicus_article # for using bookdown features like \@ref()
+
+    header-includes:
+    - \usepackage{caption}
 ---
@@
 section_land_use_categories = 2.2
+
+#Appendices
+table_ForCchanges = "Appendix A"
+table_ForCfieldmapping = "Appendix B"
 
@@
 However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified ## fields (Table ##), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified ## fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
 
@@
 
+(`r table_ForCchanges`)
+
 ## New variables
@@
 We added a total of `r n_new_variables` new variables to the set of named and defined variables, counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
-All of these variables are relevant to C accounting under IPCC, and hence to EFDB (Fig. `r fig_variable_mapping`)
+All of these variables are relevant to C accounting under IPCC, and hence to EFDB (Fig. `r fig_variable_mapping`).
 This included eleven increment variables, adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
@@
 ## Mapping ForC to EFDB
+
+`r table_ForCfieldmapping`
+
 ### Carbon cycle variables
@@
 \clearpage
-# Appendix A 
-<!--remove number-->
+# Appendix A. Updates to ForC {-}
 
-```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-
-ForC_variables_included <- read.csv("figures_tables/ForC_variables_included.csv")
-names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
-ForC_variables_included <- ForC_variables_included[c(1:5)] #include all fields but TO_DO
-
-knitr::kable(ForC_variables_included, format = "latex",
-             longtable = TRUE,
-             caption = "\\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). ") %>%
-  row_spec(0, bold = TRUE) %>%
-kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
-
-```
-'Required' field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
-
-** '(yes)' for most recent severe disturbance; 'no' for other history events
-
-\clearpage
-# Appendix B
-<!--remove number-->
-
+\captionsetup[table]{labelformat=empty}
+Table A1: **Table of changes to ForC fields.**
 ```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
@@
 knitr::kable(record_of_changes, format = "latex",
-             longtable = TRUE,
-             caption = "\\textbf{Table of changes to ForC fields.}") %>%
+             longtable = TRUE) %>%
   kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
@@
 ```
+
+\clearpage
+# Appendix B. Mapping ForC to EFDB {-}
+
+Table B1: **Mapping of ForC fields to EFDB.** See footnotes at end of table (STILL NEED TO BE PROPERLY INSERTED).
+```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+
+ForC_variables_included <- read.csv("figures_tables/ForC_variables_included.csv")
+names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
+ForC_variables_included <- ForC_variables_included[c(1:5)] #include all fields but TO_DO
+
+knitr::kable(ForC_variables_included, format = "latex",
+             longtable = TRUE) %>%
+  row_spec(0, bold = TRUE) %>%
+kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
+
+```
+'Required' field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
+
+** '(yes)' for most recent severe disturbance; 'no' for other history events

### 2022-04-16 doc/manuscript/ESSDmanuscript.tex
appendix arrangement

@@
 To support export of data to EFDB, and to improve the overall quality of
-the ForC database, we added or modified \#\# fields (Table \#\#),
+the ForC database, we added or modified \#\# fields (Appendix A),
 defined 15 new variables, implemented enhanced quality control, manually
@@
 \subsection{New or modified fields}
+
+(Appendix A)
 
@@
 these variables are relevant to C accounting under IPCC, and hence to
-EFDB (Fig. 1) This included eleven increment variables, adding to only
+EFDB (Fig. 1). This included eleven increment variables, adding to only
 one previously defined increment variable (aboveground biomass
@@
 \subsection{Mapping ForC to EFDB}
+
+Appendix B
 
@@
 
-\section{Appendix A}
+\section*{Appendix A. Updates to ForC}
+\addcontentsline{toc}{section}{Appendix A. Updates to ForC}
+
+\captionsetup[table]{labelformat=empty}
+
+Table A1: \textbf{Table of changes to ForC fields.}
 
 \begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\caption{\label{tab:table_ForCfieldmapping}\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted). }\\
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+\endfirsthead
+\multicolumn{5}{@{}l}{\textit{(continued)}}\\
+\hline
+\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
+\hline
+\endhead
+Sites & coordinates.precision & Precision of geographic coordinates, as reported by source or estimated from maps. & field added & allow identification of records with poor coordinate precision\\
+\hline
+Measurements & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
+\hline
+ & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
+& replaces `stat` and `stat.name` & cleaner format; ability to handle assymetrical 95 percent confidence intervals\\
+\hline
+ & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
+\hline
+ & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
+& field added & track units conversion, allow back-calculation of OM if conversion factor deemed inappropriate\\
+\hline
+PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
+\hline
+ & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
+\hline
+Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC\\
+\hline
+ & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
+\hline
+ & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
+\hline
+ & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
+\hline
+ & source.type & citation source type & field added & field required by IPCC\\
+\hline
+ & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
+\hline
+ & EFDB.ready & Indicates whether data have been checked for export to EFDB. & field added & housekeeping\\
+\hline
+\end{longtabu}
+
+\clearpage
+
+\section*{Appendix B. Mapping ForC to EFDB}
+\addcontentsline{toc}{section}{Appendix B. Mapping ForC to EFDB}
+
+Table B1: \textbf{Mapping of ForC fields to EFDB.} See footnotes at end
+of table (STILL NEED TO BE PROPERLY INSERTED).
+
+\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
 \hline
@@
 \endfirsthead
-\caption[]{\textbf{Mapping of ForC fields to EFDB.} See footnotes at end of table (still need to be properly inserted).  \textit{(continued)}}\\
+\multicolumn{5}{@{}l}{\textit{(continued)}}\\
 \hline
@@
 
-\clearpage
-
-\section{Appendix B}
-
-\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\caption{\label{tab:table_ForCchanges}\textbf{Table of changes to ForC fields.}}\\
-\hline
-\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
-\hline
-\endfirsthead
-\caption[]{\textbf{Table of changes to ForC fields.} \textit{(continued)}}\\
-\hline
-\textbf{Table} & \textbf{Column} & \textbf{Description} & \textbf{Changes} & \textbf{Motivation}\\
-\hline
-\endhead
-Sites & coordinates.precision & Precision of geographic coordinates, as reported by source or estimated from maps. & field added & allow identification of records with poor coordinate precision\\
-\hline
-Measurements & data.location.within.source & Location of data within the source listed in citation.ID. & field added & facilitate review, ensure traceability\\
-\hline
- & sd, se, lower95\%CI, upper 95\%CI & Standard deviation, standard error, and lower and upper 95 percent confidence intvervals, respectively.
-& replaces `stat` and `stat.name` & cleaner format; ability to handle assymetrical 95 percent confidence intervals\\
-\hline
- & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
-\hline
- & C.conversion.factor & Assumed/ measured C content of organic matter used to convert organic matter to C.
-& field added & track units conversion, allow back-calculation of OM if conversion factor deemed inappropriate\\
-\hline
-PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
-\hline
- & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
-\hline
-Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC\\
-\hline
- & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
-\hline
- & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
-\hline
- & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
-\hline
- & source.type & citation source type & field added & field required by IPCC\\
-\hline
- & pdf.in.repository & Indicates whether pdf of original study has been retrieved and saved in ForC's reference repository & field added & housekeeping\\
-\hline
- & EFDB.ready & Indicates whether data have been checked for export to EFDB. & field added & housekeeping\\
-\hline
-\end{longtabu}
-
 \codedataavailability{use this to add a statement when having data sets
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-\section{Mapping ForC to EFDB}
-
-CURRENT TABLE 3 GOES HERE
-
-\section{Updates to ForC}
-
-CURRENT TABLE 4 GOES HERE
-\noappendix
 

### 2022-04-16 doc/manuscript/ESSDmanuscript.Rmd
describe new or modified fields

@@
 n_new_variables = 11 + 2 + 2 #delta. + woody.mortality + O horizon
+# get n fields modified
+sampling_details <- read.csv(paste0(path_to_ForC,"/ForC/database_management_records/record_of_changes.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+n_fields_modified =  length(sampling_details[,"Column"]) + 4 
 ```
@@
 
-Previous versions of ForC [@refs] contained most of the information required for transfer to EFDB, and, more broadly, to inform C accounting under IPCC guidelines. 
+Previous versions of ForC [@refs] contained most of the information required by EFDB, and, more broadly, to inform C accounting under IPCC guidelines. 
 However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified ## fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
 
@@
 
-(`r table_ForCchanges`)
+We added or modified a total of `r n_fields_modified` fields (`r table_ForCchanges`).
+Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
+For the latter, we *used an R script to automatically retrieve information based on the DOI*<!--needs info from Valentine here-->.
 
@@
 
-# ForC to EFDB data transfer
+# Transfer of data from ForC to EFDB
 

### 2022-04-16 doc/manuscript/ESSDmanuscript.tex
describe new or modified fields

@@
 Previous versions of ForC \citep{refs} contained most of the information
-required for transfer to EFDB, and, more broadly, to inform C accounting
-under IPCC guidelines. However, modest changes to the structure and
-contents of ForC were needed in order to provide all information
-required by EFDB and to improve ForC's capacity to serve as a repository
-of valuable information for forest C accounting under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of
-the ForC database, we added or modified \#\# fields (Appendix A),
-defined 15 new variables, implemented enhanced quality control, manually
-reviewed \textgreater\# records to obtain additional required
-information, and added \# new records.
+required by EFDB, and, more broadly, to inform C accounting under IPCC
+guidelines. However, modest changes to the structure and contents of
+ForC were needed in order to provide all information required by EFDB
+and to improve ForC's capacity to serve as a repository of valuable
+information for forest C accounting under IPCC guidelines. To support
+export of data to EFDB, and to improve the overall quality of the ForC
+database, we added or modified 18 fields (Appendix A), defined 15 new
+variables, implemented enhanced quality control, manually reviewed
+\textgreater\# records to obtain additional required information, and
+added \# new records.
 
@@
 
-(Appendix A)
+We added or modified a total of 18 fields (Appendix A). Most notably,
+these included improvement of the representation of uncertainty,
+recording of original units and organic matter to C conversion factors,
+and expanding the information recorded in the citations table. For the
+latter, we \emph{used an R script to automatically retrieve information
+based on the DOI}.
 
@@
 
-\section{ForC to EFDB data transfer}
+\section{Transfer of data from ForC to EFDB}
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
finish first draft of section 3 (ForC updates)

@@
 n_fields_modified =  length(sampling_details[,"Column"]) + 4 
+n_records_checked = NA # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
+n_records_added = NA
+
 ```
@@
 
-Previous versions of ForC [@refs] contained most of the information required by EFDB, and, more broadly, to inform C accounting under IPCC guidelines. 
+Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C accounting under IPCC guidelines.
 However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
@@
 
-*This section describes changes relative to ForC v2.0 [@anderson-teixeira_forc_2018].*
+This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
 
@@
 Prior to releasing ForC v4.0, we executed several quality control measures.
-First, we implemented a system of continuous integration using GitHub Actions (*sensu* Kim et al. in review) to run some automatic checks any time the master data files are updated.
-Second, to improve information on geographic coordinates, we flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
+First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_using_inreview] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
+Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
 Third, to identify erroneous climate data... *(Issue #212)[https://github.com/forc-db/ForC/issues/212]*.
@@
 
-EFDB data submissions require information that was not recorded in previous versions of ForC.
-It was therefore necessary to return to original publications to retrieve relevant information, including the following:
+EFDB data submissions required information that was not recorded in previous versions of ForC, but rather was recorded in fields (`r table_ForCchanges`).
+It was therefore necessary to return to original publications to retrieve relevant information, including 
+(1) whether data were presented in tables / text or digitized from figures (EFDB will not accept digitized data, whereas ForC does), 
+(2) estimates in original units, 
+and (3) confidence intervals (when not already in ForC).
+We also checked that existing ForC records were complete and correct.
 
-*Data location within source*-  Because EFDB will not accept data which has been digitized (from graphs) while ForC does, it was necessary to review and note the location of the data within the original source
+Manual review of records was the limiting step for data transfer to EFDB. 
+We prioritized review of
+(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021],
+(2) studies with confidence intervals recorded in ForC (because uncertainty estimates are important to the IPCC),
+(3) original publications containing large numbers of EFDB-relevant records,
+and (4) records from tropical regions.
+The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@refs], ground-based data on tropical forest C cycling tend to be more scarce due to a variety of challenges [@delima_making_2022], and *tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on previously existing data [@ref].* <!-- I'm almost certain this is true, but need to confirm/ find ref.-->
 
 ## Addition of new records
+
+In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
+These included **##** records from **##** studies [@piponiot_distribution_2022; @refs<!--others?-->] that were not previously included in ForC.
+In addition, we created new records for **##** EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
finish first draft of section 3 (ForC updates)

@@
 
-Previous versions of ForC \citep{refs} contained most of the information
-required by EFDB, and, more broadly, to inform C accounting under IPCC
-guidelines. However, modest changes to the structure and contents of
-ForC were needed in order to provide all information required by EFDB
-and to improve ForC's capacity to serve as a repository of valuable
-information for forest C accounting under IPCC guidelines. To support
-export of data to EFDB, and to improve the overall quality of the ForC
-database, we added or modified 18 fields (Appendix A), defined 15 new
-variables, implemented enhanced quality control, manually reviewed
-\textgreater\# records to obtain additional required information, and
-added \# new records.
-
-\emph{This section describes changes relative to ForC v2.0
-\citep{anderson-teixeira_forc_2018}.}
+Previous versions of ForC
+\citep{anderson-teixeira_carbon_2016, anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}
+contained most of the information required by EFDB, and, more broadly,
+to inform C accounting under IPCC guidelines. However, modest changes to
+the structure and contents of ForC were needed in order to provide all
+information required by EFDB and to improve ForC's capacity to serve as
+a repository of valuable information for forest C accounting under IPCC
+guidelines. To support export of data to EFDB, and to improve the
+overall quality of the ForC database, we added or modified 18 fields
+(Appendix A), defined 15 new variables, implemented enhanced quality
+control, manually reviewed \textgreater\# records to obtain additional
+required information, and added \# new records.
+
+This section describes changes relative to ForC v3.0
+\citep{anderson-teixeira_carbon_2021}.
 
@@
 measures. First, we implemented a system of continuous integration using
-GitHub Actions (\emph{sensu} Kim et al.~in review) to run some automatic
-checks any time the master data files are updated. Second, to improve
-information on geographic coordinates, we flagged and reviewed records
-with suspected low precision \emph{(Issue
-\#29){[}https://github.com/forc-db/ForC/issues/229{]}}. Third, to
-identify erroneous climate data\ldots{} \emph{(Issue
+GitHub Actions \citep[\emph{sensu}][]{kim_using_inreview} to run some
+automatic checks any time the master data files are updated, including
+outlier tests and checks for completeness and naming consistency of
+records across data files. Second, to improve information on geographic
+coordinates, we created a field to record coordinate precision (Appendix
+A), and flagged and reviewed records with suspected low precision
+\emph{(Issue \#29){[}https://github.com/forc-db/ForC/issues/229{]}}.
+Third, to identify erroneous climate data\ldots{} \emph{(Issue
 \#212){[}https://github.com/forc-db/ForC/issues/212{]}}.
@@
 
-EFDB data submissions require information that was not recorded in
-previous versions of ForC. It was therefore necessary to return to
-original publications to retrieve relevant information, including the
-following:
-
-\emph{Data location within source}- Because EFDB will not accept data
-which has been digitized (from graphs) while ForC does, it was necessary
-to review and note the location of the data within the original source
+EFDB data submissions required information that was not recorded in
+previous versions of ForC, but rather was recorded in fields (Appendix
+A). It was therefore necessary to return to original publications to
+retrieve relevant information, including (1) whether data were presented
+in tables / text or digitized from figures (EFDB will not accept
+digitized data, whereas ForC does), (2) estimates in original units, and
+(3) confidence intervals (when not already in ForC). We also checked
+that existing ForC records were complete and correct.
+
+Manual review of records was the limiting step for data transfer to
+EFDB. We prioritized review of (1) records from the Forest Global Earth
+Observatory
+\citep[ForestGEO,][]{anderson-teixeira_ctfsforestgeo_2015, davies_forestgeo_2021},
+(2) studies with confidence intervals recorded in ForC (because
+uncertainty estimates are important to the IPCC), (3) original
+publications containing large numbers of EFDB-relevant records, and (4)
+records from tropical regions. The latter criteria was motivated by the
+fact that although tropical forest is the single most important biome
+for climate change mitigation \citep{refs}, ground-based data on
+tropical forest C cycling tend to be more scarce due to a variety of
+challenges \citep{delima_making_2022}, and \emph{tropical countries are
+more likely to apply Tier 1 methodology that bases forest C budgets on
+previously existing data \citep{ref}.}
 
 \subsection{Addition of new records}
+
+In addition to reviewing existing records, we added a total of NA new
+records to ForC. These included \textbf{\#\#} records from \textbf{\#\#}
+studies \citep{piponiot_distribution_2022, refs} that were not
+previously included in ForC. In addition, we created new records for
+\textbf{\#\#} EFDB-relevant estimates presented in the original
+publication that were not yet present in ForC.
 
@@
 
-As of April 16, 2022, ForC (v4.0) contained NA independent records (NA
+As of April 18, 2022, ForC (v4.0) contained NA independent records (NA
 total), NA of which were for variables relevant to EFDB (Fig. 1). These
@@
 
-As of April 16, 2022, we had reviewed and sent NA, NA of which have been
+As of April 18, 2022, we had reviewed and sent NA, NA of which have been
 reviewed, accepted, and posted (Fig. 2-3, Table 2). \emph{{[}DETAILS{]}}

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
added code to pull in some statistics

@@
 #Statistics
-n_records_ForC= NA # total n records in ForC
-n_plots_ForC = NA # total n plots in ForC
-n_areas_ForC = NA # total n geographic areas in ForC
-n_independent_records_ForC= NA #suspected duplicates removed
+ForC_measurements <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_measurements.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+n_records_ForC= length(ForC_measurements[,"measurement.ID"])  # total n records in ForC
+
+ForC_plots <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_plots.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+n_plots_ForC = length(ForC_plots[,"plot.ID"]) # total n plots in ForC
+
+ForC_sites <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_sites.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+n_areas_ForC = max(ForC_sites$geographic.area, na.rm = TRUE) # total n geographic areas in ForC
+
+n_independent_records_ForC= NA #suspected duplicates removed. This is the number of records fin ForC simplified.
 n_records_EFDB_variables= NA #n independent records relevant to EFDB
@@
 n_records_checked = NA # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
-n_records_added = NA
+n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
added code to pull in some statistics

@@
 \citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. For
-C currently contains NA records from NA plots in NA distinct
+C currently contains 39838 records from 10585 plots in 1531 distinct
 geographical areas, along with records of stand age and disturbance
@@
 
-In addition to reviewing existing records, we added a total of NA new
+In addition to reviewing existing records, we added a total of 305 new
 records to ForC. These included \textbf{\#\#} records from \textbf{\#\#}
@@
 
-As of April 18, 2022, ForC (v4.0) contained NA independent records (NA
-total), NA of which were for variables relevant to EFDB (Fig. 1). These
-records were distributed across all forested continents and ecozones,
-with particularly high concentrations in \emph{{[}ecozone(s){]}} and low
-concentrations in \emph{{[}ecozone(s){]}} (Fig. 2). The most widely
-represented forest type was \emph{{[}type{]}}, followed by
-\emph{{[}type{]}} and \emph{{[}type{]}} (Fig. 3).
+As of April 18, 2022, ForC (v4.0) contained NA independent records
+(39838 total), NA of which were for variables relevant to EFDB (Fig. 1).
+These records were distributed across all forested continents and
+ecozones, with particularly high concentrations in
+\emph{{[}ecozone(s){]}} and low concentrations in
+\emph{{[}ecozone(s){]}} (Fig. 2). The most widely represented forest
+type was \emph{{[}type{]}}, followed by \emph{{[}type{]}} and
+\emph{{[}type{]}} (Fig. 3).
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
writing section 4

@@
 section_land_use_categories = 2.2
+section_C_variable_mapping = "4.1.1"
+section_land_use_mapping = "4.1.2"
 
@@
 n_records_sent_to_EFDB = NA # n records sent to EFDB
-n_records_posted_EFDB = NA # n records accepted and posted in EFDB
+n_records_posted_EFDB = 73 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
+Then get number from upper left box where it says "displayed records")
 
@@
 However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added # new records.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added `r n_records_added` new records.
 
@@
 It was therefore necessary to return to original publications to retrieve relevant information, including 
-(1) whether data were presented in tables / text or digitized from figures (EFDB will not accept digitized data, whereas ForC does), 
-(2) estimates in original units, 
-and (3) confidence intervals (when not already in ForC).
+(1) whether records of interest were presented in tables / text or digitized from figures (EFDB will not accept digitized data, whereas ForC does), 
+(2) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, ForC would accept the sum of these as a valid estimate of total biomass, whereas EFDB would not)
+(3) estimates in original units, 
+and (4) confidence intervals (when not already in ForC).
 We also checked that existing ForC records were complete and correct.
@@
 
-To transfer data from ForC into the EFDB, we populated EFDB's bulk import form ("EFDB bulk import.xlsx"), using R scripts to restructure complete, reviewed records for entry into this form.
+To transfer complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
+Criteria for data transfer were that (1) records had been checked against the original study and determined to be complete and correct (`EFDB.ready` field in `ForC_citations` table), (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables (`data.location.within.source` field in `ForC_measurements` table), and (3) the records had not previously been sent to EFDB.
+Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for inclusion in EFDB. 
+Complete records needed to be reviewed by the EFDB editorial board prior to posting in the database -- a process that lags behind records transfer and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
 
@@
 
-`r table_ForCfieldmapping`
+The mapping of ForC fields into EFDB fields is documented in `r table_ForCfieldmapping`.
+
+For the majority of fields, contents of the field in ForC was transferred directly into an EFDB field, either as the only contents of that field or as part of a composite record.
+For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field `Region/Regionalconditions` (`r table_ForCfieldmapping`).
+In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field `Comments from Data Provider`).
+For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
+For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (`Gases` field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
+There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) determination of land use categories (section `r section_land_use_mapping`).
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
writing section 4

@@
 control, manually reviewed \textgreater\# records to obtain additional
-required information, and added \# new records.
+required information, and added 305 new records.
 
@@
 A). It was therefore necessary to return to original publications to
-retrieve relevant information, including (1) whether data were presented
-in tables / text or digitized from figures (EFDB will not accept
-digitized data, whereas ForC does), (2) estimates in original units, and
-(3) confidence intervals (when not already in ForC). We also checked
-that existing ForC records were complete and correct.
+retrieve relevant information, including (1) whether records of interest
+were presented in tables / text or digitized from figures (EFDB will not
+accept digitized data, whereas ForC does), (2) whether records of
+interest were presented directly, as opposed to having been calculated
+from related variables (for example, if a study presents aboveground
+biomass and root biomass but not total biomass, ForC would accept the
+sum of these as a valid estimate of total biomass, whereas EFDB would
+not) (3) estimates in original units, and (4) confidence intervals (when
+not already in ForC). We also checked that existing ForC records were
+complete and correct.
 
@@
 
-To transfer data from ForC into the EFDB, we populated EFDB's bulk
-import form (``EFDB bulk import.xlsx''), using R scripts to restructure
-complete, reviewed records for entry into this form.
+To transfer complete, reviewed ForC records into EFDB, we created R
+scripts to restructure ForC records and populate EFDB's bulk import form
+(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
+records had been checked against the original study and determined to be
+complete and correct (\texttt{EFDB.ready} field in
+\texttt{ForC\_citations} table), (2) the original study presented values
+in tables or text, as opposed to the values having been digitized from
+graphs or calculated based on related variables
+(\texttt{data.location.within.source} field in
+\texttt{ForC\_measurements} table), and (3) the records had not
+previously been sent to EFDB. Once converted into EFDB format, the
+records were reviewed and then sent to the IPCC's Technical Support Unit
+for inclusion in EFDB. Complete records needed to be reviewed by the
+EFDB editorial board prior to posting in the database -- a process that
+lags behind records transfer and had not yet been completed for all
+records sent as of April 18, 2022.
 
@@
 
-Appendix B
+The mapping of ForC fields into EFDB fields is documented in Appendix B.
+
+For the majority of fields, contents of the field in ForC was
+transferred directly into an EFDB field, either as the only contents of
+that field or as part of a composite record. For example, ten ForC
+fields describing site location, climate, and edaphic properties all
+mapped into the EFDB field \texttt{Region/Regionalconditions} (Appendix
+B). In cases where original studies did not present 95\% confidence
+intervals (required by IPCC when available) but did present information
+required to calculate these (standard error or n and standard
+deviation), we calculated the 95\% confidence intervals and populated
+the EFDB field with this information (noting the calculation in the EFDB
+field \texttt{Comments\ from\ Data\ Provider}). For some fields, simple
+conditional logic was used to populate EFDB fields based on ForC
+records. For example, for stock variables presented in the original
+publication in units of dry organic matter mass (as opposed to C),
+several greenhouse gasses (CO\textsubscript{2}, CO, CH\textsubscript{4},
+NO, NO\textsubscript{2}, N\textsubscript{2}O) were entered in the EFDB
+field indicating the greenhouse gases to which the record could be
+pertinent (\texttt{Gases} field) because these values could be used in
+calculations of greenhouse gas emissions from biomass burning
+\citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse gas
+would be CO\textsubscript{2}. There were two cases in which more complex
+mapping was required: (1) mapping of C cycle variables (section 4.1.1)
+and (2) determination of land use categories (section 4.1.2).
 
@@
 
-As of April 18, 2022, we had reviewed and sent NA, NA of which have been
+As of April 18, 2022, we had reviewed and sent NA, 73 of which have been
 reviewed, accepted, and posted (Fig. 2-3, Table 2). \emph{{[}DETAILS{]}}

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
describe mapping of C cycle variables

@@
 section_land_use_categories = 2.2
+section_new_variables = 3.2
 section_C_variable_mapping = "4.1.1"
@@
 
-The mapping of ForC fields into EFDB fields is documented in `r table_ForCfieldmapping`.
+The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`, with details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
 
 For the majority of fields, contents of the field in ForC was transferred directly into an EFDB field, either as the only contents of that field or as part of a composite record.
-For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field `Region/Regionalconditions` (`r table_ForCfieldmapping`).
+For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field `Region/Regional conditions` (`r table_ForCfieldmapping`).
 In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field `Comments from Data Provider`).
@@
 
-Mapping of variables is shown in Fig. `r fig_variable_mapping`.
+With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into the C pools and stock, stock changes, increments, and losses (Fig.
+`r fig_variable_mapping`).
+As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented. 
+It it important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
+For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
+Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
+Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the IPCC-EFDB-integration repository ([https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
describe mapping of C cycle variables

@@
 
-The mapping of ForC fields into EFDB fields is documented in Appendix B.
+The mapping of ForC fields into EFDB fields is summarized in Appendix B,
+with details documented in the public GitHub repository associated with
+the project, IPCC-EFDB-integration repository within the ForC-db
+organization (file \emph{ForC-EFDB\_mapping.csv} available at
+\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).
 
@@
 fields describing site location, climate, and edaphic properties all
-mapped into the EFDB field \texttt{Region/Regionalconditions} (Appendix
-B). In cases where original studies did not present 95\% confidence
-intervals (required by IPCC when available) but did present information
-required to calculate these (standard error or n and standard
-deviation), we calculated the 95\% confidence intervals and populated
-the EFDB field with this information (noting the calculation in the EFDB
-field \texttt{Comments\ from\ Data\ Provider}). For some fields, simple
-conditional logic was used to populate EFDB fields based on ForC
-records. For example, for stock variables presented in the original
-publication in units of dry organic matter mass (as opposed to C),
-several greenhouse gasses (CO\textsubscript{2}, CO, CH\textsubscript{4},
-NO, NO\textsubscript{2}, N\textsubscript{2}O) were entered in the EFDB
-field indicating the greenhouse gases to which the record could be
-pertinent (\texttt{Gases} field) because these values could be used in
-calculations of greenhouse gas emissions from biomass burning
-\citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse gas
-would be CO\textsubscript{2}. There were two cases in which more complex
-mapping was required: (1) mapping of C cycle variables (section 4.1.1)
-and (2) determination of land use categories (section 4.1.2).
+mapped into the EFDB field \texttt{Region/Regional\ conditions}
+(Appendix B). In cases where original studies did not present 95\%
+confidence intervals (required by IPCC when available) but did present
+information required to calculate these (standard error or n and
+standard deviation), we calculated the 95\% confidence intervals and
+populated the EFDB field with this information (noting the calculation
+in the EFDB field \texttt{Comments\ from\ Data\ Provider}). For some
+fields, simple conditional logic was used to populate EFDB fields based
+on ForC records. For example, for stock variables presented in the
+original publication in units of dry organic matter mass (as opposed to
+C), several greenhouse gasses (CO\textsubscript{2}, CO,
+CH\textsubscript{4}, NO, NO\textsubscript{2}, N\textsubscript{2}O) were
+entered in the EFDB field indicating the greenhouse gases to which the
+record could be pertinent (\texttt{Gases} field) because these values
+could be used in calculations of greenhouse gas emissions from biomass
+burning \citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse
+gas would be CO\textsubscript{2}. There were two cases in which more
+complex mapping was required: (1) mapping of C cycle variables (section
+4.1.1) and (2) determination of land use categories (section 4.1.2).
 
@@
 
-Mapping of variables is shown in Fig. 1.
+With input from the IPCC's Technical Support Unit, we reviewed the list
+of ForC variables to identify those that were relevant to EFDB and to
+appropriately map them into the C pools and stock, stock changes,
+increments, and losses (Fig. 1). As described in section 3.2, we also
+defined 15 new EFDB-relevant variables that were not previously
+represented. It it important to note that the correspondence of ForC
+variables to IPCC criteria often depends upon measurement protocols
+(``important sources of estimate variation'' in Table 1). For example,
+ForC records of biomass and dead wood vary in the minimum stem diameter
+censused, such that some records would match the IPCC criteria whereas
+others would not. Information on minimum diameters censused and other
+important sources of methodological variation are recorded as covariates
+in ForC and mapped into the EFDB field \emph{Other Properties} (Appendix
+B). Details on the mapping of ForC variables to EFDB -- including
+associated covariates, IPCC pools (Table 1) and relevant equations
+\citep{ipcc_2006_2006} -- are documented in the file
+ForC\_variables\_mapping.csv in the IPCC-EFDB-integration repository
+(\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
various changes

@@
 The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest carbon stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
+ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
 For C currently contains `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
@@
 For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
-For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig.
-`r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006].
+For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006].
 Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools.
@@
   row_spec(0, bold = TRUE)
-```
-
-```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, and that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/C_variable_mapping.png")
 ```
@@
 
-We added a total of `r n_new_variables` new variables to the set of named and defined variables, counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
-All of these variables are relevant to C accounting under IPCC, and hence to EFDB (Fig. `r fig_variable_mapping`).
+We added a total of `r n_new_variables` new EFDB-relevant variables to the set of named and defined variables (Fig.
+`r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
 This included eleven increment variables, adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
@@
 Although most of these variables currently lack records in ForC, the structure exists such that records can be populated over time. 
+Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).
 
-Further, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including *litter* (OL)).
+```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
+Most, but not all, EFDB-relevant ForC variables are shown here.
+Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon."}
+#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
+knitr::include_graphics("figures_tables/C_variable_mapping.png")
+```
 
@@
 
-With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into the C pools and stock, stock changes, increments, and losses (Fig.
+With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig.
 `r fig_variable_mapping`).
-As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented. 
+For each C pools (Table `r table_pools`), we identified variables representing organic matter or C stocks, stock changes (a.k.a.
+"net annual increments" by IPCC, "increments" in ForC), gains (a.k.a.
+"gross annual increments" by IPCC, "fluxes" in ForC), and losses ("fluxes" in ForC).
+As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
 It it important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
@@
 Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).
-
-## Review and posting of data
 
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB`, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Fig.
+As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB` records to EFDB, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Fig.
 `r fig_map`-`r fig_dominant_veg`, Table `r table_variables`).

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
various changes

@@
 fluxes
-\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. For
-C currently contains 39838 records from 10585 plots in 1531 distinct
+\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
+includes data ingested from individual publications and relevant
+databases, including the Global Reforestation Opportunity Assessment
+(GROA) database \citep[database doi:
+10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
+respiration database
+\citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
+For C currently contains 39838 records from 10585 plots in 1531 distinct
 geographical areas, along with records of stand age and disturbance
@@
 \(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
-takes the difference between gains and losses (influx and outflux
-variables in Fig. 1), or using the ``Stock-Difference Method'', which
-computes \(\Delta C\) based on C stocks at two points in time
-\citep{ipcc_2006_2006}. Thus, C cycle variables relevant to the IPCC
-methodology and to EFDB include C stocks, increments, and fluxes in the
-IPCC-defined pools.
+takes the difference between gains and losses, or using the
+``Stock-Difference Method'', which computes \(\Delta C\) based on C
+stocks at two points in time \citep{ipcc_2006_2006}. Thus, C cycle
+variables relevant to the IPCC methodology and to EFDB include C stocks,
+increments, and fluxes in the IPCC-defined pools.
 
@@
 \end{table}
-
-\begin{figure}
-\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; ForC variables corresponding to the stock, increment, influx and outflux; and relationships among them.} Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, and that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.}\label{fig:fig_variable_mapping}
-\end{figure}
 
@@
 
-We added a total of 15 new variables to the set of named and defined
-variables, counting each pair of variables with units in C (ending in
-\texttt{\_C}) or organic matter (ending in \texttt{\_OM}) as one. All of
-these variables are relevant to C accounting under IPCC, and hence to
-EFDB (Fig. 1). This included eleven increment variables, adding to only
-one previously defined increment variable (aboveground biomass
+We added a total of 15 new EFDB-relevant variables to the set of named
+and defined variables (Fig. 1), counting each pair of variables with
+units in C (ending in \texttt{\_C}) or organic matter (ending in
+\texttt{\_OM}) as one. This included eleven increment variables, adding
+to only one previously defined increment variable (aboveground biomass
 increment, \emph{delta.agb}). These are directly related to C stocks as
@@
 lack records in ForC, the structure exists such that records can be
-populated over time.
-
-Further, to provide better definition of the previously existing
-variable \emph{organic.layer}, which has a nebulous definition that
-reflects the varied definitions adopted by original studies, we added
-two clearly defined variables: \emph{litter} (relatively undecomposed
-plant material/ OL horizon), and \emph{O.horizon} (entire O-horizon,
-including \emph{litter} (OL)).
+populated over time. Finally, to provide better definition of the
+previously existing variable \emph{organic.layer}, which has a nebulous
+definition that reflects the varied definitions adopted by original
+studies, we added two clearly defined variables: \emph{litter}
+(relatively undecomposed plant material/ OL horizon), and
+\emph{O.horizon} (entire O-horizon, including OL).
+
+\begin{figure}
+\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
+Most, but not all, EFDB-relevant ForC variables are shown here.
+Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.}\label{fig:fig_variable_mapping}
+\end{figure}
 
@@
 of ForC variables to identify those that were relevant to EFDB and to
-appropriately map them into the C pools and stock, stock changes,
-increments, and losses (Fig. 1). As described in section 3.2, we also
-defined 15 new EFDB-relevant variables that were not previously
-represented. It it important to note that the correspondence of ForC
-variables to IPCC criteria often depends upon measurement protocols
+appropriately map them into EFDB (Fig. 1). For each C pools (Table 1),
+we identified variables representing organic matter or C stocks, stock
+changes (a.k.a. ``net annual increments'' by IPCC, ``increments'' in
+ForC), gains (a.k.a. ``gross annual increments'' by IPCC, ``fluxes'' in
+ForC), and losses (``fluxes'' in ForC). As described in section 3.2, we
+also defined 15 new EFDB-relevant variables that were not previously
+represented in ForC. It it important to note that the correspondence of
+ForC variables to IPCC criteria often depends upon measurement protocols
 (``important sources of estimate variation'' in Table 1). For example,
@@
 
-\subsection{Review and posting of data}
-
 \section{Results}
@@
 
-As of April 18, 2022, we had reviewed and sent NA, 73 of which have been
-reviewed, accepted, and posted (Fig. 2-3, Table 2). \emph{{[}DETAILS{]}}
+As of April 18, 2022, we had reviewed and sent NA records to EFDB, 73 of
+which have been reviewed, accepted, and posted (Fig. 2-3, Table 2).
+\emph{{[}DETAILS{]}}
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
work on recommendations section

@@
 fig_map=2
-fig_dominant_veg=3
+fig_histograms=3
 
@@
 section_land_use_mapping = "4.1.2"
+section_database_needs = "6.1"
+section_new_data_needs = "6.2"
+section_data_reporting_needs = "6.3" 
+section_mismatches = 6.4
 
@@
 The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include carbon stocks, increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
+In the case of forests, emission factors include carbon stocks, net increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
 
@@
 For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006].
-Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools.
+Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
 
@@
 These records were distributed across all forested continents and ecozones, with particularly high concentrations in *[ecozone(s)]* and low concentrations in *[ecozone(s)]* (Fig. `r fig_map`).
-The most widely represented forest type was *[type]*, followed by *[type]* and *[type]* (Fig. `r fig_dominant_veg`).
+The most widely represented forest type was *[type]*, followed by *[type]* and *[type]* (Fig. `r fig_histograms`).
 
@@
 
-```{r fig_dominant_veg, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
 For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB` records to EFDB, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Fig.
-`r fig_map`-`r fig_dominant_veg`, Table `r table_variables`).
+As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB` records to EFDB, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
+`r fig_map`-`r fig_histograms`, Table `r table_variables`).
 *[DETAILS]*
@@
 
-(*strongly flag both useful variables that the EFDB does not track and useful variables that papers fail to include that EFDB needs*)
+Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work (section `r section_database_needs`), new data collection and analysis (section `r section_new_data_needs`), and reporting (section `r section_data_reporting_needs`).
+We also highlight notable mismatches between IPCC accounting methods and forest C mensuration (section `r section_mismatches`).
+
+## Database needs
+
+There is vast potential to expand forest C data in EFDB by completing the process of reviewing and sending data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
+So far, only ~`r round(n_records_sent_to_EFDB/n_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.
+
+Moreover, there are many published EFDB-relevant forest C data that are not included in ForC, with more being published on a nearly daily basis. 
+Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
+*(There are some efforts underway, including a few that Susan can specify.)*
+Such reviews are necessary to even develop a rigorous assessment of forest C data that are available, versus those that require additional data collection and analysis.
 
@@
 
-*(Paragraph highlighting important gaps in variables / regions)*
+New data collection and analysis is needed to fill notable knowledge gaps.
+While aboveground biomass stocks in particular have received -- and continue to receive -- significant research attention, other pools and variables remain poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021].
+Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
+For instance, C cycling of tropical forests -- particularly in Africa -- remains relatively poorly characterized, in large part due to substantial barriers to data collection and distribution [@delima_making_2022]
+*(add some more here?)*
 
-Several variables of value to IPCC, including standing dead wood, woody mortality, delta.agb, are not calculated and presented as frequently as are AGB and ANPP_woody, even though they can readily be derived from the same census data.
-We recommend that researchers calculate and report these, following the reporting guidelines specified below.
-Furthermore, there is an opportunity to fill data gaps by calculating these from existing census data.
-For example, the core census protocol of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfsforestgeo_2015,@davies_forestgeo_2021] collects the data required to calculate standing dead wood, woody mortality, and delta.agb, but these have not been calculated and reported for all sites for which the appropriate number of censuses are available <!--"(n=1 for standing dead wood, n=2 for woody mortality and delta.agb)" -- previous text commented out because I think it's out of date-->[but see @piponiot_distribution_2022, @refs<!--insert other refs including these variables-->].
+Several variables of value for IPCC C accounting have not been calculated and presented as would be possible given the same forest census data and minimal extra research effort.
+For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022].
+Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be easily be estimated in parallel with aboveground biomass, with the greatest barrier being availability of reliable allometries, as have been developed for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
+However, while equations for estimating root (and thereby total) biomass require improvement, they do exist *for many forest types*[@refs], and IPCC provides default recommendations of below-ground to above-ground ratios for estimation of root biomass [@ipcc_2019_2019].
+In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate accounting. 
+We recommend that, when possible, researchers calculate these, following the reporting guidelines specified in section `r section_data_reporting_needs`.
 
+*Other EFDB-relevant variables require more effort but are warrented given their importance for forest C accounting.*
 Given widespread trends of increasing tree mortality[@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.
+...
 
@@
 
-## Database needs
+## Mismatches between IPCC accounting methods and forest C mensuration
 
-There are plenty of relevant, published data that are not included in ForC. 
-Systematic review of the literature could vastly improve data coverage.
-*(There are some efforts underway, including a few that Susan can specify.)*
+Remote sensing biomass estimates include standing dead wood [@duncanson_aboveground_2021].
 
-## IPCC protocol considerations
+IPCC accounting methods cannot leverage eddy-covariance measurements, which are widely seen as the best available method for quantifying ecosystem-atmosphere gas exchange.
 
 An important challenge is that forests are changing rapidly, and data collected a decade ago may no longer be relevant, particularly in the cases of C increments and fluxes.
-
-Remote sensing biomass estimates include standing dead wood [@duncanson_aboveground_2021].
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
work on recommendations section

@@
 sources for inclusion in the database. In the case of forests, emission
-factors include carbon stocks, increments (``stock changes''), and
+factors include carbon stocks, net increments (``stock changes''), and
 fluxes (``gains'' and ``losses'') for various pools
@@
 variables relevant to the IPCC methodology and to EFDB include C stocks,
-increments, and fluxes in the IPCC-defined pools.
+net annual increments, and fluxes in the IPCC-defined pools.
 
@@
 \includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_dominant_veg}
+For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_histograms}
 \end{figure}
@@
 As of April 18, 2022, we had reviewed and sent NA records to EFDB, 73 of
-which have been reviewed, accepted, and posted (Fig. 2-3, Table 2).
+which have been reviewed, accepted, and posted (Figs. 2-3, Table 2).
 \emph{{[}DETAILS{]}}
@@
 
-(\emph{strongly flag both useful variables that the EFDB does not track
-and useful variables that papers fail to include that EFDB needs})
+Based on our experience contributing forest C data to EFDB via ForC, we
+make several recommendations as to how scientists can improve forest C
+records in EFDB through database work (section 6.1), new data collection
+and analysis (section 6.2), and reporting (section 6.3). We also
+highlight notable mismatches between IPCC accounting methods and forest
+C mensuration (section 6.4).
+
+\subsection{Database needs}
+
+There is vast potential to expand forest C data in EFDB by completing
+the process of reviewing and sending data that are already in ForC
+(Figs. 2-3). So far, only \textasciitilde NA\% of the EFDB-relevant data
+in ForC have been sent to EFDB.
+
+Moreover, there are many published EFDB-relevant forest C data that are
+not included in ForC, with more being published on a nearly daily basis.
+Coverage of particular variables or regions could be vastly improved
+through systematic review of the literature. \emph{(There are some
+efforts underway, including a few that Susan can specify.)} Such reviews
+are necessary to even develop a rigorous assessment of forest C data
+that are available, versus those that require additional data collection
+and analysis.
 
@@
 
-\emph{(Paragraph highlighting important gaps in variables / regions)}
-
-Several variables of value to IPCC, including standing dead wood, woody
-mortality, delta.agb, are not calculated and presented as frequently as
-are AGB and ANPP\_woody, even though they can readily be derived from
-the same census data. We recommend that researchers calculate and report
-these, following the reporting guidelines specified below. Furthermore,
-there is an opportunity to fill data gaps by calculating these from
-existing census data. For example, the core census protocol of the
-Forest Global Earth Observatory {[}ForestGEO;
-\citet{anderson-teixeira_ctfsforestgeo_2015},\citet{davies_forestgeo_2021}{]}
-collects the data required to calculate standing dead wood, woody
-mortality, and delta.agb, but these have not been calculated and
-reported for all sites for which the appropriate number of censuses are
-available \citep[but see][\citet{refs}]{piponiot_distribution_2022}.
-
-Given widespread trends of increasing tree mortality\citep{refs},
-including through severe natural disturbance \citep{refs}, better
-characterization of dead wood will be critical.
+New data collection and analysis is needed to fill notable knowledge
+gaps. While aboveground biomass stocks in particular have received --
+and continue to receive -- significant research attention, other pools
+and variables remain poorly quantified \citep[Table
+2,][]{anderson-teixeira_carbon_2021}. Furthermore, data distribution is
+uneven across forest types and geographical regions (Figs. 2-3). For
+instance, C cycling of tropical forests -- particularly in Africa --
+remains relatively poorly characterized, in large part due to
+substantial barriers to data collection and distribution
+\citep{delima_making_2022} \emph{(add some more here?)}
+
+Several variables of value for IPCC C accounting have not been
+calculated and presented as would be possible given the same forest
+census data and minimal extra research effort. For example, aboveground
+woody mortality (\emph{woody.mortality\_ag}) and aboveground biomass
+increment (\emph{delta.agb}) can be calculated from the same census data
+as aboveground woody productivity (\emph{ANPP\_woody}), yet the latter
+has received far more research attention, and correspondingly has far
+more records in ForC \citetext{\citealp[Table
+2,][]{anderson-teixeira_carbon_2021}; \citealp[but
+see][]{piponiot_distribution_2022}}. Similarly, live coarse root
+biomass, total biomass, and changes in both of these pools could in
+theory be easily be estimated in parallel with aboveground biomass, with
+the greatest barrier being availability of reliable allometries, as have
+been developed for aboveground biomass
+\citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022}.
+However, while equations for estimating root (and thereby total) biomass
+require improvement, they do exist \emph{for many forest
+types}\citep{refs}, and IPCC provides default recommendations of
+below-ground to above-ground ratios for estimation of root biomass
+\citep{ipcc_2019_2019}. In addition, standing dead trees are captured in
+most forest censuses and could be used to estimate standing dead wood,
+although additional data on breakage would be needed for accurate
+accounting. We recommend that, when possible, researchers calculate
+these, following the reporting guidelines specified in section 6.3.
+
+\emph{Other EFDB-relevant variables require more effort but are
+warrented given their importance for forest C accounting.} Given
+widespread trends of increasing tree mortality\citep{refs}, including
+through severe natural disturbance \citep{refs}, better characterization
+of dead wood will be critical. \ldots{}
 
@@
 
-\subsection{Database needs}
-
-There are plenty of relevant, published data that are not included in
-ForC. Systematic review of the literature could vastly improve data
-coverage. \emph{(There are some efforts underway, including a few that
-Susan can specify.)}
-
-\subsection{IPCC protocol considerations}
+\subsection{Mismatches between IPCC accounting methods and forest C
+mensuration}
+
+Remote sensing biomass estimates include standing dead wood
+\citep{duncanson_aboveground_2021}.
+
+IPCC accounting methods cannot leverage eddy-covariance measurements,
+which are widely seen as the best available method for quantifying
+ecosystem-atmosphere gas exchange.
 
@@
 cases of C increments and fluxes.
-
-Remote sensing biomass estimates include standing dead wood
-\citep{duncanson_aboveground_2021}.
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.Rmd
start recommendations table

@@
 table_variables =2
+table_recommendations = 3
 
@@
 
-We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards:
-
-  - adopt common standards for variables like min diameter of deadwood, select soil sampling increments to include a cutoff at 30.
-
-  - report 95% CIs, SE, or STD and n
-
-  - report C variables in article text, table, or SI table. EFDB cannot accept data digitized from figures
-
-  - present calculations of all variables that would be useful to IPCC.
-EFDB requires that data in the database be presented in the original article, and as such cannot accept subsequent calculations.
-For example, if aboveground biomass and total biomass are presented, but root biomass is not presented, root biomass cannot be subsequently calculated and sent to EFDB.
-Similarly, fine and coarse root biomass can't be summed; soil carbon can't be summed across depth increments, etc.
+We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards (`r table_recommendations`).
 
@@
 The latter is more efficient for getting data to EFDB, but does not get the data into ForC, where it can be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].
+
+```{r table_recommendations, eval = TRUE, echo=FALSE, warning=FALSE}
+recommendations_table <- read.csv("figures_tables/recommendations.csv")
+names(recommendations_table) <- stringr::str_replace_all(names(recommendations_table), "\\.", " ")
+
+library(knitr)
+library(kableExtra)
+knitr::kable(recommendations_table, format = "latex",
+             longtable = FALSE,
+             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....") %>%
+  kableExtra::kable_styling(full_width = TRUE) %>%
+  row_spec(0, bold = TRUE)
+```
 

### 2022-04-18 doc/manuscript/ESSDmanuscript.tex
start recommendations table

@@
 otherwise, researchers calculate and report the values according to IPCC
-standards:
-
-\begin{itemize}
-\item
-  adopt common standards for variables like min diameter of deadwood,
-  select soil sampling increments to include a cutoff at 30.
-\item
-  report 95\% CIs, SE, or STD and n
-\item
-  report C variables in article text, table, or SI table. EFDB cannot
-  accept data digitized from figures
-\item
-  present calculations of all variables that would be useful to IPCC.
-  EFDB requires that data in the database be presented in the original
-  article, and as such cannot accept subsequent calculations. For
-  example, if aboveground biomass and total biomass are presented, but
-  root biomass is not presented, root biomass cannot be subsequently
-  calculated and sent to EFDB. Similarly, fine and coarse root biomass
-  can't be summed; soil carbon can't be summed across depth increments,
-  etc.
-\end{itemize}
+standards (3).
 
@@
 or model benchmarking \citep{fer_ecosystem_2021}.
+
+\begin{table}
+
+\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
+\centering
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\hline
+\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
+\hline
+variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data.
+& Estimates of relevant variables are not always calcualted.\\
+\hline
+forest census methods & Census stems down to the smallest diameter feasible. & IPCC biomass pool definition includes all living vegetation\\
+\hline
+ & taxa to include: & IPCC biomass pool definition includes all living vegetation\\
+\hline
+dead organic matter sampling & include damage estimates on standing dead trees & \\
+\hline
+ & adopt IPCC recommendations for min diameter of deadwood & \\
+\hline
+belowground sampling & Select soil sampling increments to include a cutoff at 30 cm depth. & \\
+\hline
+reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables.
+& EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
+\hline
+reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
+\hline
+reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
+\hline
+\end{tabu}
+\end{table}
 

### 2022-04-19 doc/manuscript/ESSDmanuscript.Rmd
describing land classification

@@
 
-## Land use categories
+## Land classification
 IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
 Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
-Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ* [@ipcc_good_2003].
+Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
 Definitions of forest are allowed to vary by country, but must be applied consistently.
@@
 
-Inventories are conducted only for managed land, which is defined by IPCC to include "all forests subject to some kind of human interactions (notably commercial management, harvest of industrial round-wood (logs) and fuelwood, production and use of wood commodities, and forest managed for amenity value or environmental protection if specified by the country), with defined geographical boundaries" [@ipcc_good_2003].
-As such, managed land constitutes all or the majority of Forest Land in most countries; however, many countries have not yet reported their approach for defining managed land [@ogle_delineating_2018; @deng_comparing_2021].
+The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
+This expansive definition of managed land implies that the majority of Forest Land in most countries is managed [e.g.,].
+However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
 
@@
 For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (`Gases` field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
-There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) determination of land use categories (section `r section_land_use_mapping`).
+There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) land classification (section `r section_land_use_mapping`).
 
@@
 
-### Land use categories
+### Land classification
+
+Determination of the IPCC land-use category (i.e., Forest Land, Grassland, Wetlands, Cropland, Settlements, or Other Land; section `r  section_land_use_categories`) was made based on the categorical ForC field *dominant.life.form*, sometimes drawing upon stand age.
+Records with "woody" *dominant.life.form* were classified as Forest Land.
+Those with *dominant.life.form* of "woody+grass", which in ForC is indicative of anything from a shrub-encroached grassland to a tree-dominated savanna, were given dual classification of Forest Land and Grassland.
+This dual classification indicates that records may be relevant to either category depending on the definition of forest applied (varies by country).
+For (rare) cases where *dominant.life.form* was grass and stand age was greater than zero, indicative of early successional vegetation, we assigned a classification of Forest Land, consistent with the IPCC definition that Forest Land includes land expected to succeed to forest.
+Cases where *dominant.life.form* was grass or crop and stand age was zero were indicative of a control for studies of forest regrowth following agricultural abandonment, and were classified as Grassland and Cropland, respectively.
+
+*(now describe sub-category classification, which depends on past land use)*
 
@@
 
-The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [*IPCC 2006 full report REF*].
-This definition is applied differently across countries, and is not clearly defined by the majority of governments [@ogle_delineating_2018]. 
-Given this, and because the IPCC definition of management does not necessarily match that which would be reported in scientific publications and hence in ForC, we do not transfer any classification of land management status from ForC to the EFDB, but do provide auxiliary info that may be useful in making this determination (e.g., geographical location).
+Given the lack of public information needed to determine whether lands are classified as mangaged [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not transfer any classification of land management status from ForC to the EFDB.
+However, we do provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
 

### 2022-04-19 doc/manuscript/ESSDmanuscript.tex
describing land classification

@@
 
-\subsection{Land use categories}
+\subsection{Land classification}
 
@@
 Forest Land). Forest Land is defined as at least 10-30\% crown cover of
-trees with potential to reach a minimum height of 2-5 m \emph{in situ}
+trees with potential to reach a minimum height of 2-5 m \emph{in situ},
+and shorter-stature natural vegetation would be classified as Grassland
 \citep{ipcc_good_2003}. Definitions of forest are allowed to vary by
@@
 
-Inventories are conducted only for managed land, which is defined by
-IPCC to include ``all forests subject to some kind of human interactions
-(notably commercial management, harvest of industrial round-wood (logs)
-and fuelwood, production and use of wood commodities, and forest managed
-for amenity value or environmental protection if specified by the
-country), with defined geographical boundaries'' \citep{ipcc_good_2003}.
-As such, managed land constitutes all or the majority of Forest Land in
-most countries; however, many countries have not yet reported their
-approach for defining managed land
+The UNFCCC requires greenhouse gas reporting for all managed lands in a
+country, where management is defined as ``human interventions and
+practices have been applied to perform production, ecological or social
+functions'' \citep{ipcc_2006_2006}. This expansive definition of managed
+land implies that the majority of Forest Land in most countries is
+managed {[}e.g.,{]}. However, the definition is applied differently
+across countries, and the majority of governments have yet to report
+their approach for defining managed land or provide maps of managed land
 \citep{ogle_delineating_2018, deng_comparing_2021}.
@@
 lags behind records transfer and had not yet been completed for all
-records sent as of April 18, 2022.
+records sent as of April 19, 2022.
 
@@
 complex mapping was required: (1) mapping of C cycle variables (section
-4.1.1) and (2) determination of land use categories (section 4.1.2).
+4.1.1) and (2) land classification (section 4.1.2).
 
@@
 
-\subsubsection{Land use categories}
+\subsubsection{Land classification}
+
+Determination of the IPCC land-use category (i.e., Forest Land,
+Grassland, Wetlands, Cropland, Settlements, or Other Land; section 2.2)
+was made based on the categorical ForC field \emph{dominant.life.form},
+sometimes drawing upon stand age. Records with ``woody''
+\emph{dominant.life.form} were classified as Forest Land. Those with
+\emph{dominant.life.form} of ``woody+grass'', which in ForC is
+indicative of anything from a shrub-encroached grassland to a
+tree-dominated savanna, were given dual classification of Forest Land
+and Grassland. This dual classification indicates that records may be
+relevant to either category depending on the definition of forest
+applied (varies by country). For (rare) cases where
+\emph{dominant.life.form} was grass and stand age was greater than zero,
+indicative of early successional vegetation, we assigned a
+classification of Forest Land, consistent with the IPCC definition that
+Forest Land includes land expected to succeed to forest. Cases where
+\emph{dominant.life.form} was grass or crop and stand age was zero were
+indicative of a control for studies of forest regrowth following
+agricultural abandonment, and were classified as Grassland and Cropland,
+respectively.
+
+\emph{(now describe sub-category classification, which depends on past
+land use)}
 
@@
 
-The UNFCCC requires greenhouse gas reporting for all managed lands in a
-country, where management is defined as ``human interventions and
-practices have been applied to perform production, ecological or social
-functions'' {[}\emph{IPCC 2006 full report REF}{]}. This definition is
-applied differently across countries, and is not clearly defined by the
-majority of governments \citep{ogle_delineating_2018}. Given this, and
-because the IPCC definition of management does not necessarily match
-that which would be reported in scientific publications and hence in
-ForC, we do not transfer any classification of land management status
-from ForC to the EFDB, but do provide auxiliary info that may be useful
-in making this determination (e.g., geographical location).
+Given the lack of public information needed to determine whether lands
+are classified as mangaged
+\citep{ogle_delineating_2018, deng_comparing_2021}, and because the
+IPCC's definition of managed land is more expansive than is commonly
+applied in the scientific literature and hence in ForC, we did not
+transfer any classification of land management status from ForC to the
+EFDB. However, we do provide auxiliary information that should be useful
+in making this determination, including geographical location and
+notable disturbance events.
 
@@
 
-As of April 18, 2022, ForC (v4.0) contained NA independent records
+As of April 19, 2022, ForC (v4.0) contained NA independent records
 (39838 total), NA of which were for variables relevant to EFDB (Fig. 1).
@@
 
-As of April 18, 2022, we had reviewed and sent NA records to EFDB, 73 of
+As of April 19, 2022, we had reviewed and sent NA records to EFDB, 73 of
 which have been reviewed, accepted, and posted (Figs. 2-3, Table 2).
@@
 \hline
-forest census methods & Census stems down to the smallest diameter feasible. & IPCC biomass pool definition includes all living vegetation\\
-\hline
- & taxa to include: & IPCC biomass pool definition includes all living vegetation\\
-\hline
-dead organic matter sampling & include damage estimates on standing dead trees & \\
-\hline
- & adopt IPCC recommendations for min diameter of deadwood & \\
-\hline
-belowground sampling & Select soil sampling increments to include a cutoff at 30 cm depth. & \\
+forest census methods & Adopt IPCC guidelines (country-specific) for minimum stem size in censues in census and reporting.
+Ideally, census stem down to the smallest diameter feasible.
+& IPCC biomass pool definition includes all living vegetation\\
+\hline
+ & *taxa to include* & \\
+\hline
+dead organic matter sampling & *include damage estimates on standing dead trees* & \\
+\hline
+ & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & \\
+\hline
+belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & IPCC biomass pool definition includes all living vegetation\\
 \hline

### 2022-04-19 doc/manuscript/ESSDmanuscript.Rmd
finished description of mapping land classification

@@
 
-*(now describe sub-category classification, which depends on past land use)*
-
-Documented at https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/defining_land_subcategory.md, https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/IPCC_LandUse_mapping.csv, and in [issue #8](https://github.com/forc-db/IPCC_database_integration/issues/8).
+Classification into sub-categories was dependent upon stand age and site history (section `r  section_land_use_categories`).
+For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
+For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Cropland converted to Forest Land", respectively.
+For forests <20 years old with unspecified previous agricultural use, we assigned the sub-category "Land Converted to Forest land".
+Forests <20 years old with unknown land use prior to the study date were simply classified as "Forest Land".
+The same logic was applied for savannas, but including both Forest Land and Grassland as potentially relevant categories.
 

### 2022-04-19 doc/manuscript/ESSDmanuscript.tex
finished description of mapping land classification

@@
 
-\emph{(now describe sub-category classification, which depends on past
-land use)}
-
-Documented at
-https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB\_mapping/defining\_land\_subcategory.md,
-https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB\_mapping/IPCC\_LandUse\_mapping.csv,
-and in
-\href{https://github.com/forc-db/IPCC_database_integration/issues/8}{issue
-\#8}.
+Classification into sub-categories was dependent upon stand age and site
+history (section 2.2). For Forest Land ≥ 20 years old or of unknown
+(relatively mature) age, or Forest Land \textless{} 20 years old that
+was forest prior to a stand-clearing disturbance, the past land-use
+category was Forest Land, making the sub-category ``Forest Land
+Remaining Forest land''. For forests \textless20 years old with history
+including cultivation/ tillage or grazing, past land-use categories were
+Cropland and Grassland, respectively, making land-use subcategories were
+``Cropland converted to Forest Land'' and ``Cropland converted to Forest
+Land'', respectively. For forests \textless20 years old with unspecified
+previous agricultural use, we assigned the sub-category ``Land Converted
+to Forest land''. Forests \textless20 years old with unknown land use
+prior to the study date were simply classified as ``Forest Land''. The
+same logic was applied for savannas, but including both Forest Land and
+Grassland as potentially relevant categories.
 

### 2022-04-20 doc/manuscript/ESSDmanuscript.Rmd
various updates

@@
 author:
+  - given_name: Kristina J.
+    surname: Anderson-Teixeira
+    affiliation: "1,2 *"
+    email: teixeirak@si.edu
+    corresponding: true
+  - given_name: Valentine
+    surname: Herrmann
+    affiliation: "1"
   - given_name: Madison 
@@
     affiliation: "1"
-  - given_name: Valentine
-    surname: Herrmann
+  - given_name: Teagan
+    surname: Rogers
     affiliation: "1"
@@
     surname: Banbury Morgan
-    affiliation: "1,2" 
+    affiliation: "1,3" 
   - given_name: Ben 
     surname: Bond-Lamberty
-    affiliation: "3" 
+    affiliation: "4" 
   - given_name: Susan 
     surname: Cook-Patton
-    affiliation: "4" 
+    affiliation: "5" 
   - given_name: Helene
     surname: Muller-Landau
-    affiliation: "5" 
+    affiliation: "2" 
   - given_name: Camille
@@
     affiliation: "6"
-  - given_name: Teagan
-    surname: Rogers
-    affiliation: "1"
-  - given_name: Kristina J.
-    surname: Anderson-Teixeira
-    affiliation: "1,5 *"
-    email: teixeirak@si.edu
-    corresponding: true
+
 affiliation:
   - code: 1
-    address: Center for Conservatiton Ecology, Smithsonian Conservation Biology Institute, Front Royal, VA, United States
+    address: Center for Conservation Ecology, Smithsonian's National Zoo & Conservation Biology Institute, Front Royal, VA, United States
   - code: 2
+    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
+  - code: 3
     address: School of Geography, University of Leeds, Leeds, UK
-  - code: 3
+  - code: 4
     address: Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
-  - code: 4
+  - code: 5
     address: The Nature Conservancy; Arlington VA 22203, USA
-  - code: 5
-    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
   - code: 6
@@
   title: IPCC forest C accounting with ForC
-  author: Williams et al.
+  author: Anderson-Teixeira et al.
 competinginterests: |
@@
 
-n_independent_records_ForC= NA #suspected duplicates removed. This is the number of records fin ForC simplified.
+ForC_simplified <- read.csv(paste0(path_to_ForC,"/ForC/ForC_simplified/ForC_simplified.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+n_independent_records_ForC= length(ForC_simplified[,"measurement.ID"]) #suspected duplicates removed. This is the number of records fin ForC simplified.
+
 n_records_EFDB_variables= NA #n independent records relevant to EFDB
+
+n_records_checked = NA # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
+n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
+
 n_records_sent_to_EFDB = NA # n records sent to EFDB
+
 n_records_posted_EFDB = 73 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
@@
 n_fields_modified =  length(sampling_details[,"Column"]) + 4
-n_records_checked = NA # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
-n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
 
@@
 ForC contained data for **##** of the **##** variables relevant to EFDB (Table `r table_variables`, Fig. `r fig_variable_mapping`).
+The records were very unevenly distributed across variables ...
 
@@
 
-*Other EFDB-relevant variables require more effort but are warrented given their importance for forest C accounting.*
-Given widespread trends of increasing tree mortality[@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.
+*Other EFDB-relevant variables require more effort but are warranted given their importance for forest C accounting.*
+Given widespread trends of increasing tree mortality [@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.
 ...

### 2022-04-20 doc/manuscript/ESSDmanuscript.tex
various updates

@@
 
+\Author[1,2 *]{Kristina J.}{Anderson-Teixeira}
+\Author[1]{Valentine}{Herrmann}
 \Author[1]{Madison}{Williams}
-\Author[1]{Valentine}{Herrmann}
-\Author[1,2]{Rebecca}{Banbury Morgan}
-\Author[3]{Ben}{Bond-Lamberty}
-\Author[4]{Susan}{Cook-Patton}
-\Author[5]{Helene}{Muller-Landau}
+\Author[1]{Teagan}{Rogers}
+\Author[1,3]{Rebecca}{Banbury Morgan}
+\Author[4]{Ben}{Bond-Lamberty}
+\Author[5]{Susan}{Cook-Patton}
+\Author[2]{Helene}{Muller-Landau}
 \Author[6]{Camille}{Piponiot}
-\Author[1]{Teagan}{Rogers}
-\Author[1,5 *]{Kristina J.}{Anderson-Teixeira}
-
-\affil[1]{Center for Conservatiton Ecology, Smithsonian Conservation
-Biology Institute, Front Royal, VA, United States}
-\affil[2]{School of Geography, University of Leeds, Leeds, UK}
-\affil[3]{Joint Global Change Research Institute, Pacific Northwest
+
+\affil[1]{Center for Conservation Ecology, Smithsonian's National Zoo \&
+Conservation Biology Institute, Front Royal, VA, United States}
+\affil[2]{Forest Global Earth Observatory, Smithsonian Tropical Research
+Institute, Panama, Republic of Panama}
+\affil[3]{School of Geography, University of Leeds, Leeds, UK}
+\affil[4]{Joint Global Change Research Institute, Pacific Northwest
 National Laboratory, College Park, MD, United States}
-\affil[4]{The Nature Conservancy; Arlington VA 22203, USA}
-\affil[5]{Forest Global Earth Observatory, Smithsonian Tropical Research
-Institute, Panama, Republic of Panama}
+\affil[5]{The Nature Conservancy; Arlington VA 22203, USA}
 \affil[6]{CIRAD, Montpellier, France}
@@
 
-\runningauthor{Williams et al.}
+\runningauthor{Anderson-Teixeira et al.}
 
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-For C currently contains 39838 records from 10585 plots in 1531 distinct
+For C currently contains 39846 records from 10591 plots in 1531 distinct
 geographical areas, along with records of stand age and disturbance
@@
 control, manually reviewed \textgreater\# records to obtain additional
-required information, and added 305 new records.
+required information, and added 313 new records.
 
@@
 
-In addition to reviewing existing records, we added a total of 305 new
+In addition to reviewing existing records, we added a total of 313 new
 records to ForC. These included \textbf{\#\#} records from \textbf{\#\#}
@@
 lags behind records transfer and had not yet been completed for all
-records sent as of April 19, 2022.
+records sent as of April 20, 2022.
 
@@
 
-As of April 19, 2022, ForC (v4.0) contained NA independent records
-(39838 total), NA of which were for variables relevant to EFDB (Fig. 1).
+As of April 20, 2022, ForC (v4.0) contained 32540 independent records
+(39846 total), NA of which were for variables relevant to EFDB (Fig. 1).
 These records were distributed across all forested continents and
@@
 ForC contained data for \textbf{\#\#} of the \textbf{\#\#} variables
-relevant to EFDB (Table 2, Fig. 1).
+relevant to EFDB (Table 2, Fig. 1). The records were very unevenly
+distributed across variables \ldots{}
 
@@
 
-As of April 19, 2022, we had reviewed and sent NA records to EFDB, 73 of
+As of April 20, 2022, we had reviewed and sent NA records to EFDB, 73 of
 which have been reviewed, accepted, and posted (Figs. 2-3, Table 2).
@@
 \emph{Other EFDB-relevant variables require more effort but are
-warrented given their importance for forest C accounting.} Given
-widespread trends of increasing tree mortality\citep{refs}, including
+warranted given their importance for forest C accounting.} Given
+widespread trends of increasing tree mortality \citep{refs}, including
 through severe natural disturbance \citep{refs}, better characterization

### 2022-04-20 scripts/3_create_list_of_ all_data_sources_sent.R
give a try to the table #35
let's give this a try, @teixeirak . You'll need to update your rmd as follow: (ignore first column in kable, but use it to indicate what needs to be bolded - that way, if we add/remove,/change order, you won't need to update that)
knitr::kable(C_variables_table[,-1], format = "latex",
             longtable = TRUE,
             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1. ") %>%
  kableExtra::kable_styling(fu

@@ -1,4 +1,6 @@
-# Generate a summary list, in excel format, of all data sources; and for each data source the number of data submitted, clustered by carbon pool, with univocal reference to the relevant EFDB entry form in which data have been complied."
+# Generate a summary list of all data sources; and for each data source the number of data submitted, clustered by carbon pool, with univocal reference to the relevant EFDB entry form in which data have been complied.
+
+# also, generate a table of records of ForC variables relevant to, and sent to, EFDB (see https://github.com/forc-db/IPCC-EFDB-integration/issues/35 )
 
 
 
@@ -10,13 +12,21 @@ rm(list = ls())
 # load data ####
 
 # ## citation data
-# CITATIONS <- read.csv("https://raw.githubusercontent.com/forc-db/ForC/master/data/ForC_citations.csv", stringsAsFactors = F)
+CITATIONS <- read.csv("https://raw.githubusercontent.com/forc-db/ForC/master/data/ForC_citations.csv", stringsAsFactors = F)
+MEASUREMENTS <- read.csv("https://raw.githubusercontent.com/forc-db/ForC/master/data/ForC_measurements.csv", stringsAsFactors = F)
+VARIABLES <- read.csv("https://raw.githubusercontent.com/forc-db/ForC/master/data/ForC_variables.csv", stringsAsFactors = F)
+
+ForC_simplified <- read.csv("https://raw.githubusercontent.com/forc-db/ForC/master/ForC_simplified/ForC_simplified.csv", stringsAsFactors = F)
+
+
+V_mapping <- read.csv("doc/ForC-EFDB_mapping/ForC_variables_mapping.csv")
+trace_of_measurement_IDs <- read.csv("data/3-EFDB-forms-ready/trace_of_measurement_ID_processed.csv")
 
 ## list of files sent
 list_files_sent <- list.files("data/4-transferred-to-EFDB/", pattern = ".txt", full.names = T)
 
 
-# Create the table with all we need ####
+# Create the table with all we need about the data we submitted ####
 summary_table <- NULL
 
 for(f in list_files_sent) {
@@ -45,3 +55,78 @@ summary_table <- summary_table[order(summary_table$Full.Technical.Reference),]
 
 # save table ####
[…]

### 2022-04-20 doc/manuscript/ESSDmanuscript.Rmd
knit in new table
#36

@@
 library(kableExtra)
-knitr::kable(C_variables_table, format = "latex",
+knitr::kable(C_variables_table[,-1], format = "latex",
              longtable = TRUE,
@@
   row_spec(0, bold = TRUE)  %>%
-  row_spec(1, bold = TRUE)  %>% #biomass
-  row_spec(6, bold = TRUE)  %>% #agb
-  row_spec(14, bold = TRUE) %>% #root
-  row_spec(25, bold = TRUE)  %>% #dead wood
-  row_spec(33, bold = TRUE)  %>% #litter
-  row_spec(40, bold = TRUE)  %>% #som
-  row_spec(44, bold = TRUE)  #total
+  row_spec(which(C_variables_table$bold), bold = TRUE) #total
 ```

### 2022-04-20 doc/manuscript/ESSDmanuscript.Rmd
misc

@@
 
-n_records_checked = NA # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
+n_records_reviewed_or_added= NA
 n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
+n_records_checked = n_records_reviewed_or_added - n_records_added # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
 
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed and sent `r n_records_sent_to_EFDB` records to EFDB, `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
+As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were sent to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
 `r fig_map`-`r fig_histograms`, Table `r table_variables`).

### 2022-04-20 doc/manuscript/ESSDmanuscript.tex
knit in new table
#36
misc

@@
 \hline
-biomass &  &  &  & \\
-\hline
-delta.biomass &  &  &  & \\
-\hline
-NPP\_woody &  &  &  & \\
-\hline
-woody.mortality &  &  &  & \\
+biomass & 1095 & 847 & 93 & 48\\
+\hline
+delta.biomass & 0 & 0 & 0 & 0\\
+\hline
+NPP\_woody & 136 & 93 & 0 & 0\\
+\hline
+woody.mortality & 0 & 0 & 0 & 0\\
 \hline
@@
 \hline
-biomass\_ag &  &  &  & \\
-\hline
-biomass\_ag\_woody &  &  &  & \\
-\hline
-biomass\_ag\_foliage &  &  &  & \\
-\hline
-delta.agb &  &  &  & \\
-\hline
-ANPP\_woody &  &  &  & \\
-\hline
-ANPP\_foliage &  &  &  & \\
-\hline
-woody.mortality\_ag &  &  &  & \\
+biomass\_ag & 9038 & 7721 & 1230 & 693\\
+\hline
+biomass\_ag\_woody & 460 & 366 & 10 & 10\\
+\hline
+biomass\_ag\_foliage & 601 & 502 & 49 & 27\\
+\hline
+delta.agb & 166 & 128 & 123 & 123\\
+\hline
+ANPP\_woody & 299 & 242 & 0 & 0\\
+\hline
+ANPP\_foliage & 0 & 0 & 0 & 0\\
+\hline
+woody.mortality\_ag & 113 & 59 & 17 & 17\\
 \hline
@@
 \hline
-biomass\_root &  &  &  & \\
-\hline
-biomass\_root\_fine &  &  &  & \\
-\hline
-biomass\_root\_coarse &  &  &  & \\
-\hline
-delta.biomass\_root &  &  &  & \\
-\hline
-delta.biomass\_root\_coarse &  &  &  & \\
-\hline
-delta.biomass\_root\_fine &  &  &  & \\
-\hline
-woody.mortality\_root &  &  &  & \\
-\hline
-BNPP\_root\_fine &  &  &  & \\
-\hline
-BNPP\_root.turnover\_fine &  &  &  & \\
-\hline
-BNPP\_root\_coarse &  &  &  & \\
+biomass\_root & 4629 & 4180 & 123 & 55\\
+\hline
+biomass\_root\_fine & 931 & 594 & 18 & 18\\
+\hline
+biomass\_root\_coarse & 599 & 410 & 12 & 7\\
+\hline
+delta.biomass\_root & 0 & 0 & 0 & 0\\
+\hline
+delta.biomass\_root\_coarse & 0 & 0 & 0 & 0\\
+\hline
+delta.biomass\_root\_fine & 0 & 0 & 0 & 0\\
+\hline
+woody.mortality\_root & 0 & 0 & 0 & 0\\
+\hline
+BNPP\_root\_fine & 489 & 332 & 0 & 0\\
+\hline
+BNPP\_root.turnover\_fine & 91 & 56 & 0 & 0\\
+\hline
+BNPP\_root\_coarse & 329 & 250 & 0 & 0\\
 \hline
@@
 \hline
-deadwood &  &  &  & \\
-\hline
-deadwood\_standing &  &  &  & \\
-\hline
-deadwood\_down &  &  &  & \\
-\hline
-delta.deadwood &  &  &  & \\
-\hline
-delta.deadwood\_standing &  &  &  & \\
-\hline
-delta.deadwood\_down &  &  &  & \\
-\hline
-R\_het\_deadwood & 0 &  &  & \\
+deadwood & 438 & 301 & 94 & 61\\
+\hline
+deadwood\_standing & 152 & 120 & 17 & 17\\
+\hline
+deadwood\_down & 424 & 368 & 51 & 27\\
+\hline
+delta.deadwood & 0 & 0 & 0 & 0\\
+\hline
+delta.deadwood\_standing & 0 & 0 & 0 & 0\\
+\hline
+delta.deadwood\_down & 0 & 0 & 0 & 0\\
+\hline
+R\_het\_deadwood & 0 & 0 & 0 & 0\\
 \hline
@@
 \hline
-O.horizon &  &  &  & \\
-\hline
-delta.O.horizon &  &  &  & \\
-\hline
-litter &  &  &  & \\
-\hline
-delta.litter &  &  &  & \\
-\hline
-NPP\_litterfall &  &  &  & \\
-\hline
-R\_het\_litter &  &  &  & \\
+O.horizon & 38 & 38 & 38 & 38\\
+\hline
+delta.O.horizon & 4 & 4 & 4 & 4\\
+\hline
+litter & 30 & 30 & 23 & 23\\
+\hline
+delta.litter & 0 & 0 & 0 & 0\\
+\hline
+NPP\_litter & 0 & 0 & 0 & 0\\
+\hline
+R\_het\_litter & 167 & 143 & 0 & 0\\
 \hline
@@
 \hline
-SOM / SOC &  &  &  & \\
-\hline
-delta.SOM / delta.SOC &  &  &  & \\
-\hline
-R\_het\_soil &  &  &  & \\
+SOM / SOC & 693 & 396 & 79 & 46\\
+\hline
+delta.SOM / delta.SOC & 0 & 0 & 0 & 0\\
+\hline
+R\_het\_soil & 0 & 0 & 0 & 0\\
 \hline
@@
 
-As of April 20, 2022, we had reviewed and sent NA records to EFDB, 73 of
-which have been reviewed, accepted, and posted (Figs. 2-3, Table 2).
+As of April 20, 2022, we had reviewed or added NA EFDB-relevant records,
+NA records of which were sent to EFDB, and 73 of which have been
+reviewed, accepted, and posted (Figs. 2-3, Table 2).
 \emph{{[}DETAILS{]}}

### 2022-04-21 scripts/3_create_list_of_ all_data_sources_sent.R
Update 3_create_list_of_ all_data_sources_sent.R

@@ -118,7 +118,7 @@ varibales_in_order$pool <- NULL
 
 # add a row for the total
 
-varibales_in_order <- rbind(varibales_in_order, c(TRUE, "TOTAL", t(colSums(varibales_in_order[, -c(1,2)]))))
+varibales_in_order <- rbind(varibales_in_order, c(TRUE, "TOTAL", t(colSums(varibales_in_order[, -c(1,2)], na.rm = T))))
 
 
 colnames(varibales_in_order) <- gsub("_", " ", colnames(varibales_in_order))

### 2022-04-22 doc/manuscript/ESSDmanuscript.Rmd
set up auto population of statistics, writing (data reporting needs)

@@
 
-n_records_EFDB_variables= NA #n independent records relevant to EFDB
+C_variables <- read.csv(paste0(path_to_dir,"/doc/manuscript/figures_tables/C_variables.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+n_records_EFDB_variables= C_variables[length(C_variables[,"variable"]), "n independent records in ForC"] #n independent records relevant to EFDB
 
-n_records_reviewed_or_added= NA
+n_records_reviewed_or_added= C_variables[length(C_variables[,"variable"]), "n reviewed"]
 n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
@@
 
-n_records_sent_to_EFDB = NA # n records sent to EFDB
+n_records_sent_to_EFDB = C_variables[length(C_variables[,"variable"]), "n sent to EFDB"] # n records sent to EFDB
 
@@
 
-We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards (`r table_recommendations`).
+We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards (Table `r table_recommendations`).
+It is particularly noteworthy that simple decisions on the presentation of results will determine whether the data meet the criteria for inclusion in EFDB.
+Some examples are as follows:
+(1) presenting data only in a figure makes it ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion;
+(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies data from inclusion;
+(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
+It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
 
-For data synthesis projects, compilation can only be useful to the EFDB if they include all the required, along with transparent description on the methodology applied to derive emission factors (or have a brief description and a reference to the original source) and the original emission factor values are present (not modified/rounded).
+For those compiling published data (e.g., for meta-analyses), the data set can have added value if all information required by IPCC is extracted from original publications. 
+This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
+The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.
 
-**Contributing data to ForC and/or EFDB directly will ensure its broader impact.**
-The latter is more efficient for getting data to EFDB, but does not get the data into ForC, where it can be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].
+Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipline developed here.
+For individual publications, the former option will generally be more efficient.
+However, by getting the data into ForC as well as EFDB, the latter option will allow the data to be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].
 

### 2022-04-22 doc/manuscript/ESSDmanuscript.tex
set up auto population of statistics, writing (data reporting needs)

@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-For C currently contains 39846 records from 10591 plots in 1531 distinct
+For C currently contains 39856 records from 10590 plots in 1531 distinct
 geographical areas, along with records of stand age and disturbance
@@
 control, manually reviewed \textgreater\# records to obtain additional
-required information, and added 313 new records.
+required information, and added 325 new records.
 
@@
 
-In addition to reviewing existing records, we added a total of 313 new
+In addition to reviewing existing records, we added a total of 325 new
 records to ForC. These included \textbf{\#\#} records from \textbf{\#\#}
@@
 lags behind records transfer and had not yet been completed for all
-records sent as of April 20, 2022.
+records sent as of April 22, 2022.
 
@@
 
-As of April 20, 2022, ForC (v4.0) contained 32540 independent records
-(39846 total), NA of which were for variables relevant to EFDB (Fig. 1).
-These records were distributed across all forested continents and
+As of April 22, 2022, ForC (v4.0) contained 32550 independent records
+(39856 total), 17190 of which were for variables relevant to EFDB (Fig.
+1). These records were distributed across all forested continents and
 ecozones, with particularly high concentrations in
@@
 \hline
-biomass\_ag & 9038 & 7721 & 1230 & 693\\
+biomass\_ag & 9038 & 7724 & 1232 & 693\\
 \hline
@@
 \hline
-woody.mortality\_ag & 113 & 59 & 17 & 17\\
+woody.mortality\_ag & 113 & 61 & 17 & 17\\
 \hline
@@
 \hline
-BNPP\_root\_fine & 489 & 332 & 0 & 0\\
+BNPP\_root\_fine & 489 & 333 & 0 & 0\\
 \hline
@@
 \hline
-deadwood & 438 & 301 & 94 & 61\\
+deadwood & 438 & 303 & 94 & 61\\
 \hline
@@
 \hline
-SOM / SOC & 693 & 396 & 79 & 46\\
+SOM / SOC & 693 & 398 & 81 & 46\\
 \hline
@@
 \hline
-\textbf{TOTAL} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
+\textbf{TOTAL} & \textbf{20922} & \textbf{17190} & \textbf{1985} & \textbf{1214}\\
 \hline
@@
 
-As of April 20, 2022, we had reviewed or added NA EFDB-relevant records,
-NA records of which were sent to EFDB, and 73 of which have been
-reviewed, accepted, and posted (Figs. 2-3, Table 2).
+As of April 22, 2022, we had reviewed or added 1985 EFDB-relevant
+records, 1214 records of which were sent to EFDB, and 73 of which have
+been reviewed, accepted, and posted (Figs. 2-3, Table 2).
 \emph{{[}DETAILS{]}}
@@
 the process of reviewing and sending data that are already in ForC
-(Figs. 2-3). So far, only \textasciitilde NA\% of the EFDB-relevant data
+(Figs. 2-3). So far, only \textasciitilde7\% of the EFDB-relevant data
 in ForC have been sent to EFDB.
@@
 otherwise, researchers calculate and report the values according to IPCC
-standards (3).
-
-For data synthesis projects, compilation can only be useful to the EFDB
-if they include all the required, along with transparent description on
-the methodology applied to derive emission factors (or have a brief
-description and a reference to the original source) and the original
-emission factor values are present (not modified/rounded).
-
-\textbf{Contributing data to ForC and/or EFDB directly will ensure its
-broader impact.} The latter is more efficient for getting data to EFDB,
-but does not get the data into ForC, where it can be more broadly
-useful--for example, being used for basic science
+standards (Table 3). It is particularly noteworthy that simple decisions
+on the presentation of results will determine whether the data meet the
+criteria for inclusion in EFDB. Some examples are as follows: (1)
+presenting data only in a figure makes it ineligible for inclusion in
+EFDB, whereas presentation in a table or supplementary data file allows
+inclusion; (2) direct presentation of all relevant variables allows
+inclusion, whereas presenting only components of variables of interest
+(e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or
+requiring simple mathematical operations to obtain a variable of
+interest (e.g., \emph{delta.agb} = \emph{ANPP\_woody} -
+\emph{woody.mortality.agb}) disqualifies data from inclusion; (3)
+matching IPCC-defined thresholds for defining C pools (Table 1), which
+may vary by country, can make the data far more relevant for IPCC
+accounting (e.g., using a 10 cm cutoff between dead wood and litter,
+presenting soil C to a depth of 30 cm). It should also be emphasized
+that reporting of 95\% confidence intervals (or other metrics of error),
+when applicable, is highly desirable and makes the data more relevant to
+IPCC.
+
+For those compiling published data (e.g., for meta-analyses), the data
+set can have added value if all information required by IPCC is
+extracted from original publications. This includes -- but is not
+limited to -- retaining original values as presented without
+modification or rounding, noting whether data were digitized, recording
+confidence intervals, and recording all required fields (as indicated in
+the EFDB's bulk import template). The significant effort required to map
+a database into EFDB has been accomplished here (Appendix B), and we
+welcome other researchers to use the ForC template.
+
+Once EFDB-relevant data are available in peer-reviewed publications,
+they may be submitted directly to EFDB or may use the ForC - EFDB data
+pipline developed here. For individual publications, the former option
+will generally be more efficient. However, by getting the data into ForC
+as well as EFDB, the latter option will allow the data to be more
+broadly useful--for example, being used for basic science
 \citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021}
@@
 Ideally, census stem down to the smallest diameter feasible.
-& IPCC biomass pool definition includes all living vegetation\\
+& IPCC biomass pool definition includes all living vegetation, but understory may be excluded when contribution is minor.\\
 \hline

### 2022-04-22 doc/manuscript/ESSDmanuscript.Rmd
move abstract out of front matter
this is so that statistics can be automatically inserted

@@
 abstract: |
-  Forests are critical for climate change mitigation and consitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
-  Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
-  Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
-  Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-  Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
-  We begin by reconciling terminology and mapping ForC fields into EFDB.
-  This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-  At the time of writing, ForC contained ## values that would qualify for inclusion in the EFDB, ## of which have been transferred to date.
-  (Some analysis of representation/ gaps.)
-  <!-- -->
-  In the future, forest C estimates in EFDB can be improved through targetted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
 
@@
 ```
+
+**Abstract.**
+Forests are critical for climate change mitigation and consitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
+Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
+Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
+Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
+Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
+We begin by reconciling terminology and mapping ForC fields into EFDB.
+This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
+At the time of writing, ForC contained ## values that would qualify for inclusion in the EFDB, ## of which have been transferred to date.
+(Some analysis of representation/ gaps.)
+<!-- -->
+In the future, forest C estimates in EFDB can be improved through targetted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
 

### 2022-04-22 doc/manuscript/ESSDmanuscript.tex
move abstract out of front matter
this is so that statistics can be automatically inserted

@@
 
-\begin{abstract}
-Forests are critical for climate change mitigation and consitute a
-substantial portion of planned emissions reductions under the 2015 Paris
-Agreement. Yet, the efficacy of greenhouse gas mitigation planning and
-reporting is dependent upon the quality of available emission factors
-data, including forest carbon (C) stocks and changes therein. Tens of
-thousands of relevant forest C estimates have been published, yet are
-not readily accesible to the practitioners compiling national greenhouse
-gas inventories. Many of these data have, however, been compiled in the
-Global Forest C database (ForC; https://forc-db.github.io/) and stand to
-be of value to greenhouse gas accounting if made available through the
-Emission Factor Database (EFDB) of the International Panel on Climate
-Change (IPCC). Here, we develop and document a process for
-semi-automated transfer of data from ForC into the EFDB, assess the data
-available and transferred to date, and provide recommendations for
-improving forest data collection, analysis, and reporting to improve
-accounting of forest-sector greenhoouse gas emissions and removals. We
-begin by reconciling terminology and mapping ForC fields into EFDB. This
-process required some updates to the ForC database structure, leading to
-the release of a new version of ForC (v4.0; described here). At the time
-of writing, ForC contained \#\# values that would qualify for inclusion
-in the EFDB, \#\# of which have been transferred to date. (Some analysis
-of representation/ gaps.) In the future, forest C estimates in EFDB can
-be improved through targetted research to fill critical gaps, reporting
-of information required by IPCC, and continued submission of data from
-scientific publications to the EFDB.
-\end{abstract}
+\textbf{Abstract.} Forests are critical for climate change mitigation
+and consitute a substantial portion of planned emissions reductions
+under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
+mitigation planning and reporting is dependent upon the quality of
+available emission factors data, including forest carbon (C) stocks and
+changes therein. Tens of thousands of relevant forest C estimates have
+been published, yet are not readily accesible to the practitioners
+compiling national greenhouse gas inventories. Many of these data have,
+however, been compiled in the Global Forest C database (ForC;
+https://forc-db.github.io/) and stand to be of value to greenhouse gas
+accounting if made available through the Emission Factor Database (EFDB)
+of the International Panel on Climate Change (IPCC). Here, we develop
+and document a process for semi-automated transfer of data from ForC
+into the EFDB, assess the data available and transferred to date, and
+provide recommendations for improving forest data collection, analysis,
+and reporting to improve accounting of forest-sector greenhoouse gas
+emissions and removals. We begin by reconciling terminology and mapping
+ForC fields into EFDB. This process required some updates to the ForC
+database structure, leading to the release of a new version of ForC
+(v4.0; described here). At the time of writing, ForC contained \#\#
+values that would qualify for inclusion in the EFDB, \#\# of which have
+been transferred to date. (Some analysis of representation/ gaps.) In
+the future, forest C estimates in EFDB can be improved through targetted
+research to fill critical gaps, reporting of information required by
+IPCC, and continued submission of data from scientific publications to
+the EFDB.
 

### 2022-04-22 doc/manuscript/ESSDmanuscript.Rmd
revise abstract

@@
 C_variables <- read.csv(paste0(path_to_dir,"/doc/manuscript/figures_tables/C_variables.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_records_EFDB_variables= C_variables[length(C_variables[,"variable"]), "n independent records in ForC"] #n independent records relevant to EFDB
+n_ind_records_EFDB_variables= C_variables[length(C_variables[,"variable"]), "n independent records in ForC"] #n independent records relevant to EFDB
 
@@
 This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-At the time of writing, ForC contained ## values that would qualify for inclusion in the EFDB, ## of which have been transferred to date.
-(Some analysis of representation/ gaps.)
-<!-- -->
-In the future, forest C estimates in EFDB can be improved through targetted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
+At the time of writing, ForC contained ~`r n_ind_records_EFDB_variables` independent records that would be relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
+Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil carbon, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
+Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
+ForC represents a diversity of stand ages, although records for young stands are primarily limited to C stocks, as opposed to net increments of fluxes.
+This distribution of records is generally reflected in the subset of records that have been submitted to EFDB to date.
+In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
 
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_records_EFDB_variables` of which were for variables relevant to EFDB (Fig.
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for variables relevant to EFDB (Fig.
 `r fig_variable_mapping`).
@@
 There is vast potential to expand forest C data in EFDB by completing the process of reviewing and sending data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
-So far, only ~`r round(n_records_sent_to_EFDB/n_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.
+So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.
 

### 2022-04-22 doc/manuscript/ESSDmanuscript.tex
revise abstract

@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). At the time of writing, ForC contained \#\#
-values that would qualify for inclusion in the EFDB, \#\# of which have
-been transferred to date. (Some analysis of representation/ gaps.) In
-the future, forest C estimates in EFDB can be improved through targetted
-research to fill critical gaps, reporting of information required by
-IPCC, and continued submission of data from scientific publications to
-the EFDB.
+(v4.0; described here). At the time of writing, ForC contained
+\textasciitilde17190 independent records that would be relevant to EFDB,
+1214 of which have been submitted to date. Among the data in ForC, there
+is disproportionate representation of biomass (particularly aboveground)
+stocks, with far fewer records for dead organic matter and soil carbon,
+and relatively few or no records for net annual increments or C fluxes
+into (gains) or out of (losses) the IPCC-defined C pools. Geographic
+representation is also quite uneven, with the highest densities of
+relevant records in temperate forests, and with relatively scant
+representation of tropical forests in Africa and Asia. ForC represents a
+diversity of stand ages, although records for young stands are primarily
+limited to C stocks, as opposed to net increments of fluxes. This
+distribution of records is generally reflected in the subset of records
+that have been submitted to EFDB to date. In the future, forest C
+estimates in EFDB can be improved through targeted research to fill
+critical gaps, reporting of information required by IPCC, and continued
+submission of data from scientific publications to the EFDB.
 

### 2022-04-26 doc/manuscript/ESSDmanuscript.tex
reposition table
add in-prep note

@@
 
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
+\emph{THIS IS AN IN-PREP MANUSCRIPT.}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
 \textbf{Abstract.} Forests are critical for climate change mitigation
@@
 lags behind records transfer and had not yet been completed for all
-records sent as of April 22, 2022.
+records sent as of April 26, 2022.
 
@@
 
-As of April 22, 2022, ForC (v4.0) contained 32550 independent records
+As of April 26, 2022, ForC (v4.0) contained 32550 independent records
 (39856 total), 17190 of which were for variables relevant to EFDB (Fig.
@@
 
-As of April 22, 2022, we had reviewed or added 1985 EFDB-relevant
+As of April 26, 2022, we had reviewed or added 1985 EFDB-relevant
 records, 1214 records of which were sent to EFDB, and 73 of which have
@@
 
+\begin{table}
+
+\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
+\centering
+\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\hline
+\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
+\hline
+variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data.
+& Estimates of relevant variables are not always calcualted.\\
+\hline
+forest census methods & Adopt IPCC guidelines (country-specific) for minimum stem size in censues in census and reporting.
+Ideally, census stem down to the smallest diameter feasible.
+& IPCC biomass pool definition includes all living vegetation, but understory may be excluded when contribution is minor.\\
+\hline
+ & *taxa to include* & \\
+\hline
+dead organic matter sampling & *include damage estimates on standing dead trees* & \\
+\hline
+ & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & \\
+\hline
+belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & IPCC biomass pool definition includes all living vegetation\\
+\hline
+reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables.
+& EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
+\hline
+reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
+\hline
+reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
+\hline
+\end{tabu}
+\end{table}
+
 For those compiling published data (e.g., for meta-analyses), the data
@@
 or model benchmarking \citep{fer_ecosystem_2021}.
-
-\begin{table}
-
-\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
-\centering
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\hline
-\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
-\hline
-variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data.
-& Estimates of relevant variables are not always calcualted.\\
-\hline
-forest census methods & Adopt IPCC guidelines (country-specific) for minimum stem size in censues in census and reporting.
-Ideally, census stem down to the smallest diameter feasible.
-& IPCC biomass pool definition includes all living vegetation, but understory may be excluded when contribution is minor.\\
-\hline
- & *taxa to include* & \\
-\hline
-dead organic matter sampling & *include damage estimates on standing dead trees* & \\
-\hline
- & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & \\
-\hline
-belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & IPCC biomass pool definition includes all living vegetation\\
-\hline
-reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables.
-& EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
-\hline
-reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
-\hline
-reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
-\hline
-\end{tabu}
-\end{table}
 

### 2022-04-27 doc/manuscript/ESSDmanuscript.Rmd
reposition table
add in-prep note
improve table formatting

@@
     - \usepackage{caption}
+    #- \usepackage{booktabs}
+    #- \usepackage{makecell}
 ---
@@
 ```
+
+****
+*THIS IS AN IN-PREP MANUSCRIPT.*
+
+****
 
@@
              caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
-  kableExtra::kable_styling(full_width = TRUE) %>%
-  #column_spec(1, width = ".7cm") %>% #This doesn't seem to work
-  row_spec(0, bold = TRUE)
+  row_spec(0, bold = TRUE) %>%
+  kableExtra::kable_styling(full_width = TRUE) 
+  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
 ```
@@
 
+\newpage
 ```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 knitr::kable(C_variables_table[,-1], format = "latex",  
-             longtable = TRUE,
+             longtable = T,
              caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
 ") %>%
-  kableExtra::kable_styling(full_width = TRUE) %>%
   row_spec(0, bold = TRUE)  %>%
-  row_spec(which(C_variables_table$bold), bold = TRUE) #total
+  row_spec(which(C_variables_table$bold), bold = TRUE) %>% #total
+  kable_styling(font_size = 8) %>%
+  #kableExtra::kable_styling(full_width = TRUE) 
+  kable_styling(latex_options = c("hold_position", "repeat_header"), protect_latex = T) 
 ```
@@
 
-For those compiling published data (e.g., for meta-analyses), the data set can have added value if all information required by IPCC is extracted from original publications. 
-This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
-The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.
-
-Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipline developed here.
-For individual publications, the former option will generally be more efficient.
-However, by getting the data into ForC as well as EFDB, the latter option will allow the data to be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].
-
 ```{r table_recommendations, eval = TRUE, echo=FALSE, warning=FALSE}
@@
              caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....") %>%
-  kableExtra::kable_styling(full_width = TRUE) %>%
-  row_spec(0, bold = TRUE)
+  row_spec(0, bold = TRUE)%>%
+  kable_styling(font_size = 10) %>%
+  kableExtra::kable_styling(full_width = TRUE) 
+  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
 ```
+
+For those compiling published data (e.g., for meta-analyses), the data set can have added value if all information required by IPCC is extracted from original publications. 
+This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
+The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.
+
+Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipline developed here.
+For individual publications, the former option will generally be more efficient.
+However, by getting the data into ForC as well as EFDB, the latter option will allow the data to be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].
 
@@
   kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
+  kable_styling(font_size = 8) %>%
   row_spec(0, bold = TRUE)
@@
   row_spec(0, bold = TRUE) %>%
+  kable_styling(font_size = 8) %>%
 kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))

### 2022-04-27 doc/manuscript/ESSDmanuscript.tex
improve table formatting

@@
 lags behind records transfer and had not yet been completed for all
-records sent as of April 26, 2022.
+records sent as of April 27, 2022.
 
@@
 
-As of April 26, 2022, ForC (v4.0) contained 32550 independent records
+As of April 27, 2022, ForC (v4.0) contained 32550 independent records
 (39856 total), 17190 of which were for variables relevant to EFDB (Fig.
@@
 
-\begin{longtabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
+\newpage
+\begingroup\fontsize{8}{10}\selectfont
+
+\begin{longtable}[t]{l|l|l|l|l}
 \caption{\label{tab:table_variables}\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
@@
 \hline
+\endfirsthead
+\caption[]{\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
+\textit{(continued)}}\\
+\hline
+\textbf{variable} & \textbf{n in ForC} & \textbf{n independent records in ForC} & \textbf{n reviewed} & \textbf{n sent to EFDB}\\
+\hline
+\endhead
 \textbf{Biomass} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
@@
 \hline
-\end{longtabu}
+\end{longtable}
+\endgroup{}
 
@@
 
-As of April 26, 2022, we had reviewed or added 1985 EFDB-relevant
+As of April 27, 2022, we had reviewed or added 1985 EFDB-relevant
 records, 1214 records of which were sent to EFDB, and 73 of which have
@@
 \centering
+\fontsize{10}{12}\selectfont
 \begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
@@
 Table A1: \textbf{Table of changes to ForC fields.}
+\begingroup\fontsize{8}{10}\selectfont
 
@@
 \end{longtabu}
+\endgroup{}
 
@@
 of table (STILL NEED TO BE PROPERLY INSERTED).
+\begingroup\fontsize{8}{10}\selectfont
 

### 2022-04-27 doc/manuscript/ESSDmanuscript.Rmd
minor

@@
              longtable = T,
-             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
-") %>%
+             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.}") %>%
   row_spec(0, bold = TRUE)  %>%

### 2022-04-27 doc/manuscript/ESSDmanuscript.tex
minor

@@
 \begin{longtable}[t]{l|l|l|l|l}
-\caption{\label{tab:table_variables}\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
-}\\
+\caption{\label{tab:table_variables}\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.}}\\
 \hline
@@
 \endfirsthead
-\caption[]{\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} Relationships of variables to IPCC-defined forest C pools (Table 1) and to each other are illustrated in Figure 1.
-\textit{(continued)}}\\
+\caption[]{\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} \textit{(continued)}}\\
 \hline

### 2022-04-27 doc/manuscript/ESSDmanuscript.tex
knit with updated table

@@
 (v4.0; described here). At the time of writing, ForC contained
-\textasciitilde17190 independent records that would be relevant to EFDB,
+\textasciitilde17204 independent records that would be relevant to EFDB,
 1214 of which have been submitted to date. Among the data in ForC, there
@@
 As of April 27, 2022, ForC (v4.0) contained 32550 independent records
-(39856 total), 17190 of which were for variables relevant to EFDB (Fig.
+(39856 total), 17204 of which were for variables relevant to EFDB (Fig.
 1). These records were distributed across all forested continents and
@@
 \hline
-biomass\_ag & 9038 & 7724 & 1232 & 693\\
+biomass\_ag & 9050 & 7737 & 1251 & 693\\
 \hline
@@
 \hline
-ANPP\_foliage & 0 & 0 & 0 & 0\\
-\hline
-woody.mortality\_ag & 113 & 61 & 17 & 17\\
+woody.mortality\_ag & 112 & 62 & 30 & 17\\
 \hline
@@
 \hline
-deadwood & 438 & 303 & 94 & 61\\
+deadwood & 437 & 303 & 103 & 61\\
 \hline
@@
 \hline
-SOM / SOC & 693 & 398 & 81 & 46\\
+SOM / SOC & 693 & 398 & 89 & 46\\
 \hline
@@
 \hline
-\textbf{TOTAL} & \textbf{20922} & \textbf{17190} & \textbf{1985} & \textbf{1214}\\
+\textbf{TOTAL} & \textbf{20932} & \textbf{17204} & \textbf{2034} & \textbf{1214}\\
 \hline
@@
 
-As of April 27, 2022, we had reviewed or added 1985 EFDB-relevant
+As of April 27, 2022, we had reviewed or added 2034 EFDB-relevant
 records, 1214 records of which were sent to EFDB, and 73 of which have

### 2022-04-27 doc/manuscript/ESSDmanuscript.Rmd
add FAO ecozone codes
https://www.fao.org/3/ad652e/ad652e10.htm

closes #33

@@
 ```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: ...."}
+Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems."}
 #this is for a two-column figure. for one-column figure, use out.width = "8.3cm"

### 2022-04-27 doc/manuscript/ESSDmanuscript.tex
add FAO ecozone codes
https://www.fao.org/3/ad652e/ad652e10.htm

closes #33

@@
 \includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: ....}\label{fig:fig_map}
+Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.}\label{fig:fig_map}
 \end{figure}

### 2022-04-29 doc/manuscript/ESSDmanuscript.Rmd
small updates

@@
 In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
+Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C accounting.
 
@@
 In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
-These included **##** records from **##** studies [@piponiot_distribution_2022; @refs<!--others?-->] that were not previously included in ForC.
+These included **##** records from **##** studies [@piponiot_distribution_2022; @lutz_largediameter_2021; @refs<!--others?-->] that were not previously included in ForC.
 In addition, we created new records for **##** EFDB-relevant estimates presented in the original publication that were not yet present in ForC.

### 2022-04-29 doc/manuscript/ESSDmanuscript.tex
small updates

@@
 critical gaps, reporting of information required by IPCC, and continued
-submission of data from scientific publications to the EFDB.
+submission of data from scientific publications to the EFDB. Given that
+climate change is rapidly impacting the world's forests, timely
+reporting of recent estimates will be especially critical to accurate
+forest C accounting.
 
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-For C currently contains 39856 records from 10590 plots in 1531 distinct
+For C currently contains 39860 records from 10590 plots in 1531 distinct
 geographical areas, along with records of stand age and disturbance
@@
 control, manually reviewed \textgreater\# records to obtain additional
-required information, and added 325 new records.
+required information, and added 329 new records.
 
@@
 
-In addition to reviewing existing records, we added a total of 325 new
+In addition to reviewing existing records, we added a total of 329 new
 records to ForC. These included \textbf{\#\#} records from \textbf{\#\#}
-studies \citep{piponiot_distribution_2022, refs} that were not
-previously included in ForC. In addition, we created new records for
-\textbf{\#\#} EFDB-relevant estimates presented in the original
-publication that were not yet present in ForC.
+studies
+\citep{piponiot_distribution_2022, lutz_largediameter_2021, refs} that
+were not previously included in ForC. In addition, we created new
+records for \textbf{\#\#} EFDB-relevant estimates presented in the
+original publication that were not yet present in ForC.
 
@@
 lags behind records transfer and had not yet been completed for all
-records sent as of April 27, 2022.
+records sent as of April 29, 2022.
 
@@
 
-As of April 27, 2022, ForC (v4.0) contained 32550 independent records
-(39856 total), 17204 of which were for variables relevant to EFDB (Fig.
+As of April 29, 2022, ForC (v4.0) contained 32609 independent records
+(39860 total), 17204 of which were for variables relevant to EFDB (Fig.
 1). These records were distributed across all forested continents and
@@
 
-As of April 27, 2022, we had reviewed or added 2034 EFDB-relevant
+As of April 29, 2022, we had reviewed or added 2034 EFDB-relevant
 records, 1214 records of which were sent to EFDB, and 73 of which have

### 2022-05-24 doc/manuscript/ESSDmanuscript.Rmd
paper
Update ESSDmanuscript.Rmd

@@
 ForC contained data for **##** of the **##** variables relevant to EFDB (Table `r table_variables`, Fig. `r fig_variable_mapping`).
-The records were very unevenly distributed across variables ...
+The records were very unevenly distributed across variables. The most records colected were for the Aboveground biomass variable(biomass_ag).This variable
 
@@
 It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
+Reports which had the most successful data transfers used EFDB variables and had clear tables showing their results.
 

### 2023-05-05 doc/manuscript/ESSDmanuscript.Rmd
edits through section 2.2.1 (biomass)

@@
 **Abstract.**
-Forests are critical for climate change mitigation and consitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
+Forests are critical for climate change mitigation and constitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
 Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
@@
 At the time of writing, ForC contained ~`r n_ind_records_EFDB_variables` independent records that would be relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
-Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil carbon, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
+Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
 Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
@@
 
-Reflecting their strong influence on Earth's climate, forests play a central role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
+Reflecting their strong influence on Earth's climate, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
 Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options, with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
@@
 The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
-This guidance has improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
-For example, the year following the release of the latest IPCC guidelines, @cook-patton_mapping_2020 found that the latest default rates may underestimate rates of C accumulation in regrowth forests by 32% on average and fail to capture eight-fold variation within ecozones.
-In addition, @cuni-sanchez_high_2021 found that aboveground C stocks in mature African tropical montane forests were two-thirds higher than the IPCC default values for these forests.
-This rapid evolution of scientific information on the climate mitigation potential of forests is beneficial to climate mitigation efforts, but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
+A tiered approach to accounting is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
+Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
+For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default values underestimated C sequestration by 32% on average and failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
+In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
+This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
 Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.
 
-To improve the data accessible for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
+To improve data accessibility for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
 The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
 The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include carbon stocks, net increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
+In the case of forests, emission factors include C stocks, net increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
 
-The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest carbon stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
+The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
 ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
@@
 Here, we
-(1) review IPCC methods and definitions for tallying forest C;
+(1) review IPCC methods and definitions for forest C accounting in the context of typical forest C estimation methodologies;
 (2) describe mapping of ForC to IPCC's EFDB;
@@
 The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
+For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
 For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006].
@@
 
-Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies.
+Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- scientific studies.
 IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
 While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
-In this section, we define and review the IPCC definitions in the context of typical forest C estimation methodologies.
 
@@
 Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
+IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify (issue 40)-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
 A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
-In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs].
-Beyond the minimum DBH sampled, forest censuses and biomass estimates also differ in their inclusion of life forms other than dicot trees -- including lianas, ferns, palms, and bamboo -- which in some places can reach large sizes and/or constitute a large fraction of forest C.
-*[explain IPCC guidance]*
-Further, it is important to note that this excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
+In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs_from_ForC], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs_from_ForC].
+While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although thhese do tend to be censused when they consitute a large proportion of the biomass [@refs_from_ForC].
+Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
 
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometries to convert DBH measurements to biomass.
-Selection of allometries has an enormous influence on estimates of biomass stocks, increments, of fluxes [@clark_landscapescale_2000;@clark_net_2001].
-While trusted and standardized allometric equations are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
-*[explain IPCC guidance]*
+A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass.
+Selection of allometric models has an enormous influence on estimates of biomass stocks, increments, of fluxes [@clark_landscapescale_2000;@clark_net_2001].
+While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
+IPCC Tier 1 values currently draw on studies applying a variety of allometric models [e.g., @requenasuarez_estimating_2019; @rozendaal_aboveground_2022].
 
 Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
-The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain.
+The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain [@ref].
 Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
-Field methods for estimating root biomass are highly variable.
+Field methods for estimating root biomass are highly variable [@ref].
 IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].
@@
 
+\newpage
 ```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.

### 2023-05-05 doc/manuscript/ESSDmanuscript.tex
edits through section 2.2.1 (biomass)

@@
 \textbf{Abstract.} Forests are critical for climate change mitigation
-and consitute a substantial portion of planned emissions reductions
+and constitute a substantial portion of planned emissions reductions
 under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
@@
 is disproportionate representation of biomass (particularly aboveground)
-stocks, with far fewer records for dead organic matter and soil carbon,
-and relatively few or no records for net annual increments or C fluxes
-into (gains) or out of (losses) the IPCC-defined C pools. Geographic
+stocks, with far fewer records for dead organic matter and soil C, and
+relatively few or no records for net annual increments or C fluxes into
+(gains) or out of (losses) the IPCC-defined C pools. Geographic
 representation is also quite uneven, with the highest densities of
@@
 Reflecting their strong influence on Earth's climate, forests play a
-central role in international plans for climate change mitigation under
-the Paris Agreement \citep{unfccc_adoption_2015}. Forest conservation,
-reforestation, and improved sustainable management all have significant
--- and relatively cost-effective -- potential as climate change
-mitigation options, with conservation and reforestation having the
-fourth and fifth largest net emission reduction potentials or all
-mitigation options \citep{ipcc_summary_2022}. As of 2016, forest-based
-mitigation accounted for 26\% of total planned greenhouse gas mitigation
-within Nationally Determined Contributions under the Paris Agreement
-\citep{grassi_key_2017}. Yet, envisioned forest-based climate change
-mitigation initiatives do not always correspond to actual emission
-reductions through on-the-ground implementation
+substantial role in international plans for climate change mitigation
+under the Paris Agreement \citep{unfccc_adoption_2015}. Forest
+conservation, reforestation, and improved sustainable management all
+have significant -- and relatively cost-effective -- potential as
+climate change mitigation options, with conservation and reforestation
+having the fourth and fifth largest net emission reduction potentials or
+all mitigation options \citep{ipcc_summary_2022}. As of 2016,
+forest-based mitigation accounted for 26\% of total planned greenhouse
+gas mitigation within Nationally Determined Contributions under the
+Paris Agreement \citep{grassi_key_2017}. Yet, envisioned forest-based
+climate change mitigation initiatives do not always correspond to actual
+emission reductions through on-the-ground implementation
 \citep[e.g.,][]{badgley_systematic_2022}. One critical need for ensuring
@@
 greenhouse gas (mainly CO\textsubscript{2}) exchanges between forest
-land and the atmosphere \citep{ipcc_agriculture_2006, ipcc_2019_2019}.
-This guidance has improved over the years as more of the relevant
+land and the atmosphere \citep{ipcc_agriculture_2006, ipcc_2019_2019}. A
+tiered approach to accounting is employed, where the lowest tier (Tier
+1) represents the simplest approach and relies on default parameter
+values -- for example, forest carbon (C) stocks values by ecozone
+\citep{fao_global_2012} and forest age class derived as the average of
+published estimates \citep{ipcc_2019_2019, rozendaal_aboveground_2022}.
+Tier 1 values have improved over the years as more of the relevant
 underlying data has become available
@@
 For example, the year following the release of the latest IPCC
-guidelines, \citet{cook-patton_mapping_2020} found that the latest
-default rates may underestimate rates of C accumulation in regrowth
-forests by 32\% on average and fail to capture eight-fold variation
-within ecozones. In addition, \citet{cuni-sanchez_high_2021} found that
-aboveground C stocks in mature African tropical montane forests were
-two-thirds higher than the IPCC default values for these forests. This
-rapid evolution of scientific information on the climate mitigation
-potential of forests is beneficial to climate mitigation efforts, but
-requires improved mechanisms for communicating the latest information
-from scientific researchers to the practitioners who need reliable
-estimates for greenhouse gas mitigation planning. Moreover, high
-variability of forest C cycling within ecozones
+guidelines, a more thorough analysis of C accumulation in regrowth
+forests found that IPCC's Tier 1 default values underestimated C
+sequestration by 32\% on average and failed to capture eight-fold
+variation within ecozones \citep{cook-patton_mapping_2020}. In addition,
+it was revealed that C stocks in mature African tropical montane forests
+were two-thirds higher than the IPCC Tier 1 values for these forests
+\citep{cuni-sanchez_high_2021}. This rapid evolution of scientific
+information on C cycling in forests is valuable for informing climate
+change mitigation efforts but requires improved mechanisms for
+communicating the latest information from scientific researchers to the
+practitioners who need reliable estimates for greenhouse gas mitigation
+planning. Moreover, high variability of forest C cycling within ecozones
 \citep[e.g.,][]{cook-patton_mapping_2020, cuni-sanchez_high_2021}
@@
 
-To improve the data accessible for C accounting, the IPCC created the
+To improve data accessibility for C accounting, the IPCC created the
 Emission Factor Database (EFDB;
@@
 sources for inclusion in the database. In the case of forests, emission
-factors include carbon stocks, net increments (``stock changes''), and
-fluxes (``gains'' and ``losses'') for various pools
+factors include C stocks, net increments (``stock changes''), and fluxes
+(``gains'' and ``losses'') for various pools
 \citep{ipcc_2006_2006, ipcc_2019_2019}.
@@
 (\url{https://forc-db.github.io/}), is the largest collection of
-published estimates of forest carbon stocks, increments, and annual
-fluxes
+published estimates of forest C stocks, increments, and annual fluxes
 \citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-For C currently contains 39860 records from 10590 plots in 1531 distinct
+For C currently contains 39855 records from 10589 plots in 1535 distinct
 geographical areas, along with records of stand age and disturbance
@@
 
-Here, we (1) review IPCC methods and definitions for tallying forest C;
-(2) describe mapping of ForC to IPCC's EFDB; (3) describe updates to
-ForC (ForC v4.0), most of which were implemented to facilitate data
-transfer to EFDB; (4) summarize the data in ForC that's relevant to EFDB
-and records that have been transferred to date; and (5) provide
+Here, we (1) review IPCC methods and definitions for forest C accounting
+in the context of typical forest C estimation methodologies; (2)
+describe mapping of ForC to IPCC's EFDB; (3) describe updates to ForC
+(ForC v4.0), most of which were implemented to facilitate data transfer
+to EFDB; (4) summarize the data in ForC that's relevant to EFDB and
+records that have been transferred to date; and (5) provide
 recommendations for improving data collection, analysis, database, and
@@
 yr\textsuperscript{-1}) are calculated as the sum of changes in various
-pools (section 2.1), plus any harvested wood products. For each pool,
-\(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
-takes the difference between gains and losses, or using the
-``Stock-Difference Method'', which computes \(\Delta C\) based on C
+pools (described in section 2.1), plus any harvested wood products. For
+each pool, \(\Delta C\) may be calculated using the ``Gain-Loss
+Method'', which takes the difference between gains and losses, or using
+the ``Stock-Difference Method'', which computes \(\Delta C\) based on C
 stocks at two points in time \citep{ipcc_2006_2006}. Thus, C cycle
@@
 no single standard for measuring or reporting that is adhered to by all
--- or even most -- studies. IPCC parses forest C pools into biomass
-(aboveground and belowground), dead organic matter (dead wood and
-litter), and soil organic matter (Table 1). While there is some
+-- or even most -- scientific studies. IPCC parses forest C pools into
+biomass (aboveground and belowground), dead organic matter (dead wood
+and litter), and soil organic matter (Table 1). While there is some
 flexibility around the components included in each pool, each national
-inventory must apply these in a consistent manner. In this section, we
-define and review the IPCC definitions in the context of typical forest
-C estimation methodologies.
+inventory must apply these in a consistent manner.
 
@@
 \hline
- &  & include non-dicot trees? & ?\\
+ &  & include non-dicot trees? & yes\\
 \hline
@@
 \hline
- &  & biomass allometry & ?\\
+ &  & biomass allometry & Tier 1 defaults draw on a variety of allometries\\
 \hline
@@
  &  & min root diameter & may exclude fine roots\\
-\hline
- &  & sampling depth & ?\\
 \hline
@@
 savannas, small trees and shrubs contribute a much larger proportion of
-C stocks and fluxes \citep{piponiot_distribution_2022, refs}, and,
-correspondingly, biomass estimates for these ecosystems tend include
-smaller size classes \citep[e.g.,][]{refs}. Beyond the minimum DBH
-sampled, forest censuses and biomass estimates also differ in their
-inclusion of life forms other than dicot trees -- including lianas,
-ferns, palms, and bamboo -- which in some places can reach large sizes
-and/or constitute a large fraction of forest C. \emph{{[}explain IPCC
-guidance{]}} Further, it is important to note that this excludes
-standing dead wood, which is included in remote sensing biomass
-estimates \citep{duncanson_aboveground_2021}.
+C stocks and fluxes \citep{piponiot_distribution_2022, refs_from_ForC},
+and, correspondingly, biomass estimates for these ecosystems tend
+include smaller size classes \citep[e.g.,][]{refs_from_ForC}. While IPCC
+guidance specifies that all living vegetation should be included in
+biomass estimates, forest censuses and biomass estimates do not
+consistently include life forms other than dicot trees (e.g., lianas,
+ferns, palms, bamboo), although thhese do tend to be censused when they
+consitute a large proportion of the biomass \citep{refs_from_ForC}.
+Further, it is important to note that the IPCC definition of aboveground
+biomass excludes standing dead wood, which is included in remote sensing
+biomass estimates \citep{duncanson_aboveground_2021}.
 
 A universal challenge in estimating biomass (living or dead) from forest
-census data is applying appropriate allometries to convert DBH
-measurements to biomass. Selection of allometries has an enormous
+census data is applying appropriate allometric models to convert DBH
+measurements to biomass. Selection of allometric models has an enormous
 influence on estimates of biomass stocks, increments, of fluxes
 \citep{clark_landscapescale_2000, clark_net_2001}. While trusted and
-standardized allometric equations are becoming increasingly available
+standardized allometric models are becoming increasingly available
 \citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022},
-large uncertainties remain. \emph{{[}explain IPCC guidance{]}}
+large uncertainties remain. IPCC Tier 1 values currently draw on studies
+applying a variety of allometric models
+\citep[e.g.,][]{requenasuarez_estimating_2019, rozendaal_aboveground_2022}.
 
@@
 same methodological sources of variation, but tending to be far more
-uncertain. Fine root biomass generally constitutes a much smaller C pool
-\citep[typically \textless5 t C
+uncertain \citep{ref}. Fine root biomass generally constitutes a much
+smaller C pool \citep[typically \textless5 t C
 ha\textsuperscript{-1},][]{anderson-teixeira_carbon_2021}, and IPCC
@@
 \citep{ipcc_2006_2006}, which can be a painstaking process. Field
-methods for estimating root biomass are highly variable. IPCC's default
-method for Tier 1 estimates is to apply a ratio of belowground to
-aboveground biomass, with default factors defined based on ecological
-zone, continent, and forest age \citep{ipcc_2006_2006, ipcc_2019_2019}.
+methods for estimating root biomass are highly variable \citep{ref}.
+IPCC's default method for Tier 1 estimates is to apply a ratio of
+belowground to aboveground biomass, with default factors defined based
+on ecological zone, continent, and forest age
+\citep{ipcc_2006_2006, ipcc_2019_2019}.
 
@@
 lags behind records transfer and had not yet been completed for all
-records sent as of April 29, 2022.
+records sent as of May 05, 2023.
 
@@
 
-As of April 29, 2022, ForC (v4.0) contained 32609 independent records
-(39860 total), 17204 of which were for variables relevant to EFDB (Fig.
+As of May 05, 2023, ForC (v4.0) contained 32693 independent records
+(39855 total), 17204 of which were for variables relevant to EFDB (Fig.
 1). These records were distributed across all forested continents and
@@
 
+\newpage
 \begin{figure}
@@
 relevant to EFDB (Table 2, Fig. 1). The records were very unevenly
-distributed across variables \ldots{}
+distributed across variables. The most records colected were for the
+Aboveground biomass variable(biomass\_ag).This variable
 
@@
 
-As of April 29, 2022, we had reviewed or added 2034 EFDB-relevant
-records, 1214 records of which were sent to EFDB, and 73 of which have
-been reviewed, accepted, and posted (Figs. 2-3, Table 2).
+As of May 05, 2023, we had reviewed or added 2034 EFDB-relevant records,
+1214 records of which were sent to EFDB, and 73 of which have been
+reviewed, accepted, and posted (Figs. 2-3, Table 2).
 \emph{{[}DETAILS{]}}
@@
 when applicable, is highly desirable and makes the data more relevant to
-IPCC.
+IPCC. Reports which had the most successful data transfers used EFDB
+variables and had clear tables showing their results.
 

### 2023-05-05 doc/manuscript/ESSDmanuscript.Rmd
check/ update through end of section 2

@@
 This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
-While dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance; @carmona_coarse_2002], aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
-In turn, they are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
+Dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance, @carmona_coarse_2002].
+However, aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
+In turn, dead wood pools are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
 
@@
 
-Soil organic matter/ carbon (SOM/ SOC), which *(statistic on how much C is typial)* [@ref], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
-Live fine roots (suggested diameter cutoff of 2 mm) may be included with soil organic matter when it is not feasible to distinguish them empirically.
+Soil organic matter/ carbon (SOM/ SOC), which is typically >100 t C and can exceed 300 t C in the top two meters of soil [@sanderman_soil_2017], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
+Live fine roots may be included with soil organic matter when it is not feasible to distinguish them empirically.
 The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.
@@
 The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
-This expansive definition of managed land implies that the majority of Forest Land in most countries is managed [e.g.,].
+This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
 However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].

### 2023-05-05 doc/manuscript/ESSDmanuscript.tex
check/ update through end of section 2

@@
 \hline
-aboveground biomass & all biomass of living vegetation & min dbh & may exclude understory if minor component\\
+aboveground biomass & all biomass of living vegetation & minimum size censused & may exclude understory if minor component\\
 \hline
@@
 \hline
- &  & biomass allometry & Tier 1 defaults draw on a variety of allometries\\
+ &  & biomass allometry & Tier 1 defaults draw on a variety of allometric models\\
 \hline
@@
 \hline
- &  & min root diameter & may exclude fine roots\\
-\hline
-dead wood & all non-living woody biomass above a specified diameter, aboveground or belowground & min diameter & 10 cm default, but may be chosen by country\\
+ &  & minimum root diameter & may exclude fine roots; suggested minimum diameter of 2 mm for fine roots\\
+\hline
+dead wood & all non-living woody biomass above a specified diameter, aboveground or belowground & minimum diameter & 10 cm default, but may be chosen by country\\
 \hline
@@
 \hline
-litter & all non-living biomass smaller than dead wood but larger than soil organic matter, in various states of decomposition both above or within the mineral or organic soil & max diameter (= min diameter for deadwood) & 10 cm default, but may be chosen by country\\
-\hline
- &  & min size (= size limit for soil organic matter) & ?\\
+litter & all non-living biomass smaller than dead wood but larger than soil organic matter, in various states of decomposition both above or within the mineral or organic soil & maximum diameter (= minimum diameter for deadwood) & 10 cm default, but may be chosen by country\\
+\hline
+ &  & minimum size (= size limit for soil organic matter) & suggested 2 mm\\
 \hline
@@
 fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a
-diameter specified by the country). While dead wood stocks and fluxes
-can be quite variable across forests
-\citep{anderson-teixeira_carbon_2021}, and can at times be the dominant
-pool in a forest ecosystem {[}e.g., following a severe natural
-disturbance; \citet{carmona_coarse_2002}{]}, aboveground dead wood
-remains relatively poorly characterized at a global scale
-\citep{anderson-teixeira_carbon_2021}, and belowground dead wood is
-rarely studied \citep{merganicova_dadwood_2012}. In turn, they are
-poorly characterized in large-scale forest C budgets
+diameter specified by the country). Dead wood stocks and fluxes can be
+quite variable across forests \citep{anderson-teixeira_carbon_2021}, and
+can at times be the dominant pool in a forest ecosystem \citep[e.g.,
+following a severe natural disturbance,][]{carmona_coarse_2002}.
+However, aboveground dead wood remains relatively poorly characterized
+at a global scale \citep{anderson-teixeira_carbon_2021}, and belowground
+dead wood is rarely studied \citep{merganicova_dadwood_2012}. In turn,
+dead wood pools are poorly characterized in large-scale forest C budgets
 \citep{pan_large_2011, harris_global_2021}, and IPCC's latest Tier 1
@@
 
-Soil organic matter/ carbon (SOM/ SOC), which \emph{(statistic on how
-much C is typial)} \citep{ref}, is defined by IPCC as ``organic carbon
-in mineral and organic soils (including peat) to a specified depth
-chosen by the country and applied consistently through the time series''
-\citep{ipcc_good_2003, ipcc_2006_2006}. Live fine roots (suggested
-diameter cutoff of 2 mm) may be included with soil organic matter when
-it is not feasible to distinguish them empirically. The greatest source
-of methodological variation in measuring SOM/ SOC is sampling depth,
-which has a suggested default of 30 cm but may vary by country provided
-that consistent criteria are applied.
+Soil organic matter/ carbon (SOM/ SOC), which is typically
+\textgreater100 t C and can exceed 300 t C in the top two meters of soil
+\citep{sanderman_soil_2017}, is defined by IPCC as ``organic carbon in
+mineral and organic soils (including peat) to a specified depth chosen
+by the country and applied consistently through the time series''
+\citep{ipcc_good_2003, ipcc_2006_2006}. Live fine roots may be included
+with soil organic matter when it is not feasible to distinguish them
+empirically. The greatest source of methodological variation in
+measuring SOM/ SOC is sampling depth, which has a suggested default of
+30 cm but may vary by country provided that consistent criteria are
+applied.
 
@@
 land implies that the majority of Forest Land in most countries is
-managed {[}e.g.,{]}. However, the definition is applied differently
-across countries, and the majority of governments have yet to report
-their approach for defining managed land or provide maps of managed land
+managed. However, the definition is applied differently across
+countries, and the majority of governments have yet to report their
+approach for defining managed land or provide maps of managed land
 \citep{ogle_delineating_2018, deng_comparing_2021}.

### 2023-05-05 doc/manuscript/ESSDmanuscript.Rmd
update through section 3.1

@@
 However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed ># records to obtain additional required information, and added `r n_records_added` new records.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
 
@@
 Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we *used an R script to automatically retrieve information based on the DOI*<!--needs info from Valentine here-->.
+For the latter, we *used an R script to automatically retrieve information based on the DOI ([issue 41](https://github.com/forc-db/IPCC-EFDB-integration/issues/41))*<!--needs info from Valentine here-->.
 

### 2023-05-05 doc/manuscript/ESSDmanuscript.tex
update through section 3.1

@@
 (Appendix A), defined 15 new variables, implemented enhanced quality
-control, manually reviewed \textgreater\# records to obtain additional
+control, manually reviewed \textgreater1705 records to obtain additional
 required information, and added 329 new records.
@@
 latter, we \emph{used an R script to automatically retrieve information
-based on the DOI}.
+based on the DOI
+(\href{https://github.com/forc-db/IPCC-EFDB-integration/issues/41}{issue
+41})}.
 

### 2023-05-06 doc/manuscript/ESSDmanuscript.Rmd
check/update through end of section 3.3

@@
 This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-At the time of writing, ForC contained ~`r n_ind_records_EFDB_variables` independent records that would be relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records that would be relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
 Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
@@
 ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
-For C currently contains `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
 As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
@@
 `r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
-This included eleven increment variables, adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
+The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
 These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
 Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
-Although most of these variables currently lack records in ForC, the structure exists such that records can be populated over time. 
+Although most of these variables lacked records in ForC as of `r format(Sys.time(), '%B %d, %Y')`, their addition gave the structure such that records can be populated over time. 
 Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).
@@
 Prior to releasing ForC v4.0, we executed several quality control measures.
-First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_using_inreview] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
-Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision *(Issue #29)[https://github.com/forc-db/ForC/issues/229]*.
-Third, to identify erroneous climate data... *(Issue #212)[https://github.com/forc-db/ForC/issues/212]*.
+First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_implementing_2022] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
+Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
+Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version **##** <!--need this from Becky--> [@climwin_ref] based on site coordinates. 
+Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or transfer to EFDB.
 
-Because ForC v4.0 contained known duplicate records, we used R scripts to remove likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
+Because ForC v4.0 contained known duplicate records, we used R scripts to identify likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
 Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
-All records sent to EFCB were ensured to be original through manual review, as detailed below.
+All records sent to EFDB were ensured to be independent and original through manual review, as detailed below.
 

### 2023-05-06 doc/manuscript/ESSDmanuscript.tex
check/update through end of section 3.3

@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). At the time of writing, ForC contained
+(v4.0; described here). As of May 06, 2023, ForC contained
 \textasciitilde17204 independent records that would be relevant to EFDB,
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-For C currently contains 39855 records from 10589 plots in 1535 distinct
-geographical areas, along with records of stand age and disturbance
-history. As such, ForC is positioned to improve forest C accounting
-through the transfer of data to EFDB. The purpose of this publication is
-to document that process and provide recommendations for future
-improvements.
+As of May 06, 2023, ForC contained 39855 records from 10589 plots in
+1535 distinct geographical areas, along with records of stand age and
+disturbance history. As such, ForC is positioned to improve forest C
+accounting through the transfer of data to EFDB. The purpose of this
+publication is to document that process and provide recommendations for
+future improvements.
 
@@
 units in C (ending in \texttt{\_C}) or organic matter (ending in
-\texttt{\_OM}) as one. This included eleven increment variables, adding
-to only one previously defined increment variable (aboveground biomass
-increment, \emph{delta.agb}). These are directly related to C stocks as
-previously defined in ForC, with ``\emph{delta.}'' added in front of the
-variable name. Further, we added variables capturing the belowground
-component of woody mortality (\emph{woody.mortality\_root}) and the
-combined aboveground and belowground components of woody mortality
-(\emph{woody.mortality}). Although most of these variables currently
-lack records in ForC, the structure exists such that records can be
+\texttt{\_OM}) as one. The majority of these were increment variables
+(n=11), adding to only one previously defined increment variable
+(aboveground biomass increment, \emph{delta.agb}). These are directly
+related to C stocks as previously defined in ForC, with
+``\emph{delta.}'' added in front of the variable name. Further, we added
+variables capturing the belowground component of woody mortality
+(\emph{woody.mortality\_root}) and the combined aboveground and
+belowground components of woody mortality (\emph{woody.mortality}).
+Although most of these variables lacked records in ForC as of May 06,
+2023, their addition gave the structure such that records can be
 populated over time. Finally, to provide better definition of the
@@
 measures. First, we implemented a system of continuous integration using
-GitHub Actions \citep[\emph{sensu}][]{kim_using_inreview} to run some
+GitHub Actions \citep[\emph{sensu}][]{kim_implementing_2022} to run some
 automatic checks any time the master data files are updated, including
@@
 coordinates, we created a field to record coordinate precision (Appendix
-A), and flagged and reviewed records with suspected low precision
-\emph{(Issue \#29){[}https://github.com/forc-db/ForC/issues/229{]}}.
-Third, to identify erroneous climate data\ldots{} \emph{(Issue
-\#212){[}https://github.com/forc-db/ForC/issues/212{]}}.
+A), and flagged and reviewed records with suspected low precision.
+Third, to identify erroneous climate data, we compared ForC climate
+values to those extracted from WorldClim version \textbf{\#\#}
+\citep{climwin_ref} based on site coordinates. Records deviating from
+WorldClim values by more variable-specific thresholds (\textgreater5°C
+for mean annual temperature, \textgreater7.5°C for mean temperatures of
+the warmest and coldest months, or \textgreater1 for log(mean annual
+precipitation in mm)) were flagged as requiring review prior to use in
+analysis or transfer to EFDB.
 
 Because ForC v4.0 contained known duplicate records, we used R scripts
-to remove likely duplicates, as detailed in
+to identify likely duplicates, as detailed in
 \citet{anderson-teixeira_carbon_2021}. Henceforth, we refer to the set
 of records with likely duplicates removed as ``independent records''.
-All records sent to EFCB were ensured to be original through manual
-review, as detailed below.
+All records sent to EFDB were ensured to be independent and original
+through manual review, as detailed below.
 
@@
 lags behind records transfer and had not yet been completed for all
-records sent as of May 05, 2023.
+records sent as of May 06, 2023.
 
@@
 
-As of May 05, 2023, ForC (v4.0) contained 32693 independent records
+As of May 06, 2023, ForC (v4.0) contained 32693 independent records
 (39855 total), 17204 of which were for variables relevant to EFDB (Fig.
@@
 
-As of May 05, 2023, we had reviewed or added 2034 EFDB-relevant records,
+As of May 06, 2023, we had reviewed or added 2034 EFDB-relevant records,
 1214 records of which were sent to EFDB, and 73 of which have been

### 2023-05-06 doc/manuscript/ESSDmanuscript.Rmd
check/edit through end of section 3.5

@@
 
-EFDB data submissions required information that was not recorded in previous versions of ForC, but rather was recorded in fields (`r table_ForCchanges`).
+EFDB data submissions required information that was not recorded in previous versions of ForC, but for which new fields were created for EFDB compatibility (`r table_ForCchanges`).
 It was therefore necessary to return to original publications to retrieve relevant information, including 
-(1) whether records of interest were presented in tables / text or digitized from figures (EFDB will not accept digitized data, whereas ForC does), 
-(2) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, ForC would accept the sum of these as a valid estimate of total biomass, whereas EFDB would not)
-(3) estimates in original units, 
-and (4) confidence intervals (when not already in ForC).
+(1) estimates in original units, 
+(2) confidence intervals (when not already in ForC),
+(3) whether records of interest were presented in tables or text or digitized from figures (EFDB will not accept digitized data), 
+(4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass)
 We also checked that existing ForC records were complete and correct.
@@
 and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@refs], ground-based data on tropical forest C cycling tend to be more scarce due to a variety of challenges [@delima_making_2022], and *tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on previously existing data [@ref].* <!-- I'm almost certain this is true, but need to confirm/ find ref.-->
+The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@refs], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@refs; @delima_making_2022], and *tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on previously existing data [@ref].* <!-- I'm almost certain this is true, but need to confirm/ find ref.-->
 
@@
 In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
-These included **##** records from **##** studies [@piponiot_distribution_2022; @lutz_largediameter_2021; @refs<!--others?-->] that were not previously included in ForC.
-In addition, we created new records for **##** EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
+These included 104 records<!--100 from Piponiot, 4 from Lutz--> from two studies [@piponiot_distribution_2022; @lutz_largediameter_2021] that were not previously included in ForC.
+In addition, we created new records for `r n_records_added-104` EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
 

### 2023-05-06 doc/manuscript/ESSDmanuscript.tex
check/edit through end of section 3.5

@@
 EFDB data submissions required information that was not recorded in
-previous versions of ForC, but rather was recorded in fields (Appendix
-A). It was therefore necessary to return to original publications to
-retrieve relevant information, including (1) whether records of interest
-were presented in tables / text or digitized from figures (EFDB will not
-accept digitized data, whereas ForC does), (2) whether records of
-interest were presented directly, as opposed to having been calculated
-from related variables (for example, if a study presents aboveground
-biomass and root biomass but not total biomass, ForC would accept the
-sum of these as a valid estimate of total biomass, whereas EFDB would
-not) (3) estimates in original units, and (4) confidence intervals (when
-not already in ForC). We also checked that existing ForC records were
-complete and correct.
+previous versions of ForC, but for which new fields were created for
+EFDB compatibility (Appendix A). It was therefore necessary to return to
+original publications to retrieve relevant information, including (1)
+estimates in original units, (2) confidence intervals (when not already
+in ForC), (3) whether records of interest were presented in tables or
+text or digitized from figures (EFDB will not accept digitized data),
+(4) whether records of interest were presented directly, as opposed to
+having been calculated from related variables (for example, if a study
+presents aboveground biomass and root biomass but not total biomass,
+EFDB would not accept the sum of these as a valid record of total
+biomass) We also checked that existing ForC records were complete and
+correct.
 
@@
 for climate change mitigation \citep{refs}, ground-based data on
-tropical forest C cycling tend to be more scarce due to a variety of
-challenges \citep{delima_making_2022}, and \emph{tropical countries are
-more likely to apply Tier 1 methodology that bases forest C budgets on
-previously existing data \citep{ref}.}
+tropical forest C cycling tends to be more scarce due to a variety of
+challenges \citep{refs, delima_making_2022}, and \emph{tropical
+countries are more likely to apply Tier 1 methodology that bases forest
+C budgets on previously existing data \citep{ref}.}
 
@@
 In addition to reviewing existing records, we added a total of 329 new
-records to ForC. These included \textbf{\#\#} records from \textbf{\#\#}
-studies
-\citep{piponiot_distribution_2022, lutz_largediameter_2021, refs} that
-were not previously included in ForC. In addition, we created new
-records for \textbf{\#\#} EFDB-relevant estimates presented in the
-original publication that were not yet present in ForC.
+records to ForC. These included 104 records from two studies
+\citep{piponiot_distribution_2022, lutz_largediameter_2021} that were
+not previously included in ForC. In addition, we created new records for
+225 EFDB-relevant estimates presented in the original publication that
+were not yet present in ForC.
 

### 2023-05-07 doc/manuscript/ESSDmanuscript.Rmd
update through end of section 5.1

@@
 C_variables <- read.csv(paste0(path_to_dir,"/doc/manuscript/figures_tables/C_variables.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+n_relevant_variables = nrow(C_variables) - 7 # n variables relevant = rows in C_variables - headers and total
+n_relevant_variables_with_no_data =  sum(C_variables$`n in ForC` == 0,  na.rm = TRUE)
+n_relevant_variables_with_data= n_relevant_variables - n_relevant_variables_with_no_data
+
 n_ind_records_EFDB_variables= C_variables[length(C_variables[,"variable"]), "n independent records in ForC"] #n independent records relevant to EFDB
+n_ind_records_agb <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag"]
+n_ind_records_agb_components <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag_foliage"]
+n_ind_records_biomass <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass"]
+n_ind_records_rootbiomass <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root_coarse"]
+n_ind_records_deadwood <- C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_down"]+C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_standing"]
+n_ind_records_OL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "litter"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "O.horizon"]
+n_ind_records_SOC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "SOM / SOC"] 
+n_ind_records_deltaagb <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.agb"] 
+n_ind_records_deltaOL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.O.horizon"] 
+flux_variables_with_records <- c("NPP_woody", "ANPP_woody", "ANPP_woody_ag", "BNPP_root_fine","BNPP_root.turnover_fine","BNPP_root_coarse" )
+n_ind_records_flux <- sum(C_variables$`n independent records in ForC`[C_variables$`variable` %in% flux_variables_with_records])
 
@@
 To transfer complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data transfer were that (1) records had been checked against the original study and determined to be complete and correct (`EFDB.ready` field in `ForC_citations` table), (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables (`data.location.within.source` field in `ForC_measurements` table), and (3) the records had not previously been sent to EFDB.
+Criteria for data transfer were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been sent to EFDB.
 Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for inclusion in EFDB. 
@@
 
-The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`, with details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
-
+The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`.
 For the majority of fields, contents of the field in ForC was transferred directly into an EFDB field, either as the only contents of that field or as part of a composite record.
-For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field `Region/Regional conditions` (`r table_ForCfieldmapping`).
-In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field `Comments from Data Provider`).
+For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).
+In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*).
 For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
-For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (`Gases` field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
+For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
 There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) land classification (section `r section_land_use_mapping`).
@@
 `r fig_variable_mapping`).
-For each C pools (Table `r table_pools`), we identified variables representing organic matter or C stocks, stock changes (a.k.a.
+For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, stock changes (a.k.a.
 "net annual increments" by IPCC, "increments" in ForC), gains (a.k.a.
@@
 Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
-Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the IPCC-EFDB-integration repository ([https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
+Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
 
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for variables relevant to EFDB (Fig.
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the the `r n_relevant_variables` variables relevant to EFDB (Fig.
 `r fig_variable_mapping`).
-These records were distributed across all forested continents and ecozones, with particularly high concentrations in *[ecozone(s)]* and low concentrations in *[ecozone(s)]* (Fig. `r fig_map`).
-The most widely represented forest type was *[type]*, followed by *[type]* and *[type]* (Fig. `r fig_histograms`).
+These records were distributed across all forested continents and ecozones, albeit unevenly (Fig. `r fig_map`).
+The largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig. `r fig_histograms`c).
+Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
+`r fig_histograms`b).
+Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
+The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
+In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).
 
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
 Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems."}
@@
 \newpage
-```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of independent records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
 For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
@@
 
-ForC contained data for **##** of the **##** variables relevant to EFDB (Table `r table_variables`, Fig. `r fig_variable_mapping`).
-The records were very unevenly distributed across variables. The most records colected were for the Aboveground biomass variable(biomass_ag).This variable
+ForC contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables relevant to EFDB (Table `r table_variables`, Fig. `r fig_variable_mapping`).
+The records were very unevenly distributed across variables. 
+The variable with most records was aboveground biomass, representing `r round(n_ind_records_agb/n_ind_records_EFDB_variables,2)*100`% of all independent records relevant to EFDB, and aboveground biomass components (woody biomass or foliage) representing an additional `r round(n_ind_records_agb_components/n_ind_records_EFDB_variables,2)*100`%.
+A total of `r round(n_ind_records_rootbiomass/n_ind_records_EFDB_variables,2)*100`% of relevant records were for root biomass (including fine and coarse root components), while `r round(n_ind_records_biomass/n_ind_records_EFDB_variables,2)*100`% described total biomass.
+The non-living pools were less represented, with `r round(n_ind_records_deadwood/n_ind_records_EFDB_variables,2)*100`% of relevant were for dead wood (including standing and fallen components), `r round(n_ind_records_OL/n_ind_records_EFDB_variables*100,1)`% for litter, and `r round(n_ind_records_SOC/n_ind_records_EFDB_variables*100,1)`% for soil carbon.
+
+Increment and flux variables were poorly represented (Table `r table_variables`).
+The increment variable with most records was the aboveground biomass increment, representing `r round(n_ind_records_deltaagb/n_ind_records_EFDB_variables*100,1)`% of all independent records relevant to EFDB.
+The only other relevant increment variable with any records was the O horizon (litter) increment, with just `r n_ind_records_deltaOL` records.
+Relevant flux variable records (n=`r n_ind_records_flux`) were limited to the biomass pools (aboveground, belowground, or total) and together constituted `r round(n_ind_records_flux/n_ind_records_EFDB_variables,2)*100`% of ForC's independent records relevant to EFDB.
 
@@
 
-Table B1: **Mapping of ForC fields to EFDB.** See footnotes at end of table (STILL NEED TO BE PROPERLY INSERTED).
+Table B1: **Mapping of ForC fields to EFDB.** Details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
+See footnotes at end of table (STILL NEED TO BE PROPERLY INSERTED).
 ```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}

### 2023-05-07 doc/manuscript/ESSDmanuscript.tex
update through end of section 5.1

@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 06, 2023, ForC contained
+(v4.0; described here). As of May 07, 2023, ForC contained
 \textasciitilde17204 independent records that would be relevant to EFDB,
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 06, 2023, ForC contained 39855 records from 10589 plots in
+As of May 07, 2023, ForC contained 39855 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 belowground components of woody mortality (\emph{woody.mortality}).
-Although most of these variables lacked records in ForC as of May 06,
+Although most of these variables lacked records in ForC as of May 07,
 2023, their addition gave the structure such that records can be
@@
 records had been checked against the original study and determined to be
-complete and correct (\texttt{EFDB.ready} field in
-\texttt{ForC\_citations} table), (2) the original study presented values
-in tables or text, as opposed to the values having been digitized from
-graphs or calculated based on related variables
-(\texttt{data.location.within.source} field in
-\texttt{ForC\_measurements} table), and (3) the records had not
-previously been sent to EFDB. Once converted into EFDB format, the
-records were reviewed and then sent to the IPCC's Technical Support Unit
-for inclusion in EFDB. Complete records needed to be reviewed by the
-EFDB editorial board prior to posting in the database -- a process that
-lags behind records transfer and had not yet been completed for all
-records sent as of May 06, 2023.
+complete and correct, and as originally presented, (2) the original
+study presented values in tables or text, as opposed to the values
+having been digitized from graphs or calculated based on related
+variables, and (3) the records had not previously been sent to EFDB.
+Once converted into EFDB format, the records were reviewed and then sent
+to the IPCC's Technical Support Unit for inclusion in EFDB. Complete
+records needed to be reviewed by the EFDB editorial board prior to
+posting in the database -- a process that lags behind records transfer
+and had not yet been completed for all records sent as of May 07, 2023.
 
@@
 
-The mapping of ForC fields into EFDB fields is summarized in Appendix B,
-with details documented in the public GitHub repository associated with
-the project, IPCC-EFDB-integration repository within the ForC-db
-organization (file \emph{ForC-EFDB\_mapping.csv} available at
-\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).
-
+The mapping of ForC fields into EFDB fields is summarized in Appendix B.
 For the majority of fields, contents of the field in ForC was
@@
 fields describing site location, climate, and edaphic properties all
-mapped into the EFDB field \texttt{Region/Regional\ conditions}
-(Appendix B). In cases where original studies did not present 95\%
-confidence intervals (required by IPCC when available) but did present
-information required to calculate these (standard error or n and
-standard deviation), we calculated the 95\% confidence intervals and
-populated the EFDB field with this information (noting the calculation
-in the EFDB field \texttt{Comments\ from\ Data\ Provider}). For some
-fields, simple conditional logic was used to populate EFDB fields based
-on ForC records. For example, for stock variables presented in the
-original publication in units of dry organic matter mass (as opposed to
-C), several greenhouse gasses (CO\textsubscript{2}, CO,
-CH\textsubscript{4}, NO, NO\textsubscript{2}, N\textsubscript{2}O) were
-entered in the EFDB field indicating the greenhouse gases to which the
-record could be pertinent (\texttt{Gases} field) because these values
-could be used in calculations of greenhouse gas emissions from biomass
-burning \citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse
-gas would be CO\textsubscript{2}. There were two cases in which more
-complex mapping was required: (1) mapping of C cycle variables (section
-4.1.1) and (2) land classification (section 4.1.2).
+mapped into the EFDB field \emph{Region/Regional conditions} (Appendix
+B). In cases where original studies did not present 95\% confidence
+intervals (required by IPCC when available) but did present information
+required to calculate these (standard error or n and standard
+deviation), we calculated the 95\% confidence intervals and populated
+the EFDB field with this information (noting the calculation in the EFDB
+field \emph{Comments from Data Provider}). For some fields, simple
+conditional logic was used to populate EFDB fields based on ForC
+records. For example, for stock variables presented in the original
+publication in units of dry organic matter mass (as opposed to C),
+several greenhouse gasses (CO\textsubscript{2}, CO, CH\textsubscript{4},
+NO, NO\textsubscript{2}, N\textsubscript{2}O) were entered in the EFDB
+field indicating the greenhouse gases to which the record could be
+pertinent (\emph{Gases} field) because these values could be used in
+calculations of greenhouse gas emissions from biomass burning
+\citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse gas
+would be CO\textsubscript{2}. There were two cases in which more complex
+mapping was required: (1) mapping of C cycle variables (section 4.1.1)
+and (2) land classification (section 4.1.2).
 
@@
 of ForC variables to identify those that were relevant to EFDB and to
-appropriately map them into EFDB (Fig. 1). For each C pools (Table 1),
-we identified variables representing organic matter or C stocks, stock
+appropriately map them into EFDB (Fig. 1). For each C pool (Table 1), we
+identified variables representing organic matter or C stocks, stock
 changes (a.k.a. ``net annual increments'' by IPCC, ``increments'' in
@@
 \citep{ipcc_2006_2006} -- are documented in the file
-ForC\_variables\_mapping.csv in the IPCC-EFDB-integration repository
-(\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).
+ForC\_variables\_mapping.csv in the GitHub repository associated with
+this publication IPCC-EFDB-integration repository in ForC-db
+organization (\url{https://github.com/forc-db/IPCC-EFDB-integration}).
 
@@
 
-As of May 06, 2023, ForC (v4.0) contained 32693 independent records
-(39855 total), 17204 of which were for variables relevant to EFDB (Fig.
-1). These records were distributed across all forested continents and
-ecozones, with particularly high concentrations in
-\emph{{[}ecozone(s){]}} and low concentrations in
-\emph{{[}ecozone(s){]}} (Fig. 2). The most widely represented forest
-type was \emph{{[}type{]}}, followed by \emph{{[}type{]}} and
-\emph{{[}type{]}} (Fig. 3).
+As of May 07, 2023, ForC (v4.0) contained 32693 independent records
+(39855 total), 17204 of which were for the the 36 variables relevant to
+EFDB (Fig. 1). These records were distributed across all forested
+continents and ecozones (Fig. 2). The largest number of records came
+from Asia, followed by North America, South America, and Europe, with
+relatively few records from Africa, Australia, and Oceania (Fig. 3c).
+Categorized by FAO ecozone, the greatest numbers of records came from
+subtropical humid forests, temperate mountain systems, and tropical rain
+forests, each with \textgreater2,000 independent records (Fig. 3b).
+Boreal coniferous forests, temperate continental forests, subtropical
+mountain systems, and tropical moist deciduous forests had
+\textgreater1,000 independent records each, while other ecozones all had
+\textless1,000 records. The most widely represented forest type was
+needleleaf evergreen, followed by broadleaf deciduous and broadleaf
+evergreen (Fig. 3a). In terms of stand age, the most represented age
+class was 20-100 years, followed by \textless20 years and then
+\textgreater100 years (Fig. 3d).
 
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
 Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.}\label{fig:fig_map}
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of independent records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
 For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_histograms}
@@
 
-ForC contained data for \textbf{\#\#} of the \textbf{\#\#} variables
-relevant to EFDB (Table 2, Fig. 1). The records were very unevenly
-distributed across variables. The most records colected were for the
-Aboveground biomass variable(biomass\_ag).This variable
+ForC contained records for 22 of the 36 variables relevant to EFDB
+(Table 2, Fig. 1). The records were very unevenly distributed across
+variables. The variable with most records was aboveground biomass,
+representing 45\% of all independent records relevant to EFDB, and
+aboveground biomass components (woody biomass or foliage) representing
+an additional 5\%. A total of 30\% of relevant records were for root
+biomass (including fine and coarse root components), while 5\% described
+total biomass. The non-living pools were less represented, with 5\% of
+relevant were for dead wood (including standing and fallen components),
+0.4\% for litter, and 2.3\% for soil carbon.
+
+Increment and flux variables were poorly represented (Table 2). The
+increment variable with most records was the aboveground biomass
+increment, representing 0.7\% of all independent records relevant to
+EFDB. The only other relevant increment variable with any records was
+the O horizon (litter) increment, with just 4 records. Relevant flux
+variable records (n=974) were limited to the biomass pools (aboveground,
+belowground, or total) and together constituted 6\% of ForC's
+independent records relevant to EFDB.
 
@@
 
-As of May 06, 2023, we had reviewed or added 2034 EFDB-relevant records,
+As of May 07, 2023, we had reviewed or added 2034 EFDB-relevant records,
 1214 records of which were sent to EFDB, and 73 of which have been
@@
 
-Table B1: \textbf{Mapping of ForC fields to EFDB.} See footnotes at end
-of table (STILL NEED TO BE PROPERLY INSERTED).
+Table B1: \textbf{Mapping of ForC fields to EFDB.} Details documented in
+the public GitHub repository associated with the project,
+IPCC-EFDB-integration repository within the ForC-db organization (file
+\emph{ForC-EFDB\_mapping.csv} available at
+\url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).
+See footnotes at end of table (STILL NEED TO BE PROPERLY INSERTED).
 \begingroup\fontsize{8}{10}\selectfont

### 2023-05-08 doc/manuscript/ESSDmanuscript.Rmd
worked on section 5.2 (not complete)
requires updating after we solve issues #42 and #43.

@@
 
-n_records_posted_EFDB = 73 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
+n_records_posted_EFDB = 376 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
 Then get number from upper left box where it says "displayed records")
@@
 `r fig_map`-`r fig_histograms`, Table `r table_variables`).
+The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
+The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review the records, and it is not expected that many -- if any -- records will be rejected.
+
+<!-- this paragraph will need to be revisited after we solve issues 42 and 43-->
+The records transferred from ForC to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
+However, the density of these records was very unevenly distributed across continents, biomes, and forest types and was not proportional to the numbers of relevant records in ForC (Fig.
+`r fig_histograms`).
+Rather, the largest number of records came from North America, followed by Asia and South America (Fig.
+`r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.
+`r fig_histograms`b).
+In terms of dominant vegetation, by far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
+The largest records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
+
+In terms of variables transferred, ...
 *[DETAILS]*

### 2023-05-08 doc/manuscript/ESSDmanuscript.tex
worked on section 5.2 (not complete)
requires updating after we solve issues #42 and #43.

@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 07, 2023, ForC contained
+(v4.0; described here). As of May 08, 2023, ForC contained
 \textasciitilde17204 independent records that would be relevant to EFDB,
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 07, 2023, ForC contained 39855 records from 10589 plots in
+As of May 08, 2023, ForC contained 39855 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 belowground components of woody mortality (\emph{woody.mortality}).
-Although most of these variables lacked records in ForC as of May 07,
+Although most of these variables lacked records in ForC as of May 08,
 2023, their addition gave the structure such that records can be
@@
 posting in the database -- a process that lags behind records transfer
-and had not yet been completed for all records sent as of May 07, 2023.
+and had not yet been completed for all records sent as of May 08, 2023.
 
@@
 
-As of May 07, 2023, ForC (v4.0) contained 32693 independent records
+As of May 08, 2023, ForC (v4.0) contained 32693 independent records
 (39855 total), 17204 of which were for the the 36 variables relevant to
 EFDB (Fig. 1). These records were distributed across all forested
-continents and ecozones (Fig. 2). The largest number of records came
-from Asia, followed by North America, South America, and Europe, with
-relatively few records from Africa, Australia, and Oceania (Fig. 3c).
-Categorized by FAO ecozone, the greatest numbers of records came from
-subtropical humid forests, temperate mountain systems, and tropical rain
-forests, each with \textgreater2,000 independent records (Fig. 3b).
-Boreal coniferous forests, temperate continental forests, subtropical
-mountain systems, and tropical moist deciduous forests had
+continents and ecozones, albeit unevenly (Fig. 2). The largest number of
+records came from Asia, followed by North America, South America, and
+Europe, with relatively few records from Africa, Australia, and Oceania
+(Fig. 3c). Categorized by FAO ecozone, the greatest numbers of records
+came from subtropical humid forests, temperate mountain systems, and
+tropical rain forests, each with \textgreater2,000 independent records
+(Fig. 3b). Boreal coniferous forests, temperate continental forests,
+subtropical mountain systems, and tropical moist deciduous forests had
 \textgreater1,000 independent records each, while other ecozones all had
@@
 
-As of May 07, 2023, we had reviewed or added 2034 EFDB-relevant records,
-1214 records of which were sent to EFDB, and 73 of which have been
-reviewed, accepted, and posted (Figs. 2-3, Table 2).
-\emph{{[}DETAILS{]}}
+As of May 08, 2023, we had reviewed or added 2034 EFDB-relevant records,
+1214 records of which were sent to EFDB, and 376 of which have been
+reviewed, accepted, and posted (Figs. 2-3, Table 2). The 40\%
+attenuation between records reviewed and those sent to EFDB was
+attributable to the presence of digitized records and records where a
+variable's value had been calculated as the sum or difference of related
+variables rather than presented directly in the text. The discrepancy
+between the number of records sent and that posted to EFDB is primarily
+attributable to the time required for the IPCC to review the records,
+and it is not expected that many -- if any -- records will be rejected.
+
+The records transferred from ForC to EFDB were broadly distributed
+across Earth's forests (Fig. 2). However, the density of these records
+was very unevenly distributed across continents, biomes, and forest
+types and was not proportional to the numbers of relevant records in
+ForC (Fig. 3). Rather, the largest number of records came from North
+America, followed by Asia and South America (Fig. 3c), with the most
+represented FAO ecozones being boreal coniferous forest, temperate
+continental forest, and temperate mountain systems, followed by tropical
+rain forests and moist deciduous forests (Fig. 3b).\\
+In terms of dominant vegetation, by far the most records came from
+needleleaf evergreen forests, followed by broadleaf deciduous and
+broadleaf evergreen (Fig. 3b). The largest records came from mature
+forests (\textgreater100 years), followed by young and intermediate-aged
+stands (Fig. 3d).
+
+In terms of variables transferred, \ldots{} \emph{{[}DETAILS{]}}
 

### 2023-05-08 doc/manuscript/ESSDmanuscript.Rmd
provide info on web scraping Citation #41
@teixeirak, I was not able to knit because the .rmd is looking for some file(s) in your machine. Let me know if what I edited provides enough detail.

@@
 Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we *used an R script to automatically retrieve information based on the DOI ([issue 41](https://github.com/forc-db/IPCC-EFDB-integration/issues/41))*<!--needs info from Valentine here-->.
+For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[ @rvest]).
+That information was manually retrieved when the web scraping failed.
 

### 2023-05-08 doc/manuscript/ESSDmanuscript.Rmd
update C_variable_mapping figure and caption
#43

@@
 ```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
-Most, but not all, EFDB-relevant ForC variables are shown here.
-Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon."}
+Most, but not all, EFDB-relevant ForC variables are shown here (excluded: C concentration in tree wood, annual branch turnover, understory biomass, total ecosystem C in all pools excluding soils and net annual increment thereof).
+Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon."}
 #this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
@@
 
-Moreover, there are many published EFDB-relevant forest C data that are not included in ForC, with more being published on a nearly daily basis. 
+Moreover, there are innumerous published EFDB-relevant forest C data that are not included in ForC, with more being published on a nearly daily basis. 
 Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
 *(There are some efforts underway, including a few that Susan can specify.)*
-Such reviews are necessary to even develop a rigorous assessment of forest C data that are available, versus those that require additional data collection and analysis.
+Such reviews are necessary to assess the availability of published EFDB-relevant forest C records, versus those that require additional data collection and analysis.
 

### 2023-05-08 doc/manuscript/ESSDmanuscript.tex
update C_variable_mapping figure and caption
#43

@@
 (v4.0; described here). As of May 08, 2023, ForC contained
-\textasciitilde17204 independent records that would be relevant to EFDB,
-1214 of which have been submitted to date. Among the data in ForC, there
+\textasciitilde17293 independent records that would be relevant to EFDB,
+1321 of which have been submitted to date. Among the data in ForC, there
 is disproportionate representation of biomass (particularly aboveground)
@@
 (Appendix A), defined 15 new variables, implemented enhanced quality
-control, manually reviewed \textgreater1705 records to obtain additional
+control, manually reviewed \textgreater1798 records to obtain additional
 required information, and added 329 new records.
@@
 and expanding the information recorded in the citations table. For the
-latter, we \emph{used an R script to automatically retrieve information
-based on the DOI
-(\href{https://github.com/forc-db/IPCC-EFDB-integration/issues/41}{issue
-41})}.
+latter, we used an R script to automatically harvest (scrape) the URL,
+citation, abstract and language of the publications, based on their DOI,
+using R package \texttt{rvest}\citep{rvest}). That information was
+manually retrieved when the web scraping failed.
 
@@
 \includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
-Most, but not all, EFDB-relevant ForC variables are shown here.
-Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., min DBH).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below census min DBH, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.}\label{fig:fig_variable_mapping}
+Most, but not all, EFDB-relevant ForC variables are shown here (excluded: C concentration in tree wood, annual branch turnover, understory biomass, total ecosystem C in all pools excluding soils and net annual increment thereof).
+Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.}\label{fig:fig_variable_mapping}
 \end{figure}
@@
 As of May 08, 2023, ForC (v4.0) contained 32693 independent records
-(39855 total), 17204 of which were for the the 36 variables relevant to
+(39855 total), 17293 of which were for the the 36 variables relevant to
 EFDB (Fig. 1). These records were distributed across all forested
@@
 increment variable with most records was the aboveground biomass
-increment, representing 0.7\% of all independent records relevant to
+increment, representing 0.9\% of all independent records relevant to
 EFDB. The only other relevant increment variable with any records was
 the O horizon (litter) increment, with just 4 records. Relevant flux
-variable records (n=974) were limited to the biomass pools (aboveground,
+variable records (n=972) were limited to the biomass pools (aboveground,
 belowground, or total) and together constituted 6\% of ForC's
@@
 \hline
-biomass & 1095 & 847 & 93 & 48\\
+biomass & 1094 & 850 & 95 & 50\\
 \hline
@@
 \hline
-biomass\_ag & 9050 & 7737 & 1251 & 693\\
+biomass\_ag & 9050 & 7750 & 1267 & 723\\
 \hline
@@
 \hline
-biomass\_ag\_foliage & 601 & 502 & 49 & 27\\
-\hline
-delta.agb & 166 & 128 & 123 & 123\\
+biomass\_ag\_foliage & 601 & 520 & 73 & 45\\
+\hline
+delta.agb & 166 & 150 & 145 & 123\\
 \hline
@@
 \hline
-woody.mortality\_ag & 112 & 62 & 30 & 17\\
+woody.mortality\_ag & 112 & 75 & 47 & 50\\
 \hline
@@
 \hline
-biomass\_root & 4629 & 4180 & 123 & 55\\
-\hline
-biomass\_root\_fine & 931 & 594 & 18 & 18\\
-\hline
-biomass\_root\_coarse & 599 & 410 & 12 & 7\\
+biomass\_root & 4629 & 4185 & 125 & 57\\
+\hline
+biomass\_root\_fine & 930 & 595 & 18 & 18\\
+\hline
+biomass\_root\_coarse & 599 & 413 & 12 & 7\\
 \hline
@@
 \hline
-BNPP\_root\_fine & 489 & 333 & 0 & 0\\
+BNPP\_root\_fine & 488 & 331 & 0 & 0\\
 \hline
@@
 \hline
-deadwood & 437 & 303 & 103 & 61\\
-\hline
-deadwood\_standing & 152 & 120 & 17 & 17\\
-\hline
-deadwood\_down & 424 & 368 & 51 & 27\\
+deadwood & 438 & 304 & 104 & 70\\
+\hline
+deadwood\_standing & 153 & 121 & 18 & 17\\
+\hline
+deadwood\_down & 425 & 369 & 52 & 28\\
 \hline
@@
 \hline
-O.horizon & 38 & 38 & 38 & 38\\
+O.horizon & 45 & 45 & 45 & 40\\
 \hline
@@
 \hline
-SOM / SOC & 693 & 398 & 89 & 46\\
+SOM / SOC & 693 & 401 & 89 & 56\\
 \hline
@@
 \hline
-\textbf{TOTAL} & \textbf{20932} & \textbf{17204} & \textbf{2034} & \textbf{1214}\\
+\textbf{TOTAL} & \textbf{20939} & \textbf{17293} & \textbf{2127} & \textbf{1321}\\
 \hline
@@
 
-As of May 08, 2023, we had reviewed or added 2034 EFDB-relevant records,
-1214 records of which were sent to EFDB, and 376 of which have been
-reviewed, accepted, and posted (Figs. 2-3, Table 2). The 40\%
+As of May 08, 2023, we had reviewed or added 2127 EFDB-relevant records,
+1321 records of which were sent to EFDB, and 376 of which have been
+reviewed, accepted, and posted (Figs. 2-3, Table 2). The 38\%
 attenuation between records reviewed and those sent to EFDB was
@@
 continental forest, and temperate mountain systems, followed by tropical
-rain forests and moist deciduous forests (Fig. 3b).\\
-In terms of dominant vegetation, by far the most records came from
-needleleaf evergreen forests, followed by broadleaf deciduous and
-broadleaf evergreen (Fig. 3b). The largest records came from mature
-forests (\textgreater100 years), followed by young and intermediate-aged
-stands (Fig. 3d).
+rain forests and moist deciduous forests (Fig. 3b). In terms of dominant
+vegetation, by far the most records came from needleleaf evergreen
+forests, followed by broadleaf deciduous and broadleaf evergreen (Fig.
+3b). The largest records came from mature forests (\textgreater100
+years), followed by young and intermediate-aged stands (Fig. 3d).
 
@@
 the process of reviewing and sending data that are already in ForC
-(Figs. 2-3). So far, only \textasciitilde7\% of the EFDB-relevant data
+(Figs. 2-3). So far, only \textasciitilde8\% of the EFDB-relevant data
 in ForC have been sent to EFDB.
 
-Moreover, there are many published EFDB-relevant forest C data that are
-not included in ForC, with more being published on a nearly daily basis.
-Coverage of particular variables or regions could be vastly improved
-through systematic review of the literature. \emph{(There are some
-efforts underway, including a few that Susan can specify.)} Such reviews
-are necessary to even develop a rigorous assessment of forest C data
-that are available, versus those that require additional data collection
-and analysis.
+Moreover, there are innumerous published EFDB-relevant forest C data
+that are not included in ForC, with more being published on a nearly
+daily basis. Coverage of particular variables or regions could be vastly
+improved through systematic review of the literature. \emph{(There are
+some efforts underway, including a few that Susan can specify.)} Such
+reviews are necessary to assess the availability of published
+EFDB-relevant forest C records, versus those that require additional
+data collection and analysis.
 

### 2023-05-08 doc/manuscript/ESSDmanuscript.Rmd
small updates

@@
 C_variables <- read.csv(paste0(path_to_dir,"/doc/manuscript/figures_tables/C_variables.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_relevant_variables = nrow(C_variables) - 7 # n variables relevant = rows in C_variables - headers and total
+n_relevant_variables = nrow(C_variables) - 8 + 2  # n variables relevant = rows in C_variables - headers and total + 2 that are grouped with similar variables
 n_relevant_variables_with_no_data =  sum(C_variables$`n in ForC` == 0,  na.rm = TRUE)
@@
 ```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
-Most, but not all, EFDB-relevant ForC variables are shown here (excluded: C concentration in tree wood, annual branch turnover, understory biomass, total ecosystem C in all pools excluding soils and net annual increment thereof).
+Most, but not all, EFDB-relevant ForC variables are shown here.
 Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
@@
 So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.
+Moreover, the process of selecting studies to review for transfer has yet to be optimized for representation across geographic regions, forest types, and variables, with the result that records transferred to date are unevenly distributed across these categories (Figs.
+`r fig_map`-`r fig_histograms`, Table `r table_variables`).
+Future work reviewing records for transfer will be most valuable if priority is given to those from currently under-represented regions and forest types, *to records from countries relying on existing records (as opposed to regular collection of new data through forest inventories)*, to the variables most needed by EFDB users, and to more contemporary records.
 

### 2023-05-08 scripts/3_create_list_of_ all_data_sources_sent.R
Rerun after latest fix + group variables under pretty name #43 #44
+ fix typo in object name

@@ -24,7 +24,7 @@ trace_of_measurement_IDs <- read.csv("data/3-EFDB-forms-ready/trace_of_measureme
 
 ## list of files sent
 list_files_sent <- list.files("data/4-transferred-to-EFDB/", pattern = ".txt", full.names = T)
-
+list_files_sent <- list_files_sent[list_files_sent!= "data/4-transferred-to-EFDB/notes_batch_6.txt"]
 
 # Create the table with all we need about the data we submitted ####
 summary_table <- NULL
@@ -67,66 +67,82 @@ ForC_simplified <- ForC_simplified[ForC_simplified$variable.name %in% gsub("_OM|
 
 
 ## simplify variable name
+variables_in_order <- read.csv("doc/manuscript/figures_tables/C_variables_order_template.csv")
+
 MEASUREMENTS$variable.name_simply <- gsub("_OM|_C", "", MEASUREMENTS$variable.name)
 
 ForC_simplified$variable.name_simply <- gsub("_OM|_C", "", ForC_simplified$variable.name)
 
+all(setdiff(variables_in_order$variable.name, MEASUREMENTS$variable.name_simply) %in% c("delta.biomass", "woody.mortality", "delta.biomass_root", "delta.biomass_root_coarse", 
+                                                                                         "delta.biomass_root_fine", "woody.mortality_root", "delta.deadwood", 
+                                                                                         "delta.deadwood_standing", "delta.deadwood_down", "R_het_deadwood", 
+                                                                                         "delta.litter", "delta.total.ecosystem_2", "delta.soil", "R_het_soil"
+))  # should be TRUE (all the ones that are not in n_in_ForC but are in C_variables_order_template.csv should be in the list pasted here)
+
+setdiff(MEASUREMENTS$variable.name_simply, variables_in_order$variable.name) # should be empty
+
+
+## combine variables that need to be grouped
+MEASUREMENTS$variable_pretty <- variables_in_order$variable[match(MEASUREMENTS$variable.name_simply, variables_in_order$variable.name)]
+
[…]

### 2023-05-08 doc/manuscript/ESSDmanuscript.tex

@@
 \includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
-Most, but not all, EFDB-relevant ForC variables are shown here (excluded: C concentration in tree wood, annual branch turnover, understory biomass, total ecosystem C in all pools excluding soils and net annual increment thereof).
+Most, but not all, EFDB-relevant ForC variables are shown here.
 Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
@@
 As of May 08, 2023, ForC (v4.0) contained 32693 independent records
-(39855 total), 17293 of which were for the the 36 variables relevant to
+(39855 total), 17293 of which were for the the 37 variables relevant to
 EFDB (Fig. 1). These records were distributed across all forested
@@
 
-ForC contained records for 22 of the 36 variables relevant to EFDB
+ForC contained records for 23 of the 37 variables relevant to EFDB
 (Table 2, Fig. 1). The records were very unevenly distributed across
@@
 (Figs. 2-3). So far, only \textasciitilde8\% of the EFDB-relevant data
-in ForC have been sent to EFDB.
+in ForC have been sent to EFDB. Moreover, the process of selecting
+studies to review for transfer has yet to be optimized for
+representation across geographic regions, forest types, and variables,
+with the result that records transferred to date are unevenly
+distributed across these categories (Figs. 2-3, Table 2). Future work
+reviewing records for transfer will be most valuable if priority is
+given to those from currently under-represented regions and forest
+types, \emph{to records from countries relying on existing records (as
+opposed to regular collection of new data through forest inventories)},
+to the variables most needed by EFDB users, and to more contemporary
+records.
 

### 2023-05-08 doc/manuscript/ESSDmanuscript.Rmd
incorporate updated results
 #42

@@
 n_ind_records_OL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "litter"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "O.horizon"]
+n_ind_records_TotEcosystemC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "total.ecosystem_2"] 
 n_ind_records_SOC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "SOM / SOC"] 
@@
 n_ind_records_deltaOL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.O.horizon"] 
-flux_variables_with_records <- c("NPP_woody", "ANPP_woody", "ANPP_woody_ag", "BNPP_root_fine","BNPP_root.turnover_fine","BNPP_root_coarse" )
+flux_variables_with_records <- c("NPP_woody", "ANPP_woody", "ANPP_woody_stem", "ANPP_woody_branch", "woody.mortality_ag", "BNPP_root", "BNPP_root_fine","BNPP_root.turnover_fine","BNPP_root_coarse" , "ANPP_litterfall", "NPP_litter", "R_het_litter", "R_het_soil")
 n_ind_records_flux <- sum(C_variables$`n independent records in ForC`[C_variables$`variable` %in% flux_variables_with_records])
@@
 A total of `r round(n_ind_records_rootbiomass/n_ind_records_EFDB_variables,2)*100`% of relevant records were for root biomass (including fine and coarse root components), while `r round(n_ind_records_biomass/n_ind_records_EFDB_variables,2)*100`% described total biomass.
-The non-living pools were less represented, with `r round(n_ind_records_deadwood/n_ind_records_EFDB_variables,2)*100`% of relevant were for dead wood (including standing and fallen components), `r round(n_ind_records_OL/n_ind_records_EFDB_variables*100,1)`% for litter, and `r round(n_ind_records_SOC/n_ind_records_EFDB_variables*100,1)`% for soil carbon.
+The non-living pools were less represented, with `r round(n_ind_records_deadwood/n_ind_records_EFDB_variables,2)*100`% of relevant were for dead wood (including standing and fallen components), `r round(n_ind_records_OL/n_ind_records_EFDB_variables*100,1)`% for litter, `r round(n_ind_records_TotEcosystemC/n_ind_records_EFDB_variables*100,1)`% for total ecosystem C excluding soils, and `r round(n_ind_records_SOC/n_ind_records_EFDB_variables*100,1)`% for soil carbon.
 
@@
 The only other relevant increment variable with any records was the O horizon (litter) increment, with just `r n_ind_records_deltaOL` records.
-Relevant flux variable records (n=`r n_ind_records_flux`) were limited to the biomass pools (aboveground, belowground, or total) and together constituted `r round(n_ind_records_flux/n_ind_records_EFDB_variables,2)*100`% of ForC's independent records relevant to EFDB.
+Relevant flux variable records (n=`r n_ind_records_flux`) together constituted `r round(n_ind_records_flux/n_ind_records_EFDB_variables,2)*100`% of ForC's independent records relevant to EFDB.
 
@@
 
-<!-- this paragraph will need to be revisited after we solve issues 42 and 43-->
 The records transferred from ForC to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
@@
 `r fig_histograms`).
-Rather, the largest number of records came from North America, followed by Asia and South America (Fig.
+Rather, the largest number of records came from North America, followed by Asia, South America, and Africa (Fig.
 `r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.

### 2023-05-08 doc/manuscript/ESSDmanuscript.tex
incorporate updated results
 #42

--- 2023-05-08 doc/manuscript/ESSDmanuscript.tex
@@
 (v4.0; described here). As of May 08, 2023, ForC contained
-\textasciitilde17293 independent records that would be relevant to EFDB,
-1321 of which have been submitted to date. Among the data in ForC, there
+\textasciitilde19323 independent records that would be relevant to EFDB,
+1438 of which have been submitted to date. Among the data in ForC, there
 is disproportionate representation of biomass (particularly aboveground)
@@
 (Appendix A), defined 15 new variables, implemented enhanced quality
-control, manually reviewed \textgreater1798 records to obtain additional
+control, manually reviewed \textgreater1954 records to obtain additional
 required information, and added 329 new records.
@@
 As of May 08, 2023, ForC (v4.0) contained 32693 independent records
-(39855 total), 17293 of which were for the the 36 variables relevant to
+(39855 total), 19323 of which were for the the 45 variables relevant to
 EFDB (Fig. 1). These records were distributed across all forested
@@
 
-ForC contained records for 22 of the 36 variables relevant to EFDB
+ForC contained records for 31 of the 45 variables relevant to EFDB
 (Table 2, Fig. 1). The records were very unevenly distributed across
 variables. The variable with most records was aboveground biomass,
-representing 45\% of all independent records relevant to EFDB, and
+representing 42\% of all independent records relevant to EFDB, and
 aboveground biomass components (woody biomass or foliage) representing
-an additional 5\%. A total of 30\% of relevant records were for root
-biomass (including fine and coarse root components), while 5\% described
-total biomass. The non-living pools were less represented, with 5\% of
+an additional 5\%. A total of 27\% of relevant records were for root
+biomass (including fine and coarse root components), while 4\% described
+total biomass. The non-living pools were less represented, with 4\% of
 relevant were for dead wood (including standing and fallen components),
-0.4\% for litter, and 2.3\% for soil carbon.
+0.4\% for litter, 0.3\% for total ecosystem C excluding soils, and 2.1\%
+for soil carbon.
 
@@
 increment variable with most records was the aboveground biomass
-increment, representing 0.9\% of all independent records relevant to
+increment, representing 0.8\% of all independent records relevant to
 EFDB. The only other relevant increment variable with any records was
 the O horizon (litter) increment, with just 4 records. Relevant flux
-variable records (n=972) were limited to the biomass pools (aboveground,
-belowground, or total) and together constituted 6\% of ForC's
+variable records (n=2751) together constituted 14\% of ForC's
 independent records relevant to EFDB.
@@
 \hline
-biomass\_ag & 9050 & 7750 & 1267 & 723\\
+biomass\_ag & 9456 & 8155 & 1348 & 764\\
 \hline
@@
 \hline
+ANPP\_woody\_stem & 949 & 622 & 60 & 61\\
+\hline
+ANPP\_woody\_branch & 243 & 200 & 4 & 4\\
+\hline
 woody.mortality\_ag & 112 & 75 & 47 & 50\\
 \hline
+stem\_pC & 9 & 0 & 0 & 0\\
+\hline
 \textbf{Belowground biomass} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
@@
 \hline
+BNPP\_root & 577 & 416 & 0 & 0\\
+\hline
 BNPP\_root\_fine & 488 & 331 & 0 & 0\\
@@
 \hline
-NPP\_litter & 0 & 0 & 0 & 0\\
+ANPP\_litterfall & 294 & 253 & 11 & 11\\
+\hline
+NPP\_litter & 94 & 70 & 0 & 0\\
 \hline
@@
 \hline
+\textbf{Total Ecosystem C (excl. soils)} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
+\hline
+total.ecosystem\_2 & 64 & 64 & 0 & 0\\
+\hline
+delta.total.ecosystem\_2 & 0 & 0 & 0 & 0\\
+\hline
 \textbf{Soil organic matter} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
@@
 \hline
-\textbf{TOTAL} & \textbf{20939} & \textbf{17293} & \textbf{2127} & \textbf{1321}\\
+\textbf{TOTAL} & \textbf{23575} & \textbf{19323} & \textbf{2283} & \textbf{1438}\\
 \hline
@@
 
-As of May 08, 2023, we had reviewed or added 2127 EFDB-relevant records,
-1321 records of which were sent to EFDB, and 376 of which have been
-reviewed, accepted, and posted (Figs. 2-3, Table 2). The 38\%
+As of May 08, 2023, we had reviewed or added 2283 EFDB-relevant records,
+1438 records of which were sent to EFDB, and 376 of which have been
+reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
 attenuation between records reviewed and those sent to EFDB was
@@
 and it is not expected that many -- if any -- records will be rejected.
-
-The records transferred from ForC to EFDB were broadly distributed
-across Earth's forests (Fig. 2). However, the density of these records
-was very unevenly distributed across continents, biomes, and forest
-types and was not proportional to the numbers of relevant records in
-ForC (Fig. 3). Rather, the largest number of records came from North
-America, followed by Asia and South America (Fig. 3c), with the most
-represented FAO ecozones being boreal coniferous forest, temperate
-continental forest, and temperate mountain systems, followed by tropical
-rain forests and moist deciduous forests (Fig. 3b). In terms of dominant
-vegetation, by far the most records came from needleleaf evergreen
-forests, followed by broadleaf deciduous and broadleaf evergreen (Fig.
-3b). The largest records came from mature forests (\textgreater100
-years), followed by young and intermediate-aged stands (Fig. 3d).
-
-In terms of variables transferred, \ldots{} \emph{{[}DETAILS{]}}
-
-\section{Recommendations}
-
-Based on our experience contributing forest C data to EFDB via ForC, we
-make several recommendations as to how scientists can improve forest C
-records in EFDB through database work (section 6.1), new data collection
-and analysis (section 6.2), and reporting (section 6.3). We also
-highlight notable mismatches between IPCC accounting methods and forest
-C mensuration (section 6.4).
-
-\subsection{Database needs}
-
-There is vast potential to expand forest C data in EFDB by completing
-the process of reviewing and sending data that are already in ForC
-(Figs. 2-3). So far, only \textasciitilde8\% of the EFDB-relevant data
-in ForC have been sent to EFDB.
-
-Moreover, there are innumerous published EFDB-relevant forest C data
-that are not included in ForC, with more being published on a nearly
-daily basis. Coverage of particular variables or regions could be vastly
-improved through systematic review of the literature. \emph{(There are
-some efforts underway, including a few that Susan can specify.)} Such
-reviews are necessary to assess the availability of published
-EFDB-relevant forest C records, versus those that require additional
-data collection and analysis.
-
-\subsection{Data collection and analysis needs}
-
-New data collection and analysis is needed to fill notable knowledge
-gaps. While aboveground biomass stocks in particular have received --
-and continue to receive -- significant research attention, other pools
-and variables remain poorly quantified \citep[Table
-2,][]{anderson-teixeira_carbon_2021}. Furthermore, data distribution is
-uneven across forest types and geographical regions (Figs. 2-3). For
-instance, C cycling of tropical forests -- particularly in Africa --
-remains relatively poorly characterized, in large part due to
-substantial barriers to data collection and distribution
-\citep{delima_making_2022} \emph{(add some more here?)}
-
-Several variables of value for IPCC C accounting have not been
-calculated and presented as would be possible given the same forest
-census data and minimal extra research effort. For example, aboveground
-woody mortality (\emph{woody.mortality\_ag}) and aboveground biomass
-increment (\emph{delta.agb}) can be calculated from the same census data
-as aboveground woody productivity (\emph{ANPP\_woody}), yet the latter
-has received far more research attention, and correspondingly has far
-more records in ForC \citetext{\citealp[Table
-2,][]{anderson-teixeira_carbon_2021}; \citealp[but
-see][]{piponiot_distribution_2022}}. Similarly, live coarse root
-biomass, total biomass, and changes in both of these pools could in
-theory be easily be estimated in parallel with aboveground biomass, with
-the greatest barrier being availability of reliable allometries, as have
-been developed for aboveground biomass
-\citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022}.
-However, while equations for estimating root (and thereby total) biomass
-require improvement, they do exist \emph{for many forest
-types}\citep{refs}, and IPCC provides default recommendations of
-below-ground to above-ground ratios for estimation of root biomass
-\citep{ipcc_2019_2019}. In addition, standing dead trees are captured in
-most forest censuses and could be used to estimate standing dead wood,
-although additional data on breakage would be needed for accurate
-accounting. We recommend that, when possible, researchers calculate
-these, following the reporting guidelines specified in section 6.3.
-
-\emph{Other EFDB-relevant variables require more effort but are
-warranted given their importance for forest C accounting.} Given
-widespread trends of increasing tree mortality \citep{refs}, including
-through severe natural disturbance \citep{refs}, better characterization
-of dead wood will be critical. \ldots{}
-
-\subsection{Data reporting needs}
-
-We recommend that, unless they have some specific reason to do
-otherwise, researchers calculate and report the values according to IPCC
-standards (Table 3). It is particularly noteworthy that simple decisions
-on the presentation of results will determine whether the data meet the
-criteria for inclusion in EFDB. Some examples are as follows: (1)
-presenting data only in a figure makes it ineligible for inclusion in
-EFDB, whereas presentation in a table or supplementary data file allows
-inclusion; (2) direct presentation of all relevant variables allows
-inclusion, whereas presenting only components of variables of interest
-(e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or
-requiring simple mathematical operations to obtain a variable of
-interest (e.g., \emph{delta.agb} = \emph{ANPP\_woody} -
-\emph{woody.mortality.agb}) disqualifies data from inclusion; (3)
-matching IPCC-defined thresholds for defining C pools (Table 1), which
-may vary by country, can make the data far more relevant for IPCC
-accounting (e.g., using a 10 cm cutoff between dead wood and litter,
-presenting soil C to a depth of 30 cm). It should also be emphasized
-that reporting of 95\% confidence intervals (or other metrics of error),
-when applicable, is highly desirable and makes the data more relevant to
-IPCC. Reports which had the most successful data transfers used EFDB
-variables and had clear tables showing their results.
-
-\begin{table}
-
-\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
-\centering
-\fontsize{10}{12}\selectfont
-\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X}
-\hline
-\textbf{criteria} & \textbf{recommendation} & \textbf{rationale}\\
-\hline
-variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data.
-& Estimates of relevant variables are not always calcualted.\\
-\hline
-forest census methods & Adopt IPCC guidelines (country-specific) for minimum stem size in censues in census and reporting.
-Ideally, census stem down to the smallest diameter feasible.
-& IPCC biomass pool definition includes all living vegetation, but understory may be excluded when contribution is minor.\\
-\hline
- & *taxa to include* & \\
-\hline
-dead organic matter sampling & *include damage estimates on standing dead trees* & \\
-\hline
- & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & \\
-\hline
-belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & IPCC biomass pool definition includes all living vegetation\\
-\hline
-reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables.
-& EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
-\hline
-reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
-\hline
-reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
-\hline
-\end{tabu}
-\end{table}
-
-For those compiling published data (e.g., for meta-analyses), the data
-set can have added value if all information required by IPCC is
-extracted from original publications. This includes -- but is not
-limited to -- retaining original values as presented without
-modification or rounding, noting whether data were digitized, recording
-confidence intervals, and recording all required fields (as indicated in
-the EFDB's bulk import template). The significant effort required to map
-a database into EFDB has been accomplished here (Appendix B), and we
-welcome other researchers to use the ForC template.
-
-Once EFDB-relevant data are available in peer-reviewed publications,
-they may be submitted directly to EFDB or may use the ForC - EFDB data
-pipline developed here. For individual publications, the former option
-will generally be more efficient. However, by getting the data into ForC
-as well as EFDB, the latter option will allow the data to be more
-broadly useful--for example, being used for basic science
-\citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021}
-or model benchmarking \citep{fer_ecosystem_2021}.
-
-\subsection{Mismatches between IPCC accounting methods and forest C
-mensuration}
-
-Remote sensing biomass estimates include standing dead wood
-\citep{duncanson_aboveground_2021}.
-
-IPCC accounting methods cannot leverage eddy-covariance measurements,
-which are widely seen as the best available method for quantifying
-ecosystem-atmosphere gas exchange.
-
-An important challenge is that forests are changing rapidly, and data
-collected a decade ago may no longer be relevant, particularly in the
-cases of C increments and fluxes.
-
-\section{Conclusions}
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
update through start of section 6
note unresolved problem #46

@@
 n_records_sent_to_EFDB = C_variables[length(C_variables[,"variable"]), "n sent to EFDB"] # n records sent to EFDB
+n_relevant_variables_no_records_sent =  sum(C_variables$`n sent to EFDB` == 0,  na.rm = TRUE)
+n_relevant_variables_records_sent= n_relevant_variables - n_relevant_variables_no_records_sent
+n_records_sent_biomass<- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass"]
+n_records_sent_agb <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_ag"]
+n_records_sent_agb_components <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_ag_foliage"]
+n_records_sent_rootbiomass <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_root"] 
+n_records_sent_rootbiomass_components <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_root_coarse"]
+n_records_sent_deadwood<- C_variables$`n sent to EFDB`[C_variables$`variable` == "deadwood"] 
+n_records_sent_deadwood_components<- C_variables$`n sent to EFDB`[C_variables$`variable` == "deadwood_down"]+C_variables$`n sent to EFDB`[C_variables$`variable` == "deadwood_standing"]
+n_records_sent_litter<- C_variables$`n sent to EFDB`[C_variables$`variable` == "litter"] +C_variables$`n sent to EFDB`[C_variables$`variable` == "O.horizon"]
+n_records_sent_soil<- C_variables$`n sent to EFDB`[C_variables$`variable` == "SOM / SOC"] 
+n_records_sent_stocks = n_records_sent_biomass+n_records_sent_agb+n_records_sent_agb_components+n_records_sent_rootbiomass+n_records_sent_rootbiomass_components+n_records_sent_deadwood+n_records_sent_deadwood_components+ n_records_sent_litter+ n_records_sent_soil
+n_records_sent_delta.agb<- C_variables$`n sent to EFDB`[C_variables$`variable` == "delta.agb"]
+n_records_sent_delta.litter<- C_variables$`n sent to EFDB`[C_variables$`variable` == "delta.O.horizon"]+C_variables$`n sent to EFDB`[C_variables$`variable` == "delta.litter"]
+n_records_sent_increments <- n_records_sent_delta.agb+n_records_sent_delta.litter
+n_records_sent_fluxes = n_records_sent_to_EFDB-n_records_sent_stocks-n_records_sent_increments
 
@@
 Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[ @rvest]).
+For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[ @wickham_rvest_2022]. 
 That information was manually retrieved when the web scraping failed.
@@
 
-In terms of variables transferred, ...
-*[DETAILS]*
+In terms of variables transferred, records were sent for `r n_relevant_variables_records_sent` variables, including variables from each C pool (Table `r table_variables`).
+The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records sent were for C stocks, including
+`r round(n_records_sent_biomass/n_records_sent_to_EFDB,2)*100`% for total biomass, 
+`r round(n_records_sent_agb/n_records_sent_to_EFDB,2)*100`% for aboveground biomass,
+`r round(n_records_sent_agb_components/n_records_sent_to_EFDB,2)*100`% for components of aboveground biomass (wood or foliage), 
+`r round(n_records_sent_rootbiomass/n_records_sent_to_EFDB,2)*100`% for root biomass, 
+`r round(n_records_sent_rootbiomass_components/n_records_sent_to_EFDB,2)*100`% for components of root biomass (coarse or fine roots), 
+`r round(n_records_sent_deadwood/n_records_sent_to_EFDB,2)*100`% for dead wood, 
+`r round(n_records_sent_deadwood_components/n_records_sent_to_EFDB,2)*100`% for components of dead wood (standing or fallen), 
+`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (OL layer or entire O horizon),
+and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
+Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, including
+`r round(n_records_sent_delta.agb/n_records_sent_to_EFDB*100,2)`% for aboveground biomass increments 
+and `r round(n_records_sent_delta.litter/n_records_sent_to_EFDB*100,2)`% for organic horizon increments.
+The remaining (`r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`%) of records sent described fluxes,
+all of which were either inputs or outputs to the aboveground biomass pool,
+and a subset of which also described inputs to the dead wood or litter pool (Table `r table_variables`, Fig. `r fig_variable_mapping`).
 
@@
 So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.
-Moreover, the process of selecting studies to review for transfer has yet to be optimized for representation across geographic regions, forest types, and variables, with the result that records transferred to date are unevenly distributed across these categories (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`).
-Future work reviewing records for transfer will be most valuable if priority is given to those from currently under-represented regions and forest types, *to records from countries relying on existing records (as opposed to regular collection of new data through forest inventories)*, to the variables most needed by EFDB users, and to more contemporary records.
+Although this process requires manual review of records, the addition of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily transferred to EFDB using the system developed here.
+Future efforts to review studies for transfer should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
+`r fig_map`-`r fig_histograms`, Table `r table_variables`), *to records from countries relying on existing records (as opposed to regular collection of new data through forest inventories)*, to the variables most needed by EFDB users, and to the more contemporary records.
 
-Moreover, there are innumerous published EFDB-relevant forest C data that are not included in ForC, with more being published on a nearly daily basis. 
+In addition to the large potential to expand EFDB using records already in ForC, there are innumerous published EFDB-relevant forest C data that are not currently included in ForC, with more being published on a nearly daily basis.
 Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
-*(There are some efforts underway, including a few that Susan can specify.)*
-Such reviews are necessary to assess the availability of published EFDB-relevant forest C records, versus those that require additional data collection and analysis.
+Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022], ... *others?  (SUSAN)*. 
+Such reviews valuable for assessing the availability of published EFDB-relevant forest C records, versus those that require additional data collection and analysis.
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.tex
update through start of section 6
note unresolved problem #46

--- 2023-05-08 doc/manuscript/ESSDmanuscript.tex
@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 08, 2023, ForC contained
-\textasciitilde17293 independent records that would be relevant to EFDB,
-1321 of which have been submitted to date. Among the data in ForC, there
+(v4.0; described here). As of May 12, 2023, ForC contained
+\textasciitilde19323 independent records that would be relevant to EFDB,
+1438 of which have been submitted to date. Among the data in ForC, there
 is disproportionate representation of biomass (particularly aboveground)
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 08, 2023, ForC contained 39855 records from 10589 plots in
+As of May 12, 2023, ForC contained 39855 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 citation, abstract and language of the publications, based on their DOI,
-using R package \texttt{rvest}\citep{rvest}). That information was
-manually retrieved when the web scraping failed.
+using R package \texttt{rvest}\citep{wickham_rvest_2022}. That
+information was manually retrieved when the web scraping failed.
 
@@
 belowground components of woody mortality (\emph{woody.mortality}).
-Although most of these variables lacked records in ForC as of May 08,
+Although most of these variables lacked records in ForC as of May 12,
 2023, their addition gave the structure such that records can be
@@
 posting in the database -- a process that lags behind records transfer
-and had not yet been completed for all records sent as of May 08, 2023.
+and had not yet been completed for all records sent as of May 12, 2023.
 
@@
 
-As of May 08, 2023, ForC (v4.0) contained 32693 independent records
-(39855 total), 17293 of which were for the the 36 variables relevant to
+As of May 12, 2023, ForC (v4.0) contained 32693 independent records
+(39855 total), 19323 of which were for the the 45 variables relevant to
 EFDB (Fig. 1). These records were distributed across all forested
@@
 
-As of May 08, 2023, we had reviewed or added 2127 EFDB-relevant records,
-1321 records of which were sent to EFDB, and 376 of which have been
-reviewed, accepted, and posted (Figs. 2-3, Table 2). The 38\%
+As of May 12, 2023, we had reviewed or added 2283 EFDB-relevant records,
+1438 records of which were sent to EFDB, and 376 of which have been
+reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
 attenuation between records reviewed and those sent to EFDB was
@@
 ForC (Fig. 3). Rather, the largest number of records came from North
-America, followed by Asia and South America (Fig. 3c), with the most
-represented FAO ecozones being boreal coniferous forest, temperate
+America, followed by Asia, South America, and Africa (Fig. 3c), with the
+most represented FAO ecozones being boreal coniferous forest, temperate
 continental forest, and temperate mountain systems, followed by tropical
@@
 
-In terms of variables transferred, \ldots{} \emph{{[}DETAILS{]}}
+In terms of variables transferred, records were sent for 21 variables,
+including variables from each C pool (Table 2). The majority (82\%) of
+records sent were for C stocks, including 3\% for total biomass, 53\%
+for aboveground biomass, 4\% for components of aboveground biomass (wood
+or foliage), 4\% for root biomass, 2\% for components of root biomass
+(coarse or fine roots), 5\% for dead wood, 3\% for components of dead
+wood (standing or fallen), 4\% for litter (OL layer or entire O
+horizon), and 4\% for SOM/ SOC. Increment records totaled 9\% of records
+sent, including 8.55\% for aboveground biomass increments and 0.28\% for
+organic horizon increments. The remaining (9\%) of records sent
+described fluxes, all of which were either inputs or outputs to the
+aboveground biomass pool, and a subset of which also described inputs to
+the dead wood or litter pool (Table 2, Fig. 1).
 
@@
 the process of reviewing and sending data that are already in ForC
-(Figs. 2-3). So far, only \textasciitilde8\% of the EFDB-relevant data
-in ForC have been sent to EFDB.
-
-Moreover, there are innumerous published EFDB-relevant forest C data
-that are not included in ForC, with more being published on a nearly
-daily basis. Coverage of particular variables or regions could be vastly
-improved through systematic review of the literature. \emph{(There are
-some efforts underway, including a few that Susan can specify.)} Such
-reviews are necessary to assess the availability of published
-EFDB-relevant forest C records, versus those that require additional
-data collection and analysis.
+(Figs. 2-3). So far, only \textasciitilde7\% of the EFDB-relevant data
+in ForC have been sent to EFDB. Although this process requires manual
+review of records, the addition of new records to EFDB is hugely
+facilitated by the fact that most pertinent information for each record
+is already entered in ForC and can be easily transferred to EFDB using
+the system developed here. Future efforts to review studies for transfer
+should optimize for representation across geographic regions, forest
+types, and variables, giving priority to those from currently
+under-represented regions and forest types (Figs. 2-3, Table 2),
+\emph{to records from countries relying on existing records (as opposed
+to regular collection of new data through forest inventories)}, to the
+variables most needed by EFDB users, and to the more contemporary
+records.
+
+In addition to the large potential to expand EFDB using records already
+in ForC, there are innumerous published EFDB-relevant forest C data that
+are not currently included in ForC, with more being published on a
+nearly daily basis. Coverage of particular variables or regions could be
+vastly improved through systematic review of the literature. Indeed,
+recent efforts have compiled large databases of relevant data from
+monoculture plantation forests \citep{bukoski_rates_2022}, \ldots{}
+\emph{others? (SUSAN)}. Such reviews valuable for assessing the
+availability of published EFDB-relevant forest C records, versus those
+that require additional data collection and analysis.
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
fix variable counts
partially fixes #46

@@
 C_variables <- read.csv(paste0(path_to_dir,"/doc/manuscript/figures_tables/C_variables.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_relevant_variables = nrow(C_variables) - 8 + 2  # n variables relevant = rows in C_variables - headers and total + 2 that are grouped with similar variables
+n_relevant_variables = nrow(C_variables) - 8  # n variables relevant = rows in C_variables - headers and total 
 n_relevant_variables_with_no_data =  sum(C_variables$`n in ForC` == 0,  na.rm = TRUE)
@@
 n_records_sent_increments <- n_records_sent_delta.agb+n_records_sent_delta.litter
-n_records_sent_fluxes = n_records_sent_to_EFDB-n_records_sent_stocks-n_records_sent_increments
+n_records_sent_fluxes <- sum(C_variables$`n sent to EFDB`[C_variables$`variable` %in% flux_variables_with_records])
+n_records_sent_fluxes_check = n_records_sent_to_EFDB-n_records_sent_stocks-n_records_sent_increments
 
@@
 
-We added a total of `r n_new_variables` new EFDB-relevant variables to the set of named and defined variables (Fig.
+To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables (Fig.
 `r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
@@
 
-ForC contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables relevant to EFDB (Table `r table_variables`, Fig. `r fig_variable_mapping`).
+ForC contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Table `r table_variables`, Fig.
+`r fig_variable_mapping`).
 The records were very unevenly distributed across variables. 
@@
              longtable = T,
-             caption = "\\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.}") %>%
+             caption = "\\textbf{Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.}") %>%
   row_spec(0, bold = TRUE)  %>%
@@
 
-In terms of variables transferred, records were sent for `r n_relevant_variables_records_sent` variables, including variables from each C pool (Table `r table_variables`).
+In terms of variables transferred, records were sent for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
 The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records sent were for C stocks, including
@@
 `r round(n_records_sent_deadwood_components/n_records_sent_to_EFDB,2)*100`% for components of dead wood (standing or fallen), 
-`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (OL layer or entire O horizon),
+`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (entire O horizon or OL layer component),
 and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.

### 2023-05-12 doc/manuscript/ESSDmanuscript.tex
fix variable counts
partially fixes #46

@@
 
-We added a total of 15 new EFDB-relevant variables to the set of named
-and defined variables (Fig. 1), counting each pair of variables with
-units in C (ending in \texttt{\_C}) or organic matter (ending in
-\texttt{\_OM}) as one. The majority of these were increment variables
-(n=11), adding to only one previously defined increment variable
-(aboveground biomass increment, \emph{delta.agb}). These are directly
-related to C stocks as previously defined in ForC, with
-``\emph{delta.}'' added in front of the variable name. Further, we added
-variables capturing the belowground component of woody mortality
-(\emph{woody.mortality\_root}) and the combined aboveground and
-belowground components of woody mortality (\emph{woody.mortality}).
-Although most of these variables lacked records in ForC as of May 12,
-2023, their addition gave the structure such that records can be
-populated over time. Finally, to provide better definition of the
-previously existing variable \emph{organic.layer}, which has a nebulous
-definition that reflects the varied definitions adopted by original
-studies, we added two clearly defined variables: \emph{litter}
-(relatively undecomposed plant material/ OL horizon), and
+To create structure for EFDB-relevant records, we added a total of 15
+new variables to the set of named and defined variables (Fig. 1),
+counting each pair of variables with units in C (ending in \texttt{\_C})
+or organic matter (ending in \texttt{\_OM}) as one. The majority of
+these were increment variables (n=11), adding to only one previously
+defined increment variable (aboveground biomass increment,
+\emph{delta.agb}). These are directly related to C stocks as previously
+defined in ForC, with ``\emph{delta.}'' added in front of the variable
+name. Further, we added variables capturing the belowground component of
+woody mortality (\emph{woody.mortality\_root}) and the combined
+aboveground and belowground components of woody mortality
+(\emph{woody.mortality}). Although most of these variables lacked
+records in ForC as of May 12, 2023, their addition gave the structure
+such that records can be populated over time. Finally, to provide better
+definition of the previously existing variable \emph{organic.layer},
+which has a nebulous definition that reflects the varied definitions
+adopted by original studies, we added two clearly defined variables:
+\emph{litter} (relatively undecomposed plant material/ OL horizon), and
 \emph{O.horizon} (entire O-horizon, including OL).
@@
 As of May 12, 2023, ForC (v4.0) contained 32693 independent records
-(39855 total), 19323 of which were for the the 45 variables relevant to
+(39855 total), 19323 of which were for the the 43 variables relevant to
 EFDB (Fig. 1). These records were distributed across all forested
@@
 
-ForC contained records for 31 of the 45 variables relevant to EFDB
-(Table 2, Fig. 1). The records were very unevenly distributed across
-variables. The variable with most records was aboveground biomass,
-representing 42\% of all independent records relevant to EFDB, and
-aboveground biomass components (woody biomass or foliage) representing
-an additional 5\%. A total of 27\% of relevant records were for root
-biomass (including fine and coarse root components), while 4\% described
-total biomass. The non-living pools were less represented, with 4\% of
-relevant were for dead wood (including standing and fallen components),
-0.4\% for litter, 0.3\% for total ecosystem C excluding soils, and 2.1\%
-for soil carbon.
+ForC contained records for 29 of the 43 variables (or closely-related
+variable groups) relevant to EFDB (Table 2, Fig. 1). The records were
+very unevenly distributed across variables. The variable with most
+records was aboveground biomass, representing 42\% of all independent
+records relevant to EFDB, and aboveground biomass components (woody
+biomass or foliage) representing an additional 5\%. A total of 27\% of
+relevant records were for root biomass (including fine and coarse root
+components), while 4\% described total biomass. The non-living pools
+were less represented, with 4\% of relevant were for dead wood
+(including standing and fallen components), 0.4\% for litter, 0.3\% for
+total ecosystem C excluding soils, and 2.1\% for soil carbon.
 
@@
 \begin{longtable}[t]{l|l|l|l|l}
-\caption{\label{tab:table_variables}\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.}}\\
+\caption{\label{tab:table_variables}\textbf{Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.}}\\
 \hline
@@
 \endfirsthead
-\caption[]{\textbf{Numbers of records of ForC variables relevant to, and sent to, EFDB.} \textit{(continued)}}\\
+\caption[]{\textbf{Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.} \textit{(continued)}}\\
 \hline
@@
 
-In terms of variables transferred, records were sent for 21 variables,
-including variables from each C pool (Table 2). The majority (82\%) of
-records sent were for C stocks, including 3\% for total biomass, 53\%
-for aboveground biomass, 4\% for components of aboveground biomass (wood
-or foliage), 4\% for root biomass, 2\% for components of root biomass
-(coarse or fine roots), 5\% for dead wood, 3\% for components of dead
-wood (standing or fallen), 4\% for litter (OL layer or entire O
-horizon), and 4\% for SOM/ SOC. Increment records totaled 9\% of records
-sent, including 8.55\% for aboveground biomass increments and 0.28\% for
-organic horizon increments. The remaining (9\%) of records sent
-described fluxes, all of which were either inputs or outputs to the
-aboveground biomass pool, and a subset of which also described inputs to
-the dead wood or litter pool (Table 2, Fig. 1).
+In terms of variables transferred, records were sent for 19 variables
+(or closely-related variable groups), including variables from each C
+pool (Table 2). The majority (82\%) of records sent were for C stocks,
+including 3\% for total biomass, 53\% for aboveground biomass, 4\% for
+components of aboveground biomass (wood or foliage), 4\% for root
+biomass, 2\% for components of root biomass (coarse or fine roots), 5\%
+for dead wood, 3\% for components of dead wood (standing or fallen), 4\%
+for litter (entire O horizon or OL layer component), and 4\% for SOM/
+SOC. Increment records totaled 9\% of records sent, including 8.55\% for
+aboveground biomass increments and 0.28\% for organic horizon
+increments. The remaining (9\%) of records sent described fluxes, all of
+which were either inputs or outputs to the aboveground biomass pool, and
+a subset of which also described inputs to the dead wood or litter pool
+(Table 2, Fig. 1).
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
resolve rounding problem
#46

@@
 and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
-Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, including
-`r round(n_records_sent_delta.agb/n_records_sent_to_EFDB*100,2)`% for aboveground biomass increments 
-and `r round(n_records_sent_delta.litter/n_records_sent_to_EFDB*100,2)`% for organic horizon increments.
+Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, all for aboveground biomass excepting `r n_records_sent_delta.litter` records for organic horizon increments.
 The remaining (`r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`%) of records sent described fluxes,

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
update through end of 6.1

@@
 and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
-Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, all for aboveground biomass excepting `r n_records_sent_delta.litter` records for organic horizon increments.
-The remaining (`r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`%) of records sent described fluxes,
-all of which were either inputs or outputs to the aboveground biomass pool,
-and a subset of which also described inputs to the dead wood or litter pool (Table `r table_variables`, Fig. `r fig_variable_mapping`).
+Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, virtually all for aboveground biomass (excepting `r n_records_sent_delta.litter` records for delta.O.horizon).
+The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records sent described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Table `r table_variables`, Fig.
+`r fig_variable_mapping`).
 
@@
 Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022], ... *others?  (SUSAN)*. 
-Such reviews valuable for assessing the availability of published EFDB-relevant forest C records, versus those that require additional data collection and analysis.
+Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.tex
resolve rounding problem
#46
update through end of 6.1

@@
 for litter (entire O horizon or OL layer component), and 4\% for SOM/
-SOC. Increment records totaled 9\% of records sent, including 8.55\% for
-aboveground biomass increments and 0.28\% for organic horizon
-increments. The remaining (9\%) of records sent described fluxes, all of
-which were either inputs or outputs to the aboveground biomass pool, and
-a subset of which also described inputs to the dead wood or litter pool
-(Table 2, Fig. 1).
+SOC. Increment records totaled 9\% of records sent, virtually all for
+aboveground biomass (excepting 4 records for delta.O.horizon). The
+remaining 9\% of records sent described fluxes, all of which were either
+inputs or outputs to the aboveground biomass pool, a subset of which
+also described inputs to the dead wood or litter pool (Table 2, Fig. 1).
 
@@
 monoculture plantation forests \citep{bukoski_rates_2022}, \ldots{}
-\emph{others? (SUSAN)}. Such reviews valuable for assessing the
-availability of published EFDB-relevant forest C records, versus those
-that require additional data collection and analysis.
+\emph{others? (SUSAN)}. Beyond expanding collections of relevant forest
+C records, such reviews are valuable for assessing the availability of
+published records and identifying variables and regions that require
+additional data collection and analysis.
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
edit through end of section 6.3

@@
 New data collection and analysis is needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- significant research attention, other pools and variables remain poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021].
+While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021], production of an accurate global map of forest C stocks remains an ongoing challenge [@refs].
+Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@refs].
 Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
-For instance, C cycling of tropical forests -- particularly in Africa -- remains relatively poorly characterized, in large part due to substantial barriers to data collection and distribution [@delima_making_2022]
-*(add some more here?)*
+For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
+Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@refs].
 
-Several variables of value for IPCC C accounting have not been calculated and presented as would be possible given the same forest census data and minimal extra research effort.
+Several EFDB-relevant variables have not been calculated and presented as frequently as would be possible given existing forest census data and minimal extra research effort.
 For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022].
-Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be easily be estimated in parallel with aboveground biomass, with the greatest barrier being availability of reliable allometries, as have been developed for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
-However, while equations for estimating root (and thereby total) biomass require improvement, they do exist *for many forest types*[@refs], and IPCC provides default recommendations of below-ground to above-ground ratios for estimation of root biomass [@ipcc_2019_2019].
+Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
+However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017].
 In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate accounting. 
-We recommend that, when possible, researchers calculate these, following the reporting guidelines specified in section `r section_data_reporting_needs`.
+We recommend that, when possible, researchers calculate and report these variables, following the reporting guidelines specified in section `r section_data_reporting_needs`.
 
-*Other EFDB-relevant variables require more effort but are warranted given their importance for forest C accounting.*
-Given widespread trends of increasing tree mortality [@refs], including through severe natural disturbance [@refs], better characterization of dead wood will be critical.
-...
+Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for forest C accounting.
+Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@refs], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@refs].
+Given observations of increasing tree mortality in some forested regions [@refs], including through severe natural disturbance [@refs], better characterization of forest dead wood will be critical.
+Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@refs].
+Significant investment in ground-based forest research will be critical to filling these gaps.
 
@@
 
-We recommend that, unless they have some specific reason to do otherwise, researchers calculate and report the values according to IPCC standards (Table `r table_recommendations`).
-It is particularly noteworthy that simple decisions on the presentation of results will determine whether the data meet the criteria for inclusion in EFDB.
+We recommend that, in order to make research valuable to C accounting under IPCC guidelines, researchers calculate and report results according to IPCC standards (Table `r table_recommendations`).
+It is particularly noteworthy that simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
 Some examples are as follows:
-(1) presenting data only in a figure makes it ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion;
-(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies data from inclusion;
+(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion;
+(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;
 (3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
@@
              longtable = FALSE,
-             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....") %>%
+             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}") %>%
   row_spec(0, bold = TRUE)%>%
@@
 
-For those compiling published data (e.g., for meta-analyses), the data set can have added value if all information required by IPCC is extracted from original publications. 
+For those compiling published records (e.g., for meta-analyses), the data set can have added value if all information required by EFDB is extracted from original publications. 
 This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
@@
 
-Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipline developed here.
+Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipeline developed here.
 For individual publications, the former option will generally be more efficient.
-However, by getting the data into ForC as well as EFDB, the latter option will allow the data to be more broadly useful--for example, being used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021] or model benchmarking [@fer_ecosystem_2021].
+However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mititagation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.tex
edit through end of section 6.3

@@
 gaps. While aboveground biomass stocks in particular have received --
-and continue to receive -- significant research attention, other pools
-and variables remain poorly quantified \citep[Table
-2,][]{anderson-teixeira_carbon_2021}. Furthermore, data distribution is
-uneven across forest types and geographical regions (Figs. 2-3). For
-instance, C cycling of tropical forests -- particularly in Africa --
-remains relatively poorly characterized, in large part due to
-substantial barriers to data collection and distribution
-\citep{delima_making_2022} \emph{(add some more here?)}
-
-Several variables of value for IPCC C accounting have not been
-calculated and presented as would be possible given the same forest
-census data and minimal extra research effort. For example, aboveground
-woody mortality (\emph{woody.mortality\_ag}) and aboveground biomass
-increment (\emph{delta.agb}) can be calculated from the same census data
-as aboveground woody productivity (\emph{ANPP\_woody}), yet the latter
-has received far more research attention, and correspondingly has far
-more records in ForC \citetext{\citealp[Table
+and continue to receive -- by far the most research attention
+\citep[Table 2,][]{anderson-teixeira_carbon_2021}, production of an
+accurate global map of forest C stocks remains an ongoing challenge
+\citep{refs}. Other pools and variables remain very poorly quantified
+\citep[Table 2,][]{anderson-teixeira_carbon_2021}, introducing
+substantive uncertainties into global forest C budgets \citep{refs}.
+Furthermore, data distribution is uneven across forest types and
+geographical regions (Figs. 2-3). For instance, data on C cycling of
+tropical forests -- particularly in Africa -- remains relatively sparse,
+in large part due to substantial barriers to data collection and
+distribution \citep{delima_making_2022}. Significant investment in
+research and researchers focused on ground-based measurement of forest C
+in such regions will be important to filling knowledge gaps in forest C
+cycling \citep{delima_making_2022, refs}.
+
+Several EFDB-relevant variables have not been calculated and presented
+as frequently as would be possible given existing forest census data and
+minimal extra research effort. For example, aboveground woody mortality
+(\emph{woody.mortality\_ag}) and aboveground biomass increment
+(\emph{delta.agb}) can be calculated from the same census data as
+aboveground woody productivity (\emph{ANPP\_woody}), yet the latter has
+received far more research attention, and correspondingly has far more
+records in ForC \citetext{\citealp[Table
 2,][]{anderson-teixeira_carbon_2021}; \citealp[but
@@
 biomass, total biomass, and changes in both of these pools could in
-theory be easily be estimated in parallel with aboveground biomass, with
-the greatest barrier being availability of reliable allometries, as have
-been developed for aboveground biomass
+theory be estimated in parallel with aboveground biomass, with the
+greatest barrier being that allometric models for estimating root
+biomass are not as reliable or easily available as are those for
+aboveground biomass
 \citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022}.
 However, while equations for estimating root (and thereby total) biomass
-require improvement, they do exist \emph{for many forest
-types}\citep{refs}, and IPCC provides default recommendations of
-below-ground to above-ground ratios for estimation of root biomass
-\citep{ipcc_2019_2019}. In addition, standing dead trees are captured in
-most forest censuses and could be used to estimate standing dead wood,
-although additional data on breakage would be needed for accurate
-accounting. We recommend that, when possible, researchers calculate
-these, following the reporting guidelines specified in section 6.3.
-
-\emph{Other EFDB-relevant variables require more effort but are
-warranted given their importance for forest C accounting.} Given
-widespread trends of increasing tree mortality \citep{refs}, including
-through severe natural disturbance \citep{refs}, better characterization
-of dead wood will be critical. \ldots{}
+require improvement, they do exist for many forest types
+\citep[e.g.,][]{brassard_coarse_2011, chojnacky_updated_2014, waring_overlooking_2017}.
+In addition, standing dead trees are captured in most forest censuses
+and could be used to estimate standing dead wood, although additional
+data on breakage would be needed for accurate accounting. We recommend
+that, when possible, researchers calculate and report these variables,
+following the reporting guidelines specified in section 6.3.
+
+Filling knowledge gaps in other EFDB-relevant variables will require
+more effort, but this effort is warranted given their importance for
+forest C accounting. Although aboveground biomass is the most studied
+variable considered here (Table 2) and is the target of satellite
+missions \citep{refs}, significant ground-based research effort is
+required to create accurate global maps of forest biomass and changes
+therein \citep{refs}. Given observations of increasing tree mortality in
+some forested regions \citep{refs}, including through severe natural
+disturbance \citep{refs}, better characterization of forest dead wood
+will be critical. Additionally, C stocks in forest organic horizons and
+soils can be quite substantial and highly uncertain in many parts of the
+world \citep{refs}. Significant investment in ground-based forest
+research will be critical to filling these gaps.
 
@@
 
-We recommend that, unless they have some specific reason to do
-otherwise, researchers calculate and report the values according to IPCC
-standards (Table 3). It is particularly noteworthy that simple decisions
-on the presentation of results will determine whether the data meet the
-criteria for inclusion in EFDB. Some examples are as follows: (1)
-presenting data only in a figure makes it ineligible for inclusion in
-EFDB, whereas presentation in a table or supplementary data file allows
-inclusion; (2) direct presentation of all relevant variables allows
-inclusion, whereas presenting only components of variables of interest
-(e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or
-requiring simple mathematical operations to obtain a variable of
-interest (e.g., \emph{delta.agb} = \emph{ANPP\_woody} -
-\emph{woody.mortality.agb}) disqualifies data from inclusion; (3)
-matching IPCC-defined thresholds for defining C pools (Table 1), which
-may vary by country, can make the data far more relevant for IPCC
-accounting (e.g., using a 10 cm cutoff between dead wood and litter,
-presenting soil C to a depth of 30 cm). It should also be emphasized
-that reporting of 95\% confidence intervals (or other metrics of error),
-when applicable, is highly desirable and makes the data more relevant to
-IPCC. Reports which had the most successful data transfers used EFDB
-variables and had clear tables showing their results.
+We recommend that, in order to make research valuable to C accounting
+under IPCC guidelines, researchers calculate and report results
+according to IPCC standards (Table 3). It is particularly noteworthy
+that simple decisions on the presentation of results will determine
+whether the records meet the criteria for inclusion in EFDB. Some
+examples are as follows: (1) presenting data only in a figure makes them
+ineligible for inclusion in EFDB, whereas presentation in a table or
+supplementary data file allows inclusion; (2) direct presentation of all
+relevant variables allows inclusion, whereas presenting only components
+of variables of interest (e.g., parsing litter into fine woody debris,
+OL, OF, and OH layers) or requiring simple mathematical operations to
+obtain a variable of interest (e.g., \emph{delta.agb} =
+\emph{ANPP\_woody} - \emph{woody.mortality.agb}) disqualifies records
+from inclusion; (3) matching IPCC-defined thresholds for defining C
+pools (Table 1), which may vary by country, can make the data far more
+relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead
+wood and litter, presenting soil C to a depth of 30 cm). It should also
+be emphasized that reporting of 95\% confidence intervals (or other
+metrics of error), when applicable, is highly desirable and makes the
+data more relevant to IPCC. Reports which had the most successful data
+transfers used EFDB variables and had clear tables showing their
+results.
 
@@
 
-\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance. } ....}
+\caption{\label{tab:table_recommendations}\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}}
 \centering
@@
 \hline
-variables to include & When possible, calculate and present all relevant variables that can be readily caluclated based on available data.
-& Estimates of relevant variables are not always calcualted.\\
+variables to include & When possible, calculate and present all relevant variables that can be readily estimated based on available data. & Estimates of relevant variables are not always calcualted.\\
 \hline
@@
 \hline
- & *taxa to include* & \\
-\hline
-dead organic matter sampling & *include damage estimates on standing dead trees* & \\
-\hline
- & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & \\
-\hline
-belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & IPCC biomass pool definition includes all living vegetation\\
-\hline
-reporting variiables & Present each variable individually, as opposed to requiring that variables of interest be calculated from related variables.
+ & Census all taxa crontributing signficantly to biomass & IPCC biomass pool definition includes all living vegetation.\\
+\hline
+dead organic matter sampling & Adopt IPCC recommendations for minimum diameter of deadwood (country-specific, default 10 cm). & Diameter cutoff must be applied consistently by each country.\\
+\hline
+belowground sampling & Select and report soil sampling increments to include a cutoff at 30 cm depth (or country-specific depth). & Diameter cutoff must be applied consistently by each country.\\
+\hline
+reporting variables & Present each EFDB- relevant variable individually, as opposed to requiring summation of related variables.
 & EFDB requires that values in the database be presented in the original article, and cannot accept subsequent calculations.\\
@@
 \hline
-reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals, when possible.\\
+reporting confidence intervals & Report 95\% confidence intervals, standard error, or standard deviation and sample size. & EFDB requires confidence invervals whenever possible.\\
 \hline
@@
 
-For those compiling published data (e.g., for meta-analyses), the data
-set can have added value if all information required by IPCC is
+For those compiling published records (e.g., for meta-analyses), the
+data set can have added value if all information required by EFDB is
 extracted from original publications. This includes -- but is not
@@
 they may be submitted directly to EFDB or may use the ForC - EFDB data
-pipline developed here. For individual publications, the former option
-will generally be more efficient. However, by getting the data into ForC
-as well as EFDB, the latter option will allow the data to be more
-broadly useful--for example, being used for basic science
-\citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021}
-or model benchmarking \citep{fer_ecosystem_2021}.
+pipeline developed here. For individual publications, the former option
+will generally be more efficient. However, data incorporated into ForC
+as well as EFDB will be more broadly useful; for example, these data may
+be used for basic science
+\citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021},
+analyses of forest-based climate change mititagation potential
+\citep[e.g.,][]{cook-patton_mapping_2020, goldstein_protecting_2020},
+and model benchmarking \citep{fer_ecosystem_2021}.
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
delete mismatches section
Not critical, and I don't feel like writing this! :-)

@@
 section_data_reporting_needs = "6.3"
-section_mismatches = 6.4
 
@@
 Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work (section `r section_database_needs`), new data collection and analysis (section `r section_new_data_needs`), and reporting (section `r section_data_reporting_needs`).
-We also highlight notable mismatches between IPCC accounting methods and forest C mensuration (section `r section_mismatches`).
 
@@
 It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
-Reports which had the most successful data transfers used EFDB variables and had clear tables showing their results.
 
@@
 
-## Mismatches between IPCC accounting methods and forest C mensuration
-
-Remote sensing biomass estimates include standing dead wood [@duncanson_aboveground_2021].
-
-IPCC accounting methods cannot leverage eddy-covariance measurements, which are widely seen as the best available method for quantifying ecosystem-atmosphere gas exchange.
-
-An important challenge is that forests are changing rapidly, and data collected a decade ago may no longer be relevant, particularly in the cases of C increments and fluxes.
-
 # Conclusions
@@
 You can modify the section name with  `\conclusions[modified heading if necessary]`.-->
+
+An important challenge is that forests are changing rapidly, and data collected a decade ago may no longer be relevant, particularly in the cases of C increments and fluxes.
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.tex
delete mismatches section
Not critical, and I don't feel like writing this! :-)

@@
 records in EFDB through database work (section 6.1), new data collection
-and analysis (section 6.2), and reporting (section 6.3). We also
-highlight notable mismatches between IPCC accounting methods and forest
-C mensuration (section 6.4).
+and analysis (section 6.2), and reporting (section 6.3).
 
@@
 metrics of error), when applicable, is highly desirable and makes the
-data more relevant to IPCC. Reports which had the most successful data
-transfers used EFDB variables and had clear tables showing their
-results.
+data more relevant to IPCC.
 
@@
 
-\subsection{Mismatches between IPCC accounting methods and forest C
-mensuration}
-
-Remote sensing biomass estimates include standing dead wood
-\citep{duncanson_aboveground_2021}.
-
-IPCC accounting methods cannot leverage eddy-covariance measurements,
-which are widely seen as the best available method for quantifying
-ecosystem-atmosphere gas exchange.
+\section{Conclusions}
 
@@
 cases of C increments and fluxes.
-
-\section{Conclusions}
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.Rmd
draft conclusions

@@
 This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records that would be relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
 Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
 Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
-ForC represents a diversity of stand ages, although records for young stands are primarily limited to C stocks, as opposed to net increments of fluxes.
 This distribution of records is generally reflected in the subset of records that have been submitted to EFDB to date.
@@
 For individual publications, the former option will generally be more efficient.
-However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mititagation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].
+However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mitigation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].
 
 # Conclusions
-
 <!--The conclusion goes here.
@@
 
-An important challenge is that forests are changing rapidly, and data collected a decade ago may no longer be relevant, particularly in the cases of C increments and fluxes.
+The ForC database contains large numbers of records that could potentially be useful for C accounting under IPCC guidelines.
+Here we have developed a framework for transferring these records to the EFDB, making them more accessible for forest C accounting under IPCC guidelines [@ipcc_2006_2006;@ipcc_2019_2019].
+As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
+Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
+The records transferred to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs.
+`r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.
+
+Going forward, forest researchers can make their research more useful for forest C accounting under IPCC guidelines by calculating and reporting results in ways that are compatible with IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
+In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest C accounting.
+This challenge is heightened by the fact that forests are changing rapidly [@refs], and data collected a decade or more in the past may no longer be relevant.
+This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.
+We view this publication as one step towards that goal.
 

### 2023-05-12 doc/manuscript/ESSDmanuscript.tex
draft conclusions

@@
 (v4.0; described here). As of May 12, 2023, ForC contained
-\textasciitilde19323 independent records that would be relevant to EFDB,
-1438 of which have been submitted to date. Among the data in ForC, there
-is disproportionate representation of biomass (particularly aboveground)
+\textasciitilde19323 independent records relevant to EFDB, 1438 of which
+have been submitted to date. Among the data in ForC, there is
+disproportionate representation of biomass (particularly aboveground)
 stocks, with far fewer records for dead organic matter and soil C, and
@@
 relevant records in temperate forests, and with relatively scant
-representation of tropical forests in Africa and Asia. ForC represents a
-diversity of stand ages, although records for young stands are primarily
-limited to C stocks, as opposed to net increments of fluxes. This
-distribution of records is generally reflected in the subset of records
-that have been submitted to EFDB to date. In the future, forest C
-estimates in EFDB can be improved through targeted research to fill
-critical gaps, reporting of information required by IPCC, and continued
-submission of data from scientific publications to the EFDB. Given that
-climate change is rapidly impacting the world's forests, timely
-reporting of recent estimates will be especially critical to accurate
-forest C accounting.
+representation of tropical forests in Africa and Asia. This distribution
+of records is generally reflected in the subset of records that have
+been submitted to EFDB to date. In the future, forest C estimates in
+EFDB can be improved through targeted research to fill critical gaps,
+reporting of information required by IPCC, and continued submission of
+data from scientific publications to the EFDB. Given that climate change
+is rapidly impacting the world's forests, timely reporting of recent
+estimates will be especially critical to accurate forest C accounting.
 
@@
 \citep[e.g.,][]{banburymorgan_global_2021, anderson-teixeira_carbon_2021},
-analyses of forest-based climate change mititagation potential
+analyses of forest-based climate change mitigation potential
 \citep[e.g.,][]{cook-patton_mapping_2020, goldstein_protecting_2020},
@@
 
-An important challenge is that forests are changing rapidly, and data
-collected a decade ago may no longer be relevant, particularly in the
-cases of C increments and fluxes.
+The ForC database contains large numbers of records that could
+potentially be useful for C accounting under IPCC guidelines. Here we
+have developed a framework for transferring these records to the EFDB,
+making them more accessible for forest C accounting under IPCC
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 12, 2023,
+we have submitted 1438 records to EFDB. Although this represents just
+7\% of relevant records in ForC, it substantially increases the number
+of forest land records in EFDB. The records transferred to EFDB and
+present in ForC are very unevenly distributed across variables, regions,
+and forest types (Figs. 2-3, Table 2), reflecting broader patterns in
+allocation of research effort.
+
+Going forward, forest researchers can make their research more useful
+for forest C accounting under IPCC guidelines by calculating and
+reporting results in ways that are compatible with IPCC guidelines
+(Tables 1, 3). In addition, substantial investments in research and
+researchers focused on ground-based measurement of forest C will be
+required to fill knowledge gaps and thereby increase the accuracy of
+forest C accounting. This challenge is heightened by the fact that
+forests are changing rapidly \citep{refs}, and data collected a decade
+or more in the past may no longer be relevant. This heightens the need
+for an efficient system of making forest C data accessible for national
+greenhouse gas inventories. We view this publication as one step towards
+that goal.
 

### 2023-05-13 doc/manuscript/ESSDmanuscript.Rmd
fill in end statements

@@
   codedata: |
-    use this to add a statement when having data sets and software code available
+    All code and data are openly available.
+The ForC database and associated code is available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD).
+The data and code associated with data transfer to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration).
 authorcontribution: |
-  (fill this in)
+  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data transfer to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC; HML and CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
 disclaimer: |
 acknowledgements: |
-  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have helped to build the database.
-  Funding for this study was provided by Bezos Earth Fund to The Nature Conservancy, the Institute for Global Environmental Strategies, WLS(?)
+  We gratefully acknowledge the substantial contributions of Valentyna Slivinska and Sandro Federici for collaboration on the conception, design, and technical review of this project.
+  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have contributed to building the database.
+  Funding for this study was provided by the Institute for Global Environmental Strategies, a Bezos Earth Fund award to The Nature Conservancy, and a Smithsonian Working Land and Seascapes grant.
 appendix: 
@@
 \clearpage
+
+\clearpage
+
 # Appendix A. Updates to ForC {-}
@@
 \clearpage
+
 # Appendix B. Mapping ForC to EFDB {-}

### 2023-05-13 doc/manuscript/ESSDmanuscript.tex
fill in end statements

@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 12, 2023, ForC contained
+(v4.0; described here). As of May 13, 2023, ForC contained
 \textasciitilde19323 independent records relevant to EFDB, 1438 of which
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 12, 2023, ForC contained 39855 records from 10589 plots in
+As of May 13, 2023, ForC contained 39855 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 (\emph{woody.mortality}). Although most of these variables lacked
-records in ForC as of May 12, 2023, their addition gave the structure
+records in ForC as of May 13, 2023, their addition gave the structure
 such that records can be populated over time. Finally, to provide better
@@
 posting in the database -- a process that lags behind records transfer
-and had not yet been completed for all records sent as of May 12, 2023.
+and had not yet been completed for all records sent as of May 13, 2023.
 
@@
 
-As of May 12, 2023, ForC (v4.0) contained 32693 independent records
+As of May 13, 2023, ForC (v4.0) contained 32693 independent records
 (39855 total), 19323 of which were for the the 43 variables relevant to
@@
 
-As of May 12, 2023, we had reviewed or added 2283 EFDB-relevant records,
+As of May 13, 2023, we had reviewed or added 2283 EFDB-relevant records,
 1438 records of which were sent to EFDB, and 376 of which have been
@@
 making them more accessible for forest C accounting under IPCC
-guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 12, 2023,
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 13, 2023,
 we have submitted 1438 records to EFDB. Although this represents just
@@
 
+\clearpage
+
 \section*{Appendix A. Updates to ForC}
@@
 
-\codedataavailability{use this to add a statement when having data sets
-and software code
-available} %% use this section when having data sets and software code available
+\codedataavailability{All code and data are openly available. The ForC
+database and associated code is available via the ForC repository within
+the ForC-db organization on GitHub (https://github.com/forc-db/ForC),
+and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD). The
+data and code associated with data transfer to EFDB and preparation of
+this manuscript are available via the the IPCC-EFDB-integration
+repository within the ForC-db organization on GitHub
+(https://github.com/forc-db/IPCC-EFDB-integration).} %% use this section when having data sets and software code available
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontribution{(fill this in)} %% optional section
+\authorcontribution{KAT and VH conceived and designed the project; VH
+wrote the scripts for database management, data transfer to EFDB, and
+the analyses presented here; MW, TR, and RBM added and reviewed ForC
+data, BBL and SCP contributed large databases to ForC; HML and CP
+provided methodological expertise; KAT, VH, and MW prepared the first
+draft of the manuscript; all authors reviewed the results and approved
+the final version of the manuscript.} %% optional section
 
@@
 \begin{acknowledgements}
-Thank you to all researchers who collected and published the data
-contained in ForC, and to all research assistants and collaborators who
-have helped to build the database. Funding for this study was provided
-by Bezos Earth Fund to The Nature Conservancy, the Institute for Global
-Environmental Strategies, WLS(?)
+We gratefully acknowledge the substantial contributions of Valentyna
+Slivinska and Sandro Federici for collaboration on the conception,
+design, and technical review of this project. Thank you to all
+researchers who collected and published the data contained in ForC, and
+to all research assistants and collaborators who have contributed to
+building the database. Funding for this study was provided by the
+Institute for Global Environmental Strategies, a Bezos Earth Fund award
+to The Nature Conservancy, and a Smithsonian Working Land and Seascapes
+grant.
 \end{acknowledgements}

### 2023-05-15 doc/manuscript/ESSDmanuscript.Rmd
cleanup, adding refs
#39

@@
 and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@refs], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@refs; @delima_making_2022], and *tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on previously existing data [@ref].* <!-- I'm almost certain this is true, but need to confirm/ find ref.-->
+The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@refs], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@refs; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
 
@@
 Future efforts to review studies for transfer should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), *to records from countries relying on existing records (as opposed to regular collection of new data through forest inventories)*, to the variables most needed by EFDB users, and to the more contemporary records.
+`r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
 
@@
 New data collection and analysis is needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021], production of an accurate global map of forest C stocks remains an ongoing challenge [@refs].
+While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021], production of an accurate global map of forest C stocks remains an ongoing challenge [@refs; @labrière_forest_2023].
 Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@refs].
@@
 Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for forest C accounting.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@refs], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@refs].
+Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
 Given observations of increasing tree mortality in some forested regions [@refs], including through severe natural disturbance [@refs], better characterization of forest dead wood will be critical.
@@
 
-\clearpage
-
 # Appendix A. Updates to ForC {-}

### 2023-05-15 doc/manuscript/ESSDmanuscript.tex
cleanup, adding refs
#39

@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 13, 2023, ForC contained
+(v4.0; described here). As of May 15, 2023, ForC contained
 \textasciitilde19323 independent records relevant to EFDB, 1438 of which
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 13, 2023, ForC contained 39855 records from 10589 plots in
+As of May 15, 2023, ForC contained 39855 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 (\emph{woody.mortality}). Although most of these variables lacked
-records in ForC as of May 13, 2023, their addition gave the structure
+records in ForC as of May 15, 2023, their addition gave the structure
 such that records can be populated over time. Finally, to provide better
@@
 tropical forest C cycling tends to be more scarce due to a variety of
-challenges \citep{refs, delima_making_2022}, and \emph{tropical
-countries are more likely to apply Tier 1 methodology that bases forest
-C budgets on previously existing data \citep{ref}.}
+challenges \citep{refs, delima_making_2022}, and tropical countries are
+more likely to apply Tier 1 methodology that bases forest C budgets on
+internationally defined IPCC default values
+\citep{romijn_assessing_2015}.
 
@@
 posting in the database -- a process that lags behind records transfer
-and had not yet been completed for all records sent as of May 13, 2023.
+and had not yet been completed for all records sent as of May 15, 2023.
 
@@
 
-As of May 13, 2023, ForC (v4.0) contained 32693 independent records
+As of May 15, 2023, ForC (v4.0) contained 32693 independent records
 (39855 total), 19323 of which were for the the 43 variables relevant to
@@
 
-As of May 13, 2023, we had reviewed or added 2283 EFDB-relevant records,
+As of May 15, 2023, we had reviewed or added 2283 EFDB-relevant records,
 1438 records of which were sent to EFDB, and 376 of which have been
@@
 types, and variables, giving priority to those from currently
-under-represented regions and forest types (Figs. 2-3, Table 2),
-\emph{to records from countries relying on existing records (as opposed
-to regular collection of new data through forest inventories)}, to the
-variables most needed by EFDB users, and to the more contemporary
-records.
+under-represented regions and forest types (Figs. 2-3, Table 2), to
+records from countries relying on existing data for their greenhouse gas
+inventories (Tier 1 or 2 methodology), to the variables most needed by
+EFDB users, and to the more contemporary records.
 
@@
 accurate global map of forest C stocks remains an ongoing challenge
-\citep{refs}. Other pools and variables remain very poorly quantified
-\citep[Table 2,][]{anderson-teixeira_carbon_2021}, introducing
-substantive uncertainties into global forest C budgets \citep{refs}.
-Furthermore, data distribution is uneven across forest types and
-geographical regions (Figs. 2-3). For instance, data on C cycling of
-tropical forests -- particularly in Africa -- remains relatively sparse,
-in large part due to substantial barriers to data collection and
-distribution \citep{delima_making_2022}. Significant investment in
-research and researchers focused on ground-based measurement of forest C
-in such regions will be important to filling knowledge gaps in forest C
-cycling \citep{delima_making_2022, refs}.
+\citep{refs, labrière_forest_2023}. Other pools and variables remain
+very poorly quantified \citep[Table
+2,][]{anderson-teixeira_carbon_2021}, introducing substantive
+uncertainties into global forest C budgets \citep{refs}. Furthermore,
+data distribution is uneven across forest types and geographical regions
+(Figs. 2-3). For instance, data on C cycling of tropical forests --
+particularly in Africa -- remains relatively sparse, in large part due
+to substantial barriers to data collection and distribution
+\citep{delima_making_2022}. Significant investment in research and
+researchers focused on ground-based measurement of forest C in such
+regions will be important to filling knowledge gaps in forest C cycling
+\citep{delima_making_2022, refs}.
 
@@
 variable considered here (Table 2) and is the target of satellite
-missions \citep{refs}, significant ground-based research effort is
-required to create accurate global maps of forest biomass and changes
-therein \citep{refs}. Given observations of increasing tree mortality in
-some forested regions \citep{refs}, including through severe natural
-disturbance \citep{refs}, better characterization of forest dead wood
-will be critical. Additionally, C stocks in forest organic horizons and
-soils can be quite substantial and highly uncertain in many parts of the
-world \citep{refs}. Significant investment in ground-based forest
-research will be critical to filling these gaps.
+missions
+\citep{dubayah_global_2020, quegan_european_2019, nisar_nasaisro_2018},
+significant ground-based research effort is required to create accurate
+global maps of forest biomass and changes therein
+\citep{duncanson_importance_2019, labrière_forest_2023}. Given
+observations of increasing tree mortality in some forested regions
+\citep{refs}, including through severe natural disturbance \citep{refs},
+better characterization of forest dead wood will be critical.
+Additionally, C stocks in forest organic horizons and soils can be quite
+substantial and highly uncertain in many parts of the world
+\citep{refs}. Significant investment in ground-based forest research
+will be critical to filling these gaps.
 
@@
 making them more accessible for forest C accounting under IPCC
-guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 13, 2023,
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 15, 2023,
 we have submitted 1438 records to EFDB. Although this represents just
@@
 
-\clearpage
-
 \section*{Appendix A. Updates to ForC}

### 2023-05-15 doc/manuscript/ESSDmanuscript.Rmd
refs
#39

@@
 A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
-In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs_from_ForC], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs_from_ForC].
+In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @refs_from_ForC], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs_from_ForC].
 While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although thhese do tend to be censused when they consitute a large proportion of the biomass [@refs_from_ForC].
@@
 New data collection and analysis is needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021], production of an accurate global map of forest C stocks remains an ongoing challenge [@refs; @labrière_forest_2023].
-Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@refs].
+While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
+Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
 Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
 For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
-Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@refs].
+Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labrière_forest_2023].
 
@@
 Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
-Given observations of increasing tree mortality in some forested regions [@refs], including through severe natural disturbance [@refs], better characterization of forest dead wood will be critical.
-Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@refs].
+Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
+Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
 Significant investment in ground-based forest research will be critical to filling these gaps.
@@
 In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest C accounting.
-This challenge is heightened by the fact that forests are changing rapidly [@refs], and data collected a decade or more in the past may no longer be relevant.
+This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may no longer be relevant.
 This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.

### 2023-05-15 doc/manuscript/ESSDmanuscript.tex
refs
#39

@@
 savannas, small trees and shrubs contribute a much larger proportion of
-C stocks and fluxes \citep{piponiot_distribution_2022, refs_from_ForC},
+C stocks and fluxes
+\citep{lutz_global_2018, piponiot_distribution_2022, refs_from_ForC},
 and, correspondingly, biomass estimates for these ecosystems tend
@@
 and continue to receive -- by far the most research attention
-\citep[Table 2,][]{anderson-teixeira_carbon_2021}, production of an
-accurate global map of forest C stocks remains an ongoing challenge
-\citep{refs, labrière_forest_2023}. Other pools and variables remain
-very poorly quantified \citep[Table
+\citep[Table
+2,][]{anderson-teixeira_carbon_2021, dubayah_global_2020, quegan_european_2019, nisar_nasaisro_2018},
+production of an accurate global map of forest C stocks remains an
+ongoing challenge \citep{araza_decade_2023}. Other pools and variables
+remain very poorly quantified \citep[Table
 2,][]{anderson-teixeira_carbon_2021}, introducing substantive
-uncertainties into global forest C budgets \citep{refs}. Furthermore,
-data distribution is uneven across forest types and geographical regions
+uncertainties into global forest C budgets
+\citep{pan_large_2011, harris_global_2021}. Furthermore, data
+distribution is uneven across forest types and geographical regions
 (Figs. 2-3). For instance, data on C cycling of tropical forests --
@@
 regions will be important to filling knowledge gaps in forest C cycling
-\citep{delima_making_2022, refs}.
+\citep{delima_making_2022, araza_decade_2023, labrière_forest_2023}.
 
@@
 observations of increasing tree mortality in some forested regions
-\citep{refs}, including through severe natural disturbance \citep{refs},
-better characterization of forest dead wood will be critical.
-Additionally, C stocks in forest organic horizons and soils can be quite
-substantial and highly uncertain in many parts of the world
-\citep{refs}. Significant investment in ground-based forest research
-will be critical to filling these gaps.
+\citep{mcdowell_pervasive_2020}, better characterization of forest dead
+wood will be critical. Additionally, C stocks in forest organic horizons
+and soils can be quite substantial and highly uncertain in many parts of
+the world \citep{tifafi_large_2018}. Significant investment in
+ground-based forest research will be critical to filling these gaps.
 
@@
 forest C accounting. This challenge is heightened by the fact that
-forests are changing rapidly \citep{refs}, and data collected a decade
-or more in the past may no longer be relevant. This heightens the need
-for an efficient system of making forest C data accessible for national
-greenhouse gas inventories. We view this publication as one step towards
-that goal.
+forests are changing rapidly \citep[e.g.,][]{mcdowell_pervasive_2020},
+and data collected a decade or more in the past may no longer be
+relevant. This heightens the need for an efficient system of making
+forest C data accessible for national greenhouse gas inventories. We
+view this publication as one step towards that goal.
 

### 2023-05-15 doc/manuscript/ESSDmanuscript.Rmd
minor clean-up
#39

@@
 A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
-In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @refs_from_ForC], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs_from_ForC].
-While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although thhese do tend to be censused when they consitute a large proportion of the biomass [@refs_from_ForC].
+In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.
+While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [@refs_from_ForC].
 Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

### 2023-05-15 doc/manuscript/ESSDmanuscript.Rmd
Helene declined authorship

@@
     affiliation: "5" 
-  - given_name: Helene
-    surname: Muller-Landau
-    affiliation: "2" 
   - given_name: Camille
@@
 authorcontribution: |
-  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data transfer to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC; HML and CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
+  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data transfer to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC; CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
 disclaimer: |

### 2023-05-15 doc/manuscript/ESSDmanuscript.tex
Helene declined authorship

@@
 \Author[5]{Susan}{Cook-Patton}
-\Author[2]{Helene}{Muller-Landau}
 \Author[6]{Camille}{Piponiot}
@@
 C stocks and fluxes
-\citep{lutz_global_2018, piponiot_distribution_2022, refs_from_ForC},
+\citep{lutz_global_2018, piponiot_distribution_2022, hughes_biomass_1999},
 and, correspondingly, biomass estimates for these ecosystems tend
-include smaller size classes \citep[e.g.,][]{refs_from_ForC}. While IPCC
-guidance specifies that all living vegetation should be included in
-biomass estimates, forest censuses and biomass estimates do not
-consistently include life forms other than dicot trees (e.g., lianas,
-ferns, palms, bamboo), although thhese do tend to be censused when they
-consitute a large proportion of the biomass \citep{refs_from_ForC}.
-Further, it is important to note that the IPCC definition of aboveground
-biomass excludes standing dead wood, which is included in remote sensing
-biomass estimates \citep{duncanson_aboveground_2021}.
+include smaller size classes. While IPCC guidance specifies that all
+living vegetation should be included in biomass estimates, forest
+censuses and biomass estimates do not consistently include life forms
+other than dicot trees (e.g., lianas, ferns, palms, bamboo), although
+these do tend to be censused when they constitute a large proportion of
+the biomass \citep{refs_from_ForC}. Further, it is important to note
+that the IPCC definition of aboveground biomass excludes standing dead
+wood, which is included in remote sensing biomass estimates
+\citep{duncanson_aboveground_2021}.
 
@@
 the analyses presented here; MW, TR, and RBM added and reviewed ForC
-data, BBL and SCP contributed large databases to ForC; HML and CP
-provided methodological expertise; KAT, VH, and MW prepared the first
-draft of the manuscript; all authors reviewed the results and approved
-the final version of the manuscript.} %% optional section
+data, BBL and SCP contributed large databases to ForC; CP provided
+methodological expertise; KAT, VH, and MW prepared the first draft of
+the manuscript; all authors reviewed the results and approved the final
+version of the manuscript.} %% optional section
 

### 2023-05-15 doc/manuscript/ESSDmanuscript.Rmd
add climwin version
#39

@@
 Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
-Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version **##** <!--need this from Becky--> [@climwin_ref] based on site coordinates. 
+Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version 2.1 [@vandepol_identifying_2016; @bailey_climwin_2016] based on site coordinates. 
 Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or transfer to EFDB.

### 2023-05-16 doc/manuscript/ESSDmanuscript.Rmd
add ref, incorporate small ForC updates

@@
 In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.
-While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [@refs_from_ForC].
+While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [e.g., @fukushima_recovery_2007].
 Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

### 2023-05-16 doc/manuscript/ESSDmanuscript.tex
add ref, incorporate small ForC updates

@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 15, 2023, ForC contained
+(v4.0; described here). As of May 16, 2023, ForC contained
 \textasciitilde19323 independent records relevant to EFDB, 1438 of which
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 15, 2023, ForC contained 39855 records from 10589 plots in
+As of May 16, 2023, ForC contained 39848 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 these do tend to be censused when they constitute a large proportion of
-the biomass \citep{refs_from_ForC}. Further, it is important to note
-that the IPCC definition of aboveground biomass excludes standing dead
-wood, which is included in remote sensing biomass estimates
-\citep{duncanson_aboveground_2021}.
+the biomass \citep[e.g.,][]{fukushima_recovery_2007}. Further, it is
+important to note that the IPCC definition of aboveground biomass
+excludes standing dead wood, which is included in remote sensing biomass
+estimates \citep{duncanson_aboveground_2021}.
 
@@
 (\emph{woody.mortality}). Although most of these variables lacked
-records in ForC as of May 15, 2023, their addition gave the structure
+records in ForC as of May 16, 2023, their addition gave the structure
 such that records can be populated over time. Finally, to provide better
@@
 Third, to identify erroneous climate data, we compared ForC climate
-values to those extracted from WorldClim version \textbf{\#\#}
-\citep{climwin_ref} based on site coordinates. Records deviating from
-WorldClim values by more variable-specific thresholds (\textgreater5°C
-for mean annual temperature, \textgreater7.5°C for mean temperatures of
-the warmest and coldest months, or \textgreater1 for log(mean annual
-precipitation in mm)) were flagged as requiring review prior to use in
-analysis or transfer to EFDB.
+values to those extracted from WorldClim version 2.1
+\citep{vandepol_identifying_2016, bailey_climwin_2016} based on site
+coordinates. Records deviating from WorldClim values by more
+variable-specific thresholds (\textgreater5°C for mean annual
+temperature, \textgreater7.5°C for mean temperatures of the warmest and
+coldest months, or \textgreater1 for log(mean annual precipitation in
+mm)) were flagged as requiring review prior to use in analysis or
+transfer to EFDB.
 
@@
 posting in the database -- a process that lags behind records transfer
-and had not yet been completed for all records sent as of May 15, 2023.
+and had not yet been completed for all records sent as of May 16, 2023.
 
@@
 
-As of May 15, 2023, ForC (v4.0) contained 32693 independent records
-(39855 total), 19323 of which were for the the 43 variables relevant to
+As of May 16, 2023, ForC (v4.0) contained 32693 independent records
+(39848 total), 19323 of which were for the the 43 variables relevant to
 EFDB (Fig. 1). These records were distributed across all forested
@@
 
-As of May 15, 2023, we had reviewed or added 2283 EFDB-relevant records,
+As of May 16, 2023, we had reviewed or added 2283 EFDB-relevant records,
 1438 records of which were sent to EFDB, and 376 of which have been
@@
 making them more accessible for forest C accounting under IPCC
-guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 15, 2023,
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 16, 2023,
 we have submitted 1438 records to EFDB. Although this represents just

### 2023-05-23 doc/manuscript/ESSDmanuscript.Rmd
incorporate Sandro's comments (send in Word doc)
plus  a few other changes, including title

#39

@@
 ---
-title: Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)
+title: Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
 journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
@@
 **Abstract.**
-Forests are critical for climate change mitigation and constitute a substantial portion of planned emissions reductions under the 2015 Paris Agreement.
+Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
 Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
@@
 Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-Here, we develop and document a process for semi-automated transfer of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhoouse gas emissions and removals.
+Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhouse gas emissions and removals.
 We begin by reconciling terminology and mapping ForC fields into EFDB.
@@
 
-Forests are critical to management of atmospheric concentrations of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
+Forests are critical to management of the atmospheric concentration of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
 In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
@@
 Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
+Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
 Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and on human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
@@
 A tiered approach to accounting is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
-Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances.
+Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances and datasets grow and become more widely accessible.
 For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default values underestimated C sequestration by 32% on average and failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
@@
 
-To improve data accessibility for C accounting, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
+To improve data accessibility for preparing greenhouse gas estimates, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
 The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
@@
 As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
-As such, ForC is positioned to improve forest C accounting through the transfer of data to EFDB.
+As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
 The purpose of this publication is to document that process and provide recommendations for future improvements.
@@
 Here, we
-(1) review IPCC methods and definitions for forest C accounting in the context of typical forest C estimation methodologies;
+(1) review IPCC methods and definitions applied to estimate CO2 emissions and removals from forest in the context of typical forest C estimation methodologies;
 (2) describe mapping of ForC to IPCC's EFDB;
 (3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data transfer to EFDB;
-(4) summarize the data in ForC that's relevant to EFDB and records that have been transferred to date; and
+(4) summarize the data in ForC that's relevant to EFDB and records that have been submitted to date; and
 (5) provide recommendations for improving data collection, analysis, database, and accounting.
@@
 
-Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
+Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
 As noted above, live fine roots may be included in litter when difficult to separate empirically.
@@
 
-Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C accounting under IPCC guidelines.
-However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C accounting under IPCC guidelines.
+Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement.
+However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C inventories under IPCC guidelines.
 To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
@@
 
-# Transfer of data from ForC to EFDB
+# Submission of ForC data to EFDB
 
-To transfer complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data transfer were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been sent to EFDB.
+To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
+Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been sent to EFDB.
 Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for inclusion in EFDB. 
@@
 As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
-It it important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
+It is important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
 For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
@@
 For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
-For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Cropland converted to Forest Land", respectively.
+For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Grassland converted to Forest Land", respectively.
 For forests <20 years old with unspecified previous agricultural use, we assigned the sub-category "Land Converted to Forest land".
@@
 
-Given the lack of public information needed to determine whether lands are classified as mangaged [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not transfer any classification of land management status from ForC to the EFDB.
+Given the lack of public information needed to determine whether lands are classified as managed [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not transfer any classification of land management status from ForC to the EFDB.
 However, we do provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the the `r n_relevant_variables` variables relevant to EFDB (Fig.
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig.
 `r fig_variable_mapping`).
@@
 
-## Data transfers to EFDB
+## Data submissions to EFDB
 
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were sent to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
+As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were submitted to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
 `r fig_map`-`r fig_histograms`, Table `r table_variables`).
@@
 
-There is vast potential to expand forest C data in EFDB by completing the process of reviewing and sending data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
-So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been sent to EFDB.
-Although this process requires manual review of records, the addition of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily transferred to EFDB using the system developed here.
+There is vast potential to expand forest C data in EFDB by completing the process of reviewing and submitting data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
+So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been submitted to EFDB.
+Although this process requires manual review of records, the submission of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
 Future efforts to review studies for transfer should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
@@
 
-Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for forest C accounting.
+Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
 Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
@@
 
-We recommend that, in order to make research valuable to C accounting under IPCC guidelines, researchers calculate and report results according to IPCC standards (Table `r table_recommendations`).
+We recommend that, in order to make research valuable to estimate C stock changes according to methods provided in the IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
 It is particularly noteworthy that simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
@@
 
-The ForC database contains large numbers of records that could potentially be useful for C accounting under IPCC guidelines.
-Here we have developed a framework for transferring these records to the EFDB, making them more accessible for forest C accounting under IPCC guidelines [@ipcc_2006_2006;@ipcc_2019_2019].
+The ForC database contains large numbers of records that could potentially be useful for estimating C stock changes applying methodological guidance provided by the IPCC.
+Here we have developed a framework for submitting these records to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
 As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
 Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
-The records transferred to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs.
+The records submitted to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs.
 `r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.
 
-Going forward, forest researchers can make their research more useful for forest C accounting under IPCC guidelines by calculating and reporting results in ways that are compatible with IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
-In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest C accounting.
+Going forward, forest researchers can make their research more useful for forest C accounting under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
+In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under teh Paris Agreement.
 This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may no longer be relevant.
 This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.
-We view this publication as one step towards that goal.
+We view the system developed here for submitting ForC data to the IPCC EFDB as one important step towards that goal.
 

### 2023-05-23 doc/manuscript/ESSDmanuscript.tex
incorporate Sandro's comments (send in Word doc)
plus  a few other changes, including title

#39

@@
 
-\title{Informing IPCC accounting of forest carbon using the global
-forest carbon database (ForC v4.0)}
+\title{Informing forest carbon inventories under the Paris Agreement
+using the Global Forest Carbon Database (ForC v4.0)}
 
@@
 \textbf{Abstract.} Forests are critical for climate change mitigation
-and constitute a substantial portion of planned emissions reductions
+and constitute a substantial portion of planned net emissions reductions
 under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
@@
 of the International Panel on Climate Change (IPCC). Here, we develop
-and document a process for semi-automated transfer of data from ForC
+and document a process for semi-automated submission of data from ForC
 into the EFDB, assess the data available and transferred to date, and
 provide recommendations for improving forest data collection, analysis,
-and reporting to improve accounting of forest-sector greenhoouse gas
+and reporting to improve accounting of forest-sector greenhouse gas
 emissions and removals. We begin by reconciling terminology and mapping
@@
 database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 16, 2023, ForC contained
+(v4.0; described here). As of May 23, 2023, ForC contained
 \textasciitilde19323 independent records relevant to EFDB, 1438 of which
@@
 
-Forests are critical to management of atmospheric concentrations of the
-greenhouse gas carbon dioxide (CO\textsubscript{2}), and thereby climate
-change. In recent decades, CO\textsubscript{2} uptake by forests,
-woodlands, and savannas has exceeded releases from deforestation and
-other severe disturbances, resulting in a net carbon CO\textsubscript{2}
-sink of \textasciitilde0.88 Gt C yr\textsuperscript{-1} \citep[all
-biomes with trees,][]{xu_changes_2021} to \textasciitilde1.6 Gt C
-yr\textsuperscript{-1} \citep[forests only,][]{harris_global_2021}. This
-has offset an estimated 10\% to 18\% of anthropogenic
-CO\textsubscript{2} emissions from fossil fuels and cement
-\citep{xu_changes_2021, harris_global_2021}, dramatically slowing the
-pace of atmospheric CO\textsubscript{2} accumulation and climate change.
-Going into the future, the fate of this important CO\textsubscript{2}
-sink is highly uncertain, depending both upon forest responses to
-climate change, which are likely to reduce the sink strength
-\citep{mcdowell_pervasive_2020, hammond_global_2022}, and on human
-conservation, restoration, and management of forests
+Forests are critical to management of the atmospheric concentration of
+the greenhouse gas carbon dioxide (CO\textsubscript{2}), and thereby
+climate change. In recent decades, CO\textsubscript{2} uptake by
+forests, woodlands, and savannas has exceeded releases from
+deforestation and other severe disturbances, resulting in a net carbon
+CO\textsubscript{2} sink of \textasciitilde0.88 Gt C
+yr\textsuperscript{-1} \citep[all biomes with trees,][]{xu_changes_2021}
+to \textasciitilde1.6 Gt C yr\textsuperscript{-1} \citep[forests
+only,][]{harris_global_2021}. This has offset an estimated 10\% to 18\%
+of anthropogenic CO\textsubscript{2} emissions from fossil fuels and
+cement \citep{xu_changes_2021, harris_global_2021}, dramatically slowing
+the pace of atmospheric CO\textsubscript{2} accumulation and associated
+climate change. Going into the future, the fate of this important
+CO\textsubscript{2} sink is highly uncertain, depending both upon forest
+responses to climate change, which are likely to reduce the sink
+strength \citep{mcdowell_pervasive_2020, hammond_global_2022}, and on
+human conservation, restoration, and management of forests
 \citep{ipcc_climate_2019, ipcc_climate_2022}.
@@
 \citep{requenasuarez_estimating_2019, rozendaal_aboveground_2022}, but
-there remains room for continuous improvement as the science advances.
-For example, the year following the release of the latest IPCC
-guidelines, a more thorough analysis of C accumulation in regrowth
-forests found that IPCC's Tier 1 default values underestimated C
-sequestration by 32\% on average and failed to capture eight-fold
-variation within ecozones \citep{cook-patton_mapping_2020}. In addition,
-it was revealed that C stocks in mature African tropical montane forests
-were two-thirds higher than the IPCC Tier 1 values for these forests
+there remains room for continuous improvement as the science advances
+and datasets grow and become more widely accessible. For example, the
+year following the release of the latest IPCC guidelines, a more
+thorough analysis of C accumulation in regrowth forests found that
+IPCC's Tier 1 default values underestimated C sequestration by 32\% on
+average and failed to capture eight-fold variation within ecozones
+\citep{cook-patton_mapping_2020}. In addition, it was revealed that C
+stocks in mature African tropical montane forests were two-thirds higher
+than the IPCC Tier 1 values for these forests
 \citep{cuni-sanchez_high_2021}. This rapid evolution of scientific
@@
 
-To improve data accessibility for C accounting, the IPCC created the
-Emission Factor Database (EFDB;
+To improve data accessibility for preparing greenhouse gas estimates,
+the IPCC created the Emission Factor Database (EFDB;
 \url{https://www.ipcc-nggip.iges.or.jp/EFDB/main.php}), which is
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 16, 2023, ForC contained 39848 records from 10589 plots in
+As of May 23, 2023, ForC contained 39848 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
-disturbance history. As such, ForC is positioned to improve forest C
-accounting through the transfer of data to EFDB. The purpose of this
+disturbance history. As such, ForC is positioned to improve
+forest-related estimates of CO\textsubscript{2} emissions and removals
+through the submission of data to the EFDB. The purpose of this
 publication is to document that process and provide recommendations for
@@
 
-Here, we (1) review IPCC methods and definitions for forest C accounting
-in the context of typical forest C estimation methodologies; (2)
-describe mapping of ForC to IPCC's EFDB; (3) describe updates to ForC
-(ForC v4.0), most of which were implemented to facilitate data transfer
-to EFDB; (4) summarize the data in ForC that's relevant to EFDB and
-records that have been transferred to date; and (5) provide
-recommendations for improving data collection, analysis, database, and
-accounting.
+Here, we (1) review IPCC methods and definitions applied to estimate CO2
+emissions and removals from forest in the context of typical forest C
+estimation methodologies; (2) describe mapping of ForC to IPCC's EFDB;
+(3) describe updates to ForC (ForC v4.0), most of which were implemented
+to facilitate data transfer to EFDB; (4) summarize the data in ForC
+that's relevant to EFDB and records that have been submitted to date;
+and (5) provide recommendations for improving data collection, analysis,
+database, and accounting.
 
@@
 can exceed 100 t C ha\textsuperscript{-1}
-\citep{anderson-teixeira_carbon_2021}, is defined by IPCCC as including
+\citep{anderson-teixeira_carbon_2021}, is defined by IPCC as including
 ``all non-living biomass with a diameter less than a minimum diameter
@@
 contained most of the information required by EFDB, and, more broadly,
-to inform C accounting under IPCC guidelines. However, modest changes to
-the structure and contents of ForC were needed in order to provide all
-information required by EFDB and to improve ForC's capacity to serve as
-a repository of valuable information for forest C accounting under IPCC
-guidelines. To support export of data to EFDB, and to improve the
-overall quality of the ForC database, we added or modified 18 fields
-(Appendix A), defined 15 new variables, implemented enhanced quality
-control, manually reviewed \textgreater1954 records to obtain additional
-required information, and added 329 new records.
+to inform C stock change calculations under the Paris Agreement.
+However, modest changes to the structure and contents of ForC were
+needed in order to provide all information required by EFDB and to
+improve ForC's capacity to serve as a repository of valuable information
+for forest C inventories under IPCC guidelines. To support export of
+data to EFDB, and to improve the overall quality of the ForC database,
+we added or modified 18 fields (Appendix A), defined 15 new variables,
+implemented enhanced quality control, manually reviewed \textgreater1954
+records to obtain additional required information, and added 329 new
+records.
 
@@
 (\emph{woody.mortality}). Although most of these variables lacked
-records in ForC as of May 16, 2023, their addition gave the structure
+records in ForC as of May 23, 2023, their addition gave the structure
 such that records can be populated over time. Finally, to provide better
@@
 
-\section{Transfer of data from ForC to EFDB}
-
-To transfer complete, reviewed ForC records into EFDB, we created R
+\section{Submission of ForC data to EFDB}
+
+To submit complete, reviewed ForC records into EFDB, we created R
 scripts to restructure ForC records and populate EFDB's bulk import form
-(``EFDB bulk import.xlsx''). Criteria for data transfer were that (1)
+(``EFDB bulk import.xlsx''). Criteria for data submission were that (1)
 records had been checked against the original study and determined to be
@@
 posting in the database -- a process that lags behind records transfer
-and had not yet been completed for all records sent as of May 16, 2023.
+and had not yet been completed for all records sent as of May 23, 2023.
 
@@
 also defined 15 new EFDB-relevant variables that were not previously
-represented in ForC. It it important to note that the correspondence of
+represented in ForC. It is important to note that the correspondence of
 ForC variables to IPCC criteria often depends upon measurement protocols
@@
 Cropland and Grassland, respectively, making land-use subcategories were
-``Cropland converted to Forest Land'' and ``Cropland converted to Forest
-Land'', respectively. For forests \textless20 years old with unspecified
-previous agricultural use, we assigned the sub-category ``Land Converted
-to Forest land''. Forests \textless20 years old with unknown land use
-prior to the study date were simply classified as ``Forest Land''. The
-same logic was applied for savannas, but including both Forest Land and
-Grassland as potentially relevant categories.
+``Cropland converted to Forest Land'' and ``Grassland converted to
+Forest Land'', respectively. For forests \textless20 years old with
+unspecified previous agricultural use, we assigned the sub-category
+``Land Converted to Forest land''. Forests \textless20 years old with
+unknown land use prior to the study date were simply classified as
+``Forest Land''. The same logic was applied for savannas, but including
+both Forest Land and Grassland as potentially relevant categories.
 
 Given the lack of public information needed to determine whether lands
-are classified as mangaged
+are classified as managed
 \citep{ogle_delineating_2018, deng_comparing_2021}, and because the
@@
 
-As of May 16, 2023, ForC (v4.0) contained 32693 independent records
-(39848 total), 19323 of which were for the the 43 variables relevant to
-EFDB (Fig. 1). These records were distributed across all forested
-continents and ecozones, albeit unevenly (Fig. 2). The largest number of
-records came from Asia, followed by North America, South America, and
-Europe, with relatively few records from Africa, Australia, and Oceania
-(Fig. 3c). Categorized by FAO ecozone, the greatest numbers of records
-came from subtropical humid forests, temperate mountain systems, and
-tropical rain forests, each with \textgreater2,000 independent records
-(Fig. 3b). Boreal coniferous forests, temperate continental forests,
-subtropical mountain systems, and tropical moist deciduous forests had
+As of May 23, 2023, ForC (v4.0) contained 32693 independent records
+(39848 total), 19323 of which were for the 43 variables relevant to EFDB
+(Fig. 1). These records were distributed across all forested continents
+and ecozones, albeit unevenly (Fig. 2). The largest number of records
+came from Asia, followed by North America, South America, and Europe,
+with relatively few records from Africa, Australia, and Oceania (Fig.
+3c). Categorized by FAO ecozone, the greatest numbers of records came
+from subtropical humid forests, temperate mountain systems, and tropical
+rain forests, each with \textgreater2,000 independent records (Fig. 3b).
+Boreal coniferous forests, temperate continental forests, subtropical
+mountain systems, and tropical moist deciduous forests had
 \textgreater1,000 independent records each, while other ecozones all had
@@
 
-\subsection{Data transfers to EFDB}
-
-As of May 16, 2023, we had reviewed or added 2283 EFDB-relevant records,
-1438 records of which were sent to EFDB, and 376 of which have been
+\subsection{Data submissions to EFDB}
+
+As of May 23, 2023, we had reviewed or added 2283 EFDB-relevant records,
+1438 records of which were submitted to EFDB, and 376 of which have been
 reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
@@
 There is vast potential to expand forest C data in EFDB by completing
-the process of reviewing and sending data that are already in ForC
+the process of reviewing and submitting data that are already in ForC
 (Figs. 2-3). So far, only \textasciitilde7\% of the EFDB-relevant data
-in ForC have been sent to EFDB. Although this process requires manual
-review of records, the addition of new records to EFDB is hugely
-facilitated by the fact that most pertinent information for each record
-is already entered in ForC and can be easily transferred to EFDB using
-the system developed here. Future efforts to review studies for transfer
-should optimize for representation across geographic regions, forest
-types, and variables, giving priority to those from currently
-under-represented regions and forest types (Figs. 2-3, Table 2), to
-records from countries relying on existing data for their greenhouse gas
-inventories (Tier 1 or 2 methodology), to the variables most needed by
-EFDB users, and to the more contemporary records.
+in ForC have been submitted to EFDB. Although this process requires
+manual review of records, the submission of new records to EFDB is
+hugely facilitated by the fact that most pertinent information for each
+record is already entered in ForC and can be easily prepared for
+submission to EFDB using the system developed here. Future efforts to
+review studies for transfer should optimize for representation across
+geographic regions, forest types, and variables, giving priority to
+those from currently under-represented regions and forest types (Figs.
+2-3, Table 2), to records from countries relying on existing data for
+their greenhouse gas inventories (Tier 1 or 2 methodology), to the
+variables most needed by EFDB users, and to the more contemporary
+records.
 
@@
 more effort, but this effort is warranted given their importance for
-forest C accounting. Although aboveground biomass is the most studied
-variable considered here (Table 2) and is the target of satellite
-missions
+estimating forest C stock chnages. Although aboveground biomass is the
+most studied variable considered here (Table 2) and is the target of
+satellite missions
 \citep{dubayah_global_2020, quegan_european_2019, nisar_nasaisro_2018},
@@
 
-We recommend that, in order to make research valuable to C accounting
-under IPCC guidelines, researchers calculate and report results
-according to IPCC standards (Table 3). It is particularly noteworthy
-that simple decisions on the presentation of results will determine
-whether the records meet the criteria for inclusion in EFDB. Some
-examples are as follows: (1) presenting data only in a figure makes them
-ineligible for inclusion in EFDB, whereas presentation in a table or
-supplementary data file allows inclusion; (2) direct presentation of all
-relevant variables allows inclusion, whereas presenting only components
-of variables of interest (e.g., parsing litter into fine woody debris,
-OL, OF, and OH layers) or requiring simple mathematical operations to
-obtain a variable of interest (e.g., \emph{delta.agb} =
-\emph{ANPP\_woody} - \emph{woody.mortality.agb}) disqualifies records
-from inclusion; (3) matching IPCC-defined thresholds for defining C
-pools (Table 1), which may vary by country, can make the data far more
-relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead
-wood and litter, presenting soil C to a depth of 30 cm). It should also
-be emphasized that reporting of 95\% confidence intervals (or other
-metrics of error), when applicable, is highly desirable and makes the
-data more relevant to IPCC.
+We recommend that, in order to make research valuable to estimate C
+stock changes according to methods provided in the IPCC guidelines,
+researchers calculate and report results according to IPCC good practice
+(Table 3). It is particularly noteworthy that simple decisions on the
+presentation of results will determine whether the records meet the
+criteria for inclusion in EFDB. Some examples are as follows: (1)
+presenting data only in a figure makes them ineligible for inclusion in
+EFDB, whereas presentation in a table or supplementary data file allows
+inclusion; (2) direct presentation of all relevant variables allows
+inclusion, whereas presenting only components of variables of interest
+(e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or
+requiring simple mathematical operations to obtain a variable of
+interest (e.g., \emph{delta.agb} = \emph{ANPP\_woody} -
+\emph{woody.mortality.agb}) disqualifies records from inclusion; (3)
+matching IPCC-defined thresholds for defining C pools (Table 1), which
+may vary by country, can make the data far more relevant for IPCC
+accounting (e.g., using a 10 cm cutoff between dead wood and litter,
+presenting soil C to a depth of 30 cm). It should also be emphasized
+that reporting of 95\% confidence intervals (or other metrics of error),
+when applicable, is highly desirable and makes the data more relevant to
+IPCC.
 
@@
 The ForC database contains large numbers of records that could
-potentially be useful for C accounting under IPCC guidelines. Here we
-have developed a framework for transferring these records to the EFDB,
-making them more accessible for forest C accounting under IPCC
-guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 16, 2023,
+potentially be useful for estimating C stock changes applying
+methodological guidance provided by the IPCC. Here we have developed a
+framework for submitting these records to the EFDB, thus making those
+data more accessible for reporting CO\textsubscript{2} emissions and
+removals from forest land consistent with good practice in the IPCC
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 23, 2023,
 we have submitted 1438 records to EFDB. Although this represents just
 7\% of relevant records in ForC, it substantially increases the number
-of forest land records in EFDB. The records transferred to EFDB and
+of forest land records in EFDB. The records submitted to EFDB and
 present in ForC are very unevenly distributed across variables, regions,
@@
 for forest C accounting under IPCC guidelines by calculating and
-reporting results in ways that are compatible with IPCC guidelines
-(Tables 1, 3). In addition, substantial investments in research and
-researchers focused on ground-based measurement of forest C will be
-required to fill knowledge gaps and thereby increase the accuracy of
-forest C accounting. This challenge is heightened by the fact that
-forests are changing rapidly \citep[e.g.,][]{mcdowell_pervasive_2020},
-and data collected a decade or more in the past may no longer be
-relevant. This heightens the need for an efficient system of making
-forest C data accessible for national greenhouse gas inventories. We
-view this publication as one step towards that goal.
+reporting results in ways that are consistent with methodologies
+provided in the IPCC guidelines (Tables 1, 3). In addition, substantial
+investments in research and researchers focused on ground-based
+measurement of forest C will be required to fill knowledge gaps and
+thereby increase the accuracy of forest CO\textsubscript{2} inventories
+for forest lands under teh Paris Agreement. This challenge is heightened
+by the fact that forests are changing rapidly
+\citep[e.g.,][]{mcdowell_pervasive_2020}, and data collected a decade or
+more in the past may no longer be relevant. This heightens the need for
+an efficient system of making forest C data accessible for national
+greenhouse gas inventories. We view the system developed here for
+submitting ForC data to the IPCC EFDB as one important step towards that
+goal.
 

### 2023-05-23 doc/manuscript/ESSDmanuscript.Rmd
rewording "accounting" and "transfer" (except in display items)
#39 , #51

@@
 running:
-  title: IPCC forest C accounting with ForC
+  title: Informing forest CO~2~ inventories with ForC
   author: Anderson-Teixeira et al.
@@
 The ForC database and associated code is available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD).
-The data and code associated with data transfer to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration).
+The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration).
 authorcontribution: |
-  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data transfer to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC; CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
+  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC; CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
 disclaimer: |
@@
 Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
-Tens of thousands of relevant forest C estimates have been published, yet are not readily accesible to the practitioners compiling national greenhouse gas inventories. 
-Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas accounting if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and transferred to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve accounting of forest-sector greenhouse gas emissions and removals.
+Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
+Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas inventories if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
+Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
 We begin by reconciling terminology and mapping ForC fields into EFDB.
@@
 In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
-Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C accounting.
+Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
 
@@
 Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
-The IPCC inventory guidelines include specific instructions for accounting for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
-A tiered approach to accounting is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
+The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
+A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
 Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances and datasets grow and become more widely accessible.
@@
 Here, we
-(1) review IPCC methods and definitions applied to estimate CO2 emissions and removals from forest in the context of typical forest C estimation methodologies;
+(1) review IPCC methods and definitions applied to estimate CO~2~ emissions and removals from forest in the context of typical forest C estimation methodologies;
 (2) describe mapping of ForC to IPCC's EFDB;
-(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data transfer to EFDB;
+(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data submission to EFDB;
 (4) summarize the data in ForC that's relevant to EFDB and records that have been submitted to date; and
-(5) provide recommendations for improving data collection, analysis, database, and accounting.
+(5) provide recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
 
@@
 
-```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
+```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
 Most, but not all, EFDB-relevant ForC variables are shown here.
@@
 Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version 2.1 [@vandepol_identifying_2016; @bailey_climwin_2016] based on site coordinates. 
-Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or transfer to EFDB.
+Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or submission to EFDB.
 
@@
 
-Manual review of records was the limiting step for data transfer to EFDB. 
+Manual review of records was the limiting step for data submission to EFDB. 
 We prioritized review of
@@
 To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been sent to EFDB.
-Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for inclusion in EFDB. 
-Complete records needed to be reviewed by the EFDB editorial board prior to posting in the database -- a process that lags behind records transfer and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
+Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been submitted to EFDB.
+Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB. 
+Complete records needed to be reviewed by the EFDB editorial board and then posted in the database -- a process that lags behind submission of records and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
 
@@
 The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`.
-For the majority of fields, contents of the field in ForC was transferred directly into an EFDB field, either as the only contents of that field or as part of a composite record.
+For the majority of fields, contents of the field in ForC was copied directly into an EFDB field, either as the only contents of that field or as part of a composite record.
 For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).
@@
 
-Given the lack of public information needed to determine whether lands are classified as managed [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not transfer any classification of land management status from ForC to the EFDB.
-However, we do provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
+Given the lack of public information needed to determine whether lands are classified as managed [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not include any classification of land management status from ForC in the records submitted to EFDB.
+However, we did provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
 
@@
 
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
 Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems."}
@@
 \newpage
-```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of independent records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
 For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
@@
 
-The records transferred from ForC to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
+The ForC records submitted to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
 However, the density of these records was very unevenly distributed across continents, biomes, and forest types and was not proportional to the numbers of relevant records in ForC (Fig.
@@
 
-In terms of variables transferred, records were sent for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
+In terms of variables, records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
 The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records sent were for C stocks, including
@@
 Although this process requires manual review of records, the submission of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
-Future efforts to review studies for transfer should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
+Future efforts to review studies for submission should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
 `r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
@@
 However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017].
-In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate accounting. 
+In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate estimates. 
 We recommend that, when possible, researchers calculate and report these variables, following the reporting guidelines specified in section `r section_data_reporting_needs`.
@@
 (2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;
-(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant for IPCC accounting (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
+(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant estimating forest C stock changes according to IPCC guidelines (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
 It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
@@
 
-Going forward, forest researchers can make their research more useful for forest C accounting under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
+Going forward, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
 In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under teh Paris Agreement.

### 2023-05-23 doc/manuscript/ESSDmanuscript.tex
rewording "accounting" and "transfer" (except in display items)
#39 , #51

@@
 
-\runningtitle{IPCC forest C accounting with ForC}
+\runningtitle{Informing forest CO\textsubscript{2} inventories with
+ForC}
 
@@
 changes therein. Tens of thousands of relevant forest C estimates have
-been published, yet are not readily accesible to the practitioners
+been published, yet are not readily accessible to the practitioners
 compiling national greenhouse gas inventories. Many of these data have,
@@
 https://forc-db.github.io/) and stand to be of value to greenhouse gas
-accounting if made available through the Emission Factor Database (EFDB)
-of the International Panel on Climate Change (IPCC). Here, we develop
-and document a process for semi-automated submission of data from ForC
-into the EFDB, assess the data available and transferred to date, and
-provide recommendations for improving forest data collection, analysis,
-and reporting to improve accounting of forest-sector greenhouse gas
-emissions and removals. We begin by reconciling terminology and mapping
-ForC fields into EFDB. This process required some updates to the ForC
-database structure, leading to the release of a new version of ForC
-(v4.0; described here). As of May 23, 2023, ForC contained
-\textasciitilde19323 independent records relevant to EFDB, 1438 of which
-have been submitted to date. Among the data in ForC, there is
-disproportionate representation of biomass (particularly aboveground)
+inventories if made available through the Emission Factor Database
+(EFDB) of the International Panel on Climate Change (IPCC). Here, we
+develop and document a process for semi-automated submission of data
+from ForC into the EFDB, assess the data available and submitted to
+date, and provide recommendations for improving forest data collection,
+analysis, and reporting to improve inventories of forest-sector
+greenhouse gas emissions and removals. We begin by reconciling
+terminology and mapping ForC fields into EFDB. This process required
+some updates to the ForC database structure, leading to the release of a
+new version of ForC (v4.0; described here). As of May 23, 2023, ForC
+contained \textasciitilde19323 independent records relevant to EFDB,
+1438 of which have been submitted to date. Among the data in ForC, there
+is disproportionate representation of biomass (particularly aboveground)
 stocks, with far fewer records for dead organic matter and soil C, and
@@
 is rapidly impacting the world's forests, timely reporting of recent
-estimates will be especially critical to accurate forest C accounting.
+estimates will be especially critical to accurate forest C inventories.
 
@@
 the world's forest land \citep{ogle_delineating_2018}. The IPCC
-inventory guidelines include specific instructions for accounting for
+inventory guidelines include specific instructions for inventories for
 greenhouse gas (mainly CO\textsubscript{2}) exchanges between forest
 land and the atmosphere \citep{ipcc_agriculture_2006, ipcc_2019_2019}. A
-tiered approach to accounting is employed, where the lowest tier (Tier
-1) represents the simplest approach and relies on default parameter
-values -- for example, forest carbon (C) stocks values by ecozone
+tiered approach is employed, where the lowest tier (Tier 1) represents
+the simplest approach and relies on default parameter values -- for
+example, forest carbon (C) stocks values by ecozone
 \citep{fao_global_2012} and forest age class derived as the average of
@@
 
-Here, we (1) review IPCC methods and definitions applied to estimate CO2
-emissions and removals from forest in the context of typical forest C
-estimation methodologies; (2) describe mapping of ForC to IPCC's EFDB;
-(3) describe updates to ForC (ForC v4.0), most of which were implemented
-to facilitate data transfer to EFDB; (4) summarize the data in ForC
-that's relevant to EFDB and records that have been submitted to date;
-and (5) provide recommendations for improving data collection, analysis,
-database, and accounting.
+Here, we (1) review IPCC methods and definitions applied to estimate
+CO\textsubscript{2} emissions and removals from forest in the context of
+typical forest C estimation methodologies; (2) describe mapping of ForC
+to IPCC's EFDB; (3) describe updates to ForC (ForC v4.0), most of which
+were implemented to facilitate data submission to EFDB; (4) summarize
+the data in ForC that's relevant to EFDB and records that have been
+submitted to date; and (5) provide recommendations as to how the
+scientific community can better provide useful data for forest C
+inventories under the Paris Agreement.
 
@@
 \begin{figure}
-\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools quantified under IPCC accounting; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
+\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
 Most, but not all, EFDB-relevant ForC variables are shown here.
@@
 mm)) were flagged as requiring review prior to use in analysis or
-transfer to EFDB.
+submission to EFDB.
 
@@
 
-Manual review of records was the limiting step for data transfer to
+Manual review of records was the limiting step for data submission to
 EFDB. We prioritized review of (1) records from the Forest Global Earth
@@
 having been digitized from graphs or calculated based on related
-variables, and (3) the records had not previously been sent to EFDB.
-Once converted into EFDB format, the records were reviewed and then sent
-to the IPCC's Technical Support Unit for inclusion in EFDB. Complete
-records needed to be reviewed by the EFDB editorial board prior to
-posting in the database -- a process that lags behind records transfer
-and had not yet been completed for all records sent as of May 23, 2023.
+variables, and (3) the records had not previously been submitted to
+EFDB. Once converted into EFDB format, the records were reviewed and
+then sent to the IPCC's Technical Support Unit for submission to EFDB.
+Complete records needed to be reviewed by the EFDB editorial board and
+then posted in the database -- a process that lags behind submission of
+records and had not yet been completed for all records sent as of May
+23, 2023.
 
@@
 The mapping of ForC fields into EFDB fields is summarized in Appendix B.
-For the majority of fields, contents of the field in ForC was
-transferred directly into an EFDB field, either as the only contents of
-that field or as part of a composite record. For example, ten ForC
-fields describing site location, climate, and edaphic properties all
-mapped into the EFDB field \emph{Region/Regional conditions} (Appendix
-B). In cases where original studies did not present 95\% confidence
-intervals (required by IPCC when available) but did present information
-required to calculate these (standard error or n and standard
-deviation), we calculated the 95\% confidence intervals and populated
-the EFDB field with this information (noting the calculation in the EFDB
-field \emph{Comments from Data Provider}). For some fields, simple
-conditional logic was used to populate EFDB fields based on ForC
-records. For example, for stock variables presented in the original
-publication in units of dry organic matter mass (as opposed to C),
-several greenhouse gasses (CO\textsubscript{2}, CO, CH\textsubscript{4},
-NO, NO\textsubscript{2}, N\textsubscript{2}O) were entered in the EFDB
-field indicating the greenhouse gases to which the record could be
-pertinent (\emph{Gases} field) because these values could be used in
-calculations of greenhouse gas emissions from biomass burning
-\citep{ipcc_2006_2006}; otherwise, the only pertinent greenhouse gas
-would be CO\textsubscript{2}. There were two cases in which more complex
-mapping was required: (1) mapping of C cycle variables (section 4.1.1)
-and (2) land classification (section 4.1.2).
+For the majority of fields, contents of the field in ForC was copied
+directly into an EFDB field, either as the only contents of that field
+or as part of a composite record. For example, ten ForC fields
+describing site location, climate, and edaphic properties all mapped
+into the EFDB field \emph{Region/Regional conditions} (Appendix B). In
+cases where original studies did not present 95\% confidence intervals
+(required by IPCC when available) but did present information required
+to calculate these (standard error or n and standard deviation), we
+calculated the 95\% confidence intervals and populated the EFDB field
+with this information (noting the calculation in the EFDB field
+\emph{Comments from Data Provider}). For some fields, simple conditional
+logic was used to populate EFDB fields based on ForC records. For
+example, for stock variables presented in the original publication in
+units of dry organic matter mass (as opposed to C), several greenhouse
+gasses (CO\textsubscript{2}, CO, CH\textsubscript{4}, NO,
+NO\textsubscript{2}, N\textsubscript{2}O) were entered in the EFDB field
+indicating the greenhouse gases to which the record could be pertinent
+(\emph{Gases} field) because these values could be used in calculations
+of greenhouse gas emissions from biomass burning \citep{ipcc_2006_2006};
+otherwise, the only pertinent greenhouse gas would be
+CO\textsubscript{2}. There were two cases in which more complex mapping
+was required: (1) mapping of C cycle variables (section 4.1.1) and (2)
+land classification (section 4.1.2).
 
@@
 applied in the scientific literature and hence in ForC, we did not
-transfer any classification of land management status from ForC to the
-EFDB. However, we do provide auxiliary information that should be useful
-in making this determination, including geographical location and
-notable disturbance events.
+include any classification of land management status from ForC in the
+records submitted to EFDB. However, we did provide auxiliary information
+that should be useful in making this determination, including
+geographical location and notable disturbance events.
 
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and transferred to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
+\includegraphics[width=15cm]{figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent} \caption{\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
 Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.}\label{fig:fig_map}
@@
 \begin{figure}
-\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of independent records in ForC relevant to (grey) and transferred to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+\includegraphics[width=15cm]{figures_tables/Histogram_n_Relevant_and_Transferred_Records} \caption{\textbf{Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
 For FAO ecozones (b), codes are as listed in the caption of Figure 2.}\label{fig:fig_histograms}
@@
 
-The records transferred from ForC to EFDB were broadly distributed
-across Earth's forests (Fig. 2). However, the density of these records
-was very unevenly distributed across continents, biomes, and forest
-types and was not proportional to the numbers of relevant records in
-ForC (Fig. 3). Rather, the largest number of records came from North
-America, followed by Asia, South America, and Africa (Fig. 3c), with the
-most represented FAO ecozones being boreal coniferous forest, temperate
-continental forest, and temperate mountain systems, followed by tropical
-rain forests and moist deciduous forests (Fig. 3b). In terms of dominant
+The ForC records submitted to EFDB were broadly distributed across
+Earth's forests (Fig. 2). However, the density of these records was very
+unevenly distributed across continents, biomes, and forest types and was
+not proportional to the numbers of relevant records in ForC (Fig. 3).
+Rather, the largest number of records came from North America, followed
+by Asia, South America, and Africa (Fig. 3c), with the most represented
+FAO ecozones being boreal coniferous forest, temperate continental
+forest, and temperate mountain systems, followed by tropical rain
+forests and moist deciduous forests (Fig. 3b). In terms of dominant
 vegetation, by far the most records came from needleleaf evergreen
@@
 
-In terms of variables transferred, records were sent for 19 variables
-(or closely-related variable groups), including variables from each C
-pool (Table 2). The majority (82\%) of records sent were for C stocks,
+In terms of variables, records were submitted for 19 variables (or
+closely-related variable groups), including variables from each C pool
+(Table 2). The majority (82\%) of records sent were for C stocks,
 including 3\% for total biomass, 53\% for aboveground biomass, 4\% for
@@
 submission to EFDB using the system developed here. Future efforts to
-review studies for transfer should optimize for representation across
+review studies for submission should optimize for representation across
 geographic regions, forest types, and variables, giving priority to
@@
 and could be used to estimate standing dead wood, although additional
-data on breakage would be needed for accurate accounting. We recommend
+data on breakage would be needed for accurate estimates. We recommend
 that, when possible, researchers calculate and report these variables,
@@
 matching IPCC-defined thresholds for defining C pools (Table 1), which
-may vary by country, can make the data far more relevant for IPCC
-accounting (e.g., using a 10 cm cutoff between dead wood and litter,
-presenting soil C to a depth of 30 cm). It should also be emphasized
-that reporting of 95\% confidence intervals (or other metrics of error),
-when applicable, is highly desirable and makes the data more relevant to
-IPCC.
+may vary by country, can make the data far more relevant estimating
+forest C stock changes according to IPCC guidelines (e.g., using a 10 cm
+cutoff between dead wood and litter, presenting soil C to a depth of 30
+cm). It should also be emphasized that reporting of 95\% confidence
+intervals (or other metrics of error), when applicable, is highly
+desirable and makes the data more relevant to IPCC.
 
@@
 Going forward, forest researchers can make their research more useful
-for forest C accounting under IPCC guidelines by calculating and
+for forest C inventories under IPCC guidelines by calculating and
 reporting results in ways that are consistent with methodologies
@@
 and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD). The
-data and code associated with data transfer to EFDB and preparation of
+data and code associated with data submission to EFDB and preparation of
 this manuscript are available via the the IPCC-EFDB-integration
@@
 \authorcontribution{KAT and VH conceived and designed the project; VH
-wrote the scripts for database management, data transfer to EFDB, and
+wrote the scripts for database management, data submission to EFDB, and
 the analyses presented here; MW, TR, and RBM added and reviewed ForC

### 2023-05-23 doc/manuscript/ESSDmanuscript.Rmd
change "transferred" or "sent" to "submitted" #51
@teixeirak, FYI, I also edidted the .RMD.
Not sure why but values in C_variables.csv have changed. let me know if that was expected to you. (for biomass_ag)

@@
 
-n_records_sent_to_EFDB = C_variables[length(C_variables[,"variable"]), "n sent to EFDB"] # n records sent to EFDB
-n_relevant_variables_no_records_sent =  sum(C_variables$`n sent to EFDB` == 0,  na.rm = TRUE)
+n_records_sent_to_EFDB = C_variables[length(C_variables[,"variable"]), "n submitted to EFDB"] # n records sent to EFDB
+n_relevant_variables_no_records_sent =  sum(C_variables$`n submitted to EFDB` == 0,  na.rm = TRUE)
 n_relevant_variables_records_sent= n_relevant_variables - n_relevant_variables_no_records_sent
-n_records_sent_biomass<- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass"]
-n_records_sent_agb <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_ag"]
-n_records_sent_agb_components <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_ag_foliage"]
-n_records_sent_rootbiomass <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_root"] 
-n_records_sent_rootbiomass_components <- C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n sent to EFDB`[C_variables$`variable` == "biomass_root_coarse"]
-n_records_sent_deadwood<- C_variables$`n sent to EFDB`[C_variables$`variable` == "deadwood"] 
-n_records_sent_deadwood_components<- C_variables$`n sent to EFDB`[C_variables$`variable` == "deadwood_down"]+C_variables$`n sent to EFDB`[C_variables$`variable` == "deadwood_standing"]
-n_records_sent_litter<- C_variables$`n sent to EFDB`[C_variables$`variable` == "litter"] +C_variables$`n sent to EFDB`[C_variables$`variable` == "O.horizon"]
-n_records_sent_soil<- C_variables$`n sent to EFDB`[C_variables$`variable` == "SOM / SOC"] 
+n_records_sent_biomass<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass"]
+n_records_sent_agb <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag"]
+n_records_sent_agb_components <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag_foliage"]
+n_records_sent_rootbiomass <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root"] 
+n_records_sent_rootbiomass_components <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root_coarse"]
+n_records_sent_deadwood<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood"] 
+n_records_sent_deadwood_components<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_down"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_standing"]
+n_records_sent_litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "litter"] +C_variables$`n submitted to EFDB`[C_variables$`variable` == "O.horizon"]
+n_records_sent_soil<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "SOM / SOC"] 
 n_records_sent_stocks = n_records_sent_biomass+n_records_sent_agb+n_records_sent_agb_components+n_records_sent_rootbiomass+n_records_sent_rootbiomass_components+n_records_sent_deadwood+n_records_sent_deadwood_components+ n_records_sent_litter+ n_records_sent_soil
-n_records_sent_delta.agb<- C_variables$`n sent to EFDB`[C_variables$`variable` == "delta.agb"]
-n_records_sent_delta.litter<- C_variables$`n sent to EFDB`[C_variables$`variable` == "delta.O.horizon"]+C_variables$`n sent to EFDB`[C_variables$`variable` == "delta.litter"]
+n_records_sent_delta.agb<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.agb"]
+n_records_sent_delta.litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.O.horizon"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.litter"]
 n_records_sent_increments <- n_records_sent_delta.agb+n_records_sent_delta.litter
-n_records_sent_fluxes <- sum(C_variables$`n sent to EFDB`[C_variables$`variable` %in% flux_variables_with_records])
+n_records_sent_fluxes <- sum(C_variables$`n submitted to EFDB`[C_variables$`variable` %in% flux_variables_with_records])
 n_records_sent_fluxes_check = n_records_sent_to_EFDB-n_records_sent_stocks-n_records_sent_increments

### 2023-05-23 scripts/1_ForC_to_IPCC_formatting.R

@@ -76,11 +76,11 @@ ForC_simplified <- ForC_simplified[ForC_simplified$variable.name %in% V_mapping$
 # ---- right, here, make the histogram for the paper ---- #
 ## see this issue:https://github.com/forc-db/IPCC-EFDB-integration/issues/34
 ## grey is number of records in ForC that are relevant to EFDB (n records for the variables we send)
-## black is what is already transferred to EFDB
+## black is what is already submitted to EFDB
 ## x axis is dominant vegetation type
 
 
-ForBarplot <- data.frame(data_type = rep(c("relevant", "transferred"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg), FAO.ecozone = c(ForC_simplified$FAO.ecozone, ForC_simplified_already_sent$FAO.ecozone), continent = c(ForC_simplified$continent, ForC_simplified_already_sent$continent), stand.age = c(ForC_simplified$stand.age, ForC_simplified_already_sent$stand.age))
+ForBarplot <- data.frame(data_type = rep(c("relevant", "submitted"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg), FAO.ecozone = c(ForC_simplified$FAO.ecozone, ForC_simplified_already_sent$FAO.ecozone), continent = c(ForC_simplified$continent, ForC_simplified_already_sent$continent), stand.age = c(ForC_simplified$stand.age, ForC_simplified_already_sent$stand.age))
 
 ForBarplot$stand.age <- addNA(cut(as.numeric(ForBarplot$stand.age), breaks = c(0, 20, 100, 10000), labels = c("<20 yrs", "20-100 yrs", ">100 yrs" )))
 levels(ForBarplot$stand.age)[4] <- "Unclassified"

### 2023-05-23 scripts/3_create_list_of_ all_data_sources_sent.R

@@ -97,7 +97,7 @@ n_independent <- as.data.frame(table(ForC_simplified$variable_pretty[ForC_simpli
 
 n_reviewed <- as.data.frame(table(ForC_simplified$variable_pretty[ForC_simplified$citation.ID %in% CITATIONS$citation.ID[CITATIONS$EFDB.ready == 1]]), responseName = "n_reviewed")
 
-n_sent_to_EFDB <- as.data.frame(table(MEASUREMENTS$variable_pretty[MEASUREMENTS$measurement.ID %in% trace_of_measurement_IDs$measurement.ID]), responseName = "n_sent_to_EFDB")
+n_submitted_to_EFDB <- as.data.frame(table(MEASUREMENTS$variable_pretty[MEASUREMENTS$measurement.ID %in% trace_of_measurement_IDs$measurement.ID]), responseName = "n_submitted_to_EFDB")
 
 
 
@@ -106,7 +106,7 @@ n_sent_to_EFDB <- as.data.frame(table(MEASUREMENTS$variable_pretty[MEASUREMENTS$
 variables_in_order$n_in_ForC <- n_in_ForC$n_in_ForC[match(variables_in_order$variable, n_in_ForC$Var1)]
 variables_in_order$n_independent_records_in_ForC <- n_independent$n_independent_records_in_ForC[match(variables_in_order$variable, n_independent$Var1)]
 variables_in_order$n_reviewed <- n_reviewed$n_reviewed[match(variables_in_order$variable, n_reviewed$Var1)]
-variables_in_order$n_sent_to_EFDB <- n_sent_to_EFDB$n_sent_to_EFDB[match(variables_in_order$variable, n_sent_to_EFDB$Var1)]
+variables_in_order$n_submitted_to_EFDB <- n_submitted_to_EFDB$n_submitted_to_EFDB[match(variables_in_order$variable, n_submitted_to_EFDB$Var1)]
 
 
 

### 2023-05-23 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd

@@
 Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify (issue 40)-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
+IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component (defined as ...<!--issue 49-->), and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
 A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].

### 2023-05-23 doc/manuscript/ESSDmanuscript.Rmd
exclude R_het_soil, knit in new display items
closes #51
closes #45

@@
 n_ind_records_deltaOL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.O.horizon"] 
-flux_variables_with_records <- c("NPP_woody", "ANPP_woody", "ANPP_woody_stem", "ANPP_woody_branch", "woody.mortality_ag", "BNPP_root", "BNPP_root_fine","BNPP_root.turnover_fine","BNPP_root_coarse" , "ANPP_litterfall", "NPP_litter", "R_het_litter", "R_het_soil")
+flux_variables_with_records <- c("NPP_woody", "ANPP_woody", "ANPP_woody_stem", "ANPP_woody_branch", "woody.mortality_ag", "BNPP_root", "BNPP_root_fine","BNPP_root.turnover_fine","BNPP_root_coarse" , "ANPP_litterfall", "NPP_litter", "R_het_litter")
 n_ind_records_flux <- sum(C_variables$`n independent records in ForC`[C_variables$`variable` %in% flux_variables_with_records])
@@
 Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon."}
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
+**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable."}
 #this is for a two-column figure. for one-column figure, use out.width = "8.3cm"

### 2023-05-23 doc/manuscript/ESSDmanuscript.tex
exclude R_het_soil, knit in new display items
closes #51
closes #45

@@
 new version of ForC (v4.0; described here). As of May 23, 2023, ForC
-contained \textasciitilde19323 independent records relevant to EFDB,
+contained \textasciitilde19316 independent records relevant to EFDB,
 1438 of which have been submitted to date. Among the data in ForC, there
@@
 understory may be excluded the understory if it constitutes a ``minor''
-component, \emph{where quantitative definitions of ``understory'' and
-``minor'' are not provided}, but where a commonly applied minimum size
-sampling threshold for mature forests would be 10 cm stem diameter at
-breast height (DBH). A recent study characterizing the contributions of
-trees in different DBH classes to ecosystem C stocks and fluxes found
+component (defined as \ldots{}), and where a commonly applied minimum
+size sampling threshold for mature forests would be 10 cm stem diameter
+at breast height (DBH). A recent study characterizing the contributions
+of trees in different DBH classes to ecosystem C stocks and fluxes found
 that trees 1 - 10 cm DBH contributed up to \textasciitilde8\%
@@
 we added or modified 18 fields (Appendix A), defined 15 new variables,
-implemented enhanced quality control, manually reviewed \textgreater1954
+implemented enhanced quality control, manually reviewed \textgreater1963
 records to obtain additional required information, and added 329 new
@@
 Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.}\label{fig:fig_variable_mapping}
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
+**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable.}\label{fig:fig_variable_mapping}
 \end{figure}
@@
 As of May 23, 2023, ForC (v4.0) contained 32693 independent records
-(39848 total), 19323 of which were for the 43 variables relevant to EFDB
+(39848 total), 19316 of which were for the 42 variables relevant to EFDB
 (Fig. 1). These records were distributed across all forested continents
@@
 
-ForC contained records for 29 of the 43 variables (or closely-related
+ForC contained records for 29 of the 42 variables (or closely-related
 variable groups) relevant to EFDB (Table 2, Fig. 1). The records were
@@
 \hline
-\textbf{variable} & \textbf{n in ForC} & \textbf{n independent records in ForC} & \textbf{n reviewed} & \textbf{n sent to EFDB}\\
+\textbf{variable} & \textbf{n in ForC} & \textbf{n independent records in ForC} & \textbf{n reviewed} & \textbf{n submitted to EFDB}\\
 \hline
@@
 \hline
-biomass\_ag & 9456 & 8155 & 1348 & 764\\
+biomass\_ag & 9449 & 8148 & 1357 & 764\\
 \hline
@@
 \hline
-R\_het\_soil & 0 & 0 & 0 & 0\\
-\hline
-\textbf{TOTAL} & \textbf{23575} & \textbf{19323} & \textbf{2283} & \textbf{1438}\\
+\textbf{TOTAL} & \textbf{23568} & \textbf{19316} & \textbf{2292} & \textbf{1438}\\
 \hline
@@
 
-As of May 23, 2023, we had reviewed or added 2283 EFDB-relevant records,
+As of May 23, 2023, we had reviewed or added 2292 EFDB-relevant records,
 1438 records of which were submitted to EFDB, and 376 of which have been

### 2023-06-02 doc/manuscript/ESSDmanuscript.Rmd

@@
 Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
-Yet, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
+However, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
 Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
@@
 Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
-This distribution of records is generally reflected in the subset of records that have been submitted to EFDB to date.
 In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
@@
 Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
-Going into the future, the fate of this important CO~2~ sink is highly uncertain, depending both upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and on human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
+The future of this important CO~2~ sink is highly uncertain, and depends upon both forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
 
-Reflecting their strong influence on Earth's climate, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
+Forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
 Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options, with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
@@
 A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
-Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for continuous improvement as the science advances and datasets grow and become more widely accessible.
+Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
 For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default values underestimated C sequestration by 32% on average and failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
@@
 The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
-ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
+ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
 As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
@@
 (3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data submission to EFDB;
-(4) summarize the data in ForC that's relevant to EFDB and records that have been submitted to date; and
+(4) summarize the data in ForC relevant to EFDB and records that have been submitted to date; and
 (5) provide recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
@@
 
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass.
-Selection of allometric models has an enormous influence on estimates of biomass stocks, increments, of fluxes [@clark_landscapescale_2000;@clark_net_2001].
+A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001].
 While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
@@
 Dead organic matter includes all non-living biomass that is not within the mineral soil layer and smaller than the litter size threshold. 
-It's inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
+Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
 
@@
 
+This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
 Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement.
@@
 To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
-
-This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
 
@@
 Some examples are as follows:
-(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion;
+(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion while supporting FAIR (http://dx.doi.org/10.1038/d41586-019-01720-7) goals;
 (2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;

### 2023-06-02 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd
@bpbond , FYI, this text in contributions section refers to Susan Cook-Patton's contribution of the GROA database. I added a bit of clarification.

@@
 authorcontribution: |
-  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC; CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
+  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
 disclaimer: |

### 2023-06-02 doc/manuscript/ESSDmanuscript.Rmd
Camille declined authorship

@@
     affiliation: "5"
-  - given_name: Camille
-    surname: Piponiot
-    affiliation: "6"
 
@@
     address: The Nature Conservancy; Arlington VA 22203, USA
-  - code: 6
-    address: CIRAD, Montpellier, France
+
 abstract: |

### 2023-06-08 doc/manuscript/ESSDmanuscript.Rmd
add comments from Avni Malhotra
@bpbond , FYI

@@
   Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have contributed to building the database.
+  Thank you to Avni Malhotra for helpful comments on an earlier draft of this manuscript.
   Funding for this study was provided by the Institute for Global Environmental Strategies, a Bezos Earth Fund award to The Nature Conservancy, and a Smithsonian Working Land and Seascapes grant.
@@
 Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
-The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain [@ref].
+The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, and tending to be very uncertain [e.g., @keller_biomass_2001].
 Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
-Field methods for estimating root biomass are highly variable [@ref].
+Field methods for estimating root biomass are highly variable [@freschet_starting_2021].
 IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].

### 2023-06-08 doc/manuscript/ESSDmanuscript.tex
add comments from Avni Malhotra
@bpbond , FYI

@@
 \Author[5]{Susan}{Cook-Patton}
-\Author[6]{Camille}{Piponiot}
 
@@
 \affil[5]{The Nature Conservancy; Arlington VA 22203, USA}
-\affil[6]{CIRAD, Montpellier, France}
 
@@
 and constitute a substantial portion of planned net emissions reductions
-under the 2015 Paris Agreement. Yet, the efficacy of greenhouse gas
+under the 2015 Paris Agreement. However, the efficacy of greenhouse gas
 mitigation planning and reporting is dependent upon the quality of
@@
 some updates to the ForC database structure, leading to the release of a
-new version of ForC (v4.0; described here). As of May 23, 2023, ForC
+new version of ForC (v4.0; described here). As of June 08, 2023, ForC
 contained \textasciitilde19316 independent records relevant to EFDB,
@@
 relevant records in temperate forests, and with relatively scant
-representation of tropical forests in Africa and Asia. This distribution
-of records is generally reflected in the subset of records that have
-been submitted to EFDB to date. In the future, forest C estimates in
-EFDB can be improved through targeted research to fill critical gaps,
-reporting of information required by IPCC, and continued submission of
-data from scientific publications to the EFDB. Given that climate change
-is rapidly impacting the world's forests, timely reporting of recent
-estimates will be especially critical to accurate forest C inventories.
+representation of tropical forests in Africa and Asia. In the future,
+forest C estimates in EFDB can be improved through targeted research to
+fill critical gaps, reporting of information required by IPCC, and
+continued submission of data from scientific publications to the EFDB.
+Given that climate change is rapidly impacting the world's forests,
+timely reporting of recent estimates will be especially critical to
+accurate forest C inventories.
 
@@
 the pace of atmospheric CO\textsubscript{2} accumulation and associated
-climate change. Going into the future, the fate of this important
-CO\textsubscript{2} sink is highly uncertain, depending both upon forest
-responses to climate change, which are likely to reduce the sink
-strength \citep{mcdowell_pervasive_2020, hammond_global_2022}, and on
-human conservation, restoration, and management of forests
+climate change. The future of this important CO\textsubscript{2} sink is
+highly uncertain, and depends upon both forest responses to climate
+change, which are likely to reduce the sink strength
+\citep{mcdowell_pervasive_2020, hammond_global_2022}, and human
+conservation, restoration, and management of forests
 \citep{ipcc_climate_2019, ipcc_climate_2022}.
 
-Reflecting their strong influence on Earth's climate, forests play a
-substantial role in international plans for climate change mitigation
-under the Paris Agreement \citep{unfccc_adoption_2015}. Forest
-conservation, reforestation, and improved sustainable management all
-have significant -- and relatively cost-effective -- potential as
-climate change mitigation options, with conservation and reforestation
-having the fourth and fifth largest net emission reduction potentials or
-all mitigation options \citep{ipcc_summary_2022}. As of 2016,
-forest-based mitigation accounted for 26\% of total planned greenhouse
-gas mitigation within Nationally Determined Contributions under the
-Paris Agreement \citep{grassi_key_2017}. Yet, envisioned forest-based
-climate change mitigation initiatives do not always correspond to actual
-emission reductions through on-the-ground implementation
+Forests play a substantial role in international plans for climate
+change mitigation under the Paris Agreement
+\citep{unfccc_adoption_2015}. Forest conservation, reforestation, and
+improved sustainable management all have significant -- and relatively
+cost-effective -- potential as climate change mitigation options, with
+conservation and reforestation having the fourth and fifth largest net
+emission reduction potentials or all mitigation options
+\citep{ipcc_summary_2022}. As of 2016, forest-based mitigation accounted
+for 26\% of total planned greenhouse gas mitigation within Nationally
+Determined Contributions under the Paris Agreement
+\citep{grassi_key_2017}. Yet, envisioned forest-based climate change
+mitigation initiatives do not always correspond to actual emission
+reductions through on-the-ground implementation
 \citep[e.g.,][]{badgley_systematic_2022}. One critical need for ensuring
@@
 \citep{requenasuarez_estimating_2019, rozendaal_aboveground_2022}, but
-there remains room for continuous improvement as the science advances
-and datasets grow and become more widely accessible. For example, the
-year following the release of the latest IPCC guidelines, a more
-thorough analysis of C accumulation in regrowth forests found that
-IPCC's Tier 1 default values underestimated C sequestration by 32\% on
-average and failed to capture eight-fold variation within ecozones
-\citep{cook-patton_mapping_2020}. In addition, it was revealed that C
-stocks in mature African tropical montane forests were two-thirds higher
-than the IPCC Tier 1 values for these forests
+there remains room for improvement as datasets grow and become more
+widely accessible. For example, the year following the release of the
+latest IPCC guidelines, a more thorough analysis of C accumulation in
+regrowth forests found that IPCC's Tier 1 default values underestimated
+C sequestration by 32\% on average and failed to capture eight-fold
+variation within ecozones \citep{cook-patton_mapping_2020}. In addition,
+it was revealed that C stocks in mature African tropical montane forests
+were two-thirds higher than the IPCC Tier 1 values for these forests
 \citep{cuni-sanchez_high_2021}. This rapid evolution of scientific
@@
 (GROA) database \citep[database doi:
-10.5281/zenodo.3983644]{cook-patton_mapping_2020}, the global soil
+10.5281/zenodo.3983644]{cook-patton_mapping_2020}, and the global soil
 respiration database
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of May 23, 2023, ForC contained 39848 records from 10589 plots in
+As of June 08, 2023, ForC contained 39848 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 were implemented to facilitate data submission to EFDB; (4) summarize
-the data in ForC that's relevant to EFDB and records that have been
-submitted to date; and (5) provide recommendations as to how the
-scientific community can better provide useful data for forest C
-inventories under the Paris Agreement.
+the data in ForC relevant to EFDB and records that have been submitted
+to date; and (5) provide recommendations as to how the scientific
+community can better provide useful data for forest C inventories under
+the Paris Agreement.
 
@@
 \hline
- &  & minimum root diameter & may exclude fine roots; suggested minimum diameter of 2 mm for fine roots\\
+ &  & minimum root diameter & may exclude fine roots; suggested diameter cutoff of 2 mm for fine roots\\
 \hline
@@
 census data is applying appropriate allometric models to convert DBH
-measurements to biomass. Selection of allometric models has an enormous
-influence on estimates of biomass stocks, increments, of fluxes
+measurements to biomass, and such selection has an enormous influence on
+estimates of biomass stocks, increments, and fluxes
 \citep{clark_landscapescale_2000, clark_net_2001}. While trusted and
@@
 methodologically linked to aboveground biomass estimates, sharing the
-same methodological sources of variation, but tending to be far more
-uncertain \citep{ref}. Fine root biomass generally constitutes a much
-smaller C pool \citep[typically \textless5 t C
-ha\textsuperscript{-1},][]{anderson-teixeira_carbon_2021}, and IPCC
+same methodological sources of variation, and tending to be very
+uncertain \citep[e.g.,][]{keller_biomass_2001}. Fine root biomass
+generally constitutes a much smaller C pool \citep[typically \textless5
+t C ha\textsuperscript{-1},][]{anderson-teixeira_carbon_2021}, and IPCC
 guidance is that it can be excluded when fine roots cannot be
@@
 \citep{ipcc_2006_2006}, which can be a painstaking process. Field
-methods for estimating root biomass are highly variable \citep{ref}.
-IPCC's default method for Tier 1 estimates is to apply a ratio of
-belowground to aboveground biomass, with default factors defined based
-on ecological zone, continent, and forest age
-\citep{ipcc_2006_2006, ipcc_2019_2019}.
+methods for estimating root biomass are highly variable
+\citep{freschet_starting_2021}. IPCC's default method for Tier 1
+estimates is to apply a ratio of belowground to aboveground biomass,
+with default factors defined based on ecological zone, continent, and
+forest age \citep{ipcc_2006_2006, ipcc_2019_2019}.
 
@@
 Dead organic matter includes all non-living biomass that is not within
-the mineral soil layer and smaller than the litter size threshold. It's
+the mineral soil layer and smaller than the litter size threshold. Its
 inclusion in inventories is not required under Tier 1 methodology for
@@
 
-Previous versions of ForC
+This section describes changes relative to ForC v3.0
+\citep{anderson-teixeira_carbon_2021}. Previous versions of ForC
 \citep{anderson-teixeira_carbon_2016, anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}
@@
 
-This section describes changes relative to ForC v3.0
-\citep{anderson-teixeira_carbon_2021}.
-
 \subsection{New or modified fields}
@@
 (\emph{woody.mortality}). Although most of these variables lacked
-records in ForC as of May 23, 2023, their addition gave the structure
+records in ForC as of June 08, 2023, their addition gave the structure
 such that records can be populated over time. Finally, to provide better
@@
 then posted in the database -- a process that lags behind submission of
-records and had not yet been completed for all records sent as of May
-23, 2023.
+records and had not yet been completed for all records sent as of June
+08, 2023.
 
@@
 
-As of May 23, 2023, ForC (v4.0) contained 32693 independent records
+As of June 08, 2023, ForC (v4.0) contained 32693 independent records
 (39848 total), 19316 of which were for the 42 variables relevant to EFDB
@@
 
-As of May 23, 2023, we had reviewed or added 2292 EFDB-relevant records,
-1438 records of which were submitted to EFDB, and 376 of which have been
-reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
+As of June 08, 2023, we had reviewed or added 2292 EFDB-relevant
+records, 1438 records of which were submitted to EFDB, and 376 of which
+have been reviewed, accepted, and posted (Figs. 2-3, Table 2). The 37\%
 attenuation between records reviewed and those sent to EFDB was
@@
 EFDB, whereas presentation in a table or supplementary data file allows
-inclusion; (2) direct presentation of all relevant variables allows
-inclusion, whereas presenting only components of variables of interest
-(e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or
-requiring simple mathematical operations to obtain a variable of
-interest (e.g., \emph{delta.agb} = \emph{ANPP\_woody} -
-\emph{woody.mortality.agb}) disqualifies records from inclusion; (3)
-matching IPCC-defined thresholds for defining C pools (Table 1), which
-may vary by country, can make the data far more relevant estimating
-forest C stock changes according to IPCC guidelines (e.g., using a 10 cm
-cutoff between dead wood and litter, presenting soil C to a depth of 30
-cm). It should also be emphasized that reporting of 95\% confidence
-intervals (or other metrics of error), when applicable, is highly
-desirable and makes the data more relevant to IPCC.
+inclusion while supporting FAIR
+(http://dx.doi.org/10.1038/d41586-019-01720-7) goals; (2) direct
+presentation of all relevant variables allows inclusion, whereas
+presenting only components of variables of interest (e.g., parsing
+litter into fine woody debris, OL, OF, and OH layers) or requiring
+simple mathematical operations to obtain a variable of interest (e.g.,
+\emph{delta.agb} = \emph{ANPP\_woody} - \emph{woody.mortality.agb})
+disqualifies records from inclusion; (3) matching IPCC-defined
+thresholds for defining C pools (Table 1), which may vary by country,
+can make the data far more relevant estimating forest C stock changes
+according to IPCC guidelines (e.g., using a 10 cm cutoff between dead
+wood and litter, presenting soil C to a depth of 30 cm). It should also
+be emphasized that reporting of 95\% confidence intervals (or other
+metrics of error), when applicable, is highly desirable and makes the
+data more relevant to IPCC.
 
@@
 \hline
-reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not acceptvalues digitized from figures.\\
+reporting estimates & Report all relevant values in tables, text, or supplementary tables/ data files, as opposed to in figures only. & EFDB does not accept values digitized from figures.\\
 \hline
@@
 removals from forest land consistent with good practice in the IPCC
-guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of May 23, 2023,
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of June 08, 2023,
 we have submitted 1438 records to EFDB. Although this represents just
@@
 the analyses presented here; MW, TR, and RBM added and reviewed ForC
-data, BBL and SCP contributed large databases to ForC; CP provided
-methodological expertise; KAT, VH, and MW prepared the first draft of
-the manuscript; all authors reviewed the results and approved the final
-version of the manuscript.} %% optional section
+data, BBL and SCP contributed large databases to ForC (EFDB and GROA,
+respectively); CP provided methodological expertise; KAT, VH, and MW
+prepared the first draft of the manuscript; all authors reviewed the
+results and approved the final version of the
+manuscript.} %% optional section
 
@@
 to all research assistants and collaborators who have contributed to
-building the database. Funding for this study was provided by the
-Institute for Global Environmental Strategies, a Bezos Earth Fund award
-to The Nature Conservancy, and a Smithsonian Working Land and Seascapes
-grant.
+building the database. Thank you to Avni Malhotra for helpful comments
+on an earlier draft of this manuscript. Funding for this study was
+provided by the Institute for Global Environmental Strategies, a Bezos
+Earth Fund award to The Nature Conservancy, and a Smithsonian Working
+Land and Seascapes grant.
 \end{acknowledgements}

### 2023-06-08 doc/manuscript/ESSDmanuscript.Rmd
address issue #49
closes #49

@@
 Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component (defined as ...<!--issue 49-->), and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
-A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
+IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
+A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can generally be considered a minor component of aboveground biomass for these forests.
 In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.

### 2023-06-08 doc/manuscript/ESSDmanuscript.tex
address issue #49
closes #49

@@
 understory may be excluded the understory if it constitutes a ``minor''
-component (defined as \ldots{}), and where a commonly applied minimum
-size sampling threshold for mature forests would be 10 cm stem diameter
-at breast height (DBH). A recent study characterizing the contributions
-of trees in different DBH classes to ecosystem C stocks and fluxes found
-that trees 1 - 10 cm DBH contributed up to \textasciitilde8\%
-aboveground biomass, \textasciitilde17\% aboveground woody net primary
-productivity (\(ANPP_{woody.stem}\)), and \textasciitilde20\% woody
-mortality (\(M_{woody}\)) of mature closed-canopy forests worldwide
-\citep{piponiot_distribution_2022}. In regrowth forests, woodlands, or
-savannas, small trees and shrubs contribute a much larger proportion of
-C stocks and fluxes
+component \citep[defined as \textless{} 25 - 30 \% of emissions/removals
+for the overall category,][]{ipcc_2006_2006}, and where a commonly
+applied minimum size sampling threshold for mature forests would be 10
+cm stem diameter at breast height (DBH). A recent study characterizing
+the contributions of trees in different DBH classes to ecosystem C
+stocks and fluxes found that trees 1 - 10 cm DBH contributed up to
+\textasciitilde8\% aboveground biomass, \textasciitilde17\% aboveground
+woody net primary productivity (\(ANPP_{woody.stem}\)), and
+\textasciitilde20\% woody mortality (\(M_{woody}\)) of mature
+closed-canopy forests worldwide \citep{piponiot_distribution_2022}, and
+therefore stems \textless{} 10 cm DBH can generally be considered a
+minor component of aboveground biomass for these forests. In regrowth
+forests, woodlands, or savannas, small trees and shrubs contribute a
+much larger proportion of C stocks and fluxes
 \citep{lutz_global_2018, piponiot_distribution_2022, hughes_biomass_1999},

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd

@@
 IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
-A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can generally be considered a minor component of aboveground biomass for these forests.
+A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can usually be considered a minor component of aboveground biomass for these forests.
 In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
revise language on gains & losses
#48

@@
 The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include C stocks, net increments ("stock changes"), and fluxes ("gains" and "losses") for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
+In the case of forests, emission factors include C stocks, net annual increments, and annual fluxes for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
 
@@
 `r fig_variable_mapping`).
-For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, stock changes (a.k.a.
-"net annual increments" by IPCC, "increments" in ForC), gains (a.k.a.
-"gross annual increments" by IPCC, "fluxes" in ForC), and losses ("fluxes" in ForC).
+For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, net annual increments, influxes (a.k.a. "gross annual increments" by IPCC), and outfluxes.
 As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 

### 2023-06-09 doc/manuscript/ESSDmanuscript.tex
revise language on gains & losses
#48

@@
 some updates to the ForC database structure, leading to the release of a
-new version of ForC (v4.0; described here). As of June 08, 2023, ForC
+new version of ForC (v4.0; described here). As of June 09, 2023, ForC
 contained \textasciitilde19316 independent records relevant to EFDB,
@@
 sources for inclusion in the database. In the case of forests, emission
-factors include C stocks, net increments (``stock changes''), and fluxes
-(``gains'' and ``losses'') for various pools
-\citep{ipcc_2006_2006, ipcc_2019_2019}.
+factors include C stocks, net annual increments, and annual fluxes for
+various pools \citep{ipcc_2006_2006, ipcc_2019_2019}.
 
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of June 08, 2023, ForC contained 39848 records from 10589 plots in
+As of June 09, 2023, ForC contained 39848 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 closed-canopy forests worldwide \citep{piponiot_distribution_2022}, and
-therefore stems \textless{} 10 cm DBH can generally be considered a
-minor component of aboveground biomass for these forests. In regrowth
-forests, woodlands, or savannas, small trees and shrubs contribute a
-much larger proportion of C stocks and fluxes
+therefore stems \textless{} 10 cm DBH can usually be considered a minor
+component of aboveground biomass for these forests. In regrowth forests,
+woodlands, or savannas, small trees and shrubs contribute a much larger
+proportion of C stocks and fluxes
 \citep{lutz_global_2018, piponiot_distribution_2022, hughes_biomass_1999},
@@
 (\emph{woody.mortality}). Although most of these variables lacked
-records in ForC as of June 08, 2023, their addition gave the structure
+records in ForC as of June 09, 2023, their addition gave the structure
 such that records can be populated over time. Finally, to provide better
@@
 records and had not yet been completed for all records sent as of June
-08, 2023.
+09, 2023.
 
@@
 appropriately map them into EFDB (Fig. 1). For each C pool (Table 1), we
-identified variables representing organic matter or C stocks, stock
-changes (a.k.a. ``net annual increments'' by IPCC, ``increments'' in
-ForC), gains (a.k.a. ``gross annual increments'' by IPCC, ``fluxes'' in
-ForC), and losses (``fluxes'' in ForC). As described in section 3.2, we
-also defined 15 new EFDB-relevant variables that were not previously
-represented in ForC. It is important to note that the correspondence of
-ForC variables to IPCC criteria often depends upon measurement protocols
-(``important sources of estimate variation'' in Table 1). For example,
-ForC records of biomass and dead wood vary in the minimum stem diameter
-censused, such that some records would match the IPCC criteria whereas
-others would not. Information on minimum diameters censused and other
-important sources of methodological variation are recorded as covariates
-in ForC and mapped into the EFDB field \emph{Other Properties} (Appendix
-B). Details on the mapping of ForC variables to EFDB -- including
-associated covariates, IPCC pools (Table 1) and relevant equations
+identified variables representing organic matter or C stocks, net annual
+increments, influxes (a.k.a. ``gross annual increments'' by IPCC), and
+outfluxes. As described in section 3.2, we also defined 15 new
+EFDB-relevant variables that were not previously represented in ForC. It
+is important to note that the correspondence of ForC variables to IPCC
+criteria often depends upon measurement protocols (``important sources
+of estimate variation'' in Table 1). For example, ForC records of
+biomass and dead wood vary in the minimum stem diameter censused, such
+that some records would match the IPCC criteria whereas others would
+not. Information on minimum diameters censused and other important
+sources of methodological variation are recorded as covariates in ForC
+and mapped into the EFDB field \emph{Other Properties} (Appendix B).
+Details on the mapping of ForC variables to EFDB -- including associated
+covariates, IPCC pools (Table 1) and relevant equations
 \citep{ipcc_2006_2006} -- are documented in the file
@@
 
-As of June 08, 2023, ForC (v4.0) contained 32693 independent records
+As of June 09, 2023, ForC (v4.0) contained 32693 independent records
 (39848 total), 19316 of which were for the 42 variables relevant to EFDB
@@
 
-As of June 08, 2023, we had reviewed or added 2292 EFDB-relevant
+As of June 09, 2023, we had reviewed or added 2292 EFDB-relevant
 records, 1438 records of which were submitted to EFDB, and 376 of which
@@
 removals from forest land consistent with good practice in the IPCC
-guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of June 08, 2023,
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of June 09, 2023,
 we have submitted 1438 records to EFDB. Although this represents just

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
further revise language on gains & losses (#48)
#48

@@
 As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
-Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into (gains) or out of (losses) the IPCC-defined C pools.
+Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into or out of the IPCC-defined C pools.
 Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
@@
 For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
-For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006].
 Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
@@
 
-```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
+```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
 Most, but not all, EFDB-relevant ForC variables are shown here.

### 2023-06-09 doc/manuscript/ESSDmanuscript.tex
further revise language on gains & losses (#48)
#48

@@
 relatively few or no records for net annual increments or C fluxes into
-(gains) or out of (losses) the IPCC-defined C pools. Geographic
-representation is also quite uneven, with the highest densities of
-relevant records in temperate forests, and with relatively scant
-representation of tropical forests in Africa and Asia. In the future,
-forest C estimates in EFDB can be improved through targeted research to
-fill critical gaps, reporting of information required by IPCC, and
-continued submission of data from scientific publications to the EFDB.
-Given that climate change is rapidly impacting the world's forests,
-timely reporting of recent estimates will be especially critical to
-accurate forest C inventories.
+or out of the IPCC-defined C pools. Geographic representation is also
+quite uneven, with the highest densities of relevant records in
+temperate forests, and with relatively scant representation of tropical
+forests in Africa and Asia. In the future, forest C estimates in EFDB
+can be improved through targeted research to fill critical gaps,
+reporting of information required by IPCC, and continued submission of
+data from scientific publications to the EFDB. Given that climate change
+is rapidly impacting the world's forests, timely reporting of recent
+estimates will be especially critical to accurate forest C inventories.
 
@@
 yr\textsuperscript{-1}) are calculated as the sum of changes in various
-pools (described in section 2.1), plus any harvested wood products. For
-each pool, \(\Delta C\) may be calculated using the ``Gain-Loss
-Method'', which takes the difference between gains and losses, or using
-the ``Stock-Difference Method'', which computes \(\Delta C\) based on C
-stocks at two points in time \citep{ipcc_2006_2006}. Thus, C cycle
-variables relevant to the IPCC methodology and to EFDB include C stocks,
-net annual increments, and fluxes in the IPCC-defined pools.
+pools (described in section 2.1), plus any harvested wood products.
+Thus, C cycle variables relevant to the IPCC methodology and to EFDB
+include C stocks, net annual increments, and fluxes in the IPCC-defined
+pools.
 
@@
 \begin{figure}
-\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, stock change (net annual increment), gain (influx), and loss.
+\includegraphics[width=14cm]{figures_tables/C_variable_mapping} \caption{\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
 Most, but not all, EFDB-relevant ForC variables are shown here.

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
integrate Susan's comments
#39

@@
 This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have been submitted to date.
+As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have undergone necessary review and been submitted to date.
 Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into or out of the IPCC-defined C pools.
@@
 Forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
-Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options, with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
+Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
 As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
@@
 Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
-For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default values underestimated C sequestration by 32% on average and failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
+For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
 In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
@@
 Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
+IPCC's guidance is that the understory may be excluded if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
 A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can usually be considered a minor component of aboveground biomass for these forests.
@@
 
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001].
+A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001;@calders_laser_2022].
 While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
@@
 
-Dead organic matter includes all non-living biomass that is not within the mineral soil layer and smaller than the litter size threshold. 
+Dead organic matter includes all non-living biomass larger than the litter size threshold. 
 Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
@@
 As noted above, live fine roots may be included in litter when difficult to separate empirically.
-The definition includes the entire O horizion, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
+The definition includes the entire O horizon, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
 This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
@@
 and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@refs], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@refs; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
+The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@refs; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
 
@@
 Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
-Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022], ... *others?  (SUSAN)*. 
+Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
 Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
@@
 Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
-However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017].
+However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017; @mokany_critical_2006].
 In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate estimates. 
@@
 Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
+Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023;@calders_laser_2022].
 Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.

### 2023-06-09 doc/manuscript/ESSDmanuscript.tex
integrate Susan's comments
#39

@@
 contained \textasciitilde19316 independent records relevant to EFDB,
-1438 of which have been submitted to date. Among the data in ForC, there
-is disproportionate representation of biomass (particularly aboveground)
-stocks, with far fewer records for dead organic matter and soil C, and
-relatively few or no records for net annual increments or C fluxes into
-or out of the IPCC-defined C pools. Geographic representation is also
-quite uneven, with the highest densities of relevant records in
-temperate forests, and with relatively scant representation of tropical
-forests in Africa and Asia. In the future, forest C estimates in EFDB
-can be improved through targeted research to fill critical gaps,
-reporting of information required by IPCC, and continued submission of
-data from scientific publications to the EFDB. Given that climate change
-is rapidly impacting the world's forests, timely reporting of recent
-estimates will be especially critical to accurate forest C inventories.
+1438 of which have undergone necessary review and been submitted to
+date. Among the data in ForC, there is disproportionate representation
+of biomass (particularly aboveground) stocks, with far fewer records for
+dead organic matter and soil C, and relatively few or no records for net
+annual increments or C fluxes into or out of the IPCC-defined C pools.
+Geographic representation is also quite uneven, with the highest
+densities of relevant records in temperate forests, and with relatively
+scant representation of tropical forests in Africa and Asia. In the
+future, forest C estimates in EFDB can be improved through targeted
+research to fill critical gaps, reporting of information required by
+IPCC, and continued submission of data from scientific publications to
+the EFDB. Given that climate change is rapidly impacting the world's
+forests, timely reporting of recent estimates will be especially
+critical to accurate forest C inventories.
 
@@
 improved sustainable management all have significant -- and relatively
-cost-effective -- potential as climate change mitigation options, with
-conservation and reforestation having the fourth and fifth largest net
-emission reduction potentials or all mitigation options
-\citep{ipcc_summary_2022}. As of 2016, forest-based mitigation accounted
-for 26\% of total planned greenhouse gas mitigation within Nationally
-Determined Contributions under the Paris Agreement
+cost-effective -- potential as climate change mitigation options
+\citep{roe_landbased_2021}, with conservation and reforestation having
+the fourth and fifth largest net emission reduction potentials or all
+mitigation options \citep{ipcc_summary_2022}. As of 2016, forest-based
+mitigation accounted for 26\% of total planned greenhouse gas mitigation
+within Nationally Determined Contributions under the Paris Agreement
 \citep{grassi_key_2017}. Yet, envisioned forest-based climate change
@@
 latest IPCC guidelines, a more thorough analysis of C accumulation in
-regrowth forests found that IPCC's Tier 1 default values underestimated
-C sequestration by 32\% on average and failed to capture eight-fold
-variation within ecozones \citep{cook-patton_mapping_2020}. In addition,
-it was revealed that C stocks in mature African tropical montane forests
-were two-thirds higher than the IPCC Tier 1 values for these forests
-\citep{cuni-sanchez_high_2021}. This rapid evolution of scientific
-information on C cycling in forests is valuable for informing climate
-change mitigation efforts but requires improved mechanisms for
+regrowth forests found that IPCC's Tier 1 default failed to capture
+eight-fold variation within ecozones \citep{cook-patton_mapping_2020}.
+In addition, it was revealed that C stocks in mature African tropical
+montane forests were two-thirds higher than the IPCC Tier 1 values for
+these forests \citep{cuni-sanchez_high_2021}. This rapid evolution of
+scientific information on C cycling in forests is valuable for informing
+climate change mitigation efforts but requires improved mechanisms for
 communicating the latest information from scientific researchers to the
@@
 \citep{ipcc_good_2003, ipcc_2006_2006}. IPCC's guidance is that the
-understory may be excluded the understory if it constitutes a ``minor''
-component \citep[defined as \textless{} 25 - 30 \% of emissions/removals
-for the overall category,][]{ipcc_2006_2006}, and where a commonly
-applied minimum size sampling threshold for mature forests would be 10
-cm stem diameter at breast height (DBH). A recent study characterizing
-the contributions of trees in different DBH classes to ecosystem C
-stocks and fluxes found that trees 1 - 10 cm DBH contributed up to
+understory may be excluded if it constitutes a ``minor'' component
+\citep[defined as \textless{} 25 - 30 \% of emissions/removals for the
+overall category,][]{ipcc_2006_2006}, and where a commonly applied
+minimum size sampling threshold for mature forests would be 10 cm stem
+diameter at breast height (DBH). A recent study characterizing the
+contributions of trees in different DBH classes to ecosystem C stocks
+and fluxes found that trees 1 - 10 cm DBH contributed up to
 \textasciitilde8\% aboveground biomass, \textasciitilde17\% aboveground
@@
 estimates of biomass stocks, increments, and fluxes
-\citep{clark_landscapescale_2000, clark_net_2001}. While trusted and
-standardized allometric models are becoming increasingly available
+\citep{clark_landscapescale_2000, clark_net_2001, calders_laser_2022}.
+While trusted and standardized allometric models are becoming
+increasingly available
 \citep{chave_improved_2014, rejou-mechain_biomass_2017, gonzalez-akre_allodb_2022},
@@
 
-Dead organic matter includes all non-living biomass that is not within
-the mineral soil layer and smaller than the litter size threshold. Its
-inclusion in inventories is not required under Tier 1 methodology for
-Forest Land remaining Forest Land (see section 2.2), but is required for
-land that has transitioned to or from forest within the past 20 years
-\citep{ipcc_2006_2006}.
+Dead organic matter includes all non-living biomass larger than the
+litter size threshold. Its inclusion in inventories is not required
+under Tier 1 methodology for Forest Land remaining Forest Land (see
+section 2.2), but is required for land that has transitioned to or from
+forest within the past 20 years \citep{ipcc_2006_2006}.
 
@@
 may be included in litter when difficult to separate empirically. The
-definition includes the entire O horizion, including litter (OL), fumic
+definition includes the entire O horizon, including litter (OL), fumic
 (OF), and humic (OH) layers, in addition to litter embedded within the
@@
 fact that although tropical forest is the single most important biome
-for climate change mitigation \citep{refs}, ground-based data on
-tropical forest C cycling tends to be more scarce due to a variety of
+for climate change mitigation
+\citep{griscom_natural_2017, griscom_national_2020}, ground-based data
+on tropical forest C cycling tends to be more scarce due to a variety of
 challenges \citep{refs, delima_making_2022}, and tropical countries are
@@
 recent efforts have compiled large databases of relevant data from
-monoculture plantation forests \citep{bukoski_rates_2022}, \ldots{}
-\emph{others? (SUSAN)}. Beyond expanding collections of relevant forest
-C records, such reviews are valuable for assessing the availability of
-published records and identifying variables and regions that require
-additional data collection and analysis.
+monoculture plantation forests \citep{bukoski_rates_2022} and mixed
+species plantation forests
+\citep{feng_multispecies_2022, warner_higher_2022}, and such a
+compilation is in works for agroforestry (Susan Cook-Patton, unpublished
+data). Beyond expanding collections of relevant forest C records, such
+reviews are valuable for assessing the availability of published records
+and identifying variables and regions that require additional data
+collection and analysis.
 
@@
 require improvement, they do exist for many forest types
-\citep[e.g.,][]{brassard_coarse_2011, chojnacky_updated_2014, waring_overlooking_2017}.
+\citep[e.g.,][]{brassard_coarse_2011, chojnacky_updated_2014, waring_overlooking_2017, mokany_critical_2006}.
 In addition, standing dead trees are captured in most forest censuses
@@
 global maps of forest biomass and changes therein
-\citep{duncanson_importance_2019, labrière_forest_2023}. Given
-observations of increasing tree mortality in some forested regions
+\citep{duncanson_importance_2019, labrière_forest_2023, calders_laser_2022}.
+Given observations of increasing tree mortality in some forested regions
 \citep{mcdowell_pervasive_2020}, better characterization of forest dead

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
Update ESSDmanuscript.Rmd
#50

@@
 The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
-The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review the records, and it is not expected that many -- if any -- records will be rejected.
+The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review and post the records, and also because a minority of records were deemed not applicable to EFDB by the review panel.
 

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
filling in/fixing references
includes temporary hack to fix labrière citation (manually changed citation key to labriere)

#39

@@
 and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@refs; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
+The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@anderson-teixeira_carbon_2021; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
 
@@
 For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
-Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labrière_forest_2023].
+Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labriere_forest_2023].
 
@@
 Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023;@calders_laser_2022].
+Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labriere_forest_2023;@calders_laser_2022].
 Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
@@
 \captionsetup[table]{labelformat=empty}
-Table A1: **Table of changes to ForC fields.**
+Table A1: **Table of changes to ForC fields.** These are changes implemented between releases of v3.0 and v4.0. 
 ```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}

### 2023-06-09 doc/manuscript/ESSDmanuscript.tex
filling in/fixing references
includes temporary hack to fix labrière citation (manually changed citation key to labriere)

#39

@@
 on tropical forest C cycling tends to be more scarce due to a variety of
-challenges \citep{refs, delima_making_2022}, and tropical countries are
-more likely to apply Tier 1 methodology that bases forest C budgets on
-internationally defined IPCC default values
+challenges \citep{anderson-teixeira_carbon_2021, delima_making_2022},
+and tropical countries are more likely to apply Tier 1 methodology that
+bases forest C budgets on internationally defined IPCC default values
 \citep{romijn_assessing_2015}.
@@
 between the number of records sent and that posted to EFDB is primarily
-attributable to the time required for the IPCC to review the records,
-and it is not expected that many -- if any -- records will be rejected.
+attributable to the time required for the IPCC to review and post the
+records, and also because a minority of records were deemed not
+applicable to EFDB by the review panel.
 
@@
 regions will be important to filling knowledge gaps in forest C cycling
-\citep{delima_making_2022, araza_decade_2023, labrière_forest_2023}.
+\citep{delima_making_2022, araza_decade_2023, labriere_forest_2023}.
 
@@
 global maps of forest biomass and changes therein
-\citep{duncanson_importance_2019, labrière_forest_2023, calders_laser_2022}.
+\citep{duncanson_importance_2019, labriere_forest_2023, calders_laser_2022}.
 Given observations of increasing tree mortality in some forested regions
@@
 
-Table A1: \textbf{Table of changes to ForC fields.}
+Table A1: \textbf{Table of changes to ForC fields.} These are changes
+implemented between releases of v3.0 and v4.0.
 \begingroup\fontsize{8}{10}\selectfont

### 2023-06-09 doc/manuscript/ESSDmanuscript.tex
check/update appendix a

@@
 
-As of June 09, 2023, ForC (v4.0) contained 32693 independent records
+As of June 09, 2023, ForC (v4.0) contained 32686 independent records
 (39848 total), 19316 of which were for the 42 variables relevant to EFDB
@@
 \hline
- & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC with original units, reduce errors/improve reproducibility\\
+ & mean.in.original.units, original.units & mean value and units presented in original publication & fields added & provide IPCC's EFDB with original units, reduce errors/improve reproducibility\\
 \hline
@@
 \hline
-PFT & description & Definition of the pftcode at the community level. Differs from individual level in that properly describes mixed plant functional types. & field added & \\
-\hline
- & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & \\
-\hline
-Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC\\
-\hline
- & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections. & field added & field required by IPCC\\
-\hline
- & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC\\
-\hline
- & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC\\
-\hline
- & source.type & citation source type & field added & field required by IPCC\\
+PFT & description & Definition of the pftcode at the community level.
+Differs from individual level in that properly describes mixed plant functional types.
+& field added & clarify PFT at community and individual levels\\
+\hline
+ & description.individual & Definition of the pftcode at the individual plant level. & field name change (previously `description`) & clarify PFT at community and individual levels\\
+\hline
+Citations & citation.citation & Full citation. Most of these records are automatically generated in R based upon DOI lookup. & field added & field required by IPCC's EFDB\\
+\hline
+ & citation.language & Language of original publication, automatically generated based on the title and abstract, with some manual entries and corrections.
+& field added & field required by IPCC's EFDB\\
+\hline
+ & citation.url & URL of original publication, generally retrieved automatically via URL lookup. & field added & field required by IPCC's EFDB\\
+\hline
+ & citation.abstract & Abstract, generally retrieved automatically via DOI lookup. & field added & field required by IPCC's EFDB\\
+\hline
+ & source.type & citation source type & field added & field required by IPCC's EFDB\\
 \hline

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
rename file (disambiguation)

@@
 
-ForC_variables_included <- read.csv("figures_tables/ForC_variables_included.csv")
+ForC_variables_included <- read.csv("figures_tables/ForC_fields_included.csv")
 names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")

### 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
fix footnotes in appendix b
#39

@@
 Table B1: **Mapping of ForC fields to EFDB.** Details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
-See footnotes at end of table (STILL NEED TO BE PROPERLY INSERTED).
 ```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
-ForC_variables_included <- ForC_variables_included[c(1:5)] #include all fields but TO_DO
+column_names <- names(ForC_variables_included) # Define column names
+column_names[5] <- "Required*" # Add asterisk for footnote
+names(ForC_variables_included) <- column_names# Set column names of the matrix
+ForC_variables_included <- ForC_variables_included[c(1:5)]
 
@@
 ```
-'Required' field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
-
-** '(yes)' for most recent severe disturbance; 'no' for other history events
+\* *Required* field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.

### 2023-06-09 doc/manuscript/ESSDmanuscript.tex
fix footnotes in appendix b
#39

@@
 \url{https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv}).
-See footnotes at end of table (STILL NEED TO BE PROPERLY INSERTED).
 \begingroup\fontsize{8}{10}\selectfont
@@
 \hline
-\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required}\\
+\textbf{ForC table} & \textbf{ForC field} & \textbf{EFDB field} & \textbf{Usage} & \textbf{Required*}\\
 \hline
@@
 \hline
-History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & (yes)/no**\\
+History & date, hist.cat, hist.type & 1996 Source/Sink Categories, 2006 Source/Sink Categories, Abatement/Control technologies & used to determine distmrs.type for Source/Sink Categories, generate list of events for Abatement/Control technologies & most recent severe disturbance: (yes) / other history events: no\\
 \hline
@@
 
-`Required' field indicates whether the field is required by EFDB: yes =
-value required; (yes) = input required, missing value acceptable if not
-reported; auto = present within ForC infrasructure, and therefore will
-always be exported to EFDB ; (no) = not required for EFDB, but required
-for ForC and therefore will always be exported to EFDB; no = not
-required, but exported to EFDB when a value is present.
-
-** `(yes)' for most recent severe disturbance; `no' for other history
-events
+* \emph{Required} field indicates whether the field is required by EFDB:
+yes = value required; (yes) = input required, missing value acceptable
+if not reported; auto = present within ForC infrasructure, and therefore
+will always be exported to EFDB ; (no) = not required for EFDB, but
+required for ForC and therefore will always be exported to EFDB; no =
+not required, but exported to EFDB when a value is present.
 

### 2023-06-09 doc/manuscript/ForC-EFDB manuscript for ESSD.Rmd
rename manuscript file
refine language on funding

--- 2023-06-09 doc/manuscript/ESSDmanuscript.Rmd
@@
   Thank you to Avni Malhotra for helpful comments on an earlier draft of this manuscript.
-  Funding for this study was provided by the Institute for Global Environmental Strategies, a Bezos Earth Fund award to The Nature Conservancy, and a Smithsonian Working Land and Seascapes grant.
+  Funding for this study was provided by the Smithsonian (Forest Global Earth Observatory, Smithsonian Working Land and Seascapes); a Bezos Earth Fund grant to the Nature Conservancy, with a sub-grant to NZCBI; and the Institute for Global Environmental Strategies.
 appendix: 
@@
 ```
-
-****
-*THIS IS AN IN-PREP MANUSCRIPT.*
-
-****
 

### 2023-06-09 doc/manuscript/ForC-EFDB-manuscript-for-ESSD.tex
rename manuscript file
refine language on funding

--- 2023-06-09 doc/manuscript/ESSDmanuscript.tex
@@
 
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
-\emph{THIS IS AN IN-PREP MANUSCRIPT.}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
 \textbf{Abstract.} Forests are critical for climate change mitigation
@@
 on an earlier draft of this manuscript. Funding for this study was
-provided by the Institute for Global Environmental Strategies, a Bezos
-Earth Fund award to The Nature Conservancy, and a Smithsonian Working
-Land and Seascapes grant.
+provided by the Smithsonian (Forest Global Earth Observatory,
+Smithsonian Working Land and Seascapes); a Bezos Earth Fund grant to the
+Nature Conservancy, with a sub-grant to NZCBI; and the Institute for
+Global Environmental Strategies.
 \end{acknowledgements}

### 2023-06-09 doc/manuscript/ForC-EFDB manuscript for ESSD.Rmd
fix Ogle reference, add Zenodo DOI

@@
 The ForC database and associated code is available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD).
-The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration).
+The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).
 authorcontribution: |

### 2023-06-09 doc/manuscript/ForC-EFDB-manuscript-for-ESSD.tex
fix Ogle reference, add Zenodo DOI

@@
 repository within the ForC-db organization on GitHub
-(https://github.com/forc-db/IPCC-EFDB-integration).} %% use this section when having data sets and software code available
+(https://github.com/forc-db/IPCC-EFDB-integration) and archived in
+Zenodo (DOI:
+10.5281/zenodo.8021474).} %% use this section when having data sets and software code available
 

### 2023-06-09 doc/manuscript/ForC-EFDB manuscript for ESSD.Rmd
refine data statement

@@
     All code and data are openly available.
-The ForC database and associated code is available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD).
+The ForC database and associated code are available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) is archived in Zenodo (DOI: TBD).
 The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).

### 2023-06-09 doc/manuscript/ForC-EFDB-manuscript-for-ESSD.tex
refine data statement

@@
 \codedataavailability{All code and data are openly available. The ForC
-database and associated code is available via the ForC repository within
-the ForC-db organization on GitHub (https://github.com/forc-db/ForC),
-and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD). The
-data and code associated with data submission to EFDB and preparation of
-this manuscript are available via the the IPCC-EFDB-integration
-repository within the ForC-db organization on GitHub
-(https://github.com/forc-db/IPCC-EFDB-integration) and archived in
-Zenodo (DOI:
+database and associated code are available via the ForC repository
+within the ForC-db organization on GitHub
+(https://github.com/forc-db/ForC), and the version used here (ForC v4.0)
+is archived in Zenodo (DOI: TBD). The data and code associated with data
+submission to EFDB and preparation of this manuscript are available via
+the the IPCC-EFDB-integration repository within the ForC-db organization
+on GitHub (https://github.com/forc-db/IPCC-EFDB-integration) and
+archived in Zenodo (DOI:
 10.5281/zenodo.8021474).} %% use this section when having data sets and software code available

### 2023-10-26 doc/manuscript/ESSD/Copernicus template-rticles/Copernicus-template-rticles.tex
stash ESSD files, create first-pass word file

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% copernicus_rticles_template (flag for rticles template detection - do not remove!)

%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.

%% 2-column papers and discussion papers
\documentclass[gc, manuscript]{copernicus}

%% Journal abbreviations (please use the same for discussion papers and final revised papers)

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% ASTRA Proceedings (ap)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Bone and Joint Infection (jbji)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Polarforschung - Journal of the German Society for Polar Research (polf)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

% Pandoc citation processing

% The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
%
\usepackage{algorithmic}
\usepackage{algorithm}

\begin{document}

\title{Template for preparing your manuscript submission to Copernicus
journals using RMarkdown}

\Author[1, *]{Daniel}{Nüst}
\Author[2]{Josiah}{Carberry}
\Author[1, *]{Markus}{Konkol}
\Author[3, †]{Nikolaus}{Copernicus}

\affil[1]{Institute for Geoinformatics, University of Münster, 48149
Münster, Germany}
\affil[2]{Psychoceramics, Wesleyan University, Middletown, CT, United
States}
\affil[3]{University of Ferrara, Ferrara, Italy}
\affil[†]{deceased, 24 May 1543}
\affil[*]{These authors contributed equally to this work.}

\runningtitle{R Markdown Template for Copernicus}

\runningauthor{Nüst et al.}

\correspondence{Daniel\ Nüst\ (daniel.nuest@uni-muenster.de)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
The abstract goes here. It can also be on \emph{multiple lines}.
\end{abstract}

\copyrightstatement{The author's copyright for this publication is
transferred to institution/company.}

\introduction[Introduction]

Introduction text goes here. You can change the name of the section if
neccessary using
\texttt{\textbackslash{}introduction{[}modified\ heading{]}}.

The following settings can or must be configured in the header of this
file and are bespoke for Copernicus manuscripts:

\begin{itemize}
\item
  The \texttt{journal} you are submitting to using the official
  abbreviation. You can use the function
  \texttt{rticles::copernicus\_journal\_abbreviations(name\ =\ \textquotesingle{}...\textquotesingle{})}
  to search the existing journals.
\item
  Specific sections of the manuscript:

  \begin{itemize}
  \item
    \texttt{running} with \texttt{title} and \texttt{author}
  \item
    \texttt{competinginterests}
  \item
    \texttt{copyrightstatement} (optional)
  \item
    \texttt{availability} (strongly recommended if any used), one of
    \texttt{code}, \texttt{data}, or \texttt{codedata}
  \item
    \texttt{authorcontribution}
  \item
    \texttt{disclaimer}
  \item
    \texttt{acknowledgements}
  \end{itemize}
\end{itemize}

See the defaults and examples from the skeleton and the official
Copernicus documentation for details.

\textbf{Important}: Always double-check with the official manuscript
preparation guidelines at
\url{https://publications.copernicus.org/for_authors/manuscript_preparation.html},
especially the sections ``Technical instructions for LaTeX'' and
``Manuscript composition''. Please contact Daniel Nüst,
\texttt{daniel.nuest@uni-muenster.de}, with any problems.

\section{Content section one}

\subsection{Subsection Heading Here}

Subsection text here.

\subsubsection{Subsubsection Heading Here}

Subsubsection text here.

\section{Content section with citations}

See the
\href{http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html}{R
Markdown docs for bibliographies and citations}.

Copernicus supports biblatex and a sample bibliography is in file
\texttt{sample.bib}. Read \citep{Feynman1963118}, and
\citep[see][]{Dirac1953888}.

\section{Content section with R code chunks}

You should always use \texttt{echo\ =\ FALSE} on R Markdown code blocks
as they add formatting and styling not desired by Copernicus. The hidden
workflow results in 42.

You can add verbatim code snippets without extra styles by using
\texttt{\textasciigrave{}\textasciigrave{}\textasciigrave{}} without
additional instructions.

\begin{verbatim}
sum <- 1 + 41
\end{verbatim}

\section{Content section with list}

If you want to insert a list, you must

\begin{itemize}
\item
  leave
\item
  empty lines
\item
  between each list item
\end{itemize}

because the \texttt{\textbackslash{}tightlist} format used by R Markdown
is not supported in the Copernicus template. Example:

\begin{verbatim}
- leave

- empty lines

- between each list item
\end{verbatim}

\section{Examples from the official template}

\subsection{FIGURES}

When figures and tables are placed at the end of the MS (article in
one-column style), please add \clearpage between bibliography and first
table and/or figure as well as between each table and/or figure.

\subsubsection{ONE-COLUMN FIGURES}

Include a 12cm width figure of Nikolaus Copernicus from
\href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia}
with caption using R Markdown.

\begin{figure}
\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:unnamed-chunk-2}
\end{figure}

\subsubsection{TWO-COLUMN FIGURES}

You can also include a larger figure.

\begin{figure}
\includegraphics[width=12cm]{Nikolaus_Kopernikus} \caption{two column figure}\label{fig:unnamed-chunk-3}
\end{figure}

\subsection{TABLES}

You can ad \LaTeX table in an R Markdown document to meet the template
requirements.

\subsubsection{ONE-COLUMN TABLE}

\begin{table}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table Footnotes}
\end{table}

\subsubsection{TWO-COLUMN TABLE}

\begin{table*}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table footnotes}
\end{table*}

\subsection{MATHEMATICAL EXPRESSIONS}

All papers typeset by Copernicus Publications follow the math
typesetting regulations given by the IUPAC Green Book (IUPAC:
Quantities, Units and Symbols in Physical Chemistry, 2nd Edn., Blackwell
Science, available at:
http://old.iupac.org/publications/books/gbook/green\_book\_2ed.pdf,
1993).

Physical quantities/variables are typeset in italic font (t for time, T
for Temperature)

Indices which are not defined are typeset in italic font (x, y, z, a, b,
c)

Items/objects which are defined are typeset in roman font (Car A, Car B)

Descriptions/specifications which are defined by itself are typeset in
roman font (abs, rel, ref, tot, net, ice)

Abbreviations from 2 letters are typeset in roman font (RH, LAI)

Vectors are identified in bold italic font using \vec{x}

Matrices are identified in bold roman font

Multiplication signs are typeset using the LaTeX commands
\texttt{\textbackslash{}times} (for vector products, grids, and
exponential notations) or \texttt{\textbackslash{}cdot}

The character * should not be applied as mutliplication sign

\subsection{EQUATIONS}

\subsubsection{Single-row equation}

Unnumbered equations (i.e.~using \texttt{\$\$} and getting inline
preview in RStudio) are not supported by Copernicus.

\begin{equation}
1 \times 1 \cdot 1 = 42
\end{equation}

\begin{equation}
A = \pi r^2
\end{equation}

\begin{equation}
x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
\end{equation}

\subsubsection{Multiline equation}

\begin{align}
& 3 + 5 = 8\\
& 3 + 5 = 8\\
& 3 + 5 = 8
\end{align}

\subsection{MATRICES}

\[
\begin{matrix}
x & y & z\\
x & y & z\\
x & y & z\\
\end{matrix}
\]

\subsection{ALGORITHM}

If you want to use algorithms, you can either enable the required
packages in the header (the default, see \texttt{algorithms:\ true}), or
make sure yourself that the \LaTeX packages \texttt{algorithms} and
\texttt{algorithmicx} are installed so that \texttt{algorithm.sty}
respectively \texttt{algorithmic.sty} can be loaded by the Copernicus
template. Copernicus staff will remove all undesirable packages from
your LaTeX source code, so please stick to using the header option,
which only adds the two acceptable packages.

\begin{algorithm}
\caption{Algorithm Caption}
\label{a1}
\begin{algorithmic}
\STATE $i\gets 10$
\IF {$i\geq 5$} 
        \STATE $i\gets i-1$
\ELSE
        \IF {$i\leq 3$}
                \STATE $i\gets i+2$
        \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

\subsection{CHEMICAL FORMULAS AND REACTIONS}

For formulas embedded in the text, please use
\texttt{\textbackslash{}chem\{\}}, e.g.~\chem{A \rightarrow B}.

The reaction environment creates labels including the letter R,
i.e.~(R1), (R2), etc.

\begin{itemize}
\item
  \texttt{\textbackslash{}rightarrow} should be used for normal
  (one-way) chemical reactions
\item
  \texttt{\textbackslash{}rightleftharpoons} should be used for
  equilibria
\item
  \texttt{\textbackslash{}leftrightarrow} should be used for resonance
  structures
\end{itemize}

\begin{reaction}
A \rightarrow B \\
\end{reaction}
\begin{reaction}
Coper \rightleftharpoons nicus \\
\end{reaction}
\begin{reaction}
Publi \leftrightarrow cations
\end{reaction}

\subsection{PHYSICAL UNITS}

Please use \texttt{\textbackslash{}unit\{\}} (allows to save the
math/\texttt{\$} environment) and apply the exponential notation, for
example \(3.14\,\unit{km\,h^{-1}}\) (using LaTeX mode:
\texttt{\textbackslash{}(\ 3.14\textbackslash{},\textbackslash{}unit\{...\}\ \textbackslash{})})
or \unit{0.872\,m\,s^{-1}} (using only
\texttt{\textbackslash{}unit\{0.872\textbackslash{},m\textbackslash{},s\^{}\{-1\}\}}).

\conclusions[Conclusions]

The conclusion goes here. You can modify the section name with
\texttt{\textbackslash{}conclusions{[}modified\ heading\ if\ necessary{]}}.

\codedataavailability{use this to add a statement when having data sets
and software code
available} %% use this section when having data sets and software code available

\sampleavailability{use this section when having geoscientific samples
available} %% use this section when having geoscientific samples available

\videosupplement{use this section when having video supplements
available} %% use this section when having geoscientific samples available

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Figures and tables in appendices}

Regarding figures and tables in appendices, the following two options
are possible depending on your general handling of figures and tables in
the manuscript environment:

\subsection{Option 1}

If you sorted all figures and tables into the sections of the text,
please also sort the appendix figures and appendix tables into the
respective appendix sections. They will be correctly named
automatically.

\subsection{Option 2}

If you put all figures after the reference list, please insert appendix
tables and figures after the normal tables and figures.

To rename them correctly to A1, A2, etc., please add the following
commands in front of them: \texttt{\textbackslash{}appendixfigures}
needs to be added in front of appendix figures
\texttt{\textbackslash{}appendixtables} needs to be added in front of
appendix tables

Please add \texttt{\textbackslash{}clearpage} between each table and/or
figure. Further guidelines on figures and tables can be found below.
\noappendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontribution{Daniel wrote the package. Josiah thought about
poterry. Markus filled in for a second author.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\competinginterests{The authors declare no competing
interests.} %% this section is mandatory even if you declare that no competing interests are present

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\disclaimer{We like Copernicus.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
Thanks to the rticles contributors!
\end{acknowledgements}

%% REFERENCES
%% DN: pre-configured to BibTeX for rticles

%% The reference list is compiled as follows:
%%
%% \begin{thebibliography}{}
%%
%% \bibitem[AUTHOR(YEAR)]{LABEL1}
%% REFERENCE 1
%%
%% \bibitem[AUTHOR(YEAR)]{LABEL2}
%% REFERENCE 2
%%
%% \end{thebibliography}

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{sample.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

%% LITERATURE CITATIONS
%%
%% command                        & example result
%% \citet{jones90}|               & Jones et al. (1990)
%% \citep{jones90}|               & (Jones et al., 1990)
%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
%% \citeauthor{jones90}|          & Jones et al.
%% \citeyear{jones90}|            & 1990

\end{document}

### 2023-10-26 doc/manuscript/ESSD/ForC-EFDB manuscript for ESSD.Rmd
stash ESSD files, create first-pass word file

--- 2023-06-09 doc/manuscript/ForC-EFDB manuscript for ESSD.Rmd
@@
     All code and data are openly available.
-The ForC database and associated code is available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD).
+The ForC database and associated code are available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) is archived in Zenodo [@anderson-teixeira_forcdb_2023, DOI: 10.5281/zenodo.8020861].
 The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).
@@
 
-The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
+The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023].
 ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
stash ESSD files, create first-pass word file

--- 2023-10-26 doc/manuscript/ESSD/ForC-EFDB manuscript for ESSD.Rmd
@@
 ---
-title: Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
-journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
-author:
-  - given_name: Kristina J.
-    surname: Anderson-Teixeira
-    affiliation: "1,2 *"
-    email: teixeirak@si.edu
-    corresponding: true
-  - given_name: Valentine
-    surname: Herrmann
-    affiliation: "1"
-  - given_name: Madison 
-    surname: Williams
-    affiliation: "1"
-  - given_name: Teagan
-    surname: Rogers
-    affiliation: "1"
-  - given_name: Rebecca
-    surname: Banbury Morgan
-    affiliation: "1,3" 
-  - given_name: Ben 
-    surname: Bond-Lamberty
-    affiliation: "4" 
-  - given_name: Susan 
-    surname: Cook-Patton
-    affiliation: "5"
+title: ''
+output: 
+    #pdf_document: 
+    bookdown::word_document2: 
+      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+bibliography: references.bib
+csl: global-change-biology.csl
 
-affiliation:
-  - code: 1
-    address: Center for Conservation Ecology, Smithsonian's National Zoo & Conservation Biology Institute, Front Royal, VA, United States
-  - code: 2
-    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
-  - code: 3
-    address: School of Geography, University of Leeds, Leeds, UK
-  - code: 4
-    address: Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
-  - code: 5
-    address: The Nature Conservancy; Arlington VA 22203, USA
-
-abstract: |
-
-bibliography: references.bib
-running:
-  title: Informing forest CO~2~ inventories with ForC
-  author: Anderson-Teixeira et al.
-competinginterests: |
-  The authors declare no competing interests.
-# OPTIONAL:
-algorithms: true
-# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
-# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
-copyrightstatement: |
-### The author's copyright for this publication is transferred to institution/company. 
-### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
-availability:
-  #  use this to add a statement when having only software code available
-  #  use this to add a statement when having only data sets available
-  #code: |
-  #data: |
-  codedata: |
-    All code and data are openly available.
-The ForC database and associated code are available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) is archived in Zenodo [@anderson-teixeira_forcdb_2023, DOI: 10.5281/zenodo.8020861].
-The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).
-authorcontribution: |
-  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
-disclaimer: |
-acknowledgements: |
-  We gratefully acknowledge the substantial contributions of Valentyna Slivinska and Sandro Federici for collaboration on the conception, design, and technical review of this project.
-  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have contributed to building the database.
-  Thank you to Avni Malhotra for helpful comments on an earlier draft of this manuscript.
-  Funding for this study was provided by the Smithsonian (Forest Global Earth Observatory, Smithsonian Working Land and Seascapes); a Bezos Earth Fund grant to the Nature Conservancy, with a sub-grant to NZCBI; and the Institute for Global Environmental Strategies.
-appendix: 
-output:
-  rticles::copernicus_article: default
-  bookdown::pdf_book:
-    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
-
-    header-includes:
-    - \usepackage{caption}
-    #- \usepackage{booktabs}
-    #- \usepackage{makecell}
 ---
 
-<!--**Important**: Always double-check with the official manuscript preparation guidelines at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html), especially the sections "Technical instructions for LaTeX" and "Manuscript composition".
-Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.-->
+**Title:** Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
+
+*running title: Informing forest CO~2~ inventories with ForC*
+
+**Authors:**
+
+Kristina J. Anderson-Teixeira^1,2^*
+
+Valentine Herrmann^1^
+
+Madison Williams^1^
+
+Teagan Rogers^1^
+
+Rebecca Banbury Morgan^1,3^
+
+Ben Bond-Lamberty^4^
+
+Susan Cook-Patton^5^
+
+**Affiliations:**
+1. Center for Conservation Ecology, Smithsonian's National Zoo & Conservation Biology Institute, Front Royal, VA, United States
+
+2. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
+
+3. School of Geography, University of Leeds, Leeds, UK
+
+4. Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
+
+5.The Nature Conservancy; Arlington VA 22203, USA
+
+*corresponding author: teixeirak@si.edu
 
@@
 
-\introduction[Introduction]
+\clearpage
+
+# Introduction
 
@@
 Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[ @wickham_rvest_2022]. 
+For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[@wickham_rvest_2022]. 
 That information was manually retrieved when the web scraping failed.
@@
 
-```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
+![**Figure `r fig_variable_mapping`.
+Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.** For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
 Most, but not all, EFDB-relevant ForC variables are shown here.
@@
 Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
-**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/C_variable_mapping.png")
-```
+**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable.](figures_tables/C_variable_mapping.png)
 
@@
 
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png")
-```
+![**Figure `r fig_map`.
+Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.** Symbols are colored according to the number of records at each site.
+Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
 
 \newpage
-```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
-```
+![**Figure `r fig_histograms`.
+Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.** For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)
 
@@
 # Conclusions
-<!--The conclusion goes here.
-You can modify the section name with  `\conclusions[modified heading if necessary]`.-->
 
@@
 We view the system developed here for submitting ForC data to the IPCC EFDB as one important step towards that goal.
+
+# Acknowledgements
+  We gratefully acknowledge the substantial contributions of Valentyna Slivinska and Sandro Federici for collaboration on the conception, design, and technical review of this project.
+  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have contributed to building the database.
+  Thank you to Avni Malhotra for helpful comments on an earlier draft of this manuscript.
+  Funding for this study was provided by the Smithsonian (Forest Global Earth Observatory, Smithsonian Working Land and Seascapes); a Bezos Earth Fund grant to the Nature Conservancy, with a sub-grant to NZCBI; and the Institute for Global Environmental Strategies.
+
+# Code and Data Availability
+    All code and data are openly available.
+The ForC database and associated code are available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) is archived in Zenodo [@anderson-teixeira_forcdb_2023, DOI: 10.5281/zenodo.8020861].
+The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).
+
+# Author contributions 
+  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
 
@@
 \* *Required* field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
+
+\clearpage
+# References

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
first pass at formatting for P3
#58

@@
 output: 
-    #pdf_document: 
     bookdown::word_document2: 
@@
 bibliography: references.bib
-csl: global-change-biology.csl
+csl: new-phytologist.csl
 
@@
 **Title:** Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
-
-*running title: Informing forest CO~2~ inventories with ForC*
 
@@
 **Affiliations:**
+
 1. Center for Conservation Ecology, Smithsonian's National Zoo & Conservation Biology Institute, Front Royal, VA, United States
@@
 
-5.The Nature Conservancy; Arlington VA 22203, USA
+5. The Nature Conservancy; Arlington VA 22203, USA
 
@@
 #Appendices
-table_ForCchanges = "Appendix A"
-table_ForCfieldmapping = "Appendix B"
+appendix_ForC_updates = "Methods S1"
+appendix_IPCC_definitions = "Notes S1"
+table_ForCchanges = "Table S1"
+table_ForCfieldmapping = "Table S2"
 
@@
 
-**Abstract.**
+\newpage
+
+# Summary {-}
+<!--The Summary for research articles must not exceed 250 words and should be organised using four bullet points to indicate (1) the research conducted or aims, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion.
+It should not contain citations of other papers.
+For reviews, keep to the word limit, but bullet points do not apply.
+-->
+
+*(NEEDS REVISION)*
+
 Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
@@
 
-\clearpage
+**Keywords:**
+<!--Five to eight keywords (in alphabetical order) must be given at the end of the Summary.-->
+
+# Societal Impact Statement {-}
+
+<!--Please provide a Societal Impact Statement to accompany your manuscript.
+Societal Impact Statements should outline the wider impact or relevance of the work for people, society or the planet, including any implications for policy and/or practice.
+This engaging 100 word statement should be written in plain language to be accessible for non-experts.-->
+
+\newpage
 
@@
 
-# IPCC methods and definitions
-
-The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
-Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
-
-## Carbon pools
-
-Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- scientific studies.
-IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
-While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
-
-```{r table_pools, eval = TRUE, echo=FALSE, warning=FALSE}
-C_pools_table <- read.csv("figures_tables/C_pools.csv")
-names(C_pools_table) <- stringr::str_replace_all(names(C_pools_table), "\\.", " ")
-
-library(knitr)
-library(kableExtra)
-knitr::kable(C_pools_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
-  row_spec(0, bold = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-### Biomass
-Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].
-
-Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
-A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can usually be considered a minor component of aboveground biomass for these forests.
-In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.
-While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [e.g., @fukushima_recovery_2007].
-Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
-
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001;@calders_laser_2022].
-While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
-IPCC Tier 1 values currently draw on studies applying a variety of allometric models [e.g., @requenasuarez_estimating_2019; @rozendaal_aboveground_2022].
-
-Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
-The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, and tending to be very uncertain [e.g., @keller_biomass_2001].
-Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
-Field methods for estimating root biomass are highly variable [@freschet_starting_2021].
-IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].
-
-### Dead Organic Matter
-
-Dead organic matter includes all non-living biomass larger than the litter size threshold. 
-Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
-
-Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_good_2003;@ipcc_2006_2006].
-This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
-Dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance, @carmona_coarse_2002].
-However, aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
-In turn, dead wood pools are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
-As noted above, live fine roots may be included in litter when difficult to separate empirically.
-The definition includes the entire O horizon, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
-This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
-Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-### Soil Organic Matter/ Carbon
-
-Soil organic matter/ carbon (SOM/ SOC), which is typically >100 t C and can exceed 300 t C in the top two meters of soil [@sanderman_soil_2017], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
-Live fine roots may be included with soil organic matter when it is not feasible to distinguish them empirically.
-The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.
-
-## Land classification
-IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
-Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
-Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
-Definitions of forest are allowed to vary by country, but must be applied consistently.
-Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
-
-The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
-This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
-However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
-
-# Updates to ForC (ForC v4.0)
-
-This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
-Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement.
-However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C inventories under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
-
-## New or modified fields
-
-We added or modified a total of `r n_fields_modified` fields (`r table_ForCchanges`).
-Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[@wickham_rvest_2022]. 
-That information was manually retrieved when the web scraping failed.
-
-## New variables
-
-To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables (Fig.
-`r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
-The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
-These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
-Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
-Although most of these variables lacked records in ForC as of `r format(Sys.time(), '%B %d, %Y')`, their addition gave the structure such that records can be populated over time. 
-Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).
-
-![**Figure `r fig_variable_mapping`.
-Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.** For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
-Most, but not all, EFDB-relevant ForC variables are shown here.
-Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
-**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable.](figures_tables/C_variable_mapping.png)
-
-## Quality control measures
-Prior to releasing ForC v4.0, we executed several quality control measures.
-First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_implementing_2022] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
-Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
-Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version 2.1 [@vandepol_identifying_2016; @bailey_climwin_2016] based on site coordinates. 
-Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or submission to EFDB.
-
-Because ForC v4.0 contained known duplicate records, we used R scripts to identify likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
-Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
-All records sent to EFDB were ensured to be independent and original through manual review, as detailed below.
-
-## Manual review of records to be sent to EFDB
-
-EFDB data submissions required information that was not recorded in previous versions of ForC, but for which new fields were created for EFDB compatibility (`r table_ForCchanges`).
-It was therefore necessary to return to original publications to retrieve relevant information, including 
-(1) estimates in original units, 
-(2) confidence intervals (when not already in ForC),
-(3) whether records of interest were presented in tables or text or digitized from figures (EFDB will not accept digitized data), 
-(4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass)
-We also checked that existing ForC records were complete and correct.
-
-Manual review of records was the limiting step for data submission to EFDB. 
-We prioritized review of
-(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021],
-(2) studies with confidence intervals recorded in ForC (because uncertainty estimates are important to the IPCC),
-(3) original publications containing large numbers of EFDB-relevant records,
-and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@anderson-teixeira_carbon_2021; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
-
-## Addition of new records
-
-In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
-These included 104 records<!--100 from Piponiot, 4 from Lutz--> from two studies [@piponiot_distribution_2022; @lutz_largediameter_2021] that were not previously included in ForC.
-In addition, we created new records for `r n_records_added-104` EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
-
-# Submission of ForC data to EFDB
-
-To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been submitted to EFDB.
-Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB. 
-Complete records needed to be reviewed by the EFDB editorial board and then posted in the database -- a process that lags behind submission of records and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
+# Materials and Methods
 
@@
 
+```{r table_pools, eval = TRUE, echo=FALSE, warning=FALSE}
+C_pools_table <- read.csv("figures_tables/C_pools.csv")
+names(C_pools_table) <- stringr::str_replace_all(names(C_pools_table), "\\.", " ")
+
+library(knitr)
+library(kableExtra)
+knitr::kable(C_pools_table, format = "latex",
+             longtable = FALSE,
+             caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
+  row_spec(0, bold = TRUE) %>%
+  kableExtra::kable_styling(full_width = TRUE) 
+  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
+
+![**Figure `r fig_variable_mapping`.
+Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.** For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
+Most, but not all, EFDB-relevant ForC variables are shown here.
+Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
+**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable.](figures_tables/C_variable_mapping.png)
+
 ### Land classification
@@
 However, we did provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
+
+## Updating ForC
+
+Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement.
+However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C inventories under IPCC guidelines.
+To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
+Having implemented these changes to ForC v3.0 [@anderson-teixeira_carbon_2021], which are described in detail in `r appendix_ForC_updates`, we released a new major version: ForC v4.0.
+
+## Submission of ForC data to EFDB
+
+To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
+Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been submitted to EFDB.
+Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB. 
+Complete records needed to be reviewed by the EFDB editorial board and then posted in the database -- a process that lags behind submission of records and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
 
@@
 
-# Recommendations
+# Discussion
+
+** RECOMENDATIONS **
 
@@
 
-# Code and Data Availability
+# Author Contribution
+  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
+
+# Data Availability Statement 
 All code and data are openly available.
@@
 
-# Author contributions 
-  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
+\clearpage
 
-\clearpage
+# `r appendix_IPCC_definitions`. IPCC methods and definitions
+
+The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
+For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
+Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
+
+## Carbon pools
+
+Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- scientific studies.
+IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
+While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
+
+### Biomass
+Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].
+
+Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
+IPCC's guidance is that the understory may be excluded if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
+A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can usually be considered a minor component of aboveground biomass for these forests.
+In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.
+While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [e.g., @fukushima_recovery_2007].
+Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
+
+A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001;@calders_laser_2022].
+While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
+IPCC Tier 1 values currently draw on studies applying a variety of allometric models [e.g., @requenasuarez_estimating_2019; @rozendaal_aboveground_2022].
+
+Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
+The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, and tending to be very uncertain [e.g., @keller_biomass_2001].
+Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
+Field methods for estimating root biomass are highly variable [@freschet_starting_2021].
+IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].
+
+### Dead Organic Matter
+
+Dead organic matter includes all non-living biomass larger than the litter size threshold. 
+Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
+
+Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_good_2003;@ipcc_2006_2006].
+This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
+Dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance, @carmona_coarse_2002].
+However, aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
+In turn, dead wood pools are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
+
+Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
+As noted above, live fine roots may be included in litter when difficult to separate empirically.
+The definition includes the entire O horizon, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
+This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
+Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
+
+### Soil Organic Matter/ Carbon
+
+Soil organic matter/ carbon (SOM/ SOC), which is typically >100 t C and can exceed 300 t C in the top two meters of soil [@sanderman_soil_2017], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
+Live fine roots may be included with soil organic matter when it is not feasible to distinguish them empirically.
+The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.
+
+## Land classification
+IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
+Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
+Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
+Definitions of forest are allowed to vary by country, but must be applied consistently.
+Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
+
+The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
+This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
+However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
+
+# `r appendix_ForC_updates`. Updates to ForC (ForC v4.0)
+
+This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
+
+## New or modified fields
+
+We added or modified a total of `r n_fields_modified` fields (`r table_ForCchanges`).
+Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
+For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[@wickham_rvest_2022]. 
+That information was manually retrieved when the web scraping failed.
+
+## New variables
+
+To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables (Fig.
+`r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
+The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
+These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
+Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
+Although most of these variables lacked records in ForC as of `r format(Sys.time(), '%B %d, %Y')`, their addition gave the structure such that records can be populated over time. 
+Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).
+
+## Quality control measures
+Prior to releasing ForC v4.0, we executed several quality control measures.
+First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_implementing_2022] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
+Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
+Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version 2.1 [@vandepol_identifying_2016; @bailey_climwin_2016] based on site coordinates. 
+Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or submission to EFDB.
+
+Because ForC v4.0 contained known duplicate records, we used R scripts to identify likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
+Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
+All records sent to EFDB were ensured to be independent and original through manual review, as detailed below.
+
+## Manual review of records to be sent to EFDB
+
+EFDB data submissions required information that was not recorded in previous versions of ForC, but for which new fields were created for EFDB compatibility (`r table_ForCchanges`).
+It was therefore necessary to return to original publications to retrieve relevant information, including 
+(1) estimates in original units, 
+(2) confidence intervals (when not already in ForC),
+(3) whether records of interest were presented in tables or text or digitized from figures (EFDB will not accept digitized data), 
+(4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass)
+We also checked that existing ForC records were complete and correct.
+
+Manual review of records was the limiting step for data submission to EFDB. 
+We prioritized review of
+(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021],
+(2) studies with confidence intervals recorded in ForC (because uncertainty estimates are important to the IPCC),
+(3) original publications containing large numbers of EFDB-relevant records,
+and (4) records from tropical regions.
+The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@anderson-teixeira_carbon_2021; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
+
+## Addition of new records
+
+In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
+These included 104 records<!--100 from Piponiot, 4 from Lutz--> from two studies [@piponiot_distribution_2022; @lutz_largediameter_2021] that were not previously included in ForC.
+In addition, we created new records for `r n_records_added-104` EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
 

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
update refs

@@
 For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
-Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labriere_forest_2023].
+Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labrière_forest_2023].
 
@@
 Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labriere_forest_2023;@calders_laser_2022].
+Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023;@calders_laser_2022].
 Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
@@
 Some examples are as follows:
-(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion while supporting FAIR (http://dx.doi.org/10.1038/d41586-019-01720-7) goals;
+(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion while supporting FAIR goals [@stall_make_2019];
 (2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
Update ForC-EFDB manuscript.Rmd

@@
 *corresponding author: teixeirak@si.edu
+
+<!--
+special issue: Anthromes and terrestrial carbon: from the deep past to net-zero
+special issue description: https://www.newphytologist.org/news/view/332#submission-procedure
+journal: Plants, People, Planet
+author guidelines: https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/hub/journal/25722611/about/author-guidelines
+Manuscript Category: Methods and Techniques
+Length: usually in the region of 3000–4000 words
+Display Items: 6
+-->
 

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
separate out SI material

@@
 \clearpage
-
-# `r appendix_IPCC_definitions`. IPCC methods and definitions
-
-The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
-Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
-
-## Carbon pools
-
-Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- scientific studies.
-IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
-While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
-
-### Biomass
-Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].
-
-Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
-A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can usually be considered a minor component of aboveground biomass for these forests.
-In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.
-While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [e.g., @fukushima_recovery_2007].
-Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
-
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001;@calders_laser_2022].
-While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
-IPCC Tier 1 values currently draw on studies applying a variety of allometric models [e.g., @requenasuarez_estimating_2019; @rozendaal_aboveground_2022].
-
-Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
-The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, and tending to be very uncertain [e.g., @keller_biomass_2001].
-Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
-Field methods for estimating root biomass are highly variable [@freschet_starting_2021].
-IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].
-
-### Dead Organic Matter
-
-Dead organic matter includes all non-living biomass larger than the litter size threshold. 
-Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
-
-Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_good_2003;@ipcc_2006_2006].
-This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
-Dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance, @carmona_coarse_2002].
-However, aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
-In turn, dead wood pools are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
-As noted above, live fine roots may be included in litter when difficult to separate empirically.
-The definition includes the entire O horizon, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
-This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
-Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-### Soil Organic Matter/ Carbon
-
-Soil organic matter/ carbon (SOM/ SOC), which is typically >100 t C and can exceed 300 t C in the top two meters of soil [@sanderman_soil_2017], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
-Live fine roots may be included with soil organic matter when it is not feasible to distinguish them empirically.
-The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.
-
-## Land classification
-IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
-Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
-Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
-Definitions of forest are allowed to vary by country, but must be applied consistently.
-Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
-
-The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
-This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
-However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
-
-# `r appendix_ForC_updates`. Updates to ForC (ForC v4.0)
-
-This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
-
-## New or modified fields
-
-We added or modified a total of `r n_fields_modified` fields (`r table_ForCchanges`).
-Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[@wickham_rvest_2022]. 
-That information was manually retrieved when the web scraping failed.
-
-## New variables
-
-To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables (Fig.
-`r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
-The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
-These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
-Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
-Although most of these variables lacked records in ForC as of `r format(Sys.time(), '%B %d, %Y')`, their addition gave the structure such that records can be populated over time. 
-Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).
-
-## Quality control measures
-Prior to releasing ForC v4.0, we executed several quality control measures.
-First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_implementing_2022] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
-Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
-Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version 2.1 [@vandepol_identifying_2016; @bailey_climwin_2016] based on site coordinates. 
-Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or submission to EFDB.
-
-Because ForC v4.0 contained known duplicate records, we used R scripts to identify likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
-Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
-All records sent to EFDB were ensured to be independent and original through manual review, as detailed below.
-
-## Manual review of records to be sent to EFDB
-
-EFDB data submissions required information that was not recorded in previous versions of ForC, but for which new fields were created for EFDB compatibility (`r table_ForCchanges`).
-It was therefore necessary to return to original publications to retrieve relevant information, including 
-(1) estimates in original units, 
-(2) confidence intervals (when not already in ForC),
-(3) whether records of interest were presented in tables or text or digitized from figures (EFDB will not accept digitized data), 
-(4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass)
-We also checked that existing ForC records were complete and correct.
-
-Manual review of records was the limiting step for data submission to EFDB. 
-We prioritized review of
-(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021],
-(2) studies with confidence intervals recorded in ForC (because uncertainty estimates are important to the IPCC),
-(3) original publications containing large numbers of EFDB-relevant records,
-and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@anderson-teixeira_carbon_2021; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
-
-## Addition of new records
-
-In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
-These included 104 records<!--100 from Piponiot, 4 from Lutz--> from two studies [@piponiot_distribution_2022; @lutz_largediameter_2021] that were not previously included in ForC.
-In addition, we created new records for `r n_records_added-104` EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
-
-# Appendix A. Updates to ForC {-}
-
-\captionsetup[table]{labelformat=empty}
-Table A1: **Table of changes to ForC fields.** These are changes implemented between releases of v3.0 and v4.0. 
-```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
-library (readr)
-urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
-record_of_changes <-read_csv(url(urlfile))
-names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
-record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
-#record_of_changes<- record_of_changes[which(!grep("housekeeping",record_of_changes$Motivation)) , c(1:5)]  #trying to exclude rows for which Motivation = "housekeeping", but this is not working
-record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
-
-library(knitr)
-library(kableExtra)
-knitr::kable(record_of_changes, format = "latex",
-             longtable = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
-  kable_styling(font_size = 8) %>%
-  row_spec(0, bold = TRUE)
-```
-
-\clearpage
-
-# Appendix B. Mapping ForC to EFDB {-}
-
-Table B1: **Mapping of ForC fields to EFDB.** Details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
-```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-
-ForC_variables_included <- read.csv("figures_tables/ForC_fields_included.csv")
-names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
-column_names <- names(ForC_variables_included) # Define column names
-column_names[5] <- "Required*" # Add asterisk for footnote
-names(ForC_variables_included) <- column_names# Set column names of the matrix
-ForC_variables_included <- ForC_variables_included[c(1:5)]
-
-knitr::kable(ForC_variables_included, format = "latex",
-             longtable = TRUE) %>%
-  row_spec(0, bold = TRUE) %>%
-  kable_styling(font_size = 8) %>%
-kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
-
-```
-\* *Required* field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
-
-\clearpage
 # References

### 2023-10-26 doc/manuscript/SI_ForCv40.Rmd
separate out SI material

--- 2023-06-09 doc/manuscript/ForC-EFDB manuscript for ESSD.Rmd
@@
 ---
-title: Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
-journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
-author:
-  - given_name: Kristina J.
-    surname: Anderson-Teixeira
-    affiliation: "1,2 *"
-    email: teixeirak@si.edu
-    corresponding: true
-  - given_name: Valentine
-    surname: Herrmann
-    affiliation: "1"
-  - given_name: Madison 
-    surname: Williams
-    affiliation: "1"
-  - given_name: Teagan
-    surname: Rogers
-    affiliation: "1"
-  - given_name: Rebecca
-    surname: Banbury Morgan
-    affiliation: "1,3" 
-  - given_name: Ben 
-    surname: Bond-Lamberty
-    affiliation: "4" 
-  - given_name: Susan 
-    surname: Cook-Patton
-    affiliation: "5"
+title: ''
+output: 
+    bookdown::word_document2: 
+      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+bibliography: references.bib
+csl: new-phytologist.csl
 
-affiliation:
-  - code: 1
-    address: Center for Conservation Ecology, Smithsonian's National Zoo & Conservation Biology Institute, Front Royal, VA, United States
-  - code: 2
-    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
-  - code: 3
-    address: School of Geography, University of Leeds, Leeds, UK
-  - code: 4
-    address: Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
-  - code: 5
-    address: The Nature Conservancy; Arlington VA 22203, USA
-
-abstract: |
-
-bibliography: references.bib
-running:
-  title: Informing forest CO~2~ inventories with ForC
-  author: Anderson-Teixeira et al.
-competinginterests: |
-  The authors declare no competing interests.
-# OPTIONAL:
-algorithms: true
-# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
-# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
-copyrightstatement: |
-### The author's copyright for this publication is transferred to institution/company. 
-### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
-availability:
-  #  use this to add a statement when having only software code available
-  #  use this to add a statement when having only data sets available
-  #code: |
-  #data: |
-  codedata: |
-    All code and data are openly available.
-The ForC database and associated code is available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD).
-The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).
-authorcontribution: |
-  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
-disclaimer: |
-acknowledgements: |
-  We gratefully acknowledge the substantial contributions of Valentyna Slivinska and Sandro Federici for collaboration on the conception, design, and technical review of this project.
-  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have contributed to building the database.
-  Thank you to Avni Malhotra for helpful comments on an earlier draft of this manuscript.
-  Funding for this study was provided by the Smithsonian (Forest Global Earth Observatory, Smithsonian Working Land and Seascapes); a Bezos Earth Fund grant to the Nature Conservancy, with a sub-grant to NZCBI; and the Institute for Global Environmental Strategies.
-appendix: 
-output:
-  rticles::copernicus_article: default
-  bookdown::pdf_book:
-    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
-
-    header-includes:
-    - \usepackage{caption}
-    #- \usepackage{booktabs}
-    #- \usepackage{makecell}
 ---
-
-<!--**Important**: Always double-check with the official manuscript preparation guidelines at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html), especially the sections "Technical instructions for LaTeX" and "Manuscript composition".
-Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.-->
 
@@
 
-#Cited sections (listed here because these can change)
-section_C_pools = 2.1 
-section_land_use_categories = 2.2
-section_new_variables = 3.2
-section_C_variable_mapping = "4.1.1"
-section_land_use_mapping = "4.1.2"
-section_database_needs = "6.1"
-section_new_data_needs = "6.2"
-section_data_reporting_needs = "6.3"
-
 #Appendices
-table_ForCchanges = "Appendix A"
-table_ForCfieldmapping = "Appendix B"
+appendix_ForC_updates = "Methods S1"
+appendix_IPCC_definitions = "Notes S1"
+table_ForCchanges = "Table S1"
+table_ForCfieldmapping = "Table S2"
 
@@
 
-**Abstract.**
-Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
-However, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
-Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
-Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas inventories if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
-We begin by reconciling terminology and mapping ForC fields into EFDB.
-This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have undergone necessary review and been submitted to date.
-Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into or out of the IPCC-defined C pools.
-Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
-In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
-Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
+# `r appendix_ForC_updates`. Updates to ForC (ForC v4.0) {-}
 
-\introduction[Introduction]
+Here, we describe changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021], which were implemented prior to the release of ForC v4.0.
 
-Forests are critical to management of the atmospheric concentration of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
-In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
-Table 2--> [forests only, @harris_global_2021].
-This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
-Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
-The future of this important CO~2~ sink is highly uncertain, and depends upon both forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
-
-Forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
-Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
-As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
-Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
-One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
-
-The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
-Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
-The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
-A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
-Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
-For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
-In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
-This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
-Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.
-
-To improve data accessibility for preparing greenhouse gas estimates, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
-The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
-The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include C stocks, net annual increments, and annual fluxes for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
-
-The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
-ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
-As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
-The purpose of this publication is to document that process and provide recommendations for future improvements.
-
-Here, we
-(1) review IPCC methods and definitions applied to estimate CO~2~ emissions and removals from forest in the context of typical forest C estimation methodologies;
-(2) describe mapping of ForC to IPCC's EFDB;
-(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data submission to EFDB;
-(4) summarize the data in ForC relevant to EFDB and records that have been submitted to date; and
-(5) provide recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
-
-# IPCC methods and definitions
-
-The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
-Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
-
-## Carbon pools
-
-Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- scientific studies.
-IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
-While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
-
-```{r table_pools, eval = TRUE, echo=FALSE, warning=FALSE}
-C_pools_table <- read.csv("figures_tables/C_pools.csv")
-names(C_pools_table) <- stringr::str_replace_all(names(C_pools_table), "\\.", " ")
-
-library(knitr)
-library(kableExtra)
-knitr::kable(C_pools_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
-  row_spec(0, bold = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-### Biomass
-Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].
-
-Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
-A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can usually be considered a minor component of aboveground biomass for these forests.
-In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.
-While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [e.g., @fukushima_recovery_2007].
-Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
-
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001;@calders_laser_2022].
-While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
-IPCC Tier 1 values currently draw on studies applying a variety of allometric models [e.g., @requenasuarez_estimating_2019; @rozendaal_aboveground_2022].
-
-Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
-The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, and tending to be very uncertain [e.g., @keller_biomass_2001].
-Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
-Field methods for estimating root biomass are highly variable [@freschet_starting_2021].
-IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].
-
-### Dead Organic Matter
-
-Dead organic matter includes all non-living biomass larger than the litter size threshold. 
-Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
-
-Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_good_2003;@ipcc_2006_2006].
-This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
-Dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance, @carmona_coarse_2002].
-However, aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
-In turn, dead wood pools are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
-As noted above, live fine roots may be included in litter when difficult to separate empirically.
-The definition includes the entire O horizon, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
-This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
-Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-### Soil Organic Matter/ Carbon
-
-Soil organic matter/ carbon (SOM/ SOC), which is typically >100 t C and can exceed 300 t C in the top two meters of soil [@sanderman_soil_2017], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
-Live fine roots may be included with soil organic matter when it is not feasible to distinguish them empirically.
-The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.
-
-## Land classification
-IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
-Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
-Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
-Definitions of forest are allowed to vary by country, but must be applied consistently.
-Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
-
-The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
-This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
-However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
-
-# Updates to ForC (ForC v4.0)
-
-This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
-Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement.
-However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C inventories under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
-
-## New or modified fields
+## New or modified fields {-}
 
@@
 Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[ @wickham_rvest_2022]. 
+For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[@wickham_rvest_2022]. 
 That information was manually retrieved when the web scraping failed.
 
-## New variables
+## New variables {-}
 
-To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables (Fig.
-`r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
+To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables, counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
 The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
@@
 
-```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
-Most, but not all, EFDB-relevant ForC variables are shown here.
-Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
-**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/C_variable_mapping.png")
-```
-
-## Quality control measures
+## Quality control measures {-}
 Prior to releasing ForC v4.0, we executed several quality control measures.
@@
 
-## Manual review of records to be sent to EFDB
+## Manual review of records to be sent to EFDB {-}
 
@@
 
-## Addition of new records
+## Addition of new records {-}
 
@@
 
-# Submission of ForC data to EFDB
-
-To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been submitted to EFDB.
-Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB. 
-Complete records needed to be reviewed by the EFDB editorial board and then posted in the database -- a process that lags behind submission of records and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
-
-## Mapping ForC to EFDB
-
-The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`.
-For the majority of fields, contents of the field in ForC was copied directly into an EFDB field, either as the only contents of that field or as part of a composite record.
-For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).
-In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*).
-For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
-For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
-There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) land classification (section `r section_land_use_mapping`).
-
-### Carbon cycle variables
-
-With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig.
-`r fig_variable_mapping`).
-For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, net annual increments, influxes (a.k.a. "gross annual increments" by IPCC), and outfluxes.
-As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
-It is important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
-For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
-Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
-Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
-
-### Land classification
-
-Determination of the IPCC land-use category (i.e., Forest Land, Grassland, Wetlands, Cropland, Settlements, or Other Land; section `r  section_land_use_categories`) was made based on the categorical ForC field *dominant.life.form*, sometimes drawing upon stand age.
-Records with "woody" *dominant.life.form* were classified as Forest Land.
-Those with *dominant.life.form* of "woody+grass", which in ForC is indicative of anything from a shrub-encroached grassland to a tree-dominated savanna, were given dual classification of Forest Land and Grassland.
-This dual classification indicates that records may be relevant to either category depending on the definition of forest applied (varies by country).
-For (rare) cases where *dominant.life.form* was grass and stand age was greater than zero, indicative of early successional vegetation, we assigned a classification of Forest Land, consistent with the IPCC definition that Forest Land includes land expected to succeed to forest.
-Cases where *dominant.life.form* was grass or crop and stand age was zero were indicative of a control for studies of forest regrowth following agricultural abandonment, and were classified as Grassland and Cropland, respectively.
-
-Classification into sub-categories was dependent upon stand age and site history (section `r  section_land_use_categories`).
-For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
-For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Grassland converted to Forest Land", respectively.
-For forests <20 years old with unspecified previous agricultural use, we assigned the sub-category "Land Converted to Forest land".
-Forests <20 years old with unknown land use prior to the study date were simply classified as "Forest Land".
-The same logic was applied for savannas, but including both Forest Land and Grassland as potentially relevant categories.
-
-Given the lack of public information needed to determine whether lands are classified as managed [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not include any classification of land management status from ForC in the records submitted to EFDB.
-However, we did provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
-
-# Results
-
-## ForC v4.0 contents
-
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig.
-`r fig_variable_mapping`).
-These records were distributed across all forested continents and ecozones, albeit unevenly (Fig. `r fig_map`).
-The largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig. `r fig_histograms`c).
-Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
-`r fig_histograms`b).
-Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
-The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
-In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).
-
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png")
-```
-
-\newpage
-```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
-```
-
-ForC contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
-The records were very unevenly distributed across variables. 
-The variable with most records was aboveground biomass, representing `r round(n_ind_records_agb/n_ind_records_EFDB_variables,2)*100`% of all independent records relevant to EFDB, and aboveground biomass components (woody biomass or foliage) representing an additional `r round(n_ind_records_agb_components/n_ind_records_EFDB_variables,2)*100`%.
-A total of `r round(n_ind_records_rootbiomass/n_ind_records_EFDB_variables,2)*100`% of relevant records were for root biomass (including fine and coarse root components), while `r round(n_ind_records_biomass/n_ind_records_EFDB_variables,2)*100`% described total biomass.
-The non-living pools were less represented, with `r round(n_ind_records_deadwood/n_ind_records_EFDB_variables,2)*100`% of relevant were for dead wood (including standing and fallen components), `r round(n_ind_records_OL/n_ind_records_EFDB_variables*100,1)`% for litter, `r round(n_ind_records_TotEcosystemC/n_ind_records_EFDB_variables*100,1)`% for total ecosystem C excluding soils, and `r round(n_ind_records_SOC/n_ind_records_EFDB_variables*100,1)`% for soil carbon.
-
-Increment and flux variables were poorly represented (Table `r table_variables`).
-The increment variable with most records was the aboveground biomass increment, representing `r round(n_ind_records_deltaagb/n_ind_records_EFDB_variables*100,1)`% of all independent records relevant to EFDB.
-The only other relevant increment variable with any records was the O horizon (litter) increment, with just `r n_ind_records_deltaOL` records.
-Relevant flux variable records (n=`r n_ind_records_flux`) together constituted `r round(n_ind_records_flux/n_ind_records_EFDB_variables,2)*100`% of ForC's independent records relevant to EFDB.
-
-\newpage
-```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
-C_variables_table <- read.csv("figures_tables/C_variables.csv")
-names(C_variables_table) <- stringr::str_replace_all(names(C_variables_table), "\\.", " ")
-C_variables_table[is.na(C_variables_table)]<- "" #replace NAs with blanks
-
-library(knitr)
-library(kableExtra)
-knitr::kable(C_variables_table[,-1], format = "latex",  
-             longtable = T,
-             caption = "\\textbf{Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.}") %>%
-  row_spec(0, bold = TRUE)  %>%
-  row_spec(which(C_variables_table$bold), bold = TRUE) %>% #total
-  kable_styling(font_size = 8) %>%
-  #kableExtra::kable_styling(full_width = TRUE) 
-  kable_styling(latex_options = c("hold_position", "repeat_header"), protect_latex = T) 
-```
-
-## Data submissions to EFDB
-
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were submitted to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`).
-The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
-The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review and post the records, and also because a minority of records were deemed not applicable to EFDB by the review panel.
-
-The ForC records submitted to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
-However, the density of these records was very unevenly distributed across continents, biomes, and forest types and was not proportional to the numbers of relevant records in ForC (Fig.
-`r fig_histograms`).
-Rather, the largest number of records came from North America, followed by Asia, South America, and Africa (Fig.
-`r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.
-`r fig_histograms`b).
-In terms of dominant vegetation, by far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
-The largest records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
-
-In terms of variables, records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
-The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records sent were for C stocks, including
-`r round(n_records_sent_biomass/n_records_sent_to_EFDB,2)*100`% for total biomass, 
-`r round(n_records_sent_agb/n_records_sent_to_EFDB,2)*100`% for aboveground biomass,
-`r round(n_records_sent_agb_components/n_records_sent_to_EFDB,2)*100`% for components of aboveground biomass (wood or foliage), 
-`r round(n_records_sent_rootbiomass/n_records_sent_to_EFDB,2)*100`% for root biomass, 
-`r round(n_records_sent_rootbiomass_components/n_records_sent_to_EFDB,2)*100`% for components of root biomass (coarse or fine roots), 
-`r round(n_records_sent_deadwood/n_records_sent_to_EFDB,2)*100`% for dead wood, 
-`r round(n_records_sent_deadwood_components/n_records_sent_to_EFDB,2)*100`% for components of dead wood (standing or fallen), 
-`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (entire O horizon or OL layer component),
-and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
-Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, virtually all for aboveground biomass (excepting `r n_records_sent_delta.litter` records for delta.O.horizon).
-The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records sent described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
-
-# Recommendations
-
-Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work (section `r section_database_needs`), new data collection and analysis (section `r section_new_data_needs`), and reporting (section `r section_data_reporting_needs`).
-
-## Database needs
-
-There is vast potential to expand forest C data in EFDB by completing the process of reviewing and submitting data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
-So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been submitted to EFDB.
-Although this process requires manual review of records, the submission of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
-Future efforts to review studies for submission should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
-
-In addition to the large potential to expand EFDB using records already in ForC, there are innumerous published EFDB-relevant forest C data that are not currently included in ForC, with more being published on a nearly daily basis.
-Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
-Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
-Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
-
-## Data collection and analysis needs
-
-New data collection and analysis is needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
-Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
-Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
-For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
-Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labriere_forest_2023].
-
-Several EFDB-relevant variables have not been calculated and presented as frequently as would be possible given existing forest census data and minimal extra research effort.
-For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022].
-Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
-However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017; @mokany_critical_2006].
-In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate estimates. 
-We recommend that, when possible, researchers calculate and report these variables, following the reporting guidelines specified in section `r section_data_reporting_needs`.
-
-Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labriere_forest_2023;@calders_laser_2022].
-Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
-Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
-Significant investment in ground-based forest research will be critical to filling these gaps.
-
-## Data reporting needs
-
-We recommend that, in order to make research valuable to estimate C stock changes according to methods provided in the IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
-It is particularly noteworthy that simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
-Some examples are as follows:
-(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion while supporting FAIR (http://dx.doi.org/10.1038/d41586-019-01720-7) goals;
-(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;
-(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant estimating forest C stock changes according to IPCC guidelines (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
-It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
-
-```{r table_recommendations, eval = TRUE, echo=FALSE, warning=FALSE}
-recommendations_table <- read.csv("figures_tables/recommendations.csv")
-names(recommendations_table) <- stringr::str_replace_all(names(recommendations_table), "\\.", " ")
-
-library(knitr)
-library(kableExtra)
-knitr::kable(recommendations_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}") %>%
-  row_spec(0, bold = TRUE)%>%
-  kable_styling(font_size = 10) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-For those compiling published records (e.g., for meta-analyses), the data set can have added value if all information required by EFDB is extracted from original publications. 
-This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
-The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.
-
-Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipeline developed here.
-For individual publications, the former option will generally be more efficient.
-However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mitigation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].
-
-# Conclusions
-<!--The conclusion goes here.
-You can modify the section name with  `\conclusions[modified heading if necessary]`.-->
-
-The ForC database contains large numbers of records that could potentially be useful for estimating C stock changes applying methodological guidance provided by the IPCC.
-Here we have developed a framework for submitting these records to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
-As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
-Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
-The records submitted to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.
-
-Going forward, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
-In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under teh Paris Agreement.
-This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may no longer be relevant.
-This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.
-We view the system developed here for submitting ForC data to the IPCC EFDB as one important step towards that goal.
-
-\clearpage
-
-# Appendix A. Updates to ForC {-}
-
-\captionsetup[table]{labelformat=empty}
-Table A1: **Table of changes to ForC fields.** These are changes implemented between releases of v3.0 and v4.0. 
-```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
-library (readr)
-urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
-record_of_changes <-read_csv(url(urlfile))
-names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
-record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
-#record_of_changes<- record_of_changes[which(!grep("housekeeping",record_of_changes$Motivation)) , c(1:5)]  #trying to exclude rows for which Motivation = "housekeeping", but this is not working
-record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
-
-library(knitr)
-library(kableExtra)
-knitr::kable(record_of_changes, format = "latex",
-             longtable = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
-  kable_styling(font_size = 8) %>%
-  row_spec(0, bold = TRUE)
-```
-
-\clearpage
-
-# Appendix B. Mapping ForC to EFDB {-}
-
-Table B1: **Mapping of ForC fields to EFDB.** Details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
-```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-
-ForC_variables_included <- read.csv("figures_tables/ForC_fields_included.csv")
-names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
-column_names <- names(ForC_variables_included) # Define column names
-column_names[5] <- "Required*" # Add asterisk for footnote
-names(ForC_variables_included) <- column_names# Set column names of the matrix
-ForC_variables_included <- ForC_variables_included[c(1:5)]
-
-knitr::kable(ForC_variables_included, format = "latex",
-             longtable = TRUE) %>%
-  row_spec(0, bold = TRUE) %>%
-  kable_styling(font_size = 8) %>%
-kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
-
-```
-\* *Required* field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
+## References

### 2023-10-26 doc/manuscript/SI_IPCCdefs.Rmd
separate out SI material

--- 2023-06-09 doc/manuscript/ForC-EFDB manuscript for ESSD.Rmd
@@
 table_pools =1
-table_variables =2
-table_recommendations = 3
-
-#Figures
-fig_variable_mapping=1
-fig_map=2
-fig_histograms=3
-
-#Cited sections (listed here because these can change)
-section_C_pools = 2.1 
-section_land_use_categories = 2.2
-section_new_variables = 3.2
-section_C_variable_mapping = "4.1.1"
-section_land_use_mapping = "4.1.2"
-section_database_needs = "6.1"
-section_new_data_needs = "6.2"
-section_data_reporting_needs = "6.3"
 
 #Appendices
-table_ForCchanges = "Appendix A"
-table_ForCfieldmapping = "Appendix B"
-
-#READING IN FILES TO GET FIGURES/ TABLES/ STATS 
-path_to_dir <- dirname(dirname(getwd())) # get path to repo 
-path_to_ForC= "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db" #Krista's laptop
-# path_to_ForC= "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db" #Madison, to knit, you'll need to enter the path to ForC-db on your computer, then un-comment this line.
-
-# MOVE FIGURE FILES 
-#world map from GitHub
-do.call(file.remove, list(list.files(paste0(path_to_dir, "/doc/manuscript/figures_tables"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE))) # first remove old file
-flist <- list.files(paste0(path_to_ForC, "/ForC/figures/"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE)
-file.copy(flist, paste0(path_to_dir, "/doc/manuscript/figures_tables"))
-
-#Statistics
-ForC_measurements <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_measurements.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_records_ForC= length(ForC_measurements[,"measurement.ID"])  # total n records in ForC
-
-ForC_plots <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_plots.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_plots_ForC = length(ForC_plots[,"plot.ID"]) # total n plots in ForC
-
-ForC_sites <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_sites.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_areas_ForC = max(ForC_sites$geographic.area, na.rm = TRUE) # total n geographic areas in ForC
-
-ForC_simplified <- read.csv(paste0(path_to_ForC,"/ForC/ForC_simplified/ForC_simplified.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_independent_records_ForC= length(ForC_simplified[,"measurement.ID"]) #suspected duplicates removed. This is the number of records fin ForC simplified.
-
-C_variables <- read.csv(paste0(path_to_dir,"/doc/manuscript/figures_tables/C_variables.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_relevant_variables = nrow(C_variables) - 8  # n variables relevant = rows in C_variables - headers and total 
-n_relevant_variables_with_no_data =  sum(C_variables$`n in ForC` == 0,  na.rm = TRUE)
-n_relevant_variables_with_data= n_relevant_variables - n_relevant_variables_with_no_data
-
-n_ind_records_EFDB_variables= C_variables[length(C_variables[,"variable"]), "n independent records in ForC"] #n independent records relevant to EFDB
-n_ind_records_agb <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag"]
-n_ind_records_agb_components <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag_foliage"]
-n_ind_records_biomass <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass"]
-n_ind_records_rootbiomass <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root_coarse"]
-n_ind_records_deadwood <- C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_down"]+C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_standing"]
-n_ind_records_OL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "litter"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "O.horizon"]
-n_ind_records_TotEcosystemC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "total.ecosystem_2"] 
-n_ind_records_SOC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "SOM / SOC"] 
-n_ind_records_deltaagb <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.agb"] 
-n_ind_records_deltaOL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.O.horizon"] 
-flux_variables_with_records <- c("NPP_woody", "ANPP_woody", "ANPP_woody_stem", "ANPP_woody_branch", "woody.mortality_ag", "BNPP_root", "BNPP_root_fine","BNPP_root.turnover_fine","BNPP_root_coarse" , "ANPP_litterfall", "NPP_litter", "R_het_litter")
-n_ind_records_flux <- sum(C_variables$`n independent records in ForC`[C_variables$`variable` %in% flux_variables_with_records])
-
-n_records_reviewed_or_added= C_variables[length(C_variables[,"variable"]), "n reviewed"]
-n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
-n_records_checked = n_records_reviewed_or_added - n_records_added # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
-
-n_records_sent_to_EFDB = C_variables[length(C_variables[,"variable"]), "n submitted to EFDB"] # n records sent to EFDB
-n_relevant_variables_no_records_sent =  sum(C_variables$`n submitted to EFDB` == 0,  na.rm = TRUE)
-n_relevant_variables_records_sent= n_relevant_variables - n_relevant_variables_no_records_sent
-n_records_sent_biomass<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass"]
-n_records_sent_agb <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag"]
-n_records_sent_agb_components <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag_foliage"]
-n_records_sent_rootbiomass <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root"] 
-n_records_sent_rootbiomass_components <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root_coarse"]
-n_records_sent_deadwood<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood"] 
-n_records_sent_deadwood_components<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_down"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_standing"]
-n_records_sent_litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "litter"] +C_variables$`n submitted to EFDB`[C_variables$`variable` == "O.horizon"]
-n_records_sent_soil<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "SOM / SOC"] 
-n_records_sent_stocks = n_records_sent_biomass+n_records_sent_agb+n_records_sent_agb_components+n_records_sent_rootbiomass+n_records_sent_rootbiomass_components+n_records_sent_deadwood+n_records_sent_deadwood_components+ n_records_sent_litter+ n_records_sent_soil
-n_records_sent_delta.agb<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.agb"]
-n_records_sent_delta.litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.O.horizon"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.litter"]
-n_records_sent_increments <- n_records_sent_delta.agb+n_records_sent_delta.litter
-n_records_sent_fluxes <- sum(C_variables$`n submitted to EFDB`[C_variables$`variable` %in% flux_variables_with_records])
-n_records_sent_fluxes_check = n_records_sent_to_EFDB-n_records_sent_stocks-n_records_sent_increments
-
-n_records_posted_EFDB = 376 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
-Then get number from upper left box where it says "displayed records")
-
-n_new_variables = 11 + 2 + 2 #delta. + woody.mortality + O horizon
-# get n fields modified
-sampling_details <- read.csv(paste0(path_to_ForC,"/ForC/database_management_records/record_of_changes.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_fields_modified =  length(sampling_details[,"Column"]) + 4
+appendix_ForC_updates = "Methods S1"
+appendix_IPCC_definitions = "Notes S1"
+table_ForCchanges = "Table S1"
+table_ForCfieldmapping = "Table S2"
 
@@
 
-**Abstract.**
-Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
-However, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
-Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
-Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas inventories if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
-We begin by reconciling terminology and mapping ForC fields into EFDB.
-This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have undergone necessary review and been submitted to date.
-Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into or out of the IPCC-defined C pools.
-Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
-In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
-Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
-
-\introduction[Introduction]
-
-Forests are critical to management of the atmospheric concentration of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
-In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
-Table 2--> [forests only, @harris_global_2021].
-This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
-Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
-The future of this important CO~2~ sink is highly uncertain, and depends upon both forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
-
-Forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
-Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
-As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
-Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
-One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
-
-The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
-Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
-The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
-A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
-Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
-For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
-In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
-This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
-Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.
-
-To improve data accessibility for preparing greenhouse gas estimates, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
-The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
-The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include C stocks, net annual increments, and annual fluxes for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
-
-The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
-ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
-As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
-The purpose of this publication is to document that process and provide recommendations for future improvements.
-
-Here, we
-(1) review IPCC methods and definitions applied to estimate CO~2~ emissions and removals from forest in the context of typical forest C estimation methodologies;
-(2) describe mapping of ForC to IPCC's EFDB;
-(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data submission to EFDB;
-(4) summarize the data in ForC relevant to EFDB and records that have been submitted to date; and
-(5) provide recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
-
-# IPCC methods and definitions
+# `r appendix_IPCC_definitions`. IPCC methods and definitions {-}
 
 The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
+For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (see **Carbon Pools** section below), plus any harvested wood products.
 Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
 
-## Carbon pools
+## Carbon Pools {-}
 
@@
 
-```{r table_pools, eval = TRUE, echo=FALSE, warning=FALSE}
-C_pools_table <- read.csv("figures_tables/C_pools.csv")
-names(C_pools_table) <- stringr::str_replace_all(names(C_pools_table), "\\.", " ")
-
-library(knitr)
-library(kableExtra)
-knitr::kable(C_pools_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
-  row_spec(0, bold = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-### Biomass
+### Biomass {-}
 Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].
@@
 
-### Dead Organic Matter
+### Dead Organic Matter {-}
 
 Dead organic matter includes all non-living biomass larger than the litter size threshold. 
-Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
+Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see **Land Classification** section below), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
 
@@
 
-### Soil Organic Matter/ Carbon
+### Soil Organic Matter/ Carbon {-}
 
@@
 
-## Land classification
+## Land classification {-}
 IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
@@
 
-# Updates to ForC (ForC v4.0)
-
-This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
-Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement.
-However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C inventories under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
-
-## New or modified fields
-
-We added or modified a total of `r n_fields_modified` fields (`r table_ForCchanges`).
-Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[ @wickham_rvest_2022]. 
-That information was manually retrieved when the web scraping failed.
-
-## New variables
-
-To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables (Fig.
-`r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
-The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
-These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
-Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
-Although most of these variables lacked records in ForC as of `r format(Sys.time(), '%B %d, %Y')`, their addition gave the structure such that records can be populated over time. 
-Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).
-
-```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
-Most, but not all, EFDB-relevant ForC variables are shown here.
-Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
-**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/C_variable_mapping.png")
-```
-
-## Quality control measures
-Prior to releasing ForC v4.0, we executed several quality control measures.
-First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_implementing_2022] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
-Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
-Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version 2.1 [@vandepol_identifying_2016; @bailey_climwin_2016] based on site coordinates. 
-Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or submission to EFDB.
-
-Because ForC v4.0 contained known duplicate records, we used R scripts to identify likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
-Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
-All records sent to EFDB were ensured to be independent and original through manual review, as detailed below.
-
-## Manual review of records to be sent to EFDB
-
-EFDB data submissions required information that was not recorded in previous versions of ForC, but for which new fields were created for EFDB compatibility (`r table_ForCchanges`).
-It was therefore necessary to return to original publications to retrieve relevant information, including 
-(1) estimates in original units, 
-(2) confidence intervals (when not already in ForC),
-(3) whether records of interest were presented in tables or text or digitized from figures (EFDB will not accept digitized data), 
-(4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass)
-We also checked that existing ForC records were complete and correct.
-
-Manual review of records was the limiting step for data submission to EFDB. 
-We prioritized review of
-(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021],
-(2) studies with confidence intervals recorded in ForC (because uncertainty estimates are important to the IPCC),
-(3) original publications containing large numbers of EFDB-relevant records,
-and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@anderson-teixeira_carbon_2021; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
-
-## Addition of new records
-
-In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
-These included 104 records<!--100 from Piponiot, 4 from Lutz--> from two studies [@piponiot_distribution_2022; @lutz_largediameter_2021] that were not previously included in ForC.
-In addition, we created new records for `r n_records_added-104` EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
-
-# Submission of ForC data to EFDB
-
-To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been submitted to EFDB.
-Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB. 
-Complete records needed to be reviewed by the EFDB editorial board and then posted in the database -- a process that lags behind submission of records and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
-
-## Mapping ForC to EFDB
-
-The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`.
-For the majority of fields, contents of the field in ForC was copied directly into an EFDB field, either as the only contents of that field or as part of a composite record.
-For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).
-In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*).
-For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
-For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
-There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) land classification (section `r section_land_use_mapping`).
-
-### Carbon cycle variables
-
-With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig.
-`r fig_variable_mapping`).
-For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, net annual increments, influxes (a.k.a. "gross annual increments" by IPCC), and outfluxes.
-As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
-It is important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
-For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
-Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
-Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
-
-### Land classification
-
-Determination of the IPCC land-use category (i.e., Forest Land, Grassland, Wetlands, Cropland, Settlements, or Other Land; section `r  section_land_use_categories`) was made based on the categorical ForC field *dominant.life.form*, sometimes drawing upon stand age.
-Records with "woody" *dominant.life.form* were classified as Forest Land.
-Those with *dominant.life.form* of "woody+grass", which in ForC is indicative of anything from a shrub-encroached grassland to a tree-dominated savanna, were given dual classification of Forest Land and Grassland.
-This dual classification indicates that records may be relevant to either category depending on the definition of forest applied (varies by country).
-For (rare) cases where *dominant.life.form* was grass and stand age was greater than zero, indicative of early successional vegetation, we assigned a classification of Forest Land, consistent with the IPCC definition that Forest Land includes land expected to succeed to forest.
-Cases where *dominant.life.form* was grass or crop and stand age was zero were indicative of a control for studies of forest regrowth following agricultural abandonment, and were classified as Grassland and Cropland, respectively.
-
-Classification into sub-categories was dependent upon stand age and site history (section `r  section_land_use_categories`).
-For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
-For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Grassland converted to Forest Land", respectively.
-For forests <20 years old with unspecified previous agricultural use, we assigned the sub-category "Land Converted to Forest land".
-Forests <20 years old with unknown land use prior to the study date were simply classified as "Forest Land".
-The same logic was applied for savannas, but including both Forest Land and Grassland as potentially relevant categories.
-
-Given the lack of public information needed to determine whether lands are classified as managed [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not include any classification of land management status from ForC in the records submitted to EFDB.
-However, we did provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
-
-# Results
-
-## ForC v4.0 contents
-
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig.
-`r fig_variable_mapping`).
-These records were distributed across all forested continents and ecozones, albeit unevenly (Fig. `r fig_map`).
-The largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig. `r fig_histograms`c).
-Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
-`r fig_histograms`b).
-Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
-The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
-In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).
-
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png")
-```
-
-\newpage
-```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
-```
-
-ForC contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
-The records were very unevenly distributed across variables. 
-The variable with most records was aboveground biomass, representing `r round(n_ind_records_agb/n_ind_records_EFDB_variables,2)*100`% of all independent records relevant to EFDB, and aboveground biomass components (woody biomass or foliage) representing an additional `r round(n_ind_records_agb_components/n_ind_records_EFDB_variables,2)*100`%.
-A total of `r round(n_ind_records_rootbiomass/n_ind_records_EFDB_variables,2)*100`% of relevant records were for root biomass (including fine and coarse root components), while `r round(n_ind_records_biomass/n_ind_records_EFDB_variables,2)*100`% described total biomass.
-The non-living pools were less represented, with `r round(n_ind_records_deadwood/n_ind_records_EFDB_variables,2)*100`% of relevant were for dead wood (including standing and fallen components), `r round(n_ind_records_OL/n_ind_records_EFDB_variables*100,1)`% for litter, `r round(n_ind_records_TotEcosystemC/n_ind_records_EFDB_variables*100,1)`% for total ecosystem C excluding soils, and `r round(n_ind_records_SOC/n_ind_records_EFDB_variables*100,1)`% for soil carbon.
-
-Increment and flux variables were poorly represented (Table `r table_variables`).
-The increment variable with most records was the aboveground biomass increment, representing `r round(n_ind_records_deltaagb/n_ind_records_EFDB_variables*100,1)`% of all independent records relevant to EFDB.
-The only other relevant increment variable with any records was the O horizon (litter) increment, with just `r n_ind_records_deltaOL` records.
-Relevant flux variable records (n=`r n_ind_records_flux`) together constituted `r round(n_ind_records_flux/n_ind_records_EFDB_variables,2)*100`% of ForC's independent records relevant to EFDB.
-
-\newpage
-```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
-C_variables_table <- read.csv("figures_tables/C_variables.csv")
-names(C_variables_table) <- stringr::str_replace_all(names(C_variables_table), "\\.", " ")
-C_variables_table[is.na(C_variables_table)]<- "" #replace NAs with blanks
-
-library(knitr)
-library(kableExtra)
-knitr::kable(C_variables_table[,-1], format = "latex",  
-             longtable = T,
-             caption = "\\textbf{Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.}") %>%
-  row_spec(0, bold = TRUE)  %>%
-  row_spec(which(C_variables_table$bold), bold = TRUE) %>% #total
-  kable_styling(font_size = 8) %>%
-  #kableExtra::kable_styling(full_width = TRUE) 
-  kable_styling(latex_options = c("hold_position", "repeat_header"), protect_latex = T) 
-```
-
-## Data submissions to EFDB
-
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were submitted to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`).
-The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
-The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review and post the records, and also because a minority of records were deemed not applicable to EFDB by the review panel.
-
-The ForC records submitted to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
-However, the density of these records was very unevenly distributed across continents, biomes, and forest types and was not proportional to the numbers of relevant records in ForC (Fig.
-`r fig_histograms`).
-Rather, the largest number of records came from North America, followed by Asia, South America, and Africa (Fig.
-`r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.
-`r fig_histograms`b).
-In terms of dominant vegetation, by far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
-The largest records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
-
-In terms of variables, records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
-The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records sent were for C stocks, including
-`r round(n_records_sent_biomass/n_records_sent_to_EFDB,2)*100`% for total biomass, 
-`r round(n_records_sent_agb/n_records_sent_to_EFDB,2)*100`% for aboveground biomass,
-`r round(n_records_sent_agb_components/n_records_sent_to_EFDB,2)*100`% for components of aboveground biomass (wood or foliage), 
-`r round(n_records_sent_rootbiomass/n_records_sent_to_EFDB,2)*100`% for root biomass, 
-`r round(n_records_sent_rootbiomass_components/n_records_sent_to_EFDB,2)*100`% for components of root biomass (coarse or fine roots), 
-`r round(n_records_sent_deadwood/n_records_sent_to_EFDB,2)*100`% for dead wood, 
-`r round(n_records_sent_deadwood_components/n_records_sent_to_EFDB,2)*100`% for components of dead wood (standing or fallen), 
-`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (entire O horizon or OL layer component),
-and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
-Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, virtually all for aboveground biomass (excepting `r n_records_sent_delta.litter` records for delta.O.horizon).
-The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records sent described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
-
-# Recommendations
-
-Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work (section `r section_database_needs`), new data collection and analysis (section `r section_new_data_needs`), and reporting (section `r section_data_reporting_needs`).
-
-## Database needs
-
-There is vast potential to expand forest C data in EFDB by completing the process of reviewing and submitting data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
-So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been submitted to EFDB.
-Although this process requires manual review of records, the submission of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
-Future efforts to review studies for submission should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
-
-In addition to the large potential to expand EFDB using records already in ForC, there are innumerous published EFDB-relevant forest C data that are not currently included in ForC, with more being published on a nearly daily basis.
-Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
-Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
-Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
-
-## Data collection and analysis needs
-
-New data collection and analysis is needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
-Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
-Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
-For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
-Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labriere_forest_2023].
-
-Several EFDB-relevant variables have not been calculated and presented as frequently as would be possible given existing forest census data and minimal extra research effort.
-For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022].
-Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
-However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017; @mokany_critical_2006].
-In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate estimates. 
-We recommend that, when possible, researchers calculate and report these variables, following the reporting guidelines specified in section `r section_data_reporting_needs`.
-
-Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labriere_forest_2023;@calders_laser_2022].
-Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
-Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
-Significant investment in ground-based forest research will be critical to filling these gaps.
-
-## Data reporting needs
-
-We recommend that, in order to make research valuable to estimate C stock changes according to methods provided in the IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
-It is particularly noteworthy that simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
-Some examples are as follows:
-(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion while supporting FAIR (http://dx.doi.org/10.1038/d41586-019-01720-7) goals;
-(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;
-(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant estimating forest C stock changes according to IPCC guidelines (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
-It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
-
-```{r table_recommendations, eval = TRUE, echo=FALSE, warning=FALSE}
-recommendations_table <- read.csv("figures_tables/recommendations.csv")
-names(recommendations_table) <- stringr::str_replace_all(names(recommendations_table), "\\.", " ")
-
-library(knitr)
-library(kableExtra)
-knitr::kable(recommendations_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}") %>%
-  row_spec(0, bold = TRUE)%>%
-  kable_styling(font_size = 10) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-For those compiling published records (e.g., for meta-analyses), the data set can have added value if all information required by EFDB is extracted from original publications. 
-This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
-The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.
-
-Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipeline developed here.
-For individual publications, the former option will generally be more efficient.
-However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mitigation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].
-
-# Conclusions
-<!--The conclusion goes here.
-You can modify the section name with  `\conclusions[modified heading if necessary]`.-->
-
-The ForC database contains large numbers of records that could potentially be useful for estimating C stock changes applying methodological guidance provided by the IPCC.
-Here we have developed a framework for submitting these records to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
-As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
-Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
-The records submitted to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.
-
-Going forward, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
-In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under teh Paris Agreement.
-This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may no longer be relevant.
-This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.
-We view the system developed here for submitting ForC data to the IPCC EFDB as one important step towards that goal.
-
-\clearpage
-
-# Appendix A. Updates to ForC {-}
-
-\captionsetup[table]{labelformat=empty}
-Table A1: **Table of changes to ForC fields.** These are changes implemented between releases of v3.0 and v4.0. 
-```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
-library (readr)
-urlfile="https://raw.githubusercontent.com/forc-db/ForC/master/database_management_records/record_of_changes.csv"
-record_of_changes <-read_csv(url(urlfile))
-names(record_of_changes) <- stringr::str_replace_all(names(record_of_changes), "\\.", " ")
-record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
-#record_of_changes<- record_of_changes[which(!grep("housekeeping",record_of_changes$Motivation)) , c(1:5)]  #trying to exclude rows for which Motivation = "housekeeping", but this is not working
-record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
-
-library(knitr)
-library(kableExtra)
-knitr::kable(record_of_changes, format = "latex",
-             longtable = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
-  kable_styling(font_size = 8) %>%
-  row_spec(0, bold = TRUE)
-```
-
-\clearpage
-
-# Appendix B. Mapping ForC to EFDB {-}
-
-Table B1: **Mapping of ForC fields to EFDB.** Details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
-```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-
-ForC_variables_included <- read.csv("figures_tables/ForC_fields_included.csv")
-names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
-column_names <- names(ForC_variables_included) # Define column names
-column_names[5] <- "Required*" # Add asterisk for footnote
-names(ForC_variables_included) <- column_names# Set column names of the matrix
-ForC_variables_included <- ForC_variables_included[c(1:5)]
-
-knitr::kable(ForC_variables_included, format = "latex",
-             longtable = TRUE) %>%
-  row_spec(0, bold = TRUE) %>%
-  kable_styling(font_size = 8) %>%
-kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
-
-```
-\* *Required* field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
+## References {-}

### 2023-10-26 doc/manuscript/SI_tables.Rmd
separate out SI material

--- 2023-06-09 doc/manuscript/ForC-EFDB manuscript for ESSD.Rmd
@@
 ---
-title: Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
-journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'essd')`"
-author:
-  - given_name: Kristina J.
-    surname: Anderson-Teixeira
-    affiliation: "1,2 *"
-    email: teixeirak@si.edu
-    corresponding: true
-  - given_name: Valentine
-    surname: Herrmann
-    affiliation: "1"
-  - given_name: Madison 
-    surname: Williams
-    affiliation: "1"
-  - given_name: Teagan
-    surname: Rogers
-    affiliation: "1"
-  - given_name: Rebecca
-    surname: Banbury Morgan
-    affiliation: "1,3" 
-  - given_name: Ben 
-    surname: Bond-Lamberty
-    affiliation: "4" 
-  - given_name: Susan 
-    surname: Cook-Patton
-    affiliation: "5"
+title: ''
+output: 
+    pdf_document
+bibliography: references.bib
+csl: new-phytologist.csl
 
-affiliation:
-  - code: 1
-    address: Center for Conservation Ecology, Smithsonian's National Zoo & Conservation Biology Institute, Front Royal, VA, United States
-  - code: 2
-    address: Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Panama, Republic of Panama
-  - code: 3
-    address: School of Geography, University of Leeds, Leeds, UK
-  - code: 4
-    address: Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
-  - code: 5
-    address: The Nature Conservancy; Arlington VA 22203, USA
-
-abstract: |
-
-bibliography: references.bib
-running:
-  title: Informing forest CO~2~ inventories with ForC
-  author: Anderson-Teixeira et al.
-competinginterests: |
-  The authors declare no competing interests.
-# OPTIONAL:
-algorithms: true
-# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
-# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
-copyrightstatement: |
-### The author's copyright for this publication is transferred to institution/company. 
-### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
-availability:
-  #  use this to add a statement when having only software code available
-  #  use this to add a statement when having only data sets available
-  #code: |
-  #data: |
-  codedata: |
-    All code and data are openly available.
-The ForC database and associated code is available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD).
-The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).
-authorcontribution: |
-  KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
-disclaimer: |
-acknowledgements: |
-  We gratefully acknowledge the substantial contributions of Valentyna Slivinska and Sandro Federici for collaboration on the conception, design, and technical review of this project.
-  Thank you to all researchers who collected and published the data contained in ForC, and to all research assistants and collaborators who have contributed to building the database.
-  Thank you to Avni Malhotra for helpful comments on an earlier draft of this manuscript.
-  Funding for this study was provided by the Smithsonian (Forest Global Earth Observatory, Smithsonian Working Land and Seascapes); a Bezos Earth Fund grant to the Nature Conservancy, with a sub-grant to NZCBI; and the Institute for Global Environmental Strategies.
-appendix: 
-output:
-  rticles::copernicus_article: default
-  bookdown::pdf_book:
-    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
-
-    header-includes:
-    - \usepackage{caption}
-    #- \usepackage{booktabs}
-    #- \usepackage{makecell}
 ---
 
-<!--**Important**: Always double-check with the official manuscript preparation guidelines at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html), especially the sections "Technical instructions for LaTeX" and "Manuscript composition".
-Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.-->
-
 ```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
-#Tables
-table_pools =1
-table_variables =2
-table_recommendations = 3
-
-#Figures
-fig_variable_mapping=1
-fig_map=2
-fig_histograms=3
-
-#Cited sections (listed here because these can change)
-section_C_pools = 2.1 
-section_land_use_categories = 2.2
-section_new_variables = 3.2
-section_C_variable_mapping = "4.1.1"
-section_land_use_mapping = "4.1.2"
-section_database_needs = "6.1"
-section_new_data_needs = "6.2"
-section_data_reporting_needs = "6.3"
 
 #Appendices
-table_ForCchanges = "Appendix A"
-table_ForCfieldmapping = "Appendix B"
-
-#READING IN FILES TO GET FIGURES/ TABLES/ STATS 
-path_to_dir <- dirname(dirname(getwd())) # get path to repo 
-path_to_ForC= "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db" #Krista's laptop
-# path_to_ForC= "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db" #Madison, to knit, you'll need to enter the path to ForC-db on your computer, then un-comment this line.
-
-# MOVE FIGURE FILES 
-#world map from GitHub
-do.call(file.remove, list(list.files(paste0(path_to_dir, "/doc/manuscript/figures_tables"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE))) # first remove old file
-flist <- list.files(paste0(path_to_ForC, "/ForC/figures/"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE)
-file.copy(flist, paste0(path_to_dir, "/doc/manuscript/figures_tables"))
-
-#Statistics
-ForC_measurements <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_measurements.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_records_ForC= length(ForC_measurements[,"measurement.ID"])  # total n records in ForC
-
-ForC_plots <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_plots.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_plots_ForC = length(ForC_plots[,"plot.ID"]) # total n plots in ForC
-
-ForC_sites <- read.csv(paste0(path_to_ForC,"/ForC/data/ForC_sites.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_areas_ForC = max(ForC_sites$geographic.area, na.rm = TRUE) # total n geographic areas in ForC
-
-ForC_simplified <- read.csv(paste0(path_to_ForC,"/ForC/ForC_simplified/ForC_simplified.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_independent_records_ForC= length(ForC_simplified[,"measurement.ID"]) #suspected duplicates removed. This is the number of records fin ForC simplified.
-
-C_variables <- read.csv(paste0(path_to_dir,"/doc/manuscript/figures_tables/C_variables.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_relevant_variables = nrow(C_variables) - 8  # n variables relevant = rows in C_variables - headers and total 
-n_relevant_variables_with_no_data =  sum(C_variables$`n in ForC` == 0,  na.rm = TRUE)
-n_relevant_variables_with_data= n_relevant_variables - n_relevant_variables_with_no_data
-
-n_ind_records_EFDB_variables= C_variables[length(C_variables[,"variable"]), "n independent records in ForC"] #n independent records relevant to EFDB
-n_ind_records_agb <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag"]
-n_ind_records_agb_components <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_ag_foliage"]
-n_ind_records_biomass <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass"]
-n_ind_records_rootbiomass <- C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n independent records in ForC`[C_variables$`variable` == "biomass_root_coarse"]
-n_ind_records_deadwood <- C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_down"]+C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_standing"]
-n_ind_records_OL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "litter"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "O.horizon"]
-n_ind_records_TotEcosystemC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "total.ecosystem_2"] 
-n_ind_records_SOC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "SOM / SOC"] 
-n_ind_records_deltaagb <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.agb"] 
-n_ind_records_deltaOL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "delta.O.horizon"] 
-flux_variables_with_records <- c("NPP_woody", "ANPP_woody", "ANPP_woody_stem", "ANPP_woody_branch", "woody.mortality_ag", "BNPP_root", "BNPP_root_fine","BNPP_root.turnover_fine","BNPP_root_coarse" , "ANPP_litterfall", "NPP_litter", "R_het_litter")
-n_ind_records_flux <- sum(C_variables$`n independent records in ForC`[C_variables$`variable` %in% flux_variables_with_records])
-
-n_records_reviewed_or_added= C_variables[length(C_variables[,"variable"]), "n reviewed"]
-n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
-n_records_checked = n_records_reviewed_or_added - n_records_added # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
-
-n_records_sent_to_EFDB = C_variables[length(C_variables[,"variable"]), "n submitted to EFDB"] # n records sent to EFDB
-n_relevant_variables_no_records_sent =  sum(C_variables$`n submitted to EFDB` == 0,  na.rm = TRUE)
-n_relevant_variables_records_sent= n_relevant_variables - n_relevant_variables_no_records_sent
-n_records_sent_biomass<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass"]
-n_records_sent_agb <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag"]
-n_records_sent_agb_components <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag_woody"] + C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_ag_foliage"]
-n_records_sent_rootbiomass <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root"] 
-n_records_sent_rootbiomass_components <- C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root_fine"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "biomass_root_coarse"]
-n_records_sent_deadwood<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood"] 
-n_records_sent_deadwood_components<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_down"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_standing"]
-n_records_sent_litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "litter"] +C_variables$`n submitted to EFDB`[C_variables$`variable` == "O.horizon"]
-n_records_sent_soil<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "SOM / SOC"] 
-n_records_sent_stocks = n_records_sent_biomass+n_records_sent_agb+n_records_sent_agb_components+n_records_sent_rootbiomass+n_records_sent_rootbiomass_components+n_records_sent_deadwood+n_records_sent_deadwood_components+ n_records_sent_litter+ n_records_sent_soil
-n_records_sent_delta.agb<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.agb"]
-n_records_sent_delta.litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.O.horizon"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.litter"]
-n_records_sent_increments <- n_records_sent_delta.agb+n_records_sent_delta.litter
-n_records_sent_fluxes <- sum(C_variables$`n submitted to EFDB`[C_variables$`variable` %in% flux_variables_with_records])
-n_records_sent_fluxes_check = n_records_sent_to_EFDB-n_records_sent_stocks-n_records_sent_increments
-
-n_records_posted_EFDB = 376 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
-Then get number from upper left box where it says "displayed records")
-
-n_new_variables = 11 + 2 + 2 #delta. + woody.mortality + O horizon
-# get n fields modified
-sampling_details <- read.csv(paste0(path_to_ForC,"/ForC/database_management_records/record_of_changes.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_fields_modified =  length(sampling_details[,"Column"]) + 4
+appendix_ForC_updates = "Methods S1"
+appendix_IPCC_definitions = "Notes S1"
+table_ForCchanges = "Table S1"
+table_ForCfieldmapping = "Table S2"
 
@@
 
-**Abstract.**
-Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
-However, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
-Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
-Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas inventories if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
-We begin by reconciling terminology and mapping ForC fields into EFDB.
-This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have undergone necessary review and been submitted to date.
-Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into or out of the IPCC-defined C pools.
-Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
-In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
-Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
-
-\introduction[Introduction]
-
-Forests are critical to management of the atmospheric concentration of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
-In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
-Table 2--> [forests only, @harris_global_2021].
-This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
-Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
-The future of this important CO~2~ sink is highly uncertain, and depends upon both forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
-
-Forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
-Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
-As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
-Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
-One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
-
-The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
-Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
-The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
-A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
-Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
-For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
-In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
-This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
-Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.
-
-To improve data accessibility for preparing greenhouse gas estimates, the IPCC created the Emission Factor Database (EFDB; [https://www.ipcc-nggip.iges.or.jp/EFDB/main.php](https://www.ipcc-nggip.iges.or.jp/EFDB/main.php)), which is intended as a recognized library of emission factors and other parameters that can be used for estimating greenhouse gas emissions and removals.
-The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
-The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include C stocks, net annual increments, and annual fluxes for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
-
-The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021].
-ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
-As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
-The purpose of this publication is to document that process and provide recommendations for future improvements.
-
-Here, we
-(1) review IPCC methods and definitions applied to estimate CO~2~ emissions and removals from forest in the context of typical forest C estimation methodologies;
-(2) describe mapping of ForC to IPCC's EFDB;
-(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data submission to EFDB;
-(4) summarize the data in ForC relevant to EFDB and records that have been submitted to date; and
-(5) provide recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
-
-# IPCC methods and definitions
-
-The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (described in section `r section_C_pools`), plus any harvested wood products.
-Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
-
-## Carbon pools
-
-Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- scientific studies.
-IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
-While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
-
-```{r table_pools, eval = TRUE, echo=FALSE, warning=FALSE}
-C_pools_table <- read.csv("figures_tables/C_pools.csv")
-names(C_pools_table) <- stringr::str_replace_all(names(C_pools_table), "\\.", " ")
-
-library(knitr)
-library(kableExtra)
-knitr::kable(C_pools_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
-  row_spec(0, bold = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-### Biomass
-Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].
-
-Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
-IPCC's guidance is that the understory may be excluded if it constitutes a "minor" component [defined as < 25 - 30 % of emissions/removals for the overall category, @ipcc_2006_2006], and where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
-A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022], and therefore stems < 10 cm DBH can usually be considered a minor component of aboveground biomass for these forests.
-In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @hughes_biomass_1999], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes.
-While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although these do tend to be censused when they constitute a large proportion of the biomass [e.g., @fukushima_recovery_2007].
-Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].
-
-A universal challenge in estimating biomass (living or dead) from forest census data is applying appropriate allometric models to convert DBH measurements to biomass, and such selection has an enormous influence on estimates of biomass stocks, increments, and fluxes [@clark_landscapescale_2000;@clark_net_2001;@calders_laser_2022].
-While trusted and standardized allometric models are becoming increasingly available [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022], large uncertainties remain.
-IPCC Tier 1 values currently draw on studies applying a variety of allometric models [e.g., @requenasuarez_estimating_2019; @rozendaal_aboveground_2022].
-
-Belowground biomass is defined as "all biomass of live roots" [@ipcc_good_2003;@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
-The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, and tending to be very uncertain [e.g., @keller_biomass_2001].
-Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.
-Field methods for estimating root biomass are highly variable [@freschet_starting_2021].
-IPCC's default method for Tier 1 estimates is to apply a ratio of belowground to aboveground biomass, with default factors defined based on ecological zone, continent, and forest age<!--IPCC table 4.4--> [@ipcc_2006_2006;@ipcc_2019_2019].
-
-### Dead Organic Matter
-
-Dead organic matter includes all non-living biomass larger than the litter size threshold. 
-Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see section `r section_land_use_categories`), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
-
-Dead wood, which is typically <50 t C ha^-1^ but can exceed 150 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as "all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil" [@ipcc_good_2003;@ipcc_2006_2006].
-This pool includes standing and fallen dead wood, stumps, and dead roots of diameter ≥10 cm (or a diameter specified by the country).
-Dead wood stocks and fluxes can be quite variable across forests [@anderson-teixeira_carbon_2021], and can at times be the dominant pool in a forest ecosystem [e.g., following a severe natural disturbance, @carmona_coarse_2002].
-However, aboveground dead wood remains relatively poorly characterized at a global scale [@anderson-teixeira_carbon_2021], and belowground dead wood is rarely studied [@merganicova_dadwood_2012].
-In turn, dead wood pools are poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-31 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-Litter, which is typically <40 t C ha^-1^ but can exceed 100 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by IPCC as including "all non-living  biomass  with  a  diameter  less  than  a  minimum  diameter  chosen  by  the  country  (for  example 10 cm), lying dead, in various states of decomposition above the mineral or organic soil" [@ipcc_good_2003;@ipcc_2006_2006].
-As noted above, live fine roots may be included in litter when difficult to separate empirically.
-The definition includes the entire O horizon, including litter (OL), fumic (OF), and humic (OH) layers, in addition to litter embedded within the soil.
-This definition contrasts with empirical studies that focus on aboveground litter, often including only the OL layer in the definition of litter, and do not always specify the components included.
-Similar to dead wood, litter is poorly characterized in large-scale forest C budgets [@pan_large_2011;@harris_global_2021], and IPCC's latest Tier 1 default values are based on just 1-7 references per climate zone [Table 2.2 in @ipcc_2019_2019].<!--table 2.2-->
-
-### Soil Organic Matter/ Carbon
-
-Soil organic matter/ carbon (SOM/ SOC), which is typically >100 t C and can exceed 300 t C in the top two meters of soil [@sanderman_soil_2017], is defined by IPCC as "organic carbon in mineral and organic soils (including peat) to a specified depth chosen by the country and applied consistently through the time series" [@ipcc_good_2003;@ipcc_2006_2006].
-Live fine roots may be included with soil organic matter when it is not feasible to distinguish them empirically.
-The greatest source of methodological variation in measuring SOM/ SOC is sampling depth, which has a suggested default of 30 cm but may vary by country provided that consistent criteria are applied.
-
-## Land classification
-IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
-Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
-Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
-Definitions of forest are allowed to vary by country, but must be applied consistently.
-Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
-
-The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
-This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
-However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
-
-# Updates to ForC (ForC v4.0)
-
-This section describes changes relative to ForC v3.0 [@anderson-teixeira_carbon_2021].
-Previous versions of ForC [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021] contained most of the information required by EFDB, and, more broadly, to inform C stock change calculations under the Paris Agreement.
-However, modest changes to the structure and contents of ForC were needed in order to provide all information required by EFDB and to improve ForC's capacity to serve as a repository of valuable information for forest C inventories under IPCC guidelines.
-To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
-
-## New or modified fields
-
-We added or modified a total of `r n_fields_modified` fields (`r table_ForCchanges`).
-Most notably, these included improvement of the representation of uncertainty, recording of original units and organic matter to C conversion factors, and expanding the information recorded in the citations table.
-For the latter, we used an R script to automatically harvest (scrape) the URL, citation, abstract and language of the publications, based on their DOI, using R package `rvest`[ @wickham_rvest_2022]. 
-That information was manually retrieved when the web scraping failed.
-
-## New variables
-
-To create structure for EFDB-relevant records, we added a total of `r n_new_variables` new variables to the set of named and defined variables (Fig.
-`r fig_variable_mapping`), counting each pair of variables with units in C (ending in  `_C`) or organic matter (ending in `_OM`) as one.
-The majority of these were increment variables (n=11), adding to only one previously defined increment variable (aboveground biomass increment, *delta.agb*). 
-These are directly related to C stocks as previously defined in ForC, with "*delta.*" added in front of the variable name.
-Further, we added variables capturing the belowground component of woody mortality (*woody.mortality_root*) and the combined aboveground and belowground components of woody mortality (*woody.mortality*).
-Although most of these variables lacked records in ForC as of `r format(Sys.time(), '%B %d, %Y')`, their addition gave the structure such that records can be populated over time. 
-Finally, to provide better definition of the previously existing variable *organic.layer*, which has a nebulous definition that reflects the varied definitions adopted by original studies, we added two clearly defined variables: *litter* (relatively undecomposed plant material/ OL horizon), and *O.horizon* (entire O-horizon, including OL).
-
-```{r fig_variable_mapping, out.width = "14cm", echo = FALSE, fig.cap = "\\textbf{Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.} For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
-Most, but not all, EFDB-relevant ForC variables are shown here.
-Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
-**This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/C_variable_mapping.png")
-```
-
-## Quality control measures
-Prior to releasing ForC v4.0, we executed several quality control measures.
-First, we implemented a system of continuous integration using GitHub Actions [*sensu* @kim_implementing_2022] to run some automatic checks any time the master data files are updated, including outlier tests and checks for completeness and naming consistency of records across data files.
-Second, to improve information on geographic coordinates, we created a field to record coordinate precision (`r table_ForCchanges`), and flagged and reviewed records with suspected low precision.
-Third, to identify erroneous climate data, we compared ForC climate values to those extracted from WorldClim version 2.1 [@vandepol_identifying_2016; @bailey_climwin_2016] based on site coordinates. 
-Records deviating from WorldClim values by more variable-specific thresholds (>5°C for mean annual temperature, >7.5°C for mean temperatures of the warmest and coldest months, or >1 for log(mean annual precipitation in mm)) were flagged as requiring review prior to use in analysis or submission to EFDB.
-
-Because ForC v4.0 contained known duplicate records, we used R scripts to identify likely duplicates, as detailed in @anderson-teixeira_carbon_2021.
-Henceforth, we refer to the set of records with likely duplicates removed as "independent records".
-All records sent to EFDB were ensured to be independent and original through manual review, as detailed below.
-
-## Manual review of records to be sent to EFDB
-
-EFDB data submissions required information that was not recorded in previous versions of ForC, but for which new fields were created for EFDB compatibility (`r table_ForCchanges`).
-It was therefore necessary to return to original publications to retrieve relevant information, including 
-(1) estimates in original units, 
-(2) confidence intervals (when not already in ForC),
-(3) whether records of interest were presented in tables or text or digitized from figures (EFDB will not accept digitized data), 
-(4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass)
-We also checked that existing ForC records were complete and correct.
-
-Manual review of records was the limiting step for data submission to EFDB. 
-We prioritized review of
-(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021],
-(2) studies with confidence intervals recorded in ForC (because uncertainty estimates are important to the IPCC),
-(3) original publications containing large numbers of EFDB-relevant records,
-and (4) records from tropical regions.
-The latter criteria was motivated by the fact that although tropical forest is the single most important biome for climate change mitigation [@griscom_natural_2017;@griscom_national_2020], ground-based data on tropical forest C cycling tends to be more scarce due to a variety of challenges [@anderson-teixeira_carbon_2021; @delima_making_2022], and tropical countries are more likely to apply Tier 1 methodology that bases forest C budgets on internationally defined IPCC default values [@romijn_assessing_2015].
-
-## Addition of new records
-
-In addition to reviewing existing records, we added a total of `r n_records_added` new records to ForC.
-These included 104 records<!--100 from Piponiot, 4 from Lutz--> from two studies [@piponiot_distribution_2022; @lutz_largediameter_2021] that were not previously included in ForC.
-In addition, we created new records for `r n_records_added-104` EFDB-relevant estimates presented in the original publication that were not yet present in ForC.
-
-# Submission of ForC data to EFDB
-
-To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been submitted to EFDB.
-Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB. 
-Complete records needed to be reviewed by the EFDB editorial board and then posted in the database -- a process that lags behind submission of records and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
-
-## Mapping ForC to EFDB
-
-The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`.
-For the majority of fields, contents of the field in ForC was copied directly into an EFDB field, either as the only contents of that field or as part of a composite record.
-For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).
-In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*).
-For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
-For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
-There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) land classification (section `r section_land_use_mapping`).
-
-### Carbon cycle variables
-
-With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig.
-`r fig_variable_mapping`).
-For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, net annual increments, influxes (a.k.a. "gross annual increments" by IPCC), and outfluxes.
-As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
-It is important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
-For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
-Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
-Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
-
-### Land classification
-
-Determination of the IPCC land-use category (i.e., Forest Land, Grassland, Wetlands, Cropland, Settlements, or Other Land; section `r  section_land_use_categories`) was made based on the categorical ForC field *dominant.life.form*, sometimes drawing upon stand age.
-Records with "woody" *dominant.life.form* were classified as Forest Land.
-Those with *dominant.life.form* of "woody+grass", which in ForC is indicative of anything from a shrub-encroached grassland to a tree-dominated savanna, were given dual classification of Forest Land and Grassland.
-This dual classification indicates that records may be relevant to either category depending on the definition of forest applied (varies by country).
-For (rare) cases where *dominant.life.form* was grass and stand age was greater than zero, indicative of early successional vegetation, we assigned a classification of Forest Land, consistent with the IPCC definition that Forest Land includes land expected to succeed to forest.
-Cases where *dominant.life.form* was grass or crop and stand age was zero were indicative of a control for studies of forest regrowth following agricultural abandonment, and were classified as Grassland and Cropland, respectively.
-
-Classification into sub-categories was dependent upon stand age and site history (section `r  section_land_use_categories`).
-For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
-For forests <20 years old with history including cultivation/ tillage or grazing, past land-use categories were Cropland and Grassland, respectively, making land-use subcategories were "Cropland converted to Forest Land" and "Grassland converted to Forest Land", respectively.
-For forests <20 years old with unspecified previous agricultural use, we assigned the sub-category "Land Converted to Forest land".
-Forests <20 years old with unknown land use prior to the study date were simply classified as "Forest Land".
-The same logic was applied for savannas, but including both Forest Land and Grassland as potentially relevant categories.
-
-Given the lack of public information needed to determine whether lands are classified as managed [@ogle_delineating_2018; @deng_comparing_2021], and because the IPCC's definition of managed land is more expansive than is commonly applied in the scientific literature and hence in ForC, we did not include any classification of land management status from ForC in the records submitted to EFDB.
-However, we did provide auxiliary information that should be useful in making this determination, including geographical location and notable disturbance events.
-
-# Results
-
-## ForC v4.0 contents
-
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig.
-`r fig_variable_mapping`).
-These records were distributed across all forested continents and ecozones, albeit unevenly (Fig. `r fig_map`).
-The largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig. `r fig_histograms`c).
-Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
-`r fig_histograms`b).
-Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
-The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
-In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).
-
-```{r fig_map, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.} Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png")
-```
-
-\newpage
-```{r fig_histograms, out.width = "15cm", echo = FALSE, fig.cap = "\\textbf{Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.} For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-For FAO ecozones (b), codes are as listed in the caption of Figure 2."}
-#this is for a two-column figure. for one-column figure, use out.width = "8.3cm"
-knitr::include_graphics("figures_tables/Histogram_n_Relevant_and_Transferred_Records.png")
-```
-
-ForC contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
-The records were very unevenly distributed across variables. 
-The variable with most records was aboveground biomass, representing `r round(n_ind_records_agb/n_ind_records_EFDB_variables,2)*100`% of all independent records relevant to EFDB, and aboveground biomass components (woody biomass or foliage) representing an additional `r round(n_ind_records_agb_components/n_ind_records_EFDB_variables,2)*100`%.
-A total of `r round(n_ind_records_rootbiomass/n_ind_records_EFDB_variables,2)*100`% of relevant records were for root biomass (including fine and coarse root components), while `r round(n_ind_records_biomass/n_ind_records_EFDB_variables,2)*100`% described total biomass.
-The non-living pools were less represented, with `r round(n_ind_records_deadwood/n_ind_records_EFDB_variables,2)*100`% of relevant were for dead wood (including standing and fallen components), `r round(n_ind_records_OL/n_ind_records_EFDB_variables*100,1)`% for litter, `r round(n_ind_records_TotEcosystemC/n_ind_records_EFDB_variables*100,1)`% for total ecosystem C excluding soils, and `r round(n_ind_records_SOC/n_ind_records_EFDB_variables*100,1)`% for soil carbon.
-
-Increment and flux variables were poorly represented (Table `r table_variables`).
-The increment variable with most records was the aboveground biomass increment, representing `r round(n_ind_records_deltaagb/n_ind_records_EFDB_variables*100,1)`% of all independent records relevant to EFDB.
-The only other relevant increment variable with any records was the O horizon (litter) increment, with just `r n_ind_records_deltaOL` records.
-Relevant flux variable records (n=`r n_ind_records_flux`) together constituted `r round(n_ind_records_flux/n_ind_records_EFDB_variables,2)*100`% of ForC's independent records relevant to EFDB.
-
-\newpage
-```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
-C_variables_table <- read.csv("figures_tables/C_variables.csv")
-names(C_variables_table) <- stringr::str_replace_all(names(C_variables_table), "\\.", " ")
-C_variables_table[is.na(C_variables_table)]<- "" #replace NAs with blanks
-
-library(knitr)
-library(kableExtra)
-knitr::kable(C_variables_table[,-1], format = "latex",  
-             longtable = T,
-             caption = "\\textbf{Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.}") %>%
-  row_spec(0, bold = TRUE)  %>%
-  row_spec(which(C_variables_table$bold), bold = TRUE) %>% #total
-  kable_styling(font_size = 8) %>%
-  #kableExtra::kable_styling(full_width = TRUE) 
-  kable_styling(latex_options = c("hold_position", "repeat_header"), protect_latex = T) 
-```
-
-## Data submissions to EFDB
-
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were submitted to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`).
-The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
-The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review and post the records, and also because a minority of records were deemed not applicable to EFDB by the review panel.
-
-The ForC records submitted to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
-However, the density of these records was very unevenly distributed across continents, biomes, and forest types and was not proportional to the numbers of relevant records in ForC (Fig.
-`r fig_histograms`).
-Rather, the largest number of records came from North America, followed by Asia, South America, and Africa (Fig.
-`r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.
-`r fig_histograms`b).
-In terms of dominant vegetation, by far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
-The largest records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
-
-In terms of variables, records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
-The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records sent were for C stocks, including
-`r round(n_records_sent_biomass/n_records_sent_to_EFDB,2)*100`% for total biomass, 
-`r round(n_records_sent_agb/n_records_sent_to_EFDB,2)*100`% for aboveground biomass,
-`r round(n_records_sent_agb_components/n_records_sent_to_EFDB,2)*100`% for components of aboveground biomass (wood or foliage), 
-`r round(n_records_sent_rootbiomass/n_records_sent_to_EFDB,2)*100`% for root biomass, 
-`r round(n_records_sent_rootbiomass_components/n_records_sent_to_EFDB,2)*100`% for components of root biomass (coarse or fine roots), 
-`r round(n_records_sent_deadwood/n_records_sent_to_EFDB,2)*100`% for dead wood, 
-`r round(n_records_sent_deadwood_components/n_records_sent_to_EFDB,2)*100`% for components of dead wood (standing or fallen), 
-`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (entire O horizon or OL layer component),
-and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
-Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, virtually all for aboveground biomass (excepting `r n_records_sent_delta.litter` records for delta.O.horizon).
-The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records sent described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
-
-# Recommendations
-
-Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work (section `r section_database_needs`), new data collection and analysis (section `r section_new_data_needs`), and reporting (section `r section_data_reporting_needs`).
-
-## Database needs
-
-There is vast potential to expand forest C data in EFDB by completing the process of reviewing and submitting data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
-So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been submitted to EFDB.
-Although this process requires manual review of records, the submission of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
-Future efforts to review studies for submission should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
-
-In addition to the large potential to expand EFDB using records already in ForC, there are innumerous published EFDB-relevant forest C data that are not currently included in ForC, with more being published on a nearly daily basis.
-Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
-Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
-Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
-
-## Data collection and analysis needs
-
-New data collection and analysis is needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
-Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
-Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
-For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
-Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labriere_forest_2023].
-
-Several EFDB-relevant variables have not been calculated and presented as frequently as would be possible given existing forest census data and minimal extra research effort.
-For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022].
-Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
-However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017; @mokany_critical_2006].
-In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate estimates. 
-We recommend that, when possible, researchers calculate and report these variables, following the reporting guidelines specified in section `r section_data_reporting_needs`.
-
-Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labriere_forest_2023;@calders_laser_2022].
-Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
-Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
-Significant investment in ground-based forest research will be critical to filling these gaps.
-
-## Data reporting needs
-
-We recommend that, in order to make research valuable to estimate C stock changes according to methods provided in the IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
-It is particularly noteworthy that simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
-Some examples are as follows:
-(1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion while supporting FAIR (http://dx.doi.org/10.1038/d41586-019-01720-7) goals;
-(2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;
-(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant estimating forest C stock changes according to IPCC guidelines (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
-It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
-
-```{r table_recommendations, eval = TRUE, echo=FALSE, warning=FALSE}
-recommendations_table <- read.csv("figures_tables/recommendations.csv")
-names(recommendations_table) <- stringr::str_replace_all(names(recommendations_table), "\\.", " ")
-
-library(knitr)
-library(kableExtra)
-knitr::kable(recommendations_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}") %>%
-  row_spec(0, bold = TRUE)%>%
-  kable_styling(font_size = 10) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
-```
-
-For those compiling published records (e.g., for meta-analyses), the data set can have added value if all information required by EFDB is extracted from original publications. 
-This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
-The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.
-
-Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipeline developed here.
-For individual publications, the former option will generally be more efficient.
-However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mitigation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].
-
-# Conclusions
-<!--The conclusion goes here.
-You can modify the section name with  `\conclusions[modified heading if necessary]`.-->
-
-The ForC database contains large numbers of records that could potentially be useful for estimating C stock changes applying methodological guidance provided by the IPCC.
-Here we have developed a framework for submitting these records to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
-As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
-Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
-The records submitted to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.
-
-Going forward, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
-In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under teh Paris Agreement.
-This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may no longer be relevant.
-This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.
-We view the system developed here for submitting ForC data to the IPCC EFDB as one important step towards that goal.
-
-\clearpage
-
-# Appendix A. Updates to ForC {-}
-
-\captionsetup[table]{labelformat=empty}
-Table A1: **Table of changes to ForC fields.** These are changes implemented between releases of v3.0 and v4.0. 
+**`r table_ForCchanges`. Updates to ForC field implemented between releases of v3.0 and v4.0.** 
 ```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
@@
 
-# Appendix B. Mapping ForC to EFDB {-}
-
-Table B1: **Mapping of ForC fields to EFDB.** Details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
+**`r table_ForCfieldmapping`.
+Mapping of ForC fields to EFDB.** Details documented in the public GitHub repository associated with the project, IPCC-EFDB-integration repository within the ForC-db organization (file *ForC-EFDB_mapping.csv* available at [https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv](https://github.com/forc-db/IPCC-EFDB-integration/blob/main/doc/ForC-EFDB_mapping/ForC-EFDB_mapping.csv)).
 ```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
revise abstract

@@
 
-*(NEEDS REVISION)*
-
-Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
-However, the efficacy of greenhouse gas mitigation planning and reporting is dependent upon the quality of available emission factors data, including forest carbon (C) stocks and changes therein.
-Tens of thousands of relevant forest C estimates have been published, yet are not readily accessible to the practitioners compiling national greenhouse gas inventories. 
-Many of these data have, however, been compiled in the Global Forest C database (ForC; https://forc-db.github.io/) and stand to be of value to greenhouse gas inventories if made available through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
-Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
-We begin by reconciling terminology and mapping ForC fields into EFDB.
-This process required some updates to the ForC database structure, leading to the release of a new version of ForC (v4.0; described here).
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which have undergone necessary review and been submitted to date.
-Among the data in ForC, there is disproportionate representation of biomass (particularly aboveground) stocks, with far fewer records for dead organic matter and soil C, and relatively few or no records for net annual increments or C fluxes into or out of the IPCC-defined C pools.
-Geographic representation is also quite uneven, with the highest densities of relevant records in temperate forests, and with relatively scant representation of tropical forests in Africa and Asia. 
-In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
+<!-- (1) the research conducted or aims, including the rationale-->
+- Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
+The efficacy of greenhouse gas mitigation planning and reporting depends on the accessibility and quality of data on forest carbon (C) stocks and changes therein.
+Tens of thousands of relevant forest C estimates have been published and compiled in the Global Forest C database (ForC) and could be valuable for this purpose if made accessible through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
+<!-- (2) methods-->
+- Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
+<!-- (3) key results-->
+- As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which had undergone necessary review and been submitted.
+Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests). 
+<!-- (4) the main conclusion, including the key points of discussion-->
+- In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
 Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
convert tables (now they show up)

@@
 
+\newpage
+**Table `r table_pools`. IPCC-defined forest carbon pools with definitions and measurement methods.** <!--Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).-->
 ```{r table_pools, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 
-library(knitr)
-library(kableExtra)
-knitr::kable(C_pools_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).") %>%
-  row_spec(0, bold = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 8) # set font
+
+ft <- flextable(C_pools_table)
+ft <- autofit(ft)
+ft <- width(ft, 1, width = 1)
+ft <- width(ft, 2:4, width = 2)
+ft <- bold(ft, i = 1, j = NULL, bold = TRUE, part = "header") # put title in bold
+ft
 ```
@@
 \newpage
+**Table `r table_variables`. Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.**
 ```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 
-library(knitr)
-library(kableExtra)
-knitr::kable(C_variables_table[,-1], format = "latex",  
-             longtable = T,
-             caption = "\\textbf{Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.}") %>%
-  row_spec(0, bold = TRUE)  %>%
-  row_spec(which(C_variables_table$bold), bold = TRUE) %>% #total
-  kable_styling(font_size = 8) %>%
-  #kableExtra::kable_styling(full_width = TRUE) 
-  kable_styling(latex_options = c("hold_position", "repeat_header"), protect_latex = T) 
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 8) # set font
+
+ft <- flextable(C_variables_table[,-1])
+ft <- autofit(ft)
+ft <- width(ft, 1, width = 2)
+ft <- width(ft, 2:5, width = 1)
+ft <- bold(ft, i = 1, j = NULL, bold = TRUE, part = "header") # put title in bold
+ft <- bold(ft, i = which(C_variables_table$bold), j = NULL, bold = TRUE, part = "body") # put categories in bold
+ft
 ```
@@
 
+\newpage
+**Table `r table_recommendations`. Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.**
 ```{r table_recommendations, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 
-library(knitr)
-library(kableExtra)
-knitr::kable(recommendations_table, format = "latex",
-             longtable = FALSE,
-             caption = "\\textbf{Recommended best practices for reporting forest C estimates of value to national greenhouse gas inventories under IPCC guidance.}") %>%
-  row_spec(0, bold = TRUE)%>%
-  kable_styling(font_size = 10) %>%
-  kableExtra::kable_styling(full_width = TRUE) 
-  #kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 8) # set font
+
+ft <- flextable(recommendations_table)
+ft <- autofit(ft)
+ft <- width(ft, 1, width = 1)
+ft <- width(ft, 2:3, width = 3)
+ft <- bold(ft, i = 1, j = NULL, bold = TRUE, part = "header") # put title in bold
+ft
 ```

### 2023-10-26 doc/manuscript/ForC-EFDB manuscript.Rmd
updates clarifying organization

@@
 
+Major steps included
+(1) mapping ForC into EFDB, including aligning ForC and IPCC definitions (summarized in `r appendix_IPCC_definitions`),
+(2) revising ForC v3.0 to support semi-automated submissions to EFDB, yielding ForC v4.0 (detailed in `r appendix_ForC_updates`), and
+(3) submitting data to EFDB. 
+These steps are detailed below.
+
 ## Mapping ForC to EFDB
 
-The mapping of ForC fields into EFDB fields is summarized in `r table_ForCfieldmapping`.
+With input from the IPCC's Technical Support Unit and referencing IPCC guidance [@refs], we mapped ForC fields into EFDB fields, as summarized in `r table_ForCfieldmapping`.
 For the majority of fields, contents of the field in ForC was copied directly into an EFDB field, either as the only contents of that field or as part of a composite record.
@@
 For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
-There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (section `r section_C_variable_mapping`) and (2) land classification (section `r section_land_use_mapping`).
+There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (detailed below in **Carbon cycle variables**) and (2) land classification (detailed below in **Land classification**).
 
@@
 
-With input from the IPCC's Technical Support Unit, we reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig.
-`r fig_variable_mapping`).
+We reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig. `r fig_variable_mapping`).
 For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, net annual increments, influxes (a.k.a. "gross annual increments" by IPCC), and outfluxes.

### 2023-10-30 doc/manuscript/ForC-EFDB manuscript.Rmd
coauthor suggestions to abstract
These are from Susan. Sandro's are in comments.

@@
 <!-- (1) the research conducted or aims, including the rationale-->
-- Forests are critical for climate change mitigation and constitute a substantial portion of planned net emissions reductions under the 2015 Paris Agreement.
-The efficacy of greenhouse gas mitigation planning and reporting depends on the accessibility and quality of data on forest carbon (C) stocks and changes therein.
-Tens of thousands of relevant forest C estimates have been published and compiled in the Global Forest C database (ForC) and could be valuable for this purpose if made accessible through the Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC).
+- Forests are recognized as a critical source of additional climate change mitigation needed to meet the goals of the Paris Agreement. 
+However, the efficacy climate change mitigation planning and reporting depends on having high quality of data on forest carbon (C) stocks and changes.
+The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data, but needs comprehensive and well-documented data to be so.
 <!-- (2) methods-->
-- Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB, assess the data available and submitted to date, and provide recommendations for improving forest data collection, analysis, and reporting to improve inventories of forest-sector greenhouse gas emissions and removals.
+- The Global Forest C database (ForC) represents a compilation of tens of thousands of forest C estimates that is fully traceable and publicly available.
+Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB to help ensure that the EFDB has access to the best available data.
+We then assess the data available and provide recommendations for improving forest data collection, analysis, and reporting.
 <!-- (3) key results-->
@@
 Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests). 
+However, XX% of ForC records cannot be transferred because the original publication lacked the necessary information required by the IPCC.
 <!-- (4) the main conclusion, including the key points of discussion-->
-- In the future, forest C estimates in EFDB can be improved through targeted research to fill critical gaps, reporting of information required by IPCC, and continued submission of data from scientific publications to the EFDB.
-Given that climate change is rapidly impacting the world's forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
+- In the future, forest C estimates should be targeted towards other carbon pools and tropical forests, and studies should ensure that they report all necessary information to enable data integration into the EFDB.
+Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
+
+<!--suggestions from Sandro: “In the next future, forest C IPCC default values can be improved based on data populated in the EFDB, being thus of high relevance any process for the population of the EFDB with large datasets, especially to fill critical gaps in it.
+Given that climate change is rapidly impacting the world’s forests, timely uploading in the EFDB of recent data is critical to support quality of estimates C stock gains and losses reported in national GHG inventories.”-->
 

### 2023-10-31 doc/manuscript/ForC-EFDB manuscript.Rmd
change title, refine abstract

@@
 
-**Title:** Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)
+**Title:** Informing forest carbon inventories under the Paris Agreement using ground-based forest monitoring data
 
@@
 <!-- (1) the research conducted or aims, including the rationale-->
-- Forests are recognized as a critical source of additional climate change mitigation needed to meet the goals of the Paris Agreement. 
-However, the efficacy climate change mitigation planning and reporting depends on having high quality of data on forest carbon (C) stocks and changes.
-The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data, but needs comprehensive and well-documented data to be so.
+- Forests are recognized as critical to climate change mitigation under the Paris Agreement. 
+<!--Forests are recognized as a critical source of additional climate change mitigation needed to meet the goals of the Paris Agreement. -->
+However, the efficacy of climate change mitigation planning and reporting depends on having high quality data on forest carbon (C) stocks and changes.
+The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data.<!--, but needs comprehensive and well-documented data to be so.-->
 <!-- (2) methods-->
-- The Global Forest C database (ForC) represents a compilation of tens of thousands of forest C estimates that is fully traceable and publicly available.
-Here, we develop and document a process for semi-automated submission of data from ForC into the EFDB to help ensure that the EFDB has access to the best available data.
-We then assess the data available and provide recommendations for improving forest data collection, analysis, and reporting.
+- The Global Forest C database (ForC) is a compilation of tens of thousands of ground-based forest C estimates that is fully traceable and publicly available.
+Here, we develop and document a process for semi-automated submission of data from ForC into EFDB, <!-- to help ensure that the EFDB has access to the best available data.
+--> assess the data currently available through ForC, and provide recommendations for improving forest data collection, analysis, and reporting.
 <!-- (3) key results-->
@@
 Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests). 
-However, XX% of ForC records cannot be transferred because the original publication lacked the necessary information required by the IPCC.
+However, `r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_sent_to_EFDB,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked information required by the IPCC.
 <!-- (4) the main conclusion, including the key points of discussion-->
-- In the future, forest C estimates should be targeted towards other carbon pools and tropical forests, and studies should ensure that they report all necessary information to enable data integration into the EFDB.
+- In the future, ground-based forest C estimates should be target gaps in the record, and studies should ensure that they report all information necessary for inclusion in EFDB. 
 Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
@@
 Given that climate change is rapidly impacting the world’s forests, timely uploading in the EFDB of recent data is critical to support quality of estimates C stock gains and losses reported in national GHG inventories.”-->
+
+<!--
+# 200 word summary submitted to Anthromes special issue
+The efficacy of climate change mitigation planning and reporting depends on high quality data on forest carbon (C) stocks and changes.
+The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended as a definitive source for such data.
+The Global Forest C database (ForC) is a compilation of tens of thousands of ground-based forest C estimates that is fully traceable and publicly available.
+Here, we develop and document a process for semi-automated submission of data from ForC into EFDB,  assess the data currently available through ForC, and provide recommendations for improving forest data collection, analysis, and reporting.
+ForC contained ~19,316 independent records relevant to EFDB, 1,438 of which had undergone necessary review and been submitted.
+Records were unevenly distributed across variables and geographic regions.
+59% of ForC records reviewed could not be submitted because the original publication lacked information required by the IPCC.
+In the future, ground-based forest C estimates should be target gaps in the record, and studies should ensure that they report all information necessary for inclusion in EFDB.
+Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
+-->
 

### 2024-01-25 doc/manuscript/ForC-EFDB manuscript.Rmd
add final number posted to EFDB
#50

@@
 
-n_records_posted_EFDB = 376 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
+n_records_posted_EFDB = 1068 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
 Then get number from upper left box where it says "displayed records")
@@
 Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB.
-Complete records needed to be reviewed by the EFDB editorial board and then posted in the database -- a process that lags behind submission of records and had not yet been completed for all records sent as of `r format(Sys.time(), '%B %d, %Y')`.
 
@@
 The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
-The discrepancy between the number of records sent and that posted to EFDB is primarily attributable to the time required for the IPCC to review and post the records, and also because a minority of records were deemed not applicable to EFDB by the review panel.
+The discrepancy between the number of records sent and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
 

### 2024-01-26 doc/manuscript/ForC-EFDB manuscript.Rmd
word-smithing introduction

@@
 However, the efficacy of climate change mitigation planning and reporting depends on having high quality data on forest carbon (C) stocks and changes.
-The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data.<!--, but needs comprehensive and well-documented data to be so.-->
+The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data, but needs comprehensive and well-documented data to be so.
 <!-- (2) methods-->
@@
 <!--
-# 200 word summary submitted to Anthromes special issue
+# 200 word summary submitted to Anthromes special issue (invited for full submission)
 The efficacy of climate change mitigation planning and reporting depends on high quality data on forest carbon (C) stocks and changes.
@@
 
+- helps bridge scientific research accounting under Paris agreement
+
 \newpage
@@
 Forests are critical to management of the atmospheric concentration of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
-In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 Gt C yr^-1^ [all biomes with trees, @xu_changes_2021] to ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al.
-Table 2--> [forests only, @harris_global_2021].
+In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021].
+<!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^-->
+<!--1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests-->
 This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
 Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
-The future of this important CO~2~ sink is highly uncertain, and depends upon both forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020;@hammond_global_2022], and human conservation, restoration, and management of forests [@ipcc_climate_2019;@ipcc_climate_2022].
+Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
+The future of this important CO~2~ sink is highly uncertain, depending upon both direct human impacts on forests [i.e., deforestation/ degradation vs.
+conservation/ restoration, @ipcc_climate_2019;@ipcc_climate_2022] and forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].
 
@@
 
-The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change [UNFCCC, @ipcc_2006_2006;@ipcc_2019_2019].
+The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change (UNFCCC) [@ipcc_2006_2006;@ipcc_2019_2019].
 Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
@@
 Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
-For example, the year following the release of the latest IPCC guidelines, a more thorough analysis of C accumulation in regrowth forests found that IPCC's Tier 1 default failed to capture eight-fold variation within ecozones [@cook-patton_mapping_2020].
-In addition, it was revealed that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
+For example, following the 2019 release of the latest IPCC guidelines, it was revealed that IPCC's Tier 1 default failed to capture eight-fold variation of C accumulation in regrowth forests within ecozones [@cook-patton_mapping_2020] and that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
 This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
@@
 The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
-The EFDB encourages researchers to submit estimates of emission factors or other related parameters from peer-reviewed journal papers or other accepted sources for inclusion in the database. 
-In the case of forests, emission factors include C stocks, net annual increments, and annual fluxes for various pools [@ipcc_2006_2006;@ipcc_2019_2019].
+The EFDB encourages researchers to submit estimates of emission factors or other related parameters [e.g., C stocks, net annual increments, and annual fluxes for various pools [@ipcc_2006_2006;@ipcc_2019_2019] from peer-reviewed journal articles or other accepted sources for inclusion in the database.
+Needed is an efficient system for submission of data to the EFDB.
 
@@
 As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
+
+*(The following needs work:)*
+
 The purpose of this publication is to document that process and provide recommendations for future improvements.

### 2024-01-29 doc/manuscript/ForC-EFDB manuscript.Rmd
revise intro

@@
 n_records_posted_EFDB = 1068 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
-Then get number from upper left box where it says "displayed records")
+Then get number from upper left box where it says "displayed records".
+Updated Jan 2024)
+percent_forest_records = round(n_records_posted_EFDB/5644*100,0); #as of Jan 2024
 
@@
 <!-- (1) the research conducted or aims, including the rationale-->
-- Forests are recognized as critical to climate change mitigation under the Paris Agreement. 
+- *(Humans have been influencing Earth's climate via interactions with forests for thousands of years, and now)* Forests are recognized as critical to climate change mitigation under the Paris Agreement.
 <!--Forests are recognized as a critical source of additional climate change mitigation needed to meet the goals of the Paris Agreement. -->
@@
 
-Forests are critical to management of the atmospheric concentration of the greenhouse gas carbon dioxide (CO~2~), and thereby climate change. 
-In recent decades, CO~2~ uptake by forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021].
+Humans have been influencing Earth's climate via ecologically transformative land use practices for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016].
+In recent decades, as anthropogenic climate change has accelerated, this relationship has come into increasing focus [@ipcc_climate_2019;@ipcc_climate_2022].
+Deforestation and forest degradation are substantial sources of the greenhouse gas carbon dioxide (CO~2~), currently accounting for >10% of anthropogenic emissions [@friedlingstein_global_2022].
+Yet at the same time, CO~2~ uptake by remaining forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021],
 <!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^-->
 <!--1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests-->
-This has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
+which has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
 Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021], slowing the pace of atmospheric CO~2~ accumulation and associated climate change.
-The future of this important CO~2~ sink is highly uncertain, depending upon both direct human impacts on forests [i.e., deforestation/ degradation vs.
-conservation/ restoration, @ipcc_climate_2019;@ipcc_climate_2022] and forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].
+Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021].
+The future of this important CO~2~ sink depends primarily upon future trajectories of direct human impacts on forests [i.e., deforestation/ degradation vs.
+conservation/ restoration, [@ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].
 
-Forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
+Given this important role of forests in climate regulation, they play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
 Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
@@
 The EFDB can be used both for efforts to tally a nation's intended or accomplished greenhouse gas reductions, or as a basis of comparison for external parties to evaluate these inventories. 
-The EFDB encourages researchers to submit estimates of emission factors or other related parameters [e.g., C stocks, net annual increments, and annual fluxes for various pools [@ipcc_2006_2006;@ipcc_2019_2019] from peer-reviewed journal articles or other accepted sources for inclusion in the database.
-Needed is an efficient system for submission of data to the EFDB.
+The EFDB encourages researchers to submit estimates of emission factors or other related parameters [e.g., C stocks, net annual increments, and annual fluxes for various pools, @ipcc_2006_2006;@ipcc_2019_2019] from peer-reviewed journal articles or other accepted sources for inclusion in the database.
+Tens of thousands of relevant forest carbon estimates have been published -- and continue to be published at an accelerating rate -- but are not readily accessible to the practitioners assembling national greenhouse gas inventories.
+To contribute to the goal of making forest C parameters available for accounting under IPCC guidelines, forest scientists need an accessible summary of EFDB's requirements and an efficient system for submission of data to the EFDB.
+
+The goal of this publication is to lower the activation energy required for forest scientists to make their estimates of forest C stocks and changes therein useful and accessible for accounting under IPCC inventory guidelines.
+To accomplish this, we document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023], updated to facilitate data submission to EFDB (ForC v4.0).
+We (1) map common scientific forest C estimation methods and definitions to those used by the IPCC;
+(2) develop a semi-automated process for preparing ForC data for submission to EFDB;
+and (3) assess the data in ForC relevant to EFDB and records that have been submitted to date;
+We conclude with recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
+
+# Materials and Methods
 
@@
 As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
-
-*(The following needs work:)*
-
-The purpose of this publication is to document that process and provide recommendations for future improvements.
-
-Here, we
-(1) review IPCC methods and definitions applied to estimate CO~2~ emissions and removals from forest in the context of typical forest C estimation methodologies;
-(2) describe mapping of ForC to IPCC's EFDB;
-(3) describe updates to ForC  (ForC v4.0), most of which were implemented to facilitate data submission to EFDB;
-(4) summarize the data in ForC relevant to EFDB and records that have been submitted to date; and
-(5) provide recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
-
-# Materials and Methods
 
@@
 The discrepancy between the number of records sent and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
+As of `r format(Sys.time(), '%B %d, %Y')`, the `r n_records_sent_to_EFDB` ForC records posted in EFDB represented `r percent_forest_records`% of the total EFDB records for forest land.
 

### 2024-01-29 doc/manuscript/ForC-EFDB manuscript.Rmd
worked on abstract & intro

@@
 <!-- (1) the research conducted or aims, including the rationale-->
-- *(Humans have been influencing Earth's climate via interactions with forests for thousands of years, and now)* Forests are recognized as critical to climate change mitigation under the Paris Agreement.
-<!--Forests are recognized as a critical source of additional climate change mitigation needed to meet the goals of the Paris Agreement. -->
-However, the efficacy of climate change mitigation planning and reporting depends on having high quality data on forest carbon (C) stocks and changes.
+- Humans have been influencing Earth's climate via transformative impacts on forests for >12,000 years, and forests are now recognized as critical to climate change mitigation under the Paris Agreement.
+However, the efficacy of climate change mitigation planning and reporting depends on having quality data on forest carbon (C) stocks and changes.
 The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data, but needs comprehensive and well-documented data to be so.
 <!-- (2) methods-->
-- The Global Forest C database (ForC) is a compilation of tens of thousands of ground-based forest C estimates that is fully traceable and publicly available.
-Here, we develop and document a process for semi-automated submission of data from ForC into EFDB, <!-- to help ensure that the EFDB has access to the best available data.
---> assess the data currently available through ForC, and provide recommendations for improving forest data collection, analysis, and reporting.
+- To facilitate submission of forest C estimates from scientific studies to EFDB, we develop and document a process for semi-automated data submission from the Global Forest C database (ForC v4.0), which is the largest compilation of ground-based forest C estimates.
+We then assess the data currently available through ForC and provide recommendations for improving forest data collection, analysis, and reporting.
 <!-- (3) key results-->
-- As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which had undergone necessary review and been submitted.
+- As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_posted_EFDB` of which had been submitted and posted to EFDB.
 Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests).
-However, `r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_sent_to_EFDB,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked information required by the IPCC.
+`r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_sent_to_EFDB,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked required information.
 <!-- (4) the main conclusion, including the key points of discussion-->
-- In the future, ground-based forest C estimates should be target gaps in the record, and studies should ensure that they report all information necessary for inclusion in EFDB. 
+- In the future, ground-based forest C estimates should target gaps in the record, and studies should ensure that they report all information necessary for inclusion in EFDB. 
 Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
@@
 Humans have been influencing Earth's climate via ecologically transformative land use practices for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016].
-In recent decades, as anthropogenic climate change has accelerated, this relationship has come into increasing focus [@ipcc_climate_2019;@ipcc_climate_2022].
+In recent decades, as both anthropogenic land transformation and climate change have accelerated, this relationship has come into increasing focus [@ipcc_climate_2019].
 Deforestation and forest degradation are substantial sources of the greenhouse gas carbon dioxide (CO~2~), currently accounting for >10% of anthropogenic emissions [@friedlingstein_global_2022].
 Yet at the same time, CO~2~ uptake by remaining forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021],
-<!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^-->
-<!--1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests-->
-which has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
-Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
+<!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^
+-1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests-->
+ofsetting an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
+Table 2)--> - `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
 Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021].
 The future of this important CO~2~ sink depends primarily upon future trajectories of direct human impacts on forests [i.e., deforestation/ degradation vs.
-conservation/ restoration, [@ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].
+conservation/ restoration, @ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].
 
-Given this important role of forests in climate regulation, they play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
+In recognition of their important role in climate regulation, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015; @grassi_key_2017].<!--As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].-->
 Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
-As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
 Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
@@
 The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change (UNFCCC) [@ipcc_2006_2006;@ipcc_2019_2019].
-Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
+Under this guidance, greenhouse gas inventories include all managed land, which includes most of the world's forest land [@ogle_delineating_2018; @ellis_anthropogenic_2010].
 The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
@@
 
-The goal of this publication is to lower the activation energy required for forest scientists to make their estimates of forest C stocks and changes therein useful and accessible for accounting under IPCC inventory guidelines.
-To accomplish this, we document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023], updated to facilitate data submission to EFDB (ForC v4.0).
+The goal of this publication is to facilitate submission of forest C estimates from scientific studies to EFDB.
+We document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023].
 We (1) map common scientific forest C estimation methods and definitions to those used by the IPCC;
@@
 As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
+<!--ForC is fully traceable and publicly available.-->
 

### 2024-01-29 doc/manuscript/ForC-EFDB manuscript.Rmd
worked on methods

@@
 Under this guidance, greenhouse gas inventories include all managed land, which includes most of the world's forest land [@ogle_delineating_2018; @ellis_anthropogenic_2010].
-The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
+The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [summarized in `r appendix_IPCC_definitions`, @ipcc_agriculture_2006; @ipcc_2019_2019].
 A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
@@
 The goal of this publication is to facilitate submission of forest C estimates from scientific studies to EFDB.
-We document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023].
+We document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023], including data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
 We (1) map common scientific forest C estimation methods and definitions to those used by the IPCC;
@@
 
-The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023].
-ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
-As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
-<!--ForC is fully traceable and publicly available.-->
-
-Major steps included
+Major steps for submission of data from ForC to EFDB included
 (1) mapping ForC into EFDB, including aligning ForC and IPCC definitions (summarized in `r appendix_IPCC_definitions`),
@@
 
-With input from the IPCC's Technical Support Unit and referencing IPCC guidance [@refs], we mapped ForC fields into EFDB fields, as summarized in `r table_ForCfieldmapping`.
+With input from the IPCC's Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we mapped ForC fields into EFDB fields, as summarized in `r table_ForCfieldmapping`.
 For the majority of fields, contents of the field in ForC was copied directly into an EFDB field, either as the only contents of that field or as part of a composite record.
-For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).
-In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present information required to calculate these (standard error or n and standard deviation), we calculated the 95% confidence intervals and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*).
-For some fields, simple conditional logic was used to populate EFDB fields based on ForC records. 
-For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.
+<!--For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).-->
+For some fields, simple conditional logic was used to populate EFDB fields based on ForC records.<!--For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.-->
+In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present standard error or n and standard deviation, we calculated the 95% confidence intervals.<!--and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*)-->
 There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (detailed below in **Carbon cycle variables**) and (2) land classification (detailed below in **Land classification**).

### 2024-01-29 doc/manuscript/ForC-EFDB manuscript.Rmd
minor

@@
 
-# Introduction
+# Introduction {-}
 
@@
 The goal of this publication is to facilitate submission of forest C estimates from scientific studies to EFDB.
-We document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023], including data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
+We document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023], including data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the Global Soil Respiration Database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
 We (1) map common scientific forest C estimation methods and definitions to those used by the IPCC;
 (2) develop a semi-automated process for preparing ForC data for submission to EFDB;
-and (3) assess the data in ForC relevant to EFDB and records that have been submitted to date;
+and (3) assess the data in ForC relevant to EFDB and records that have been submitted to date.
 We conclude with recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
 
-# Materials and Methods
+# Materials and Methods {-}
 
 Major steps for submission of data from ForC to EFDB included
-(1) mapping ForC into EFDB, including aligning ForC and IPCC definitions (summarized in `r appendix_IPCC_definitions`),
+(1) mapping ForC into EFDB, including aligning ForC terms and concepts with those defined by IPCC guidelines (summarized in `r appendix_IPCC_definitions`),
 (2) revising ForC v3.0 to support semi-automated submissions to EFDB, yielding ForC v4.0 (detailed in `r appendix_ForC_updates`), and
@@
 
-## Mapping ForC to EFDB
+## 1. Mapping ForC to EFDB {-}
 
 With input from the IPCC's Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we mapped ForC fields into EFDB fields, as summarized in `r table_ForCfieldmapping`.
-For the majority of fields, contents of the field in ForC was copied directly into an EFDB field, either as the only contents of that field or as part of a composite record.
+For the majority of fields, contents of the field in ForC mapped directly into an EFDB field, either as the only contents of that field or as part of a composite record.
 <!--For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).-->
 For some fields, simple conditional logic was used to populate EFDB fields based on ForC records.<!--For example, for stock variables presented in the original publication in units of dry organic matter mass (as opposed to C), several greenhouse gasses (CO~2~, CO, CH~4~, NO, NO~2~, N~2~O) were entered in the EFDB field indicating the greenhouse gases to which the record could be pertinent (*Gases* field) because these values could be used in calculations of greenhouse gas emissions from biomass burning [@ipcc_2006_2006]; otherwise, the only pertinent greenhouse gas would be CO~2~.-->
-In cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present standard error or n and standard deviation, we calculated the 95% confidence intervals.<!--and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*)-->
+For example, in cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present standard error or n and standard deviation, we calculated the 95% confidence intervals.<!--and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*)-->
 There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (detailed below in **Carbon cycle variables**) and (2) land classification (detailed below in **Land classification**).
 
-### Carbon cycle variables
+### Carbon cycle variables {-}
 
@@
 
-### Land classification
+### Land classification {-}
 
@@
 
-## Updating ForC
+## 2. Updating ForC {-}
 
@@
 
-## Submission of ForC data to EFDB
+## 3. Submission of ForC data to EFDB {-}
 
@@
 
-# Results
+# Results {-}
 
-## ForC v4.0 contents
+## ForC v4.0 contents {-}
 
@@
 
-## Data submissions to EFDB
+## Data submissions to EFDB {-}
 
@@
 
-# Discussion
+# Discussion {-}
 
@@
 
-## Database needs
+## Database needs {-}
 
@@
 
-## Data collection and analysis needs
+## Data collection and analysis needs {-}
 
@@
 
-## Data reporting needs
+## Data reporting needs {-}
 
@@
 
-# Conclusions
+# Conclusions {-}
 
@@
 
-# Acknowledgements
+# Acknowledgements {-}
   We gratefully acknowledge the substantial contributions of Valentyna Slivinska and Sandro Federici for collaboration on the conception, design, and technical review of this project.
@@
 
-# Author Contribution
+# Author Contribution {-}
   KAT and VH conceived and designed the project; VH wrote the scripts for database management, data submission to EFDB, and the analyses presented here; MW, TR, and RBM added and reviewed ForC data, BBL and SCP contributed large databases to ForC (EFDB and GROA, respectively); CP provided methodological expertise; KAT, VH, and MW prepared the first draft of the manuscript; all authors reviewed the results and approved the final version of the manuscript.
 
-# Data Availability Statement 
+# Data Availability Statement  {-}
 All code and data are openly available.
@@
 \clearpage
-# References
+# References {-}

### 2024-01-31 doc/manuscript/ForC-EFDB manuscript.Rmd
methods

@@
 
-With input from the IPCC's Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we mapped ForC fields into EFDB fields, as summarized in `r table_ForCfieldmapping`.
-For the majority of fields, contents of the field in ForC mapped directly into an EFDB field, either as the only contents of that field or as part of a composite record.
+With input from the IPCC's Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we determined how EFDB fields should be populated using ForC fields (summarized in `r table_ForCfieldmapping`).
+Fields in EFDB included several fields describing how the record fits within IPCC's framework (source/sink category, greenhouse gas type, C pool, relevant equations), several describing the C estimate itself (variable, value, units, 95% confidence limits), a few composite fields describing biotic and abiotic conditions (e.g., vegetation type, minimum diameter included, site location, climate, edaphic properties, notable disturbances, stand age, plot information), and a few describing the source (e.g., type, citation, data provider).
+Most relevant ForC fields mapped directly into an EFDB field, either as the only contents of that field or as part of a composite record.
 <!--For example, ten ForC fields describing site location, climate, and edaphic properties all mapped into the EFDB field *Region/Regional conditions* (`r table_ForCfieldmapping`).-->

### 2024-01-31 doc/manuscript/ForC-EFDB manuscript.Rmd
Update ForC-EFDB manuscript.Rmd

@@
 
-We reviewed the list of ForC variables to identify those that were relevant to EFDB and to appropriately map them into EFDB (Fig. `r fig_variable_mapping`).
-For each C pool (Table `r table_pools`), we identified variables representing organic matter or C stocks, net annual increments, influxes (a.k.a. "gross annual increments" by IPCC), and outfluxes.
-As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC. 
+We reviewed the list of ForC variables to identify those that were relevant to the IPCC methodology and to EFDB (`r appendix_IPCC_definitions`).
+These included  organic matter or C stocks, net annual increments, influxes (a.k.a.
+"gross annual increments" by IPCC), and outfluxes for each IPCC-defied C pool (Table `r table_pools`, Fig.
+`r fig_variable_mapping`).
+As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC (`r appendix_ForC_updates`). 
 It is important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
 For example, ForC records of biomass and dead wood vary in the minimum stem diameter censused, such that some records would match the IPCC criteria whereas others would not.
-Information on minimum diameters censused and other important sources of methodological variation are recorded as covariates in ForC and mapped into the EFDB field *Other Properties* (`r table_ForCfieldmapping`).
-Details on the mapping of ForC variables to EFDB -- including associated covariates, IPCC pools (Table `r table_pools`) and relevant equations [@ipcc_2006_2006] -- are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
+Information on minimum diameters censused and other important sources of methodological variation were mapped into EFDB (`r table_ForCfieldmapping`).
+Details on the mapping of ForC variables to EFDB are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
 

### 2024-02-02 doc/manuscript/ForC-EFDB manuscript.Rmd
revised/shortened methods & results

@@
 table_pools =1
-table_variables =2
+
 table_recommendations = 3
@@
 table_ForCfieldmapping = "Table S2"
+table_variables ="S3"
 
@@
 The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change (UNFCCC) [@ipcc_2006_2006;@ipcc_2019_2019].
-Under this guidance, greenhouse gas inventories include all managed land, which includes most of the world's forest land [@ogle_delineating_2018; @ellis_anthropogenic_2010].
-The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [summarized in `r appendix_IPCC_definitions`, @ipcc_agriculture_2006; @ipcc_2019_2019].
+Under this guidance, greenhouse gas emissions to, or withdrawals from, the atmosphere are quantified on an annual basis for all managed land, which includes most of the world's forest land [@ogle_delineating_2018; @ellis_anthropogenic_2010].
+The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [`r appendix_IPCC_definitions`, @ipcc_agriculture_2006; @ipcc_2019_2019].
 A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
@@
 
-We reviewed the list of ForC variables to identify those that were relevant to the IPCC methodology and to EFDB (`r appendix_IPCC_definitions`).
+We identified ForC variables that were relevant to the IPCC methodology and to EFDB (`r appendix_IPCC_definitions`).
 These included  organic matter or C stocks, net annual increments, influxes (a.k.a.
@@
 `r fig_variable_mapping`).
-As described in section `r section_new_variables`, we also defined `r n_new_variables` new EFDB-relevant variables that were not previously represented in ForC (`r appendix_ForC_updates`). 
 It is important to note that the correspondence of ForC variables to IPCC criteria often depends upon measurement protocols ("important sources of estimate variation" in Table `r table_pools`).
@@
 
-Determination of the IPCC land-use category (i.e., Forest Land, Grassland, Wetlands, Cropland, Settlements, or Other Land; section `r  section_land_use_categories`) was made based on the categorical ForC field *dominant.life.form*, sometimes drawing upon stand age.
-Records with "woody" *dominant.life.form* were classified as Forest Land.
-Those with *dominant.life.form* of "woody+grass", which in ForC is indicative of anything from a shrub-encroached grassland to a tree-dominated savanna, were given dual classification of Forest Land and Grassland.
-This dual classification indicates that records may be relevant to either category depending on the definition of forest applied (varies by country).
-For (rare) cases where *dominant.life.form* was grass and stand age was greater than zero, indicative of early successional vegetation, we assigned a classification of Forest Land, consistent with the IPCC definition that Forest Land includes land expected to succeed to forest.
-Cases where *dominant.life.form* was grass or crop and stand age was zero were indicative of a control for studies of forest regrowth following agricultural abandonment, and were classified as Grassland and Cropland, respectively.
+Determination of the IPCC land-use category (i.e., Forest Land, Grassland, Wetlands, Cropland, Settlements, or Other Land; `r appendix_IPCC_definitions`) was made based on the dominant life form recorded in ForC.
+Woody vegetation -- including early seral vegetation -- was classified as forest, consistent with the IPCC definition that Forest Land includes land expected to succeed to forest.
+Mixtures of woody vegetation and grasses (i.e., anything from a shrub-encroached grassland to a tree-dominated savanna) were given dual classification of Forest Land and Grassland, indicating that records may be relevant to either category depending on the definition of forest applied (varies by country).
+Grass- or crop-dominated ecosystems (included in ForC as controls for studies of forest regrowth following agricultural abandonment) were classified as Grassland and Cropland, respectively.
 
-Classification into sub-categories was dependent upon stand age and site history (section `r  section_land_use_categories`).
+Classification into sub-categories was dependent upon stand age and site history.
 For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
@@
 
-To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form ("EFDB bulk import.xlsx").
-Criteria for data submission were that (1) records had been checked against the original study and determined to be complete and correct, and as originally presented, (2) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables, and (3) the records had not previously been submitted to EFDB.
+To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form.
+Criteria for data submission were that
+(1) records had been checked against the original study and determined to be complete and correct, and as originally presented,
+(2) all relevant records within the publication were included in ForC,
+(3) the original study presented values in tables or text, as opposed to the values having been digitized from graphs or calculated based on related variables,
+and (4) the records had not previously been submitted to EFDB.
 Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB.
@@
 
-## ForC v4.0 contents {-}
-
-As of `r format(Sys.time(), '%B %d, %Y')`, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig.
+As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig.
 `r fig_variable_mapping`).
-These records were distributed across all forested continents and ecozones, albeit unevenly (Fig. `r fig_map`).
-The largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig. `r fig_histograms`c).
-Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
-`r fig_histograms`b).
-Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
-The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
-In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).
-
-![**Figure `r fig_map`.
-Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.** Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
-
-\newpage
-![**Figure `r fig_histograms`.
-Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.** For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)
-
-ForC contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
-The records were very unevenly distributed across variables. 
-The variable with most records was aboveground biomass, representing `r round(n_ind_records_agb/n_ind_records_EFDB_variables,2)*100`% of all independent records relevant to EFDB, and aboveground biomass components (woody biomass or foliage) representing an additional `r round(n_ind_records_agb_components/n_ind_records_EFDB_variables,2)*100`%.
-A total of `r round(n_ind_records_rootbiomass/n_ind_records_EFDB_variables,2)*100`% of relevant records were for root biomass (including fine and coarse root components), while `r round(n_ind_records_biomass/n_ind_records_EFDB_variables,2)*100`% described total biomass.
-The non-living pools were less represented, with `r round(n_ind_records_deadwood/n_ind_records_EFDB_variables,2)*100`% of relevant were for dead wood (including standing and fallen components), `r round(n_ind_records_OL/n_ind_records_EFDB_variables*100,1)`% for litter, `r round(n_ind_records_TotEcosystemC/n_ind_records_EFDB_variables*100,1)`% for total ecosystem C excluding soils, and `r round(n_ind_records_SOC/n_ind_records_EFDB_variables*100,1)`% for soil carbon.
-
-Increment and flux variables were poorly represented (Table `r table_variables`).
-The increment variable with most records was the aboveground biomass increment, representing `r round(n_ind_records_deltaagb/n_ind_records_EFDB_variables*100,1)`% of all independent records relevant to EFDB.
-The only other relevant increment variable with any records was the O horizon (litter) increment, with just `r n_ind_records_deltaOL` records.
-Relevant flux variable records (n=`r n_ind_records_flux`) together constituted `r round(n_ind_records_flux/n_ind_records_EFDB_variables,2)*100`% of ForC's independent records relevant to EFDB.
-
-\newpage
-**Table `r table_variables`. Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.**
-```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
-C_variables_table <- read.csv("figures_tables/C_variables.csv")
-names(C_variables_table) <- stringr::str_replace_all(names(C_variables_table), "\\.", " ")
-C_variables_table[is.na(C_variables_table)]<- "" #replace NAs with blanks
-
-library(flextable)
-set_flextable_defaults(font.family = "Arial", font.size = 8) # set font
-
-ft <- flextable(C_variables_table[,-1])
-ft <- autofit(ft)
-ft <- width(ft, 1, width = 2)
-ft <- width(ft, 2:5, width = 1)
-ft <- bold(ft, i = 1, j = NULL, bold = TRUE, part = "header") # put title in bold
-ft <- bold(ft, i = which(C_variables_table$bold), j = NULL, bold = TRUE, part = "body") # put categories in bold
-ft
-```
-
-## Data submissions to EFDB {-}
-
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had reviewed or added `r n_records_reviewed_or_added` EFDB-relevant records, `r n_records_sent_to_EFDB` records of which were submitted to EFDB, and `r n_records_posted_EFDB` of which have been reviewed, accepted, and posted (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`).
-The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records reviewed and those sent to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
+We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which met the criteria for submission, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (Table `r table_variables`).
+The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records we reviewed and those submitted to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
 The discrepancy between the number of records sent and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
@@
 
-The ForC records submitted to EFDB were broadly distributed across Earth's forests (Fig. `r fig_map`).
-However, the density of these records was very unevenly distributed across continents, biomes, and forest types and was not proportional to the numbers of relevant records in ForC (Fig.
-`r fig_histograms`).
-Rather, the largest number of records came from North America, followed by Asia, South America, and Africa (Fig.
-`r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.
-`r fig_histograms`b).
-In terms of dominant vegetation, by far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
-The largest records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
-
-In terms of variables, records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
-The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records sent were for C stocks, including
+ForC v4.0 contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Fig.
+`r fig_variable_mapping`, Table `r table_variables`).
+Records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
+The records available in ForC v4.0 -- and the subset submitted to EFDB -- were very unevenly distributed across variables (Table `r table_variables`). 
+The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records submitted were for C stocks, including
 `r round(n_records_sent_biomass/n_records_sent_to_EFDB,2)*100`% for total biomass, 
@@
 and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
-Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records sent, virtually all for aboveground biomass (excepting `r n_records_sent_delta.litter` records for delta.O.horizon).
-The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records sent described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Table `r table_variables`, Fig.
-`r fig_variable_mapping`).
+Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records submitted, all but four of which described aboveground biomass.
+The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records submitted described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Fig.
+`r fig_variable_mapping`,Table `r table_variables`).
+
+ForC v4.0 records and the subset submitted to EFDB were distributed across all forested continents, biomes, and forest types, albeit very unevenly (Figs. `r fig_map`-`r fig_histograms`).
+Among the records submitted to EFDB, the largest number came from North America, followed by Asia, South America, and Africa (Fig.
+`r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.
+`r fig_histograms`b).
+In terms of dominant vegetation, by far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
+The largest number of records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
+
+<!--In ForC v4.0, the largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig.
+`r fig_histograms`c).
+Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
+`r fig_histograms`b).
+Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
+The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
+In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).-->
+
+![**Figure `r fig_map`.
+Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.** Symbols are colored according to the number of records at each site.
+Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
+
+![**Figure `r fig_histograms`.
+Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.** For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)
 

### 2024-02-02 doc/manuscript/SI_IPCCdefs.Rmd
worked on methods
revised/shortened methods & results

@@
 
-# `r appendix_IPCC_definitions`. IPCC methods and definitions {-}
+# `r appendix_IPCC_definitions`. Primer on forest land classification and carbon pools under IPCC guidelines {-}
 
-The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019].
-For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (see **Carbon Pools** section below), plus any harvested wood products.
+## Land classification {-}
+IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
+Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
+Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
+Definitions of forest are allowed to vary by country, but must be applied consistently.
+Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
+
+The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
+This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
+However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
+
+## Carbon pools {-}
+
+For each stratum of subdivision within a land-use category (see **Land classification** section below), annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (see **Carbon pools** section below), plus any harvested wood products.
 Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
-
-## Carbon Pools {-}
 
@@
 
-## Land classification {-}
-IPCC defines land-use categories to include six categories -- Forest Land, Grassland, Wetlands, Cropland, Settlements, and Other Land [@ipcc_2006_2006].
-Sub-divisions include land that has remained in a particular category for >20 years (e.g., Forest Land remaining Forest Land) and land that has been converted from one category to another in the past 20 years (e.g., Cropland converted to Forest Land).
-Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
-Definitions of forest are allowed to vary by country, but must be applied consistently.
-Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
-
-The UNFCCC requires greenhouse gas reporting for all managed lands in a country, where management is defined as "human interventions and practices have been applied to perform production, ecological or social functions" [@ipcc_2006_2006].
-This expansive definition of managed land implies that the majority of Forest Land in most countries is managed.
-However, the definition is applied differently across countries, and the majority of governments have yet to report their approach for defining managed land or provide maps of managed land [@ogle_delineating_2018; @deng_comparing_2021].
-
 ## References {-}

### 2024-02-02 doc/manuscript/SI_tables.Rmd
revised/shortened methods & results

@@
 table_ForCfieldmapping = "Table S2"
+table_variables ="Table S3"
 
@@
 \* *Required* field indicates whether the field is required by EFDB: yes = value required; (yes) = input required, missing value acceptable if not reported; auto = present within ForC infrasructure, and therefore will always be exported to EFDB ; (no) = not required for EFDB, but required for ForC and therefore will always be exported to EFDB; no = not required, but exported to EFDB when a value is present.
+
+\clearpage
+
+\newpage
+**`r table_variables`. Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.**
+```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}
+C_variables_table <- read.csv("figures_tables/C_variables.csv")
+names(C_variables_table) <- stringr::str_replace_all(names(C_variables_table), "\\.", " ")
+C_variables_table[is.na(C_variables_table)]<- "" #replace NAs with blanks
+
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 8) # set font
+
+ft <- flextable(C_variables_table[,-1])
+ft <- autofit(ft)
+ft <- width(ft, 1, width = 2)
+ft <- width(ft, 2:5, width = 1)
+ft <- bold(ft, i = 1, j = NULL, bold = TRUE, part = "header") # put title in bold
+ft <- bold(ft, i = which(C_variables_table$bold), j = NULL, bold = TRUE, part = "body") # put categories in bold
+ft
+```

### 2024-02-02 doc/manuscript/ForC-EFDB manuscript.Rmd
revising manuscript

@@
 table_pools =1
-
-table_recommendations = 3
+table_recommendations = 2
 
@@
 
-#Cited sections (listed here because these can change)
-section_C_pools = 2.1 
-section_land_use_categories = 2.2
-section_new_variables = 3.2
-section_C_variable_mapping = "4.1.1"
-section_land_use_mapping = "4.1.2"
-section_database_needs = "6.1"
-section_new_data_needs = "6.2"
-section_data_reporting_needs = "6.3"
-
 #Appendices
@@
 table_ForCfieldmapping = "Table S2"
-table_variables ="S3"
+table_variables ="Table S3"
 
@@
 Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests).
-`r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_sent_to_EFDB,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked required information.
+`r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_reviewed_or_added,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked required information.
 <!-- (4) the main conclusion, including the key points of discussion-->
@@
 
+climate change, database, forest carbon, greenhouse gas inventory, International Panel on Climate Change (IPCC), natural climate solutions
+
 # Societal Impact Statement {-}
@@
 
-- helps bridge scientific research accounting under Paris agreement
+Human interactions with forests have shaped Earth's climate for >12,000 years and will continue to do so as we target net-zero emissions.
+Accurately characterizing these climate impacts requires accurate forest carbon data available for national greenhouse gas inventories conducted under the Paris Agreement. 
+Here we develop a semi-automated process for submitting forest carbon measurements from the largest relevant scientific database to the International Panel on Climate Change's Emission Factor Database, which currently has sparse forest carbon data.
+Building this bridge from scientific research to international policy is an important step towards managing forests in a net-zero motivated future.
 
@@
 `r fig_variable_mapping`).
-We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which met the criteria for submission, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (Table `r table_variables`).
+We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which met the criteria for submission, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (`r table_variables`).
 The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records we reviewed and those submitted to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
@@
 ForC v4.0 contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Fig.
-`r fig_variable_mapping`, Table `r table_variables`).
-Records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (Table `r table_variables`).
-The records available in ForC v4.0 -- and the subset submitted to EFDB -- were very unevenly distributed across variables (Table `r table_variables`). 
+`r fig_variable_mapping`,  `r table_variables`).
+Records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (`r table_variables`).
+The records available in ForC v4.0 -- and the subset submitted to EFDB -- were very unevenly distributed across variables (`r table_variables`). 
 The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records submitted were for C stocks, including
@@
 The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records submitted described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Fig.
-`r fig_variable_mapping`,Table `r table_variables`).
+`r fig_variable_mapping`,`r table_variables`).
 
@@
 
-** RECOMENDATIONS **
+Here we have developed a framework for submitting records from ForC v4.0, the largest compilation of ground-based forest C estimates, to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
+As of `r format(Sys.time(), '%B %d, %Y')`, `r n_records_posted_EFDB` ForC records have been posted to EFDB, which represents just `r round(n_records_posted_EFDB/n_ind_records_EFDB_variables*100,0)`% of potentially relevant records in ForC v4.0, but currently comprises `r percent_forest_records`% of the total EFDB records for forest land.
+The records present in ForC and submitted to EFDB are very unevenly distributed across variables, regions, and forest types (Figs.
+`r fig_map`-`r fig_histograms`, `r table_variables`), reflecting broader patterns in allocation of research effort and pointing to scientific research and reporting needs.
 
-Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work (section `r section_database_needs`), new data collection and analysis (section `r section_new_data_needs`), and reporting (section `r section_data_reporting_needs`).
+Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work, new data collection and analysis, and reporting.
 
-## Database needs {-}
+### Database needs {-}
 
@@
 Future efforts to review studies for submission should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
+`r fig_map`-`r fig_histograms`, `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
 
@@
 
-## Data collection and analysis needs {-}
+### Data collection and analysis needs {-}
 
-New data collection and analysis is needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
-Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
-Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
+New data collection and analyses are needed to fill notable knowledge gaps.
+While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [`r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
+Other pools and variables remain very poorly quantified and highly uncertain for many parts of the world [`r table_variables`,@tifafi_large_2018; @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
+Furthermore, data distribution is very uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
 For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
@@
 Several EFDB-relevant variables have not been calculated and presented as frequently as would be possible given existing forest census data and minimal extra research effort.
-For example, aboveground woody mortality (*woody.mortality_ag*) and aboveground biomass increment (*delta.agb*) can be calculated from the same census data as aboveground woody productivity (*ANPP_woody*), yet the latter has received far more research attention, and correspondingly has far more records in ForC [Table `r table_variables`, @anderson-teixeira_carbon_2021; but see @piponiot_distribution_2022].
-Similarly, live coarse root biomass, total biomass, and changes in both of these pools could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
+For example, aboveground woody mortality and aboveground biomass increment can be calculated from the same census data as aboveground woody productivity, yet the latter has received far more research attention [`r table_variables`, @anderson-teixeira_carbon_2021].
+Similarly, live coarse root biomass, total biomass, and changes therein could in theory be estimated in parallel with aboveground biomass, with the greatest barrier being that allometric models for estimating root biomass are not as reliable or easily available as are those for aboveground biomass [@chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022].
 However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017; @mokany_critical_2006].
-In addition, standing dead trees are captured in most forest censuses and could be used to estimate standing dead wood, although additional data on breakage would be needed for accurate estimates. 
-We recommend that, when possible, researchers calculate and report these variables, following the reporting guidelines specified in section `r section_data_reporting_needs`.
+We recommend that, when possible, researchers calculate and report all EFDB-relevant variables.
 
-Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for estimating forest C stock chnages.
-Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023;@calders_laser_2022].
-Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
-Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
-Significant investment in ground-based forest research will be critical to filling these gaps.
-
-## Data reporting needs {-}
+### Data reporting needs {-}
 
@@
 This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
-The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we welcome other researchers to use the ForC template.
-
-Once EFDB-relevant data are available in peer-reviewed publications, they may be submitted directly to EFDB or may use the ForC - EFDB data pipeline developed here.
-For individual publications, the former option will generally be more efficient.
-However, data incorporated into ForC as well as EFDB will be more broadly useful; for example, these data may be used for basic science [e.g., @banburymorgan_global_2021; @anderson-teixeira_carbon_2021], analyses of forest-based climate change mitigation potential [e.g., @cook-patton_mapping_2020; @goldstein_protecting_2020], and model benchmarking [@fer_ecosystem_2021].
+The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we hope that it will prove useful as a model for other efforts.
 
@@
 
-The ForC database contains large numbers of records that could potentially be useful for estimating C stock changes applying methodological guidance provided by the IPCC.
-Here we have developed a framework for submitting these records to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
-As of `r format(Sys.time(), '%B %d, %Y')`, we have submitted `r n_records_sent_to_EFDB` records to EFDB.
-Although this represents just `r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables*100,0)`% of relevant records in ForC, it substantially increases the number of forest land records in EFDB.
-The records submitted to EFDB and present in ForC are very unevenly distributed across variables, regions, and forest types (Figs.
-`r fig_map`-`r fig_histograms`, Table `r table_variables`), reflecting broader patterns in allocation of research effort.
-
-Going forward, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
+As human society strives to achieve net-zero greenhouse gas emissions, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
 In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under teh Paris Agreement.
@@
 
+# Supplementary Information
+`r appendix_IPCC_definitions`. Primer on forest land classification and carbon pools under IPCC guidelines
+
+`r appendix_ForC_updates`. Updates to ForC (ForC v4.0)
+
+`r table_ForCchanges`.  Updates to ForC field implemented between releases of v3.0 and v4.0
+
+`r table_ForCfieldmapping`. Mapping of ForC fields to EFDB
+
+`r table_variables`. Numbers of ForC records and EFDB submissions by variable
+
 \clearpage

### 2024-02-02 doc/manuscript/SI_ForCv40.Rmd
revising manuscript

@@
 table_pools =1
-table_variables =2
-table_recommendations = 3
+table_recommendations = 2
 
@@
 (3) whether records of interest were presented in tables or text or digitized from figures (EFDB will not accept digitized data), 
-(4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass)
+and (4) whether records of interest were presented directly, as opposed to having been calculated from related variables (for example, if a study presents aboveground biomass and root biomass but not total biomass, EFDB would not accept the sum of these as a valid record of total biomass).
 We also checked that existing ForC records were complete and correct.

### 2024-02-02 doc/manuscript/SI_tables.Rmd
revising manuscript

@@
 \newpage
-**`r table_variables`. Numbers of records of ForC variables (or closely related variable groups) relevant to, and sent to, EFDB.**
+**`r table_variables`. Numbers of ForC records and EFDB submissions by variable.**
 ```{r table_variables, eval = TRUE, echo=FALSE, warning=FALSE}

### 2024-02-13 doc/manuscript/ForC-EFDB manuscript.Rmd
BBL suggested edits

@@
 - As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_posted_EFDB` of which had been submitted and posted to EFDB.
+These represent 19% of the total EFDB records for forest land.
 Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests).
@@
 - In the future, ground-based forest C estimates should target gaps in the record, and studies should ensure that they report all information necessary for inclusion in EFDB. 
-Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.
+Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be critical to accurate forest C inventories.
 
@@
 
-Human interactions with forests have shaped Earth's climate for >12,000 years and will continue to do so as we target net-zero emissions.
+Human interactions with forests have shaped Earth's climate for millenia and will continue to do so as we target net-zero emission goals.
 Accurately characterizing these climate impacts requires accurate forest carbon data available for national greenhouse gas inventories conducted under the Paris Agreement. 
@@
 Deforestation and forest degradation are substantial sources of the greenhouse gas carbon dioxide (CO~2~), currently accounting for >10% of anthropogenic emissions [@friedlingstein_global_2022].
-Yet at the same time, CO~2~ uptake by remaining forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021],
+At the same time, CO~2~ uptake by remaining and regrowing forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021],
 <!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^
--1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests-->
+-1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests;
+BBL: might cite Ruehr et al. 2023 http://dx.doi.org/10.1038/s43017-023-00456-3 here-->
 ofsetting an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
@@
 conservation/ restoration, @ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].
+<!--here I would definitely cite the Ruehr et al. 2023 paper-->
 
-In recognition of their important role in climate regulation, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015; @grassi_key_2017].<!--As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].-->
+Accordingly, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015; @grassi_key_2017].<!--As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].-->
 Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
-Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
+However, these envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions implemented on the ground [e.g., @badgley_systematic_2022].
 One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
 
-The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change (UNFCCC) [@ipcc_2006_2006;@ipcc_2019_2019].
+To this end, the International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change (UNFCCC) [@ipcc_2006_2006;@ipcc_2019_2019].
 Under this guidance, greenhouse gas emissions to, or withdrawals from, the atmosphere are quantified on an annual basis for all managed land, which includes most of the world's forest land [@ogle_delineating_2018; @ellis_anthropogenic_2010].
@@
 A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
-Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
+Tier 1 values have improved over the years as more data have become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement.
 For example, following the 2019 release of the latest IPCC guidelines, it was revealed that IPCC's Tier 1 default failed to capture eight-fold variation of C accumulation in regrowth forests within ecozones [@cook-patton_mapping_2020] and that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
+High variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] means that it is crucial for practitioners to have access to locally-specific information, when available.
 This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning.
-Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.
 
@@
 
-The goal of this publication is to facilitate submission of forest C estimates from scientific studies to EFDB.
+Our goal is to facilitate submission of forest C estimates from scientific studies to EFDB.
 We document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023], including data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the Global Soil Respiration Database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
@@
 (3) submitting data to EFDB.
-These steps are detailed below.
 
@@
 For example, in cases where original studies did not present 95% confidence intervals (required by IPCC when available) but did present standard error or n and standard deviation, we calculated the 95% confidence intervals.<!--and populated the EFDB field with this information (noting the calculation in the EFDB field *Comments from Data Provider*)-->
-There were two cases in which more complex mapping was required: (1) mapping of C cycle variables (detailed below in **Carbon cycle variables**) and (2) land classification (detailed below in **Land classification**).
+There were two cases in which more complex mapping was required: (1) mapping of C cycle variables and (2) land classification.
 
@@
 Information on minimum diameters censused and other important sources of methodological variation were mapped into EFDB (`r table_ForCfieldmapping`).
-Details on the mapping of ForC variables to EFDB are documented in the file ForC_variables_mapping.csv in the GitHub repository associated with this publication IPCC-EFDB-integration repository in ForC-db organization ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
+Details on the mapping of ForC variables to EFDB are documented in the GitHub repository associated with this publication ([https://github.com/forc-db/IPCC-EFDB-integration](https://github.com/forc-db/IPCC-EFDB-integration)).
 
@@
 
-Classification into sub-categories was dependent upon stand age and site history.
+Classification into EFDB sub-categories was dependent upon stand age and site history.
 For Forest Land ≥ 20 years old or of unknown (relatively mature) age, or Forest Land < 20 years old that was forest prior to a stand-clearing disturbance, the past land-use category was Forest Land, making the sub-category "Forest Land Remaining Forest land".
@@
 We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which met the criteria for submission, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (`r table_variables`).
-The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records we reviewed and those submitted to EFDB was attributable to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
+The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records we reviewed and those submitted to EFDB was due to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
 The discrepancy between the number of records sent and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
@@
 `r fig_histograms`b).
-In terms of dominant vegetation, by far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
+By far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
 The largest number of records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
@@
 So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been submitted to EFDB.
-Although this process requires manual review of records, the submission of new records to EFDB is hugely facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
+Although this process requires manual review of records, the submission of new records to EFDB is greatly facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
 Future efforts to review studies for submission should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
-`r fig_map`-`r fig_histograms`, `r table_variables`), to records from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), to the variables most needed by EFDB users, and to the more contemporary records.
+`r fig_map`-`r fig_histograms`, `r table_variables`).
+Other categories of records to prioritize include those from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), the variables most needed by EFDB users, and the most up-to-date records.
 
-In addition to the large potential to expand EFDB using records already in ForC, there are innumerous published EFDB-relevant forest C data that are not currently included in ForC, with more being published on a nearly daily basis.
-Coverage of particular variables or regions could be vastly improved through systematic review of the literature.
-Indeed, recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
+In addition to the large potential to expand EFDB using records already in ForC, there are extensive EFDB-relevant forest C data that are not currently included in ForC, with more being published on a nearly daily basis.
+Coverage of particular variables or regions could be vastly improved through systematic review of the literature, although this requires focused and extensive manual effort.
+Recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
 Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
@@
 While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [`r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
-Other pools and variables remain very poorly quantified and highly uncertain for many parts of the world [`r table_variables`,@tifafi_large_2018; @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
+Other pools and variables remain poorly quantified and highly uncertain for many parts of the world [`r table_variables`,@tifafi_large_2018; @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
 Furthermore, data distribution is very uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
 For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
+More generally, belowground C measurements remain sparse globally.
 Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labrière_forest_2023].
@@
 
-We recommend that, in order to make research valuable to estimate C stock changes according to methods provided in the IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
-It is particularly noteworthy that simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
-Some examples are as follows:
+We recommend that, in order to make research most useful for estimating C stock changes following IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
+Importantly, simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
+For example:
 (1) presenting data only in a figure makes them ineligible for inclusion in EFDB, whereas presentation in a table or supplementary data file allows inclusion while supporting FAIR goals [@stall_make_2019];

### 2024-02-29 doc/manuscript/ForC-EFDB manuscript.Rmd
incorporate Susan's edits

@@
 
-climate change, database, forest carbon, greenhouse gas inventory, International Panel on Climate Change (IPCC), natural climate solutions
+climate change, database, forest carbon, greenhouse gas inventory, International Panel on Climate Change (IPCC), natural climate solutions, nature based climate solutions
 
@@
 Human interactions with forests have shaped Earth's climate for millenia and will continue to do so as we target net-zero emission goals.
-Accurately characterizing these climate impacts requires accurate forest carbon data available for national greenhouse gas inventories conducted under the Paris Agreement. 
+Accurately characterizing these climate impacts requires making accurate forest carbon data available for forest monitoring and planning. 
 Here we develop a semi-automated process for submitting forest carbon measurements from the largest relevant scientific database to the International Panel on Climate Change's Emission Factor Database, which currently has sparse forest carbon data.
@@
 
-Humans have been influencing Earth's climate via ecologically transformative land use practices for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016].
+Humans have been influencing Earth's climate via ecologically transformative impacts on ecosystems for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016].
 In recent decades, as both anthropogenic land transformation and climate change have accelerated, this relationship has come into increasing focus [@ipcc_climate_2019].
@@
 Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
+<!--SCP: You might get asked to say relative to what...the answer could be the social cost of carbon or more technological solutions. But it's probably fine to leave as is.-->
 However, these envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions implemented on the ground [e.g., @badgley_systematic_2022].
-One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
+Realistic planning and reporting is critically needed to ensure that forest-based climate change mitigation initiatives are effective, and this requires solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
 
@@
 A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
-Tier 1 values have improved over the years as more data have become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement.
+Tier 1 values have improved over the years as more data and methods have become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement.
 For example, following the 2019 release of the latest IPCC guidelines, it was revealed that IPCC's Tier 1 default failed to capture eight-fold variation of C accumulation in regrowth forests within ecozones [@cook-patton_mapping_2020] and that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
@@
 We recommend that, in order to make research most useful for estimating C stock changes following IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
-Importantly, simple decisions on the presentation of results will determine whether the records meet the criteria for inclusion in EFDB.
+Importantly, simple decisions on the presentation of results will determine whether the records meet the current criteria for inclusion in EFDB.
 For example:
@@
 For those compiling published records (e.g., for meta-analyses), the data set can have added value if all information required by EFDB is extracted from original publications.
+<!--SCP - optional addition : Realistic  planning and reporting is critically needed to ensure that forest-based climate change mitigation initiatives are effective, and this requires solid scientific data -->
 This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
@@
 As human society strives to achieve net-zero greenhouse gas emissions, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
-In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under teh Paris Agreement.
-This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may no longer be relevant.
+In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under the Paris Agreement.
+This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may be increasingly less accurate.
 This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.

### 2024-02-29 doc/manuscript/SI_tables.Rmd
finalize SI tables

@@
 output: 
-    pdf_document
+    #pdf_document
+    bookdown::word_document2: 
+      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
 bibliography: references.bib
@@
 
-**`r table_ForCchanges`. Updates to ForC field implemented between releases of v3.0 and v4.0.** 
+**`r table_ForCchanges`. Updates to ForC fields implemented between releases of v3.0 and v4.0.** 
 ```{r table_ForCchanges, eval = TRUE, echo=FALSE, warning=FALSE, message = FALSE}
@@
 record_of_changes <- record_of_changes[c(1:5)] #include all fields but date of change
-#record_of_changes<- record_of_changes[which(!grep("housekeeping",record_of_changes$Motivation)) , c(1:5)]  #trying to exclude rows for which Motivation = "housekeeping", but this is not working
+record_of_changes <- record_of_changes[!grepl("housekeeping", record_of_changes$Motivation, ignore.case = TRUE), c(1:5)]
 record_of_changes[is.na(record_of_changes)]<- "" #replace NAs with blanks
 
-library(knitr)
-library(kableExtra)
-knitr::kable(record_of_changes, format = "latex",
-             longtable = TRUE) %>%
-  kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header")) %>%
-  kable_styling(font_size = 8) %>%
-  row_spec(0, bold = TRUE)
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 8) # set font
+ft <- flextable(record_of_changes)
+ft <- autofit(ft)
+ft <- width(ft, 1, width = 1)
+ft <- width(ft, 2:5, width = 1.5)
+ft
+
 ```
@@
 ```{r table_ForCfieldmapping, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
+library(readr)
+library(stringr)
+library(flextable)
 
+# Read the data
 ForC_variables_included <- read.csv("figures_tables/ForC_fields_included.csv")
+
+# Fix column names
 names(ForC_variables_included) <- stringr::str_replace_all(names(ForC_variables_included), "\\.", " ")
@@
 names(ForC_variables_included) <- column_names# Set column names of the matrix
+
+# Subset to the first 5 columns
 ForC_variables_included <- ForC_variables_included[c(1:5)]
 
-knitr::kable(ForC_variables_included, format = "latex",
-             longtable = TRUE) %>%
-  row_spec(0, bold = TRUE) %>%
-  kable_styling(font_size = 8) %>%
-kableExtra::kable_styling(full_width = TRUE, latex_options = c("repeat_header", "hold position"))
+# Create flextable object
+ft <- flextable(ForC_variables_included)
+ft <- autofit(ft)
+ft <- width(ft, 1, width = 1)
+ft <- width(ft, 2:5, width = 1.5)
+
+# Apply styling
+#ft <- ft %>%
+#  bold(part = "header") %>%
+#  fontsize(size = 8)
+
+ft
 

### 2024-02-29 doc/manuscript/z_archive/ESSD/ForC-EFDB-manuscript-for-ESSD.tex
archive files

--- 2023-06-09 doc/manuscript/ForC-EFDB-manuscript-for-ESSD.tex
@@
 some updates to the ForC database structure, leading to the release of a
-new version of ForC (v4.0; described here). As of June 09, 2023, ForC
+new version of ForC (v4.0; described here). As of July 28, 2023, ForC
 contained \textasciitilde19316 independent records relevant to EFDB,
@@
 published estimates of forest C stocks, increments, and annual fluxes
-\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021}. ForC
-includes data ingested from individual publications and relevant
+\citep{anderson-teixeira_forc_2018, anderson-teixeira_carbon_2021, anderson-teixeira_forcdb_2023}.
+ForC includes data ingested from individual publications and relevant
 databases, including the Global Reforestation Opportunity Assessment
@@
 \citep[SRDB-V5,][]{bond-lamberty_global_2010, jian_restructured_2021}.
-As of June 09, 2023, ForC contained 39848 records from 10589 plots in
+As of July 28, 2023, ForC contained 39848 records from 10589 plots in
 1535 distinct geographical areas, along with records of stand age and
@@
 (\emph{woody.mortality}). Although most of these variables lacked
-records in ForC as of June 09, 2023, their addition gave the structure
+records in ForC as of July 28, 2023, their addition gave the structure
 such that records can be populated over time. Finally, to provide better
@@
 then posted in the database -- a process that lags behind submission of
-records and had not yet been completed for all records sent as of June
-09, 2023.
+records and had not yet been completed for all records sent as of July
+28, 2023.
 
@@
 
-As of June 09, 2023, ForC (v4.0) contained 32686 independent records
+As of July 28, 2023, ForC (v4.0) contained 32686 independent records
 (39848 total), 19316 of which were for the 42 variables relevant to EFDB
@@
 
-As of June 09, 2023, we had reviewed or added 2292 EFDB-relevant
+As of July 28, 2023, we had reviewed or added 2292 EFDB-relevant
 records, 1438 records of which were submitted to EFDB, and 376 of which
@@
 removals from forest land consistent with good practice in the IPCC
-guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of June 09, 2023,
+guidelines \citep{ipcc_2006_2006, ipcc_2019_2019}. As of July 28, 2023,
 we have submitted 1438 records to EFDB. Although this represents just
@@
 \codedataavailability{All code and data are openly available. The ForC
-database and associated code is available via the ForC repository within
-the ForC-db organization on GitHub (https://github.com/forc-db/ForC),
-and the version used here (ForC v4.0) archived in Zenodo (DOI: TBD). The
-data and code associated with data submission to EFDB and preparation of
-this manuscript are available via the the IPCC-EFDB-integration
-repository within the ForC-db organization on GitHub
+database and associated code are available via the ForC repository
+within the ForC-db organization on GitHub
+(https://github.com/forc-db/ForC), and the version used here (ForC v4.0)
+is archived in Zenodo \citep[DOI:
+10.5281/zenodo.8020861]{anderson-teixeira_forcdb_2023}. The data and
+code associated with data submission to EFDB and preparation of this
+manuscript are available via the the IPCC-EFDB-integration repository
+within the ForC-db organization on GitHub
 (https://github.com/forc-db/IPCC-EFDB-integration) and archived in

### 2024-03-01 doc/manuscript/SI_IPCCdefs.Rmd
finalize SI

@@
 Forest Land is defined as at least 10-30% crown cover of trees with potential to reach a minimum height of 2-5 m *in situ*, and shorter-stature natural vegetation would be classified as Grassland [@ipcc_good_2003].
-Definitions of forest are allowed to vary by country, but must be applied consistently.
+Definitions of forest are allowed to vary by country but must be applied consistently.
 Forest Land includes land where vegetation temporarily falls below the threshold values for forest (e.g., due to disturbance), but is expected to exceed those thresholds in the future [@ipcc_good_2003].
@@
 
-For each stratum of subdivision within a land-use category (see **Land classification** section below), annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (see **Carbon pools** section below), plus any harvested wood products.
+For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools, plus any harvested wood products.
 Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, net annual increments, and fluxes in the IPCC-defined pools.
@@
 Dead organic matter includes all non-living biomass larger than the litter size threshold. 
-Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land (see **Land Classification** section below), but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
+Its inclusion in inventories is not required under Tier 1 methodology for Forest Land remaining Forest Land but is required for land that has transitioned to or from forest within the past 20 years [@ipcc_2006_2006].
 

### 2024-03-01 doc/manuscript/ForC-EFDB manuscript.Rmd
final check

@@
 
-Teagan Rogers^1^
+Teagan Rogers^3^
 
-Rebecca Banbury Morgan^1,3^
+Rebecca Banbury Morgan^1,4^
 
-Ben Bond-Lamberty^4^
+Ben Bond-Lamberty^5^
 
-Susan Cook-Patton^5^
+Susan Cook-Patton^6^
 
@@
 
-3. School of Geography, University of Leeds, Leeds, UK
+3. Smithsonian-Mason School of Conservation, Front Royal, VA, United States
 
-4. Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
+4. School of Geography, University of Leeds, Leeds, United Kingdom
 
-5. The Nature Conservancy; Arlington VA 22203, USA
+5. Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
+
+6. The Nature Conservancy; Arlington VA 22203, United States
 
@@
 sampling_details <- read.csv(paste0(path_to_ForC,"/ForC/database_management_records/record_of_changes.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-n_fields_modified =  length(sampling_details[,"Column"]) + 4
+n_fields_modified =  length(sampling_details[,"Field"]) + 4
 
@@
 <!-- (1) the research conducted or aims, including the rationale-->
-- Humans have been influencing Earth's climate via transformative impacts on forests for >12,000 years, and forests are now recognized as critical to climate change mitigation under the Paris Agreement.
-However, the efficacy of climate change mitigation planning and reporting depends on having quality data on forest carbon (C) stocks and changes.
+- Humans have been influencing Earth's climate via transformative impacts on forests for millenia, and forests are now recognized as critical to climate change mitigation under the Paris Agreement.
+The efficacy of climate change mitigation planning and reporting depends on quality data on forest carbon (C) stocks and changes.
 The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data, but needs comprehensive and well-documented data to be so.
@@
 - As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_posted_EFDB` of which had been submitted and posted to EFDB.
-These represent 19% of the total EFDB records for forest land.
-Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests).
+These represented `r round(n_records_posted_EFDB/5644*100,0)`% of the total EFDB records for forest land.
+Records were unevenly distributed across variables<!-- (skewed towards aboveground biomass stocks)--> and geographic regions<!-- (skewed towards temperate forests)-->.
 `r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_reviewed_or_added,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked required information.
@@
 BBL: might cite Ruehr et al. 2023 http://dx.doi.org/10.1038/s43017-023-00456-3 here-->
-ofsetting an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
+offsetting an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
 Table 2)--> - `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
@@
 The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [`r appendix_IPCC_definitions`, @ipcc_agriculture_2006; @ipcc_2019_2019].
-A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
+A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone<!-- [@fao_global_2012]--> and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
 Tier 1 values have improved over the years as more data and methods have become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement.
@@
 
-We identified ForC variables that were relevant to the IPCC methodology and to EFDB (`r appendix_IPCC_definitions`).
+We identified ForC variables that were relevant to the IPCC methodology and EFDB (`r appendix_IPCC_definitions`).
 These included  organic matter or C stocks, net annual increments, influxes (a.k.a.
@@
 New data collection and analyses are needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [`r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
+While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [`r table_variables`,@nisar_nasaisro_2018; @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
 Other pools and variables remain poorly quantified and highly uncertain for many parts of the world [`r table_variables`,@tifafi_large_2018; @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].

### 2024-03-01 doc/manuscript/ForC-EFDB manuscript.Rmd
small edits

@@
 
-Rebecca Banbury Morgan^1,4^
+Rebecca Banbury Morgan^1,4,5^
 
-Ben Bond-Lamberty^5^
+Ben Bond-Lamberty^6^
 
-Susan Cook-Patton^6^
+Susan Cook-Patton^7^
 
@@
 
-5. Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
+5. School of Biological Sciences, University of Bristol, Bristol, United Kingdom
 
-6. The Nature Conservancy; Arlington VA 22203, United States
+6. Joint Global Change Research Institute, Pacific Northwest National Laboratory, College Park, MD, United States
+
+7. The Nature Conservancy; Arlington VA 22203, United States
 
@@
 BBL: might cite Ruehr et al. 2023 http://dx.doi.org/10.1038/s43017-023-00456-3 here-->
-offsetting an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
+that offsets an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
 Table 2)--> - `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
@@
 Accordingly, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015; @grassi_key_2017].<!--As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].-->
-Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
+Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021]. 
 <!--SCP: You might get asked to say relative to what...the answer could be the social cost of carbon or more technological solutions. But it's probably fine to leave as is.-->
-However, these envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions implemented on the ground [e.g., @badgley_systematic_2022].
-Realistic planning and reporting is critically needed to ensure that forest-based climate change mitigation initiatives are effective, and this requires solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].
+Conservation and reforestation have the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
+However, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions implemented on the ground [e.g., @badgley_systematic_2022].
+Realistic planning and reporting is critically needed to ensure that forest-based climate change mitigation initiatives are effective, and this requires solid scientific data and accounting frameworks [@anderson-teixeira_effective_2022; @deng_comparing_2021].
 
@@
 To support export of data to EFDB, and to improve the overall quality of the ForC database, we added or modified `r n_fields_modified` fields (`r table_ForCchanges`), defined `r n_new_variables` new variables, implemented enhanced quality control, manually reviewed >`r n_records_checked` records to obtain additional required information, and added `r n_records_added` new records.
-Having implemented these changes to ForC v3.0 [@anderson-teixeira_carbon_2021], which are described in detail in `r appendix_ForC_updates`, we released a new major version: ForC v4.0.
+Having implemented these changes, which are described in detail in `r appendix_ForC_updates`, to ForC v3.0 [@anderson-teixeira_carbon_2021], we released a new major version: ForC v4.0.
 
@@
 
-# Supplementary Information
-`r appendix_IPCC_definitions`. Primer on forest land classification and carbon pools under IPCC guidelines
-
-`r appendix_ForC_updates`. Updates to ForC (ForC v4.0)
+# Supplementary Information {-}
 
@@
 
+`r appendix_IPCC_definitions`. Primer on forest land classification and carbon pools under IPCC guidelines
+
+`r appendix_ForC_updates`. Updates to ForC (ForC v4.0)
+
 \clearpage

### 2024-06-26 doc/manuscript/response-to-reviews.Rmd
add decision letter, response to reviews document

---
title: ''
output: word_document
bibliography: references.bib #This can be generated by exporting BibTex file from Zotero
csl: new-phytologist.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor:

Below please find our responses to reviews for our manuscript ' Informing forest carbon inventories under the Paris Agreement using ground-based forest monitoring data' (PPP-MT-2024-00903). We have addressed all points, as detailed below.

Sincerely,

Kristina Anderson-Teixeira (on behalf of all coauthors)

\newpage

# Response to Reviews

## Referee: 1

The manuscript presents a process developed for preparing data for submission to the Emission Factor Database (EFDB) and methods to assess the data in the Global Forest Carbon Database (ForC version 4.0) relevant to EFDB. The variables relevant to IPCC and EFDB were identified and listed.
Results describe the databases and analyse some basic elements of the records they contain. Data contribution to ForC is mapped.

The aims of the study are clearly identified and very relevant to the community, and the idea itself of improving scientist’s contribution to international databases appreciable. Thus, there are good reasons to publish.

**Thank you.**

The results section is however disconcerting, partly because it is not clearly structured, and partly because it is quite unclear what point authors wanted to make out of the few graphs presented.
The first paragraph is about records with deficiencies, a fraction being “deemed not applicable to EFDB…”. This is not informative at all in absence of a series of well-identified examples. Quite verbose but not that useful.
Another example, the map (Fig 2) is fine as an illustration or proof of the global participation to the efforts of reporting, but it doesn’t help identify the gaps in the current reporting. Some vegetation types are well represented, others have a serious deficit. Identifying and quantifying this, for instance in terms of report per area, would have been a more eloquent and useful than a map. The imbalance in the data goes for the below/aboveground compartments too.
Fig. 3 isn’t conclusive either because ecozones have different sizes, so the number of reports on itself is not a sufficient to evaluate the coverage. The representativity of the data within an ecozone depends on the variability between records: if they all cover the same age classes, again, the representativity is bad.

Consequently, the “Database needs” as discussed L297 are not convincingly demonstrated by the results which are limited to just frequencies. Overall the discussion seems somewhat verbose.

Some parts of the writing are a bit frustrating, as they suggest that authors have some relevant information that they do not share:

L156 “We identified ForC variables that were relevant to the IPCC…”
Great, but how? The reasoning behind the identification is certainly of interest to scientists or practitioners.

L226 “To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB’s bulk import form.”
What are those scripts?
Why not share these scripts?

L300. So far, only ~7% of the EFDB-relevant data in ForC have been submitted to EFDB.
How do you get that number? What are those 7% and what consequences does this have?
Providing concrete examples of mistakes to avoid would have been very appreciable, and made the text sound less theoretical.

The discussion section “Data reporting needs” should be named “Recommendations of good practices for reporting C stocks and stock changes”, or something more in line with the content itself.

The summary lists up a point on human’s influence on earth’s climate, which is also heavily elaborated in the introduction. The emphasis on human’s influence on forests and climate are disproportionated and distracting (i.e. not appropriate here), since the manuscript is not dedicated to studying it as it solely focuses on scientists reporting C stocks evaluations. A simple statement that reporting C stocks is an obligation but still poses difficulties would be more honest and efficient. Removing the first paragraph of the introduction, which is totally disconnected, would allow to shorten a bit the overall too long introduction.

L239: “The 37% attenuation between records we reviewed and those submitted…”
Why “attenuation”?, it seems to be a mere discrepancy, or deficit.

**We have changed this word to "discrepancy".**

In conclusion, although the aims and principle seem simple, the results are both vague and expeditive, resulting in a rather unconvincing article. Some efforts to make results more relevant, and some more details on key aspects would help.

# Referee: 2

This manuscript is a timely and relevant paper about developing a standardized protocol to input forest carbon data into the IPCC emissions database. The authors identified two main problems: the lack of authors submitting their published forest carbon data into the IPCC Emission Factor Database, and some papers not fully reporting all information (pools, fluxes, methodology) necessary to submit in the EFDB). This team has developed a robust and standardized methodology for first aggregating forest carbon data into their already developed (but modified for this new methodology) ForC database, followed by inputting ForC data into the IPCC database. The lack of available, standardized forest carbon data is a real problem as the need to monitor ecosystem carbon stocks and fluxes becomes more crucial in the face of climate change– the methodology developed in this paper has high societal impact to better fill in data gaps in a globally important IPCC database.

**Thank you.**

This paper is high quality and well written with reliable findings. The authors used appropriate techniques for both the ForC database and the methodology for semi-automated data submission into the EFDB. Figure 1 is easy to follow and shows how detailed and robust the ForC to EFDB framework is.

**Thank you.**

I appreciate the authors recommendations in the discussion section (they are all relevant, warranted, and important) but I do think some of these recommendations could be more strongly worded. The authors have described large challenges with populating the EFDB, and have developed a robust solution (and identified other, easier solutions). I would like to see the authors clear and strong recommendations in this section. Some of the suggestions would require substantial increases in funding and infrastructure (line 328-333), which are huge challenges. I think that section has been well discussed in the discussion. However, the next paragraph discusses the lack of variables available because they have either not been calculated or just not presented in papers (although they may have been available to the authors) (lines 334-346). I think recommendations in this section could be stronger and more specific – it is very easy for authors to provide data they have already collected/calculated, or to complete relatively simple allometric calculations. Are there ways this could be required or incentivized (this question may be out of scope for this paper)? A more specific recommendation in the next paragraph (lines 348-362) would also be warranted.

### 2024-07-03 doc/manuscript/response-to-reviews.Rmd
worked on response to reviews

@@
 knitr::opts_chunk$set(echo = TRUE)
+
+#Tables
+table_pools =1
+table_recommendations = 2
+
+#Figures
+fig_variable_mapping=1
+fig_map=2
+fig_climate=3
+fig_histograms=4
+
+#Appendices
+appendix_ForC_updates = "Methods S1"
+appendix_IPCC_definitions = "Notes S1"
+table_ForCchanges = "Table S1"
+table_ForCfieldmapping = "Table S2"
+table_variables ="Table S3"
 ```
@@
 The results section is however disconcerting, partly because it is not clearly structured, and partly because it is quite unclear what point authors wanted to make out of the few graphs presented.
+
+*(NEEDS RESPONSE)*
+
 The first paragraph is about records with deficiencies, a fraction being “deemed not applicable to EFDB…”.
@@
 Quite verbose but not that useful.
+
+*(NEEDS RESPONSE. We did not receive this information directly. We will have to figure out which records are missing)*
+
 Another example, the map (Fig 2) is fine as an illustration or proof of the global participation to the efforts of reporting, but it doesn’t help identify the gaps in the current reporting.
@@
 The imbalance in the data goes for the below/aboveground compartments too.
+
+**We have made some of the suggested revisions (detailed below) to improve the assessment of data representativeness and consider this to be an important improvement.
+At the same time, we also wish to emphasize that the goal of this paper was to document the process and provide an overview of the types of data submitted, as opposed to submitting data that adequately represents all forest types and documenting their representativeness.
+The latter -- to the extent that it would even be possible -- would be a massive project far beyond the scope of this paper.
+Therefore, while improving the assessment of data representativeness, we believe it would be overkill to analyze all aspects of representativeness.**
+
+**We have added a graph to the map (Fig.
+`r fig_map`) showing the proportional breakdown by FAO ecozone of (1) area with tree cover, (2) sites with records submitted to EFDB, (3) records submitted to EFDB.
+We have also added a statement on relative representation by biome: "Relative to forested area, submitted data under-represented the tropics and over-represented temperate and boreal biomes (Fig.
+`r fig_map`)."**
+
+**To assess the data distribution within in the global climate space, we have added a Whittaker-style diagram showing ForC sites and those with submitted data within the MAT-MAP climate space (new Fig.
+`r fig_climate`).
+We have also added a statement on climate representativity: "Submitted data were concentrated in climates that are relatively common over Earth's land area (Fig.
+`r fig_climate`)." **
+
+**Regarding imbalance among variable types, `r table_variables` provides a complete record of the numbers of records in ForC and submitted to EFDB.
+We feel that further analyzing by categories such as ecozone or age would be overkill, make the manuscript too long, and of little interest to most readers.**
+
 Fig.
-3 isn’t conclusive either because ecozones have different sizes, so the number of reports on itself is not a sufficient to evaluate the coverage.
+3 *(now 4)* isn’t conclusive either because ecozones have different sizes, so the number of reports on itself is not a sufficient to evaluate the coverage.
 The representativity of the data within an ecozone depends on the variability between records: if they all cover the same age classes, again, the representativity is bad.
+
+*https://github.com/forc-db/IPCC-EFDB-integration/issues/63*
+
+*Again, we feel that deeper analysis of the representativity would be overkill....*
 
@@
 
+*(NEEDS RESPONSE)*
+
 Some parts of the writing are a bit frustrating, as they suggest that authors have some relevant information that they do not share:
+
+**We'd like to emphasize that all data, code, planning documents, and even our internal discussions are documented in the associated public GitHub repository (https://github.com/forc-db/IPCC-EFDB-integration), as explained in our data availability statement.
+We have done our best to present relevant information while keeping the manuscript concise and as interesting as possible.
+We appreciate the reviewer having pointed out places where more information is desirable, and have modified the text to include it.**
 
@@
 Great, but how? The reasoning behind the identification is certainly of interest to scientists or practitioners.
+
+**We have modified this statement to read, "Working in consultation with the EFDB Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we identified ForC variables that were relevant to the IPCC methodology and EFDB (Fig.
+`r fig_variable_mapping`, `r appendix_IPCC_definitions`)."**
 
@@
 
+**These scripts are publicly available in the associated GitHub repository (https://github.com/forc-db/IPCC-EFDB-integration) and also archived in Zenodo.
+This is explained in our data availability statement, but that fact is easy to miss.
+To ensure that this is clear to readers, we have modified this statement to read: "To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see Data Availability Statement).**
+
 L300. So far, only ~7% of the EFDB-relevant data in ForC have been submitted to EFDB.
@@
 
+*(Provide a bit more detail, addressing questions.)*
+
 The discussion section “Data reporting needs” should be named “Recommendations of good practices for reporting C stocks and stock changes”, or something more in line with the content itself.
+
+*(NEEDS RESPONSE)*
 
@@
 Removing the first paragraph of the introduction, which is totally disconnected, would allow to shorten a bit the overall too long introduction.
+
+*(NEEDS RESPONSE)*
 
@@
 Some efforts to make results more relevant, and some more details on key aspects would help.
+
+*(NEEDS RESPONSE)*
 
@@
 A more specific recommendation in the next paragraph (lines 348-362) would also be warranted.
+
+*(NEEDS RESPONSE)*
+
+\clearpage
+# References {-}

### 2024-07-03 doc/manuscript/ForC-EFDB manuscript.Rmd
post-submission fixes
update Teagan's last name
worked on response to reviews

@@
 
-Teagan Rogers^3^
+Teagan Tinuviel^3^ (ORCID: 0000-0002-8828-9640)
 
@@
 fig_map=2
-fig_histograms=3
+fig_climate=3
+fig_histograms=4
 
@@
 # MOVE FIGURE FILES 
-#world map from GitHub
+#world map from ForC GitHub
 do.call(file.remove, list(list.files(paste0(path_to_dir, "/doc/manuscript/figures_tables"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE))) # first remove old file
 flist <- list.files(paste0(path_to_ForC, "/ForC/figures/"), "World_Map_of_sites_with_FAO_and_IPCC_data_sent.png", full.names = TRUE)
+file.copy(flist, paste0(path_to_dir, "/doc/manuscript/figures_tables"))
+
+#climate diagram from ForC GitHub
+do.call(file.remove, list(list.files(paste0(path_to_dir, "/doc/manuscript/figures_tables"), "Climate_of_forC_sites_IPCC.png", full.names = TRUE))) # first remove old file
+flist <- list.files(paste0(path_to_ForC, "/ForC/figures/"), "Climate_of_forC_sites_IPCC.png", full.names = TRUE)
 file.copy(flist, paste0(path_to_dir, "/doc/manuscript/figures_tables"))
@@
 
-Human interactions with forests have shaped Earth's climate for millenia and will continue to do so as we target net-zero emission goals.
+Human interactions with forests have shaped Earth's climate for millennia and will continue to do so as we target net-zero emission goals.
 Accurately characterizing these climate impacts requires making accurate forest carbon data available for forest monitoring and planning. 
@@
 
-With input from the IPCC's Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we determined how EFDB fields should be populated using ForC fields (summarized in `r table_ForCfieldmapping`).
+With input from the EFDB Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we determined how EFDB fields should be populated using ForC fields (summarized in `r table_ForCfieldmapping`).
 Fields in EFDB included several fields describing how the record fits within IPCC's framework (source/sink category, greenhouse gas type, C pool, relevant equations), several describing the C estimate itself (variable, value, units, 95% confidence limits), a few composite fields describing biotic and abiotic conditions (e.g., vegetation type, minimum diameter included, site location, climate, edaphic properties, notable disturbances, stand age, plot information), and a few describing the source (e.g., type, citation, data provider).
@@
 
-We identified ForC variables that were relevant to the IPCC methodology and EFDB (`r appendix_IPCC_definitions`).
+Working in consultation with the EFDB Technical Support Unit and referencing IPCC guidance [@ipcc_good_2003; @ipcc_2006_2006; @ipcc_2019_2019], we identified ForC variables that were relevant to the IPCC methodology and EFDB (Fig.
+`r fig_variable_mapping`, `r appendix_IPCC_definitions`).
 These included  organic matter or C stocks, net annual increments, influxes (a.k.a.
@@
 
-To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form.
+To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see **Data Availability Statement**).
 Criteria for data submission were that
@@
 We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which met the criteria for submission, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (`r table_variables`).
-The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% attenuation between records we reviewed and those submitted to EFDB was due to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
+The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% discrepancy between records we reviewed and those submitted to EFDB was due to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
 The discrepancy between the number of records sent and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
@@
 ForC v4.0 records and the subset submitted to EFDB were distributed across all forested continents, biomes, and forest types, albeit very unevenly (Figs. `r fig_map`-`r fig_histograms`).
+Relative to forested area, submitted data under-represented the tropics and over-represented temperate and boreal biomes (Fig. `r fig_map`).
+Submitted data were concentrated in climates that are relatively common over Earth's land area (Fig. `r fig_climate`).
 Among the records submitted to EFDB, the largest number came from North America, followed by Asia, South America, and Africa (Fig.
@@
 ![**Figure `r fig_map`.
-Map of sites in ForC shaded by number of independent records relevant to (circles) and submitted to (triangles) EFDB.** Symbols are colored according to the number of records at each site.
-Underlying map shows FAO ecozones, which are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
+Map of sites in ForC shaded by number of independent records relevant (circles) and submitted (triangles) to EFDB.**  Underlying map shows FAO ecozones, and symbols are colored according to the number of records at each site.
+Inset stacked bar chart shows proportional representation by FAO ecozone for tree cover area (from SYNMAP,@jung_exploiting_2006, relevant ForC records, and submitted records.
+FAO ecozones are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
+
+![**Figure `r fig_climate`.
+Distribution of ForC records relevant (dots) and submitted (triangles) to EFDB within the global climate space of mean annual temperature (MAT) and mean annual precipitation (MAP).** Climate data are from *WorldClim 2 (0.5 resolution, Fick & Hijmans, 2017*.
+Background colors indicate the global land area with each MAT-MAP combination.
+GitHub issue 61:https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.15353](figures_tables/Climate_of_forC_sites_IPCC.png)
 

### 2024-08-01 doc/manuscript/response-to-reviews.Rmd
add figures and text responding to reviewer 1 comments
#63 , #62 , #61

@@
 **We have added a graph to the map (Fig.
-`r fig_map`) showing the proportional breakdown by FAO ecozone of (1) area with tree cover, (2) sites with records submitted to EFDB, (3) records submitted to EFDB.
-We have also added a statement on relative representation by biome: "Relative to forested area, submitted data under-represented the tropics and over-represented temperate and boreal biomes (Fig.
+`r fig_map`) showing the proportional breakdown by FAO ecozone of (1) area with tree cover, (2) sites and records submitted to EFDB, (3) sites and records submitted to EFDB.
+We have also added a statement on relative representation by biome: "Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, mirroring the composition of ForC (Fig.
 `r fig_map`)."**
@@
 `r fig_climate`).
-We have also added a statement on climate representativity: "Submitted data were concentrated in climates that are relatively common over Earth's land area (Fig.
+We have also added a statement on climate representativity: "Submitted data covered all forested climates with substantial land area except the colder boreal regions (boreal tundra woodland; Figs.
+`r fig_map`, `r fig_climate`) and were concentrated in climates that are relatively common over Earth's land area (Fig.
 `r fig_climate`)." **
@@
 
-*https://github.com/forc-db/IPCC-EFDB-integration/issues/63*
+**To evaluate the coverage, we have modified this figure to include bars representing the relative area (% of global) of each category and added some text summarizing the results/ citing this figure (3rd paragraph in Results).**
 
-*Again, we feel that deeper analysis of the representativity would be overkill....*
+**Again, we feel that deeper analysis (e.g., age x ecozone) of the representativity would be overkill and of little interest to most readers.**
 

### 2024-08-01 doc/manuscript/ForC-EFDB manuscript.Rmd
add figures and text responding to reviewer 1 comments
#63 , #62 , #61

@@
 
+nocite: |
+  @harris_updated_2014; @jung_exploiting_2006; @besnard_mapping_2021
 ---
@@
 
-ForC v4.0 records and the subset submitted to EFDB were distributed across all forested continents, biomes, and forest types, albeit very unevenly (Figs. `r fig_map`-`r fig_histograms`).
-Relative to forested area, submitted data under-represented the tropics and over-represented temperate and boreal biomes (Fig. `r fig_map`).
-Submitted data were concentrated in climates that are relatively common over Earth's land area (Fig. `r fig_climate`).
-Among the records submitted to EFDB, the largest number came from North America, followed by Asia, South America, and Africa (Fig.
-`r fig_histograms`c), with the most represented FAO ecozones being boreal coniferous forest, temperate continental forest, and temperate mountain systems, followed by tropical rain forests and moist deciduous forests (Fig.
+ForC v4.0 records and the subset submitted to EFDB were distributed across all forested continents, biomes, and forest types, albeit unevenly (Figs. `r fig_map`-`r fig_histograms`).
+Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, mirroring the composition of ForC (Figs. `r fig_map`, `r fig_histograms`b).
+Submitted data covered all forested climates with substantial land area except the colder boreal regions (boreal tundra woodland; Figs.
+`r fig_map`, `r fig_climate`) and were concentrated in climates that are relatively common over Earth's land area (Fig.
+`r fig_climate`).
+Among the records submitted to EFDB, the largest number came from North America, followed by Asia, South America, and Africa, with strongly disproportionate representation of North America and under-representation of South America, Africa, Europe, Australia, and Oceania (Fig.
+`r fig_histograms`c).
+In terms of FAO ecozones, boreal coniferous forest, temperate continental forest, and temperate mountain systems had the most records and were the best represented relative to their global area, whereas other ecozones were poorly represented relative to their global area (Fig.
 `r fig_histograms`b).
-By far the most records came from needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
-The largest number of records came from mature forests (>100 years), followed by young and intermediate-aged stands (Fig. `r fig_histograms`d).
+By far the most records and best representation relative to global area were for needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
+The largest number of records came from mature forests (>100 years), followed by young and intermediate-aged stands, but young (<20 year) stands were best represented relative to their global area (Fig.
+`r fig_histograms`d).
 
@@
 Map of sites in ForC shaded by number of independent records relevant (circles) and submitted (triangles) to EFDB.**  Underlying map shows FAO ecozones, and symbols are colored according to the number of records at each site.
-Inset stacked bar chart shows proportional representation by FAO ecozone for tree cover area (from SYNMAP,@jung_exploiting_2006, relevant ForC records, and submitted records.
-FAO ecozones are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
+Inset stacked bar chart shows proportional representation by FAO ecozone for tree cover area (from SYNMAP, Jung *et al*, 2006), relevant ForC records, and submitted records.
+FAO ecozones (obtained from FAO’s GEOnetwork: http://www.fao.org:80/geonetwork) are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
 
 ![**Figure `r fig_climate`.
-Distribution of ForC records relevant (dots) and submitted (triangles) to EFDB within the global climate space of mean annual temperature (MAT) and mean annual precipitation (MAP).** Climate data are from *WorldClim 2 (0.5 resolution, Fick & Hijmans, 2017*.
-Background colors indicate the global land area with each MAT-MAP combination.
-GitHub issue 61:https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.15353](figures_tables/Climate_of_forC_sites_IPCC.png)
+Distribution of ForC records relevant (dots) and submitted (triangles) to EFDB within the global climate space of mean annual temperature (MAT) and mean annual precipitation (MAP).** Climate data are from the CRU TS3.10 dataset (0.5° resolution, 1990-2014, Harris *et al*, 2014).
+Background colors indicate the global land area with each MAT-MAP combination.](figures_tables/Climate_of_forC_sites_IPCC.png)
 
 ![**Figure `r fig_histograms`.
-Histograms of number of independent records in ForC relevant to (grey) and submitted to (black) EFDB, organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.** For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)
+Histograms of number of independent records in ForC relevant (grey) and submitted (black) to EFDB (both left axis), along with the relative global area of each type (red, right axis), organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.** For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+Global coverage of each was obtained from SYNMAP (Jung *et al*, 2006).
+For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.
+The relative area of forests by age class was obtained from the global forest age database of Besnard *et al.* (2021), updated in 2024 (DOI: xxx).
+](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)
 

### 2024-08-02 doc/manuscript/ForC-EFDB manuscript.Rmd
Update ForC-EFDB manuscript.Rmd
#56

@@
 Thank you to Avni Malhotra for helpful comments on an earlier draft of this manuscript.
-Funding for this study was provided by the Smithsonian (Forest Global Earth Observatory, Smithsonian Working Land and Seascapes); a Bezos Earth Fund grant to the Nature Conservancy, with a sub-grant to NZCBI; and the Institute for Global Environmental Strategies.
+Funding for this study was provided by the Smithsonian (Forest Global Earth Observatory, Smithsonian Working Land and Seascapes) and a Bezos Earth Fund grant to the Nature Conservancy with a sub-grant to NZCBI.
 

### 2024-08-02 doc/manuscript/ForC-EFDB manuscript.Rmd
describe reasons for data rejection; remove litter, delta.litter
#64

@@
 n_ind_records_deadwood <- C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_down"]+C_variables$`n independent records in ForC`[C_variables$`variable` == "deadwood_standing"]
-n_ind_records_OL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "litter"] +C_variables$`n independent records in ForC`[C_variables$`variable` == "O.horizon"]
+n_ind_records_OL <- C_variables$`n independent records in ForC`[C_variables$`variable` == "O.horizon"]
 n_ind_records_TotEcosystemC <- C_variables$`n independent records in ForC`[C_variables$`variable` == "total.ecosystem_2"] 
@@
 
-n_records_reviewed_or_added= C_variables[length(C_variables[,"variable"]), "n reviewed"]
+n_records_rejected_variables = 23+5  # 23 litter, 5 NEE
+
+n_records_reviewed_or_added= C_variables[length(C_variables[,"variable"]), "n reviewed"] + n_records_rejected_variables
 n_records_added = max(ForC_measurements$measurement.ID) -48956 # Madison's entries start with measurement.ID = 48956
-n_records_checked = n_records_reviewed_or_added - n_records_added # this will be hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
+n_records_checked = n_records_reviewed_or_added - n_records_added  # this is hard to get at, but we can at least estimate a minimum (= previously existing records in records sent to EFDB)
 
-n_records_sent_to_EFDB = C_variables[length(C_variables[,"variable"]), "n submitted to EFDB"] # n records sent to EFDB
+n_records_sent_to_EFDB =  C_variables[length(C_variables[,"variable"]), "n submitted to EFDB"] + n_records_rejected_variables  # n records sent to EFDB
 n_relevant_variables_no_records_sent =  sum(C_variables$`n submitted to EFDB` == 0,  na.rm = TRUE)
@@
 n_records_sent_deadwood_components<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_down"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "deadwood_standing"]
-n_records_sent_litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "litter"] +C_variables$`n submitted to EFDB`[C_variables$`variable` == "O.horizon"]
+n_records_sent_litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "O.horizon"]
 n_records_sent_soil<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "SOM / SOC"] 
@@
 n_records_sent_delta.agb<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.agb"]
-n_records_sent_delta.litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.O.horizon"]+C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.litter"]
+n_records_sent_delta.litter<- C_variables$`n submitted to EFDB`[C_variables$`variable` == "delta.O.horizon"]
 n_records_sent_increments <- n_records_sent_delta.agb+n_records_sent_delta.litter
@@
 Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
-Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter horizon (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
+Additional caveats are as follows: (a,b) branch fall and mortality of stems below the minimum stem diameter censused, which are necessary for a full accounting of dead organic matter production but typically assumed negligible for calculations of biomass change, are excluded by common measurement practice (a) or ForC variable definition (b); (c) assumes that leaf production equals leaf fall, or that changes in foliage biomass are negligble; (d,e) belowground components excluded by common measurement practice (d) or ForC variable definition (e); (f) excludes movement of dead wood into litter through breakage or size reduction; (g) measurements often limited to litter layer (OL) and may exclude larger branches and stems classified as litter and/or the more decomposed layers of the O horizon.
 **This variable is techically EFDB-relevant but not selected for submission because their is no corresponding influx variable.](figures_tables/C_variable_mapping.png)
@@
 `r fig_variable_mapping`).
-We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which met the criteria for submission, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (`r table_variables`).
+We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which had been submitted, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (`r table_variables`).
 The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% discrepancy between records we reviewed and those submitted to EFDB was due to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
-The discrepancy between the number of records sent and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
+The discrepancy between the number of records submitted and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
+Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and vaguely described disturbance histories (e.g., "moderately"/ "severely" burned).
 As of `r format(Sys.time(), '%B %d, %Y')`, the `r n_records_sent_to_EFDB` ForC records posted in EFDB represented `r percent_forest_records`% of the total EFDB records for forest land.
@@
 `r round(n_records_sent_deadwood_components/n_records_sent_to_EFDB,2)*100`% for components of dead wood (standing or fallen), 
-`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (entire O horizon or OL layer component),
+`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (entire O horizon),
 and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.

### 2024-08-02 doc/manuscript/SI_ForCv40.Rmd
describe reasons for data rejection; remove litter, delta.litter
#64

@@
 #Tables
+#Tables
 table_pools =1
@@
 
-n_records_posted_EFDB = 376 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
-Then get number from upper left box where it says "displayed records")
+n_records_posted_EFDB = 1068 # n records accepted and posted in EFDB (easy to determine by entering "ForC" in the "Other Properties" search field of EFDB: https://www.ipcc-nggip.iges.or.jp/EFDB/find_ef.php.
+Then get number from upper left box where it says "displayed records".
+Updated Jan 2024)
+percent_forest_records = round(n_records_posted_EFDB/5644*100,0); #as of Jan 2024
 

### 2024-08-02 doc/manuscript/response-to-reviews.Rmd
add reasons for data rejection
#64

@@
 
-*(NEEDS RESPONSE. We did not receive this information directly. We will have to figure out which records are missing)*
+**We have added a sentence detailing some of the reasons for rejection: "Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and vaguely described disturbance histories (e.g., "moderately"/ "severely" burned)."**
+
+**We also note that the authors of this study were not privy to the discussions of the review panel and do not have detailed information on all reasons for rejection of each record.**
 

### 2024-08-02 doc/manuscript/response-to-reviews.Rmd
work on response to reviews / minor updates

@@
 
-*(NEEDS RESPONSE)*
+**We believe that the responses to specific concerns (detailed below) help to clarify the message.
+This section is intended as a straightforward, objective assessment of the data available through ForC and submitted to EFDB, and is not intended to make a particular point.**
 
@@
 
-*(NEEDS RESPONSE)*
+**We believe that the changes to the results section and figures now convincingly demonstrate the claims in the discussion.**
+
+*(VERBOSITY CRITICISM NEEDS RESPONSE)*
 

### 2024-08-02 doc/manuscript/ForC-EFDB manuscript.Rmd
work on response to reviews / minor updates

@@
 Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and vaguely described disturbance histories (e.g., "moderately"/ "severely" burned).
-As of `r format(Sys.time(), '%B %d, %Y')`, the `r n_records_sent_to_EFDB` ForC records posted in EFDB represented `r percent_forest_records`% of the total EFDB records for forest land.
+As of `r format(Sys.time(), '%B %d, %Y')`, the `r n_records_sent_to_EFDB` ForC records posted in EFDB represented  ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC and `r percent_forest_records`% of the total EFDB records for forest land.
 

### 2024-08-05 doc/manuscript/ForC-EFDB manuscript.Rmd
add Besnard data reference

@@
 nocite: |
-  @harris_updated_2014; @jung_exploiting_2006; @besnard_mapping_2021
+  @harris_updated_2014; @jung_exploiting_2006; @besnard_mapping_2021; @besnard_global_2024
 ---
@@
 For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.
-The relative area of forests by age class was obtained from the global forest age database of Besnard *et al.* (2021), updated in 2024 (DOI: xxx).
+The relative area of forests by age class was obtained from the global forest age database of Besnard *et al.* (2021, 2024).
 ](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)

### 2024-08-06 doc/manuscript/response-to-reviews.Rmd
Worked on responses to R1/ improving results section

--- 2024-08-01 doc/manuscript/response-to-reviews.Rmd
@@
 
-*(NEEDS RESPONSE)*
+**To help clarify the message, we have responded to specific concerns (detailed below) and done some reorganization and rewording to make it less verbose.
+Our aim has been to present an objective overview of the ForC data contributed to EFDB and the potential for future submissions to improve EFDB data volume and composition.**
 
@@
 
-*(NEEDS RESPONSE. We did not receive this information directly. We will have to figure out which records are missing)*
+**We have added a sentence detailing some of the reasons for rejection: "Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and vaguely described disturbance histories (e.g., "moderately"/ "severely" burned)." We also note that the authors of this study were not privy to the discussions of the review panel and do not have detailed information on all reasons for rejection of each record.**
+
+**In addition, we have reworded and reorganized this paragraph (and the following) to make it less verbose.**
 
@@
 
-**We have made some of the suggested revisions (detailed below) to improve the assessment of data representativeness and consider this to be an important improvement.
+<!--**We have made many of the suggested revisions (detailed below) to improve the assessment of data representativeness and consider this to be an important improvement.
 At the same time, we also wish to emphasize that the goal of this paper was to document the process and provide an overview of the types of data submitted, as opposed to submitting data that adequately represents all forest types and documenting their representativeness.
 The latter -- to the extent that it would even be possible -- would be a massive project far beyond the scope of this paper.
-Therefore, while improving the assessment of data representativeness, we believe it would be overkill to analyze all aspects of representativeness.**
+Therefore, while improving the assessment of data representativeness, we believe it would be overkill to analyze all aspects of representativeness.**-->
 
@@
 
-**We have changed this word to "discrepancy".**
+**This sentence has been reworded.**
 

### 2024-08-06 doc/manuscript/ForC-EFDB manuscript.Rmd
Worked on responses to R1/ improving results section

@@
 Schematic illustrating the carbon pools defined under IPCC Guidelines for national greenhouse gas inventories; corresponding ForC variables, and relationships among them.** For each C pool, we show ForC variables corresponding to the stock, net annual increment, influx, and outflux.
-Most, but not all, EFDB-relevant ForC variables are shown here.
+Most, but not all, of the `r n_relevant_variables` EFDB-relevant ForC variables are shown here.
 Correspondence of ForC variables to IPCC criteria often depends upon measurement protocols (e.g., minimum stem diameter censused).
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, ForC (v4.0) contained `r n_independent_records_ForC` independent records (`r n_records_ForC` total), `r n_ind_records_EFDB_variables` of which were for the `r n_relevant_variables` variables relevant to EFDB (Fig.
-`r fig_variable_mapping`).
-We had reviewed or added `r n_records_reviewed_or_added` records for submission to EFDB, `r n_records_sent_to_EFDB` of which had been submitted, and `r n_records_posted_EFDB` of which had been reviewed by EFDB's review panel, accepted, and posted (`r table_variables`).
-The `r round((n_records_reviewed_or_added -n_records_sent_to_EFDB )/n_records_reviewed_or_added,2)*100`% discrepancy between records we reviewed and those submitted to EFDB was due to the presence of digitized records and records where a variable's value had been calculated as the sum or difference of related variables rather than presented directly in the text.
-The discrepancy between the number of records submitted and that posted to EFDB is because some records (`r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`%) were deemed not applicable to EFDB by the review panel.
-Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and vaguely described disturbance histories (e.g., "moderately"/ "severely" burned).
-As of `r format(Sys.time(), '%B %d, %Y')`, the `r n_records_sent_to_EFDB` ForC records posted in EFDB represented  ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC and `r percent_forest_records`% of the total EFDB records for forest land.
+As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had submitted only a fraction (~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`%) of relevant ForC records to EFDB, but these already comprised a substantial portion (`r percent_forest_records`%) of EFDB records for forest land.
+Specifically, of the `r n_records_ForC` records in ForC (v4.0), ~`r n_ind_records_EFDB_variables` were independent records of EFDB-relevant variables (Fig.
+`r fig_variable_mapping`,  `r table_variables`).
+We reviewed `r n_records_checked` records for submission to EFDB, of which
+`r round((n_records_checked -(n_records_sent_to_EFDB - n_records_added))/n_records_checked,2)*100`% were inappropriate, mostly because values were digitized from graphs or calculated from related variables rather than directly presented.
+We also added `r n_records_added` new records, mostly because EFDB required that all relevant records in a publication be submitted together. 
+Of `r n_records_sent_to_EFDB` submitted records, EFDB's review panel accepted and posted `r n_records_posted_EFDB` but deemed `r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`% not applicable.
+Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and imprecisely described disturbance histories (e.g., "moderately"/ "severely" burned).
+An iterative process of data submission review by the EFDB Technical Support Unit and review panel greatly improved but did not perfect (100% acceptance rate) our process.
 
-ForC v4.0 contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Fig.
+In terms of variables, ForC v4.0 contained records for `r n_relevant_variables_with_data` of the `r n_relevant_variables` variables (or closely-related variable groups) relevant to EFDB (Fig.
 `r fig_variable_mapping`,  `r table_variables`).
-Records were submitted for `r n_relevant_variables_records_sent` variables (or closely-related variable groups), including variables from each C pool (`r table_variables`).
-The records available in ForC v4.0 -- and the subset submitted to EFDB -- were very unevenly distributed across variables (`r table_variables`). 
+The records submitted to EFDB (`r n_relevant_variables_records_sent` variables) were very unevenly distributed across variables, mirroring the composition of ForC (`r table_variables`). 
 The majority (`r round(n_records_sent_stocks/n_records_sent_to_EFDB, 2)*100`%) of records submitted were for C stocks, including
@@
 `r round(n_records_sent_agb_components/n_records_sent_to_EFDB,2)*100`% for components of aboveground biomass (wood or foliage), 
-`r round(n_records_sent_rootbiomass/n_records_sent_to_EFDB,2)*100`% for root biomass, 
-`r round(n_records_sent_rootbiomass_components/n_records_sent_to_EFDB,2)*100`% for components of root biomass (coarse or fine roots), 
-`r round(n_records_sent_deadwood/n_records_sent_to_EFDB,2)*100`% for dead wood, 
-`r round(n_records_sent_deadwood_components/n_records_sent_to_EFDB,2)*100`% for components of dead wood (standing or fallen), 
-`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (entire O horizon),
-and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for SOM/ SOC.
-Increment records totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records submitted, all but four of which described aboveground biomass.
-The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records submitted described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, a subset of which also described inputs to the dead wood or litter pool (Fig.
+`r round((n_records_sent_rootbiomass+n_records_sent_rootbiomass_components)/n_records_sent_to_EFDB,2)*100`% for root biomass or its components, 
+`r round((n_records_sent_deadwood+n_records_sent_deadwood_components)/n_records_sent_to_EFDB,2)*100`% for dead wood or its components, 
+`r round(n_records_sent_litter/n_records_sent_to_EFDB,2)*100`% for litter (O horizon),
+and `r round(n_records_sent_soil/n_records_sent_to_EFDB,2)*100`% for soils.
+Increment records (mostly for aboveground biomass) totaled `r round(n_records_sent_increments/n_records_sent_to_EFDB*100, 0)`% of records submitted.
+The remaining `r round(n_records_sent_fluxes/n_records_sent_to_EFDB*100, 0)`% of records submitted described fluxes, all of which were either inputs or outputs to the aboveground biomass pool, with some also describing inputs to dead wood or litter pools (Fig.
 `r fig_variable_mapping`,`r table_variables`).
+Many of the EFDB-relevant variables remain poorly represented in ForC (`r table_variables`).
 
-ForC v4.0 records and the subset submitted to EFDB were distributed across all forested continents, biomes, and forest types, albeit unevenly (Figs. `r fig_map`-`r fig_histograms`).
+ForC v4.0 records -- and the subset submitted to EFDB -- were unevenly distributed across all forested continents, biomes, and forest types (Figs. `r fig_map`-`r fig_histograms`).
 Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, mirroring the composition of ForC (Figs. `r fig_map`, `r fig_histograms`b).
@@
 By far the most records and best representation relative to global area were for needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
-The largest number of records came from mature forests (>100 years), followed by young and intermediate-aged stands, but young (<20 year) stands were best represented relative to their global area (Fig.
+The largest number of submitted records came from mature forests (>100 years), but young (<20 year) stands were best-represented relative to their global area, while 20-100 year stands were under-represented (Fig.
 `r fig_histograms`d).
+ForC contains the records needed to provide more balanced geographical and forest type representation via selective data submission (Figs. `r fig_map`-`r fig_histograms`).
 

### 2024-08-07 doc/manuscript/response-to-reviews.Rmd
misc improvements in response to reviews

@@
 
-**To help clarify the message, we have responded to specific concerns (detailed below) and done some reorganization and rewording to make it less verbose.
-Our aim has been to present an objective overview of the ForC data contributed to EFDB and the potential for future submissions to improve EFDB data volume and composition.**
+**We have reorganized and reworded to clarify the message and make the section less verbose.
+We have aimed to present an objective overview of the ForC data contributed to EFDB and the potential for future submissions to improve EFDB data volume and composition.
+We have also responded to the specific concerns (detailed below).
+**
 
@@
 
-**We have added a sentence detailing some of the reasons for rejection: "Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and vaguely described disturbance histories (e.g., "moderately"/ "severely" burned)." We also note that the authors of this study were not privy to the discussions of the review panel and do not have detailed information on all reasons for rejection of each record.**
+**We have added a sentence detailing some of the reasons for rejection: "Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and vaguely described disturbance histories (e.g., "moderately"/ "severely" burned)." We also note that the authors of this study were not privy to the discussions of the review panel and do not have detailed information on all reasons for record rejection.**
 
-**In addition, we have reworded and reorganized this paragraph (and the following) to make it less verbose.**
+**In addition, we have reworded and reorganized this paragraph to make it less verbose.**
 
@@
 `r fig_map`) showing the proportional breakdown by FAO ecozone of (1) area with tree cover, (2) sites and records submitted to EFDB, (3) sites and records submitted to EFDB.
-We have also added a statement on relative representation by biome: "Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, mirroring the composition of ForC (Fig.
+We have also added a statement on relative representation by biome: "Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, reflecting the composition of ForC (Fig.
 `r fig_map`)."**
@@
 `r fig_climate`).
-We have also added a statement on climate representativity: "Submitted data covered all forested climates with substantial land area except the colder boreal regions (boreal tundra woodland; Figs.
-`r fig_map`, `r fig_climate`) and were concentrated in climates that are relatively common over Earth's land area (Fig.
-`r fig_climate`)." **
+We have also added a statement on climate representativity: "Submitted data were concentrated in climates that are relatively common over Earth's land area and covered all forested climates with substantial land area except the colder boreal regions (boreal tundra woodland; Figs.
+`r fig_map`, `r fig_climate`)." **
 
@@
 
-**To evaluate the coverage, we have modified this figure to include bars representing the relative area (% of global) of each category and added some text summarizing the results/ citing this figure (3rd paragraph in Results).**
+**To evaluate the coverage, we have modified this figure to include bars representing the relative area (% of global) of each category and added some text summarizing the results/ citing this figure (2nd paragraph in Results).**
 
@@
 
-Consequently, the “Database needs” as discussed L297 are not convincingly demonstrated by the results which are limited to just frequencies. Overall the discussion seems somewhat verbose.
+Consequently, the “Database needs” as discussed L297 are not convincingly demonstrated by the results which are limited to just frequencies.
 
 **We believe that the changes to the results section and figures now convincingly demonstrate the claims in the discussion.**
+
+Overall the discussion seems somewhat verbose.
 
@@
 
-*(Provide a bit more detail, addressing questions.)*
+**We now clarify what is meant by the 7% in the first sentence of the Results section: "...we had submitted to EFBD only a fraction (~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`%) of potentially relevant ForC records (i.e., records for EFDB-relevant variables)..."**
+
+**We have also made several changes to give more information on our process.
+For example, in the methods (section 3), we now refer readers to Methods S1 for a description of how we prioritized ForC records for review.
+We also added a description of our iterative process of submission, review, and improvement of our process: "Submissions were sent in six batches between April 2021 and June 2022, and feedback on each batch from the EFDB team was used to improve our system for subsequent submissions." (methods, section 3)...
+"The iterative process of review by EFDB's Technical Support Unit and review panel significantly improved our process, though it did not achieve a 100% acceptance rate." (results, end of 1st paragraph)**
+
+**We also revised the discussion paragraph on which this comment was focused, adding a sentence to help pu the 7% in context: "These are unevenly distributed across regions, forest types, and variables (Figs.
+`r fig_map`-`r fig_histograms`, `r table_variables`), reflecting the uneven composition of ForC [@anderson-teixeira_carbon_2021] and the fact that our effort was focused on developing an accurate and efficient data submission system rather than optimizing the composition of submitted data."**
+
+**We now provide some concrete examples of mistakes to avoid in the first paragraph of the results (see above).**
 
@@
 
-*(NEEDS RESPONSE)*
+**We have renamed this section as suggested, and have also renamed the other two to better describe the contents ("Potential for forest C databases to contribute to EFDB", "Forest C data collection and analysis needs").**
 
@@
 
-*(NEEDS RESPONSE)*
+**We have removed most of the first paragraph, as we agreed that it was distracting.
+That said, we are retaining key statements on human's influence in order to emphasize the relevance of this paper to the special issue "Anthromes and terrestrial carbon: from the deep past to net-zero".
+This was motivated by a prior request from the editors.**
 
@@
 
-*(NEEDS RESPONSE)*
+**We believe that our responses to the comments above has greatly improved this manuscript. Thank you for your feedback.**
 
@@
 
-*(NEEDS RESPONSE)*
+**We appreciate this feedback. We have made some revisions so that the discussion gives more clear and strong recommendations:**
+
+**1.
+In the section "Potential for forest C databases to contribute to EFDB", we have added a couple sentences on the practicalities of additional database work: "While submitting records from ForC or other existing databases to EFDB is by far the most efficient approach to expanding forest C data in EFDB, it does require time-consuming database work.
+As the IPCC is not able to pay for data, such work will require support from research and government institutions, grant programs, and non-governmental organizations."**
+
+**2.
+In the section "Forest C data collection and analysis needs", we have added a some details on how reporting of EFDB-relevant data might be increased: "We recommend that, when possible and relevant to their own study, researchers calculate and report all EFDB-relevant variables -- a goal that could be facilitated by development of relevant analytical tools [sensu @chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022] and encouraged by scientific journals."**
+
+**3.In the section "Recommendations of good practices for reporting forest C stocks and stock changes", we have added a suggestion on how reporting might be improved: "Improved reporting -- and potentially data submission to EFDB -- could be encouraged by journal editors, reviewers, and funding agencies."**
 

### 2024-08-07 doc/manuscript/ForC-EFDB manuscript.Rmd
re-ordered results paragraphs
misc improvements in response to reviews

@@
 <!-- (3) key results-->
-- As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_posted_EFDB` of which had been submitted and posted to EFDB.
+- As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` records potentially relevant to EFDB, `r n_records_posted_EFDB` of which had been submitted and posted to EFDB.
 These represented `r round(n_records_posted_EFDB/5644*100,0)`% of the total EFDB records for forest land.
@@
 Here, we develop and document a process for semi-automated submission of data from ForC into EFDB,  assess the data currently available through ForC, and provide recommendations for improving forest data collection, analysis, and reporting.
-ForC contained ~19,316 independent records relevant to EFDB, 1,438 of which had undergone necessary review and been submitted.
+ForC contained ~19,316 independent records potentially relevant to EFDB, 1,438 of which had undergone necessary review and been submitted.
 Records were unevenly distributed across variables and geographic regions.
@@
 
-Humans have been influencing Earth's climate via ecologically transformative impacts on ecosystems for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016].
-In recent decades, as both anthropogenic land transformation and climate change have accelerated, this relationship has come into increasing focus [@ipcc_climate_2019].
-Deforestation and forest degradation are substantial sources of the greenhouse gas carbon dioxide (CO~2~), currently accounting for >10% of anthropogenic emissions [@friedlingstein_global_2022].
-At the same time, CO~2~ uptake by remaining and regrowing forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021],
+Humans have been influencing Earth's climate via ecologically transformative impacts on ecosystems for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016],
+a relationship that has come into increasing focus in recent decades as anthropogenic land transformation and climate change have accelerated [@ipcc_climate_2019].
+<!--Deforestation and forest degradation are substantial sources of the greenhouse gas carbon dioxide (CO~2~), currently accounting for >10% of anthropogenic emissions [@friedlingstein_global_2022].
+At the same time, CO~2~ uptake by remaining and regrowing forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021]
 <!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^
 -1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests;
-BBL: might cite Ruehr et al. 2023 http://dx.doi.org/10.1038/s43017-023-00456-3 here-->
+BBL: might cite Ruehr et al. 2023 http://dx.doi.org/10.1038/s43017-023-00456-3 here->
 that offsets an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al.
-Table 2)--> - `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
-Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021].
+Table 2)-> - `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al.
+Table 2)-> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021].
 The future of this important CO~2~ sink depends primarily upon future trajectories of direct human impacts on forests [i.e., deforestation/ degradation vs.
-conservation/ restoration, @ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].
-<!--here I would definitely cite the Ruehr et al. 2023 paper-->
-
-Accordingly, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015; @grassi_key_2017].<!--As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].-->
+conservation/ restoration, @ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].-->
+Forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015; @grassi_key_2017].<!--As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].-->
 Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021]. 
@@
 (2) develop a semi-automated process for preparing ForC data for submission to EFDB;
-and (3) assess the data in ForC relevant to EFDB and records that have been submitted to date.
+and (3) assess the data in ForC potentially relevant to EFDB and records that have been submitted to date.
 We conclude with recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.
@@
 
-To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see **Data Availability Statement**).
+To submit complete, reviewed ForC records tp EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see **Data Availability Statement**).
 Criteria for data submission were that
@@
 and (4) the records had not previously been submitted to EFDB.
-Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for submission to EFDB.
+Review of potentially relevant records was prioritized as described in `r appendix_ForC_updates`.
+Once converted into EFDB format, the records were reviewed and then sent to the IPCC's Technical Support Unit for their review and submission to EFDB.
+Submissions were sent in six batches between April 2021 and June 2022, and feedback on each batch from the EFDB team was used to improve our system for subsequent submissions.
 
@@
 
-As of `r format(Sys.time(), '%B %d, %Y')`<!--inserts date upon knit-->, we had submitted only a fraction (~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`%) of relevant ForC records to EFDB, but these already comprised a substantial portion (`r percent_forest_records`%) of EFDB records for forest land.
+As of `r format(Sys.time(), '%B %Y')`<!--inserts date upon knit-->, we had submitted to EFBD only a fraction (~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`%) of potentially relevant ForC records (i.e., records for EFDB-relevant variables), but these already comprised a substantial portion (`r percent_forest_records`%) of EFDB records for forest land.
 Specifically, of the `r n_records_ForC` records in ForC (v4.0), ~`r n_ind_records_EFDB_variables` were independent records of EFDB-relevant variables (Fig.
@@
 Of `r n_records_sent_to_EFDB` submitted records, EFDB's review panel accepted and posted `r n_records_posted_EFDB` but deemed `r round((1-(n_records_posted_EFDB/n_records_sent_to_EFDB)),2)*100`% not applicable.
-Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and imprecisely described disturbance histories (e.g., "moderately"/ "severely" burned).
-An iterative process of data submission review by the EFDB Technical Support Unit and review panel greatly improved but did not perfect (100% acceptance rate) our process.
+Reasons included -- but were not limited to -- non-applicability to the IPCC methodology of the variable submitted (e.g., net ecosystem CO~2~ exchange, litter - OL layer; subsequently excluded from lists of relevant variables and counts of potentially relevant records), inadequate information on the quality of data and thus on its replicability (e.g., confidence interval/uncertainty), and imprecisely described disturbance histories (e.g., "moderately"/ "severely" burned).
+The iterative process of review by EFDB's Technical Support Unit and review panel significantly improved our process, though it did not achieve a 100% acceptance rate.
+
+ForC v4.0 records, including the subset submitted to EFDB, were unevenly distributed across all forested continents, biomes, and forest types (Figs. `r fig_map`-`r fig_histograms`).
+Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, reflecting the composition of ForC (Figs. `r fig_map`, `r fig_histograms`b).
+Submitted data were concentrated in climates that are relatively common over Earth's land area and covered all forested climates with substantial land area except the colder boreal regions (boreal tundra woodland; Figs.
+`r fig_map`, `r fig_climate`).
+Among the records submitted to EFDB, the largest number came from North America, followed by Asia, South America, and Africa, with strongly disproportionate representation of North America and under-representation of South America, Africa, Europe, Australia, and Oceania (Fig.
+`r fig_histograms`c).
+In terms of FAO ecozones, boreal coniferous forest, temperate continental forest, and temperate mountain systems had the most records and were the best represented relative to their global area, whereas other ecozones were poorly represented (Fig.
+`r fig_histograms`b).
+By far the most records and best representation relative to global area were for needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
+The largest number of submitted records came from mature forests (>100 years), but young (<20 year) stands were best represented relative to their global area, while 20-100 year stands were under-represented (Fig.
+`r fig_histograms`d).
+ForC contained the records needed to provide more balanced geographical and forest type representation via selective data submission (Figs. `r fig_map`-`r fig_histograms`).
+
+<!--In ForC v4.0, the largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig.
+`r fig_histograms`c).
+Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
+`r fig_histograms`b).
+Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
+The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
+In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).-->
+
+![**Figure `r fig_map`.
+Map of sites in ForC shaded by number of independent records potentially relevant (circles) and submitted (triangles) to EFDB.**  Underlying map shows FAO ecozones, and symbols are colored according to the number of records at each site.
+Inset stacked bar chart shows proportional representation by FAO ecozone for tree cover area (from SYNMAP, Jung *et al*, 2006), potentially relevant ForC records, and submitted records.
+FAO ecozones (obtained from FAO’s GEOnetwork: http://www.fao.org:80/geonetwork) are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
+
+![**Figure `r fig_climate`.
+Distribution of ForC records potentially relevant (dots) and submitted (triangles) to EFDB within the global climate space of mean annual temperature (MAT) and mean annual precipitation (MAP).** Climate data are from the CRU TS3.10 dataset (0.5° resolution, 1990-2014, Harris *et al*, 2014).
+Background colors indicate the global land area with each MAT-MAP combination.](figures_tables/Climate_of_forC_sites_IPCC.png)
+
+![**Figure `r fig_histograms`.
+Histograms of number of independent records in ForC potentially relevant (grey) and submitted (black) to EFDB (both left axis), along with the relative global area of each type (red, right axis), organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.** For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
+Global coverage of each was obtained from SYNMAP (Jung *et al*, 2006).
+For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.
+The relative area of forests by age class was obtained from the global forest age database of Besnard *et al.* (2021, 2024).
+](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)
 
@@
 
-ForC v4.0 records -- and the subset submitted to EFDB -- were unevenly distributed across all forested continents, biomes, and forest types (Figs. `r fig_map`-`r fig_histograms`).
-Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, mirroring the composition of ForC (Figs. `r fig_map`, `r fig_histograms`b).
-Submitted data covered all forested climates with substantial land area except the colder boreal regions (boreal tundra woodland; Figs.
-`r fig_map`, `r fig_climate`) and were concentrated in climates that are relatively common over Earth's land area (Fig.
-`r fig_climate`).
-Among the records submitted to EFDB, the largest number came from North America, followed by Asia, South America, and Africa, with strongly disproportionate representation of North America and under-representation of South America, Africa, Europe, Australia, and Oceania (Fig.
-`r fig_histograms`c).
-In terms of FAO ecozones, boreal coniferous forest, temperate continental forest, and temperate mountain systems had the most records and were the best represented relative to their global area, whereas other ecozones were poorly represented relative to their global area (Fig.
-`r fig_histograms`b).
-By far the most records and best representation relative to global area were for needleleaf evergreen forests, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`b). 
-The largest number of submitted records came from mature forests (>100 years), but young (<20 year) stands were best-represented relative to their global area, while 20-100 year stands were under-represented (Fig.
-`r fig_histograms`d).
-ForC contains the records needed to provide more balanced geographical and forest type representation via selective data submission (Figs. `r fig_map`-`r fig_histograms`).
-
-<!--In ForC v4.0, the largest number of records came from Asia, followed by North America, South America, and Europe, with relatively few records from Africa, Australia, and Oceania (Fig.
-`r fig_histograms`c).
-Categorized by FAO ecozone, the greatest numbers of records came from subtropical humid forests, temperate mountain systems, and tropical rain forests, each with >2,000 independent records (Fig.
-`r fig_histograms`b).
-Boreal coniferous forests, temperate continental forests, subtropical mountain systems, and tropical moist deciduous forests had >1,000 independent records each, while other ecozones all had <1,000 records.
-The most widely represented forest type was needleleaf evergreen, followed by broadleaf deciduous and broadleaf evergreen (Fig. `r fig_histograms`a).
-In terms of stand age, the most represented age class was 20-100 years, followed by <20 years and then >100 years (Fig. `r fig_histograms`d).-->
-
-![**Figure `r fig_map`.
-Map of sites in ForC shaded by number of independent records relevant (circles) and submitted (triangles) to EFDB.**  Underlying map shows FAO ecozones, and symbols are colored according to the number of records at each site.
-Inset stacked bar chart shows proportional representation by FAO ecozone for tree cover area (from SYNMAP, Jung *et al*, 2006), relevant ForC records, and submitted records.
-FAO ecozones (obtained from FAO’s GEOnetwork: http://www.fao.org:80/geonetwork) are coded as follows: Ba-Boreal coniferous forest, Bb-Boreal tundra woodland, BM-Boreal mountain systems, P-Polar, SBSh-Subtropical steppe, SBWh-Subtropical desert, SCf-Subtropical humid forest, SCs-Subtropical dry forest, SM-Subtropical mountain systems, TAr-Tropical rain forest,  TAwa-Tropical moist deciduous forest, TAwb-Tropical dry forest, TBSh-Tropical shrubland, TBWh-Tropical desert, TeBSk-Temperate steppe, TeBWk-Temperate desert, TeDc-Temperate continental forest, TeDo-Temperate oceanic forest, TeM-Temperate mountain systems, TM-Tropical mountain systems.](figures_tables/World_Map_of_sites_with_FAO_and_IPCC_data_sent.png)
-
-![**Figure `r fig_climate`.
-Distribution of ForC records relevant (dots) and submitted (triangles) to EFDB within the global climate space of mean annual temperature (MAT) and mean annual precipitation (MAP).** Climate data are from the CRU TS3.10 dataset (0.5° resolution, 1990-2014, Harris *et al*, 2014).
-Background colors indicate the global land area with each MAT-MAP combination.](figures_tables/Climate_of_forC_sites_IPCC.png)
-
-![**Figure `r fig_histograms`.
-Histograms of number of independent records in ForC relevant (grey) and submitted (black) to EFDB (both left axis), along with the relative global area of each type (red, right axis), organized by (a) dominant vegetation type, (b) FAO ecozone, (c) continent, and (d) stand age.** For dominant vegetation (a), 'Other' includes deciduous needleleaf, mixed broadleaf- needleleaf, non-woody vegetation (e.g., early successional), and incompletely classified or mixed forest types.
-Global coverage of each was obtained from SYNMAP (Jung *et al*, 2006).
-For FAO ecozones (b), codes are as listed in the caption of Figure `r fig_map`.
-The relative area of forests by age class was obtained from the global forest age database of Besnard *et al.* (2021, 2024).
-](figures_tables/Histogram_n_Relevant_and_Transferred_Records.png)
-
 # Discussion {-}
 
-Here we have developed a framework for submitting records from ForC v4.0, the largest compilation of ground-based forest C estimates, to the EFDB, thus making those data more accessible for reporting CO~2~ emissions and removals from forest land consistent with good practice in the IPCC guidelines  [@ipcc_2006_2006;@ipcc_2019_2019].
-As of `r format(Sys.time(), '%B %d, %Y')`, `r n_records_posted_EFDB` ForC records have been posted to EFDB, which represents just `r round(n_records_posted_EFDB/n_ind_records_EFDB_variables*100,0)`% of potentially relevant records in ForC v4.0, but currently comprises `r percent_forest_records`% of the total EFDB records for forest land.
-The records present in ForC and submitted to EFDB are very unevenly distributed across variables, regions, and forest types (Figs.
-`r fig_map`-`r fig_histograms`, `r table_variables`), reflecting broader patterns in allocation of research effort and pointing to scientific research and reporting needs.
+We developed a framework for submitting data from ForC v4.0, the largest compilation of ground-based forest C estimates, to the EFDB, making the data more accessible for reporting CO~2~ emissions and removals from forest land in line with IPCC guidelines [@ipcc_2006_2006;@ipcc_2019_2019].
+Although the process of preparing ForC data for EFDB submission requires manual review of records, most of the pertinent information is already entered in ForC, and complete records can be automatically converted to EFDB's format using the system developed here.
+As of `r format(Sys.time(), '%B %Y')`, `r n_records_posted_EFDB` ForC records have been posted to EFDB, which<!--represents just `r round(n_records_posted_EFDB/n_ind_records_EFDB_variables*100,0)`% of potentially relevant records in ForC v4.0, but currently--> comprises `r percent_forest_records`% of the total EFDB records for forest land.
 
-Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work, new data collection and analysis, and reporting.
+Based on our experience contributing forest C data to EFDB via ForC, we make several recommendations as to how scientists can improve forest C records in EFDB through database work, new data collection and analysis, and improved reporting.
 
-### Database needs {-}
+### Potential for forest C databases to contribute to EFDB {-}
 
-There is vast potential to expand forest C data in EFDB by completing the process of reviewing and submitting data that are already in ForC (Figs. `r fig_map`-`r fig_histograms`).
-So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the EFDB-relevant data in ForC have been submitted to EFDB.
-Although this process requires manual review of records, the submission of new records to EFDB is greatly facilitated by the fact that most pertinent information for each record is already entered in ForC and can be easily prepared for submission to EFDB using the system developed here.
-Future efforts to review studies for submission should optimize for representation across geographic regions, forest types, and variables, giving priority to those from currently under-represented regions and forest types (Figs.
+There is vast potential to expand forest C data in EFDB by reviewing and submitting records from ForC (Figs. `r fig_map`-`r fig_histograms`).
+So far, only ~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`% of the potentially relevant records in ForC have been submitted to EFDB.
+These are unevenly distributed across regions, forest types, and variables (Figs.
+`r fig_map`-`r fig_histograms`, `r table_variables`), reflecting the uneven composition of ForC [@anderson-teixeira_carbon_2021] and the fact that our effort was focused on developing an accurate and efficient data submission system rather than optimizing the composition of submitted data.
+Future efforts to submit ForC records to EFDB should optimize for representation across geographic regions, forest types, and variables, giving priority to those that are currently under-represented (Figs.
 `r fig_map`-`r fig_histograms`, `r table_variables`).
-Other categories of records to prioritize include those from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology), the variables most needed by EFDB users, and the most up-to-date records.
+Other categories of records to prioritize include those from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology)<!--, the variables most needed by EFDB users, --> and more recent records.
 
-In addition to the large potential to expand EFDB using records already in ForC, there are extensive EFDB-relevant forest C data that are not currently included in ForC, with more being published on a nearly daily basis.
-Coverage of particular variables or regions could be vastly improved through systematic review of the literature, although this requires focused and extensive manual effort.
+In addition to the large potential to expand EFDB using records already in ForC, there is an every-growing collection of EFDB-relevant forest C data not currently included in ForC. 
+ForC's coverage of particular variables or regions could be vastly improved through systematic review of the literature, although this requires focused and extensive manual effort.
 Recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
-Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
+These could be submitted to EFDB via ForC (as they have compatible structures) or directly using a process parallel to that described here.
+<!--Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
+-->
 
-### Data collection and analysis needs {-}
+While submitting records to EFDB from ForC or other databases containing most of the required information is by far the most efficient approach to expanding forest C data in EFDB, it does require time-consuming database work (e.g., review of records based on original publications, adding missing information).
+As the IPCC is not able to pay for data, such work will require support from research and government institutions, funding agencies, and non-governmental organizations.
+
+### Forest C data collection and analysis needs {-}
 
@@
 However, while equations for estimating root (and thereby total) biomass require improvement, they do exist for many forest types [e.g., @brassard_coarse_2011; @chojnacky_updated_2014;  @waring_overlooking_2017; @mokany_critical_2006].
-We recommend that, when possible, researchers calculate and report all EFDB-relevant variables.
+We recommend that, when possible and relevant to their own study, researchers calculate and report all EFDB-relevant variables -- a goal that could be facilitated by development of relevant analytical tools [*sensu* @chave_improved_2014;@rejou-mechain_biomass_2017;@gonzalez-akre_allodb_2022] and encouraged by scientific journals.
 
-### Data reporting needs {-}
+### Recommendations of good practices for reporting forest C stocks and stock changes {-}
 
 We recommend that, in order to make research most useful for estimating C stock changes following IPCC guidelines, researchers calculate and report results according to IPCC good practice (Table `r table_recommendations`).
-Importantly, simple decisions on the presentation of results will determine whether the records meet the current criteria for inclusion in EFDB.
+Importantly, simple decisions on the presentation of results will determine whether the records meet criteria for inclusion in EFDB.
 For example:
@@
 (2) direct presentation of all relevant variables allows inclusion, whereas presenting only components of variables of interest (e.g., parsing litter into fine woody debris, OL, OF, and OH layers) or requiring simple mathematical operations to obtain a variable of interest (e.g., *delta.agb* = *ANPP_woody* - *woody.mortality.agb*) disqualifies records from inclusion;
-(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant estimating forest C stock changes according to IPCC guidelines (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
+(3) matching IPCC-defined thresholds for defining C pools (Table `r table_pools`), which may vary by country, can make the data far more relevant for estimating forest C stock changes according to IPCC guidelines (e.g., using a 10 cm cutoff between dead wood and litter, presenting soil C to a depth of 30 cm).
 It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
+Improved reporting -- and potentially data submission to EFDB -- could be encouraged by journal editors, reviewers, and funding agencies.
 
@@
 
-For those compiling published records (e.g., for meta-analyses), the data set can have added value if all information required by EFDB is extracted from original publications.
+Researchers compiling published records (e.g., for meta-analyses) can increase the value of their data set by ensuring that all information required by EFDB is extracted from original publications.
 <!--SCP - optional addition : Realistic  planning and reporting is critically needed to ensure that forest-based climate change mitigation initiatives are effective, and this requires solid scientific data -->

### 2024-08-07 doc/manuscript/SI_ForCv40.Rmd
misc improvements in response to reviews

@@
 We prioritized review of
-(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021],
+(1) records from the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021], which are familiar to our team and generally high quality,
 (2) studies with confidence intervals recorded in ForC (because uncertainty estimates are important to the IPCC),

### 2024-08-07 doc/manuscript/ForC-EFDB manuscript.Rmd
revised discussion

@@
 Other categories of records to prioritize include those from countries relying on existing data for their greenhouse gas inventories (Tier 1 or 2 methodology)<!--, the variables most needed by EFDB users, --> and more recent records.
-
-In addition to the large potential to expand EFDB using records already in ForC, there is an every-growing collection of EFDB-relevant forest C data not currently included in ForC. 
-ForC's coverage of particular variables or regions could be vastly improved through systematic review of the literature, although this requires focused and extensive manual effort.
-Recent efforts have compiled large databases of relevant data from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
-These could be submitted to EFDB via ForC (as they have compatible structures) or directly using a process parallel to that described here.
+Coverage of many variables or regions could be vastly improved through systematic review of the literature, although this requires focused and extensive manual effort.
+Recent efforts have compiled large databases of relevant data, compatible with ForC, from monoculture plantation forests [@bukoski_rates_2022] and mixed species plantation forests [@feng_multispecies_2022;@warner_higher_2022], and such a compilation is in works for agroforestry (Susan Cook-Patton, unpublished data).
 <!--Beyond expanding collections of relevant forest C records, such reviews are valuable for assessing the availability of published records and identifying variables and regions that require additional data collection and analysis.
@@
 
-While submitting records to EFDB from ForC or other databases containing most of the required information is by far the most efficient approach to expanding forest C data in EFDB, it does require time-consuming database work (e.g., review of records based on original publications, adding missing information).
+While submitting records to EFDB from ForC or other databases is by far the most efficient approach to expanding forest C data in EFDB, it does require time-consuming database work (e.g., review of records based on original publications, adding missing information).
 As the IPCC is not able to pay for data, such work will require support from research and government institutions, funding agencies, and non-governmental organizations.
@@
 
-New data collection and analyses are needed to fill notable knowledge gaps.
-While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [`r table_variables`,@nisar_nasaisro_2018; @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
-Other pools and variables remain poorly quantified and highly uncertain for many parts of the world [`r table_variables`,@tifafi_large_2018; @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
-Furthermore, data distribution is very uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
-For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
-More generally, belowground C measurements remain sparse globally.
+While further database work could vastly improve EFDB's coverage of forest C data, new data collection and analyses will be essential to providing accurate global data on forest C stocks and changes therein.
+Even for for aboveground biomass stocks, which have received by far the most research attention [`r table_variables`,@nisar_nasaisro_2018; @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023].
+Other variables -- particularly those characterizing belowground C pools -- remain poorly quantified and highly uncertain for many parts of the world [`r table_variables`,@tifafi_large_2018; @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
+Furthermore, data distribution is very uneven across forest types and geographical regions (Figs.
+`r fig_map`-`r fig_histograms`); for instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
 Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labrière_forest_2023].
@@
 It should also be emphasized that reporting of 95% confidence intervals (or other metrics of error), when applicable, is highly desirable and makes the data more relevant to IPCC.
+
+Researchers compiling published records (e.g., for meta-analyses) can increase the value of their data set by ensuring that all information required by EFDB is extracted from original publications.
+<!--SCP - optional addition : Realistic  planning and reporting is critically needed to ensure that forest-based climate change mitigation initiatives are effective, and this requires solid scientific data -->
+This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
 Improved reporting -- and potentially data submission to EFDB -- could be encouraged by journal editors, reviewers, and funding agencies.
@@
 ```
-
-Researchers compiling published records (e.g., for meta-analyses) can increase the value of their data set by ensuring that all information required by EFDB is extracted from original publications.
-<!--SCP - optional addition : Realistic  planning and reporting is critically needed to ensure that forest-based climate change mitigation initiatives are effective, and this requires solid scientific data -->
-This includes -- but is not limited to -- retaining original values as presented without modification or rounding, noting whether data were digitized, recording confidence intervals, and recording all required fields (as indicated in the EFDB's bulk import template).
-The significant effort required to map a database into EFDB has been accomplished here (`r table_ForCfieldmapping`), and we hope that it will prove useful as a model for other efforts.
 

### 2024-08-07 doc/manuscript/response-to-reviews.Rmd
revised discussion
fix response to reviews

@@
 We have addressed all points, as detailed below.
+The most important changes have been:
+
+1.
+In order to give a better picture of the representativeness of the submitted data (requested by reviewer 1), we revised two of the figures (global map, histogram of submitted data) and added one new figure (a Whittaker-style diagram showing ForC sites and those with submitted data within the MAT-MAP climate space (new Fig.
+`r fig_climate`).
+
+2. We have given some more specifics about the process (requested by reviewer 1), as well as specific recommendations (requested by reviewer 2).
+
+3. We have revised the text to make it clearer and less verbose (requested by reviewer 1).
+
+We believe these changes have signficantly improved the manuscript. Thank you for your consideration, and we look forward to receiving your decision.
 
@@
 
-*(VERBOSITY CRITICISM NEEDS RESPONSE)*
+**We have reviewed and revised the discussion, removing extra verbosity where possible.**
 
@@
 
-**We now clarify what is meant by the 7% in the first sentence of the Results section: "...we had submitted to EFBD only a fraction (~`r round(n_records_sent_to_EFDB/n_ind_records_EFDB_variables,2)*100`%) of potentially relevant ForC records (i.e., records for EFDB-relevant variables)..."**
+**We now clarify what is meant by the 7% in the first sentence of the Results section: "...we had submitted to EFBD only a fraction (~7%) of potentially relevant ForC records (i.e., records for EFDB-relevant variables)..."**
 

### 2024-08-08 scripts/1_ForC_to_IPCC_formatting.R
update histogram figure #63
@teixeirak let me know what you think

@@ -81,7 +81,7 @@ ForC_simplified <- ForC_simplified[ForC_simplified$variable.name %in% V_mapping$
 ## x axis is dominant vegetation type
 load("scripts/X_area_of_forest_categories_for_barplot.RData")
 
-ForBarplot <- data.frame(data_type = rep(c("relevant", "submitted"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg), FAO.ecozone = c(ForC_simplified$FAO.ecozone, ForC_simplified_already_sent$FAO.ecozone), continent = c(ForC_simplified$continent, ForC_simplified_already_sent$continent), stand.age = c(ForC_simplified$stand.age, ForC_simplified_already_sent$stand.age))
+ForBarplot <- data.frame(data_type = rep(c("potentially relevant", "submitted"),  times= c(nrow(ForC_simplified), nrow(ForC_simplified_already_sent))), dominant.veg = c(ForC_simplified$dominant.veg, ForC_simplified_already_sent$dominant.veg), FAO.ecozone = c(ForC_simplified$FAO.ecozone, ForC_simplified_already_sent$FAO.ecozone), continent = c(ForC_simplified$continent, ForC_simplified_already_sent$continent), stand.age = c(ForC_simplified$stand.age, ForC_simplified_already_sent$stand.age))
 
 ForBarplot$stand.age <- addNA(cut(as.numeric(ForBarplot$stand.age), breaks = c(0, 20, 100, 10000), labels = c("<20 yrs", "20-100 yrs", ">100 yrs" )))
 levels(ForBarplot$stand.age)[4] <- "Unclassified"
@@ -93,10 +93,10 @@ ForBarplot$FAO.ecozone[is.na(ForBarplot$FAO.ecozone)] <- "Unclassified"
 
 # ForBarplot$continent <- gsub(" ", "\n", ForBarplot$continent)
 
-labels <- list(A = list("dominant.veg", "Dominant vegetation", F, dom_veg_areas),
-               B = list("FAO.ecozone", "FAO ecozone", T, FAO_ecozone_areas),
-               C = list("continent", "Continent", F, continents_areas),
-               D = list("stand.age", "Stand age", F, stand_age_area))
+labels <- list(A = list("dominant.veg", "(a) Dominant vegetation", F, dom_veg_areas),
[…]

### 2024-08-08 scripts/1_ForC_to_IPCC_formatting.R
fix stand age order #63
@teixeirak, FYI

@@ -145,7 +145,7 @@ for(i in 1:length(labels)){
 }
 
 all_data <- all_data %>% mutate(panel = factor(panel, levels = c("(a) Dominant vegetation", "(b) FAO ecozone", "(c) Continent", "(d) Stand age")),
-                                name = factor(name, levels= (c(all_data %>% filter(!panel %in% "Stand age", !name %in%  c("Other", "Unclassified")) %>% arrange(panel, -value) %>% pull(name) %>% unique(), all_data %>% filter(panel %in% "Stand age", !name %in%  c("Other", "Unclassified")) %>% pull(name) %>% unique(), c("Other", "Unclassified")) %>% unique())),
+                                name = factor(name, levels= (c(all_data %>% filter(!panel %in% "(d) Stand age", !name %in%  c("Other", "Unclassified")) %>% arrange(panel, -value) %>% pull(name) %>% unique(), all_data %>% filter(panel %in% "(d) Stand age", !name %in%  c("Other", "Unclassified")) %>% pull(name) %>% unique(), c("Other", "Unclassified")) %>% unique())),
                                 group = factor(ifelse(what %in% "global area", what, "data"), levels = c("global area", "data")),
                                 what = factor(what, levels = c("potentially relevant", "submitted","global area")))
 

### 2024-08-19 doc/manuscript/ForC-EFDB manuscript.Rmd
final check, with small changes

@@
 These included  organic matter or C stocks, net annual increments, influxes (a.k.a.
-"gross annual increments" by IPCC), and outfluxes for each IPCC-defied C pool (Table `r table_pools`, Fig.
+"gross annual increments" by IPCC), and outfluxes for each IPCC-defined C pool (Table `r table_pools`, Fig.
 `r fig_variable_mapping`).
@@
 As human society strives to achieve net-zero greenhouse gas emissions, forest researchers can make their research more useful for forest C inventories under IPCC guidelines by calculating and reporting results in ways that are consistent with methodologies provided in the IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
-In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest CO~2~ inventories for forest lands under the Paris Agreement.
-This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may be increasingly less accurate.
+In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of C inventories for forest lands under the Paris Agreement.
+This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], making data collected a decade or more in the past increasingly unreliable.
 This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.

### 2024-08-20 doc/manuscript/ForC-EFDB manuscript.Rmd
fixed typo

@@
 
-To submit complete, reviewed ForC records tp EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see **Data Availability Statement**).
+To submit complete, reviewed ForC records to EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see **Data Availability Statement**).
 Criteria for data submission were that

### 2024-09-03 doc/manuscript/ForC-EFDB manuscript.Rmd
update DOI for ForC v4.0
#54

@@
 All code and data are openly available.
-The ForC database and associated code are available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) is archived in Zenodo [@anderson-teixeira_forcdb_2023, DOI: 10.5281/zenodo.8020861].
+The ForC database and associated code are available via the ForC repository within the ForC-db organization on GitHub (https://github.com/forc-db/ForC), and the version used here (ForC v4.0) is archived in Zenodo [@anderson-teixeira_forcdb_2023, DOI: 10.5281/zenodo.13647060].
 The data and code associated with data submission to EFDB and preparation of this manuscript are available via the the IPCC-EFDB-integration repository within the ForC-db organization on GitHub  (https://github.com/forc-db/IPCC-EFDB-integration) and archived in Zenodo (DOI: 10.5281/zenodo.8021474).

### 2024-09-03 doc/manuscript/2024-08-19 submission/response-to-reviews.Rmd
folder organization

--- 2024-08-07 doc/manuscript/response-to-reviews.Rmd
@@
 1.
-In order to give a better picture of the representativeness of the submitted data (requested by reviewer 1), we revised two of the figures (global map, histogram of submitted data) and added one new figure (a Whittaker-style diagram showing ForC sites and those with submitted data within the MAT-MAP climate space (new Fig.
+In order to give a better picture of the representativeness of the submitted data (requested by reviewer 1), we revised two of the figures (global map, histogram of submitted data) and added one new figure (a Whittaker-style diagram showing ForC sites and those with submitted data within the MAT-MAP climate space; new Fig.
 `r fig_climate`).
@@
 
-We believe these changes have signficantly improved the manuscript. Thank you for your consideration, and we look forward to receiving your decision.
+We believe these changes have significantly improved the manuscript. Thank you for your consideration, and we look forward to receiving your decision.
 
@@
 **We have added a graph to the map (Fig.
-`r fig_map`) showing the proportional breakdown by FAO ecozone of (1) area with tree cover, (2) sites and records submitted to EFDB, (3) sites and records submitted to EFDB.
+`r fig_map`) showing the proportional breakdown by FAO ecozone of (1) area with tree cover, (2) sites and records potentially relevant to EFDB, (3) sites and records submitted to EFDB.
 We have also added a statement on relative representation by biome: "Relative to forested area, submitted data under-represented the tropics and over-represented temperate regions, reflecting the composition of ForC (Fig.
@@
 Fig.
-3 *(now 4)* isn’t conclusive either because ecozones have different sizes, so the number of reports on itself is not a sufficient to evaluate the coverage.
+3 **(now 4)** isn’t conclusive either because ecozones have different sizes, so the number of reports on itself is not a sufficient to evaluate the coverage.
 The representativity of the data within an ecozone depends on the variability between records: if they all cover the same age classes, again, the representativity is bad.
@@
 **These scripts are publicly available in the associated GitHub repository (https://github.com/forc-db/IPCC-EFDB-integration) and also archived in Zenodo.
-This is explained in our data availability statement, but that fact is easy to miss.
-To ensure that this is clear to readers, we have modified this statement to read: "To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see Data Availability Statement).**
+This is explained in our data availability statement, but that fact was easy to miss.
+To ensure that this is clear to readers, we have modified this statement to read: "To submit complete, reviewed ForC records into EFDB, we created R scripts to restructure ForC records and populate EFDB's bulk import form (publicly available; see Data Availability Statement)."**
 
